### 2023-05-30 article/article.Rmd
agregar article rticles

---
title: Comprehensive assessment of the climate-induced water scarcity over continental Chile
author:
  - name: Francisco Zambrano
    email: francisco.zambrano@umayor.com
    affiliation: Hemere Centro de Observación de la Tierra,Universidad Mayor
    correspondingauthor: true
    footnote: 1
  - name: Bob Security
    email: bob@example.com
    affiliation: Another University
  - name: Cat Memes
    email: cat@example.com
    affiliation: Another University
    footnote: 2
  - name: Derek Zoolander
    email: derek@example.com
    affiliation: Some Institute of Technology
    footnote: 2
address:
  - code: Some Institute of Technology
    organization: Big Wig University
    addressline: 1 main street
    city: Gotham
    state: State
    postcode: 123456
    country: United States
  - code: Another University
    organization: Department
    addressline: A street 29
    postcode: 2054 NX
    city: Manchester,
    country: The Netherlands
footnote:
  - code: 1
    text: "This is the first author footnote."
  - code: 2
    text: "Another author footnote."
abstract: |
  This is the abstract.

  It consists of two paragraphs.
keywords: 
  - keyword1
  - keyword2
journal: "An awesome journal"
date: "`r Sys.Date()`"
classoption: preprint, 3p
bibliography: references.bib
biblio-style: "apalike"
linenumbers: false
numbersections: true
# Use a CSL with `citation_package = "default"`
# csl: https://www.zotero.org/styles/elsevier-harvard
output: 
  rticles::elsevier_article:
    keep_tex: true
    citation_package: natbib
---

```{r setup,include=FALSE}
library(gt)
library(dplyr)
library(tidyr)
library(purrr)
library(kableExtra)
options(gt.html_tag_check = FALSE)
```

# Version

This Rmd-skeleton uses the mdpi Latex template published 2019/02. 
However, the official template gets more frequently updated than the 'rticles'
package. Therefore, please make sure prior to paper submission, that you're 
using the most recent .cls, .tex and .bst files 
(available [here](http://www.mdpi.com/authors/latex)).

# Introduction

In 2021, the sixth assessment report (AR6) from the working group I of the IPCC was released [@IPCC2021]. Chapter 11 [@IPCCCH112021] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes. The evidence has been strengthened since AR5 [@IPCC2013]. There is high confidence that the increasing global warning can expand the land area affected by increasing drought frequency and severity [@IPCCCH112021]. Chile has been facing a persistent rainfall deficit lasting for more than ten years [@Garreaud2017] which has impacted the hydrological system [@Boisier2018], and consequently the vegetation development [@Zambrano2020].

Precipitation is the primary driver of drought that impacts hydrological regimes and vegetation productivity. Thus, it is commonly classified as meteorological, hydrological, and agricultural [@Wilhite1985]. Lately, it has been argued that this definition does not fully address the ecological dimensions [@Crausbay2017]. @Crausbay2017 proposed the ecological drought definition as "an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems". The AR6 [@IPCC2021] state that even if global warming is stabilized at 1.5°-2°C many parts of the world will be impacted by more severe agricultural and ecological drought. Central Chile has suffered from crop productivity failure, highlighting the growing season 2007-2008 and 2008-2009 [@Zambrano2016;@Zambrano2018], which impacted an extensive surface. But, in 2019-2020, the drought intensity reached an extreme condition at North 34°S  not seen -at least- for more than 40 years [@Zambrano2020], affecting forest, grassland, and croplands areas. The prolonged lack of precipitation within Central Chile is producing changes in the ecosystem that should study.

Satellite remote sensing [@West2019;@AghaKouchak2015] is the primary method to evaluate how meteorological drought impacts vegetation dynamics. Since the 90's multiple vegetation drought indices have been derived (VCI,[@Kogan1990]; TCI, [@Kogan1995];zNDVI, [@Peters2002]; VegDri, [@Brown2008]) that have allowed making spatiotemporal analysis. Although we can calculate those indices for any time during the year (depending on satellite revisit), there are relevant during the stage vegetation has more activity, the growing season [@Mishra2015].  Although modeling phenology is a complex task, satellites offer strategies that help to address it [@Younes2021; @Vrieling2018;@Cai2017]. Also, the land cover dynamics product MCD12Q2 from the USGS [@Friedl2019] provides some phenology metrics. Some authors have proposed indices aggregated during the season. @Meroni2017 accumulating the fractional active photosynthetic active radiation(FAPAR) between the start (SOS) and the end of the season (EOS) in the Sahel, calculate the zCFAPAR. @Zambrano2018 used the same approach but with the NDVI (Normalized Difference Vegetation Index), derivating the zcNDVI within Central Chile. Besides, land use land cover  (LULC) change can be driven by drought [@Tran2019;@Akinyemi2021]. To analyze those changes, multiple time-series LULC products exist as the MCD12Q1 [@Friedl2019] and the ESA CCI-LC [@ESA2017]. For Chile, 2014 was made a high-resolution land cover at 30m of spatial resolution [@Zhao2016]. The LULC product with the vegetation drought index can help evaluate the impact of drought on the ecosystem.

Vegetation drought indices are proxies of productivity [@Paruelo2016;@Schucknecht2017]. The main environmental variables that affect it are water supply and demand [@Mishra2015].  We measure them by precipitation and evapotranspiration (ET), commonly collected from weather stations. Usually, in developing countries (i.e., Chile), incomplete records or gaps present a challenge. But, there are satellite estimates of these variables. To evaluate drought, the World Meteorological Organization (WMO; [@WMO2012]) has proposed the Standardized Precipitation Index (SPI; [@McKee1993]), a multiscalar drought index, which has been used worldwide. For Chile, @Zambrano2017 derived and evaluated it from the product of the Climate Hazards Group InfraRed Precipitation with Station data (CHIRPS; [@Funk2015]). For water demand, it is used ET. The vegetation biomass productivity is strongly related to ET [@FAO66]. The atmospheric evaporative demand (AED) represents the maximum ET rate from a land surface (without water restriction), also known as reference ET. The recommended method for its calculation is the FAO Penman-Monteith [@Pereira2015;@Allen2005]. Due to climate change, AED is increasing, driving ET rise [@IPCCCH112021]. But, it is not always true [@Milly2016]. For example, regions where AET is highest have the lowest ET. The MOD16 product [@Running2021;@Mu2011] provides AET and ET satellite estimates and has been used to derive drought indices [@Mu2013]. Soil moisture (SM) is an Essential Climate Variable (ECV) that modulates vegetative growth. The climate change initiative (CCI) from the European Space Agency (ESA) delivers the ESA CCI SM product [@Dorigo2017] (current version 6.1), which has been helpful to monitor drought [@Zhang2019]. Besides, total water storage can be retrieved by the Gravity Recovery and Climate Experiment (GRACE), which allows analyzing water availability changes [@Ahmed2014;@Ma2017]. The water demand and supply by remote sensing can help evaluate how they have impacted vegetation productivity.

The study aims to analyze the drought impact on vegetation through Central Chile for 2000-2020, using satellite data as proxies of productivity and water demand and supply. We will evaluate LULC change and use the persistent classes within 2001-2019 (> 80%) to analyze the zcNDVI index and its interconnection with precipitation deficit, AED, ET, and vegetation cover. Finally, we will investigate if the observed changes are linked to the TWS and SM.

# Study area

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{../output/figs/map_study_con_landcover.png}
\caption{ (\textbf{a}) Location of Central Chile and zones north (NCCH), central (CCCH), and south (SCCH) Central Chile. (\textbf{b}) Topography reference map. (\textbf{c}) Land cover classes for 2019. (\textbf{d}) Persistent land cover classess (> 80\%) for 2001-2019.}
\end{figure}

# Materials and Methods

## Satellite data

## zcNDVI and cluster zones

## Landcover change and persistence

## Vegetation Cover

## Evapotranspiration, Soil moisture and water storage

## Data analysis

# Results

## zcNDVI and cluster zones
<!-- \begin{figure}[H] -->
<!-- \includegraphics[width=\textwidth]{../figs/map_zcNDVI_ts_clusters.png} -->
<!-- \caption{ } -->
<!-- \end{figure} -->

## Landcover Change and persistent

```{r,echo=FALSE,fig.width =10,fig.pos='H'}
dataLCV_ts <- readRDS('../data/processed_data/timeseries_landcover_zone_LCclass_2001-2021.rds')  
dataLCV_trend <- readRDS('../data/processed_data/trends_landcover_2001-2021.rds')

dataSpark <- c('Shrubland', 'Savanna', 'Grassland', 'Barren land','Forest','Cropland') %>% 
  map(function(class){
    a <- dataLCV_ts |> 
      complete(LC_type) |> 
      ungroup() |> 
      filter(LC_type == class) |>   
      group_by(zone,LC_type) |> 
      select(zone,LC_type,sup_km2) |> 
      group_split() |> 
      setNames(c('a','b','c','d','e')) |> 
      map(function(x) pull(x,sup_km2))
    map(a,\(x) ifelse(is.na(x),0,x))
})

dataLCV_trend  |> 
  kbl(booktabs = TRUE,digits = 1,align ='r','latex',position='!ht',
      caption='Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface.') %>%
  kable_styling(latex_options="scale_down") %>% 
  kable_paper(full_width = FALSE) %>%
  column_spec(2, image = spec_plot(dataSpark[[1]],same_lim = FALSE)
     ) %>%
  column_spec(3, image = spec_plot(dataSpark[[2]],same_lim = FALSE)
    ) %>%
  column_spec(4, image = spec_plot(dataSpark[[3]],same_lim = FALSE)
    ) %>%
  column_spec(5, image = spec_plot(dataSpark[[4]],same_lim = FALSE)
    ) %>%
  column_spec(6, image = spec_plot(dataSpark[[5]],same_lim = FALSE)
    ) %>%
  column_spec(7, image = spec_plot(dataSpark[[6]],same_lim = FALSE)
    ) %>% 
  add_header_above(c(" ",'Trend of change [$km^2 year^{-1}$]' = 6),escape =FALSE)
```

<!-- \begin{figure}[H] -->
<!-- \centering -->
<!-- \includegraphics[width=\textwidth]{../output/figs/LC_perZoneAndCluster.png} -->
<!-- \caption{ } -->
<!-- \end{figure} -->

## SPI, AED, and ET

## SM and Total Storage

# Discussion

Authors should discuss the results and how they can be interpreted in perspective of previous studies and of the working hypotheses. The findings and their implications should be discussed in the broadest context possible. Future research directions may also be highlighted.

# Conclusion

This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.

# References {-}

### 2023-05-30 article/article.tex
agregar article rticles

\documentclass[preprint,
3p]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%

\usepackage[hyphens]{url}

  \journal{An awesome journal} % Sets Journal name

\usepackage{graphicx}
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
% TODO: Currently lineno needs to be loaded after amsmath because of conflict
% https://github.com/latex-lineno/lineno/issues/5
\usepackage{lineno} % add
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{apalike}

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Comprehensive assessment of the climate-induced water scarcity over continental Chile},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\begin{frontmatter}

  \title{Comprehensive assessment of the climate-induced water scarcity
over continental Chile}
    \author[Hemere Centro de Observación de la Tierra,Universidad
Mayor]{Francisco Zambrano%
  \corref{cor1}%
  \fnref{1}}
   \ead{francisco.zambrano@umayor.com} 
    \author[Another University]{Bob Security%
  %
  }
   \ead{bob@example.com} 
    \author[Another University]{Cat Memes%
  %
  \fnref{2}}
   \ead{cat@example.com} 
    \author[Some Institute of Technology]{Derek Zoolander%
  %
  \fnref{2}}
   \ead{derek@example.com} 
      \affiliation[Some Institute of Technology]{
    organization={Big Wig University},addressline={1 main
street},city={Gotham},postcode={123456},state={State},country={United
States},}
    \affiliation[Another University]{
    organization={Department},addressline={A street
29},city={Manchester,},postcode={2054 NX},country={The Netherlands},}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{This is the first author footnote.}
    \fntext[2]{Another author footnote.}

  \begin{abstract}
  This is the abstract.

  It consists of two paragraphs.
  \end{abstract}
    \begin{keyword}
    keyword1 \sep 
    keyword2
  \end{keyword}

 \end{frontmatter}

\hypertarget{version}{%
\section{Version}\label{version}}

This Rmd-skeleton uses the mdpi Latex template published 2019/02.
However, the official template gets more frequently updated than the
`rticles' package. Therefore, please make sure prior to paper
submission, that you're using the most recent .cls, .tex and .bst files
(available \href{http://www.mdpi.com/authors/latex}{here}).

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

In 2021, the sixth assessment report (AR6) from the working group I of
the IPCC was released \citep{IPCC2021}. Chapter 11 \citep{IPCCCH112021}
indicates that human-induced greenhouse gas emissions have increased the
frequency and/or intensity of some weather and climate extremes. The
evidence has been strengthened since AR5 \citep{IPCC2013}. There is high
confidence that the increasing global warning can expand the land area
affected by increasing drought frequency and severity
\citep{IPCCCH112021}. Chile has been facing a persistent rainfall
deficit lasting for more than ten years \citep{Garreaud2017} which has
impacted the hydrological system \citep{Boisier2018}, and consequently
the vegetation development \citep{Zambrano2020}.

Precipitation is the primary driver of drought that impacts hydrological
regimes and vegetation productivity. Thus, it is commonly classified as
meteorological, hydrological, and agricultural \citep{Wilhite1985}.
Lately, it has been argued that this definition does not fully address
the ecological dimensions \citep{Crausbay2017}. \citet{Crausbay2017}
proposed the ecological drought definition as ``an episodic deficit in
water availability that drives ecosystems beyond thresholds of
vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems''. The AR6 \citep{IPCC2021} state that even
if global warming is stabilized at 1.5°-2°C many parts of the world will
be impacted by more severe agricultural and ecological drought. Central
Chile has suffered from crop productivity failure, highlighting the
growing season 2007-2008 and 2008-2009
\citep{Zambrano2016, Zambrano2018}, which impacted an extensive surface.
But, in 2019-2020, the drought intensity reached an extreme condition at
North 34°S not seen -at least- for more than 40 years
\citep{Zambrano2020}, affecting forest, grassland, and croplands areas.
The prolonged lack of precipitation within Central Chile is producing
changes in the ecosystem that should study.

Satellite remote sensing \citep{West2019, AghaKouchak2015} is the
primary method to evaluate how meteorological drought impacts vegetation
dynamics. Since the 90's multiple vegetation drought indices have been
derived (VCI,\citep{Kogan1990}; TCI, \citep{Kogan1995};zNDVI,
\citep{Peters2002}; VegDri, \citep{Brown2008}) that have allowed making
spatiotemporal analysis. Although we can calculate those indices for any
time during the year (depending on satellite revisit), there are
relevant during the stage vegetation has more activity, the growing
season \citep{Mishra2015}. Although modeling phenology is a complex
task, satellites offer strategies that help to address it
\citep{Younes2021, Vrieling2018, Cai2017}. Also, the land cover dynamics
product MCD12Q2 from the USGS \citep{Friedl2019} provides some phenology
metrics. Some authors have proposed indices aggregated during the
season. \citet{Meroni2017} accumulating the fractional active
photosynthetic active radiation(FAPAR) between the start (SOS) and the
end of the season (EOS) in the Sahel, calculate the zCFAPAR.
\citet{Zambrano2018} used the same approach but with the NDVI
(Normalized Difference Vegetation Index), derivating the zcNDVI within
Central Chile. Besides, land use land cover (LULC) change can be driven
by drought \citep{Tran2019, Akinyemi2021}. To analyze those changes,
multiple time-series LULC products exist as the MCD12Q1
\citep{Friedl2019} and the ESA CCI-LC \citep{ESA2017}. For Chile, 2014
was made a high-resolution land cover at 30m of spatial resolution
\citep{Zhao2016}. The LULC product with the vegetation drought index can
help evaluate the impact of drought on the ecosystem.

Vegetation drought indices are proxies of productivity
\citep{Paruelo2016, Schucknecht2017}. The main environmental variables
that affect it are water supply and demand \citep{Mishra2015}. We
measure them by precipitation and evapotranspiration (ET), commonly
collected from weather stations. Usually, in developing countries (i.e.,
Chile), incomplete records or gaps present a challenge. But, there are
satellite estimates of these variables. To evaluate drought, the World
Meteorological Organization (WMO; \citep{WMO2012}) has proposed the
Standardized Precipitation Index (SPI; \citep{McKee1993}), a multiscalar
drought index, which has been used worldwide. For Chile,
\citet{Zambrano2017} derived and evaluated it from the product of the
Climate Hazards Group InfraRed Precipitation with Station data (CHIRPS;
\citep{Funk2015}). For water demand, it is used ET. The vegetation
biomass productivity is strongly related to ET \citep{FAO66}. The
atmospheric evaporative demand (AED) represents the maximum ET rate from
a land surface (without water restriction), also known as reference ET.
The recommended method for its calculation is the FAO Penman-Monteith
\citep{Pereira2015, Allen2005}. Due to climate change, AED is
increasing, driving ET rise \citep{IPCCCH112021}. But, it is not always
true \citep{Milly2016}. For example, regions where AET is highest have
the lowest ET. The MOD16 product \citep{Running2021, Mu2011} provides
AET and ET satellite estimates and has been used to derive drought
indices \citep{Mu2013}. Soil moisture (SM) is an Essential Climate
Variable (ECV) that modulates vegetative growth. The climate change
initiative (CCI) from the European Space Agency (ESA) delivers the ESA
CCI SM product \citep{Dorigo2017} (current version 6.1), which has been
helpful to monitor drought \citep{Zhang2019}. Besides, total water
storage can be retrieved by the Gravity Recovery and Climate Experiment
(GRACE), which allows analyzing water availability changes
\citep{Ahmed2014, Ma2017}. The water demand and supply by remote sensing
can help evaluate how they have impacted vegetation productivity.

The study aims to analyze the drought impact on vegetation through
Central Chile for 2000-2020, using satellite data as proxies of
productivity and water demand and supply. We will evaluate LULC change
and use the persistent classes within 2001-2019 (\textgreater{} 80\%) to
analyze the zcNDVI index and its interconnection with precipitation
deficit, AED, ET, and vegetation cover. Finally, we will investigate if
the observed changes are linked to the TWS and SM.

\hypertarget{study-area}{%
\section{Study area}\label{study-area}}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{../output/figs/map_study_con_landcover.png}
\caption{ (\textbf{a}) Location of Central Chile and zones north (NCCH), central (CCCH), and south (SCCH) Central Chile. (\textbf{b}) Topography reference map. (\textbf{c}) Land cover classes for 2019. (\textbf{d}) Persistent land cover classess (> 80\%) for 2001-2019.}
\end{figure}

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

\hypertarget{satellite-data}{%
\subsection{Satellite data}\label{satellite-data}}

\hypertarget{zcndvi-and-cluster-zones}{%
\subsection{zcNDVI and cluster zones}\label{zcndvi-and-cluster-zones}}

\hypertarget{landcover-change-and-persistence}{%
\subsection{Landcover change and
persistence}\label{landcover-change-and-persistence}}

\hypertarget{vegetation-cover}{%
\subsection{Vegetation Cover}\label{vegetation-cover}}

\hypertarget{evapotranspiration-soil-moisture-and-water-storage}{%
\subsection{Evapotranspiration, Soil moisture and water
storage}\label{evapotranspiration-soil-moisture-and-water-storage}}

\hypertarget{data-analysis}{%
\subsection{Data analysis}\label{data-analysis}}

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{zcndvi-and-cluster-zones-1}{%
\subsection{zcNDVI and cluster zones}\label{zcndvi-and-cluster-zones-1}}

\hypertarget{landcover-change-and-persistent}{%
\subsection{Landcover Change and
persistent}\label{landcover-change-and-persistent}}

\begin{table}[!ht]

\caption{\label{tab:unnamed-chunk-1}Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{r>{}r>{}r>{}r>{}r>{}r>{}r}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{6}{c}{Trend of change [$km^2 year^{-1}$]} \\
\cmidrule(l{3pt}r{3pt}){2-7}
zone & Shrubland & Savanna & Grassland & Barren land & Forest & Cropland\\
\midrule
norte grande & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c19134a3d.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c11281a1c.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c76e40d4c.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396cc832d9.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c22c3df0c.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c1e23c330.pdf}\\
norte chico & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c3e39b63c.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c3a55945c.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c5ea10cfa.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c3e253181.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c6e3f818e.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c381a2e04.pdf}\\
zona central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c752a96cf.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c400c5e63.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c511304bc.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c58fe26f6.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c16afaacc.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c66febecf.pdf}\\
zona sur & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c1ee86964.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c457bef18.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c688683cf.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c7f0e5ef0.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c4f819a2c.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c1671afb3.pdf}\\
zona austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c7a04eff2.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c292ba1fa.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c362a0ad3.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c32966b1b.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c211e7d1e.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396cb504520.pdf}\\
\bottomrule
\end{tabular}}
\end{table}

\hypertarget{spi-aed-and-et}{%
\subsection{SPI, AED, and ET}\label{spi-aed-and-et}}

\hypertarget{sm-and-total-storage}{%
\subsection{SM and Total Storage}\label{sm-and-total-storage}}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Authors should discuss the results and how they can be interpreted in
perspective of previous studies and of the working hypotheses. The
findings and their implications should be discussed in the broadest
context possible. Future research directions may also be highlighted.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

This section is not mandatory, but can be added to the manuscript if the
discussion is unusually long or complex.

\renewcommand\refname{References}
\bibliography{references.bib}

\end{document}

### 2023-05-30 scripts/15_IGBP_variacion_tendencia_2001-2021.R

@@ -1,4 +1,4 @@
-# Analyzing landcover variation through central Chile
+# Analyzing landcover variation through Chile
 # August 2021
 # by frzambra
 
@@ -23,8 +23,8 @@ classes <- matrix(c(1:17,rep(1,5),rep(2,2),rep(3,2),4,5,6,7,6,8,9,10),ncol=2)
 igbpRecl <- classify(igbp,rcl=classes,include.lowest = TRUE)
 igbpRecl[igbpRecl == 0] <- NA
 
-zones <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
-  st_transform(32719)
+zones <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') 
+
 data <- terra::extract(igbpRecl,zones)
 names(data) <- c('zone',2001:2021)
 
@@ -72,9 +72,9 @@ data3  |>
 library(tidy)
 library(broom)
 
-data3|> 
+data3 |> 
   ungroup() |>  
   nest_by(zone,LC_type) |>  
-  summarize(trend  = as.numeric(tidy(lm(sup_km2~year,data))[2,2])) |>  
+  summarize(trend  = as.numeric(tidy(lm(sup_km2~year,data))[2,2])) |> 
   pivot_wider(names_from = 'LC_type',values_from=3) |> 
-  saveRDS(.,'data/processed_data/trends_landcover_2001-2021.rds')  
+  saveRDS('data/processed_data/trends_landcover_2001-2021.rds')  

### 2023-05-30 article/article.Rmd
mejorar introduccion, estructura y empezar abstract

@@
     email: francisco.zambrano@umayor.com
-    affiliation: Hemere Centro de Observación de la Tierra,Universidad Mayor
+    affiliation: Hemera Centro de Observación de la Tierra,Universidad Mayor
     correspondingauthor: true
@@
 address:
-  - code: Some Institute of Technology
-    organization: Big Wig University
-    addressline: 1 main street
-    city: Gotham
-    state: State
+  - code: Hemera Centro de Observación de la Tierra
+    organization: Facultad de Ciencias, Universidad Mayor.
+    addressline: La Piramide 5750
+    city: Santiago
+    state: Huechuraba
     postcode: 123456
@@
 abstract: |
-  This is the abstract.
+  Human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes globally.
+Chile has been affected by persistent water scarcity which has been impacting the hydrological system and vegetation development.
+Central Chile it is been the focus of research studies due to the diminishing of water supply, nevertheless water deficit is expanded beyond, specifically to the south of the country.
 
-  It consists of two paragraphs.
 keywords: 
@@
   - keyword2
-journal: "An awesome journal"
+journal: "Remote Sensing of Environment"
 date: "`r Sys.Date()`"
@@
 
-In 2021, the sixth assessment report (AR6) from the working group I of the IPCC was released [@IPCC2021].
-Chapter 11 [@IPCCCH112021] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes.
-The evidence has been strengthened since AR5 [@IPCC2013].
+The sixth assessment report (AR6) from the working group I of the IPCC [@IPCC2021] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes and the evidence has been strengthened since AR5 [@IPCC2013].
 There is high confidence that the increasing global warning can expand the land area affected by increasing drought frequency and severity [@IPCCCH112021].
@@
 
-Vegetation drought indices are proxies of productivity [@Paruelo2016;@Schucknecht2017].
-The main environmental variables that affect it are water supply and demand [@Mishra2015].
-We measure them by precipitation and evapotranspiration (ET), commonly collected from weather stations.
+Vegetation drought indices are commonly used as proxies of productivity [@Paruelo2016;@Schucknecht2017].
+The main environmental variables that affect productivity are water supply and demand [@Mishra2015].
+Those are measured by precipitation and evapotranspiration (ET), commonly collected from weather stations.
 Usually, in developing countries (i.e., Chile), incomplete records or gaps present a challenge.
@@
 
-The study aims to analyze the drought impact on vegetation through Central Chile for 2000-2020, using satellite data as proxies of productivity and water demand and supply.
-We will evaluate LULC change and use the persistent classes within 2001-2019 (> 80%) to analyze the zcNDVI index and its interconnection with precipitation deficit, AED, ET, and vegetation cover.
-Finally, we will investigate if the observed changes are linked to the TWS and SM.
+We aims to analyze the climate-induced water scarcity over continental Chile for 2000-2023 by using estimated environmental variables of biomass productivity, and water demand/supply gather from earth observation products.
+The specific objective for the study are i) to evaluate LULC change for 2001-2021, ii) to derive and  assess the vegetation drought index zcNDVI as a proxy of biomass productivity, iii) analyze the interconnection of zcNDVI with drought indices of supply (i.e., precipitation), demand (i.e, ET), and vegetation cover; and iv) we will investigate if the observed changes are linked to the TWS and SM.
 
@@
 
-\begin{figure}[H]
+\begin{figure}[!ht]
 \centering
@@
 (\textbf{c}) Land cover classes for 2019.
-(\textbf{d}) Persistent land cover classess (> 80\%) for 2001-2019.}
+(\textbf{d}) Persistent land cover classess (> 80\%) for 2001-2021.}
 \end{figure}
@@
 
-## Satellite data
+## Data
 
-## zcNDVI and cluster zones
+### Satellite data
 
-## Landcover change and persistence
+### in-situ data
 
-## Vegetation Cover
+## LULC change for 2001-2021
 
-## Evapotranspiration, Soil moisture and water storage
+### Landcover change and persistence
 
-## Data analysis
+## Drought index for biomass productivity zcNDVI
+
+## Interconnection of productivity with drought indices of supply/demand and vegetation cover
+
+## Totatl water storage and Soil Moisture
 
@@
 
-## zcNDVI and cluster zones
-<!-- \begin{figure}[H] -->
-<!-- \includegraphics[width=\textwidth]{../figs/map_zcNDVI_ts_clusters.png} -->
-<!-- \caption{ } -->
-<!-- \end{figure} -->
+## LULC change for 2001-2021
 
-## Landcover Change and persistent
+```{r,echo=FALSE,fig.width =10,fig.pos='H',warning=FALSE}
 
-```{r,echo=FALSE,fig.width =10,fig.pos='H'}
 dataLCV_ts <- readRDS('../data/processed_data/timeseries_landcover_zone_LCclass_2001-2021.rds')  
@@
 
-<!-- \begin{figure}[H] -->
-<!-- \centering -->
-<!-- \includegraphics[width=\textwidth]{../output/figs/LC_perZoneAndCluster.png} -->
-<!-- \caption{ } -->
-<!-- \end{figure} -->
+### Landcover change and persistence
 
-## SPI, AED, and ET
+## Drought index for biomass productivity zcNDVI
 
-## SM and Total Storage
+## Interconnection of productivity with drought indices of supply/demand and vegetation cover
 
+## Total water storage and Soil Moisture
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=\textwidth]{../output/figs/JPL_TELLUS_GRACE-FO_L3_Monthly_Land_WET_SMA_R61v04_macrozonas.png}
+\caption{Total water storage (mm) trend for 2002-2023 in the five macrozones on continental Chile}
+\end{figure}
 # Discussion

### 2023-05-30 article/article.tex
mejorar introduccion, estructura y empezar abstract

@@
 
-  \journal{An awesome journal} % Sets Journal name
+  \journal{Remote Sensing of Environment} % Sets Journal name
 
@@
 over continental Chile}
-    \author[Hemere Centro de Observación de la Tierra,Universidad
+    \author[Hemera Centro de Observación de la Tierra,Universidad
 Mayor]{Francisco Zambrano%
@@
    \ead{derek@example.com} 
-      \affiliation[Some Institute of Technology]{
-    organization={Big Wig University},addressline={1 main
-street},city={Gotham},postcode={123456},state={State},country={United
+      \affiliation[Hemera Centro de Observación de la Tierra]{
+    organization={Facultad de Ciencias, Universidad
+Mayor.},addressline={La Piramide
+5750},city={Santiago},postcode={123456},state={Huechuraba},country={United
 States},}
@@
   \begin{abstract}
-  This is the abstract.
-
-  It consists of two paragraphs.
+  Human-induced greenhouse gas emissions have increased the frequency
+  and/or intensity of some weather and climate extremes globally. Chile
+  has been affected by persistent water scarcity which has been
+  impacting the hydrological system and vegetation development. Central
+  Chile it is been the focus of research studies due to the diminishing
+  of water supply, nevertheless water deficit is expanded beyond,
+  specifically to the south of the country.
   \end{abstract}
@@
 
-In 2021, the sixth assessment report (AR6) from the working group I of
-the IPCC was released \citep{IPCC2021}. Chapter 11 \citep{IPCCCH112021}
-indicates that human-induced greenhouse gas emissions have increased the
-frequency and/or intensity of some weather and climate extremes. The
-evidence has been strengthened since AR5 \citep{IPCC2013}. There is high
-confidence that the increasing global warning can expand the land area
-affected by increasing drought frequency and severity
-\citep{IPCCCH112021}. Chile has been facing a persistent rainfall
-deficit lasting for more than ten years \citep{Garreaud2017} which has
-impacted the hydrological system \citep{Boisier2018}, and consequently
-the vegetation development \citep{Zambrano2020}.
+The sixth assessment report (AR6) from the working group I of the IPCC
+\citep{IPCC2021} indicates that human-induced greenhouse gas emissions
+have increased the frequency and/or intensity of some weather and
+climate extremes and the evidence has been strengthened since AR5
+\citep{IPCC2013}. There is high confidence that the increasing global
+warning can expand the land area affected by increasing drought
+frequency and severity \citep{IPCCCH112021}. Chile has been facing a
+persistent rainfall deficit lasting for more than ten years
+\citep{Garreaud2017} which has impacted the hydrological system
+\citep{Boisier2018}, and consequently the vegetation development
+\citep{Zambrano2020}.
 
@@
 
-Vegetation drought indices are proxies of productivity
+Vegetation drought indices are commonly used as proxies of productivity
 \citep{Paruelo2016, Schucknecht2017}. The main environmental variables
-that affect it are water supply and demand \citep{Mishra2015}. We
-measure them by precipitation and evapotranspiration (ET), commonly
-collected from weather stations. Usually, in developing countries (i.e.,
-Chile), incomplete records or gaps present a challenge. But, there are
-satellite estimates of these variables. To evaluate drought, the World
-Meteorological Organization (WMO; \citep{WMO2012}) has proposed the
-Standardized Precipitation Index (SPI; \citep{McKee1993}), a multiscalar
-drought index, which has been used worldwide. For Chile,
-\citet{Zambrano2017} derived and evaluated it from the product of the
-Climate Hazards Group InfraRed Precipitation with Station data (CHIRPS;
-\citep{Funk2015}). For water demand, it is used ET. The vegetation
-biomass productivity is strongly related to ET \citep{FAO66}. The
-atmospheric evaporative demand (AED) represents the maximum ET rate from
-a land surface (without water restriction), also known as reference ET.
-The recommended method for its calculation is the FAO Penman-Monteith
-\citep{Pereira2015, Allen2005}. Due to climate change, AED is
-increasing, driving ET rise \citep{IPCCCH112021}. But, it is not always
-true \citep{Milly2016}. For example, regions where AET is highest have
-the lowest ET. The MOD16 product \citep{Running2021, Mu2011} provides
-AET and ET satellite estimates and has been used to derive drought
-indices \citep{Mu2013}. Soil moisture (SM) is an Essential Climate
-Variable (ECV) that modulates vegetative growth. The climate change
-initiative (CCI) from the European Space Agency (ESA) delivers the ESA
-CCI SM product \citep{Dorigo2017} (current version 6.1), which has been
-helpful to monitor drought \citep{Zhang2019}. Besides, total water
-storage can be retrieved by the Gravity Recovery and Climate Experiment
-(GRACE), which allows analyzing water availability changes
-\citep{Ahmed2014, Ma2017}. The water demand and supply by remote sensing
-can help evaluate how they have impacted vegetation productivity.
+that affect productivity are water supply and demand \citep{Mishra2015}.
+Those are measured by precipitation and evapotranspiration (ET),
+commonly collected from weather stations. Usually, in developing
+countries (i.e., Chile), incomplete records or gaps present a challenge.
+But, there are satellite estimates of these variables. To evaluate
+drought, the World Meteorological Organization (WMO; \citep{WMO2012})
+has proposed the Standardized Precipitation Index (SPI;
+\citep{McKee1993}), a multiscalar drought index, which has been used
+worldwide. For Chile, \citet{Zambrano2017} derived and evaluated it from
+the product of the Climate Hazards Group InfraRed Precipitation with
+Station data (CHIRPS; \citep{Funk2015}). For water demand, it is used
+ET. The vegetation biomass productivity is strongly related to ET
+\citep{FAO66}. The atmospheric evaporative demand (AED) represents the
+maximum ET rate from a land surface (without water restriction), also
+known as reference ET. The recommended method for its calculation is the
+FAO Penman-Monteith \citep{Pereira2015, Allen2005}. Due to climate
+change, AED is increasing, driving ET rise \citep{IPCCCH112021}. But, it
+is not always true \citep{Milly2016}. For example, regions where AET is
+highest have the lowest ET. The MOD16 product
+\citep{Running2021, Mu2011} provides AET and ET satellite estimates and
+has been used to derive drought indices \citep{Mu2013}. Soil moisture
+(SM) is an Essential Climate Variable (ECV) that modulates vegetative
+growth. The climate change initiative (CCI) from the European Space
+Agency (ESA) delivers the ESA CCI SM product \citep{Dorigo2017} (current
+version 6.1), which has been helpful to monitor drought
+\citep{Zhang2019}. Besides, total water storage can be retrieved by the
+Gravity Recovery and Climate Experiment (GRACE), which allows analyzing
+water availability changes \citep{Ahmed2014, Ma2017}. The water demand
+and supply by remote sensing can help evaluate how they have impacted
+vegetation productivity.
 
-The study aims to analyze the drought impact on vegetation through
-Central Chile for 2000-2020, using satellite data as proxies of
-productivity and water demand and supply. We will evaluate LULC change
-and use the persistent classes within 2001-2019 (\textgreater{} 80\%) to
-analyze the zcNDVI index and its interconnection with precipitation
-deficit, AED, ET, and vegetation cover. Finally, we will investigate if
-the observed changes are linked to the TWS and SM.
+We aims to analyze the climate-induced water scarcity over continental
+Chile for 2000-2023 by using estimated environmental variables of
+biomass productivity, and water demand/supply gather from earth
+observation products. The specific objective for the study are i) to
+evaluate LULC change for 2001-2021, ii) to derive and assess the
+vegetation drought index zcNDVI as a proxy of biomass productivity, iii)
+analyze the interconnection of zcNDVI with drought indices of supply
+(i.e., precipitation), demand (i.e, ET), and vegetation cover; and iv)
+we will investigate if the observed changes are linked to the TWS and
+SM.
 
@@
 
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
+
 \hypertarget{satellite-data}{%
-\subsection{Satellite data}\label{satellite-data}}
+\subsubsection{Satellite data}\label{satellite-data}}
 
-\hypertarget{zcndvi-and-cluster-zones}{%
-\subsection{zcNDVI and cluster zones}\label{zcndvi-and-cluster-zones}}
+\hypertarget{in-situ-data}{%
+\subsubsection{in-situ data}\label{in-situ-data}}
+
+\hypertarget{lulc-change-for-2001-2021}{%
+\subsection{LULC change for 2001-2021}\label{lulc-change-for-2001-2021}}
 
 \hypertarget{landcover-change-and-persistence}{%
-\subsection{Landcover change and
+\subsubsection{Landcover change and
 persistence}\label{landcover-change-and-persistence}}
 
-\hypertarget{vegetation-cover}{%
-\subsection{Vegetation Cover}\label{vegetation-cover}}
+\hypertarget{drought-index-for-biomass-productivity-zcndvi}{%
+\subsection{Drought index for biomass productivity
+zcNDVI}\label{drought-index-for-biomass-productivity-zcndvi}}
 
-\hypertarget{evapotranspiration-soil-moisture-and-water-storage}{%
-\subsection{Evapotranspiration, Soil moisture and water
-storage}\label{evapotranspiration-soil-moisture-and-water-storage}}
+\hypertarget{interconnection-of-productivity-with-drought-indices-of-supplydemand-and-vegetation-cover}{%
+\subsection{Interconnection of productivity with drought indices of
+supply/demand and vegetation
+cover}\label{interconnection-of-productivity-with-drought-indices-of-supplydemand-and-vegetation-cover}}
 
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
+\hypertarget{totatl-water-storage-and-soil-moisture}{%
+\subsection{Totatl water storage and Soil
+Moisture}\label{totatl-water-storage-and-soil-moisture}}
 
@@
 
-\hypertarget{zcndvi-and-cluster-zones-1}{%
-\subsection{zcNDVI and cluster zones}\label{zcndvi-and-cluster-zones-1}}
-
-\hypertarget{landcover-change-and-persistent}{%
-\subsection{Landcover Change and
-persistent}\label{landcover-change-and-persistent}}
+\hypertarget{lulc-change-for-2001-2021-1}{%
+\subsection{LULC change for
+2001-2021}\label{lulc-change-for-2001-2021-1}}
 
@@
 \midrule
-norte grande & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c19134a3d.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c11281a1c.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c76e40d4c.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396cc832d9.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c22c3df0c.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c1e23c330.pdf}\\
-norte chico & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c3e39b63c.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c3a55945c.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c5ea10cfa.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c3e253181.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c6e3f818e.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c381a2e04.pdf}\\
-zona central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c752a96cf.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c400c5e63.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c511304bc.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c58fe26f6.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c16afaacc.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c66febecf.pdf}\\
-zona sur & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c1ee86964.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c457bef18.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c688683cf.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c7f0e5ef0.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c4f819a2c.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c1671afb3.pdf}\\
-zona austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c7a04eff2.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c292ba1fa.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c362a0ad3.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c32966b1b.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396c211e7d1e.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_d396cb504520.pdf}\\
+norte grande & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671073045199.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671013721869.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671079e41f67.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671075ebd9d8.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710524640a1.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671012073e8a.pdf}\\
+norte chico & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671061580771.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967101832bef9.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967105771ba87.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671042ea518f.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710607baab3.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967106f91099.pdf}\\
+zona central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967105c411aeb.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967108a6119a.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671016b408b2.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967103ee6f936.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671074c30c6.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671056667823.pdf}\\
+zona sur & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967106cc832ad.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671034e3692f.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967104ac90971.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967107f7df47e.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967104e385249.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671013698bfb.pdf}\\
+zona austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710491ed82a.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710190c87cf.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967101d6ab929.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671033cd1ca4.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967101f22c71d.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710336ead1.pdf}\\
 \bottomrule
@@
 
-\hypertarget{spi-aed-and-et}{%
-\subsection{SPI, AED, and ET}\label{spi-aed-and-et}}
+\hypertarget{landcover-change-and-persistence-1}{%
+\subsubsection{Landcover change and
+persistence}\label{landcover-change-and-persistence-1}}
 
-\hypertarget{sm-and-total-storage}{%
-\subsection{SM and Total Storage}\label{sm-and-total-storage}}
+\hypertarget{drought-index-for-biomass-productivity-zcndvi-1}{%
+\subsection{Drought index for biomass productivity
+zcNDVI}\label{drought-index-for-biomass-productivity-zcndvi-1}}
+
+\hypertarget{interconnection-of-productivity-with-drought-indices-of-supplydemand-and-vegetation-cover-1}{%
+\subsection{Interconnection of productivity with drought indices of
+supply/demand and vegetation
+cover}\label{interconnection-of-productivity-with-drought-indices-of-supplydemand-and-vegetation-cover-1}}
+
+\hypertarget{total-water-storage-and-soil-moisture}{%
+\subsection{Total water storage and Soil
+Moisture}\label{total-water-storage-and-soil-moisture}}
+
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=\textwidth]{../output/figs/JPL_TELLUS_GRACE-FO_L3_Monthly_Land_WET_SMA_R61v04_macrozonas.png}
+\caption{Total water storage (mm) trend for 2002-2023 in the five macrozones on continental Chile}
+\end{figure}
 

### 2023-05-30 article/article.Rmd
escribir abstract

@@
   Human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes globally.
-Chile has been affected by persistent water scarcity which has been impacting the hydrological system and vegetation development.
-Central Chile it is been the focus of research studies due to the diminishing of water supply, nevertheless water deficit is expanded beyond, specifically to the south of the country.
+Chile has been affected by persistent water scarcity which is impacting the hydrological system and vegetation development.
+Central Chile it is been the focus of research studies due to the diminishing water supply, nevertheless water deficit is expanded beyond.
+We analyze earth observation data for 2000-2023 to do a comprehensive assessment of water scarcity in Chile.
+For the analysis, continental Chile was divided into five zones from north to south.
+We used the time series of MODIS for land cover change (LULC), fraction of vegetation cover (FVC), and to derive the drought index zcNDVI and zcET (standardized anomaly of cumulative NDVI and ET through the growing season); and CHELSA v2.1 dataset to derive the standardized precipitation evapotranspiration index (SPEI).
+We evaluate the interconnection between the productivity drought index (zcNDVI) and water demand (zcET), supply (SPEIs) and FVC.
+Finally, we analyzed the temporal correlation of the drought indices with total water storage (TWS) from GRACE (Gravity Recovery and Climate Experiment).
+Our results showed that ...
 
@@
 numbersections: true
-# Use a CSL with `citation_package = "default"`
 # csl: https://www.zotero.org/styles/elsevier-harvard
@@
     citation_package: natbib
+  word_document: default
 ---

### 2023-05-30 article/article.tex
escribir abstract

@@
   and/or intensity of some weather and climate extremes globally. Chile
-  has been affected by persistent water scarcity which has been
-  impacting the hydrological system and vegetation development. Central
-  Chile it is been the focus of research studies due to the diminishing
-  of water supply, nevertheless water deficit is expanded beyond,
-  specifically to the south of the country.
+  has been affected by persistent water scarcity which is impacting the
+  hydrological system and vegetation development. Central Chile it is
+  been the focus of research studies due to the diminishing water
+  supply, nevertheless water deficit is expanded beyond. We analyze
+  earth observation data for 2000-2023 to do a comprehensive assessment
+  of water scarcity in Chile. For the analysis, continental Chile was
+  divided into five zones from north to south. We used the time series
+  of MODIS for land cover change (LULC), fraction of vegetation cover
+  (FVC), and to derive the drought index zcNDVI and zcET (standardized
+  anomaly of cumulative NDVI and ET through the growing season); and
+  CHELSA v2.1 dataset to derive the standardized precipitation
+  evapotranspiration index (SPEI). We evaluate the interconnection
+  between the productivity drought index (zcNDVI) and water demand
+  (zcET), supply (SPEIs) and FVC. Finally, we analyzed the temporal
+  correlation of the drought indices with total water storage (TWS) from
+  GRACE (Gravity Recovery and Climate Experiment). Our results showed
+  that \ldots{}
   \end{abstract}
@@
 \midrule
-norte grande & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671073045199.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671013721869.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671079e41f67.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671075ebd9d8.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710524640a1.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671012073e8a.pdf}\\
-norte chico & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671061580771.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967101832bef9.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967105771ba87.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671042ea518f.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710607baab3.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967106f91099.pdf}\\
-zona central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967105c411aeb.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967108a6119a.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671016b408b2.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967103ee6f936.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671074c30c6.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671056667823.pdf}\\
-zona sur & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967106cc832ad.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671034e3692f.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967104ac90971.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967107f7df47e.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967104e385249.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671013698bfb.pdf}\\
-zona austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710491ed82a.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710190c87cf.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967101d6ab929.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_19671033cd1ca4.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1967101f22c71d.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_196710336ead1.pdf}\\
+norte grande & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70937146e24b.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709384faf70.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709373919fc5.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709321ad9f19.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709379ea79bd.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093c8ff3c5.pdf}\\
+norte chico & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709337b6241e.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70937f0ed48a.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70934b4bd377.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70931e4fd155.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70932e8b190a.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093305c1274.pdf}\\
+zona central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70932bbec995.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093598f66c9.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709370472832.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093ba3805d.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709341ec566.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70933b82960c.pdf}\\
+zona sur & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70936a87e29b.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70935b745f29.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709363a8f17a.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70932c57d58c.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70935fdd829a.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709328c1f26f.pdf}\\
+zona austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709343d0ce35.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093615c29b7.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70933872723d.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709369bca642.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709386fdff0.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70931a654519.pdf}\\
 \bottomrule

### 2023-10-08 article/paper_drought_Chile2023.qmd
working third paragraph introduction

---
title: Drought assessment of the impact of water supply and demand over land cover in continental Chile for 2000-2023 from ERA5-Land and MODIS datasets
#subtitle: A Short Subtitle
author:
  - name: Francisco Zambrano
    email: francisco.zambrano@umayor.cl
    affiliations: 
        - id: umayor
          name: Universidad Mayor
          department: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad
          #address: Street Address
          city: Santiago, Chile
          #state: State
          postal-code: 7500994
    attributes:
        corresponding: true
    note: This is the first author footnote.
  # - name: Bob Security
  #   email: bob@example.com
  #   affiliations:
  #       - id: another-u
  #         name: Another University
  #         department: Department Name
  #         address: Street Address
  #         city: City
  #         state: State
  #         postal-code: Postal Code
  #   note: |
  #     Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  # - name: Cat Memes
  #   email: cat@example.com
  #   affiliations:
  #       - ref: another-u
  #   note: Yet another author footnote.
  # - name: Derek Zoolander
  #   email: derek@example.com
  #   affilations:
  #       - ref: some-tech
abstract: |
  Human-induced greenhouse gas emissions have increased the frequency and/or intensity of weather and climate extremes. Central Chile has been affected by a persistent drought which is impacting the hydrological system and vegetation development. The region has been the focus of research studies due to the diminishing water supply, this persistent period of water scarcity has been defined as a “mega drought". Nevertheless, our results evidence that the water deficit has expanded beyond. Our goal is to analyze the impact of drought, measured by drought indices of water supply/demand and vegetation status, in the LULCC (land use land cover change) over continental Chile.  For the analysis, continental Chile was divided into five zones according to a latitudinal gradient: “Norte Grande”,  “Norte Chico”,  “Zona Central”, “Zona Sur”, and “Zona Austral”.  We used monthly climatic re-analysis variables for precipitation, temperature and soil moisture for 1981-2023 from ERA5-Land; and MODIS (Moderate-Resolution Imaging Spectroradiometer) product MCD12Q1 for land cover for 2001-2021, and the NDVI vegetation index from product MOD13A2 collection 6.1 for 2000-2023, both from collection 6.1. We estimated atmospheric evaporative demand (AED) by combining the Hargreaves-Samani equation with the ERA5-Land temperature. We derived the drought indices SPI (Standardized Precipitation Index), SPEI (Standardized Precipitation Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), zcSM (standardized anomaly of cumulative soil moisture), and the zcNDVI (standardized anomaly of cumulative NDVI). These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1, 3, and 6 months). We analyzed the temporal correlation of SPI, SPEI, EDDI, and zcSM with zcNDVI to have insights into the impact of water supply and demand on vegetation.   Our results showed that LULCC had an increasing trend of 412 [km2yr−1] of forest expansion in the “Zona Sur”, together with a  decreasing trend of 24 [km2yr−1] of cropland contraction in the “Zona Central” meanwhile the “Zona Sur” showed an increase of 31 [km2yr−1], and a contraction of 80 [km2yr−1] of bare soil in the “Zona Austral”. The EDDI was the less correlated index for the five macro zones and the five types of land cover, showing that the temperature in Chile has little impact on vegetation. Higher r-squared values, between 0.5 and 0.8, were obtained at “Norte Chico” and “Zona Central” for the land cover types of savanna, shrubland, grassland, and croplands for the indices SPEI and zcSM at time scales of 12 and 24 months. The forest type reaches a r-squared of ~0.5 for zcSM of 12 months. The results indicate that the “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits longer than a year, potentially explained by a low capacity of water storage in those zones that should be further investigated.
keywords: 
  - drought
  - land cover change
  - satellite
date: last-modified
bibliography: references.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      natbiboptions: numbers
      #cite-style: super
---

# Introduction

The sixth assessment report (AR6) of the IPCC [@IPCC2023] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes, and the evidence has been strengthened since AR5 [@IPCC2013]. There is high confidence that increasing global warming can expand the land area affected by increasing drought frequency and severity [@IPCCCH112021]. Furthermore, drought increases tree mortality and triggers changes in land cover and, consequently, land use, thus impacting ecosystems [@Crausbay2017]. Nevertheless, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.

Precipitation is the primary driver of drought, which impacts soil moisture, hydrological regimes, and vegetation productivity. Initially, drought was commonly classified as meteorological, hydrological, and agricultural [@Wilhite1985]. Lately, @Loon2016 and @AghaKouchak2021 have given an updated definition of drought for the Anthropocene, suggesting that it should be considered the feedback of humans' decisions and activities that drives the anthropogenic drought. Even though it has been argued that those definitions do not fully address the ecological dimensions of drought. @Crausbay2017 proposed the ecological drought definition as *"an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems"*. Moreover, many ecological studies have misinterpreted how to characterize drought, for example, sometimes considering "dry" conditions as "drought" [@Slette2019]. On the other hand, the AR6 [@IPCC2023] states that even if global warming is stabilized at 1.5°–2°C, many parts of the world will be impacted by more severe agricultural and ecological droughts. Then, there is a challenge in conducting drought research, especially its impact on ecosystems.

Chile has been facing a persistent rainfall deficit for more than a decade [@Garreaud2017], which has impacted the the vegetation development [@Zambrano2023] and hydrological system [@Boisier2018]. Current drought conditions have affected crop yields in Central Chile. Highlighting the growing seasons 2007-2008 and 2008-2009 [@Zambrano2016, @Zambrano2018], which impacted an extensive surface in [COMPLETAR]. But, in 2019–2020, the drought intensity reached an extreme condition in North 34°S not seen for at least 40 years [@Zambrano2023], thus affecting forest, grassland, and cropland areas. The prolonged lack of precipitation in Central Chile is producing changes in ecosystem dynamics that must be studied.

Satellite remote sensing [@West2019; @AghaKouchak2015] is the primary method to evaluate how meteorological drought impacts vegetation dynamics. Since the 90’s multiple vegetation drought indices have been developed, such as (VCI,@Kogan1990; TCI, @Kogan1995; zNDVI, [@Peters2002]; VegDri, [@Brown2008]) that have allowed making spatiotemporal analysis. Although we can derive these indices for any time during the year (depending on satellite revisit), they are more relevant during the the growing season [@Mishra2015]. Although modeling phenology is a complex task (e.g, ….), satellites offer strategies that help to address these issues [@Younes2021; @Vrieling2018; @Cai2017]. Moreover, land cover dynamics products such as the MCD12Q2 from the USGS [@Friedl2019] provide phenology metrics. Rather than estimating these parameters at a given point in time, some authors have proposed aggregating these indices throughout the season. @Meroni2017 proposes to accumulate the fractional active photosynthetic active radiation (FAPAR) between the start (SOS), and the end of the season (EOS) thus obtaining zCFAPAR as a proxy for biomass productivity. @Zambrano2018 used the same approach but with the NDVI (Normalized Difference Vegetation Index), deriving the zcNDVI for Central Chile. Besides, land use land cover (LULC) change can be driven by drought [@Tran2019; @Akinyemi2021]. To analyze these changes, multiple time-series LULC products exist, such as the MCD12Q1 [@Friedl2019] and the ESA CCI-LC (ESA 2017). The LULC product, together with the vegetation drought index can help evaluate the impact of drought on the ecosystem.

Vegetation drought indices are commonly used as proxies of productivity [@Paruelo2016; @Schucknecht2017]. The main environmental variables that affect productivity are water supply (i.e., precipitation) and demand (i.e., evapotranspiration)[@Mishra2015] commonly collected from weather stations. In developing countries, such as Chile, gaps in historical records are challenging. However,  satellite products can be used to fill these gaps.

To evaluate drought, the World Meteorological Organization (WMO; @WMO2012) has proposed the Standardized Precipitation Index (SPI; @Mckee1993), a multiscalar drought index that allows to monitor precipitation deficit from short- to long-term. For water supply, @Zambrano2017 has already correlated i tInfraRed Precipitation against weather  Station data (CHIRPS; @Funk2015). On the other hand, vegetation biomass productivity also strongly correlates to ET, which in turns depends on the atmospheric evaporative demand (AED) [@FAO66; @Pereira2015; @Allen2005]. Due to increasing trends in air temperatures, AED is increasing, driving a rise in ET [@IPCCCH112021]. But, it is not always true [@Milly2016]. For example, regions where AET is highest have the lowest ET.

The MOD16 product [@Running2021] provides AET and ET satellite estimates and has been used to derive drought indices [@Mu2013]. Soil moisture (SM) is an Essential Climate Variable (ECV) that modulates vegetative growth. The climate change initiative (CCI) from the European Space Agency (ESA) delivers the ESA CCI SM product [@Dorigo2017] (current version 6.1), which has been helpful to monitor drought [@Zhang2019]. Besides, total water storage can be retrieved by the Gravity Recovery and Climate Experiment (GRACE), which allows analyzing water availability changes [@Ahmed2014; @Ma2017]. The balance between water demand and supply by remote sensing can be examined against vegetation productivity, also derived from remote sensing.

We aim to analyze the impact of drought over different types of land cover classes in continental Chile for 2000-2023 by combining environmental variables of biomass productivity and water demand/supply gathered from earth observation products. The specific objective for the study are i) to calculate multi-scalar drought indices for water demand and supply for 1981-2023, ii) to analyze the relationship of a proxy o biomass (zcNDVI) with drought indices, iii) to evaluate LULC change for 2001-2021 and its relation with drought indices,  and  iv) to assess if the observed changes in drought indices and land cover are linked to the TWS.

# Study area

![(a) Chile and the five zones "norte grande", "norte chico",  "zona central", "zona sur", and "zona austral". (b) Topography reference map. (c) Land cover classes for 2021. (d) Persistent land cover classes (> 80%) for 2001-2021.](../output/figs/map_study_con_landcover.png){#fig1_study_area width=80% fig-pos='ht'}

# Materials and Methods

##	Data

###	Earth observation data

###	in-situ data

##	Drought indices for water demand and supply

##	Analysis of a biomass proxy with drought indices of supply and demand

##	LULC change for 2001-2021 and its relation with water supply and demand

![Proportion of land cover class from the persistent land cover for 2001-2021 (>80%) per macrozone.

::: {#fig-elephants layout-ncol=2 fig.pos='!ht'}
![](../output/figs/LC_pers80_per_macrozone.png){width="7cm"}

![](../output/figs/table_var_landcover_macro.png){width="8cm"}

Proportion of land cover class from the persistent land cover for 2001-2021 (>80%) per macrozone. The table on the left shows the linear change trend next to time-series plot of surface, per land cover class (IGBP MCD12Q1.006) for 2001-2021 through the fives zones on continental Chile. Blue dots on the plots indicate the maximum and red dots minimum surface reach.
:::

##	Total water storage and drought indices

Terrestrial Water Storage (TWS) is defined as the total amount of water stored on land that includes any natural or artificial water bodies, such as groundwater, soil moisture, rivers, lakes, snowpack, ice, and biomass water [@Humphrey2023; @Deng2023]. TWS changes evidence the effects of multiple water  fluxes on the hydrological cycle [@Deng2023]. These are reflected in the temporal variation of observations of the Earth's gravitational field [@Abolafia2021; @Sabzehee2023] and, in recent decades, the twin Gravity Recovery and Climate Experiment (GRACE) satellites and their Follow-On mission (GRACE-FO) have provided valuable results on globally distributed TWS anomalies [@Tapley2019;  @Ferreira2023]. The GRACE mission was launched in March 2002 and was operational until October 2017 [@Ramjeawon2022]. Then its GRACE Follow-On successor was launched in May 2018 [@Landerer2020; @Yin2022]. The information provided by these satellites is used to construct monthly maps of the Earth's average gravity field, providing details of the movement of water masses or water mass anomaly estimates relative to the long-term average gravity field [@Humphrey2023; @Wahr2004].

In this study, RL06.1_V3 GRACE mascon (mass concentration) monthly solutions were used, which are provided by the Jet Propulsion Laboratory (JPL-M, https://grace.jpl.nasa.gov). Each GRACE Tellus monthly grid at 0.5 degrees represents the deviation of surface mass for that month relative to a reference time average (2004-2009 baseline), which is subtracted from all other monthly grids to provide terrestrial water storage anomalies (TWSA) [@Ramjeawon2022; @Yin2022]. This JPL-M version of the data employs a Coastal Resolution Improvement (CRI) filter that reduces signal leakage errors across coastlines [@Wiese2019; @Wiese2016]. Although the mascon solutions greatly reduced leakage errors, a gain-factor, which is used to enhance the spatial resolution, was applied to the dataset as recommended for hydrological studies [@Ramjeawon2022; @Yin2022]. The water storage/height anomalies are given in Equivalent Water Height or Thickness units (EWH, cm) [@Sabzehee2023], and its temporal resolution monthly is from April 2002 to May 2023.

## Validation of ERA5-Land variables

#	Results

##	Biomass proxy with drought indices of supply and demand

##	LULC change for 2001-2021 and its relation with water supply and demand

##	Total water storage (TWS) and drought indices

## Validation of ERA5-Land variables

#	Discussion

# Conclusion

# References {-}

### 2023-10-08 article/paper_drought_Chile2023.tex
working third paragraph introduction

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  number,
  preprint,
  3p]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought assessment of the impact of water supply and demand over land cover in continental Chile for 2000-2023 from ERA5-Land and MODIS datasets},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{Drought assessment of the impact of water supply and demand over
land cover in continental Chile for 2000-2023 from ERA5-Land and MODIS
datasets}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl}

\affiliation[1]{organization={Universidad Mayor, Hémera Centro de
Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en
Medio Ambiente y Sustentabilidad},city={Santiago,
Chile},postcode={7500994},postcodesep={}}

\cortext[cor1]{Corresponding author}
\fntext[fn1]{This is the first author footnote.}

\begin{abstract}
Human-induced greenhouse gas emissions have increased the frequency
and/or intensity of weather and climate extremes. Central Chile has been
affected by a persistent drought which is impacting the hydrological
system and vegetation development. The region has been the focus of
research studies due to the diminishing water supply, this persistent
period of water scarcity has been defined as a ``mega drought''.
Nevertheless, our results evidence that the water deficit has expanded
beyond. Our goal is to analyze the impact of drought, measured by
drought indices of water supply/demand and vegetation status, in the
LULCC (land use land cover change) over continental Chile. For the
analysis, continental Chile was divided into five zones according to a
latitudinal gradient: ``Norte Grande'', ``Norte Chico'', ``Zona
Central'', ``Zona Sur'', and ``Zona Austral''. We used monthly climatic
re-analysis variables for precipitation, temperature and soil moisture
for 1981-2023 from ERA5-Land; and MODIS (Moderate-Resolution Imaging
Spectroradiometer) product MCD12Q1 for land cover for 2001-2021, and the
NDVI vegetation index from product MOD13A2 collection 6.1 for 2000-2023,
both from collection 6.1. We estimated atmospheric evaporative demand
(AED) by combining the Hargreaves-Samani equation with the ERA5-Land
temperature. We derived the drought indices SPI (Standardized
Precipitation Index), SPEI (Standardized Precipitation
Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), zcSM
(standardized anomaly of cumulative soil moisture), and the zcNDVI
(standardized anomaly of cumulative NDVI). These indices were calculated
for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1,
3, and 6 months). We analyzed the temporal correlation of SPI, SPEI,
EDDI, and zcSM with zcNDVI to have insights into the impact of water
supply and demand on vegetation. Our results showed that LULCC had an
increasing trend of 412 {[}km2yr−1{]} of forest expansion in the ``Zona
Sur'', together with a decreasing trend of 24 {[}km2yr−1{]} of cropland
contraction in the ``Zona Central'' meanwhile the ``Zona Sur'' showed an
increase of 31 {[}km2yr−1{]}, and a contraction of 80 {[}km2yr−1{]} of
bare soil in the ``Zona Austral''. The EDDI was the less correlated
index for the five macro zones and the five types of land cover, showing
that the temperature in Chile has little impact on vegetation. Higher
r-squared values, between 0.5 and 0.8, were obtained at ``Norte Chico''
and ``Zona Central'' for the land cover types of savanna, shrubland,
grassland, and croplands for the indices SPEI and zcSM at time scales of
12 and 24 months. The forest type reaches a r-squared of
\textasciitilde0.5 for zcSM of 12 months. The results indicate that the
``Norte Chico'' and ``Zona Central'' are the most sensitive regions to
water supply deficits longer than a year, potentially explained by a low
capacity of water storage in those zones that should be further
investigated.
\end{abstract}

\begin{keyword}
    drought \sep land cover change \sep 
    satellite
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The sixth assessment report (AR6) of the IPCC \citep{IPCC2023} indicates
that human-induced greenhouse gas emissions have increased the frequency
and/or intensity of some weather and climate extremes, and the evidence
has been strengthened since AR5 \citep{IPCC2013}. There is high
confidence that increasing global warming can expand the land area
affected by increasing drought frequency and severity
\citep{IPCCCH112021}. Furthermore, drought increases tree mortality and
triggers changes in land cover and, consequently, land use, thus
impacting ecosystems \citep{Crausbay2017}. Nevertheless, there is a lack
of understanding of how the alteration in water supply and demand is
affecting land cover transformations.

Precipitation is the primary driver of drought, which impacts soil
moisture, hydrological regimes, and vegetation productivity. Initially,
drought was commonly classified as meteorological, hydrological, and
agricultural \citep{Wilhite1985}. Lately, \citep{Loon2016} and
\citep{AghaKouchak2021} have given an updated definition of drought for
the Anthropocene, suggesting that it should be considered the feedback
of humans' decisions and activities that drives the anthropogenic
drought. Even though it has been argued that those definitions do not
fully address the ecological dimensions of drought. \citep{Crausbay2017}
proposed the ecological drought definition as \emph{``an episodic
deficit in water availability that drives ecosystems beyond thresholds
of vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems''}. Moreover, many ecological studies have
misinterpreted how to characterize drought, for example, sometimes
considering ``dry'' conditions as ``drought'' \citep{Slette2019}. On the
other hand, the AR6 \citep{IPCC2023} states that even if global warming
is stabilized at 1.5°--2°C, many parts of the world will be impacted by
more severe agricultural and ecological droughts. Then, there is a
challenge in conducting drought research, especially its impact on
ecosystems.

Chile has been facing a persistent rainfall deficit for more than a
decade \citep{Garreaud2017}, which has impacted the hydrological system
\citep{Boisier2018} and consequently the vegetation development
\citep{Zambrano2023}. Current drought conditions have affected crop
yields in Central Chile. Highlighting the growing seasons 2007-2008 and
2008-2009 \citep[\citep{Zambrano2018}]{Zambrano2016}, which impacted an
extensive surface in {[}COMPLETAR{]}. But, in 2019--2020, the drought
intensity reached an extreme condition in North 34°S not seen for at
least 40 years \citep{Zambrano2023}, thus affecting forest, grassland,
and cropland areas. The prolonged lack of precipitation in Central Chile
is producing changes in ecosystem dynamics that must be studied.

Satellite remote sensing \citep{West2019, AghaKouchak2015} is the
primary method to evaluate how meteorological drought impacts vegetation
dynamics. Since the 90's multiple vegetation drought indices have been
developed, such as (VCI,\citep{Kogan1990}; TCI, \citep{Kogan1995};
zNDVI, \citep{Peters2002}; VegDri, \citep{Brown2008}) that have allowed
making spatiotemporal analysis. Although we can derive these indices for
any time during the year (depending on satellite revisit), they are more
relevant during the the growing season \citep{Mishra2015}. Although
modeling phenology is a complex task (e.g, \ldots.), satellites offer
strategies that help to address these issues
\citep{Younes2021, Vrieling2018, Cai2017}. Moreover, land cover dynamics
products such as the MCD12Q2 from the USGS \citep{Friedl2019} provide
phenology metrics. Rather than estimating these parameters at a given
point in time, some authors have proposed aggregating these indices
throughout the season. \citep{Meroni2017} proposes to accumulate the
fractional active photosynthetic active radiation (FAPAR) between the
start (SOS), and the end of the season (EOS) thus obtaining zCFAPAR as a
proxy for biomass productivity. \citep{Zambrano2018} used the same
approach but with the NDVI (Normalized Difference Vegetation Index),
deriving the zcNDVI for Central Chile. Besides, land use land cover
(LULC) change can be driven by drought \citep{Tran2019, Akinyemi2021}.
To analyze these changes, multiple time-series LULC products exist, such
as the MCD12Q1 \citep{Friedl2019} and the ESA CCI-LC (ESA 2017). The
LULC product, together with the vegetation drought index can help
evaluate the impact of drought on the ecosystem.

Vegetation drought indices are commonly used as proxies of productivity
\citep{Paruelo2016, Schucknecht2017}. The main environmental variables
that affect productivity are water supply (i.e., precipitation) and
demand (i.e., evapotranspiration)\citep{Mishra2015} commonly collected
from weather stations. In developing countries, such as Chile, gaps in
historical records are challenging. However, satellite products can be
used to fill these gaps.

To evaluate drought, the World Meteorological Organization (WMO;
\citep{WMO2012}) has proposed the Standardized Precipitation Index (SPI;
\citep{Mckee1993}), a multiscalar drought index that allows to monitor
precipitation deficit from short- to long-term. For water supply,
\citep{Zambrano2017} has already correlated i tInfraRed Precipitation
against weather Station data (CHIRPS; \citep{Funk2015}). On the other
hand, vegetation biomass productivity also strongly correlates to ET,
which in turns depends on the atmospheric evaporative demand (AED)
\citep{FAO66, Pereira2015, Allen2005}. Due to increasing trends in air
temperatures, AED is increasing, driving a rise in ET
\citep{IPCCCH112021}. But, it is not always true \citep{Milly2016}. For
example, regions where AET is highest have the lowest ET.

The MOD16 product \citep{Running2021} provides AET and ET satellite
estimates and has been used to derive drought indices \citep{Mu2013}.
Soil moisture (SM) is an Essential Climate Variable (ECV) that modulates
vegetative growth. The climate change initiative (CCI) from the European
Space Agency (ESA) delivers the ESA CCI SM product \citep{Dorigo2017}
(current version 6.1), which has been helpful to monitor drought
\citep{Zhang2019}. Besides, total water storage can be retrieved by the
Gravity Recovery and Climate Experiment (GRACE), which allows analyzing
water availability changes \citep{Ahmed2014, Ma2017}. The balance
between water demand and supply by remote sensing can be examined
against vegetation productivity, also derived from remote sensing.

We aim to analyze the impact of drought over different types of land
cover classes in continental Chile for 2000-2023 by combining
environmental variables of biomass productivity and water demand/supply
gathered from earth observation products. The specific objective for the
study are i) to calculate multi-scalar drought indices for water demand
and supply for 1981-2023, ii) to analyze the relationship of a proxy o
biomass (zcNDVI) with drought indices, iii) to evaluate LULC change for
2001-2021 and its relation with drought indices, and iv) to assess if
the observed changes in drought indices and land cover are linked to the
TWS.

\hypertarget{study-area}{%
\section{Study area}\label{study-area}}

\begin{figure}[ht]

{\centering \includegraphics[width=0.8\textwidth,height=\textheight]{../output/figs/map_study_con_landcover.png}

}

\caption{(a) Chile and the five zones ``norte grande'', ``norte chico'',
``zona central'', ``zona sur'', and ``zona austral''. (b) Topography
reference map. (c) Land cover classes for 2021. (d) Persistent land
cover classes (\textgreater{} 80\%) for 2001-2021.}

\end{figure}

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

\hypertarget{data}{%
\subsection{Data}\label{data}}

\hypertarget{earth-observation-data}{%
\subsubsection{Earth observation data}\label{earth-observation-data}}

\hypertarget{in-situ-data}{%
\subsubsection{in-situ data}\label{in-situ-data}}

\hypertarget{drought-indices-for-water-demand-and-supply}{%
\subsection{Drought indices for water demand and
supply}\label{drought-indices-for-water-demand-and-supply}}

\hypertarget{analysis-of-a-biomass-proxy-with-drought-indices-of-supply-and-demand}{%
\subsection{Analysis of a biomass proxy with drought indices of supply
and
demand}\label{analysis-of-a-biomass-proxy-with-drought-indices-of-supply-and-demand}}

\hypertarget{lulc-change-for-2001-2021-and-its-relation-with-water-supply-and-demand}{%
\subsection{LULC change for 2001-2021 and its relation with water supply
and
demand}\label{lulc-change-for-2001-2021-and-its-relation-with-water-supply-and-demand}}

!{[}Proportion of land cover class from the persistent land cover for
2001-2021 (\textgreater80\%) per macrozone.

\begin{figure}

\begin{minipage}[t]{0.47\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics[width=7cm,height=\textheight]{../output/figs/LC_pers80_per_macrozone.png}

}

}

\end{minipage}%
%
\begin{minipage}[t]{0.53\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics[width=8cm,height=\textheight]{../output/figs/table_var_landcover_macro.png}

}

}

\end{minipage}%

\caption{\label{fig-elephants}Proportion of land cover class from the
persistent land cover for 2001-2021 (\textgreater80\%) per macrozone.
The table on the left shows the linear change trend next to time-series
plot of surface, per land cover class (IGBP MCD12Q1.006) for 2001-2021
through the fives zones on continental Chile. Blue dots on the plots
indicate the maximum and red dots minimum surface reach.}

\end{figure}

\hypertarget{total-water-storage-and-drought-indices}{%
\subsection{Total water storage and drought
indices}\label{total-water-storage-and-drought-indices}}

Terrestrial Water Storage (TWS) is defined as the total amount of water
stored on land that includes any natural or artificial water bodies,
such as groundwater, soil moisture, rivers, lakes, snowpack, ice, and
biomass water \citep{Humphrey2023, Deng2023}. TWS changes evidence the
effects of multiple water fluxes on the hydrological cycle
\citep{Deng2023}. These are reflected in the temporal variation of
observations of the Earth's gravitational field
\citep{Abolafia2021, Sabzehee2023} and, in recent decades, the twin
Gravity Recovery and Climate Experiment (GRACE) satellites and their
Follow-On mission (GRACE-FO) have provided valuable results on globally
distributed TWS anomalies \citep{Tapley2019, Ferreira2023}. The GRACE
mission was launched in March 2002 and was operational until October
2017 \citep{Ramjeawon2022}. Then its GRACE Follow-On successor was
launched in May 2018 \citep{Landerer2020, Yin2022}. The information
provided by these satellites is used to construct monthly maps of the
Earth's average gravity field, providing details of the movement of
water masses or water mass anomaly estimates relative to the long-term
average gravity field \citep{Humphrey2023, Wahr2004}.

In this study, RL06.1\_V3 GRACE mascon (mass concentration) monthly
solutions were used, which are provided by the Jet Propulsion Laboratory
(JPL-M, https://grace.jpl.nasa.gov). Each GRACE Tellus monthly grid at
0.5 degrees represents the deviation of surface mass for that month
relative to a reference time average (2004-2009 baseline), which is
subtracted from all other monthly grids to provide terrestrial water
storage anomalies (TWSA) \citep{Ramjeawon2022, Yin2022}. This JPL-M
version of the data employs a Coastal Resolution Improvement (CRI)
filter that reduces signal leakage errors across coastlines
\citep{Wiese2019, Wiese2016}. Although the mascon solutions greatly
reduced leakage errors, a gain-factor, which is used to enhance the
spatial resolution, was applied to the dataset as recommended for
hydrological studies \citep{Ramjeawon2022, Yin2022}. The water
storage/height anomalies are given in Equivalent Water Height or
Thickness units (EWH, cm) \citep{Sabzehee2023}, and its temporal
resolution monthly is from April 2002 to May 2023.

\hypertarget{validation-of-era5-land-variables}{%
\subsection{Validation of ERA5-Land
variables}\label{validation-of-era5-land-variables}}

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{biomass-proxy-with-drought-indices-of-supply-and-demand}{%
\subsection{Biomass proxy with drought indices of supply and
demand}\label{biomass-proxy-with-drought-indices-of-supply-and-demand}}

\hypertarget{lulc-change-for-2001-2021-and-its-relation-with-water-supply-and-demand-1}{%
\subsection{LULC change for 2001-2021 and its relation with water supply
and
demand}\label{lulc-change-for-2001-2021-and-its-relation-with-water-supply-and-demand-1}}

\hypertarget{total-water-storage-tws-and-drought-indices}{%
\subsection{Total water storage (TWS) and drought
indices}\label{total-water-storage-tws-and-drought-indices}}

\hypertarget{validation-of-era5-land-variables-1}{%
\subsection{Validation of ERA5-Land
variables}\label{validation-of-era5-land-variables-1}}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\renewcommand\refname{References}
  \bibliography{references.bib}

\end{document}

### 2023-10-08 article_bckp/article.Rmd
working third paragraph introduction

--- 2023-05-30 article/article.Rmd
@@
 ---
-title: Comprehensive assessment of the climate-induced water scarcity over continental Chile
+title: "Comprehensive assessment of the climate-induced water scarcity over continental
+  Chile"
 author:
@@
   affiliation: Hemera Centro de Observación de la Tierra,Universidad Mayor
-    correspondingauthor: true
+  correspondingauthor: yes
   footnote: 1
@@
   footnote: 2
+date: "`r Sys.Date()`"
+output:
+  word_document:
+    number_sections: yes
+    toc: yes
+  rticles::elsevier_article:
+    keep_tex: yes
+    citation_package: natbib
+abstract: "Human-induced greenhouse gas emissions have increased the frequency and/or
+  intensity of some weather and climate extremes globally. Chile has been affected
+  by persistent water scarcity which is impacting the hydrological system and vegetation
+  development. Central Chile it is been the focus of research studies due to the diminishing
+  water supply, nevertheless our results evidence that water deficit is expanded beyond.
+  We analyze earth observation data for 2000-2023 to make a comprehensive assessment
+  of water scarcity in Chile.  For the analysis, continental Chile was divided into
+  five zones big north, little north, central, south, and austral zone.  We used the
+  time series of MODIS for land cover change (LULC), fraction of vegetation cover
+  (FVC), and to derive the drought index zcNDVI and zcET (standardized anomaly of
+  cumulative NDVI and ET through the growing season); and CHELSA v2.1 dataset to derive
+  the standardized precipitation evapotranspiration index (SPEI). We evaluate the
+  interconnection between the productivity drought index (zcNDVI) and water demand
+  (zcET), supply (SPEIs) and FVC. Finally, we analyzed the temporal correlation of
+  the drought indices with total water storage (TWS) from GRACE (Gravity Recovery
+  and Climate Experiment). Our LULC results showed a increasing trend of 412 $[km^2\\cdot
+  yr^{-1}$] of forest in the south zone, a decresasing trend of 24 $[km^2\\cdot yr^{-1}$]
+  \ of cropland in the central zone and an increase of 31 $[km^2\\cdot yr^{-1}$] in
+  the south zone, and a diminish of 80 $[km^2\\cdot yr^{-1}$] of barrend land in the
+  austral zone. \n"
+keywords:
+- keyword1
+- keyword2
+journal: Remote Sensing of Environment
 address:
@@
   country: The Netherlands
+classoption: preprint, 3p
+bibliography: references.bib
+biblio-style: apalike
+linenumbers: no
+numbersections: yes
 footnote:
 - code: 1
-    text: "This is the first author footnote."
+  text: This is the first author footnote.
 - code: 2
-    text: "Another author footnote."
-abstract: |
-  Human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes globally.
-Chile has been affected by persistent water scarcity which is impacting the hydrological system and vegetation development.
-Central Chile it is been the focus of research studies due to the diminishing water supply, nevertheless water deficit is expanded beyond.
-We analyze earth observation data for 2000-2023 to do a comprehensive assessment of water scarcity in Chile.
-For the analysis, continental Chile was divided into five zones from north to south.
-We used the time series of MODIS for land cover change (LULC), fraction of vegetation cover (FVC), and to derive the drought index zcNDVI and zcET (standardized anomaly of cumulative NDVI and ET through the growing season); and CHELSA v2.1 dataset to derive the standardized precipitation evapotranspiration index (SPEI).
-We evaluate the interconnection between the productivity drought index (zcNDVI) and water demand (zcET), supply (SPEIs) and FVC.
-Finally, we analyzed the temporal correlation of the drought indices with total water storage (TWS) from GRACE (Gravity Recovery and Climate Experiment).
-Our results showed that ...
-
-keywords: 
-  - keyword1
-  - keyword2
-journal: "Remote Sensing of Environment"
-date: "`r Sys.Date()`"
-classoption: preprint, 3p
-bibliography: references.bib
-biblio-style: "apalike"
-linenumbers: false
-numbersections: true
-# csl: https://www.zotero.org/styles/elsevier-harvard
-output: 
-  rticles::elsevier_article:
-    keep_tex: true
-    citation_package: natbib
-  word_document: default
+  text: Another author footnote.
 ---
@@
 
-# Version
-
-This Rmd-skeleton uses the mdpi Latex template published 2019/02. 
-However, the official template gets more frequently updated than the 'rticles'
-package. Therefore, please make sure prior to paper submission, that you're 
-using the most recent .cls, .tex and .bst files 
-(available [here](http://www.mdpi.com/authors/latex)).
-
 # Introduction
 
-The sixth assessment report (AR6) from the working group I of the IPCC [@IPCC2021] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes and the evidence has been strengthened since AR5 [@IPCC2013].
+The sixth assessment report (AR6) of the IPCC [@IPCC2021] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes and the evidence has been strengthened since AR5 [@IPCC2013].
 There is high confidence that the increasing global warning can expand the land area affected by increasing drought frequency and severity [@IPCCCH112021].
@@
 But, in 2019-2020, the drought intensity reached an extreme condition at North 34°S  not seen -at least- for more than 40 years [@Zambrano2020], affecting forest, grassland, and croplands areas.
-The prolonged lack of precipitation within Central Chile is producing changes in the ecosystem that should study.
+The prolonged lack of precipitation within Central Chile is producing changes in the ecosystem that must be studied.
 
@@
 To analyze those changes, multiple time-series LULC products exist as the MCD12Q1 [@Friedl2019] and the ESA CCI-LC [@ESA2017].
-For Chile, 2014 was made a high-resolution land cover at 30m of spatial resolution [@Zhao2016].
 The LULC product with the vegetation drought index can help evaluate the impact of drought on the ecosystem.

### 2023-10-08 article_bckp/article.tex
working third paragraph introduction

--- 2023-05-30 article/article.tex
@@
   been the focus of research studies due to the diminishing water
-  supply, nevertheless water deficit is expanded beyond. We analyze
-  earth observation data for 2000-2023 to do a comprehensive assessment
-  of water scarcity in Chile. For the analysis, continental Chile was
-  divided into five zones from north to south. We used the time series
-  of MODIS for land cover change (LULC), fraction of vegetation cover
-  (FVC), and to derive the drought index zcNDVI and zcET (standardized
-  anomaly of cumulative NDVI and ET through the growing season); and
-  CHELSA v2.1 dataset to derive the standardized precipitation
-  evapotranspiration index (SPEI). We evaluate the interconnection
-  between the productivity drought index (zcNDVI) and water demand
-  (zcET), supply (SPEIs) and FVC. Finally, we analyzed the temporal
-  correlation of the drought indices with total water storage (TWS) from
-  GRACE (Gravity Recovery and Climate Experiment). Our results showed
-  that \ldots{}
+  supply, nevertheless our results evidence that water deficit is
+  expanded beyond. We analyze earth observation data for 2000-2023 to
+  make a comprehensive assessment of water scarcity in Chile. For the
+  analysis, continental Chile was divided into five zones big north,
+  little north, central, south, and austral zone. We used the time
+  series of MODIS for land cover change (LULC), fraction of vegetation
+  cover (FVC), and to derive the drought index zcNDVI and zcET
+  (standardized anomaly of cumulative NDVI and ET through the growing
+  season); and CHELSA v2.1 dataset to derive the standardized
+  precipitation evapotranspiration index (SPEI). We evaluate the
+  interconnection between the productivity drought index (zcNDVI) and
+  water demand (zcET), supply (SPEIs) and FVC. Finally, we analyzed the
+  temporal correlation of the drought indices with total water storage
+  (TWS) from GRACE (Gravity Recovery and Climate Experiment). Our LULC
+  results showed a increasing trend of 412 \([km^2\cdot yr^{-1}\){]} of
+  forest in the south zone, a decresasing trend of 24
+  \([km^2\cdot yr^{-1}\){]} of cropland in the central zone and an
+  increase of 31 \([km^2\cdot yr^{-1}\){]} in the south zone, and a
+  diminish of 80 \([km^2\cdot yr^{-1}\){]} of barrend land in the
+  austral zone.
   \end{abstract}
@@
 
-\hypertarget{version}{%
-\section{Version}\label{version}}
-
-This Rmd-skeleton uses the mdpi Latex template published 2019/02.
-However, the official template gets more frequently updated than the
-`rticles' package. Therefore, please make sure prior to paper
-submission, that you're using the most recent .cls, .tex and .bst files
-(available \href{http://www.mdpi.com/authors/latex}{here}).
-
 \hypertarget{introduction}{%
@@
 
-The sixth assessment report (AR6) from the working group I of the IPCC
-\citep{IPCC2021} indicates that human-induced greenhouse gas emissions
-have increased the frequency and/or intensity of some weather and
-climate extremes and the evidence has been strengthened since AR5
-\citep{IPCC2013}. There is high confidence that the increasing global
-warning can expand the land area affected by increasing drought
-frequency and severity \citep{IPCCCH112021}. Chile has been facing a
-persistent rainfall deficit lasting for more than ten years
-\citep{Garreaud2017} which has impacted the hydrological system
-\citep{Boisier2018}, and consequently the vegetation development
-\citep{Zambrano2020}.
+The sixth assessment report (AR6) of the IPCC \citep{IPCC2021} indicates
+that human-induced greenhouse gas emissions have increased the frequency
+and/or intensity of some weather and climate extremes and the evidence
+has been strengthened since AR5 \citep{IPCC2013}. There is high
+confidence that the increasing global warning can expand the land area
+affected by increasing drought frequency and severity
+\citep{IPCCCH112021}. Chile has been facing a persistent rainfall
+deficit lasting for more than ten years \citep{Garreaud2017} which has
+impacted the hydrological system \citep{Boisier2018}, and consequently
+the vegetation development \citep{Zambrano2020}.
 
@@
 The prolonged lack of precipitation within Central Chile is producing
-changes in the ecosystem that should study.
+changes in the ecosystem that must be studied.
 
@@
 multiple time-series LULC products exist as the MCD12Q1
-\citep{Friedl2019} and the ESA CCI-LC \citep{ESA2017}. For Chile, 2014
-was made a high-resolution land cover at 30m of spatial resolution
-\citep{Zhao2016}. The LULC product with the vegetation drought index can
-help evaluate the impact of drought on the ecosystem.
+\citep{Friedl2019} and the ESA CCI-LC \citep{ESA2017}. The LULC product
+with the vegetation drought index can help evaluate the impact of
+drought on the ecosystem.
 
@@
 \midrule
-norte grande & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70937146e24b.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709384faf70.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709373919fc5.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709321ad9f19.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709379ea79bd.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093c8ff3c5.pdf}\\
-norte chico & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709337b6241e.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70937f0ed48a.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70934b4bd377.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70931e4fd155.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70932e8b190a.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093305c1274.pdf}\\
-zona central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70932bbec995.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093598f66c9.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709370472832.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093ba3805d.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709341ec566.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70933b82960c.pdf}\\
-zona sur & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70936a87e29b.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70935b745f29.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709363a8f17a.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70932c57d58c.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70935fdd829a.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709328c1f26f.pdf}\\
-zona austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709343d0ce35.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d7093615c29b7.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70933872723d.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709369bca642.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d709386fdff0.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_1d70931a654519.pdf}\\
+big north & -2.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d410cb31f.pdf} & 0.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d4ddd8049.pdf} & -31.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d7c9fbffd.pdf} & 32.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d1a8bb73a.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d474fdeb5.pdf} & NA\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d269510e4.pdf}\\
+little north & 79.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d5d88be43.pdf} & -66.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d150a4cbe.pdf} & -100.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d404fa544.pdf} & 104.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70dca4c474.pdf} & 0.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d7e3728ac.pdf} & -13.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d4a63e6a4.pdf}\\
+central & 130.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d25a3233e.pdf} & -128.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d57d04753.pdf} & 89.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d42f71f69.pdf} & 23.3\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d1dec138d.pdf} & -66.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d1c2543ab.pdf} & -24.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d8528d9c.pdf}\\
+south & -14.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d20c19221.pdf} & -316.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d2c19997d.pdf} & -55.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d3e82c025.pdf} & 2.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d5b7a821e.pdf} & 412.4\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d79c7dc48.pdf} & 30.8\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d41d72265.pdf}\\
+austral & -44.6\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d435b8d.pdf} & 163.9\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d26854a6b.pdf} & 226.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d5a94aab.pdf} & -80.2\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d53a33e6a.pdf} & -9.1\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70d14d93133.pdf} & -1.0\includegraphics[width=0.67in, height=0.17in]{article_files/figure-latex//plot_ee70deb41525.pdf}\\
 \bottomrule

### 2023-10-09 article/paper_drought_Chile2023.qmd
first draft of the introduction

@@
 ---
-title: Drought assessment of the impact of water supply and demand over land cover in continental Chile for 2000-2023 from ERA5-Land and MODIS datasets
+title: Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from the ERA5-Land and MODIS
 #subtitle: A Short Subtitle
@@
       natbiboptions: numbers
+    include-in-header: 
+      text: '\usepackage{lineno}\linenumbers'
       #cite-style: super
@@
 
-Precipitation is the primary driver of drought, which impacts soil moisture, hydrological regimes, and vegetation productivity.
+Precipitation is the primary driver of drought and is intensified by temperature [@Luo2017].
+Drought impacts soil moisture, hydrological regimes, and vegetation productivity.
 Initially, drought was commonly classified as meteorological, hydrological, and agricultural [@Wilhite1985].
@@
 On the other hand, the AR6 [@IPCC2023] states that even if global warming is stabilized at 1.5°–2°C, many parts of the world will be impacted by more severe agricultural and ecological droughts.
-Then, there is a challenge in conducting drought research, especially its impact on ecosystems.
+Then, there is a challenge in conducting drought research, especially to evaluate its impact on ecosystems.
 
-Chile has been facing a persistent rainfall deficit for more than a decade [@Garreaud2017], which has impacted the the vegetation development [@Zambrano2023] and hydrological system [@Boisier2018].
-Current drought conditions have affected crop yields in Central Chile.
-Highlighting the growing seasons 2007-2008 and 2008-2009 [@Zambrano2016, @Zambrano2018], which impacted an extensive surface in [COMPLETAR].
-But, in 2019–2020, the drought intensity reached an extreme condition in North 34°S not seen for at least 40 years [@Zambrano2023], thus affecting forest, grassland, and cropland areas.
+Chile has been facing a persistent rainfall deficit for more than a decade [@Garreaud2017], which has impacted vegetation development [@Zambrano2023] and the hydrological system [@Boisier2018].
+Current drought conditions have affected crop productivity [@Zambrano2016; @Zambrano2018], forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], land cover change [@Fuentes2021], water supply in watersheds [@AlvarezGarreton2021], and have had economic impacts [@Fernandez2023].
+In 2019–2020, the drought severity reached an extreme condition in Central Chile (30–34°S) not seen for at least 40 years, and the evidence indicates that the impact is transversal to the land cover classes of forest, grassland, and cropland [@Zambrano2023].
 The prolonged lack of precipitation in Central Chile is producing changes in ecosystem dynamics that must be studied.
 
-Satellite remote sensing [@West2019; @AghaKouchak2015] is the primary method to evaluate how meteorological drought impacts vegetation dynamics.
-Since the 90’s multiple vegetation drought indices have been developed, such as (VCI,@Kogan1990; TCI, @Kogan1995; zNDVI, [@Peters2002]; VegDri, [@Brown2008]) that have allowed making spatiotemporal analysis.
-Although we can derive these indices for any time during the year (depending on satellite revisit), they are more relevant during the the growing season [@Mishra2015].
-Although modeling phenology is a complex task (e.g, ….), satellites offer strategies that help to address these issues [@Younes2021; @Vrieling2018; @Cai2017].
-Moreover, land cover dynamics products such as the MCD12Q2 from the USGS [@Friedl2019] provide phenology metrics.
-Rather than estimating these parameters at a given point in time, some authors have proposed aggregating these indices throughout the season.
-@Meroni2017 proposes to accumulate the fractional active photosynthetic active radiation (FAPAR) between the start (SOS), and the end of the season (EOS) thus obtaining zCFAPAR as a proxy for biomass productivity.
-@Zambrano2018 used the same approach but with the NDVI (Normalized Difference Vegetation Index), deriving the zcNDVI for Central Chile.
-Besides, land use land cover (LULC) change can be driven by drought [@Tran2019; @Akinyemi2021].
-To analyze these changes, multiple time-series LULC products exist, such as the MCD12Q1 [@Friedl2019] and the ESA CCI-LC (ESA 2017).
-The LULC product, together with the vegetation drought index can help evaluate the impact of drought on the ecosystem.
+For the spatiotemporal assessment of drought impact (i.e., by water supply and demand) on land cover changes, we need climatic realiable variables such as precipitation, temperature, soil moisture, land cover, and vegetation status.
+For developing countries like Chile, the weather networks present several disadvantages, such as gaps, a short history, and low-quality data.
+Reanalysis data, as the ERA5-Land (ERA5L) [@MunozSabater2021] provides hourly climatic information (precipitation, temperature, and soil moisture) without gaps since 1950 with global extension.
+ERA5L has already been used for drought assessment using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@Nouri2023] and for flash drought [@Wang2023] by analyzing soil mositure and evapotranspiration.
+On the other hand, satellite remote sensing [@West2019; @AghaKouchak2015] is the primary method to evaluate how drought impacts vegetation dynamics.
+Vegetation drought indices (VDI) are commonly used as proxies of productivity [@Paruelo2016; @Schucknecht2017], which can be derived from the MODIS (Moderate-Resolution Imaging Spectroradiometer).
+Besides, land use and land cover (LULC) change can be driven by drought [@Tran2019; @Akinyemi2021].
+To analyze these changes, multiple LULC products exist [@Grekousis2015], one of those that provides time series since 2001 is the MCD12Q1 [@Friedl2019] from MODIS.
+The variation in water supply and demand is finally reflected in the total water storage (TWS).
+The TWS can be retrieved by the Gravity Recovery and Climate Experiment (GRACE), which allows analyzing water availability changes at coarse resolution [@Ahmed2014; @Ma2017].
+We can use climatic reanalysis (ERA5L) and vegetation data (MODIS) to derive drought indices of supply (i.e., precipitation) and demand (i.e., temperature) and thus evaluate the impact of drought on LULC changes.
+Further, the TWS can be assessed with regard to the changes in water supply and demand to gain insight into the impact on water storage.
 
-Vegetation drought indices are commonly used as proxies of productivity [@Paruelo2016; @Schucknecht2017].
-The main environmental variables that affect productivity are water supply (i.e., precipitation) and demand (i.e., evapotranspiration)[@Mishra2015] commonly collected from weather stations.
-In developing countries, such as Chile, gaps in historical records are challenging.
-However,  satellite products can be used to fill these gaps.
+To evaluate meteorological drought (i.e., water supply), the World Meteorological Organization (WMO; @WMO2012) recommends the Standardized Precipitation Index (SPI; @Mckee1993), a multiscalar drought index that allows to monitor precipitation deficits from short- to long-term.
+Following the same approach, @Vicente-Serrano2010 incorporates into the SPI the effect of temperature through the use of potential evapotranspiration, thus proposing the SPEI (Standardized Precipitation Evapotranspiration Index).
+Similarly, to evaluate solely the evaporative demand driven by temperature, @Hobbins2016 and @McEvoy2016 came up with the Evaporative Demand Drought Index (EDDI).
+For vegetation, in a similar manner as the SPI, SPEI and EDDI;  @Zambrano2018 proposed the zcNDVI, a standardized anomaly of the cumulative Normalized Difference Vegetation Index (NDVI), which could be acumulated over the growing season or any period (e.g., months), resulting in a multiscalar drought index.
+For soil moisture, several drought indices exist, such as the Soil Moisture Deficit Index (SDMI) a normalized index [@Narasimhan2005] and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021] which is a normalized index using vegetation, land surface temperature, and a vegetation condition index (VCI, [@Kogan1995]).
+From TWS, we can estimate the standardized terrestrial water storage index (STI) [@Cui2021], a standardized anomaly that follows the methodology of the SPI, SPEI, EDDI, and zcNDVI.
+Thereby, we have drought indices for water supply, demand, and storage, which can help to make a comprehensive assessment of drought.
 
-To evaluate drought, the World Meteorological Organization (WMO; @WMO2012) has proposed the Standardized Precipitation Index (SPI; @Mckee1993), a multiscalar drought index that allows to monitor precipitation deficit from short- to long-term.
-For water supply, @Zambrano2017 has already correlated i tInfraRed Precipitation against weather  Station data (CHIRPS; @Funk2015).
-On the other hand, vegetation biomass productivity also strongly correlates to ET, which in turns depends on the atmospheric evaporative demand (AED) [@FAO66; @Pereira2015; @Allen2005].
-Due to increasing trends in air temperatures, AED is increasing, driving a rise in ET [@IPCCCH112021].
-But, it is not always true [@Milly2016].
-For example, regions where AET is highest have the lowest ET.
-
-The MOD16 product [@Running2021] provides AET and ET satellite estimates and has been used to derive drought indices [@Mu2013].
-Soil moisture (SM) is an Essential Climate Variable (ECV) that modulates vegetative growth.
-The climate change initiative (CCI) from the European Space Agency (ESA) delivers the ESA CCI SM product [@Dorigo2017] (current version 6.1), which has been helpful to monitor drought [@Zhang2019].
-Besides, total water storage can be retrieved by the Gravity Recovery and Climate Experiment (GRACE), which allows analyzing water availability changes [@Ahmed2014; @Ma2017].
-The balance between water demand and supply by remote sensing can be examined against vegetation productivity, also derived from remote sensing.
-
-We aim to analyze the impact of drought over different types of land cover classes in continental Chile for 2000-2023 by combining environmental variables of biomass productivity and water demand/supply gathered from earth observation products.
-The specific objective for the study are i) to calculate multi-scalar drought indices for water demand and supply for 1981-2023, ii) to analyze the relationship of a proxy o biomass (zcNDVI) with drought indices, iii) to evaluate LULC change for 2001-2021 and its relation with drought indices,  and  iv) to assess if the observed changes in drought indices and land cover are linked to the TWS.
+In this research, we aim to analyze the impact of drought on different types of land cover classes in continental Chile by combining environmental variables such as biomass productivity, and water demand and supply gathered from earth observation products.
+The specific objectives of the study are: i) to calculate multi-scalar drought indices for water demand and supply for 1981–2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of biomass (zcNDVI) with drought indices; and iv) to assess if the observed changes in the drought indices are linked to the TWS.
 
@@
 This JPL-M version of the data employs a Coastal Resolution Improvement (CRI) filter that reduces signal leakage errors across coastlines [@Wiese2019; @Wiese2016].
-Although the mascon solutions greatly reduced leakage errors, a gain-factor, which is used to enhance the spatial resolution, was applied to the dataset as recommended for hydrological studies [@Ramjeawon2022; @Yin2022].
-The water storage/height anomalies are given in Equivalent Water Height or Thickness units (EWH, cm) [@Sabzehee2023], and its temporal resolution monthly is from April 2002 to May 2023.
+Although the mascon solutions greatly reduced leakage errors, a gain factor, which is used to enhance the spatial resolution, was applied to the dataset as recommended for hydrological studies [@Ramjeawon2022; @Yin2022].
+The water storage and height anomalies are given in Equivalent Water Height or Thickness units (EWH, cm) [@Sabzehee2023], and their temporal resolution monthly is from April 2002 to May 2023.
 
@@
 #	Results
+
+##	Drought indices for water demand and supply
 

### 2023-10-09 article/paper_drought_Chile2023.tex
first draft of the introduction

@@
 
+\usepackage{lineno}\linenumbers
 \makeatletter
@@
 \hypersetup{
-  pdftitle={Drought assessment of the impact of water supply and demand over land cover in continental Chile for 2000-2023 from ERA5-Land and MODIS datasets},
+  pdftitle={Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from the ERA5-Land and MODIS},
   pdfauthor={Francisco Zambrano},
@@
 \begin{frontmatter}
-\title{Drought assessment of the impact of water supply and demand over
-land cover in continental Chile for 2000-2023 from ERA5-Land and MODIS
-datasets}
+\title{Assessment of drought impact over land cover in continental Chile
+by the analysis of water supply and demand from the ERA5-Land and MODIS}
 \author[1]{Francisco Zambrano%
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
-Precipitation is the primary driver of drought, which impacts soil
-moisture, hydrological regimes, and vegetation productivity. Initially,
-drought was commonly classified as meteorological, hydrological, and
-agricultural \citep{Wilhite1985}. Lately, \citep{Loon2016} and
+Precipitation is the primary driver of drought and is intensified by
+temperature \citep{Luo2017}. Drought impacts soil moisture, hydrological
+regimes, and vegetation productivity. Initially, drought was commonly
+classified as meteorological, hydrological, and agricultural
+\citep{Wilhite1985}. Lately, \citep{Loon2016} and
 \citep{AghaKouchak2021} have given an updated definition of drought for
@@
 more severe agricultural and ecological droughts. Then, there is a
-challenge in conducting drought research, especially its impact on
-ecosystems.
+challenge in conducting drought research, especially to evaluate its
+impact on ecosystems.
 
 Chile has been facing a persistent rainfall deficit for more than a
-decade \citep{Garreaud2017}, which has impacted the hydrological system
-\citep{Boisier2018} and consequently the vegetation development
-\citep{Zambrano2023}. Current drought conditions have affected crop
-yields in Central Chile. Highlighting the growing seasons 2007-2008 and
-2008-2009 \citep[\citep{Zambrano2018}]{Zambrano2016}, which impacted an
-extensive surface in {[}COMPLETAR{]}. But, in 2019--2020, the drought
-intensity reached an extreme condition in North 34°S not seen for at
-least 40 years \citep{Zambrano2023}, thus affecting forest, grassland,
-and cropland areas. The prolonged lack of precipitation in Central Chile
-is producing changes in ecosystem dynamics that must be studied.
+decade \citep{Garreaud2017}, which has impacted vegetation development
+\citep{Zambrano2023} and the hydrological system \citep{Boisier2018}.
+Current drought conditions have affected crop productivity
+\citep{Zambrano2016, Zambrano2018}, forest development
+\citep{Miranda2020, Venegas2018}, forest fire occurrence
+\citep{UrrutiaJalabert2018}, land cover change \citep{Fuentes2021},
+water supply in watersheds \citep{AlvarezGarreton2021}, and have had
+economic impacts \citep{Fernandez2023}. In 2019--2020, the drought
+severity reached an extreme condition in Central Chile (30--34°S) not
+seen for at least 40 years, and the evidence indicates that the impact
+is transversal to the land cover classes of forest, grassland, and
+cropland \citep{Zambrano2023}. The prolonged lack of precipitation in
+Central Chile is producing changes in ecosystem dynamics that must be
+studied.
 
-Satellite remote sensing \citep{West2019, AghaKouchak2015} is the
-primary method to evaluate how meteorological drought impacts vegetation
-dynamics. Since the 90's multiple vegetation drought indices have been
-developed, such as (VCI,\citep{Kogan1990}; TCI, \citep{Kogan1995};
-zNDVI, \citep{Peters2002}; VegDri, \citep{Brown2008}) that have allowed
-making spatiotemporal analysis. Although we can derive these indices for
-any time during the year (depending on satellite revisit), they are more
-relevant during the the growing season \citep{Mishra2015}. Although
-modeling phenology is a complex task (e.g, \ldots.), satellites offer
-strategies that help to address these issues
-\citep{Younes2021, Vrieling2018, Cai2017}. Moreover, land cover dynamics
-products such as the MCD12Q2 from the USGS \citep{Friedl2019} provide
-phenology metrics. Rather than estimating these parameters at a given
-point in time, some authors have proposed aggregating these indices
-throughout the season. \citep{Meroni2017} proposes to accumulate the
-fractional active photosynthetic active radiation (FAPAR) between the
-start (SOS), and the end of the season (EOS) thus obtaining zCFAPAR as a
-proxy for biomass productivity. \citep{Zambrano2018} used the same
-approach but with the NDVI (Normalized Difference Vegetation Index),
-deriving the zcNDVI for Central Chile. Besides, land use land cover
-(LULC) change can be driven by drought \citep{Tran2019, Akinyemi2021}.
-To analyze these changes, multiple time-series LULC products exist, such
-as the MCD12Q1 \citep{Friedl2019} and the ESA CCI-LC (ESA 2017). The
-LULC product, together with the vegetation drought index can help
-evaluate the impact of drought on the ecosystem.
+For the spatiotemporal assessment of drought impact (i.e., by water
+supply and demand) on land cover changes, we need climatic realiable
+variables such as precipitation, temperature, soil moisture, land cover,
+and vegetation status. For developing countries like Chile, the weather
+networks present several disadvantages, such as gaps, a short history,
+and low-quality data. Reanalysis data, as the ERA5-Land (ERA5L)
+\citep{MunozSabater2021} provides hourly climatic information
+(precipitation, temperature, and soil moisture) without gaps since 1950
+with global extension. ERA5L has already been used for drought
+assessment using the Standardized Precipitation-Evapotranspiration Index
+(SPEI) \citep{Nouri2023} and for flash drought \citep{Wang2023} by
+analyzing soil mositure and evapotranspiration. On the other hand,
+satellite remote sensing \citep{West2019, AghaKouchak2015} is the
+primary method to evaluate how drought impacts vegetation dynamics.
+Vegetation drought indices (VDI) are commonly used as proxies of
+productivity \citep{Paruelo2016, Schucknecht2017}, which can be derived
+from the MODIS (Moderate-Resolution Imaging Spectroradiometer). Besides,
+land use and land cover (LULC) change can be driven by drought
+\citep{Tran2019, Akinyemi2021}. To analyze these changes, multiple LULC
+products exist \citep{Grekousis2015}, one of those that provides time
+series since 2001 is the MCD12Q1 \citep{Friedl2019} from MODIS. The
+variation in water supply and demand is finally reflected in the total
+water storage (TWS). The TWS can be retrieved by the Gravity Recovery
+and Climate Experiment (GRACE), which allows analyzing water
+availability changes at coarse resolution \citep{Ahmed2014, Ma2017}. We
+can use climatic reanalysis (ERA5L) and vegetation data (MODIS) to
+derive drought indices of supply (i.e., precipitation) and demand (i.e.,
+temperature) and thus evaluate the impact of drought on LULC changes.
+Further, the TWS can be assessed with regard to the changes in water
+supply and demand to gain insight into the impact on water storage.
 
-Vegetation drought indices are commonly used as proxies of productivity
-\citep{Paruelo2016, Schucknecht2017}. The main environmental variables
-that affect productivity are water supply (i.e., precipitation) and
-demand (i.e., evapotranspiration)\citep{Mishra2015} commonly collected
-from weather stations. In developing countries, such as Chile, gaps in
-historical records are challenging. However, satellite products can be
-used to fill these gaps.
+To evaluate meteorological drought (i.e., water supply), the World
+Meteorological Organization (WMO; \citep{WMO2012}) recommends the
+Standardized Precipitation Index (SPI; \citep{Mckee1993}), a multiscalar
+drought index that allows to monitor precipitation deficits from short-
+to long-term. Following the same approach, \citep{Vicente-Serrano2010}
+incorporates into the SPI the effect of temperature through the use of
+potential evapotranspiration, thus proposing the SPEI (Standardized
+Precipitation Evapotranspiration Index). Similarly, to evaluate solely
+the evaporative demand driven by temperature, \citep{Hobbins2016} and
+\citep{McEvoy2016} came up with the Evaporative Demand Drought Index
+(EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI;
+~\citep{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
+cumulative Normalized Difference Vegetation Index (NDVI), which could be
+acumulated over the growing season or any period (e.g., months),
+resulting in a multiscalar drought index. For soil moisture, several
+drought indices exist, such as the Soil Moisture Deficit Index (SDMI) a
+normalized index \citep{Narasimhan2005} and the Soil Moisture
+Agricultural Drought Index (SMADI) \citep{Souza2021} which is a
+normalized index using vegetation, land surface temperature, and a
+vegetation condition index (VCI, \citep{Kogan1995}). From TWS, we can
+estimate the standardized terrestrial water storage index (STI)
+\citep{Cui2021}, a standardized anomaly that follows the methodology of
+the SPI, SPEI, EDDI, and zcNDVI. Thereby, we have drought indices for
+water supply, demand, and storage, which can help to make a
+comprehensive assessment of drought.
 
-To evaluate drought, the World Meteorological Organization (WMO;
-\citep{WMO2012}) has proposed the Standardized Precipitation Index (SPI;
-\citep{Mckee1993}), a multiscalar drought index that allows to monitor
-precipitation deficit from short- to long-term. For water supply,
-\citep{Zambrano2017} has already correlated i tInfraRed Precipitation
-against weather Station data (CHIRPS; \citep{Funk2015}). On the other
-hand, vegetation biomass productivity also strongly correlates to ET,
-which in turns depends on the atmospheric evaporative demand (AED)
-\citep{FAO66, Pereira2015, Allen2005}. Due to increasing trends in air
-temperatures, AED is increasing, driving a rise in ET
-\citep{IPCCCH112021}. But, it is not always true \citep{Milly2016}. For
-example, regions where AET is highest have the lowest ET.
-
-The MOD16 product \citep{Running2021} provides AET and ET satellite
-estimates and has been used to derive drought indices \citep{Mu2013}.
-Soil moisture (SM) is an Essential Climate Variable (ECV) that modulates
-vegetative growth. The climate change initiative (CCI) from the European
-Space Agency (ESA) delivers the ESA CCI SM product \citep{Dorigo2017}
-(current version 6.1), which has been helpful to monitor drought
-\citep{Zhang2019}. Besides, total water storage can be retrieved by the
-Gravity Recovery and Climate Experiment (GRACE), which allows analyzing
-water availability changes \citep{Ahmed2014, Ma2017}. The balance
-between water demand and supply by remote sensing can be examined
-against vegetation productivity, also derived from remote sensing.
-
-We aim to analyze the impact of drought over different types of land
-cover classes in continental Chile for 2000-2023 by combining
-environmental variables of biomass productivity and water demand/supply
-gathered from earth observation products. The specific objective for the
-study are i) to calculate multi-scalar drought indices for water demand
-and supply for 1981-2023, ii) to analyze the relationship of a proxy o
-biomass (zcNDVI) with drought indices, iii) to evaluate LULC change for
-2001-2021 and its relation with drought indices, and iv) to assess if
-the observed changes in drought indices and land cover are linked to the
-TWS.
+In this research, we aim to analyze the impact of drought on different
+types of land cover classes in continental Chile by combining
+environmental variables such as biomass productivity, and water demand
+and supply gathered from earth observation products. The specific
+objectives of the study are: i) to calculate multi-scalar drought
+indices for water demand and supply for 1981--2023; ii) to evaluate LULC
+change for 2001--2021 and its relation to drought indices; iii) to
+analyze the relationship of a proxy of biomass (zcNDVI) with drought
+indices; and iv) to assess if the observed changes in the drought
+indices are linked to the TWS.
 
@@
 \citep{Wiese2019, Wiese2016}. Although the mascon solutions greatly
-reduced leakage errors, a gain-factor, which is used to enhance the
+reduced leakage errors, a gain factor, which is used to enhance the
 spatial resolution, was applied to the dataset as recommended for
-hydrological studies \citep{Ramjeawon2022, Yin2022}. The water
-storage/height anomalies are given in Equivalent Water Height or
-Thickness units (EWH, cm) \citep{Sabzehee2023}, and its temporal
-resolution monthly is from April 2002 to May 2023.
+hydrological studies \citep{Ramjeawon2022, Yin2022}. The water storage
+and height anomalies are given in Equivalent Water Height or Thickness
+units (EWH, cm) \citep{Sabzehee2023}, and their temporal resolution
+monthly is from April 2002 to May 2023.
 
@@
 \section{Results}\label{results}}
+
+\hypertarget{drought-indices-for-water-demand-and-supply-1}{%
+\subsection{Drought indices for water demand and
+supply}\label{drought-indices-for-water-demand-and-supply-1}}
 

### 2024-01-04 article/paper_drought_Chile2023.qmd
update for igarrs2023

@@
       model: 3p
+      cite-style: number
       natbiboptions: numbers
@@
 In this research, we aim to analyze the impact of drought on different types of land cover classes in continental Chile by combining environmental variables such as biomass productivity, and water demand and supply gathered from earth observation products.
-The specific objectives of the study are: i) to calculate multi-scalar drought indices for water demand and supply for 1981–2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of biomass (zcNDVI) with drought indices; and iv) to assess if the observed changes in the drought indices are linked to the TWS.
+The specific objectives of the study are: i) to calculate multi-scalar drought indices for water demand and supply for 1981–2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of biomass (zcNDVI) with drought indices; and iv) to assess if the observed changes in the drought indices are linked to TWS.
 
@@
 
+For water supply and demand variabes, we used ERA5-Land [@Munoz2019], a reanalys dataset wich provides the evolution of land variables since 1950.
+It has a spatial resolution of 0.1°, hourly frequency and global coverage.
+We selected the variables for total precipitation, 2 meters temperature, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).
+
+To derive a proxy of vegetation productivity, we used the product MOD13A3 collection 6.1 from MODIS [@Didan2015].
+It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency.
+
 ###	in-situ data
@@
 
-::: {#fig-elephants layout-ncol=2 fig.pos='!ht'}
+::: {#fig-elephants layout-ncol=1 fig.pos='!ht'}
 ![](../output/figs/LC_pers80_per_macrozone.png){width="7cm"}

### 2024-01-04 article/paper_drought_Chile2023.tex
update for igarrs2023

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, breakable, enhanced]}{\end{tcolorbox}}\fi
 
@@
 indices; and iv) to assess if the observed changes in the drought
-indices are linked to the TWS.
+indices are linked to TWS.
 
@@
 
+For water supply and demand variabes, we used ERA5-Land
+\citep{Munoz2019}, a reanalys dataset wich provides the evolution of
+land variables since 1950. It has a spatial resolution of 0.1°, hourly
+frequency and global coverage. We selected the variables for total
+precipitation, 2 meters temperature, and volumetric soil water layers
+between 0 and 100cm of depth (layer 1 to layer 3).
+
+To derive a proxy of vegetation productivity, we used the product
+MOD13A3 collection 6.1 from MODIS \citep{Didan2015}. It provides
+vegetation indices (NDVI and EVI) at 1km of spatial resolution and
+monthly frequency.
+
 \hypertarget{in-situ-data}{%
@@
 
-\begin{minipage}[t]{0.47\linewidth}
+\begin{minipage}[t]{\linewidth}
 
@@
 \end{minipage}%
-%
-\begin{minipage}[t]{0.53\linewidth}
+\newline
+\begin{minipage}[t]{\linewidth}
 

### 2024-01-04 conference/IGARSS2024.qmd
update for igarrs2023

---
title: "DDS4Chl: A dataset of drought indices of demand and water supply for continental Chile"
format:
  ieee-pdf:
    keep-tex: false  
  ieee-html: default
title-block-banner: true
author:
  - id: dfolio
    name: Francisco Zambrano
    affiliations:
      - name: Facultad de Ciencias, Universidad Mayor
        department: Hemera Centro de Observación de la Tierra
        city: Santiago
        country: Chile
        postal-code: 8580745
        orcid: 0000-0001-6896-8534
        email: francisco.zambrano@umayor.cl
        #url: https://dfolio.fr/
        attributes:
        corresponding: true
        #photo: david-folio.png
  - id: iduran
    name: Iongel Duran
    affiliations: 
      - name: Facultad de Ciencias, Universidad Mayor
        department: Hemera Centro de Observación de la Tierra
        city: Santiago
        country: Chile
        postal-code: 8580745

note: "Template created June 23, 2023; revised `r format(Sys.Date(),format='%B %d, %Y')`."
#keywords: [IEEE, IEEEtran, journal, Quarto, Pandoc, template]
#funding: 
funding: 
  statement: "The `quarto-ieee` template is freely available under the MIT license on github: <https://github.com/dfolio/quarto-ieee>."
pageheader:
  left: IGARSS, January 2024
  right: 'D. Folio:  A Sample Article Using quarto-ieee'
#bibliography: bibliography.bib
date: 2024-01-05
pdf: https://github.com/dfolio/quarto-ieee/blob/main/template.pdf
citation: 
  container-title: GitHUB
  page: 1-3
  type: software
  issued: 2023-06-23
  url: https://github.com/dfolio/quarto-ieee
  pdf-url: https://github.com/dfolio/quarto-ieee/template.pdf
---

# **Abstract** {-}

Central Chile has been affected by a persistent drought which is impacting the hydrological system and vegetation development. The region has been the focus of research studies due to a persistent period of water scarcity which has been defined as a “mega drought.  
<br>

**Index Terms—** One, two, three, four, five

# **Introduction**

# **Methods**

# **Results**

# Conlusion

### 2024-01-04 R/02_mosaic_chile_modis_products.R

@@ -96,8 +96,8 @@ file.path(dir,datesUpdate,'/') %>% map(function(dir){
 })
 #Product MCD12Q1.061
 
-dir <- '/mnt/raster_raw/MODIS/MCD12Q1.061/'
-dir.out <- '/mnt/HDD4TB_2/data/rasters/Procesados/MODIS/IGBP.MCD12Q1.061/'
+dir <- '/mnt/md0/raster_raw/MODIS/MCD12Q1.061/'
+dir.out <- '/mnt/md0/raster_procesada/MODIS/IGBP.MCD12Q1.061/'
 dirs <- list.files(dir,full.names=TRUE)
 
 dirs %>% map(function(dir){

### 2024-01-04 scripts/08_smooth_NDVI_MOD13.R

@@ -57,7 +57,7 @@ dir_out <- '/mnt/md0/raster_procesada/MODIS_derived/NDVI_loess.MOD13A3.061/'
 
 new_names <- paste0(dir_out,'chl_',yj,'_1_km_monthly_NDVI_loess.tif')
 
-lapply(6:dim(ndvism)[3],\(i){
+lapply(7:dim(ndvism)[3],\(i){
   write_stars(ndvism[,,,i],dsn = new_names[i],type = 'Int16')
 })
 

### 2024-01-04 scripts/14_IGBP_persistencia.R

@@ -64,7 +64,7 @@ writeRaster(igbpRecl,paste0(dir.out,'IGBP',prop,'_reclassified.tif'),
 
 # landcover del último año (2021)
 
-igbp2021 <- classify(igbp[[21]],rcl=classes,include.lowest = TRUE)
-igbp2021[igbp2021 == 0] <- NA
-writeRaster(igbp2021,paste0(dir.out,'IGBP_2021_reclassified.tif'),
+igbp2022 <- classify(igbp[[22]],rcl=classes,include.lowest = TRUE)
+igbp2022[igbp2022 == 0] <- NA
+writeRaster(igbp2022,paste0(dir.out,'IGBP_2022_reclassified.tif'),
             wopt=list(datatype='INT1U'))

### 2024-01-04 scripts/15_IGBP_variacion_tendencia_2001-2021.R

@@ -25,10 +25,10 @@ igbpRecl[igbpRecl == 0] <- NA
 
 zones <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
   st_transform(32719)
-zones$macrozona <- c('little north','big north','austral','central','south')
+zones$macrozona <- c('norte grande','norte chico','austral','centro','sur')
 
 data <- terra::extract(igbpRecl,zones)
-names(data) <- c('zone',2001:2021)
+names(data) <- c('zone',2001:2022)
 
 library(dplyr)
 library(tidyr)
@@ -53,7 +53,7 @@ data2 |>
   left_join(dataSum,by='zone') |>  
   mutate(prop = n/ntot,sup_km2=n*0.2146587) -> data3
 
-saveRDS(data3,'data/processed_data/timeseries_landcover_zone_LCclass_2001-2021.rds')  
+saveRDS(data3,'data/processed_data/timeseries_landcover_zone_LCclass_2001-2022.rds')  
 
 #plot
 data3  |>  
@@ -66,7 +66,7 @@ data3  |>
   labs(y=expression(paste('Surface [',km^2,']'))) +
   # scale_y_continuous(labels = scales::percent_format(accuracy = .1),
   #                    expand = c(0,0)) +
-  facet_grid(LC_type~zone,scales ='free') +
+  facet_wrap(LC_type~zone,scales ='free_y') +
   theme_bw() +
   theme(axis.title.x = element_blank())
 
@@ -79,4 +79,4 @@ data3 |>
   nest_by(zone,LC_type) |>  
   summarize(trend  = as.numeric(tidy(lm(sup_km2~year,data))[2,2])) |> 
   pivot_wider(names_from = 'LC_type',values_from=3) |> 
-  saveRDS('data/processed_data/trends_landcover_2001-2021.rds')  
+  saveRDS('data/processed_data/trends_landcover_2001-2022.rds')  

### 2024-01-04 scripts/17_extrarer_correlations_macro_landcover.R

@@ -12,11 +12,6 @@ paleta <- read.csv('data/processed_data/paleta_colores_landcover.csv') |>
 colors <-  rgb(paleta$R,paleta$G,paleta$B,maxColorValue = 255)
 attr(colors,'names') <- paleta$class
 
-# 
-pal <- read.csv('data/processed_data/paleta_colores_landcover.csv')
-t <- pal$class
-names(t) <- pal$Name
-
 lc <- rast('/mnt/md0/raster_procesada/MODIS_derived/IGBP.MCD12Q1.061/IGBP80_reclassified.tif')
 
 dir <- '/mnt/md0/raster_procesada/analysis/correlations/'
@@ -26,7 +21,8 @@ cors <- rast(files)
 cors <- resample(cors,lc)
 
 macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
-  st_transform(32719)
+  st_transform(32719) |> 
+  mutate(macrozona = c('norte chico','norte grande','austral','centro','sur'))
 
 library(tidyverse)
 getmode <- function(v) {
@@ -34,100 +30,35 @@ getmode <- function(v) {
   uniqv[which.max(tabulate(match(v, uniqv)))]
 }
 
-cors_i <- subset(cors,seq(1,8,2))
+#obtener los indices y escalas de tiempo en dónde se alcanza la máxima correlación
+cors <- subset(cors,seq(1,8,2))
 data_ind <- map_df(1:5,function(i){
-  cors_im <- mask(cors_i,macro[i,])
+  cors_m <- mask(cors,macro[i,])
   lc_m <- mask(lc,macro[i,])
-  cors_df <- zonal(cors_im,lc_m,getmode)
-  names(cors_df)[2:5] <- c('EDDI','SPI','SPEI','zcSM')
+  cors_df <- zonal(cors_m,lc_m,getmode)
+  names(cors_df)[2:5] <- c('eddi','spi','spei','zcsm')
   cors_df |> 
     pivot_longer(-1) |> 
     rename(indice = name) |> 
-    mutate(clase = factor(lyr.1,levels = t, labels = names(t)),
+    mutate(clase = factor(lyr.1,levels = paleta$class, labels = paleta$Name),
            macro = macro[i,]$macrozona) |> 
     select(macro,indice,clase,value) 
 })
 
-data_ind_4gt <- data_ind |> 
+data_ind |> 
   filter(clase %in% c('Forest','Shrubland','Savanna','Grassland','Cropland','Barren land')) |> 
[…]

### 2024-01-04 scripts/20_figuras_zcNDVI_macro_landcover.R

@@ -6,6 +6,7 @@ library(sf)
 library(purrr)
 library(stringr)
 library(tidyverse)
+library(glue)
 
 #paleta colores landcover persistencia
 paleta <- read.csv('data/processed_data/paleta_colores_landcover.csv') |> 
@@ -19,43 +20,29 @@ pal <- read.csv('data/processed_data/paleta_colores_landcover.csv')
 t <- pal$class
 names(t) <- pal$Name
 
-lc <- rast('/mnt/md0/raster_procesada/MODIS_derived/IGBP.MCD12Q1.061/IGBP80_reclassified.tif') |> 
-  project('EPSG:4326',method = 'mode')
+lc <- rast('/mnt/md0/raster_procesada/MODIS_derived/IGBP.MCD12Q1.061/IGBP80_reclassified.tif') 
+  # project('EPSG:4326',method = 'mode') #for ERA5-Land
 
-dir <- '/mnt/md0/raster_procesada/ERA5-Land_tiff/sequia/zcSM/zcSM-12'
+dir <- '/mnt/md0/raster_procesada/MODIS_derived/zcNDVI/zcNDVI-6'
 files <- dir_ls(dir,regexp = 'tif$')
 
 index <- rast(files)
 index <- resample(index,lc)
 
-macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg')
+macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> st_transform(32719)
 
 data_index <- map_df(1:5,function(i){
   index_im <- mask(index,macro[i,])
   lc_m <- mask(lc,macro[i,])
   index_df <- zonal(index_im,lc_m,na.rm = TRUE)
   names(index_df) <- c('lyr',str_extract(names(index_df[-1]),'[0-9]{4}-[0-9]{2}-[0-9]{2}'))
-  index_df |> 
+  d <- index_df |> 
     pivot_longer(-1) |> 
     rename(date = name) |> 
     mutate(clase = factor(lyr,levels = t, labels = names(t)),
            macro = macro[i,]$macrozona) |> 
     select(macro,date,clase,value) 
+  write_rds(d,glue('data/processed_data/timesseries_zcNDVI6_macro_landcover{i}.rds'))
+  
 })
 
-write_rds(data_index,'data/processed_data/timesseries_zcSM12_macro_landcover.rds')
-
-data_index |> 
[…]

### 2024-01-04 scripts/22_tendencia_indices_sequia_landcover_macro.R

@@ -1,0 +1,83 @@
+# November 2023
+# by frzambra
+
+dir <- '/mnt/md0/raster_procesada/MODIS/'
+
+paleta <- read.csv('data/processed_data/paleta_colores_landcover.csv')
+colors <-  rgb(paleta$R,paleta$G,paleta$B,maxColorValue = 255)
+attr(colors,'names') <- paleta$Name
+
+library(terra)
+library(rnaturalearth)
+library(sf)
+library(tidyverse)
+library(fs)
+library(stars)
+
+chl <- ne_countries(country = 'chile',returnclass = 'sf') |> 
+  st_transform(32719)
+
+macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
+  st_transform(32719) |> 
+  mutate(macrozona = c('norte chico','norte grande','austral','centro','sur'))
+
+lc <- rast('/mnt/md0/raster_procesada/MODIS_derived/IGBP.MCD12Q1.061/IGBP80_reclassified.tif') 
+
+indices <- c('EDDI','SPI','SPEI','zcSM')
+scales <- c(1,3,6,12,24,36)
+scales_ndvi <- c(1,3,6,12)
+
+data_t <- map_df(indices,\(index){
+  if( index == 'zcNDVI') scales <- scales_ndvi
+  
+  trends <- map(scales,\(scale){
+    
+    if (index == 'zcNDVI'){
+      dir <- glue::glue('/mnt/md0/raster_procesada/MODIS_derived/zcNDVI/{index}-{scale}')
+    } else dir <- glue::glue('/mnt/md0/raster_procesada/ERA5-Land_tiff/sequia/{index}/{index}-{scale}')
+    
+    #files <- dir_ls(dir,regexp = paste(2001:2023,collapse='|'))
+    files <- dir_ls(dir,regexp = 'tif$')
+    index_ts <- rast(files)
+    #cores <- parallel::detectCores()-5
+
+    trend <- app(index_ts,\(y){
+      df <- data.frame(x = seq_along(y),y=y)
+      if(!(is.na(df[,2]) |> all())){
+        summary(lm(y~x,df))$coefficients[2,1]
+      } else  NA
+      })
+    return(trend)
+  })
+  
+  trends <- rast(trends)
+  names_ind <- if(index == 'zcNDVI') paste0(index,'-',c(1,3,6,12)) else paste0(index,'-',c(1,3,6,12,24,36))
+  names(trends) <- names_ind
+
[…]

### 2024-01-04 scripts/23_analysis_trend_drought_indices_and_landcover_macroXlandcover.R

@@ -1,0 +1,139 @@
+library(tidyverse)
+library(tidymodels)
+
+data_trend_lc <- read_rds('data/processed_data/trends_landcover_2001-2022.rds') |> 
+  mutate(zone = fct_recode(zone,'norte grande' = 'big north' , 'norte chico' = 'little north','centro' = 'central', 'sur' = 'south' ,'austral' = 'austral')) |> 
+  pivot_longer(-zone) |> 
+  rename(landcover = value,macro = zone,clase = name)
+
+data_trend_di <- read_rds('data/processed_data/trends_indices_macroXlandocver_2001_2023.rds')
+
+data_ana <- data_trend_di |> 
+  full_join(data_trend_lc) |> 
+  drop_na()
+
+#análisis de cluster
+library(dbscan)
+library(NbClust)
+
+res <- sapply(1:10,\(i){
+  kmeans(data_ana[,3:31],i)$tot.withinss  
+})
+
+res.nbclust <- NbClust(data_ana[,3:31], distance = "euclidean",
+                       min.nc = 2, max.nc = 9, 
+                       method = "ward.D2", index ="dunn")
+factoextra::fviz_nbclust(res.nbclust) + theme_minimal() + ggtitle("NbClust's optimal number of clusters")
+
+data_ana <- as.data.frame(data_ana)
+row.names(data_ana) <- paste0(data_ana$macro,'_',data_ana$clase)
+
+data_cluster <- data_ana[,1:2]
+
+hclust_eddi <-  data_ana[,3:8,drop = FALSE] |> scale() |>   dist() |> hclust(method = 'ward.D2')
+
+hclust_spi <-  data_ana[,9:14,drop = FALSE] |> scale() |>   dist() |> hclust(method = 'ward.D2')
+
+hclust_spei <-  data_ana[,15:20,drop = FALSE] |> scale() |>   dist() |> hclust(method = 'ward.D2')
+
+hclust_zcsm <-  data_ana[,21:26,drop = FALSE] |> scale() |>   dist() |> hclust(method = 'ward.D2')
+
+hclust_zcndvi <-  data_ana[,27:30,drop = FALSE] |> scale() |>   dist() |> hclust(method = 'ward.D2')
+
+hclust_land <-  data_ana[,31,drop = FALSE] |> scale() |>   dist() |> hclust(method = 'ward.D2')
+
[…]

### 2024-01-04 scripts/24_extraer_valores_indices_sequia_macrozona_landcover.R

@@ -1,0 +1,66 @@
+# November 2023
+# by frzambra
+
+dir <- '/mnt/md0/raster_procesada/MODIS/'
+
+paleta <- read.csv('data/processed_data/paleta_colores_landcover.csv')
+colors <-  rgb(paleta$R,paleta$G,paleta$B,maxColorValue = 255)
+attr(colors,'names') <- paleta$Name
+
+library(terra)
+library(rnaturalearth)
+library(sf)
+library(tidyverse)
+library(fs)
+library(stars)
+
+chl <- ne_countries(country = 'chile',returnclass = 'sf') |> 
+  st_transform(32719)
+
+macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
+  st_transform(32719) |> 
+  mutate(macrozona = c('norte chico','norte grande','austral','centro','sur'))
+
+lc <- rast('/mnt/md0/raster_procesada/MODIS_derived/IGBP.MCD12Q1.061/IGBP80_reclassified.tif') 
+
+lc <- project(lc,"EPSG:4326",method = 'mode')
+macro <- st_transform(macro,4326)
+
+indices <- c('EDDI','SPI','SPEI','zcSM')
+scales <- c(1,3,6,12,24,36)
+scales_ndvi <- c(1,3,6,12)
+
+data_index <- map_df(indices,\(index){
+  if( index == 'zcNDVI') scales <- scales_ndvi
+  
+  trends <- map(scales,\(scale){
+    
+    if (index == 'zcNDVI'){
+      dir <- glue::glue('/mnt/md0/raster_procesada/MODIS_derived/zcNDVI/{index}-{scale}')
+    } else dir <- glue::glue('/mnt/md0/raster_procesada/ERA5-Land_tiff/sequia/{index}/{index}-{scale}')
+    
+    files <- dir_ls(dir,regexp = 'tif$')
+    index_ts <- rast(files)
+    index_ts_r <- resample(index_ts,lc,threads = 50)
+    
+  
+  data_ts <- map_df(1:5,function(i){
+    index_m <- mask(index_ts_r,macro[i,])
+    lc_m <- mask(lc,macro[i,])
+    index_df <- zonal(index_m,lc_m,'median',na.rm = TRUE)
+    
+    trends_df |> 
+      pivot_longer(-1) |> 
+      rename(scale = name) |> 
+      mutate(clase = factor(lyr.1,levels = paleta$class, labels = paleta$Name),
[…]

### 2024-01-04 scripts/25_trend_drought_indices.R

@@ -1,0 +1,82 @@
+library(tidyverse)
+library(terra)
+library(tmap)
+library(sf)
+library(rnaturalearth)
+
+chl_b <- ne_countries(country='chile',scale = 'medium',returnclass = 'sf')
+macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') 
+
+trend_spi <- rast('/mnt/md0/raster_procesada/analysis/trends/trend_SPI_1981-2023.tif')
+#names(trend_spei) <- paste0('SPEI-',c(1,3,6,12,24,36))
+#writeRaster(trend_spei,'/mnt/md0/raster_procesada/analysis/trends/trend_SPEI_1981-2023_2.tif',overwrite = TRUE)
+names(trend_spi) <- paste0('SPI-',c(1,3,6,12,24,36))
+map_spi <- tm_shape(trend_spi*100) + 
+  tm_raster(palette = '-inferno',midpoint = 0,title = 'Trend SPI (x 100)',style = 'kmeans') +
+  tm_shape(macro) + 
+  tm_borders(col='white') +
+  tm_shape(chl_b) + 
+  tm_borders(col = 'black') +
+  tm_facets() +
+  tm_layout(panel.labels = paste0('SPI-',c(1,3,6,12,24,36)))
+tmap_save(map_spi,'output/figs/trend_raster_SPI_1981-2023.png',asp=.2)
+
+trend_spei <- rast('/mnt/md0/raster_procesada/analysis/trends/trend_SPEI_1981-2023.tif')
+#names(trend_spei) <- paste0('SPEI-',c(1,3,6,12,24,36))
+#writeRaster(trend_spei,'/mnt/md0/raster_procesada/analysis/trends/trend_SPEI_1981-2023_2.tif',overwrite = TRUE)
+
+map_spei <- tm_shape(trend_spei*100) + 
+  tm_raster(palette = '-inferno',midpoint = 0,title = 'Trend SPEI (x 100)',style = 'kmeans') +
+  tm_shape(macro) + 
+  tm_borders(col='white') +
+  tm_shape(chl_b) + 
+  tm_borders(col = 'black') +
+  tm_facets() +
+  tm_layout(panel.labels = paste0('SPEI-',c(1,3,6,12,24,36)))
+tmap_save(map_spei,'output/figs/trend_raster_SPEI_1981-2023.png',asp=.2)
[…]

### 2024-01-05 article/paper_drought_Chile2023.qmd
guardar cambios previos

@@
 ---
-title: Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from the ERA5-Land and MODIS
+title: Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from ERA5-Land and MODIS
 #subtitle: A Short Subtitle
@@
 (c) Land cover classes for 2021.
-(d) Persistent land cover classes (> 80%) for 2001-2021.](../output/figs/map_study_con_landcover.png){#fig1_study_area width=80% fig-pos='ht'}
+(d) Persistent land cover classes (> 80%) for 2001-2021.](../output/figs/map_study_con_landcover.png){fig-pos='!ht'}
 
@@
 
+\begin{table}[!ht]
+\caption{Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile.
+Red dots on the plots indicate the maximum and minimum surface}
+\includegraphics[]{../output/figs/table_var_landcover_macro.png}
+\end{table}
+
 ##	Total water storage and drought indices
@@
 Each GRACE Tellus monthly grid at 0.5 degrees represents the deviation of surface mass for that month relative to a reference time average (2004-2009 baseline), which is subtracted from all other monthly grids to provide terrestrial water storage anomalies (TWSA) [@Ramjeawon2022; @Yin2022].
-This JPL-M version of the data employs a Coastal Resolution Improvement (CRI) filter that reduces signal leakage errors across coastlines [@Wiese2019; @Wiese2016].
+This JPL-M version of the data employs a Coastal Resolution Improvement (CRI) filter that reduces signal leakage errors across coastlines [@NASA2019; @Wiese2016].
 Although the mascon solutions greatly reduced leakage errors, a gain factor, which is used to enhance the spatial resolution, was applied to the dataset as recommended for hydrological studies [@Ramjeawon2022; @Yin2022].

### 2024-01-05 conference/IGARSS2024.qmd
first version of the article igarss2023

@@
 ---
-title: "DDS4Chl: A dataset of drought indices of demand and water supply for continental Chile"
+title: "DDS4Chl: A dataset of drought indices of water demand and supply and its impact on vegetation for continental Chile"
 format:
@@
 funding: 
-  statement: "The `quarto-ieee` template is freely available under the MIT license on github: <https://github.com/dfolio/quarto-ieee>."
+  statement: "The research was funded by the National Research and Development Agency of Chile (ANID) and the projects FSEQ210022 and 11190360."
 pageheader:
@@
   right: 'D. Folio:  A Sample Article Using quarto-ieee'
-#bibliography: bibliography.bib
+bibliography: references.bib
 date: 2024-01-05
@@
 
-Central Chile has been affected by a persistent drought which is impacting the hydrological system and vegetation development.
-The region has been the focus of research studies due to a persistent period of water scarcity which has been defined as a “mega drought.
+A persistent drought is impacting Chile.
+It affects the hydrological system and vegetation development.
+Research studies have focused on the central part of the country.
+This is due to a persistent period of water scarcity.
+Scientists define this scarcity as a megadrought.
+This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December.
+The SPI only considers precipitation as a drought indicator.
+It does not account for atmospheric evaporative demand (AED), soil moisture, or their combined effect on plant growth.
+We present a developing database (DDS4Chl) of drought indices for continental Chile since 1981.
+The indices measure water demand, supply and its impact on vegetation.
+We derived the SPI for water demand.
+We also derived the Evaporative Demand Drought Index (EDDI) for water supply.
+The Standardized Precipitation Evapotranspiration Index (SPEI) shows the combined effect of AED and precipitation.
+We estimate the standardized anomaly of cumulative soil moisture at one meter (zcSM) and the standardized anomaly of cumulative NDVI (zcNDVI) to show the impact of water demand and supply.
+Finally, we present the historical linear trend of the drought indices in continental Chile.
+We cover short to long scales.
 <br>
@@
 
+The sixth assessment report (AR6) of the IPCC [@IPCC2023] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes, and the evidence has been strengthened since AR5 [@IPCC2013].
+There is high confidence that increasing global warming can expand the land area affected by increasing drought frequency and severity [@IPCCCH112021].
+Climate change enhance atmospheric evaporative demand (AED) and modified precipitation patterns.
+Thus, to monitor drought, we must consider the water supply and demand.
+We must also consider its impact on vegetation productivity.
+
+Chile has been facing a persistent rainfall deficit for more than a decade [@Garreaud2017], which has impacted vegetation development [@Zambrano2023] and the hydrological system [@Boisier2018].
+Current drought conditions have affected crop productivity [@Zambrano2016; @Zambrano2018], forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], land cover change [@Fuentes2021], water supply in watersheds [@AlvarezGarreton2021], and have had economic impacts [@Fernandez2023].
+In 2019–2020, the drought severity reached an extreme condition in Central Chile (30–34°S) not seen for at least 40 years, and the evidence indicates that the impact is transversal to the land cover classes of forest, grassland, and cropland [@Zambrano2023].
+The prolonged lack of precipitation in Central Chile is producing changes in ecosystem dynamics that must be studied.
+
+To evaluate meteorological drought (i.e., water supply), the World Meteorological Organization (WMO; @WMO2012) recommends the Standardized Precipitation Index (SPI; @McKee1993), a multiscalar drought index that allows to monitor precipitation deficits from short- to long-term.
+Following the same approach, @Vicente-Serrano2010 incorporates into the SPI the effect of temperature through the use of potential evapotranspiration, thus proposing the SPEI (Standardized Precipitation Evapotranspiration Index).
+Similarly, to evaluate solely the evaporative demand driven by temperature, @Hobbins2016 and @McEvoy2016 came up with the Evaporative Demand Drought Index (EDDI).
+To assess the impact of water supply and demand on vegetation, Zambrano (2018) proposed the zcNDVI.
+This is a standardized anomaly of the cumulative Normalized Difference Vegetation Index (NDVI).
+It's similar to the SPI, SPEI, and EDDI.
+The zcNDVI could be accumulated over the growing season or any period (e.g., months).
+It results in a multiscalar drought index.
+Several drought indices exist for soil moisture.
+These include the Soil Moisture Deficit Index (SDMI), a normalized index [@Narasimhan2005], and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021].
+SMADI is a normalized index that uses vegetation, land surface temperature, and a vegetation condition index (VCI, [@Kogan1995]).
+
 # **Methods**
+
+## Study area
+
+For the analysis, continental Chile was divided into five zones according to a latitudinal gradient: “Norte Grande”,  “Norte Chico”,  “Zona Central”, “Zona Sur”, and “Zona Austral”. (@fig-studyArea)
+
+```{r fig-studyArea}
+#| fig-pos: "!ht"
+#| fig-cap: "(a) Chile and the five zones 'norte grande', 'norte chico',  'zona central', 'zona sur', and 'zona austral'.
+(b) Topography reference map.
+(c) Land cover classes for 2021.
+(d) Persistent land cover classes (> 80%) for 2001-2021"
+
+knitr::include_graphics("../output/figs/map_study_con_landcover.png")
+```
+
+##	Data
+
+###	Earth observation data
+
+For water supply and demand variabes, we used ERA5-Land [@Munoz2019], a reanalys dataset wich provides the evolution of land variables since 1950.
+It has a spatial resolution of 0.1°, hourly frequency and global coverage.
+We selected the variables for total precipitation, 2 meters temperature, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).
+
+To derive a proxy of vegetation productivity, we used the product MOD13A3 collection 6.1 from MODIS [@Didan2015].
+It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency.
+
+## Drought indices
+
+We used a non-parametric method for the procedure of normalization and derivation of the standardized anomaly (drought indices).
+This method obtains empirically derived probabilities using an inverse normal approximation [@Hobbins2016].
+We derived the SPI, SPEI, EDDI, zcSM, and zcNDVI drought indices.
+We calculate all aggregations at 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which is calculated at 1, 3, and 6 months.
 
@@
 
-# Conlusion
+```{r}
+#| fig-pos: "!ht"
+#| fig-cap: "Linear trend of the SPI at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
+knitr::include_graphics("../output/figs/trend_raster_SPI_1981-2023.png")
+```
+
+```{r}
+#| fig-pos: "!ht"
+#| fig-cap: ""
+knitr::include_graphics("../output/figs/trend_raster_EDDI_1981-2023.png")
+```
+
+```{r}
+#| fig-pos: "!ht"
+#| fig-cap: ""
+knitr::include_graphics("../output/figs/trend_raster_zcNDVI_2001-2023.png")
+```
+
+# **Conlusion**
+
+# **References**

### 2024-01-05 conference/IGARSS2024.qmd
first version of the article igarss2023

@@
 
-```{r}
+For the SPI, in the macrozones "norte chico", "central", and "sur" the trend decrease with the time scales of the index.
+Showing that at longer scales (e.g, 36 months) it reaches a trend of -0.056 (z-score) per decade (@fig-trendSPI).
+In the case of the AED measured by the EDDI.
+It shows a positive trend, caused by a increase on AED.
+The trend on EDDI reaches 0.053 per decade from the macrozones "norte grande" and "norte chico".
+
+```{r fig-trendSPI}
 #| fig-pos: "!ht"

### 2024-01-06 conference/IGARSS2024.qmd
first draft paper IGARSS2024

--- 2024-01-05 conference/IGARSS2024.qmd
@@
     affiliations:
-      - name: Facultad de Ciencias, Universidad Mayor
+      - name: Facultad de Ciencias, Ingeniería y Tecnologías, Universidad Mayor
         department: Hemera Centro de Observación de la Tierra
@@
   - id: iduran
-    name: Iongel Duran
+    name: Iongel Duran-Llacer
     affiliations: 
-      - name: Facultad de Ciencias, Universidad Mayor
+      - name: Facultad de Ciencias, Ingeniería y Tecnologías, Universidad Mayor
         department: Hemera Centro de Observación de la Tierra
@@
 This is due to a persistent period of water scarcity.
-Scientists define this scarcity as a megadrought.
+This scarcity has been found to be a megadrought.
 This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December.
 The SPI only considers precipitation as a drought indicator.
-It does not account for atmospheric evaporative demand (AED), soil moisture, or their combined effect on plant growth.
+It does not account for atmospheric evaporative demand (AED), soil moisture, or their combined vegetation productivity.
 We present a developing database (DDS4Chl) of drought indices for continental Chile since 1981.
@@
 Finally, we present the historical linear trend of the drought indices in continental Chile.
-We cover short to long scales.
+Also we calculated the temporal correlation between the indices of water supply, water demand, and soil moisture with the zcNDVI (proxy of vegetation productivity).
+The dought index for soil moisture (zcSM) derived from ERA5-Land showed to be the best predictor of vegetation productivity around continental Chile.
 <br>
 
-**Index Terms—** One, two, three, four, five
+**Index Terms—** drought, Chile, water supply, water demand
 
@@
 
-For the analysis, continental Chile was divided into five zones according to a latitudinal gradient: “Norte Grande”,  “Norte Chico”,  “Zona Central”, “Zona Sur”, and “Zona Austral”. (@fig-studyArea)
+Continental Chile has at least seven major climatic types.
+They range from low desert in the north to alpine tundra and glaciers in the east and southeast.
+The south has an Oceanic climate and central Chile has a Mediterranean climate [@kottek2006].
+For the analysis, continental Chile was also divided into five zones according to a latitudinal gradient: "Norte Grande", "Norte Chico",  "Zona Central", "Zona Sur", and "Zona Austral".
+(@fig-studyArea)
 
@@
 (d) Persistent land cover classes (> 80%) for 2001-2021"
+#| out-width: "\\textwidth"
 
@@
 We derived the SPI, SPEI, EDDI, zcSM, and zcNDVI drought indices.
-We calculate all aggregations at 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which is calculated at 1, 3, and 6 months.
+We calculate all aggregations at 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which is calculated at 6 months.
+
+## Correlation and trend analysis
+
+For all the pixels, we apply the following procedure: i) First, we calculated the linear trend per drought index and time scale for 1981–2023 (SPI, SPEI, EDDI, zcSM) and 2001–2023 (zcNDVI).
+ii) Second, over all the pixels, we used the zcNDVI as a proxy of vegetation productivity and analyzed its correlation with the drought indices SPI, SPEI, EDDI, and zcSM.
+We selected the time scales per index that reached the maximum coefficient of determination, and then we extracted the Pearson correlation for those time scales.
+The goal is to identify which climate variable has the most significant impact on vegetation development.
+Also, identify if it is a short- or long-term effect of the climate variable on vegetation.
 
@@
 
-```{r}
+Regarding vegetation productivity, there is a negative trend on zcNDVI in the central part of the country ("norte chico" and "zona central") (@fig-zcNDVI_var).
+Indicating either a reduction of vegetation surface, biomass, or status.
+It reached its lowest values during the year 2020, mostly on the "zona central." There are other negative trends in small areas in "norte grande" and "zona asutral."
+
+```{r fig-zcNDVI_var, out.extra='style="float:right; padding:10px"'}
+#| fig-pos: "!ht"
+#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023.
+Greener colors indicate a positive trend and an increase in vegetation development; reder colors correspond to a negative trend and a decrease in vegetation development.
+(b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile.
+Each horizontal panel corresponds to a macrozone from 'norte grande' to 'zona austral'."
+#| out-width: "\\textwidth"
+knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas.png")
+```
+
+Analyzing the water supply in the macrozones "norte chico," "central," and "sur," the SPI shows a decreasing trend that increases at longer time scales due to the prolonged reduction in precipitation.
+It reaches a trend of -0.056 (z-score) per decade (@fig-trendSPI) for time scales of 36 months.
+In the case of water demand, we analyze the EDDI.
+It shows a positive trend, caused by an increase in AED.
+The trend on EDDI reaches a maximum of 0.053 per decade in the macrozones "norte grande" and "norte chico" (@fig-trendEDDI).
+The behavior is similar to SPI, having close trends with opposite signs.
+Nevertheless, the spatial patterns are different.
+The maximum trend for SPI-36 is in "norte chico" and "zona central." For EDDI-36, the maximum trend is from "northe grande" to "zona central."
+
+```{r fig-trendSPI}
 #| fig-pos: "!ht"
 #| fig-cap: "Linear trend of the SPI at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
+#| out-width: "\\textwidth"
+
 knitr::include_graphics("../output/figs/trend_raster_SPI_1981-2023.png")
@@
 
-```{r}
+```{r fig-trendEDDI}
 #| fig-pos: "!ht"
-#| fig-cap: ""
+#| out-width: "\\textwidth"
+#| fig-cap: "Linear trend of the EDDI at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
 knitr::include_graphics("../output/figs/trend_raster_EDDI_1981-2023.png")
@@
 
-```{r}
+The correlation between zcNDVI-6 and drought indices shows that time scales higher than six months are predominant over continental Chile in explaining vegetation variability (@fig-corTimeScale).
+Pearson correlations between 0.6 and 0.9 are concentrated in the "norte chico" and "zona central" for SPI and SPEI.
+Soil moisture reached a correlation of 0.6–0.9 that extended north and south with respect to SPI.
+
+```{r fig-corTimeScale}
 #| fig-pos: "!ht"
-#| fig-cap: ""
-knitr::include_graphics("../output/figs/trend_raster_zcNDVI_2001-2023.png")
+#| out-width: "\\textwidth"
+#| fig-cap: "Time scales per drought index that reach the maximum coefficient of determination"
+knitr::include_graphics("../output/figs/mapa_cor_selec_indices_zcNDVI6.png")
+```
+
+```{r fig-corPerson}
+#| fig-pos: "!ht"
+#| out-width: "\\textwidth"
+#| fig-cap: "Pearson correlation value for the time scales and drought index that reach the maximum coefficient of determination"
+knitr::include_graphics("../output/figs/mapa_cor_r_indices_zcNDVI6.png")
 ```
@@
 
+The vegetation productivity shows a high negative trend in the "norte chico" and "zona central" of continental Chile.
+This trend is in the primary case related to long-term precipitation decreases, which in the secondary case could be explained by the increase in AED.
+The dought index for soil moisture (zcSM) derived from ERA5-Land showed to be the best predictor of vegetation productivity around continental Chile.
+
 # **References**

### 2024-01-06 scripts/25_trend_drought_indices.R

@@ -8,11 +8,21 @@ chl_b <- ne_countries(country='chile',scale = 'medium',returnclass = 'sf')
 macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') 
 
 trend_spi <- rast('/mnt/md0/raster_procesada/analysis/trends/trend_SPI_1981-2023.tif')
+
+macro_ras <- rasterize(macro,trend_spi,field = 'macrozona') 
+
+data_trend_spi_macro <- zonal(trend_spi,macro_ras,'mean',na.rm=TRUE) |> as_tibble()
+
+data_trend_spi_macro |> 
+  pivot_longer(-macrozona) |> 
+  mutate(value = value*10) |> 
+  arrange(value)
+
 #names(trend_spei) <- paste0('SPEI-',c(1,3,6,12,24,36))
 #writeRaster(trend_spei,'/mnt/md0/raster_procesada/analysis/trends/trend_SPEI_1981-2023_2.tif',overwrite = TRUE)
 names(trend_spi) <- paste0('SPI-',c(1,3,6,12,24,36))
-map_spi <- tm_shape(trend_spi*100) + 
-  tm_raster(palette = '-inferno',midpoint = 0,title = 'Trend SPI (x 100)',style = 'kmeans') +
+map_spi <- tm_shape(trend_spi*10) + 
+  tm_raster(palette = '-inferno',midpoint = 0,title = 'Trend SPI (per decade)',style = 'kmeans') +
   tm_shape(macro) + 
   tm_borders(col='white') +
   tm_shape(chl_b) + 
@@ -25,8 +35,8 @@ trend_spei <- rast('/mnt/md0/raster_procesada/analysis/trends/trend_SPEI_1981-20
 #names(trend_spei) <- paste0('SPEI-',c(1,3,6,12,24,36))
 #writeRaster(trend_spei,'/mnt/md0/raster_procesada/analysis/trends/trend_SPEI_1981-2023_2.tif',overwrite = TRUE)
 
-map_spei <- tm_shape(trend_spei*100) + 
-  tm_raster(palette = '-inferno',midpoint = 0,title = 'Trend SPEI (x 100)',style = 'kmeans') +
+map_spei <- tm_shape(trend_spei*10) + 
[…]

### 2024-01-08 article/paper_drought_Chile2023.qmd
last update

@@
 In this research, we aim to analyze the impact of drought on different types of land cover classes in continental Chile by combining environmental variables such as biomass productivity, and water demand and supply gathered from earth observation products.
-The specific objectives of the study are: i) to calculate multi-scalar drought indices for water demand and supply for 1981–2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of biomass (zcNDVI) with drought indices; and iv) to assess if the observed changes in the drought indices are linked to TWS.
+The specific objectives of the study are: i) to analize the trend on multi-scalar drought indices for water demand/supply, soil moisture and vegetation productivity for 1981–2023/2001-2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of biomass (zcNDVI) with drought indices; and iv) to assess if the observed changes in the drought indices are linked to TWS.
 
@@
 
-##	Drought indices for water demand and supply
+## Drought indices
+
+##
+
+##	Drought indices for water demand/supply, soil moisture and vegetation productivity
 
@@
 
-![Proportion of land cover class from the persistent land cover for 2001-2021 (>80%) per macrozone.
-
-::: {#fig-elephants layout-ncol=1 fig.pos='!ht'}
-![](../output/figs/LC_pers80_per_macrozone.png){width="7cm"}
-
-![](../output/figs/table_var_landcover_macro.png){width="8cm"}
-
-Proportion of land cover class from the persistent land cover for 2001-2021 (>80%) per macrozone.
-The table on the left shows the linear change trend next to time-series plot of surface, per land cover class (IGBP MCD12Q1.006) for 2001-2021 through the fives zones on continental Chile.
-Blue dots on the plots indicate the maximum and red dots minimum surface reach.
-:::
+```{r LCprop}
+#| echo: false
+#| fig-pos: "!ht"
+#| fig-cap: "Proportion of land cover class from the persistent land cover for 2001-2021 (>80%) per macrozone"
+#| out-width: "\\textwidth"
+knitr::include_graphics("../output/figs/LC_pers80_per_macrozone.png")
+```
 
@@
 
-##	Drought indices for water demand and supply
+##	Trend of drought indices for water demand/supply, soil moisture, and vegetation productivity
 
-##	Biomass proxy with drought indices of supply and demand
+Regarding vegetation productivity, there is a negative trend on zcNDVI in the central part of the country ("norte chico" and "zona central") (@fig-zcNDVI_var).
+Indicating either a reduction of vegetation surface, biomass, or status.
+It reached its lowest values during the year 2020, mostly on the "zona central." There are other negative trends in small areas in "norte grande" and "zona asutral."
+
+```{r fig-zcNDVI_var}
+#| echo: false
+#| fig-pos: "!ht"
+#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023.
+Greener colors indicate a positive trend and an increase in vegetation development; reder colors correspond to a negative trend and a decrease in vegetation development.
+(b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile.
+Each horizontal panel corresponds to a macrozone from 'norte grande' to 'zona austral'."
+#| out-width: "\\textwidth"
+knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas.png")
+```
+
+Analyzing the water supply in the macrozones "norte chico," "central," and "sur," the SPI shows a decreasing trend that increases at longer time scales due to the prolonged reduction in precipitation.
+It reaches a trend of -0.056 (z-score) per decade (@fig-trendSPI) for time scales of 36 months.
+In the case of water demand, we analyze the EDDI.
+It shows a positive trend, caused by an increase in AED.
+The trend on EDDI reaches a maximum of 0.053 per decade in the macrozones "norte grande" and "norte chico" (@fig-trendEDDI).
+The behavior is similar to SPI, having close trends with opposite signs.
+Nevertheless, the spatial patterns are different.
+The maximum trend for SPI-36 is in "norte chico" and "zona central." For EDDI-36, the maximum trend is from "northe grande" to "zona central."
+
+```{r fig-trendSPI}
+#| echo: false
+#| fig-pos: "!ht"
+#| fig-cap: "Linear trend of the SPI at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
+#| out-width: "\\textwidth"
+
+knitr::include_graphics("../output/figs/trend_raster_SPI_1981-2023.png")
+```
+
+```{r fig-trendEDDI}
+#| echo: false
+#| fig-pos: "!ht"
+#| out-width: "\\textwidth"
+#| fig-cap: "Linear trend of the EDDI at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
+knitr::include_graphics("../output/figs/trend_raster_EDDI_1981-2023.png")
+```
+
+The correlation between zcNDVI-6 and drought indices shows that time scales higher than six months are predominant over continental Chile in explaining vegetation variability (@fig-corTimeScale).
+Pearson correlations between 0.6 and 0.9 are concentrated in the "norte chico" and "zona central" for SPI and SPEI.
+Soil moisture reached a correlation of 0.6–0.9 that extended north and south with respect to SPI.
+
+```{r fig-corTimeScale}
+#| echo: false
+#| fig-pos: "!ht"
+#| out-width: "\\textwidth"
+#| fig-cap: "Time scales per drought index that reach the maximum coefficient of determination"
+knitr::include_graphics("../output/figs/mapa_cor_selec_indices_zcNDVI6.png")
+```
+
+```{r fig-corPerson}
+#| echo: false
+#| fig-pos: "!ht"
+#| out-width: "\\textwidth"
+#| fig-cap: "Pearson correlation value for the time scales and drought index that reach the maximum coefficient of determination"
+knitr::include_graphics("../output/figs/mapa_cor_r_indices_zcNDVI6.png")
+```
 

### 2024-01-08 article/paper_drought_Chile2023.tex
last update

@@
 \hypersetup{
-  pdftitle={Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from the ERA5-Land and MODIS},
+  pdftitle={Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from ERA5-Land and MODIS},
   pdfauthor={Francisco Zambrano},
@@
 \title{Assessment of drought impact over land cover in continental Chile
-by the analysis of water supply and demand from the ERA5-Land and MODIS}
+by the analysis of water supply and demand from ERA5-Land and MODIS}
 \author[1]{Francisco Zambrano%
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, breakable, enhanced]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, boxrule=0pt, interior hidden, breakable]}{\end{tcolorbox}}\fi
 
@@
 and supply gathered from earth observation products. The specific
-objectives of the study are: i) to calculate multi-scalar drought
-indices for water demand and supply for 1981--2023; ii) to evaluate LULC
-change for 2001--2021 and its relation to drought indices; iii) to
-analyze the relationship of a proxy of biomass (zcNDVI) with drought
-indices; and iv) to assess if the observed changes in the drought
-indices are linked to TWS.
+objectives of the study are: i) to analize the trend on multi-scalar
+drought indices for water demand/supply, soil moisture and vegetation
+productivity for 1981--2023/2001-2023; ii) to evaluate LULC change for
+2001--2021 and its relation to drought indices; iii) to analyze the
+relationship of a proxy of biomass (zcNDVI) with drought indices; and
+iv) to assess if the observed changes in the drought indices are linked
+to TWS.
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[!ht]
 
-{\centering \includegraphics[width=0.8\textwidth,height=\textheight]{../output/figs/map_study_con_landcover.png}
+{\centering \includegraphics{../output/figs/map_study_con_landcover.png}
 
@@
 
-\hypertarget{drought-indices-for-water-demand-and-supply}{%
-\subsection{Drought indices for water demand and
-supply}\label{drought-indices-for-water-demand-and-supply}}
+\hypertarget{drought-indices}{%
+\subsection{Drought indices}\label{drought-indices}}
+
+\hypertarget{section}{%
+\subsection{}\label{section}}
+
+\hypertarget{drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}{%
+\subsection{Drought indices for water demand/supply, soil moisture and
+vegetation
+productivity}\label{drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}}
 
@@
 
-!{[}Proportion of land cover class from the persistent land cover for
-2001-2021 (\textgreater80\%) per macrozone.
+\begin{figure}[!ht]
 
-\begin{figure}
-
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics[width=7cm,height=\textheight]{../output/figs/LC_pers80_per_macrozone.png}
+{\centering \includegraphics{../output/figs/LC_pers80_per_macrozone.png}
 
@@
 
-}
-
-\end{minipage}%
-\newline
-\begin{minipage}[t]{\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics[width=8cm,height=\textheight]{../output/figs/table_var_landcover_macro.png}
-
-}
-
-}
-
-\end{minipage}%
-
-\caption{\label{fig-elephants}Proportion of land cover class from the
-persistent land cover for 2001-2021 (\textgreater80\%) per macrozone.
-The table on the left shows the linear change trend next to time-series
-plot of surface, per land cover class (IGBP MCD12Q1.006) for 2001-2021
-through the fives zones on continental Chile. Blue dots on the plots
-indicate the maximum and red dots minimum surface reach.}
+\caption{Proportion of land cover class from the persistent land cover
+for 2001-2021 (\textgreater80\%) per macrozone}
 
 \end{figure}
+
+\begin{table}[!ht]
+\caption{Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile.
+Red dots on the plots indicate the maximum and minimum surface}
+\includegraphics[]{../output/figs/table_var_landcover_macro.png}
+\end{table}
 
@@
 filter that reduces signal leakage errors across coastlines
-\citep{Wiese2019, Wiese2016}. Although the mascon solutions greatly
+\citep{NASA2019, Wiese2016}. Although the mascon solutions greatly
 reduced leakage errors, a gain factor, which is used to enhance the
@@
 
-\hypertarget{drought-indices-for-water-demand-and-supply-1}{%
-\subsection{Drought indices for water demand and
-supply}\label{drought-indices-for-water-demand-and-supply-1}}
+\hypertarget{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}{%
+\subsection{Trend of drought indices for water demand/supply, soil
+moisture, and vegetation
+productivity}\label{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}}
 
-\hypertarget{biomass-proxy-with-drought-indices-of-supply-and-demand}{%
-\subsection{Biomass proxy with drought indices of supply and
-demand}\label{biomass-proxy-with-drought-indices-of-supply-and-demand}}
+Regarding vegetation productivity, there is a negative trend on zcNDVI
+in the central part of the country (``norte chico'' and ``zona
+central'') (Figure~\ref{fig-zcNDVI_var}). Indicating either a reduction
+of vegetation surface, biomass, or status. It reached its lowest values
+during the year 2020, mostly on the ``zona central.'' There are other
+negative trends in small areas in ``norte grande'' and ``zona asutral.''
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/temporal_variation_zcNDVI6_macrozonas.png}
+
+}
+
+\caption{\label{fig-zcNDVI_var}(a) Map of the linear trend of the index
+zcNDVI-6 for 2001--2023. Greener colors indicate a positive trend and an
+increase in vegetation development; reder colors correspond to a
+negative trend and a decrease in vegetation development. (b) Temporal
+variation of zcNDVI-6 aggregated at macrozone level within continental
+Chile. Each horizontal panel corresponds to a macrozone from `norte
+grande' to `zona austral'.}
+
+\end{figure}
+
+Analyzing the water supply in the macrozones ``norte chico,''
+``central,'' and ``sur,'' the SPI shows a decreasing trend that
+increases at longer time scales due to the prolonged reduction in
+precipitation. It reaches a trend of -0.056 (z-score) per decade
+(Figure~\ref{fig-trendSPI}) for time scales of 36 months. In the case of
+water demand, we analyze the EDDI. It shows a positive trend, caused by
+an increase in AED. The trend on EDDI reaches a maximum of 0.053 per
+decade in the macrozones ``norte grande'' and ``norte chico''
+(Figure~\ref{fig-trendEDDI}). The behavior is similar to SPI, having
+close trends with opposite signs. Nevertheless, the spatial patterns are
+different. The maximum trend for SPI-36 is in ``norte chico'' and ``zona
+central.'' For EDDI-36, the maximum trend is from ``northe grande'' to
+``zona central.''
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}
+
+}
+
+\caption{\label{fig-trendSPI}Linear trend of the SPI at time scales of
+1, 3, 6, 12, 24, and 36 months for 1981-2023}
+
+\end{figure}
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}
+
+}
+
+\caption{\label{fig-trendEDDI}Linear trend of the EDDI at time scales of
+1, 3, 6, 12, 24, and 36 months for 1981-2023}
+
+\end{figure}
+
+The correlation between zcNDVI-6 and drought indices shows that time
+scales higher than six months are predominant over continental Chile in
+explaining vegetation variability (Figure~\ref{fig-corTimeScale}).
+Pearson correlations between 0.6 and 0.9 are concentrated in the ``norte
+chico'' and ``zona central'' for SPI and SPEI. Soil moisture reached a
+correlation of 0.6--0.9 that extended north and south with respect to
+SPI.
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/mapa_cor_selec_indices_zcNDVI6.png}
+
+}
+
+\caption{\label{fig-corTimeScale}Time scales per drought index that
+reach the maximum coefficient of determination}
+
+\end{figure}
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/mapa_cor_r_indices_zcNDVI6.png}
+
+}
+
+\caption{\label{fig-corPerson}Pearson correlation value for the time
+scales and drought index that reach the maximum coefficient of
+determination}
+
+\end{figure}
 

### 2024-01-08 conference/IGARSS2024.qmd
last update

@@
 date: 2024-01-05
-pdf: https://github.com/dfolio/quarto-ieee/blob/main/template.pdf
-citation: 
-  container-title: GitHUB
-  page: 1-3
-  type: software
-  issued: 2023-06-23
-  url: https://github.com/dfolio/quarto-ieee
-  pdf-url: https://github.com/dfolio/quarto-ieee/template.pdf
 ---

### 2024-01-08 scripts/25_trend_drought_indices.R

@@ -3,9 +3,39 @@ library(terra)
 library(tmap)
 library(sf)
 library(rnaturalearth)
+library(fs)
 
 chl_b <- ne_countries(country='chile',scale = 'medium',returnclass = 'sf')
 macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') 
+
+## Time series of zcNDVI for macrozones
+dir <- '/mnt/md0/raster_procesada/MODIS_derived/zcNDVI/zcNDVI-6'
+
+zcndvi <- rast(dir_ls(dir,regexp = 'tif$'))
+macro_mask <- rasterize(st_transform(macro,32719),zcndvi[[1]],field ='macrozona')
+
+df_zcndvi <- zonal(zcndvi,macro_mask,fun = 'mean',na.rm = TRUE)
+df_zcndvi |> 
+  pivot_longer(-macrozona) |> 
+  mutate(dates = str_extract(name,'[0-9]{4}-[0-9]{2}-[0-9]{2}'),
+         dates = ymd(dates),
+         macrozona = fct(macrozona,c('norte grande','norte chico','zona central','zona sur','zona austral'))) |> 
+  select(-name) |> 
+  drop_na() |> 
+  ggplot(aes(dates,value)) +
+  #geom_point(size=.2) + 
+  #geom_line(size = .2) +
+  geom_ribbon(aes(ymin = pmin(value,0),ymax = 0),fill = 'red',alpha=.7)+
+  geom_ribbon(aes(ymin = 0,ymax = pmax(value,0)),fill = 'darkgreen',alpha=.7)+
+  geom_hline(yintercept = 0,col = 'red') +
+  scale_x_date(date_breaks = "2 years", date_labels = "%Y",limits = c(as.Date("2000-01-01"),as.Date("2023-04-01")),expand = c(0,0)) +
+  geom_smooth(method = 'lm',se = FALSE,alpha=.8,lty = 'dashed',col='grey') +
+  labs(y = 'zcNDVI') +
+  facet_grid(macrozona~.) +
+  theme_minimal() +
+  theme(axis.title.x = element_blank())
+
+ggsave('output/figs/temporal_variation_zcNDVI6_macrozonas.png',scale = 1.5,bg = 'white',width=6,height=3)
 
 trend_spi <- rast('/mnt/md0/raster_procesada/analysis/trends/trend_SPI_1981-2023.tif')
[…]

### 2024-01-11 article/paper_drought_Chile2023.qmd
agregar mapa koppe-geiger al área de estudio

@@
 
-In this research, we aim to analyze the impact of drought on different types of land cover classes in continental Chile by combining environmental variables such as biomass productivity, and water demand and supply gathered from earth observation products.
-The specific objectives of the study are: i) to analize the trend on multi-scalar drought indices for water demand/supply, soil moisture and vegetation productivity for 1981–2023/2001-2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of biomass (zcNDVI) with drought indices; and iv) to assess if the observed changes in the drought indices are linked to TWS.
+In this research, we present the dataset DDS4Chl, which provides climate variables and drought indices of water demand and  supply and vegetation productivity for continental Chile since 1981 at a monthly frequency.
+Those were gathered from the earth observation products ERA5-Land and MODIS.
+Then, we used DDS4Chl to analyze the impact of drought on different types of land cover classes in continental Chile.
+The specific objectives of the study are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of vegetation productivity (zcNDVI) with drought indices of water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.
 
@@
 
-## Drought indices
-
-##
-
-##	Drought indices for water demand/supply, soil moisture and vegetation productivity
+##	Drought indices for water demand and supply, soil moisture and vegetation productivity
 

### 2024-01-11 article/paper_drought_Chile2023.tex
agregar mapa koppe-geiger al área de estudio

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, boxrule=0pt, interior hidden, breakable]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 (EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI;
-~\citep{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
+\citep{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
 cumulative Normalized Difference Vegetation Index (NDVI), which could be
@@
 
-In this research, we aim to analyze the impact of drought on different
-types of land cover classes in continental Chile by combining
-environmental variables such as biomass productivity, and water demand
-and supply gathered from earth observation products. The specific
-objectives of the study are: i) to analize the trend on multi-scalar
-drought indices for water demand/supply, soil moisture and vegetation
-productivity for 1981--2023/2001-2023; ii) to evaluate LULC change for
-2001--2021 and its relation to drought indices; iii) to analyze the
-relationship of a proxy of biomass (zcNDVI) with drought indices; and
+In this research, we present the dataset DDS4Chl, which provides climate
+variables and drought indices of water demand and~ supply and vegetation
+productivity for continental Chile since 1981 at a monthly frequency.
+Those were gathered from the earth observation products ERA5-Land and
+MODIS. Then, we used DDS4Chl to analyze the impact of drought on
+different types of land cover classes in continental Chile. The specific
+objectives of the study are: i) to analyze the trend on multi-scalar
+drought indices for water demand and supply, soil moisture, and
+vegetation productivity for 1981--2023/2001--2023; ii) to evaluate LULC
+change for 2001--2021 and its relation to drought indices; iii) to
+analyze the relationship of a proxy of vegetation productivity (zcNDVI)
+with drought indices of water demand and supply and soil moisture; and
 iv) to assess if the observed changes in the drought indices are linked
@@
 
-\hypertarget{drought-indices}{%
-\subsection{Drought indices}\label{drought-indices}}
-
-\hypertarget{section}{%
-\subsection{}\label{section}}
-
-\hypertarget{drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}{%
-\subsection{Drought indices for water demand/supply, soil moisture and
-vegetation
-productivity}\label{drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}}
+\hypertarget{drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}{%
+\subsection{Drought indices for water demand and supply, soil moisture
+and vegetation
+productivity}\label{drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}}
 

### 2024-01-11 conference/IGARSS2024.qmd
agregar mapa koppe-geiger al área de estudio

@@
 
-To evaluate meteorological drought (i.e., water supply), the World Meteorological Organization (WMO; @WMO2012) recommends the Standardized Precipitation Index (SPI; @McKee1993), a multiscalar drought index that allows to monitor precipitation deficits from short- to long-term.
+To evaluate meteorological drought (i.e., water supply), the World Meteorological Organization (WMO; @WMO2012;@Hayes2011) recommends the Standardized Precipitation Index (SPI; @McKee1993), a multiscalar drought index that allows to monitor precipitation deficits from short- to long-term.
 Following the same approach, @Vicente-Serrano2010 incorporates into the SPI the effect of temperature through the use of potential evapotranspiration, thus proposing the SPEI (Standardized Precipitation Evapotranspiration Index).

### 2024-01-11 scripts/16_map_study_landcover.R

@@ -93,13 +93,40 @@ map_globe <- tm_shape(bm_world) +
   tm_text('admin',size=.4,xmod = -.5,col='black') +
   tm_layout(frame = TRUE,bg.color = "white")
 
+# Mapa Chile con Koppen-Geiger 1991-2020
+
+kp_chile <- rast('data/processed_data/spatial/koppen_geiger_1991_2020.tif')
+rgb <- coltab(kp_chile)[[1]][2:31,]
+paleta_kg <- readr::read_csv2('data/processed_data/legend_koppen-geiger.csv')
+labs <- substr(paleta_kg$type,1,3)
+names(labs) <- 1:30
+colores <- rgb(rgb[,2],rgb[,3],rgb[,4],rgb[,5],maxColor = 255)
+names(colores) <- 1:30
+
+kp_chile <- raster::raster(kp_chile)
+vals <- values(kp_chile)
+vals[vals %in% c(6,26,10,30,27,18)] <- NA
+values(kp_chile) <- vals
+
+zones$macrozona <- c('Norte Chico','Norte Grande',
+                     'Austral','Centro','Sur')
+  
 map_chl <- tm_shape(bm2) +
   tm_rgb() +
+  tm_shape(kp_chile) +
+  tm_raster(palette = colores,
+            style = 'cat',
+            labels = labs,
+            title = 'Climate (1990-2020)') +
   tm_shape(zones) +
   tm_borders(lwd=2,col='black',lty='solid',alpha=0.8) +
-  tm_text('macrozona',just='top',xmod=c(-6,-6,3,-5,-5),ymod=c(0,-3.5,0,0,0),size=2) +
-  tm_credits('(a)',position = c('left','bottom'),fontface = 'bold',size=1.7)
-  
+  tm_text('macrozona',just='top',xmod=c(-5,-5,3,-3,-3),ymod=c(0,-1.5,3,0,0),size=1) +
+  tm_credits('(a)',position = c('left','bottom'),fontface = 'bold',size=1.7) +
+  tm_layout(legend.outside = FALSE,
+            legend.text.size = 0.7)
+
+tmap_save(map_chl,'output/figs/mapa_koppen_geiger.png')
+
 xy <- st_bbox(world)
 asp2 <- (xy$xmax - xy$xmin)/(xy$ymax - xy$ymin)
 

### 2024-01-11 scripts/28_mapa_koppen-geiger.R

@@ -1,0 +1,35 @@
+library(sf)
+library(terra)
+library(fs)
+library(tidyverse)
+
+dir <- '/mnt/data_procesada/data/rasters/Procesados/koppen_geiger_tif/1991_2020'
+
+#/1991_2020'
+
+#/2041_2070/ssp585'
+
+files <- dir_ls(dir,regexp = '0p01')
+
+chile <- read_sf('data/processed_data/spatial/chile_continental.gpkg')
+kp <- rast(files)
+kp_chile <- crop(kp,chile)
+kp_chile <- mask(kp,chile)
+kp_chile <- trim(kp_chile)
+writeRaster(kp_chile,'data/processed_data/spatial/koppen_geiger_1991_2020.tif')
+library(tmap)
+
+kp_chile <- rast('data/processed_data/spatial/koppen_geiger_1991_2020.tif')
+rgb <- coltab(kp_chile)[[1]][2:31,]
+paleta_kg <- read_csv2('data/processed_data/legend_koppen-geiger.csv')
+labs <- paleta_kg$type
+names(labs) <- 1:30
+colores <- rgb(rgb[,2],rgb[,3],rgb[,4],rgb[,5],maxColor = 255)
+names(colores) <- 1:30
+
+tm_shape(kp_chile) +
+  tm_raster(palette = colores,
+            style = 'cat',
+            labels = labs,
+            title = 'Climate (1990-2020)') +
+  tm_layout(legend.outside = TRUE)

### 2024-01-11 article/paper_drought_Chile2023.qmd
completar area de estudio en articulo

@@
 # Study area
+
+Continetal Chile has a diverse climate condition from north to south and east to west, which determines its great ecosystem diversity [@Aceituno2021].
+The Andes Mountains are a main factor in longitudinal variation [@Garreaud2009].
+Acoording to the v2 of the Koppen-Geiger realease by @Beck2023 and the landcover types from the IGBP [@Friedl2019].
+In the "Norte Grande" and "Norte Chico" predominate an arid desert climate with hot (Bwh) and cold (Bwk) temperatures.
+At south of "Norte Chico" the climate change to a arid steppe with cold temperatures (Bsk).
+Mainly the land is barrend and with a minor surface of vegetation types of Shrubland (X% ) and Grassland.
 

### 2024-01-11 article/paper_drought_Chile2023.tex
completar area de estudio en articulo

@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 \section{Study area}\label{study-area}}
+
+Continetal Chile has a diverse climate condition from north to south and
+east to west, which determines its great ecosystem diversity
+\citep{Aceituno2021}. The Andes Mountains are a main factor in
+longitudinal variation \citep{Garreaud2009}. Acoording to the v2 of the
+Koppen-Geiger realease by \citep{Beck2023} and the landcover types from
+the IGBP \citep{Friedl2019}. In the ``Norte Grande'' and ``Norte Chico''
+predominate an arid desert climate with hot (Bwh) and cold (Bwk)
+temperatures. At south of ``Norte Chico'' the climate change to a arid
+steppe with cold temperatures (Bsk). Mainly the land is barrend and with
+a minor surface of vegetation types of Shrubland (X\% ) and Grassland.
 

### 2024-01-12 article/paper_drought_Chile2023.qmd
avances articulo reunion 12 enero

--- 2024-01-11 article/paper_drought_Chile2023.qmd
@@
 bibliography: references.bib
+keep-tex: false
 format:
@@
 
-In this research, we present the dataset DDS4Chl, which provides climate variables and drought indices of water demand and  supply and vegetation productivity for continental Chile since 1981 at a monthly frequency.
+In this research, we present the raster dataset DDS4Chl, which provides climate variables and drought indices of water demand and supply and vegetation productivity for continental Chile since 1981 at a monthly frequency.
 Those were gathered from the earth observation products ERA5-Land and MODIS.
@@
 
-![(a) Chile and the five zones "norte grande", "norte chico",  "zona central", "zona sur", and "zona austral".
+Continetal Chile has a diverse climate condition from north to south and east to west [@Aceituno2021] (@fig-studyArea), which determines its great ecosystem diversity (@fig-LCprop).
+The Andes Mountains are a main factor in latitudinal variation [@Garreaud2009].
+To describe the climate and ecosystem of Chile, we use the Koppen-Geiger release by @Beck2023 and the landcover type persistance of 80% of times for 2001–2022, from the IGBP classification scheme [@Friedl2019].
+"Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures.
+At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk).
+Mainly, the land is barren, with a minor surface of vegetation types such as shrubland and grassland.
+In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
+In "Centro," there is a major proportion (~50%) of chilean matorral (shrubland and savanna), followed by grassland (~16%), forest (8%), and croplands (5%).
+The south part of "Centro" and the north part of "Austral" are dominated by an Oceanic climate (Cfb).
+Those zones are high in forest and grassland.
+The southern part of the country has a tundra climate, and in Patagonia, it is a cold semi-arid area with an extended surface of grassland.
+
+```{r fig-studyArea}
+#| echo: false
+#| fig-pos: "!ht"
+#| fig-cap: '(a) Chile with the Koppen-Geiger climate classes and the five macrozones "Norte Grande", "Norte Chico",  "Centro", "Sur", and "Austral".
 (b) Topography reference map.
-(c) Land cover classes for 2021.
-(d) Persistent land cover classes (> 80%) for 2001-2021.](../output/figs/map_study_con_landcover.png){fig-pos='!ht'}
+(c) Land cover classes for 2022.
+(d) Persistent land cover classes (> 80%) for 2001-2022'
+#| out-width: "\\textwidth"
+knitr::include_graphics("../output/figs/map_study_con_landcover.png")
+```
 
@@
 
-For water supply and demand variabes, we used ERA5-Land [@Munoz2019], a reanalys dataset wich provides the evolution of land variables since 1950.
+For water supply and demand variabes, we used ERA5-Land [@MunozSabater2021], a reanalys dataset wich provides the evolution of land variables since 1950.
 It has a spatial resolution of 0.1°, hourly frequency and global coverage.
@@
 
-##	Drought indices for water demand and supply, soil moisture and vegetation productivity
+## Trend of drought indices for water demand/supply, soil moisture, and vegetation productivity
 
-##	Analysis of a biomass proxy with drought indices of supply and demand
+##	Impact fof water supply and demand, and soil moisture in vegetation productivity within landcover types
 
-##	LULC change for 2001-2021 and its relation with water supply and demand
+##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture
 
-```{r LCprop}
+```{r fig-LCprop}
 #| echo: false
@@
 ```
-
-\begin{table}[!ht]
-\caption{Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile.
-Red dots on the plots indicate the maximum and minimum surface}
-\includegraphics[]{../output/figs/table_var_landcover_macro.png}
-\end{table}
 
@@
 
+##	Impact for water supply and demand, and soil moisture in vegetation productivity
+
 ```{r fig-corTimeScale}
@@
 
-##	LULC change for 2001-2021 and its relation with water supply and demand
+##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture
+
+\begin{table}[!ht]
+\label{tab-landcoverTrend}
+\caption{Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile.
+Red dots on the plots indicate the maximum and minimum surface}
+\includegraphics[]{../output/figs/table_var_landcover_macro.png}
+\end{table}
+
+\begin{table}[!ht]
+\label{tab-corLandcover}
+\caption{}
+\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
+\end{table}
 

### 2024-01-12 article/paper_drought_Chile2023.tex
avances articulo reunion 12 enero

--- 2024-01-11 article/paper_drought_Chile2023.tex
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners]}{\end{tcolorbox}}\fi
+    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-In this research, we present the dataset DDS4Chl, which provides climate
-variables and drought indices of water demand and~ supply and vegetation
-productivity for continental Chile since 1981 at a monthly frequency.
-Those were gathered from the earth observation products ERA5-Land and
-MODIS. Then, we used DDS4Chl to analyze the impact of drought on
-different types of land cover classes in continental Chile. The specific
-objectives of the study are: i) to analyze the trend on multi-scalar
-drought indices for water demand and supply, soil moisture, and
-vegetation productivity for 1981--2023/2001--2023; ii) to evaluate LULC
-change for 2001--2021 and its relation to drought indices; iii) to
+In this research, we present the raster dataset DDS4Chl, which provides
+climate variables and drought indices of water demand and supply and
+vegetation productivity for continental Chile since 1981 at a monthly
+frequency. Those were gathered from the earth observation products
+ERA5-Land and MODIS. Then, we used DDS4Chl to analyze the impact of
+drought on different types of land cover classes in continental Chile.
+The specific objectives of the study are: i) to analyze the trend on
+multi-scalar drought indices for water demand and supply, soil moisture,
+and vegetation productivity for 1981--2023/2001--2023; ii) to evaluate
+LULC change for 2001--2021 and its relation to drought indices; iii) to
 analyze the relationship of a proxy of vegetation productivity (zcNDVI)
@@
 
+Continetal Chile has a diverse climate condition from north to south and
+east to west \citep{Aceituno2021} (Figure~\ref{fig-studyArea}), which
+determines its great ecosystem diversity (Figure~\ref{fig-LCprop}). The
+Andes Mountains are a main factor in latitudinal variation
+\citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
+use the Koppen-Geiger release by \citep{Beck2023} and the landcover type
+persistance of 80\% of times for 2001--2022, from the IGBP
+classification scheme \citep{Friedl2019}. ``Norte Grande'' and ``Norte
+Chico'' predominate in an arid desert climate with hot (Bwh) and cold
+(Bwk) temperatures. At the south of ``Norte Chico,'' the climate changes
+to an arid steppe with cold temperatures (Bsk). Mainly, the land is
+barren, with a minor surface of vegetation types such as shrubland and
+grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
+main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
+In ``Centro,'' there is a major proportion (\textasciitilde50\%) of
+chilean matorral (shrubland and savanna), followed by grassland
+(\textasciitilde16\%), forest (8\%), and croplands (5\%). The south part
+of ``Centro'' and the north part of ``Austral'' are dominated by an
+Oceanic climate (Cfb). Those zones are high in forest and grassland. The
+southern part of the country has a tundra climate, and in Patagonia, it
+is a cold semi-arid area with an extended surface of grassland.
+
 \begin{figure}[!ht]
@@
 
-\caption{(a) Chile and the five zones ``norte grande'', ``norte chico'',
-``zona central'', ``zona sur'', and ``zona austral''. (b) Topography
-reference map. (c) Land cover classes for 2021. (d) Persistent land
-cover classes (\textgreater{} 80\%) for 2001-2021.}
+\caption{\label{fig-studyArea}(a) Chile with the Koppen-Geiger climate
+classes and the five macrozones ``Norte Grande'', ``Norte Chico'',
+``Centro'', ``Sur'', and ``Austral''. (b) Topography reference map. (c)
+Land cover classes for 2022. (d) Persistent land cover classes
+(\textgreater{} 80\%) for 2001-2022}
 
@@
 For water supply and demand variabes, we used ERA5-Land
-\citep{Munoz2019}, a reanalys dataset wich provides the evolution of
-land variables since 1950. It has a spatial resolution of 0.1°, hourly
-frequency and global coverage. We selected the variables for total
-precipitation, 2 meters temperature, and volumetric soil water layers
-between 0 and 100cm of depth (layer 1 to layer 3).
+\citep{MunozSabater2021}, a reanalys dataset wich provides the evolution
+of land variables since 1950. It has a spatial resolution of 0.1°,
+hourly frequency and global coverage. We selected the variables for
+total precipitation, 2 meters temperature, and volumetric soil water
+layers between 0 and 100cm of depth (layer 1 to layer 3).
 
@@
 
-\hypertarget{drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}{%
-\subsection{Drought indices for water demand and supply, soil moisture
-and vegetation
-productivity}\label{drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}}
+\hypertarget{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}{%
+\subsection{Trend of drought indices for water demand/supply, soil
+moisture, and vegetation
+productivity}\label{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}}
 
-\hypertarget{analysis-of-a-biomass-proxy-with-drought-indices-of-supply-and-demand}{%
-\subsection{Analysis of a biomass proxy with drought indices of supply
-and
-demand}\label{analysis-of-a-biomass-proxy-with-drought-indices-of-supply-and-demand}}
+\hypertarget{impact-fof-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-landcover-types}{%
+\subsection{Impact fof water supply and demand, and soil moisture in
+vegetation productivity within landcover
+types}\label{impact-fof-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-landcover-types}}
 
-\hypertarget{lulc-change-for-2001-2021-and-its-relation-with-water-supply-and-demand}{%
-\subsection{LULC change for 2001-2021 and its relation with water supply
-and
-demand}\label{lulc-change-for-2001-2021-and-its-relation-with-water-supply-and-demand}}
+\hypertarget{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}{%
+\subsection{LULC change for 2001-2022 and its relation with water supply
+and demand, and soil
+moisture}\label{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}}
 
@@
 
-\caption{Proportion of land cover class from the persistent land cover
-for 2001-2021 (\textgreater80\%) per macrozone}
+\caption{\label{fig-LCprop}Proportion of land cover class from the
+persistent land cover for 2001-2021 (\textgreater80\%) per macrozone}
 
 \end{figure}
-
-\begin{table}[!ht]
-\caption{Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile.
-Red dots on the plots indicate the maximum and minimum surface}
-\includegraphics[]{../output/figs/table_var_landcover_macro.png}
-\end{table}
 
@@
 
-\hypertarget{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}{%
+\hypertarget{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity-1}{%
 \subsection{Trend of drought indices for water demand/supply, soil
 moisture, and vegetation
-productivity}\label{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}}
+productivity}\label{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity-1}}
 
@@
 
+\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}{%
+\subsection{Impact for water supply and demand, and soil moisture in
+vegetation
+productivity}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}}
+
 \begin{figure}[!ht]
@@
 
-\hypertarget{lulc-change-for-2001-2021-and-its-relation-with-water-supply-and-demand-1}{%
-\subsection{LULC change for 2001-2021 and its relation with water supply
-and
-demand}\label{lulc-change-for-2001-2021-and-its-relation-with-water-supply-and-demand-1}}
+\hypertarget{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture-1}{%
+\subsection{LULC change for 2001-2022 and its relation with water supply
+and demand, and soil
+moisture}\label{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture-1}}
+
+\begin{table}[!ht]
+\label{tab-landcoverTrend}
+\caption{Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile.
+Red dots on the plots indicate the maximum and minimum surface}
+\includegraphics[]{../output/figs/table_var_landcover_macro.png}
+\end{table}
+
+\begin{table}[!ht]
+\label{tab-corLandcover}
+\caption{}
+\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
+\end{table}
 

### 2024-01-12 scripts/16_map_study_landcover.R

@@ -34,8 +34,8 @@ igbp80 <- mask(igbp80,chl)
 #igbp80[igbp80 == 10] <- NA
 
 #landcover año 2021
-igbp2021 <- rast(paste0(dir,'IGBP_2021_reclassified.tif'))
-igbp2021 <- mask(igbp2021,chl)
+igbp2022 <- rast(paste0(dir,'IGBP_2022_reclassified.tif'))
+igbp2022 <- mask(igbp2022,chl)
 
 paleta <- read.csv('data/processed_data/paleta_colores_landcover.csv')
 colors <-  rgb(paleta$R,paleta$G,paleta$B,maxColorValue = 255)
@@ -46,7 +46,7 @@ m2 <- tm_shape(bm) +
   tm_credits('(c)',position = c('left','bottom'),fontface = 'bold') +
   # tm_shape(divisiones) + 
   # tm_lines(lwd=1,col='black',lty='solid',alpha=.6)+
-  tm_shape(igbp2021) +
+  tm_shape(igbp2022) +
   tm_raster(style = 'cat',
             palette = colors,title = 'Landcover classes',
             labels = paleta$Name) +
@@ -120,12 +120,13 @@ map_chl <- tm_shape(bm2) +
             title = 'Climate (1990-2020)') +
   tm_shape(zones) +
   tm_borders(lwd=2,col='black',lty='solid',alpha=0.8) +
-  tm_text('macrozona',just='top',xmod=c(-5,-5,3,-3,-3),ymod=c(0,-1.5,3,0,0),size=1) +
+  tm_text('macrozona',just='top',xmod=c(5,-7,3,-4,-4),ymod=c(0,-3.5,1,0,0),size=1.5) +
   tm_credits('(a)',position = c('left','bottom'),fontface = 'bold',size=1.7) +
   tm_layout(legend.outside = FALSE,
-            legend.text.size = 0.7)
+            legend.text.size = 1.2,
+            legend.position = c(0.1,0.7))
 
-tmap_save(map_chl,'output/figs/mapa_koppen_geiger.png')
+#tmap_save(map_chl,'output/figs/mapa_koppen_geiger.png')
 
 xy <- st_bbox(world)
 asp2 <- (xy$xmax - xy$xmin)/(xy$ymax - xy$ymin)
@@ -139,6 +140,6 @@ w <- 0.5
 h <- asp2 * w
 vp <- viewport(x=.53, y=1.43, width = w, height=h, just=c("right", "top"))
 
[…]

### 2024-01-12 scripts/17_extrarer_correlations_macro_landcover.R

@@ -31,9 +31,9 @@ getmode <- function(v) {
 }
 
 #obtener los indices y escalas de tiempo en dónde se alcanza la máxima correlación
-cors <- subset(cors,seq(1,8,2))
+cors1 <- subset(cors,seq(1,8,2))
 data_ind <- map_df(1:5,function(i){
-  cors_m <- mask(cors,macro[i,])
+  cors_m <- mask(cors1,macro[i,])
   lc_m <- mask(lc,macro[i,])
   cors_df <- zonal(cors_m,lc_m,getmode)
   names(cors_df)[2:5] <- c('eddi','spi','spei','zcsm')
@@ -45,7 +45,7 @@ data_ind <- map_df(1:5,function(i){
     select(macro,indice,clase,value) 
 })
 
-data_ind |> 
+data_ind_4gt <- data_ind |> 
   filter(clase %in% c('Forest','Shrubland','Savanna','Grassland','Cropland','Barren land')) |> 
   mutate(value = ceiling(value),
          value = factor(value,levels = 1:6,labels = c(1,3,6,12,24,36)),
@@ -56,26 +56,13 @@ data_ind |>
   #mutate(row=row_number()) |> 
   pivot_longer(-c(clase,macro))  |> 
   mutate(clase = str_to_title(clase)) |> 
-  pivot_wider(names_from=c(clase, name), values_from=value) |> 
-  gt() %>% 
-  #opt_stylize(style = 6, color = 'gray')
-  tab_spanner_delim(
-    delim="_"
-  ) %>% 
-  fmt_missing(
-    columns=everything(),
-    missing_text=""
-  ) |> 
-  gtsave('output/figs/tabla_best_cor_macro_indice.png',expand=20,vwidth=1200)
-
-  gt(groupname_col = "macro") |> 
-  opt_stylize(style = 6, color = 'gray')
+  pivot_wider(names_from=c(clase, name), values_from=value) 
 
 #obtener los valores de correlación para los indices y escalas de tiempo en dónde se alcanza la máxima correlación
   
-cors <- subset(cors,seq(2,8,2))
+cors2 <- subset(cors,seq(2,8,2))
 data_r <- map_df(1:5,function(i){
-  cors_m <- mask(cors,macro[i,])
+  cors_m <- mask(cors2,macro[i,])
   lc_m <- mask(lc,macro[i,])
   cors_df <- zonal(cors_m,lc_m,na.rm = TRUE)
   names(cors_df)[2:5] <- c('eddi','spi','spei','zcsm')
[…]

### 2024-01-12 scripts/18_extraer_correlations_macro_landcover_meses.R

@@ -107,8 +107,13 @@ data_r_4gt <- data_r |>
   pivot_wider(names_from=c(clase, name), values_from=value) |> 
   rename_with(\(x) str_c(x,'_r'),-zone)
 
-tabla_gt <- cbind(data_ind_4gt,data_r_4gt |>  mutate(zone = fct_relevel(zone, "norte grande", "norte chico",
-                                                                  'zona central','zona sur')) |> arrange(zone)) |> 
+tabla_gt <- cbind(data_ind_4gt,data_r_4gt |>  
+                    mutate(zone = fct_relevel(zone, 
+                                              "norte grande", 
+                                              "norte chico",
+                                              'zona central',
+                                              'zona sur')) |> 
+                    arrange(zone)) |> 
   select(-zone) |> 
   rename(zone = macro)
 

### 2024-01-12 scripts/25_trend_drought_indices.R

@@ -6,7 +6,8 @@ library(rnaturalearth)
 library(fs)
 
 chl_b <- ne_countries(country='chile',scale = 'medium',returnclass = 'sf')
-macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') 
+macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
+  mutate(macrozona = c('Norte Chico','Norte Grande','Austral','Centro','Sur'))
 
 ## Time series of zcNDVI for macrozones
 dir <- '/mnt/md0/raster_procesada/MODIS_derived/zcNDVI/zcNDVI-6'
@@ -19,7 +20,8 @@ df_zcndvi |>
   pivot_longer(-macrozona) |> 
   mutate(dates = str_extract(name,'[0-9]{4}-[0-9]{2}-[0-9]{2}'),
          dates = ymd(dates),
-         macrozona = fct(macrozona,c('norte grande','norte chico','zona central','zona sur','zona austral'))) |> 
+         macrozona =str_to_title(macrozona),
+         macrozona = fct(macrozona,c('Norte Grande','Norte Chico','Centro','Sur','Austral'))) |> 
   select(-name) |> 
   drop_na() |> 
   ggplot(aes(dates,value)) +
@@ -37,33 +39,46 @@ df_zcndvi |>
 
 ggsave('output/figs/temporal_variation_zcNDVI6_macrozonas.png',scale = 1.5,bg = 'white',width=6,height=3)
 
-trend_spi <- rast('/mnt/md0/raster_procesada/analysis/trends/trend_SPI_1981-2023.tif')
+#Mapa de tendencia Mann-Kendall
+library(fs)
+dir <- '/mnt/md0/raster_procesada/analysis/trends/'
+files <- dir_ls(dir,regexp = 'mann.*SPI.*tif$')
+scales <- str_extract(basename(files),'-.*\\.') |> str_remove('-') |> str_remove('\\.')
 
-macro_ras <- rasterize(macro,trend_spi,field = 'macrozona') 
+trend_spi <- rast(files)
+trend_spi <- app(trend_spi,\(x){
+  i <- seq(1,12,2)
+  x[i]*x[i+1]
+})
 
[…]

### 2024-01-15 article/paper_drought_Chile2023.qmd
avanzar en paper metodologia

@@
 ---
-title: Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from ERA5-Land and MODIS
+title: Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile
 #subtitle: A Short Subtitle
@@
 keep-tex: false
+include-before-body:
+  - text: |
+      \captionsetup{justification=raggedright,singlelinecheck=false}
 format:
@@
       formatting: preprint
+      layout: onecolumn
       model: 3p
@@
     include-in-header: 
-      text: '\usepackage{lineno}\linenumbers'
-      #cite-style: super
+      text: '\usepackage{lineno}\linenumbers
+             \usepackage{multirow}'
 ---
@@
 The Andes Mountains are a main factor in latitudinal variation [@Garreaud2009].
-To describe the climate and ecosystem of Chile, we use the Koppen-Geiger release by @Beck2023 and the landcover type persistance of 80% of times for 2001–2022, from the IGBP classification scheme [@Friedl2019].
+To describe the climate and ecosystem of Chile, we use the Koppen-Geiger release by @Beck2023 and the landcover type persistance of 80% of times for 2001–2022, from the IGBP classification scheme [@Friedl2019] (see @sec-methods_lulc).
 "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures.
@@
 It has a spatial resolution of 0.1°, hourly frequency and global coverage.
-We selected the variables for total precipitation, 2 meters temperature, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).
+We selected the variables for total precipitation, 2 meters temperature maximum and minimum, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).
 
@@
 
+```{=latex}
+\begin{table}[!ht]
+\caption{Description of the earth observation data used }
+\label{tab-desEOD}
+\small
+\centering
+\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
+\hline
+\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
+\hline
+\multirow{4}{*}{ERA5-Land} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
+         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
+         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
+         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
+ERA5-Land* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
+        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
+         & MCD12Q1.061 & Landcover IGBP scheme & & 2001-2022 & ~ & Landcover \\ 
+\hline
+\end{tabular}
+{\raggedright *Derived from ERA5-Land with Eq. \ref{eq-AED}. \par}
+\end{table}
+```
+
 ###	in-situ data
 
-## Trend of drought indices for water demand/supply, soil moisture, and vegetation productivity
+## Software and packages used
 
-##	Impact fof water supply and demand, and soil moisture in vegetation productivity within landcover types
+For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, `R` [@R2023].
+For downloading ERA5-Land, we used the `{ecmwfr}` package [@Hufkens2019].
+For processing raster data, we used `{terra}` [@Hijmans2023] and `{stars}` [@Pebesma2023].
+For managing vectorial data, we used `{sf}` [@Pebesma2018].
+For the calculation of AED, we used `{SPEI}` [@Bergueria2023].
 
-##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture
+## Drought Indices
+
+We derived the drought indices of water supply and demand, soil moisture from the ERA5-Land dataset, and vegetation from the MODIS product, all at monthly frequency.
+
+For the indices EDDI and SPEI that use water demand, first we have to calculate the AED. For this, we used the method of Hargreaves [@Hargreaves1994]:
+
+$$AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}$$ {#eq-AED}
+
+where $Ra$ $(MJ\,m^2\, day^{-1})$ is extraterrestrial radiation;  $T$, $T_{max}$, and $T_{min}$ are mean, maximum, and minimum temperature $(°C)$.
+We calculate the centroid coordinates per pixel and use the latitude to estimate $Ra$.
+
+To evaluate water demand, we chose the $EDDI$ [@Hobbins2016; @McEvoy2016] index, which uses the $AED$.
+For supply, we used the index recommended by the World Meteorological Organization (WMO) for monitoring drought, the SPI [@Mckee1993].
+We calculated the SPEI, which used a balance between $P$ and $AED$, in this case, an auxiliary variable $D = P-AED$ is used.
+In this study, we propose the $zcSM$ (standardized anomaly of acumulated soil moisture at 1 m), which uses SM.
+Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI.
+Before using the NDVI, it was smoothed following the procedure described in @Zambrano2018 and @Zambrano2016.
+
+All the indices are multi-scalar and were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was calculated for 6 months.
+The goal is to be able to evaluate short- and long-term droughts in water demand and supply and soil moisture.
+This is particularly important for central Chile because it has suffered from a prolonged decrease in precipitation for more than 12 years [@Garreaud2020;@Boisier2018;@Garreaud2017].
+
+To calculate the drought indices, first we must calculate the accumulation of the variable.
+In this case, for generalization purposes, we will use $V$, referring to $P$, $AED$, $D$, $NDVI$, and $SM$ (Table \ref{tab-desEOD}).
+We cumulated each $V$ over the time series of $n$ values, and for the time scales $s$:
+
+$$A_{si} = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  $$ {#eq-sumvar}
+
+It corresponds to a moving window (convolution) that sums the variable for $s$ starting for the last month $n$ until the month, which could sum for $s$ months (n-s+1).
+Once the variable is cumulated over time for the scale $s$, we used a nonparametric approach following @Hobbins2016 to derive the drought indices.
+Thus, the empirically derived probabilities are obtained through an inverse normal approximation [@Abramowitz1968].
+Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i$ to derive the $P(A_i)$ probabilities across a period of interest:
+
+$$P(A_i) = \frac{i-0.33}{n+0.33'}$$ {#eq-probPai}
+
+The drought indices $SPI$, $SPEI$, $EDDI$, $zcSM$, and $zcNDVI$ are obtained following the inverse normal approximation:
+
+$$DI(A_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}$$ {#eq-DI}
+
+$DI$ is referring to the drought index calculated for the variable $V$.
+The values for the constats are: $C_0 = 2.515517$, $C_1 = 0.802853$, $C_2 = 0.010328$, $d_1 = 1.432788$, $d_2 = 0.189269$, and $d3 = 0.001308$.
+For $P(A) \leq 0.5$, W=$\sqrt{-2\cdot ln(P(A_i))}$ , and for $P(A_i) > 0.5$, replace $P(A_i)$ with $1-P(A_i)$ and reverse the sign of $DI(A_i)$.
+
+## Trend of drought indices for water demand and supply, soil moisture, and vegetation productivity
+
+To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975].
+To determine the magnitude of the trend, we used Sen's slope [@Sen1968].
+One of the advantages of applying this methodology is that the Sen's slope is not affected by outliers as regular regression does.
+We applied both to the five indices at the six time scales from 1981 to 2023 and indices SPI, EDDI, SPEI, and zcSM, and from 2000 to 2023 in the case of zcNDVI.
+
+@Vicente-Serrano2021 made a global analysis of the drought's severity trend using SPI, SPEI, and the Standardized Evapotranspiration Deficit Index (SEDI; @Vicente-Serrano2018) to evaluate AED.
+They indicate that the increase in hydrological drought has been due to anthropogenic effects rather than climate change.
+This is because the global increase in AED does not explain the change in the spatial pattern of hydrological drought.
+Also, they state that _"the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change"_.
+We will contrast this hypothesis with what is occurring in Chile.
+To achieve this, we will use the landcover class type that remain more than 80% of types for 2001-2022, and use this as a mask, where there are low changes, to evaluate the trend on zcNDVI.
+
+##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture {#sec-methods_lulc}
 
@@
 #| fig-pos: "!ht"
-#| fig-cap: "Proportion of land cover class from the persistent land cover for 2001-2021 (>80%) per macrozone"
+#| fig-cap: "Proportion of land cover class from the persistent land cover for 2001-2022 (>80%) per macrozone"
 #| out-width: "\\textwidth"
@@
 
-##	Total water storage and drought indices
+To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS.
+This product has a yearly frequency from 2011 to 2022.
+The IGBP defines 17 classes; from these, we regrouped into ten macroclasses for forest, schrublands, savannas, grasslands, wetlands, croplands, urban, snow and ice, barren, and water bodies.
+Then, we calculated the surface ocuppied per landcover class into each macrozone ("Norte Grande" to "Austral") per year.
+To determine the trend of change we used the Sen' slope [@Sen1968] over the surface per year.
 
-Terrestrial Water Storage (TWS) is defined as the total amount of water stored on land that includes any natural or artificial water bodies, such as groundwater, soil moisture, rivers, lakes, snowpack, ice, and biomass water [@Humphrey2023; @Deng2023].
+We calculated a raster mask considering the classes that remains more than 80% of years (2001-2022), which allow us to identify the areas with no landcover change for the macroclasses.
+@fig-LCprop shows the summary of proportion of surface per landcover class and macrozone, derived from the mask over continental Chile.
+
+We will explore if the trend in landcover classess is associated to trend of the drought indices.
+For this we will use the regularizations techniques of Lasso [@Tibshirani2010] and Ridge regressión [@Hoerl1970].
+Also, we will test random forest for this purpose [@Ho1995].
+
+##	Impact for water supply and demand, and soil moisture in vegetation productivity within landcover types
+
+The aim of this section is to analyze the drought indices of water demand and supply and soil moisture against vegetation to address: i) if short- or long-term time scales are most important in impacting vegetation through Chile; and ii) the strength of the correlation for the variable and the time scale.
+Then, we will summarize for each landcover class and macrozone.
+Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses: forest, cropland, grassland, savanna, and shrubland.
+
+To assess how water demand and supply and soil moisture are related to vegetation productivity (zcNDVI), we analyze the linear correlation between the indices SPI, SPEI, EDDI, and zcSM for 1, 3, 6, 12, 24, and 36-month time scales against zcNDVI.
+We followed a similar approach to that used by @Meroni2016 when using the SPI for meteorological drought against the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation) as a proxy for vegetation productivity.
+We made a pixel-to-pixel linear correlation analysis per index.
+First, we calculate the Pearson coefficient of correlation for the six time scales and let the time scale that reaches the maximum correlation be significant (p < 0.05).
+Then, we extracted the Pearson correlation value corresponding to the time scales that reached the maximum value.
+Thus, we derived two raster maps per index, the first with the time scales and the second with the correlation value.
+
+<!-- ##	Total water storage and drought indices -->
+
+<!-- Terrestrial Water Storage (TWS) is defined as the total amount of water stored on land that includes any natural or artificial water bodies, such as groundwater, soil moisture, rivers, lakes, snowpack, ice, and biomass water [@Humphrey2023; @Deng2023].
 TWS changes evidence the effects of multiple water fluxes on the hydrological cycle [@Deng2023].
@@
 The information provided by these satellites is used to construct monthly maps of the Earth's average gravity field, providing details of the movement of water masses or water mass anomaly estimates relative to the long-term average gravity field [@Humphrey2023; @Wahr2004].
+-->
 
-In this study, RL06.1_V3 GRACE mascon (mass concentration) monthly solutions were used, which are provided by the Jet Propulsion Laboratory (JPL-M, https://grace.jpl.nasa.gov).
+<!-- In this study, RL06.1_V3 GRACE mascon (mass concentration) monthly solutions were used, which are provided by the Jet Propulsion Laboratory (JPL-M, https://grace.jpl.nasa.gov).
 Each GRACE Tellus monthly grid at 0.5 degrees represents the deviation of surface mass for that month relative to a reference time average (2004-2009 baseline), which is subtracted from all other monthly grids to provide terrestrial water storage anomalies (TWSA) [@Ramjeawon2022; @Yin2022].
@@
 The water storage and height anomalies are given in Equivalent Water Height or Thickness units (EWH, cm) [@Sabzehee2023], and their temporal resolution monthly is from April 2002 to May 2023.
+-->
 

### 2024-01-15 article/paper_drought_Chile2023.tex
avanzar en paper metodologia

@@
   preprint,
-  3p]{elsarticle}
+  3p,
+  onecolumn]{elsarticle}
 
@@
 
-\usepackage{lineno}\linenumbers
+\usepackage{lineno}\linenumbers \usepackage{multirow}
 \makeatletter
@@
 \hypersetup{
-  pdftitle={Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from ERA5-Land and MODIS},
+  pdftitle={Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile},
   pdfauthor={Francisco Zambrano},
@@
 \begin{frontmatter}
-\title{Assessment of drought impact over land cover in continental Chile
-by the analysis of water supply and demand from ERA5-Land and MODIS}
+\title{Drought indices of water demand and supply, soil moisture,
+vegetation, and its impact on LULCC in continental Chile}
 \author[1]{Francisco Zambrano%
@@
 \end{frontmatter}
-    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable, sharp corners, frame hidden]}{\end{tcolorbox}}\fi
+    \captionsetup{justification=raggedright,singlelinecheck=false}
+
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 persistance of 80\% of times for 2001--2022, from the IGBP
-classification scheme \citep{Friedl2019}. ``Norte Grande'' and ``Norte
-Chico'' predominate in an arid desert climate with hot (Bwh) and cold
-(Bwk) temperatures. At the south of ``Norte Chico,'' the climate changes
-to an arid steppe with cold temperatures (Bsk). Mainly, the land is
-barren, with a minor surface of vegetation types such as shrubland and
+classification scheme \citep{Friedl2019} (see
+Section~\ref{sec-methods_lulc}). ``Norte Grande'' and ``Norte Chico''
+predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
+temperatures. At the south of ``Norte Chico,'' the climate changes to an
+arid steppe with cold temperatures (Bsk). Mainly, the land is barren,
+with a minor surface of vegetation types such as shrubland and
 grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
@@
 hourly frequency and global coverage. We selected the variables for
-total precipitation, 2 meters temperature, and volumetric soil water
-layers between 0 and 100cm of depth (layer 1 to layer 3).
+total precipitation, 2 meters temperature maximum and minimum, and
+volumetric soil water layers between 0 and 100cm of depth (layer 1 to
+layer 3).
 
@@
 
+\begin{table}[!ht]
+\caption{Description of the earth observation data used }
+\label{tab-desEOD}
+\small
+\centering
+\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
+\hline
+\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
+\hline
+\multirow{4}{*}{ERA5-Land} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
+         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
+         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
+         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
+ERA5-Land* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
+        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
+         & MCD12Q1.061 & Landcover IGBP scheme & & 2001-2022 & ~ & Landcover \\ 
+\hline
+\end{tabular}
+{\raggedright *Derived from ERA5-Land with Eq. \ref{eq-AED}. \par}
+\end{table}
+
 \hypertarget{in-situ-data}{%
@@
 
-\hypertarget{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}{%
-\subsection{Trend of drought indices for water demand/supply, soil
+\hypertarget{software-and-packages-used}{%
+\subsection{Software and packages
+used}\label{software-and-packages-used}}
+
+For the downloading, processing, and analysis of the spatio-temporal
+data, we used the open source software for statistical computing and
+graphics, \texttt{R} \citep{R2023}. For downloading ERA5-Land, we used
+the \texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing
+raster data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
+\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
+used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
+used \texttt{\{SPEI\}} \citep{Bergueria2023}.
+
+\hypertarget{drought-indices}{%
+\subsection{Drought Indices}\label{drought-indices}}
+
+We derived the drought indices of water supply and demand, soil moisture
+from the ERA5-Land dataset, and vegetation from the MODIS product, all
+at monthly frequency.
+
+For the indices EDDI and SPEI that use water demand, first we have to
+calculate the AED. For this, we used the method of Hargreaves
+\citep{Hargreaves1994}:
+
+\begin{equation}\protect\hypertarget{eq-AED}{}{AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}}\label{eq-AED}\end{equation}
+
+where \(Ra\) \((MJ\,m^2\, day^{-1})\) is extraterrestrial radiation;
+~\(T\), \(T_{max}\), and \(T_{min}\) are mean, maximum, and minimum
+temperature \((°C)\). We calculate the centroid coordinates per pixel
+and use the latitude to estimate \(Ra\).
+
+To evaluate water demand, we chose the \(EDDI\)
+\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). For
+supply, we used the index recommended by the World Meteorological
+Organization (WMO) for monitoring drought, the SPI \citep{Mckee1993}. We
+calculated the SPEI, which used a balance between \(P\) and \(AED\), in
+this case, an auxiliary variable \(D = P-AED\) is used. In this study,
+we propose the \(zcSM\) (standardized anomaly of acumulated soil
+moisture at 1 m), which uses SM. Finally, for the proxy of productivity,
+\(zcNDVI\), we used the NDVI. Before using the NDVI, it was smoothed
+following the procedure described in \citep{Zambrano2018} and
+\citep{Zambrano2016}.
+
+All the indices are multi-scalar and were calculated for time scales of
+1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was calculated
+for 6 months. The goal is to be able to evaluate short- and long-term
+droughts in water demand and supply and soil moisture. This is
+particularly important for central Chile because it has suffered from a
+prolonged decrease in precipitation for more than 12 years
+\citep{Garreaud2020, Boisier2018, Garreaud2017}.
+
+To calculate the drought indices, first we must calculate the
+accumulation of the variable. In this case, for generalization purposes,
+we will use \(V\), referring to \(P\), \(AED\), \(D\), \(NDVI\), and
+\(SM\) (Table \ref{tab-desEOD}). We cumulated each \(V\) over the time
+series of \(n\) values, and for the time scales \(s\):
+
+\begin{equation}\protect\hypertarget{eq-sumvar}{}{A_{si} = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  }\label{eq-sumvar}\end{equation}
+
+It corresponds to a moving window (convolution) that sums the variable
+for \(s\) starting for the last month \(n\) until the month, which could
+sum for \(s\) months (n-s+1). Once the variable is cumulated over time
+for the scale \(s\), we used a nonparametric approach following
+\citep{Hobbins2016} to derive the drought indices. Thus, the empirically
+derived probabilities are obtained through an inverse normal
+approximation \citep{Abramowitz1968}. Then, we used the empirical Tukey
+plotting position \citep{Wilks2011} over \(A_i\) to derive the
+\(P(A_i)\) probabilities across a period of interest:
+
+\begin{equation}\protect\hypertarget{eq-probPai}{}{P(A_i) = \frac{i-0.33}{n+0.33'}}\label{eq-probPai}\end{equation}
+
+The drought indices \(SPI\), \(SPEI\), \(EDDI\), \(zcSM\), and
+\(zcNDVI\) are obtained following the inverse normal approximation:
+
+\begin{equation}\protect\hypertarget{eq-DI}{}{DI(A_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}}\label{eq-DI}\end{equation}
+
+\(DI\) is referring to the drought index calculated for the variable
+\(V\). The values for the constats are: \(C_0 = 2.515517\),
+\(C_1 = 0.802853\), \(C_2 = 0.010328\), \(d_1 = 1.432788\),
+\(d_2 = 0.189269\), and \(d3 = 0.001308\). For \(P(A) \leq 0.5\),
+W=\(\sqrt{-2\cdot ln(P(A_i))}\) , and for \(P(A_i) > 0.5\), replace
+\(P(A_i)\) with \(1-P(A_i)\) and reverse the sign of \(DI(A_i)\).
+
+\hypertarget{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}{%
+\subsection{Trend of drought indices for water demand and supply, soil
 moisture, and vegetation
-productivity}\label{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}}
+productivity}\label{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}}
 
-\hypertarget{impact-fof-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-landcover-types}{%
-\subsection{Impact fof water supply and demand, and soil moisture in
-vegetation productivity within landcover
-types}\label{impact-fof-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-landcover-types}}
+To estimate if there are significant positive or negative trends for the
+drought indices, we used the non-parametric test of Mann-Kendall
+\citep{Kendall1975}. To determine the magnitude of the trend, we used
+Sen's slope \citep{Sen1968}. One of the advantages of applying this
+methodology is that the Sen's slope is not affected by outliers as
+regular regression does. We applied both to the five indices at the six
+time scales from 1981 to 2023 and indices SPI, EDDI, SPEI, and zcSM, and
+from 2000 to 2023 in the case of zcNDVI.
 
-\hypertarget{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}{%
+\citep{Vicente-Serrano2021} made a global analysis of the drought's
+severity trend using SPI, SPEI, and the Standardized Evapotranspiration
+Deficit Index (SEDI; \citep{Vicente-Serrano2018}) to evaluate AED. They
+indicate that the increase in hydrological drought has been due to
+anthropogenic effects rather than climate change. This is because the
+global increase in AED does not explain the change in the spatial
+pattern of hydrological drought. Also, they state that \emph{``the
+increase in hydrological droughts has been primarily observed in regions
+with high water demand and land cover change''}. We will contrast this
+hypothesis with what is occurring in Chile. To achieve this, we will use
+the landcover class type that remain more than 80\% of types for
+2001-2022, and use this as a mask, where there are low changes, to
+evaluate the trend on zcNDVI.
+
+\hypertarget{sec-methods_lulc}{%
 \subsection{LULC change for 2001-2022 and its relation with water supply
-and demand, and soil
-moisture}\label{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}}
+and demand, and soil moisture}\label{sec-methods_lulc}}
 
@@
 \caption{\label{fig-LCprop}Proportion of land cover class from the
-persistent land cover for 2001-2021 (\textgreater80\%) per macrozone}
+persistent land cover for 2001-2022 (\textgreater80\%) per macrozone}
 
@@
 
-\hypertarget{total-water-storage-and-drought-indices}{%
-\subsection{Total water storage and drought
-indices}\label{total-water-storage-and-drought-indices}}
+To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
+6.1 from MODIS. This product has a yearly frequency from 2011 to 2022.
+The IGBP defines 17 classes; from these, we regrouped into ten
+macroclasses for forest, schrublands, savannas, grasslands, wetlands,
+croplands, urban, snow and ice, barren, and water bodies. Then, we
+calculated the surface ocuppied per landcover class into each macrozone
+(``Norte Grande'' to ``Austral'') per year. To determine the trend of
+change we used the Sen' slope \citep{Sen1968} over the surface per year.
 
-Terrestrial Water Storage (TWS) is defined as the total amount of water
-stored on land that includes any natural or artificial water bodies,
-such as groundwater, soil moisture, rivers, lakes, snowpack, ice, and
-biomass water \citep{Humphrey2023, Deng2023}. TWS changes evidence the
-effects of multiple water fluxes on the hydrological cycle
-\citep{Deng2023}. These are reflected in the temporal variation of
-observations of the Earth's gravitational field
-\citep{Abolafia2021, Sabzehee2023} and, in recent decades, the twin
-Gravity Recovery and Climate Experiment (GRACE) satellites and their
-Follow-On mission (GRACE-FO) have provided valuable results on globally
-distributed TWS anomalies \citep{Tapley2019, Ferreira2023}. The GRACE
-mission was launched in March 2002 and was operational until October
-2017 \citep{Ramjeawon2022}. Then its GRACE Follow-On successor was
-launched in May 2018 \citep{Landerer2020, Yin2022}. The information
-provided by these satellites is used to construct monthly maps of the
-Earth's average gravity field, providing details of the movement of
-water masses or water mass anomaly estimates relative to the long-term
-average gravity field \citep{Humphrey2023, Wahr2004}.
+We calculated a raster mask considering the classes that remains more
+than 80\% of years (2001-2022), which allow us to identify the areas
+with no landcover change for the macroclasses. Figure~\ref{fig-LCprop}
+shows the summary of proportion of surface per landcover class and
+macrozone, derived from the mask over continental Chile.
 
-In this study, RL06.1\_V3 GRACE mascon (mass concentration) monthly
-solutions were used, which are provided by the Jet Propulsion Laboratory
-(JPL-M, https://grace.jpl.nasa.gov). Each GRACE Tellus monthly grid at
-0.5 degrees represents the deviation of surface mass for that month
-relative to a reference time average (2004-2009 baseline), which is
-subtracted from all other monthly grids to provide terrestrial water
-storage anomalies (TWSA) \citep{Ramjeawon2022, Yin2022}. This JPL-M
-version of the data employs a Coastal Resolution Improvement (CRI)
-filter that reduces signal leakage errors across coastlines
-\citep{NASA2019, Wiese2016}. Although the mascon solutions greatly
-reduced leakage errors, a gain factor, which is used to enhance the
-spatial resolution, was applied to the dataset as recommended for
-hydrological studies \citep{Ramjeawon2022, Yin2022}. The water storage
-and height anomalies are given in Equivalent Water Height or Thickness
-units (EWH, cm) \citep{Sabzehee2023}, and their temporal resolution
-monthly is from April 2002 to May 2023.
+We will explore if the trend in landcover classess is associated to
+trend of the drought indices. For this we will use the regularizations
+techniques of Lasso \citep{Tibshirani2010} and Ridge regressión
+\citep{Hoerl1970}. Also, we will test random forest for this purpose
+\citep{Ho1995}.
+
+\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-landcover-types}{%
+\subsection{Impact for water supply and demand, and soil moisture in
+vegetation productivity within landcover
+types}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-landcover-types}}
+
+The aim of this section is to analyze the drought indices of water
+demand and supply and soil moisture against vegetation to address: i) if
+short- or long-term time scales are most important in impacting
+vegetation through Chile; and ii) the strength of the correlation for
+the variable and the time scale. Then, we will summarize for each
+landcover class and macrozone. Thus, we will be able to advance in
+understanding how climate is affecting vegetation, considering the
+impact on the five macroclasses: forest, cropland, grassland, savanna,
+and shrubland.
+
+To assess how water demand and supply and soil moisture are related to
+vegetation productivity (zcNDVI), we analyze the linear correlation
+between the indices SPI, SPEI, EDDI, and zcSM for 1, 3, 6, 12, 24, and
+36-month time scales against zcNDVI. We followed a similar approach to
+that used by \citep{Meroni2016} when using the SPI for meteorological
+drought against the cumulative FAPAR (Fraction of Absorbed
+Photosynthetically Active Radiation) as a proxy for vegetation
+productivity. We made a pixel-to-pixel linear correlation analysis per
+index. First, we calculate the Pearson coefficient of correlation for
+the six time scales and let the time scale that reaches the maximum
+correlation be significant (p \textless{} 0.05). Then, we extracted the
+Pearson correlation value corresponding to the time scales that reached
+the maximum value. Thus, we derived two raster maps per index, the first
+with the time scales and the second with the correlation value.
 
@@
 
-\hypertarget{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity-1}{%
+\hypertarget{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}{%
 \subsection{Trend of drought indices for water demand/supply, soil
 moisture, and vegetation
-productivity}\label{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity-1}}
+productivity}\label{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}}
 
@@
 
-\hypertarget{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture-1}{%
+\hypertarget{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}{%
 \subsection{LULC change for 2001-2022 and its relation with water supply
 and demand, and soil
-moisture}\label{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture-1}}
+moisture}\label{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}}
 

### 2024-01-15 conference/IGARSS2024.qmd
avanzar en paper metodologia

@@
 ---
-title: "DDS4Chl: A dataset of drought indices of water demand and supply and its impact on vegetation for continental Chile"
+title: "Assessment of drought in continental Chile for 1981–2023 by climate variables of water supply and demand, soil moisture, and vegetation"
 format:
@@
 funding: 
-  statement: "The research was funded by the National Research and Development Agency of Chile (ANID) and the projects FSEQ210022 and 11190360."
+  statement: "The research was funded by the National Research and Development Agency of Chile (ANID) and the projects FSEQ210022, Fondecyt de Iniciacipon 11190360 and Fondecyt de Postdoctorado 3230678"
 pageheader:
@@
 We estimate the standardized anomaly of cumulative soil moisture at one meter (zcSM) and the standardized anomaly of cumulative NDVI (zcNDVI) to show the impact of water demand and supply.
-Finally, we present the historical linear trend of the drought indices in continental Chile.
-Also we calculated the temporal correlation between the indices of water supply, water demand, and soil moisture with the zcNDVI (proxy of vegetation productivity).
-The dought index for soil moisture (zcSM) derived from ERA5-Land showed to be the best predictor of vegetation productivity around continental Chile.
+We present the historical linear trend of the drought indices in continental Chile.
+Also we calculated the temporal correlation between the indices of water supply and demand, and soil moisture with the zcNDVI (proxy of vegetation productivity).
+The dought index for soil moisture acummulated at 12 months (zcSM-12) showed to be the best predictor of vegetation productivity around continental Chile.
 <br>
 
-**Index Terms—** drought, Chile, water supply, water demand
+**Index Terms—** drought, Chile, water supply, water demand, soil moisture
 
@@
 
+We evaluated the trend of the drought indices SPI, SPEI, EDDI, and zcSM; and their relation to vegetation productivity.
+Our goal is to assess the impact on vegetation of the climate variables of water demand and supply and soil moisture.
+
+```{r fig-studyArea}
+#| fig-pos: "!ht"
+#| fig-env: "figure*"
+#| fig-cap: '(a) Chile with the Koppen-Geiger climate classes and the five macrozones "Norte Grande", "Norte Chico",  "Centro", "Sur", and "Austral".
+(b) Topography reference map.
+(c) Land cover classes for 2022.
+(d) Persistent land cover classes (> 80%) for 2001-2022'
+#| out-width: "\\textwidth"
+
+knitr::include_graphics("../output/figs/map_study_con_landcover.png")
+```
+
 # **Methods**
@@
 They range from low desert in the north to alpine tundra and glaciers in the east and southeast.
-The south has an Oceanic climate and central Chile has a Mediterranean climate [@kottek2006].
-For the analysis, continental Chile was also divided into five zones according to a latitudinal gradient: "Norte Grande", "Norte Chico",  "Zona Central", "Zona Sur", and "Zona Austral".
+The south has an Oceanic climate and central Chile has a Mediterranean climate [@Beck2023].
+For the analysis, continental Chile was also divided into five zones according to a latitudinal gradient: "Norte Grande", "Norte Chico",  "Zona Central", "Zona Sur", and "Zona Asutral".
 (@fig-studyArea)
 
-```{r fig-studyArea}
+```{r fig-zcNDVI_var, out.extra='style="float:right; padding:10px"'}
 #| fig-pos: "!ht"
-#| fig-cap: "(a) Chile and the five zones 'norte grande', 'norte chico',  'zona central', 'zona sur', and 'zona austral'.
-(b) Topography reference map.
-(c) Land cover classes for 2021.
-(d) Persistent land cover classes (> 80%) for 2001-2021"
+#| fig-env: "figure*"
+#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023.
+Greener colors indicate a positive trend and an increase in vegetation development; reder colors correspond to a negative trend and a decrease in vegetation development.
+(b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile.
+Each horizontal panel corresponds to a macrozone from 'Norte Grande' to 'Zona Asutral'."
 #| out-width: "\\textwidth"
-
-knitr::include_graphics("../output/figs/map_study_con_landcover.png")
+knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas.png")
 ```
@@
 
-Regarding vegetation productivity, there is a negative trend on zcNDVI in the central part of the country ("norte chico" and "zona central") (@fig-zcNDVI_var).
+Regarding vegetation productivity, there is a negative trend on zcNDVI in the central part of the country ("Norte Chico" and "Zona Central") (@fig-zcNDVI_var).
 Indicating either a reduction of vegetation surface, biomass, or status.
-It reached its lowest values during the year 2020, mostly on the "zona central." There are other negative trends in small areas in "norte grande" and "zona asutral."
+It reached its lowest values during the year 2020, mostly on the "Zona Central." There are other negative trends in small areas in "Norte Grande" and "zona asutral."
 
-```{r fig-zcNDVI_var, out.extra='style="float:right; padding:10px"'}
-#| fig-pos: "!ht"
-#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023.
-Greener colors indicate a positive trend and an increase in vegetation development; reder colors correspond to a negative trend and a decrease in vegetation development.
-(b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile.
-Each horizontal panel corresponds to a macrozone from 'norte grande' to 'zona austral'."
-#| out-width: "\\textwidth"
-knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas.png")
-```
-
-Analyzing the water supply in the macrozones "norte chico," "central," and "sur," the SPI shows a decreasing trend that increases at longer time scales due to the prolonged reduction in precipitation.
+Analyzing the water supply in the macrozones "Norte Chico," "central," and "sur," the SPI shows a decreasing trend that increases at longer time scales due to the prolonged reduction in precipitation.
 It reaches a trend of -0.056 (z-score) per decade (@fig-trendSPI) for time scales of 36 months.
@@
 It shows a positive trend, caused by an increase in AED.
-The trend on EDDI reaches a maximum of 0.053 per decade in the macrozones "norte grande" and "norte chico" (@fig-trendEDDI).
+The trend on EDDI reaches a maximum of 0.053 per decade in the macrozones "Norte Grande" and "Norte Chico" (@fig-trendEDDI).
 The behavior is similar to SPI, having close trends with opposite signs.
 Nevertheless, the spatial patterns are different.
-The maximum trend for SPI-36 is in "norte chico" and "zona central." For EDDI-36, the maximum trend is from "northe grande" to "zona central."
+The maximum trend for SPI-36 is in "Norte Chico" and "Zona Central." For EDDI-36, the maximum trend is from "northe grande" to "Zona Central."
 
@@
 The correlation between zcNDVI-6 and drought indices shows that time scales higher than six months are predominant over continental Chile in explaining vegetation variability (@fig-corTimeScale).
-Pearson correlations between 0.6 and 0.9 are concentrated in the "norte chico" and "zona central" for SPI and SPEI.
-Soil moisture reached a correlation of 0.6–0.9 that extended north and south with respect to SPI.
+Pearson correlations between 0.6 and 0.9 are concentrated in the "Norte Chico" and "Zona Central" for SPI and SPEI.
+Soil moisture reached a correlation of 0.6–0.9 that extended north and south with respect to SPI (@fig-corPerson).
 
@@
 
-The vegetation productivity shows a high negative trend in the "norte chico" and "zona central" of continental Chile.
+The vegetation productivity shows a high negative trend in the "Norte Chico" and "Zona Central" of continental Chile.
 This trend is in the primary case related to long-term precipitation decreases, which in the secondary case could be explained by the increase in AED.

### 2024-01-15 R/trend_func.R

@@ -1,6 +1,6 @@
-trend_func <- function(x,start = c(1981,1),end = c(2023,11)){
+trend_func <- function(x,start = c(1981,1),end = c(2023,11),frecuency =12){
   x <- as.numeric(x)
-  t <- ts(x,start=start,end = end,frequency = 12)
+  t <- ts(x,start=start,end = end,frequency = frecuency)
   
   if (!all(is.na(t))){
     t <- imputeTS::na_mean(t)

### 2024-01-15 R/updt_modis.R

@@ -40,7 +40,7 @@ updt_modis <- function(type,product,dir_out){
       tryCatch(
       httr::GET(url = paste0(URL,dir,'/',name),
           config = httr::config(connecttimeout = 120),
-          authenticate("frzambra@gmail.com","29H1kc12"),
+          authenticate("frzambra@gmail.com","Traplozx398#"),
           write_disk(paste0(dir_out,product,'/',dir,'/',name),
                      overwrite=TRUE),
           progress()),

### 2024-01-15 scripts/21_landcover_per_macrozone.R

@@ -19,7 +19,8 @@ names(t) <- pal$Name
 lc <- rast('/mnt/md0/raster_procesada/MODIS_derived/IGBP.MCD12Q1.061/IGBP80_reclassified.tif') |> 
   project('EPSG:4326')
 
-macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg')
+macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
+  dplyr::mutate(macrozona = c('Norte Chico','Norte Grande','Austral','Centro','Sur'))
 
 LC_data <- terra::extract(lc,macro)
 
@@ -32,8 +33,8 @@ attr(colors,'names') <- paleta$Name
 LC_data |> 
   mutate(LC_type = factor(lyr.1,levels=paleta$class,labels=paleta$Name),
          zone = factor(ID,levels=1:5,labels=macro$macrozona),
-         zone = fct_relevel(zone, "norte grande", "norte chico",
-                                           'zona central','zona sur')
+         zone = fct_relevel(zone, "Norte Grande", "Norte Chico",
+                            'Centro','Sur','Austral')
          ) |> 
   select(LC_type,zone) |> 
   na.omit() |>  
@@ -59,7 +60,7 @@ summ_data2 |>
   theme_bw() +
   #coord_flip() +
   theme(
-    plot.margin = margin(5, 0, 45, 0, "pt"),
+    plot.margin = margin(3, 3, 3, 3, "pt"),
     legend.position = 'bottom',
     #axis.text = element_blank(),
     #axis.text.x = element_text(vjust=10),
@@ -67,6 +68,7 @@ summ_data2 |>
     #panel.grid = element_blank(),
     #panel.grid.minor = element_blank(),
     #plot.margin = unit(rep(-1,4), "cm")      # Adjust the margin to make in sort labels are not truncated!
-  ) -> plotZone
-ggsave('output/figs/LC_pers80_per_macrozone.png',width=15,height = 10,scale=.5)  
+  ) + 
+  guides(fill = guide_legend(nrow = 1)) -> plotZone
+ggsave('output/figs/LC_pers80_per_macrozone.png',width=7,height = 4,scale=1.5)  
 

### 2024-01-15 scripts/23_analysis_trend_drought_indices_and_landcover_macroXlandcover.R

@@ -2,155 +2,30 @@ library(tidyverse)
 library(tidymodels)
 
 data_trend_lc <- read_rds('data/processed_data/trends_landcover_2001-2022.rds') |> 
-  mutate(zone = fct_recode(zone,'norte grande' = 'big north' , 'norte chico' = 'little north','centro' = 'central', 'sur' = 'south' ,'austral' = 'austral')) |> 
   pivot_longer(-zone) |> 
-  rename(landcover = value,macro = zone,clase = name)
+  rename(landcover = value,macro = zone,class = name)
 
-data_trend_di <- read_rds('data/processed_data/trends_indices_macroXlandocver_2001_2023.rds')
+data_trend_di <- read_rds('data/processed_data/trends_indices_macroXlandocver_2001_2023.rds') |> 
+  mutate(macro = str_to_title(macro)) |> 
+  rename(class = clase)
 
 data_ana <- data_trend_di |> 
   full_join(data_trend_lc) |> 
   drop_na()
 
-data_ana |>
-  mutate(across(3:31,\(x) {
-    ifelse(x<0,0,1)
-    })) -> ou
-  
-#análisis de cluster
-library(dbscan)
-library(NbClust)
-
-res <- sapply(1:10,\(i){
-  kmeans(data_ana[,3:31],i)$tot.withinss  
-})
-
-res.nbclust <- NbClust(data_ana[,3:31], distance = "euclidean",
-                       min.nc = 2, max.nc = 9, 
-                       method = "ward.D2", index ="dunn")
-factoextra::fviz_nbclust(res.nbclust) + theme_minimal() + ggtitle("NbClust's optimal number of clusters")
-
-data_ana <- as.data.frame(data_ana)
-row.names(data_ana) <- paste0(data_ana$macro,'_',data_ana$clase)
-
-data_cluster <- data_ana[,1:2]
-
-hclust_eddi <-  data_ana[,3:8,drop = FALSE] |> scale() |>   dist() |> hclust(method = 'ward.D2')
-
-hclust_spi <-  data_ana[,9:14,drop = FALSE] |> scale() |>   dist() |> hclust(method = 'ward.D2')
-
-hclust_spei <-  data_ana[,15:20,drop = FALSE] |> scale() |>   dist() |> hclust(method = 'ward.D2')
-
[…]

### 2024-01-15 scripts/25_trend_drought_indices.R

@@ -12,7 +12,16 @@ macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |>
 ## Time series of zcNDVI for macrozones
 dir <- '/mnt/md0/raster_procesada/MODIS_derived/zcNDVI/zcNDVI-6'
 
+#persistencia de landcover
+lc_pers <- rast('/mnt/md0/raster_procesada/MODIS_derived/IGBP.MCD12Q1.061/IGBP80_reclassified.tif') 
+#nieve permanente
+lc_pers[lc_pers == 8] <- NA
+
+#aplicar mascara de landcover permanenete menos nieve permanenete  
 zcndvi <- rast(dir_ls(dir,regexp = 'tif$'))
+lc_pers <- resample(lc_pers,zcndvi,method = 'near')
+zcndvi <- mask(zcndvi,lc_pers)
+
 macro_mask <- rasterize(st_transform(macro,32719),zcndvi[[1]],field ='macrozona')
 
 df_zcndvi <- zonal(zcndvi,macro_mask,fun = 'mean',na.rm = TRUE)
@@ -153,15 +162,21 @@ trend_zcndvi <- subset(trend_zcndvi,ind_ord)
 names(trend_zcndvi) <- paste('zcNDVI',scales,sep ='-')
 trend_zcndvi <- mask(trend_zcndvi,st_transform(macro,crs(trend_zcndvi)))
 trend_zcndvi <- trim(trend_zcndvi) 
+
+#aplicar mascara de landcover persistente
+lc_pers <- crop(lc_pers,trend_zcndvi)
+trend_zcndvi <- mask(trend_zcndvi,lc_pers)
+
 map_zcNDVI <- tm_shape(trend_zcndvi) + 
-  tm_raster(palette = 'RdYlGn',midpoint = 0,title = 'Trend zcNDVI \n (per year)',style = 'kmeans') +
+  tm_raster(palette = 'RdYlGn',midpoint = 0,title = 'Trend zcNDVI \n (per year)',style = 'kmeans',colorNA = 'grey',textNA = 'Type Change' ) +
   tm_shape(macro) + 
-  tm_borders(col='white') +
+  tm_borders(col='black') +
   tm_shape(chl_b) + 
   tm_borders(col='black') +
-  tm_facets(nrow = 1) +
-  tm_layout(panel.labels = paste0('zcNDVI-',6),legend.outside = TRUE)
[…]

### 2024-01-15 scripts/29_mapas_correlaciones_indices_zcNDVI.R

@@ -1,0 +1,58 @@
+library(terra)
+library(sf)
+library(tidyverse)
+library(fs)
+library(tmap)
+library(basemaps)
+library(rnaturalearth)
+
+dir <- '/mnt/md0/raster_procesada/analysis/correlations/'
+files <- dir_ls(dir,type = 'file')
+
+chl <- ne_countries(country = 'chile',returnclass = 'sf') |> st_transform(32719)
+
+cors <- rast(files)
+cors_i <- subset(cors,seq(1,8,2))
+cors_r <- subset(cors,seq(2,8,2))
+
+set_defaults(map_token = "pk.eyJ1IjoiZnJ6YW1icmEiLCJhIjoiY2tqdmw5Z3QxMDZyZjJydG54M2RobWMyeSJ9.rl8_KzhiKaV0wgsLL2Y1WQ")
+bm <- basemap_raster(chl,map_service = 'osm',map_type="streets")
+
+names(cors_i) <- 1:4
+
+# map_i <- tm_shape(bm) + 
+#   tm_rgb() + 
+map_i <- tm_shape(bm) + 
+  tm_rgb() +
+  tm_shape(cors_i) + 
+  tm_raster(style = 'cat',
+            labels = as.character(c(1,3,6,12,24,36)),
+            palette = viridis::magma(6),
+            title =  'Time-scale \n(months)') +
+  # tm_shape(chl) +
+  # tm_borders() +
+  tm_facets(nrow = 1) + 
+  tm_layout(
+    panel.labels = c('EDDI','SPEI','SPI','zcSM'),
+            )
+names(cors_r) <- 1:4
+
+# map_r <- tm_shape(bm) + 
+#   tm_rgb() +
+map_r <- tm_shape(bm) + 
+  tm_rgb() +
+  tm_shape(cors_r) +
+  tm_raster(style='equal',palette = viridis::viridis(5, option = "B"),title = 'r',legend.hist = TRUE) +
+  tm_facets(free.scales	= FALSE) + 
+  tm_layout(
+    panel.labels = c('EDDI','SPI','SPEI','zcSM'),
+    legend.hist.bg.color = 'lightgrey',
+    legend.outside = FALSE,
+    legend.hist.width = .5,
+    legend.hist.height = .15,
+    legend.format = list(digits = 1)
+  )
+
+mapU <- tmap_arrange(map_i,map_r,nrow = 2)
[…]

### 2024-01-16 article/paper_drought_Chile2023.qmd
update article landscape figures of trend

@@
 We estimated atmospheric evaporative demand (AED) by combining the Hargreaves-Samani equation with the ERA5-Land temperature.
-We derived the drought indices SPI (Standardized Precipitation Index), SPEI (Standardized Precipitation Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), zcSM (standardized anomaly of cumulative soil moisture), and the zcNDVI (standardized anomaly of cumulative NDVI).
+We derived the drought indices SPI (Standardized Precipitation Index), SPEI (Standardized Precipitation Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), SSI (standardized anomaly of cumulative soil moisture), and the zcNDVI (standardized anomaly of cumulative NDVI).
 These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1, 3, and 6 months).
-We analyzed the temporal correlation of SPI, SPEI, EDDI, and zcSM with zcNDVI to have insights into the impact of water supply and demand on vegetation.
+We analyzed the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to have insights into the impact of water supply and demand on vegetation.
 Our results showed that LULCC had an increasing trend of 412 [km2yr−1] of forest expansion in the “Zona Sur”, together with a  decreasing trend of 24 [km2yr−1] of cropland contraction in the “Zona Central” meanwhile the “Zona Sur” showed an increase of 31 [km2yr−1], and a contraction of 80 [km2yr−1] of bare soil in the “Zona Austral”.
 The EDDI was the less correlated index for the five macro zones and the five types of land cover, showing that the temperature in Chile has little impact on vegetation.
-Higher r-squared values, between 0.5 and 0.8, were obtained at “Norte Chico” and “Zona Central” for the land cover types of savanna, shrubland, grassland, and croplands for the indices SPEI and zcSM at time scales of 12 and 24 months.
-The forest type reaches a r-squared of ~0.5 for zcSM of 12 months.
+Higher r-squared values, between 0.5 and 0.8, were obtained at “Norte Chico” and “Zona Central” for the land cover types of savanna, shrubland, grassland, and croplands for the indices SPEI and SSI at time scales of 12 and 24 months.
+The forest type reaches a r-squared of ~0.5 for SSI of 12 months.
 The results indicate that the “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits longer than a year, potentially explained by a low capacity of water storage in those zones that should be further investigated.
@@
     include-in-header: 
-      text: '\usepackage{lineno}\linenumbers
-             \usepackage{multirow}'
+      text: \usepackage{lineno}\linenumbers
+            \usepackage{multirow}
+            \usepackage{lscape}
+            \newcommand{\blandscape}{\begin{landscape}}
+            \newcommand{\elandscape}{\end{landscape}}
 ---
@@
 #| fig-pos: "!ht"
+#| fig-env: 'figure*'
 #| fig-cap: '(a) Chile with the Koppen-Geiger climate classes and the five macrozones "Norte Grande", "Norte Chico",  "Centro", "Sur", and "Austral".
@@
 
+## Software and packages used
+
+For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, `R` [@R2023].
+For downloading ERA5-Land, we used the `{ecmwfr}` package [@Hufkens2019].
+For processing raster data, we used `{terra}` [@Hijmans2023] and `{stars}` [@Pebesma2023].
+For managing vectorial data, we used `{sf}` [@Pebesma2018].
+For the calculation of AED, we used `{SPEI}` [@Bergueria2023].
+
 ##	Data
@@
 
-For water supply and demand variabes, we used ERA5-Land [@MunozSabater2021], a reanalys dataset wich provides the evolution of land variables since 1950.
-It has a spatial resolution of 0.1°, hourly frequency and global coverage.
-We selected the variables for total precipitation, 2 meters temperature maximum and minimum, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).
+For water supply and demand variables, we used ERA5-Land [@MunozSabater2021], a reanalys dataset that provides the evolution of land variables since 1950.
+It has a spatial resolution of 0.1°, hourly frequency, and global coverage.
+We selected the variables for total precipitation, 2 meter temperature maximum and minimum, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).
+The data was downloaded using the Copernicus Climate Data Store (CDS) Application Program Interface (API) implemented in `{ecmfwr}` [@Hufkens2019].
 
@@
 It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency.
+The MOD13A3.061 and MCD12Q1.061 were retrieved from the online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, https://lpdaa.usgs.gov/tools/data-pool/.
 
@@
 
-## Software and packages used
-
-For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, `R` [@R2023].
-For downloading ERA5-Land, we used the `{ecmwfr}` package [@Hufkens2019].
-For processing raster data, we used `{terra}` [@Hijmans2023] and `{stars}` [@Pebesma2023].
-For managing vectorial data, we used `{sf}` [@Pebesma2018].
-For the calculation of AED, we used `{SPEI}` [@Bergueria2023].
-
 ## Drought Indices
@@
 We calculated the SPEI, which used a balance between $P$ and $AED$, in this case, an auxiliary variable $D = P-AED$ is used.
-In this study, we propose the $zcSM$ (standardized anomaly of acumulated soil moisture at 1 m), which uses SM.
+In this study, we used the $SSI$ (standardized soil moisture index at 1 m) [@Hao2013;@AghaKouchak2014], which uses SM.
 Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI.
@@
 
-The drought indices $SPI$, $SPEI$, $EDDI$, $zcSM$, and $zcNDVI$ are obtained following the inverse normal approximation:
+The drought indices $SPI$, $SPEI$, $EDDI$, $SSI$, and $zcNDVI$ are obtained following the inverse normal approximation:
 
@@
 One of the advantages of applying this methodology is that the Sen's slope is not affected by outliers as regular regression does.
-We applied both to the five indices at the six time scales from 1981 to 2023 and indices SPI, EDDI, SPEI, and zcSM, and from 2000 to 2023 in the case of zcNDVI.
+We applied both to the five indices at the six time scales from 1981 to 2023 and indices SPI, EDDI, SPEI, and SSI, and from 2000 to 2023 in the case of zcNDVI.
 
@@
 
-To assess how water demand and supply and soil moisture are related to vegetation productivity (zcNDVI), we analyze the linear correlation between the indices SPI, SPEI, EDDI, and zcSM for 1, 3, 6, 12, 24, and 36-month time scales against zcNDVI.
+To assess how water demand and supply and soil moisture are related to vegetation productivity (zcNDVI), we analyze the linear correlation between the indices SPI, SPEI, EDDI, and SSI for 1, 3, 6, 12, 24, and 36-month time scales against zcNDVI.
 We followed a similar approach to that used by @Meroni2016 when using the SPI for meteorological drought against the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation) as a proxy for vegetation productivity.
@@
 
+\newpage
+
+\blandscape
 ```{r fig-trendSPI}
 #| echo: false
+#| layout-ncol: 2
 #| fig-pos: "!ht"
-#| fig-cap: "Linear trend of the SPI at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
+#| fig-env: "figure*"
+#| fig-cap: "Linear trend of the drought index (*) at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
+#| fig-subcap: 
+#| - "SPI (Standardized Precipitation Index)"
+#| - "SPEI (Standardized Precipitation Evapotranspiration Index)" 
+#| - "EDDI (Evaporative Demand Drought Index)"
+#| - "SSMI (Standardized Soil Moisture Index)"
 #| out-width: "\\textwidth"
@@
 knitr::include_graphics("../output/figs/trend_raster_SPI_1981-2023.png")
+knitr::include_graphics("../output/figs/trend_raster_SPEI_1981-2023.png")
+knitr::include_graphics("../output/figs/trend_raster_EDDI_1981-2023.png")
+knitr::include_graphics("../output/figs/trend_raster_zcSM_1981-2023.png")
+
 ```
 
-```{r fig-trendEDDI}
-#| echo: false
-#| fig-pos: "!ht"
-#| out-width: "\\textwidth"
-#| fig-cap: "Linear trend of the EDDI at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
-knitr::include_graphics("../output/figs/trend_raster_EDDI_1981-2023.png")
-```
+\elandscape
 

### 2024-01-16 article/paper_drought_Chile2023.tex
update article landscape figures of trend

@@
 
-\usepackage{lineno}\linenumbers \usepackage{multirow}
+\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
 \makeatletter
@@
 Precipitation Index), SPEI (Standardized Precipitation
-Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), zcSM
+Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), SSI
 (standardized anomaly of cumulative soil moisture), and the zcNDVI
@@
 3, and 6 months). We analyzed the temporal correlation of SPI, SPEI,
-EDDI, and zcSM with zcNDVI to have insights into the impact of water
+EDDI, and SSI with zcNDVI to have insights into the impact of water
 supply and demand on vegetation. Our results showed that LULCC had an
@@
 and ``Zona Central'' for the land cover types of savanna, shrubland,
-grassland, and croplands for the indices SPEI and zcSM at time scales of
+grassland, and croplands for the indices SPEI and SSI at time scales of
 12 and 24 months. The forest type reaches a r-squared of
-\textasciitilde0.5 for zcSM of 12 months. The results indicate that the
+\textasciitilde0.5 for SSI of 12 months. The results indicate that the
 ``Norte Chico'' and ``Zona Central'' are the most sensitive regions to
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, boxrule=0pt, enhanced, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\begin{figure}[!ht]
+\begin{figure*}[!ht]
 
@@
 
-\end{figure}
+\end{figure*}
 
@@
 \section{Materials and Methods}\label{materials-and-methods}}
+
+\hypertarget{software-and-packages-used}{%
+\subsection{Software and packages
+used}\label{software-and-packages-used}}
+
+For the downloading, processing, and analysis of the spatio-temporal
+data, we used the open source software for statistical computing and
+graphics, \texttt{R} \citep{R2023}. For downloading ERA5-Land, we used
+the \texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing
+raster data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
+\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
+used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
+used \texttt{\{SPEI\}} \citep{Bergueria2023}.
 
@@
 
-For water supply and demand variabes, we used ERA5-Land
-\citep{MunozSabater2021}, a reanalys dataset wich provides the evolution
+For water supply and demand variables, we used ERA5-Land
+\citep{MunozSabater2021}, a reanalys dataset that provides the evolution
 of land variables since 1950. It has a spatial resolution of 0.1°,
-hourly frequency and global coverage. We selected the variables for
-total precipitation, 2 meters temperature maximum and minimum, and
+hourly frequency, and global coverage. We selected the variables for
+total precipitation, 2 meter temperature maximum and minimum, and
 volumetric soil water layers between 0 and 100cm of depth (layer 1 to
-layer 3).
+layer 3). The data was downloaded using the Copernicus Climate Data
+Store (CDS) Application Program Interface (API) implemented in
+\texttt{\{ecmfwr\}} \citep{Hufkens2019}.
 
@@
 vegetation indices (NDVI and EVI) at 1km of spatial resolution and
-monthly frequency.
+monthly frequency. The MOD13A3.061 and MCD12Q1.061 were retrieved from
+the online Data Pool, courtesy of the NASA EOSDIS Land Processes
+Distributed Active Archive Center (LP DAAC), USGS Earth Resources
+Observation and Science (EROS) Center, Sioux Falls, South Dakota,
+https://lpdaa.usgs.gov/tools/data-pool/.
 
@@
 
-\hypertarget{software-and-packages-used}{%
-\subsection{Software and packages
-used}\label{software-and-packages-used}}
-
-For the downloading, processing, and analysis of the spatio-temporal
-data, we used the open source software for statistical computing and
-graphics, \texttt{R} \citep{R2023}. For downloading ERA5-Land, we used
-the \texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing
-raster data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
-\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
-used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
-used \texttt{\{SPEI\}} \citep{Bergueria2023}.
-
 \hypertarget{drought-indices}{%
@@
 this case, an auxiliary variable \(D = P-AED\) is used. In this study,
-we propose the \(zcSM\) (standardized anomaly of acumulated soil
-moisture at 1 m), which uses SM. Finally, for the proxy of productivity,
-\(zcNDVI\), we used the NDVI. Before using the NDVI, it was smoothed
-following the procedure described in \citep{Zambrano2018} and
-\citep{Zambrano2016}.
+we used the \(SSI\) (standardized soil moisture index at 1 m)
+\citep{Hao2013, AghaKouchak2014}, which uses SM. Finally, for the proxy
+of productivity, \(zcNDVI\), we used the NDVI. Before using the NDVI, it
+was smoothed following the procedure described in \citep{Zambrano2018}
+and \citep{Zambrano2016}.
 
@@
 
-The drought indices \(SPI\), \(SPEI\), \(EDDI\), \(zcSM\), and
-\(zcNDVI\) are obtained following the inverse normal approximation:
+The drought indices \(SPI\), \(SPEI\), \(EDDI\), \(SSI\), and \(zcNDVI\)
+are obtained following the inverse normal approximation:
 
@@
 regular regression does. We applied both to the five indices at the six
-time scales from 1981 to 2023 and indices SPI, EDDI, SPEI, and zcSM, and
+time scales from 1981 to 2023 and indices SPI, EDDI, SPEI, and SSI, and
 from 2000 to 2023 in the case of zcNDVI.
@@
 vegetation productivity (zcNDVI), we analyze the linear correlation
-between the indices SPI, SPEI, EDDI, and zcSM for 1, 3, 6, 12, 24, and
+between the indices SPI, SPEI, EDDI, and SSI for 1, 3, 6, 12, 24, and
 36-month time scales against zcNDVI. We followed a similar approach to
@@
 decade in the macrozones ``norte grande'' and ``norte chico''
-(Figure~\ref{fig-trendEDDI}). The behavior is similar to SPI, having
-close trends with opposite signs. Nevertheless, the spatial patterns are
+(\textbf{?@fig-trendEDDI}). The behavior is similar to SPI, having close
+trends with opposite signs. Nevertheless, the spatial patterns are
 different. The maximum trend for SPI-36 is in ``norte chico'' and ``zona
@@
 
-\begin{figure}[!ht]
+\newpage
 
-{\centering \includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}
+\blandscape
+
+\begin{figure}
+
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}
 
@@
 
-\caption{\label{fig-trendSPI}Linear trend of the SPI at time scales of
-1, 3, 6, 12, 24, and 36 months for 1981-2023}
+}
+
+\subcaption{\label{fig-trendSPI-1}SPI (Standardized Precipitation
+Index)}
+\end{minipage}%
+%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendSPI-2}SPEI (Standardized Precipitation
+Evapotranspiration Index)}
+\end{minipage}%
+\newline
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendSPI-3}EDDI (Evaporative Demand Drought
+Index)}
+\end{minipage}%
+%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_zcSM_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendSPI-4}SSMI (Standardized Soil Moisture
+Index)}
+\end{minipage}%
+
+\caption{\label{fig-trendSPI}Linear trend of the drought index (*) at
+time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023}
 
@@
 
-\begin{figure}[!ht]
-
-{\centering \includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}
-
-}
-
-\caption{\label{fig-trendEDDI}Linear trend of the EDDI at time scales of
-1, 3, 6, 12, 24, and 36 months for 1981-2023}
-
-\end{figure}
+\elandscape
 

### 2024-01-17 article/paper_drought_Chile2023.tex
actualizar repo oficina GEMA

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, boxrule=0pt, enhanced, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners, enhanced]}{\end{tcolorbox}}\fi
 

### 2024-01-17 scripts/30_validacion_landcover_IGBP.R

@@ -1,0 +1,51 @@
+library(terra)
+library(fs)
+library(rnaturalearth)
+
+
+dir <- '/mnt/md0/raster_procesada/Landcover'
+dir2 <- '/mnt/md0/raster_procesada/MODIS/IGBP.MCD12Q1.061'
+lc_chile <- rast(dir_ls(dir))
+
+lc_modis <- rast(dir_ls(dir2,regexp = '2013|2014'))
+lc_modis <- crop(lc_modis,ext(lc_chile))
+
+lc_chile <- resample(lc_chile,lc_modis,method = 'near')
+
+classes <- matrix(c(1:17,rep(2,5),rep(4,2),rep(4,2),3,5,1,8,1,10,9,6),ncol=2)
+
+lc_modis_r <- classify(lc_modis,rcl=classes,include.lowest = TRUE)
+
+lc2<- c(lc_modis_r,lc_chile)
+
+chl <- ne_countries(country = 'chile',scale = 'small',returnclass = 'sf') |> 
+  st_transform(32719)
+samples <- st_sample(chl,1000)
+
+df <- terra::extract(lc2,vect(samples))
+names(df) <- c('ID','IGBP2013','IGBP2014','LC_CHILE')
+
+f <- function(x){
+  if(substr(x,2,2) == '0') substr(x,1,2) else substr(x,1,1)
+}
+
+data <- df |> 
+  mutate(LC_CHILE_2 = substr(LC_CHILE,1,2)) |> 
+  rowwise(LC_CHILE_2) |> 
+  mutate(LC_CHILE_2 = f(LC_CHILE_2) |> as.numeric()) |> 
+  select(-ID,-LC_CHILE) |> 
+  ungroup() |> 
+  mutate(LC_CHILE_2 = ifelse(LC_CHILE_2 == 0, 6,LC_CHILE_2),
+         LC_CHILE_2 = ifelse(LC_CHILE_2 == 80, 8,LC_CHILE_2)) |> 
+  mutate(across(everything(),\(x) fct(sort(as.character(x)))))
+
+library(yardstick)
+
+conf_mat(data,LC_CHILE_2,IGBP2014)  
+accuracy(data,LC_CHILE_2,IGBP2014)  
+f_meas(data,LC_CHILE_2,IGBP2014)  
+
+roc_auc(data,truth = LC_CHILE_2,IGBP2013)  
+
+data |> distinct(LC_CHILE_2)
+data |> distinct(IGBP2013)

### 2024-01-17 scripts/31_landcover_per_basin.R

@@ -1,0 +1,71 @@
+# Analyzing landcover persistent (80%) per zone
+# August 2021
+# by frzambra
+
+library(sf)
+library(terra)
+
+paleta <- read.csv('data/processed_data/paleta_colores_landcover.csv') |> 
+  dplyr::filter(Name %in% c('Forest','Shrubland','Savanna','Grassland','Cropland','Barren land'))
+
+colors <-  rgb(paleta$R,paleta$G,paleta$B,maxColorValue = 255)
+attr(colors,'names') <- paleta$class
+
+# 
+pal <- read.csv('data/processed_data/paleta_colores_landcover.csv')
+t <- pal$class
+names(t) <- pal$Name
+
+chl <- ne_countries(country = 'chile',returnclass = 'sf') |> 
+  st_transform(32719)
+
+dir <- '/mnt/md0/raster_procesada/MODIS/'
+
+igbp <- rast(list.files(paste0(dir,'IGBP.MCD12Q1.061'),pattern='*.tif$',full.names=TRUE))
+igbp <- crop(igbp,chl)
+
+classes <- matrix(c(1:17,rep(1,5),rep(2,2),rep(3,2),4,5,6,7,6,8,9,10),ncol=2)
+
+igbpRecl <- classify(igbp,rcl=classes,include.lowest = TRUE)
+igbpRecl[igbpRecl == 0] <- NA
+
+cuencas <- read_sf('data/processed_data/spatial/Cuencas_BNA')
+
+data <- terra::extract(igbpRecl,cuencas)
+names(data) <- c('cuenca',2001:2022)
+
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+
+data |> 
+  pivot_longer(-cuenca,values_to = 'LC_type',names_to = 'year') |> 
+  filter(LC_type %in% c(1:4,6,9)) |>  
+  mutate(class2 = LC_type,
+         year = as.numeric(year),
+         cuenca = factor(cuenca),
+         LC_type = factor(LC_type,levels=paleta$class,labels=paleta$Name)) -> data2 
+
+data2 |> 
+  group_by(cuenca) |>  
+  filter(year == 2002) |>  
+  summarize(ntot = n()) ->dataSum
+
+data2 |> 
+  group_by(cuenca,year,LC_type) |>  
+  summarise(n=n()) |> 
+  left_join(dataSum,by='cuenca') |>  
[…]

### 2024-01-17 scripts/32_extraer_tendencias_indices_cuencas_landcover.R

@@ -1,0 +1,54 @@
+## Extraer tabla de tendencias para macrozonas}
+library(sf)
+library(terra)
+
+cuencas <- read_sf('data/processed_data/spatial/Cuencas_BNA') 
+
+dir <- '/mnt/md0/raster_procesada/analysis/trends'
+files <- dir_ls(dir,regexp = 'mann_kendall.*tif$')
+indices <- str_remove(str_remove(str_extract(files,'l_.*\\.'),'l_'),'\\.')
+indices <- indices[-19]
+trends <- rast(names(files)[-19])
+trends <- trends[[seq(2,48,2)]]
+names(trends) <- indices
+
+df_scales_1 <- terra::extract(trends[[7:24]],cuencas,fun =\(x){
+  table(x) |> sort(decreasing = TRUE) -> x
+  as.numeric(names(x[1]))
+})
+
+df_scales_2 <- terra::extract(trends[[1:6]],cuencas,fun =\(x){
+  table(x) |> sort() -> x
+  as.numeric(names(x[1]))
+})
+
+#zcNDVI
+trends <- rast(names(files)[19])
+trends <- trends[[2]]
+names(trends) <- 'zcNDVI-6'
+cuencas <- st_transform(cuencas,32719)
+df_scales_3 <- terra::extract(trends,cuencas,fun =\(x){
+  table(x) |> sort(decreasing = TRUE) -> x
+  as.numeric(names(x[1]))
+})
+
+df_scales <- cbind(st_drop_geometry(cuencas),df_scales_1,df_scales_2,df_scales_3) |> 
+  select(-starts_with('ID')) 
+
+sel_cuen <- df_scales |> 
+  group_by(COD_CUEN) |> 
+  summarize(n =n()) |> 
+  filter(n == 1 ) |> 
+  pull(COD_CUEN)
+
+df_scales <- df_scales |> 
+  filter(COD_CUEN %in% sel_cuen) |> 
+  select(-c(2:3)) |> 
+  mutate(COD_CUEN = fct(COD_CUEN)) 
+
+#guardar data.frame con los valores promedio de las tendencias (mann-kendall) para cada indice en las macrozonas de Chile
+df_scales |> 
+  pivot_longer(-c(COD_CUEN)) |> 
+  rename(index = name,trend = value) |> 
+  unnest(trend) |> 
+  write_rds('data/processed_data/df_trends_indices_cuencas.rds')

### 2024-01-17 scripts/33_analysis_trend_drought_indices_and_landcover_cuencas.R

@@ -1,0 +1,92 @@
+library(tidyverse)
+library(tidymodels)
+
+data_trend_lc <- read_rds('data/processed_data/trends_landcover_cuencas_2001-2022.rds') |> 
+  pivot_longer(-cuenca) |> 
+  rename(landcover = value,class = name)
+
+data_trend_di <- read_rds('data/processed_data/df_trends_indices_cuencas.rds') |> 
+  rename(class = clase)
+
+data_ana <- data_trend_di |> 
+  full_join(data_trend_lc) |> 
+  drop_na()
+
+data_model <- data_ana[,c(1:6,9:12,15:18,21:24,29,31)]
+data_model <- data_model |> 
+  #select(-starts_with('EDDI')) |> 
+  mutate(
+    #landcover = cut(landcover,5),
+    landcover = cut(landcover,quantile(landcover,c(0,1/3,2/3,1)),
+                    include.lowest = TRUE),
+         class = factor(class),
+         macro = factor(macro))
+
+df_split <- initial_split(data_model,prop =.9)
+df_train <- training(df_split)
+df_test <- testing(df_split)
+
+#random forest
+
+rf_spec <- rand_forest(trees = 1000, mode = "classification") |> 
+  set_args(importance = 'impurity')
+rf_wflow <- workflow(landcover ~ ., rf_spec)
+rf_fit <- fit(rf_wflow, df_train)
+
+ augment(rf_fit, new_data = df_train) %>%
+   conf_mat(truth = landcover, estimate = .pred_class)
+
+ # 
+augment(rf_fit, new_data = df_train) %>%
+   conf_mat(truth = landcover, estimate = .pred_class) %>%
+   autoplot(type = "heatmap")
+
+
+
+library(vip)
+rf_fit |> 
+  extract_fit_parsnip() |> 
+  vip(geom = "point",num_features = 20) + 
+  labs(title = "Random forest variable importance") 
+
+data_model |> 
+  pull(landcover)
+
+## Resampling
+## 
+
+data_folds <- vfold_cv(df_train, v = 5)
+
+rf_wf <- 
+  workflow() |> 
+  add_model(rf_spec) |> 
+  add_formula(landcover~.)
+
+get_rf_imp <- function(x) {
+  x |> 
+    extract_fit_parsnip() |> 
+    vip::vi()
+}
+
+ctrl_imp <- control_grid(extract = get_rf_imp)
+
[…]

### 2024-01-17 article/paper_drought_Chile2023.qmd
update mañana 17 ene

@@
 In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
-In "Centro," there is a major proportion (~50%) of chilean matorral (shrubland and savanna), followed by grassland (~16%), forest (8%), and croplands (5%).
+In "Centro," there is a major proportion (~50%) of chilean matorral (shrubland and savanna.[@Fuentes2020]), followed by grassland (~16%), forest (8%), and croplands (5%).
 The south part of "Centro" and the north part of "Austral" are dominated by an Oceanic climate (Cfb).
@@
 
-## Trend of drought indices for water demand and supply, soil moisture, and vegetation productivity
-
-To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975].
-To determine the magnitude of the trend, we used Sen's slope [@Sen1968].
-One of the advantages of applying this methodology is that the Sen's slope is not affected by outliers as regular regression does.
-We applied both to the five indices at the six time scales from 1981 to 2023 and indices SPI, EDDI, SPEI, and SSI, and from 2000 to 2023 in the case of zcNDVI.
-
-@Vicente-Serrano2021 made a global analysis of the drought's severity trend using SPI, SPEI, and the Standardized Evapotranspiration Deficit Index (SEDI; @Vicente-Serrano2018) to evaluate AED.
-They indicate that the increase in hydrological drought has been due to anthropogenic effects rather than climate change.
-This is because the global increase in AED does not explain the change in the spatial pattern of hydrological drought.
-Also, they state that _"the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change"_.
-We will contrast this hypothesis with what is occurring in Chile.
-To achieve this, we will use the landcover class type that remain more than 80% of types for 2001-2022, and use this as a mask, where there are low changes, to evaluate the trend on zcNDVI.
-
 ##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture {#sec-methods_lulc}
@@
 To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS.
-This product has a yearly frequency from 2011 to 2022.
+This product has a yearly frequency from 2001 to 2022.
 The IGBP defines 17 classes; from these, we regrouped into ten macroclasses for forest, schrublands, savannas, grasslands, wetlands, croplands, urban, snow and ice, barren, and water bodies.
-Then, we calculated the surface ocuppied per landcover class into each macrozone ("Norte Grande" to "Austral") per year.
-To determine the trend of change we used the Sen' slope [@Sen1968] over the surface per year.
+To validate the IGBP MODIS, we compare the macroclasses with the ones of a more detailed landcover map made by @Zhao2016 for Chile for the years 2013–2014 (LCChile).
+The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses derived from the IGBP MODIS.
+We chose the years 2013 and 2014 from MODIS IGBP because the LCChile was made with data acquired in 2013 and 2014.
+We resampled LCChile to the spatial resolution of the IGBP MODIS using the nearest neighbor method.
+Then, we took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCCHILE, IGP2013, and IGBP2014.
+Finally, we applied a confusion matrix and estimated the accuracy and F1 score as metrics of performance.
 
-We calculated a raster mask considering the classes that remains more than 80% of years (2001-2022), which allow us to identify the areas with no landcover change for the macroclasses.
-@fig-LCprop shows the summary of proportion of surface per landcover class and macrozone, derived from the mask over continental Chile.
+Later, with the goal of analyzing vegetation areas less affected by landcover change, we calculated a raster mask for IGBP MODIS considering the classes that remain for more than 80% of the years (2001–2022), which allows us to identify the areas with no landcover change for the macroclasses.
+@fig-LCprop shows the summary of the proportion of surface per landcover class and macrozone, derived from the mask over continental Chile.
+
+Further, we calculated the surface occupied per landcover class into the five macrozones ("Norte Grande" to "Austral") per year.
+After that, we calculated the trend's change in surface; we used the Sen' slope [@Sen1968] based on Mann-Kendall [@Kendall1975].
+This way, we obtain a matrix of trends of 5 x 5 (macrozones x landcover).
 
@@
 Also, we will test random forest for this purpose [@Ho1995].
+
+## Trend of drought indices for water demand and supply, soil moisture, and vegetation productivity
+
+To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975].
+To determine the magnitude of the trend, we used Sen's slope [@Sen1968].
+One of the advantages of applying this methodology is that the Sen's slope is not affected by outliers as regular regression does.
+We applied both to the six time scales from 1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI.
+In the case of zcNDVI (six months) was for 2000 to 2023.
+Thus, we have 31 trends.
+Also, we extracted the trend aggregated by macrozone and landcover class, obtaining a table of 31x5x5 (drought indices trends x macrozone x landcover class).
+We will use this data in @sec-methods_lulc to analyze if there is a strong relationship between the trends of drought indices and land cover surface within continental Chile.
+
+For zcNDVI, we want to remove the effect of the Andes's mountain and mask out trends that were at an elevation above 1500m because there is scarce vegetation.
+
+@Vicente-Serrano2021 made a global analysis of the drought's severity trend using SPI, SPEI, and the Standardized Evapotranspiration Deficit Index (SEDI; @Vicente-Serrano2018) to evaluate AED.
+They indicate that the increase in hydrological drought has been due to anthropogenic effects rather than climate change.
+This is because the global increase in AED did not explain the change in the spatial pattern of the hydrological drought.
+Also, they state that _"the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change"_.
+We will contrast this hypothesis with what is occurring in Chile.
+To achieve this, we will use the landcover class type that remains more than 80% of types for 2001–2022 to evaluate the trend on zcNDVI and use this as a mask where there are low changes.
 
@@
 
-##	Trend of drought indices for water demand/supply, soil moisture, and vegetation productivity
+##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture
 
-Regarding vegetation productivity, there is a negative trend on zcNDVI in the central part of the country ("norte chico" and "zona central") (@fig-zcNDVI_var).
-Indicating either a reduction of vegetation surface, biomass, or status.
-It reached its lowest values during the year 2020, mostly on the "zona central." There are other negative trends in small areas in "norte grande" and "zona asutral."
+The validation of IGBP2013, and IGBP2014 with LCChile reached an a ccuracy of xx11, xx2 and a F1 score of yy1,yy2 respectively (see SS1). (PONER ESTO EN LA METODOLOGIA)
+
+@fig-zcNDVI_var and \ref{tab-landcoverTrend} show that the macrozones with major LULCC were "Sur" and "Centro" with % and % of surface evidenced change, respectively.
+The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna -70 $km^2 year^{-1}$.
+In the "Centro" and "Sur" there are changes in the chilean matorral, with an important reduction in savanna (-136 to -318 $km^2 yr^{-1}$), and an increase in shrubland and grassland.
+Showing a change for more dense vegetation types.
+It appears to be a change from the surface occupied by cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $Km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.
+
+\begin{table}[!ht]
+\label{tab-landcoverTrend}
+\caption{The value of Sen's slope trend next to the time-series plot of surface per landcover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile.
+Values of zero indicate that there was not a significant trend.
+Red dots on the plots indicate the maximum and minimum values of surface.}
+\includegraphics[]{../output/figs/table_var_landcover_macro.png}
+\end{table}
+
+\begin{table}[!ht]
+\label{tab-corLandcover}
+\caption{}
+\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
+\end{table}
+
+##	Trend of drought indices for water demand and supply, soil moisture, and vegetation productivity
+
+Regarding vegetation productivity aggregated through the macrozones, there is a negative trend in four macrozones; only "Norte Grande" shows a positive trend of 0.07 per decade (z-score).
+zcNDVI reached a trend of -0.07 per decade in the "Centro" and "Norte Chico," followed by "Sur" with -0.05 and "Austral" with -0.03 (@fig-zcNDVI_var).
+This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023].
+It reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Centro." If we observe the spatial variation in @fig-zcNDVI_var a, despite the fact that "Norte Grande" has a positive trend, it can be noticed that the south part has a cluster with a negative trend.
+The negative trend is most noticeable in "Norte Chico" and the northern part of "Centro," but from there to the south, there is an extensive surface of change in landcover class where the trend was not calculated.
 
@@
 
-##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture
-
-\begin{table}[!ht]
-\label{tab-landcoverTrend}
-\caption{Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile.
-Red dots on the plots indicate the maximum and minimum surface}
-\includegraphics[]{../output/figs/table_var_landcover_macro.png}
-\end{table}
-
-\begin{table}[!ht]
-\label{tab-corLandcover}
-\caption{}
-\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
-\end{table}
-
 ##	Total water storage (TWS) and drought indices

### 2024-01-17 article/paper_drought_Chile2023.tex
update mañana 17 ene

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 In ``Centro,'' there is a major proportion (\textasciitilde50\%) of
-chilean matorral (shrubland and savanna), followed by grassland
-(\textasciitilde16\%), forest (8\%), and croplands (5\%). The south part
-of ``Centro'' and the north part of ``Austral'' are dominated by an
-Oceanic climate (Cfb). Those zones are high in forest and grassland. The
-southern part of the country has a tundra climate, and in Patagonia, it
-is a cold semi-arid area with an extended surface of grassland.
+chilean matorral (shrubland and savanna.\citep{Fuentes2020}), followed
+by grassland (\textasciitilde16\%), forest (8\%), and croplands (5\%).
+The south part of ``Centro'' and the north part of ``Austral'' are
+dominated by an Oceanic climate (Cfb). Those zones are high in forest
+and grassland. The southern part of the country has a tundra climate,
+and in Patagonia, it is a cold semi-arid area with an extended surface
+of grassland.
 
@@
 where \(Ra\) \((MJ\,m^2\, day^{-1})\) is extraterrestrial radiation;
-~\(T\), \(T_{max}\), and \(T_{min}\) are mean, maximum, and minimum
+\(T\), \(T_{max}\), and \(T_{min}\) are mean, maximum, and minimum
 temperature \((°C)\). We calculate the centroid coordinates per pixel
@@
 
-\hypertarget{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}{%
-\subsection{Trend of drought indices for water demand and supply, soil
-moisture, and vegetation
-productivity}\label{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}}
-
-To estimate if there are significant positive or negative trends for the
-drought indices, we used the non-parametric test of Mann-Kendall
-\citep{Kendall1975}. To determine the magnitude of the trend, we used
-Sen's slope \citep{Sen1968}. One of the advantages of applying this
-methodology is that the Sen's slope is not affected by outliers as
-regular regression does. We applied both to the five indices at the six
-time scales from 1981 to 2023 and indices SPI, EDDI, SPEI, and SSI, and
-from 2000 to 2023 in the case of zcNDVI.
-
-\citep{Vicente-Serrano2021} made a global analysis of the drought's
-severity trend using SPI, SPEI, and the Standardized Evapotranspiration
-Deficit Index (SEDI; \citep{Vicente-Serrano2018}) to evaluate AED. They
-indicate that the increase in hydrological drought has been due to
-anthropogenic effects rather than climate change. This is because the
-global increase in AED does not explain the change in the spatial
-pattern of hydrological drought. Also, they state that \emph{``the
-increase in hydrological droughts has been primarily observed in regions
-with high water demand and land cover change''}. We will contrast this
-hypothesis with what is occurring in Chile. To achieve this, we will use
-the landcover class type that remain more than 80\% of types for
-2001-2022, and use this as a mask, where there are low changes, to
-evaluate the trend on zcNDVI.
-
 \hypertarget{sec-methods_lulc}{%
@@
 To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
-6.1 from MODIS. This product has a yearly frequency from 2011 to 2022.
+6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
 The IGBP defines 17 classes; from these, we regrouped into ten
 macroclasses for forest, schrublands, savannas, grasslands, wetlands,
-croplands, urban, snow and ice, barren, and water bodies. Then, we
-calculated the surface ocuppied per landcover class into each macrozone
-(``Norte Grande'' to ``Austral'') per year. To determine the trend of
-change we used the Sen' slope \citep{Sen1968} over the surface per year.
+croplands, urban, snow and ice, barren, and water bodies. To validate
+the IGBP MODIS, we compare the macroclasses with the ones of a more
+detailed landcover map made by \citep{Zhao2016} for Chile for the years
+2013--2014 (LCChile). The later has a spatial resolution of 30 m and
+three levels of defined classes; from those, we used level 1, which fits
+with the macroclasses derived from the IGBP MODIS. We chose the years
+2013 and 2014 from MODIS IGBP because the LCChile was made with data
+acquired in 2013 and 2014. We resampled LCChile to the spatial
+resolution of the IGBP MODIS using the nearest neighbor method. Then, we
+took a random sample of 1000 points within continental Chile and
+extracted the classes that fell within each point for LCCHILE, IGP2013,
+and IGBP2014. Finally, we applied a confusion matrix and estimated the
+accuracy and F1 score as metrics of performance.
 
-We calculated a raster mask considering the classes that remains more
-than 80\% of years (2001-2022), which allow us to identify the areas
-with no landcover change for the macroclasses. Figure~\ref{fig-LCprop}
-shows the summary of proportion of surface per landcover class and
-macrozone, derived from the mask over continental Chile.
+Later, with the goal of analyzing vegetation areas less affected by
+landcover change, we calculated a raster mask for IGBP MODIS considering
+the classes that remain for more than 80\% of the years (2001--2022),
+which allows us to identify the areas with no landcover change for the
+macroclasses. Figure~\ref{fig-LCprop} shows the summary of the
+proportion of surface per landcover class and macrozone, derived from
+the mask over continental Chile.
+
+Further, we calculated the surface occupied per landcover class into the
+five macrozones (``Norte Grande'' to ``Austral'') per year. After that,
+we calculated the trend's change in surface; we used the Sen' slope
+\citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}. This way, we
+obtain a matrix of trends of 5 x 5 (macrozones x landcover).
 
@@
 \citep{Ho1995}.
+
+\hypertarget{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}{%
+\subsection{Trend of drought indices for water demand and supply, soil
+moisture, and vegetation
+productivity}\label{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}}
+
+To estimate if there are significant positive or negative trends for the
+drought indices, we used the non-parametric test of Mann-Kendall
+\citep{Kendall1975}. To determine the magnitude of the trend, we used
+Sen's slope \citep{Sen1968}. One of the advantages of applying this
+methodology is that the Sen's slope is not affected by outliers as
+regular regression does. We applied both to the six time scales from
+1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and
+SSI. In the case of zcNDVI (six months) was for 2000 to 2023. Thus, we
+have 31 trends. Also, we extracted the trend aggregated by macrozone and
+landcover class, obtaining a table of 31x5x5 (drought indices trends x
+macrozone x landcover class). We will use this data in
+Section~\ref{sec-methods_lulc} to analyze if there is a strong
+relationship between the trends of drought indices and land cover
+surface within continental Chile.
+
+For zcNDVI, we want to remove the effect of the Andes's mountain and
+mask out trends that were at an elevation above 1500m because there is
+scarce vegetation.
+
+\citep{Vicente-Serrano2021} made a global analysis of the drought's
+severity trend using SPI, SPEI, and the Standardized Evapotranspiration
+Deficit Index (SEDI; \citep{Vicente-Serrano2018}) to evaluate AED. They
+indicate that the increase in hydrological drought has been due to
+anthropogenic effects rather than climate change. This is because the
+global increase in AED did not explain the change in the spatial pattern
+of the hydrological drought. Also, they state that \emph{``the increase
+in hydrological droughts has been primarily observed in regions with
+high water demand and land cover change''}. We will contrast this
+hypothesis with what is occurring in Chile. To achieve this, we will use
+the landcover class type that remains more than 80\% of types for
+2001--2022 to evaluate the trend on zcNDVI and use this as a mask where
+there are low changes.
 
@@
 
-\hypertarget{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}{%
-\subsection{Trend of drought indices for water demand/supply, soil
+\hypertarget{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}{%
+\subsection{LULC change for 2001-2022 and its relation with water supply
+and demand, and soil
+moisture}\label{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}}
+
+Figure~\ref{fig-zcNDVI_var} and \ref{tab-landcoverTrend} show that the
+macrozones with major LULCC were ``Sur'' and ``Centro'' with \% and \%
+of surface evidenced change, respectively. The ``Norte Chico'' shows an
+increase in barrend land of 111 \(km^2 year^{-1}\) and a reduction in
+the class savanna -70 \(km^2 year^{-1}\). In the ``Centro'' and ``Sur''
+there are changes in the chilean matorral, with an important reduction
+in savanna (-136 to -318 \(km^2 yr^{-1}\)), and an increase in shrubland
+and grassland. Showing a change for more dense vegetation types. It
+appears to be a change from the surface occupied by cropland from the
+``Centro'' to the ``Sur.'' Also, there is a high increase in forest (397
+\(Km^2 yr^{-1}\)) in the ``Sur,'' replacing the savanna lost.
+
+\begin{table}[!ht]
+\label{tab-landcoverTrend}
+\caption{The value of Sen's slope trend next to the time-series plot of surface per landcover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile.
+Values of zero indicate that there was not a significant trend.
+Red dots on the plots indicate the maximum and minimum values of surface.}
+\includegraphics[]{../output/figs/table_var_landcover_macro.png}
+\end{table}
+
+\begin{table}[!ht]
+\label{tab-corLandcover}
+\caption{}
+\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
+\end{table}
+
+\hypertarget{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity-1}{%
+\subsection{Trend of drought indices for water demand and supply, soil
 moisture, and vegetation
-productivity}\label{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}}
+productivity}\label{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity-1}}
 
-Regarding vegetation productivity, there is a negative trend on zcNDVI
-in the central part of the country (``norte chico'' and ``zona
-central'') (Figure~\ref{fig-zcNDVI_var}). Indicating either a reduction
-of vegetation surface, biomass, or status. It reached its lowest values
-during the year 2020, mostly on the ``zona central.'' There are other
-negative trends in small areas in ``norte grande'' and ``zona asutral.''
+Regarding vegetation productivity aggregated through the macrozones,
+there is a negative trend in four macrozones; only ``Norte Grande''
+shows a positive trend of 0.07 per decade (z-score). zcNDVI reached a
+trend of -0.07 per decade in the ``Centro'' and ``Norte Chico,''
+followed by ``Sur'' with -0.05 and ``Austral'' with -0.03
+(Figure~\ref{fig-zcNDVI_var}). This could be associated either with a
+reduction in vegetation surface, a decrease in biomass, or browning
+\citep{Miranda2023}. It reached its lowest values since the year 2019,
+reaching an extreme condition in early 2020 and 2022 in the ``Centro.''
+If we observe the spatial variation in Figure~\ref{fig-zcNDVI_var} a,
+despite the fact that ``Norte Grande'' has a positive trend, it can be
+noticed that the south part has a cluster with a negative trend. The
+negative trend is most noticeable in ``Norte Chico'' and the northern
+part of ``Centro,'' but from there to the south, there is an extensive
+surface of change in landcover class where the trend was not calculated.
 
@@
 
-\hypertarget{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}{%
-\subsection{LULC change for 2001-2022 and its relation with water supply
-and demand, and soil
-moisture}\label{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}}
-
-\begin{table}[!ht]
-\label{tab-landcoverTrend}
-\caption{Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile.
-Red dots on the plots indicate the maximum and minimum surface}
-\includegraphics[]{../output/figs/table_var_landcover_macro.png}
-\end{table}
-
-\begin{table}[!ht]
-\label{tab-corLandcover}
-\caption{}
-\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
-\end{table}
-
 \hypertarget{total-water-storage-tws-and-drought-indices}{%

### 2024-01-17 article/paper_drought_Chile2023.qmd
avanzar trend

@@
 In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
-In "Centro," there is a major proportion (~50%) of chilean matorral (shrubland and savanna.[@Fuentes2020]), followed by grassland (~16%), forest (8%), and croplands (5%).
+In "Centro," there is a major proportion (~50%) of chilean matorral (shrubland and savanna, [@Fuentes2021]), followed by grassland (~16%), forest (8%), and croplands (5%).
 The south part of "Centro" and the north part of "Austral" are dominated by an Oceanic climate (Cfb).
@@
 Finally, we applied a confusion matrix and estimated the accuracy and F1 score as metrics of performance.
+The validation of IGBP2013 and IGBP2014 with LCChile reached similar metrics of performance, having an accuracy of 0.82 and a F1 score of 0.66 (see SS1).
 
@@
 
-The validation of IGBP2013, and IGBP2014 with LCChile reached an a ccuracy of xx11, xx2 and a F1 score of yy1,yy2 respectively (see SS1). (PONER ESTO EN LA METODOLOGIA)
-
-@fig-zcNDVI_var and \ref{tab-landcoverTrend} show that the macrozones with major LULCC were "Sur" and "Centro" with % and % of surface evidenced change, respectively.
+@fig-zcNDVI_var and \ref{tab-landcoverTrend} show that the macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively.
 The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna -70 $km^2 year^{-1}$.
@@
 
-\begin{table}[!ht]
-\label{tab-corLandcover}
-\caption{}
-\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
-\end{table}
-
 ##	Trend of drought indices for water demand and supply, soil moisture, and vegetation productivity
 
-Regarding vegetation productivity aggregated through the macrozones, there is a negative trend in four macrozones; only "Norte Grande" shows a positive trend of 0.07 per decade (z-score).
-zcNDVI reached a trend of -0.07 per decade in the "Centro" and "Norte Chico," followed by "Sur" with -0.05 and "Austral" with -0.03 (@fig-zcNDVI_var).
+Regarding vegetation productivity aggregated through the macrozones in the five landcover macroclasses.
+In "Norte Grande" there is a increase trend of 0.02 (z-index) per decade, related with types of grassland and shrubland.
+There is a negative trend in "Norte Chico" with -0.04  and "Centro" with -0.02 per decade.
+In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04.
+In "Centro," shrubland reaches -0.06, followed by grassland with -0.05, and croplands and savanna have ~-0.01 per decade.
 This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023].
-It reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Centro." If we observe the spatial variation in @fig-zcNDVI_var a, despite the fact that "Norte Grande" has a positive trend, it can be noticed that the south part has a cluster with a negative trend.
-The negative trend is most noticeable in "Norte Chico" and the northern part of "Centro," but from there to the south, there is an extensive surface of change in landcover class where the trend was not calculated.
+Vegetation reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Centro" (Mega Drought).
+The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var).
+The area most affected by a negative trend in vegetation seems to be associated with the Chilean matorral [@Fuentes2021], despite the croplands being as severely impacted by drought as native vegetation does in "Norte Chico.".
 
@@
 #| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023.
-Greener colors indicate a positive trend and an increase in vegetation development; reder colors correspond to a negative trend and a decrease in vegetation development.
+Greener colors indicate a positive trend; reder colors correspond to a negative trend and a decrease in vegetation productivity.
+Grey colors indicate either no vegetation or a change in landcover type for 2001–2022.
 (b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile.
-Each horizontal panel corresponds to a macrozone from 'norte grande' to 'zona austral'."
+Each horizontal panel corresponds to a macrozone from 'Norte Grande' to 'Austral'."
 #| out-width: "\\textwidth"
-knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas.png")
+knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png")
 ```
@@
 
+\begin{table}[!ht]
+\label{tab-corLandcover}
+\caption{}
+\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
+\end{table}
+
 ```{r fig-corTimeScale}
@@
 
-##	Total water storage (TWS) and drought indices
-
 ## Validation of ERA5-Land variables

### 2024-01-17 article/paper_drought_Chile2023.tex
avanzar trend

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, frame hidden, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 In ``Centro,'' there is a major proportion (\textasciitilde50\%) of
-chilean matorral (shrubland and savanna.\citep{Fuentes2020}), followed
+chilean matorral (shrubland and savanna, \citep{Fuentes2021}), followed
 by grassland (\textasciitilde16\%), forest (8\%), and croplands (5\%).
@@
 and IGBP2014. Finally, we applied a confusion matrix and estimated the
-accuracy and F1 score as metrics of performance.
+accuracy and F1 score as metrics of performance. The validation of
+IGBP2013 and IGBP2014 with LCChile reached similar metrics of
+performance, having an accuracy of 0.82 and a F1 score of 0.66 (see
+SS1).
 
@@
 Figure~\ref{fig-zcNDVI_var} and \ref{tab-landcoverTrend} show that the
-macrozones with major LULCC were ``Sur'' and ``Centro'' with \% and \%
-of surface evidenced change, respectively. The ``Norte Chico'' shows an
-increase in barrend land of 111 \(km^2 year^{-1}\) and a reduction in
-the class savanna -70 \(km^2 year^{-1}\). In the ``Centro'' and ``Sur''
-there are changes in the chilean matorral, with an important reduction
-in savanna (-136 to -318 \(km^2 yr^{-1}\)), and an increase in shrubland
-and grassland. Showing a change for more dense vegetation types. It
-appears to be a change from the surface occupied by cropland from the
-``Centro'' to the ``Sur.'' Also, there is a high increase in forest (397
-\(Km^2 yr^{-1}\)) in the ``Sur,'' replacing the savanna lost.
+macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'', and
+``Austral'' with 36\%, 31\%, and 34\%, respectively. The ``Norte Chico''
+shows an increase in barrend land of 111 \(km^2 year^{-1}\) and a
+reduction in the class savanna -70 \(km^2 year^{-1}\). In the ``Centro''
+and ``Sur'' there are changes in the chilean matorral, with an important
+reduction in savanna (-136 to -318 \(km^2 yr^{-1}\)), and an increase in
+shrubland and grassland. Showing a change for more dense vegetation
+types. It appears to be a change from the surface occupied by cropland
+from the ``Centro'' to the ``Sur.'' Also, there is a high increase in
+forest (397 \(Km^2 yr^{-1}\)) in the ``Sur,'' replacing the savanna
+lost.
 
@@
 
-\begin{table}[!ht]
-\label{tab-corLandcover}
-\caption{}
-\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
-\end{table}
-
 \hypertarget{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity-1}{%
@@
 
-Regarding vegetation productivity aggregated through the macrozones,
-there is a negative trend in four macrozones; only ``Norte Grande''
-shows a positive trend of 0.07 per decade (z-score). zcNDVI reached a
-trend of -0.07 per decade in the ``Centro'' and ``Norte Chico,''
-followed by ``Sur'' with -0.05 and ``Austral'' with -0.03
-(Figure~\ref{fig-zcNDVI_var}). This could be associated either with a
-reduction in vegetation surface, a decrease in biomass, or browning
-\citep{Miranda2023}. It reached its lowest values since the year 2019,
-reaching an extreme condition in early 2020 and 2022 in the ``Centro.''
-If we observe the spatial variation in Figure~\ref{fig-zcNDVI_var} a,
-despite the fact that ``Norte Grande'' has a positive trend, it can be
-noticed that the south part has a cluster with a negative trend. The
-negative trend is most noticeable in ``Norte Chico'' and the northern
-part of ``Centro,'' but from there to the south, there is an extensive
-surface of change in landcover class where the trend was not calculated.
+Regarding vegetation productivity aggregated through the macrozones in
+the five landcover macroclasses. In ``Norte Grande'' there is a increase
+trend of 0.02 (z-index) per decade, related with types of grassland and
+shrubland. There is a negative trend in ``Norte Chico'' with -0.04 and
+``Centro'' with -0.02 per decade. In the ``Norte Chico,'' savanna
+(-0.05) has the lowest trend, and the rest of the types are around
+-0.04. In ``Centro,'' shrubland reaches -0.06, followed by grassland
+with -0.05, and croplands and savanna have \textasciitilde-0.01 per
+decade. This could be associated either with a reduction in vegetation
+surface, a decrease in biomass, or browning \citep{Miranda2023}.
+Vegetation reached its lowest values since the year 2019, reaching an
+extreme condition in early 2020 and 2022 in the ``Centro'' (Mega
+Drought). The ``Sur'' and ``Austral'' show a positive trend of around
+0.016 per decade (Figure~\ref{fig-zcNDVI_var}). The area most affected
+by a negative trend in vegetation seems to be associated with the
+Chilean matorral \citep{Fuentes2021}, despite the croplands being as
+severely impacted by drought as native vegetation does in ``Norte
+Chico.''.
 
@@
 
-{\centering \includegraphics{../output/figs/temporal_variation_zcNDVI6_macrozonas.png}
+{\centering \includegraphics{../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png}
 
@@
 \caption{\label{fig-zcNDVI_var}(a) Map of the linear trend of the index
-zcNDVI-6 for 2001--2023. Greener colors indicate a positive trend and an
-increase in vegetation development; reder colors correspond to a
-negative trend and a decrease in vegetation development. (b) Temporal
-variation of zcNDVI-6 aggregated at macrozone level within continental
-Chile. Each horizontal panel corresponds to a macrozone from `norte
-grande' to `zona austral'.}
+zcNDVI-6 for 2001--2023. Greener colors indicate a positive trend; reder
+colors correspond to a negative trend and a decrease in vegetation
+productivity. Grey colors indicate either no vegetation or a change in
+landcover type for 2001--2022. (b) Temporal variation of zcNDVI-6
+aggregated at macrozone level within continental Chile. Each horizontal
+panel corresponds to a macrozone from `Norte Grande' to `Austral'.}
 
@@
 
+\begin{table}[!ht]
+\label{tab-corLandcover}
+\caption{}
+\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
+\end{table}
+
 \begin{figure}[!ht]
@@
 
-\hypertarget{total-water-storage-tws-and-drought-indices}{%
-\subsection{Total water storage (TWS) and drought
-indices}\label{total-water-storage-tws-and-drought-indices}}
-
 \hypertarget{validation-of-era5-land-variables-1}{%

### 2024-01-17 scripts/19_crear_gt_table_var_landcover_macro.R

@@ -25,26 +25,26 @@ dataLCV_trend |>
   pivot_longer(-zone,names_to = 'cropland_class') |> 
   write_rds('data/processed_data/trend_landcover_change_classXzone_2001_2022.rds')
 
-dataLCV_trend  |> 
-  kbl(booktabs = TRUE,digits = 1,align ='r','latex',position='!ht',
-      caption='Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface.') %>%
-  kable_styling(latex_options="scale_down") %>% 
-  kable_paper(full_width = FALSE) %>%
-  column_spec(2, image = spec_plot(dataSpark[[1]],same_lim = FALSE)
-  ) %>%
-  column_spec(3, image = spec_plot(dataSpark[[2]],same_lim = FALSE)
-  ) %>%
-  column_spec(4, image = spec_plot(dataSpark[[3]],same_lim = FALSE)
-  ) %>%
-  column_spec(5, image = spec_plot(dataSpark[[4]],same_lim = FALSE)
-  ) %>%
-  column_spec(6, image = spec_plot(dataSpark[[5]],same_lim = FALSE)
-  ) %>%
-  column_spec(7, image = spec_plot(dataSpark[[6]],same_lim = FALSE)
-  ) %>% 
-  add_header_above(c(" ",'Trend of change [$km^2 year^{-1}$]' = 6),escape =FALSE) |> 
-  save_kable(file = "table_2.png",
-             zoom = 1.5)
+# dataLCV_trend  |> 
+#   kbl(booktabs = TRUE,digits = 1,align ='r','latex',position='!ht',
+#       caption='Value of linear change trend next to time-series plot of surface, per landcover class (IGBP MCD12Q1.006) for 2001-2019 through Central Chile. Red dots on the plots indicate the maximum and minimum surface.') %>%
+#   kable_styling(latex_options="scale_down") %>% 
+#   kable_paper(full_width = FALSE) %>%
+#   column_spec(2, image = spec_plot(dataSpark[[1]],same_lim = FALSE)
+#   ) %>%
[…]

### 2024-01-17 scripts/25_trend_drought_indices.R

@@ -11,14 +11,15 @@ macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |>
 
 ## Time series of zcNDVI for macrozones
 dir <- '/mnt/md0/raster_procesada/MODIS_derived/zcNDVI/zcNDVI-6'
+zcndvi <- rast(dir_ls(dir,regexp = 'tif$'))
 
 #persistencia de landcover
 lc_pers <- rast('/mnt/md0/raster_procesada/MODIS_derived/IGBP.MCD12Q1.061/IGBP80_reclassified.tif') 
-#nieve permanente
-lc_pers[lc_pers == 8] <- NA
+#eliminar los que no corresponde a los 5 principales landcover
+lc_pers[lc_pers %in% c(5,7:10)] <- NA
+lc_pers <- resample(lc_pers,zcndvi)
 
-#aplicar mascara de landcover permanenete menos nieve permanenete  
-zcndvi <- rast(dir_ls(dir,regexp = 'tif$'))
+#aplicar mascara de landcover permanenete 
 lc_pers <- resample(lc_pers,zcndvi,method = 'near')
 zcndvi <- mask(zcndvi,lc_pers)
 
@@ -46,7 +47,7 @@ df_zcndvi |>
   theme_minimal() +
   theme(axis.title.x = element_blank())
 
-ggsave('output/figs/temporal_variation_zcNDVI6_macrozonas.png',scale = 1.5,bg = 'white',width=6,height=3)
+ggsave('output/figs/temporal_variation_zcNDVI6_macrozonas.png',scale = 2,bg = 'white',width=6,height=3)
 
 #Mapa de tendencia Mann-Kendall
 library(fs)
@@ -148,6 +149,7 @@ map_zcSM <- tm_shape(trend_zcsm*10) +
 tmap_save(map_zcSM,'output/figs/trend_raster_zcSM_1981-2023.png',asp=.2)
 
 ## zcNDVI
+dir <- '/mnt/md0/raster_procesada/analysis/trends/'
 files <- dir_ls(dir,regexp = 'mann.*zcNDVI.*tif$')
 scales <- str_extract(basename(files),'-.*\\.') |> str_remove('-') |> str_remove('\\.')
 
@@ -165,10 +167,11 @@ trend_zcndvi <- trim(trend_zcndvi)
 
 #aplicar mascara de landcover persistente
[…]

### 2024-01-17 scripts/30_validacion_landcover_IGBP.R

@@ -30,6 +30,7 @@ f <- function(x){
 }
 
 data <- df |> 
+  filter(IGBP2013 != 5 & IGBP2014 != 5) |> 
   mutate(LC_CHILE_2 = substr(LC_CHILE,1,2)) |> 
   rowwise(LC_CHILE_2) |> 
   mutate(LC_CHILE_2 = f(LC_CHILE_2) |> as.numeric()) |> 
@@ -40,10 +41,13 @@ data <- df |>
   mutate(across(everything(),\(x) fct(sort(as.character(x)))))
 
 library(yardstick)
-
 conf_mat(data,LC_CHILE_2,IGBP2014)  
 accuracy(data,LC_CHILE_2,IGBP2014)  
 f_meas(data,LC_CHILE_2,IGBP2014)  
+
+accuracy(data,LC_CHILE_2,IGBP2013)  
+f_meas(data,LC_CHILE_2,IGBP2013)  
+
 
 roc_auc(data,truth = LC_CHILE_2,IGBP2013)  
 

### 2024-01-17 scripts/34_trend_zcNDVI_macrozona_y_landcover.R

@@ -1,0 +1,50 @@
+#Trend zcNDVi-6 por amcrozona y landcover 
+
+#datos para landcover
+paleta <- read.csv('data/processed_data/paleta_colores_landcover.csv')
+colors <-  rgb(paleta$R,paleta$G,paleta$B,maxColorValue = 255)
+attr(colors,'names') <- paleta$Name
+
+dir <- '/mnt/md0/raster_procesada/analysis/trends'
+files <- dir_ls(dir,regexp = 'mann_kendall.*zcNDVI-6.*tif$')
+
+trends <- rast(files)[[2]]
+names(trends) <- 'zcNDVI-6'
+
+#mascara para elevaciones mayores a 1500m
+elev <- geodata::elevation_30s('chile',path = tempdir()) |> 
+  project('EPSG:32719')
+elev2 <- resample(elev,trends)
+mask <- elev2
+mask[mask<1500] <- 1
+mask[mask>=1500] <- NA
+
+macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
+  mutate(macrozona = c('Norte Chico','Norte Grande','Austral','Centro','Sur')) |> 
+  st_transform(32719)
+
+trends <- resample(trends,mask)
+trends <- mask(trends,mask)
+
+# ahora por tipo de landcover
+
+lc_pers <- rast('/mnt/md0/raster_procesada/MODIS_derived/IGBP.MCD12Q1.061/IGBP80_reclassified.tif') 
+#nieve permanente
+lc_pers <- resample(lc_pers,trends,method = 'near')
+lc_pers[lc_pers %in% c(5,7:10)] <- NA
+
+data <-  map_df(1:5,\(i){
+  macro_n <- macro[i,]
+  trends_n <- mask(trends,macro_n)
+  lc_pers_n <- mask(lc_pers,macro_n)
+  trends_df <- zonal(trends_n,lc_pers_n,'mean',na.rm = TRUE)
+  trends_df |> 
+    mutate(macro = factor(macro_n$macrozona,levels = c('Norte Grande','Norte Chico','Centro','Sur','Austral')),
+           clase = factor(lyr.1,levels = paleta$class, labels = paleta$Name)) |> 
+    select(-lyr.1)
+})
+
+data_final <- data |> 
+  pivot_wider(names_from = clase,values_from = `zcNDVI-6`) |> 
+  arrange(macro) |> 
+  select(c(1,6,5,4,3,2)) 

### 2024-01-18 article/paper_drought_Chile2023.qmd
update mañana 17 ene

@@
 For the analysis, continental Chile was divided into five zones according to a latitudinal gradient: “Norte Grande”,  “Norte Chico”,  “Zona Central”, “Zona Sur”, and “Zona Austral”.
-We used monthly climatic re-analysis variables for precipitation, temperature and soil moisture for 1981-2023 from ERA5-Land; and MODIS (Moderate-Resolution Imaging Spectroradiometer) product MCD12Q1 for land cover for 2001-2021, and the NDVI vegetation index from product MOD13A2 collection 6.1 for 2000-2023, both from collection 6.1.
-We estimated atmospheric evaporative demand (AED) by combining the Hargreaves-Samani equation with the ERA5-Land temperature.
+We used monthly climatic re-analysis variables for precipitation, temperature and soil moisture for 1981-2023 from ERA5-Land (ERA5L); and MODIS (Moderate-Resolution Imaging Spectroradiometer) product MCD12Q1 for land cover for 2001-2021, and the NDVI vegetation index from product MOD13A2 collection 6.1 for 2000-2023, both from collection 6.1.
+We estimated atmospheric evaporative demand (AED) by combining the Hargreaves-Samani equation with the ERA5L temperature.
 We derived the drought indices SPI (Standardized Precipitation Index), SPEI (Standardized Precipitation Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), SSI (standardized anomaly of cumulative soil moisture), and the zcNDVI (standardized anomaly of cumulative NDVI).
@@
 In this research, we present the raster dataset DDS4Chl, which provides climate variables and drought indices of water demand and supply and vegetation productivity for continental Chile since 1981 at a monthly frequency.
-Those were gathered from the earth observation products ERA5-Land and MODIS.
+Those were gathered from the earth observation products ERA5L and MODIS.
 Then, we used DDS4Chl to analyze the impact of drought on different types of land cover classes in continental Chile.
@@
 
-# Study area
+# Study area (AGREGAR SUPERFICIE DE LANDCOVER EN CADA MACROZONA)
 
@@
 For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, `R` [@R2023].
-For downloading ERA5-Land, we used the `{ecmwfr}` package [@Hufkens2019].
+For downloading ERA5L, we used the `{ecmwfr}` package [@Hufkens2019].
 For processing raster data, we used `{terra}` [@Hijmans2023] and `{stars}` [@Pebesma2023].
@@
 
-For water supply and demand variables, we used ERA5-Land [@MunozSabater2021], a reanalys dataset that provides the evolution of land variables since 1950.
+For water supply and demand variables, we used ERA5L [@MunozSabater2021], a reanalys dataset that provides the evolution of land variables since 1950.
 It has a spatial resolution of 0.1°, hourly frequency, and global coverage.
@@
 \hline
-\multirow{4}{*}{ERA5-Land} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
+\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
          &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
@@
          &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
-ERA5-Land* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
+ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
         \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
@@
 \end{tabular}
-{\raggedright *Derived from ERA5-Land with Eq. \ref{eq-AED}. \par}
+{\raggedright *Derived from ERA5L with Eq. \ref{eq-AED}. \par}
 \end{table}
@@
 
-We derived the drought indices of water supply and demand, soil moisture from the ERA5-Land dataset, and vegetation from the MODIS product, all at monthly frequency.
+We derived the drought indices of water supply and demand, soil moisture from the ERA5L dataset, and vegetation from the MODIS product, all at monthly frequency.
 
@@
 
-## Validation of ERA5-Land variables
+## Validation of ERA5L variables
 
@@
 Showing a change for more dense vegetation types.
-It appears to be a change from the surface occupied by cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $Km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.
+It appears to be a change from the surface occupied by cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.
 
@@
 This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023].
-Vegetation reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Centro" (Mega Drought).
+Vegetation reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Norte Chico" and Centro" (Mega Drought).
 The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var).
@@
 
-The correlation between zcNDVI-6 and drought indices shows that time scales higher than six months are predominant over continental Chile in explaining vegetation variability (@fig-corTimeScale).
-Pearson correlations between 0.6 and 0.9 are concentrated in the "norte chico" and "zona central" for SPI and SPEI.
-Soil moisture reached a correlation of 0.6–0.9 that extended north and south with respect to SPI.
+##	Impact for water supply and demand, and soil moisture in vegetation productivity (CHEQUEAR CIFRAS EN EL TEXTO)
 
-##	Impact for water supply and demand, and soil moisture in vegetation productivity
+According with what is showed @fig-corTimeScale, @fig-corPerson, and Table 3.
+Forest seems to be the most resistant type to drought, showing that only "Centro" is slightly (rsq = 0.3) impacted by a 12-month soil moisture deficit (SSI-12).
+Following by grasslands, which show to be impacted by a SSI-12 with a rsq = 0.4, and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.3) in the "Norte Chico" and to a lesser degree in the "Norte Grande." But, in the southern macrozones, this type was not affected by water demand, water supply, or soil moisture.
+The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
+For savanna in "Norte Chico," the SSI-12 and SPI-24 reached an rsq of 0.8 and 0.6, respectively.
+This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (rsq = 0.4 in "Centro" and 0.2 in "Sur").
+In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 40% and 75% in "Centro." The most affected type by climatic variation was shrubland, where in "Norte Chico" and "Centro" soil moisture and precipitation explained more than 50%, with SSI-12 being the most relevant variable, followed by SPI-36 in "Norte Chico" and SPI-24 in "Sur."
 
+```{=latex}
 \begin{table}[!ht]
 \label{tab-corLandcover}
-\caption{}
+\caption{Summarry per landcover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
+The numbers in each cell indicate the time scale that reached the maximum correlation for the landcover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
 \includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
 \end{table}
+```
 
@@
 
+1.-  Respecto a lo que indica @Vicente-Serrano2018, de que el aumento en la tendencia en severidad de la sequía (hidrológica) tiene que ver más con un aumento de la demanda de agua (LULCC) que a una tendencia en las condiciones climáticas (SPI-12).
+
+2.- Sobre los tipos de landcover más afectados por los indicadores de sequía. Asociación con el matorral chileno [@Fuentes2021]. Diferencia entre el Norte Chico y Centro y lo que pasa hacía el sur.
+
+3. Como podrian servir estos resultados para desarrollar un predictor de sequía basado.
+
+4.- Qué se podria hacer mejor en futuras investigaciones del tema.
+
+5.- Sequía, aridez, water scarcity [VanLoon2013] Si puede utilizar el mapa y proyección climática de Koppen_Geiger [@Beck2023] para dirigir esta discusión
+
 # Conclusion
 
+\newpage
+
 # References {-}

### 2024-01-18 article/paper_drought_Chile2023.tex
update mañana 17 ene

@@
 re-analysis variables for precipitation, temperature and soil moisture
-for 1981-2023 from ERA5-Land; and MODIS (Moderate-Resolution Imaging
-Spectroradiometer) product MCD12Q1 for land cover for 2001-2021, and the
-NDVI vegetation index from product MOD13A2 collection 6.1 for 2000-2023,
-both from collection 6.1. We estimated atmospheric evaporative demand
-(AED) by combining the Hargreaves-Samani equation with the ERA5-Land
-temperature. We derived the drought indices SPI (Standardized
-Precipitation Index), SPEI (Standardized Precipitation
+for 1981-2023 from ERA5-Land (ERA5L); and MODIS (Moderate-Resolution
+Imaging Spectroradiometer) product MCD12Q1 for land cover for 2001-2021,
+and the NDVI vegetation index from product MOD13A2 collection 6.1 for
+2000-2023, both from collection 6.1. We estimated atmospheric
+evaporative demand (AED) by combining the Hargreaves-Samani equation
+with the ERA5L temperature. We derived the drought indices SPI
+(Standardized Precipitation Index), SPEI (Standardized Precipitation
 Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), SSI
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, frame hidden, breakable, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 vegetation productivity for continental Chile since 1981 at a monthly
-frequency. Those were gathered from the earth observation products
-ERA5-Land and MODIS. Then, we used DDS4Chl to analyze the impact of
-drought on different types of land cover classes in continental Chile.
-The specific objectives of the study are: i) to analyze the trend on
-multi-scalar drought indices for water demand and supply, soil moisture,
-and vegetation productivity for 1981--2023/2001--2023; ii) to evaluate
-LULC change for 2001--2021 and its relation to drought indices; iii) to
+frequency. Those were gathered from the earth observation products ERA5L
+and MODIS. Then, we used DDS4Chl to analyze the impact of drought on
+different types of land cover classes in continental Chile. The specific
+objectives of the study are: i) to analyze the trend on multi-scalar
+drought indices for water demand and supply, soil moisture, and
+vegetation productivity for 1981--2023/2001--2023; ii) to evaluate LULC
+change for 2001--2021 and its relation to drought indices; iii) to
 analyze the relationship of a proxy of vegetation productivity (zcNDVI)
@@
 
-\hypertarget{study-area}{%
-\section{Study area}\label{study-area}}
+\hypertarget{study-area-agregar-superficie-de-landcover-en-cada-macrozona}{%
+\section{Study area (AGREGAR SUPERFICIE DE LANDCOVER EN CADA
+MACROZONA)}\label{study-area-agregar-superficie-de-landcover-en-cada-macrozona}}
 
@@
 data, we used the open source software for statistical computing and
-graphics, \texttt{R} \citep{R2023}. For downloading ERA5-Land, we used
-the \texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing
-raster data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
+graphics, \texttt{R} \citep{R2023}. For downloading ERA5L, we used the
+\texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing raster
+data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
 \texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
@@
 
-For water supply and demand variables, we used ERA5-Land
+For water supply and demand variables, we used ERA5L
 \citep{MunozSabater2021}, a reanalys dataset that provides the evolution
@@
 We derived the drought indices of water supply and demand, soil moisture
-from the ERA5-Land dataset, and vegetation from the MODIS product, all
-at monthly frequency.
+from the ERA5L dataset, and vegetation from the MODIS product, all at
+monthly frequency.
 
@@
 
-\hypertarget{validation-of-era5-land-variables}{%
-\subsection{Validation of ERA5-Land
-variables}\label{validation-of-era5-land-variables}}
+\hypertarget{validation-of-era5l-variables}{%
+\subsection{Validation of ERA5L
+variables}\label{validation-of-era5l-variables}}
 
@@
 from the ``Centro'' to the ``Sur.'' Also, there is a high increase in
-forest (397 \(Km^2 yr^{-1}\)) in the ``Sur,'' replacing the savanna
+forest (397 \(km^2 yr^{-1}\)) in the ``Sur,'' replacing the savanna
 lost.
@@
 Vegetation reached its lowest values since the year 2019, reaching an
-extreme condition in early 2020 and 2022 in the ``Centro'' (Mega
-Drought). The ``Sur'' and ``Austral'' show a positive trend of around
-0.016 per decade (Figure~\ref{fig-zcNDVI_var}). The area most affected
-by a negative trend in vegetation seems to be associated with the
-Chilean matorral \citep{Fuentes2021}, despite the croplands being as
-severely impacted by drought as native vegetation does in ``Norte
-Chico.''.
+extreme condition in early 2020 and 2022 in the ``Norte Chico'' and
+Centro'' (Mega Drought). The ``Sur'' and ``Austral'' show a positive
+trend of around 0.016 per decade (Figure~\ref{fig-zcNDVI_var}). The area
+most affected by a negative trend in vegetation seems to be associated
+with the Chilean matorral \citep{Fuentes2021}, despite the croplands
+being as severely impacted by drought as native vegetation does in
+``Norte Chico.''.
 
@@
 
-The correlation between zcNDVI-6 and drought indices shows that time
-scales higher than six months are predominant over continental Chile in
-explaining vegetation variability (Figure~\ref{fig-corTimeScale}).
-Pearson correlations between 0.6 and 0.9 are concentrated in the ``norte
-chico'' and ``zona central'' for SPI and SPEI. Soil moisture reached a
-correlation of 0.6--0.9 that extended north and south with respect to
-SPI.
+\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-chequear-cifras-en-el-texto}{%
+\subsection{Impact for water supply and demand, and soil moisture in
+vegetation productivity (CHEQUEAR CIFRAS EN EL
+TEXTO)}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-chequear-cifras-en-el-texto}}
 
-\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}{%
-\subsection{Impact for water supply and demand, and soil moisture in
-vegetation
-productivity}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}}
+According with what is showed Figure~\ref{fig-corTimeScale},
+Figure~\ref{fig-corPerson}, and Table 3. Forest seems to be the most
+resistant type to drought, showing that only ``Centro'' is slightly (rsq
+= 0.3) impacted by a 12-month soil moisture deficit (SSI-12). Following
+by grasslands, which show to be impacted by a SSI-12 with a rsq = 0.4,
+and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.3) in
+the ``Norte Chico'' and to a lesser degree in the ``Norte Grande.'' But,
+in the southern macrozones, this type was not affected by water demand,
+water supply, or soil moisture. The types that show to be most affected
+by variation in climate conditions are shrublands, savannas, and
+croplands. For savanna in ``Norte Chico,'' the SSI-12 and SPI-24 reached
+an rsq of 0.8 and 0.6, respectively. This value decreases to the south,
+but the SSI-12 is still the variable explaining more of the variation in
+vegetation productivity (rsq = 0.4 in ``Centro'' and 0.2 in ``Sur''). In
+the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between
+40\% and 75\% in ``Centro.'' The most affected type by climatic
+variation was shrubland, where in ``Norte Chico'' and ``Centro'' soil
+moisture and precipitation explained more than 50\%, with SSI-12 being
+the most relevant variable, followed by SPI-36 in ``Norte Chico'' and
+SPI-24 in ``Sur.''
 
@@
 \label{tab-corLandcover}
-\caption{}
+\caption{Summarry per landcover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
+The numbers in each cell indicate the time scale that reached the maximum correlation for the landcover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
 \includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
@@
 
-\hypertarget{validation-of-era5-land-variables-1}{%
-\subsection{Validation of ERA5-Land
-variables}\label{validation-of-era5-land-variables-1}}
+\hypertarget{validation-of-era5l-variables-1}{%
+\subsection{Validation of ERA5L
+variables}\label{validation-of-era5l-variables-1}}
 
@@
 
+1.- Respecto a lo que indica \citep{Vicente-Serrano2018}, de que el
+aumento en la tendencia en severidad de la sequía (hidrológica) tiene
+que ver más con un aumento de la demanda de agua (LULCC) que a una
+tendencia en las condiciones climáticas (SPI-12).
+
+2.- Sobre los tipos de landcover más afectados por los indicadores de
+sequía. Asociación con el matorral chileno \citep{Fuentes2021}.
+Diferencia entre el Norte Chico y Centro y lo que pasa hacía el sur.
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  Como podrian servir estos resultados para desarrollar un predictor de
+  sequía basado.
+\end{enumerate}
+
+4.- Qué se podria hacer mejor en futuras investigaciones del tema.
+
+5.- Sequía, aridez, water scarcity. Si puede utilizar el mapa y
+proyección climática de Koppen\_Geiger \citep{Beck2023} para dirigir
+esta discusión
+
 \hypertarget{conclusion}{%
 \section{Conclusion}\label{conclusion}}
+
+\newpage
 

### 2024-01-18 article/paper_drought_Chile2023.qmd
actualizar articulo

@@
       model: 3p
-      cite-style: number
+      cite-style: authoryear
       natbiboptions: numbers
@@
 In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
-In "Centro," there is a major proportion (~50%) of chilean matorral (shrubland and savanna, [@Fuentes2021]), followed by grassland (~16%), forest (8%), and croplands (5%).
-The south part of "Centro" and the north part of "Austral" are dominated by an Oceanic climate (Cfb).
+In "Centro," there is a major proportion (~50%) of Chilean matorral (shrubland and savanna, [@Fuentes2021]), followed by grassland (~16%), forest (8%), and croplands (5%).
+The south part of "Sur" and the north part of "Austral" are dominated by an Oceanic climate (Cfb).
 Those zones are high in forest and grassland.
-The southern part of the country has a tundra climate, and in Patagonia, it is a cold semi-arid area with an extended surface of grassland.
+The southern part of the country has a tundra climate, and in Patagonia, it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.
 
@@
 
+### Validation of ERA5L variables
+
 ## Drought Indices
@@
 We derived the drought indices of water supply and demand, soil moisture from the ERA5L dataset, and vegetation from the MODIS product, all at monthly frequency.
+
+### Atmospheric Evaporative Demand (AED)
 
@@
 
+### Non-parametric calculation of drought indices
+
 To evaluate water demand, we chose the $EDDI$ [@Hobbins2016; @McEvoy2016] index, which uses the $AED$.
@@
 Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI.
-Before using the NDVI, it was smoothed following the procedure described in @Zambrano2018 and @Zambrano2016.
+Before using the NDVI, it was smoothed using a locally-weighted polynomial regression, following the procedure described in @Zambrano2018 and @Zambrano2016.
 
@@
 
-## Validation of ERA5L variables
-
 #	Results
@@
 
-@fig-zcNDVI_var and \ref{tab-landcoverTrend} show that the macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively.
+@fig-zcNDVI_var and Table \ref{tab-landcoverTrend} show that the macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively.
 The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna -70 $km^2 year^{-1}$.
-In the "Centro" and "Sur" there are changes in the chilean matorral, with an important reduction in savanna (-136 to -318 $km^2 yr^{-1}$), and an increase in shrubland and grassland.
+In the "Centro" and "Sur" there are changes in the Chilean matorral, with an important reduction in savanna (-136 to -318 $km^2 yr^{-1}$), and an increase in shrubland and grassland.
 Showing a change for more dense vegetation types.
@@
 \begin{table}[!ht]
-\label{tab-landcoverTrend}
 \caption{The value of Sen's slope trend next to the time-series plot of surface per landcover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile.
@@
 Red dots on the plots indicate the maximum and minimum values of surface.}
+\label{tab-landcoverTrend}
 \includegraphics[]{../output/figs/table_var_landcover_macro.png}
@@
 
-Analyzing the water supply in the macrozones "norte chico," "central," and "sur," the SPI shows a decreasing trend that increases at longer time scales due to the prolonged reduction in precipitation.
-It reaches a trend of -0.056 (z-score) per decade (@fig-trendSPI) for time scales of 36 months.
-In the case of water demand, we analyze the EDDI.
-It shows a positive trend, caused by an increase in AED.
-The trend on EDDI reaches a maximum of 0.053 per decade in the macrozones "norte grande" and "norte chico" (@fig-trendEDDI).
-The behavior is similar to SPI, having close trends with opposite signs.
-Nevertheless, the spatial patterns are different.
-The maximum trend for SPI-36 is in "norte chico" and "zona central." For EDDI-36, the maximum trend is from "northe grande" to "zona central."
+Analyzing the water supply, the macrozones that have the lowest trend are "Norte Chico" and "Centro," where the SPI, SPEI, and SSI show that it decreases at longer time scales due to the prolonged reduction in precipitation.
+At 36 months, it reaches trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and SSI (@fig-trendDI).
+For "Sur," the behavior is similar, decreasing at longer scales and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI.
+On the other hand, all macrozones show an increase in the trend in all the drought indices, with "Norte Grande" having the highest at 36 months (0.042 per decade).
+Because of this, the SPEI (which uses AED) reached its lowest value in "Norte Grande," with -0.03 at 36 months.
+Despite the other macrozones, "Austral" showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend (@fig-trendDI).
+
+##	Impact for water supply and demand, and soil moisture in vegetation productivity
+
+According to what is shown in @fig-corTimeScale, @fig-corPerson, and Table \ref{tab-corLandcover}, forest seems to be the most resistant type to drought.
+Showing that only "Centro" is slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit (SSI-12).
+Grasslands are shown to be impacted by a SSI-12 with a rsq = 0.45 and by a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34, respectively) in the "Norte Chico" and to a lesser degree in the "Norte Grande." But, in further south macrozones, this type was not affected by water demand, water supply, or soil moisture.
+The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
+For savannas in "Norte Chico," the SSI-12 and SPI-24 reached an rsq of 0.74 and 0.58, respectively.
+This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (rsq = 0.45 in "Centro" and 0.2 in "Sur").
+In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45% and 66% of "Norte Chico." The most affected type by climatic variation was shrubland, where in "Norte Chico" and "Centro" soil moisture explained 59% and precipitation, 37%, with SSI-12 being the most relevant variable, followed by SPI-36 in "Norte Chico" and SPI-24 in "Sur."
 
@@
 \blandscape
-```{r fig-trendSPI}
+```{r fig-trendDI}
 #| echo: false
@@
 
-##	Impact for water supply and demand, and soil moisture in vegetation productivity (CHEQUEAR CIFRAS EN EL TEXTO)
-
-According with what is showed @fig-corTimeScale, @fig-corPerson, and Table 3.
-Forest seems to be the most resistant type to drought, showing that only "Centro" is slightly (rsq = 0.3) impacted by a 12-month soil moisture deficit (SSI-12).
-Following by grasslands, which show to be impacted by a SSI-12 with a rsq = 0.4, and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.3) in the "Norte Chico" and to a lesser degree in the "Norte Grande." But, in the southern macrozones, this type was not affected by water demand, water supply, or soil moisture.
-The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
-For savanna in "Norte Chico," the SSI-12 and SPI-24 reached an rsq of 0.8 and 0.6, respectively.
-This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (rsq = 0.4 in "Centro" and 0.2 in "Sur").
-In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 40% and 75% in "Centro." The most affected type by climatic variation was shrubland, where in "Norte Chico" and "Centro" soil moisture and precipitation explained more than 50%, with SSI-12 being the most relevant variable, followed by SPI-36 in "Norte Chico" and SPI-24 in "Sur."
-
 ```{=latex}
 \begin{table}[!ht]
-\label{tab-corLandcover}
 \caption{Summarry per landcover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
 The numbers in each cell indicate the time scale that reached the maximum correlation for the landcover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
+\label{tab-corLandcover}
 \includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
@@
 
-5.- Sequía, aridez, water scarcity [VanLoon2013] Si puede utilizar el mapa y proyección climática de Koppen_Geiger [@Beck2023] para dirigir esta discusión
+5.- Sequía, aridez, water scarcity [@VanLoon2013] Si puede utilizar el mapa y proyección climática de Koppen_Geiger [@Beck2023] para dirigir esta discusión
 

### 2024-01-18 article/paper_drought_Chile2023.tex
actualizar articulo

@@
 \documentclass[
-  number,
+  authoryear,
   preprint,
@@
 \usepackage[]{natbib}
-\bibliographystyle{elsarticle-num}
+\bibliographystyle{elsarticle-harv}
 \IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 classified as meteorological, hydrological, and agricultural
-\citep{Wilhite1985}. Lately, \citep{Loon2016} and
-\citep{AghaKouchak2021} have given an updated definition of drought for
+\citep{Wilhite1985}. Lately, \citet{Loon2016} and
+\citet{AghaKouchak2021} have given an updated definition of drought for
 the Anthropocene, suggesting that it should be considered the feedback
@@
 drought. Even though it has been argued that those definitions do not
-fully address the ecological dimensions of drought. \citep{Crausbay2017}
+fully address the ecological dimensions of drought. \citet{Crausbay2017}
 proposed the ecological drought definition as \emph{``an episodic
@@
 To evaluate meteorological drought (i.e., water supply), the World
-Meteorological Organization (WMO; \citep{WMO2012}) recommends the
-Standardized Precipitation Index (SPI; \citep{Mckee1993}), a multiscalar
+Meteorological Organization (WMO; \citet{WMO2012}) recommends the
+Standardized Precipitation Index (SPI; \citet{Mckee1993}), a multiscalar
 drought index that allows to monitor precipitation deficits from short-
-to long-term. Following the same approach, \citep{Vicente-Serrano2010}
+to long-term. Following the same approach, \citet{Vicente-Serrano2010}
 incorporates into the SPI the effect of temperature through the use of
@@
 Precipitation Evapotranspiration Index). Similarly, to evaluate solely
-the evaporative demand driven by temperature, \citep{Hobbins2016} and
-\citep{McEvoy2016} came up with the Evaporative Demand Drought Index
+the evaporative demand driven by temperature, \citet{Hobbins2016} and
+\citet{McEvoy2016} came up with the Evaporative Demand Drought Index
 (EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI;
-\citep{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
+\citet{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
 cumulative Normalized Difference Vegetation Index (NDVI), which could be
@@
 \citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
-use the Koppen-Geiger release by \citep{Beck2023} and the landcover type
+use the Koppen-Geiger release by \citet{Beck2023} and the landcover type
 persistance of 80\% of times for 2001--2022, from the IGBP
@@
 In ``Centro,'' there is a major proportion (\textasciitilde50\%) of
-chilean matorral (shrubland and savanna, \citep{Fuentes2021}), followed
+Chilean matorral (shrubland and savanna, \citep{Fuentes2021}), followed
 by grassland (\textasciitilde16\%), forest (8\%), and croplands (5\%).
-The south part of ``Centro'' and the north part of ``Austral'' are
+The south part of ``Sur'' and the north part of ``Austral'' are
 dominated by an Oceanic climate (Cfb). Those zones are high in forest
@@
 and in Patagonia, it is a cold semi-arid area with an extended surface
-of grassland.
+of grassland, forest, and, to a lesser extent, savanna.
 
@@
 
+\hypertarget{validation-of-era5l-variables}{%
+\subsubsection{Validation of ERA5L
+variables}\label{validation-of-era5l-variables}}
+
 \hypertarget{drought-indices}{%
@@
 monthly frequency.
+
+\hypertarget{atmospheric-evaporative-demand-aed}{%
+\subsubsection{Atmospheric Evaporative Demand
+(AED)}\label{atmospheric-evaporative-demand-aed}}
 
@@
 
+\hypertarget{non-parametric-calculation-of-drought-indices}{%
+\subsubsection{Non-parametric calculation of drought
+indices}\label{non-parametric-calculation-of-drought-indices}}
+
 To evaluate water demand, we chose the \(EDDI\)
@@
 of productivity, \(zcNDVI\), we used the NDVI. Before using the NDVI, it
-was smoothed following the procedure described in \citep{Zambrano2018}
-and \citep{Zambrano2016}.
+was smoothed using a locally-weighted polynomial regression, following
+the procedure described in \citet{Zambrano2018} and
+\citet{Zambrano2016}.
 
@@
 for the scale \(s\), we used a nonparametric approach following
-\citep{Hobbins2016} to derive the drought indices. Thus, the empirically
+\citet{Hobbins2016} to derive the drought indices. Thus, the empirically
 derived probabilities are obtained through an inverse normal
@@
 the IGBP MODIS, we compare the macroclasses with the ones of a more
-detailed landcover map made by \citep{Zhao2016} for Chile for the years
+detailed landcover map made by \citet{Zhao2016} for Chile for the years
 2013--2014 (LCChile). The later has a spatial resolution of 30 m and
@@
 
-\citep{Vicente-Serrano2021} made a global analysis of the drought's
+\citet{Vicente-Serrano2021} made a global analysis of the drought's
 severity trend using SPI, SPEI, and the Standardized Evapotranspiration
-Deficit Index (SEDI; \citep{Vicente-Serrano2018}) to evaluate AED. They
+Deficit Index (SEDI; \citet{Vicente-Serrano2018}) to evaluate AED. They
 indicate that the increase in hydrological drought has been due to
@@
 36-month time scales against zcNDVI. We followed a similar approach to
-that used by \citep{Meroni2016} when using the SPI for meteorological
+that used by \citet{Meroni2016} when using the SPI for meteorological
 drought against the cumulative FAPAR (Fraction of Absorbed
@@
 
-\hypertarget{validation-of-era5l-variables}{%
-\subsection{Validation of ERA5L
-variables}\label{validation-of-era5l-variables}}
-
 \hypertarget{results}{%
@@
 
-Figure~\ref{fig-zcNDVI_var} and \ref{tab-landcoverTrend} show that the
-macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'', and
-``Austral'' with 36\%, 31\%, and 34\%, respectively. The ``Norte Chico''
-shows an increase in barrend land of 111 \(km^2 year^{-1}\) and a
-reduction in the class savanna -70 \(km^2 year^{-1}\). In the ``Centro''
-and ``Sur'' there are changes in the chilean matorral, with an important
-reduction in savanna (-136 to -318 \(km^2 yr^{-1}\)), and an increase in
-shrubland and grassland. Showing a change for more dense vegetation
-types. It appears to be a change from the surface occupied by cropland
-from the ``Centro'' to the ``Sur.'' Also, there is a high increase in
-forest (397 \(km^2 yr^{-1}\)) in the ``Sur,'' replacing the savanna
-lost.
+Figure~\ref{fig-zcNDVI_var} and Table \ref{tab-landcoverTrend} show that
+the macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'',
+and ``Austral'' with 36\%, 31\%, and 34\%, respectively. The ``Norte
+Chico'' shows an increase in barrend land of 111 \(km^2 year^{-1}\) and
+a reduction in the class savanna -70 \(km^2 year^{-1}\). In the
+``Centro'' and ``Sur'' there are changes in the Chilean matorral, with
+an important reduction in savanna (-136 to -318 \(km^2 yr^{-1}\)), and
+an increase in shrubland and grassland. Showing a change for more dense
+vegetation types. It appears to be a change from the surface occupied by
+cropland from the ``Centro'' to the ``Sur.'' Also, there is a high
+increase in forest (397 \(km^2 yr^{-1}\)) in the ``Sur,'' replacing the
+savanna lost.
 
 \begin{table}[!ht]
-\label{tab-landcoverTrend}
 \caption{The value of Sen's slope trend next to the time-series plot of surface per landcover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile.
@@
 
-Analyzing the water supply in the macrozones ``norte chico,''
-``central,'' and ``sur,'' the SPI shows a decreasing trend that
-increases at longer time scales due to the prolonged reduction in
-precipitation. It reaches a trend of -0.056 (z-score) per decade
-(Figure~\ref{fig-trendSPI}) for time scales of 36 months. In the case of
-water demand, we analyze the EDDI. It shows a positive trend, caused by
-an increase in AED. The trend on EDDI reaches a maximum of 0.053 per
-decade in the macrozones ``norte grande'' and ``norte chico''
-(\textbf{?@fig-trendEDDI}). The behavior is similar to SPI, having close
-trends with opposite signs. Nevertheless, the spatial patterns are
-different. The maximum trend for SPI-36 is in ``norte chico'' and ``zona
-central.'' For EDDI-36, the maximum trend is from ``northe grande'' to
-``zona central.''
+Analyzing the water supply, the macrozones that have the lowest trend
+are ``Norte Chico'' and ``Centro,'' where the SPI, SPEI, and SSI show
+that it decreases at longer time scales due to the prolonged reduction
+in precipitation. At 36 months, it reaches trends between -0.03 and
+-0.04 (z-score) per decade for SPI, SPEI, and SSI
+(Figure~\ref{fig-trendDI}). For ``Sur,'' the behavior is similar,
+decreasing at longer scales and having between -0.016 and -0.025 per
+decade for SPI, SPEI, and SSI. On the other hand, all macrozones show an
+increase in the trend in all the drought indices, with ``Norte Grande''
+having the highest at 36 months (0.042 per decade). Because of this, the
+SPEI (which uses AED) reached its lowest value in ``Norte Grande,'' with
+-0.03 at 36 months. Despite the other macrozones, ``Austral'' showed an
+increase in all indices, being the highest for EDDI at 36 months (0.025)
+and the lowest for SSI, which shows only a minor increase in the trend
+(Figure~\ref{fig-trendDI}).
+
+\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}{%
+\subsection{Impact for water supply and demand, and soil moisture in
+vegetation
+productivity}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}}
+
+According to what is shown in Figure~\ref{fig-corTimeScale},
+Figure~\ref{fig-corPerson}, and Table \ref{tab-corLandcover}, forest
+seems to be the most resistant type to drought. Showing that only
+``Centro'' is slightly (rsq = 0.25) impacted by a 12-month soil moisture
+deficit (SSI-12). Grasslands are shown to be impacted by a SSI-12 with a
+rsq = 0.45 and by a decrease in water supply (SPI-36 and SPEI-24 with
+rsq = 0.28 and 0.34, respectively) in the ``Norte Chico'' and to a
+lesser degree in the ``Norte Grande.'' But, in further south macrozones,
+this type was not affected by water demand, water supply, or soil
+moisture. The types that show to be most affected by variation in
+climate conditions are shrublands, savannas, and croplands. For savannas
+in ``Norte Chico,'' the SSI-12 and SPI-24 reached an rsq of 0.74 and
+0.58, respectively. This value decreases to the south, but the SSI-12 is
+still the variable explaining more of the variation in vegetation
+productivity (rsq = 0.45 in ``Centro'' and 0.2 in ``Sur''). In the case
+of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45\% and
+66\% of ``Norte Chico.'' The most affected type by climatic variation
+was shrubland, where in ``Norte Chico'' and ``Centro'' soil moisture
+explained 59\% and precipitation, 37\%, with SSI-12 being the most
+relevant variable, followed by SPI-36 in ``Norte Chico'' and SPI-24 in
+``Sur.''
 
@@
 
-\subcaption{\label{fig-trendSPI-1}SPI (Standardized Precipitation
-Index)}
+\subcaption{\label{fig-trendDI-1}SPI (Standardized Precipitation Index)}
 \end{minipage}%
@@
 
-\subcaption{\label{fig-trendSPI-2}SPEI (Standardized Precipitation
+\subcaption{\label{fig-trendDI-2}SPEI (Standardized Precipitation
 Evapotranspiration Index)}
@@
 
-\subcaption{\label{fig-trendSPI-3}EDDI (Evaporative Demand Drought
+\subcaption{\label{fig-trendDI-3}EDDI (Evaporative Demand Drought
 Index)}
@@
 
-\subcaption{\label{fig-trendSPI-4}SSMI (Standardized Soil Moisture
+\subcaption{\label{fig-trendDI-4}SSMI (Standardized Soil Moisture
 Index)}
@@
 
-\caption{\label{fig-trendSPI}Linear trend of the drought index (*) at
+\caption{\label{fig-trendDI}Linear trend of the drought index (*) at
 time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023}
@@
 
-\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-chequear-cifras-en-el-texto}{%
-\subsection{Impact for water supply and demand, and soil moisture in
-vegetation productivity (CHEQUEAR CIFRAS EN EL
-TEXTO)}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-chequear-cifras-en-el-texto}}
-
-According with what is showed Figure~\ref{fig-corTimeScale},
-Figure~\ref{fig-corPerson}, and Table 3. Forest seems to be the most
-resistant type to drought, showing that only ``Centro'' is slightly (rsq
-= 0.3) impacted by a 12-month soil moisture deficit (SSI-12). Following
-by grasslands, which show to be impacted by a SSI-12 with a rsq = 0.4,
-and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.3) in
-the ``Norte Chico'' and to a lesser degree in the ``Norte Grande.'' But,
-in the southern macrozones, this type was not affected by water demand,
-water supply, or soil moisture. The types that show to be most affected
-by variation in climate conditions are shrublands, savannas, and
-croplands. For savanna in ``Norte Chico,'' the SSI-12 and SPI-24 reached
-an rsq of 0.8 and 0.6, respectively. This value decreases to the south,
-but the SSI-12 is still the variable explaining more of the variation in
-vegetation productivity (rsq = 0.4 in ``Centro'' and 0.2 in ``Sur''). In
-the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between
-40\% and 75\% in ``Centro.'' The most affected type by climatic
-variation was shrubland, where in ``Norte Chico'' and ``Centro'' soil
-moisture and precipitation explained more than 50\%, with SSI-12 being
-the most relevant variable, followed by SPI-36 in ``Norte Chico'' and
-SPI-24 in ``Sur.''
-
 \begin{table}[!ht]
-\label{tab-corLandcover}
 \caption{Summarry per landcover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
 The numbers in each cell indicate the time scale that reached the maximum correlation for the landcover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
+\label{tab-corLandcover}
 \includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
@@
 
-1.- Respecto a lo que indica \citep{Vicente-Serrano2018}, de que el
+1.- Respecto a lo que indica \citet{Vicente-Serrano2018}, de que el
 aumento en la tendencia en severidad de la sequía (hidrológica) tiene
@@
 
-5.- Sequía, aridez, water scarcity. Si puede utilizar el mapa y
-proyección climática de Koppen\_Geiger \citep{Beck2023} para dirigir
-esta discusión
+5.- Sequía, aridez, water scarcity \citep{VanLoon2013} Si puede utilizar
+el mapa y proyección climática de Koppen\_Geiger \citep{Beck2023} para
+dirigir esta discusión
 

### 2024-01-18 scripts/25_trend_drought_indices.R

@@ -50,6 +50,11 @@ df_zcndvi |>
 ggsave('output/figs/temporal_variation_zcNDVI6_macrozonas.png',scale = 2,bg = 'white',width=6,height=3)
 
 #Mapa de tendencia Mann-Kendall
+
+#breaks para escalas
+
+breaks <- seq(-0.06,.06,length.out = 6)
+
 library(fs)
 dir <- '/mnt/md0/raster_procesada/analysis/trends/'
 files <- dir_ls(dir,regexp = 'mann.*SPI.*tif$')
@@ -63,16 +68,27 @@ trend_spi <- app(trend_spi,\(x){
 
 ind_ord <- scales |> as.numeric() |> order()
 trend_spi <- subset(trend_spi,ind_ord)
-names(trend_spi) <- paste('SPI',scales,sep ='-')
+names(trend_spi) <- paste('SPI',scales[ind_ord],sep ='-')
+
+df_spi <- terra::extract(trend_spi,macro,fun = mean, na.rm = TRUE) |> 
+  mutate(macro = macro$macrozona) |> 
+  select(-ID) |> 
+  pivot_longer(-macro) |> 
+  mutate(scale = str_remove(name,'SPI-') |> as.numeric())
 
 map_spi <- tm_shape(trend_spi*10) + 
-  tm_raster(palette = '-inferno',midpoint = 0,title = 'Trend SPI \n (per decade)',style = 'kmeans') +
+  tm_raster(palette = 'inferno',
+            midpoint = 0,
+            title = 'Trend SPI \n (per decade)',
+            style = 'cont',
+            breaks = breaks) +
   tm_shape(macro) + 
   tm_borders(col='white') +
   tm_shape(chl_b) + 
   tm_borders(col = 'black') +
   tm_facets(nrow=1) +
-  tm_layout(panel.labels = paste0('SPI-',c(1,3,6,12,24,36)))
+  tm_layout(panel.labels = paste0('SPI-',c(1,3,6,12,24,36)),
+            panel.label.bg.color = 'white')
 tmap_save(map_spi,'output/figs/trend_raster_SPI_1981-2023.png',asp=.2)
 
 ## SPEI
@@ -88,16 +104,27 @@ trend_spei <- app(trend_spei,\(x){
 
 ind_ord <- scales |> as.numeric() |> order()
 trend_spei <- subset(trend_spei,ind_ord)
[…]

### 2024-01-19 article/paper_drought_Chile2023.qmd
actualizar articulo

@@
 
-# Study area (AGREGAR SUPERFICIE DE LANDCOVER EN CADA MACROZONA)
+# Study area
 
@@
 
+POR COMPLETAR
+
 ### Validation of ERA5L variables
+
+POR COMPLETAR
 
@@
 
-We derived the drought indices of water supply and demand, soil moisture from the ERA5L dataset, and vegetation from the MODIS product, all at monthly frequency.
-
 ### Atmospheric Evaporative Demand (AED)
 
-For the indices EDDI and SPEI that use water demand, first we have to calculate the AED. For this, we used the method of Hargreaves [@Hargreaves1994]:
+For the indices EDDI and SPEI that use water demand, first we have to calculate the AED. For this, we used the method of Hargreaves [@Hargreaves1994;@Hargreaves1985]:
 
@@
 
+We chose the method of Hargreaves to estimate AED because of its simplicity, which only requires temperatures and extrarrestrial radiation.
+Also, it has been recommended over other methods when the use of several climatic variables is limited [@Vicente-Serrano2014].
+
 ### Non-parametric calculation of drought indices
+
+We derived the drought indices of water supply and demand, soil moisture from the ERA5L dataset, and vegetation from the MODIS product, all at monthly frequency.
 
@@
 We calculated the SPEI, which used a balance between $P$ and $AED$, in this case, an auxiliary variable $D = P-AED$ is used.
-In this study, we used the $SSI$ (standardized soil moisture index at 1 m) [@Hao2013;@AghaKouchak2014], which uses SM.
+In this study, we used the $SSI$ (standardized soil moisture index at 1 m) [@Hao2013;@AghaKouchak2014], which uses soil moisture at 1m depth.
 Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI.
@@
 
-```{r fig-LCprop}
-#| echo: false
-#| fig-pos: "!ht"
-#| fig-cap: "Proportion of land cover class from the persistent land cover for 2001-2022 (>80%) per macrozone"
-#| out-width: "\\textwidth"
-knitr::include_graphics("../output/figs/LC_pers80_per_macrozone.png")
-```
+### Landcover macroclasess and validation
 
@@
 This product has a yearly frequency from 2001 to 2022.
-The IGBP defines 17 classes; from these, we regrouped into ten macroclasses for forest, schrublands, savannas, grasslands, wetlands, croplands, urban, snow and ice, barren, and water bodies.
-To validate the IGBP MODIS, we compare the macroclasses with the ones of a more detailed landcover map made by @Zhao2016 for Chile for the years 2013–2014 (LCChile).
-The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses derived from the IGBP MODIS.
-We chose the years 2013 and 2014 from MODIS IGBP because the LCChile was made with data acquired in 2013 and 2014.
-We resampled LCChile to the spatial resolution of the IGBP MODIS using the nearest neighbor method.
-Then, we took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCCHILE, IGP2013, and IGBP2014.
-Finally, we applied a confusion matrix and estimated the accuracy and F1 score as metrics of performance.
-The validation of IGBP2013 and IGBP2014 with LCChile reached similar metrics of performance, having an accuracy of 0.82 and a F1 score of 0.66 (see SS1).
+The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
+Thus, we have a landcover raster time-series with the ten classes for 2001 and 2023.
 
-Later, with the goal of analyzing vegetation areas less affected by landcover change, we calculated a raster mask for IGBP MODIS considering the classes that remain for more than 80% of the years (2001–2022), which allows us to identify the areas with no landcover change for the macroclasses.
-@fig-LCprop shows the summary of the proportion of surface per landcover class and macrozone, derived from the mask over continental Chile.
+To validate the landcover obtained, we compare the macroclasses with the ones of a more detailed landcover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile).
+The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses landcover.
+We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from landcover macrolcasses to validate with LCChile.
 
-Further, we calculated the surface occupied per landcover class into the five macrozones ("Norte Grande" to "Austral") per year.
+We follow the next procedure:
+
+i) resampled LCChile to the spatial resolution (500m) of the landcover macroclasses using the nearest neighbor method,
+ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values as the other two years as prediction
+iii) calculate a confusion matrix with the classes extracted in the 1000 poitns for LCChile, IGBP2013, and IGBP2014. Calculate the performance metrics of accuracy and F1.
+
+$$Accuracy = \frac{TP+TN}{TP+TN+FP+FN}=\frac{correct\,\, classifications}{all\,\, classifications}$$
+$$F1=\frac{2\cdot TP}{2\cdot TP + FP +FN}$$
+where $TP$ and $FN$ refer to true positive and false negative, correctly classified classes; $TN$ and $FP$ to true negative and false positive, wrongly classified classes.
+
+### Landcover persistence mask 2001-2022
+
+The time series of NDVI is affected by climatic conditions, vegetation development, seasonality, and changes in vegetation type.
+In this study, we want to analyze the variation in vegetation productivity in different landcover types and how it is affected by water demand, water supply, and soil moisture.
+In order to avoid changes due to a change in the landcover type, that will wrongly impact NDVI.
+We will develop a persistence mask for landcover for 2001–2023.
+Thereby, we reduce an important source of variation on a regional scale.
+
+Thus, we calculated a raster mask for IGBP MODIS considering the macroclasses that remain without change for more than 80% of the years (2001–2022) per pixel, which allows us to identify the areas with no landcover change for the macroclasses.
+
+### Landcover trend and drought indices
+
+We calculated the surface occupied per landcover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023.
 After that, we calculated the trend's change in surface; we used the Sen' slope [@Sen1968] based on Mann-Kendall [@Kendall1975].
 This way, we obtain a matrix of trends of 5 x 5 (macrozones x landcover).
-
-We will explore if the trend in landcover classess is associated to trend of the drought indices.
-For this we will use the regularizations techniques of Lasso [@Tibshirani2010] and Ridge regressión [@Hoerl1970].
-Also, we will test random forest for this purpose [@Ho1995].
+The aim is to later explore if the trend in landcover classes is associated with a trend in the drought indices.
+For this, we will use the techniques of regresion and regularization of Lasso [@Tibshirani2010] and Ridge [@Hoerl1970].
+Also, we will test random forests for this purpose [@Ho1995].
+We will choose the trend of landcover surface per macroclass and macrozone as the response variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor variables.
+With this analysis, we expect to gather insights regarding whether there is a pattern of climatic influence along Chile or if what is happening in Central Chile has to do with more localized climatic conditions.
 
@@
 
+### Mann-Kendall and Sen's slope
+
 To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975].
 To determine the magnitude of the trend, we used Sen's slope [@Sen1968].
-One of the advantages of applying this methodology is that the Sen's slope is not affected by outliers as regular regression does.
+Some of the advantages of applying this methodology are that the Sen's slope is not affected by outliers as regular regression does, and it is a non-paramteric method that is not affected by the distribution of the data.
 We applied both to the six time scales from 1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI.
@@
 
-For zcNDVI, we want to remove the effect of the Andes's mountain and mask out trends that were at an elevation above 1500m because there is scarce vegetation.
+### Trend in vegetation productivity without landcover change
 
@@
 
+## Data
+
+### Validation of ERA5L variables
+
 ##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture
 
-@fig-zcNDVI_var and Table \ref{tab-landcoverTrend} show that the macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively.
-The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna -70 $km^2 year^{-1}$.
-In the "Centro" and "Sur" there are changes in the Chilean matorral, with an important reduction in savanna (-136 to -318 $km^2 yr^{-1}$), and an increase in shrubland and grassland.
-Showing a change for more dense vegetation types.
-It appears to be a change from the surface occupied by cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.
+### Landcover macroclasess and validation
+
+The validation of IGBP2013 and IGBP2014 with LCChile reached similar metrics of performance, having an accuracy of 0.82 and a F1 score of 0.66 (see SS1).
+
+### Landcover persistence mask 2001-2022
+
+The macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend})
+
+```{r fig-LCprop}
+#| echo: false
+#| fig-pos: "!ht"
+#| fig-cap: "Proportion of land cover class from the persistent land cover for 2001-2022 (>80%) per macrozone"
+#| out-width: "\\textwidth"
+knitr::include_graphics("../output/figs/LC_pers80_per_macrozone.png")
+```
+
+@fig-LCprop shows the summary of the proportion of surface per landcover class and macrozone, derived from the mask over continental Chile.
+
+### Landcover trend and drought indices
 
@@
 \end{table}
+
+The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna -70 $km^2 year^{-1}$.
+In the "Centro" and "Sur" there are changes in the Chilean matorral, with an important reduction in savanna (-136 to -318 $km^2 yr^{-1}$), and an increase in shrubland and grassland.
+Showing a change for more dense vegetation types.
+It appears to be a change from the surface occupied by cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.
 
@@
 
-## Validation of ERA5L variables
-
 #	Discussion

### 2024-01-19 article/paper_drought_Chile2023.tex
actualizar articulo

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{study-area-agregar-superficie-de-landcover-en-cada-macrozona}{%
-\section{Study area (AGREGAR SUPERFICIE DE LANDCOVER EN CADA
-MACROZONA)}\label{study-area-agregar-superficie-de-landcover-en-cada-macrozona}}
+\hypertarget{study-area}{%
+\section{Study area}\label{study-area}}
 
@@
 
+POR COMPLETAR
+
 \hypertarget{validation-of-era5l-variables}{%
@@
 
+POR COMPLETAR
+
 \hypertarget{drought-indices}{%
 \subsection{Drought Indices}\label{drought-indices}}
-
-We derived the drought indices of water supply and demand, soil moisture
-from the ERA5L dataset, and vegetation from the MODIS product, all at
-monthly frequency.
 
@@
 calculate the AED. For this, we used the method of Hargreaves
-\citep{Hargreaves1994}:
+\citep{Hargreaves1994, Hargreaves1985}:
 
@@
 
+We chose the method of Hargreaves to estimate AED because of its
+simplicity, which only requires temperatures and extrarrestrial
+radiation. Also, it has been recommended over other methods when the use
+of several climatic variables is limited \citep{Vicente-Serrano2014}.
+
 \hypertarget{non-parametric-calculation-of-drought-indices}{%
@@
 indices}\label{non-parametric-calculation-of-drought-indices}}
+
+We derived the drought indices of water supply and demand, soil moisture
+from the ERA5L dataset, and vegetation from the MODIS product, all at
+monthly frequency.
 
@@
 we used the \(SSI\) (standardized soil moisture index at 1 m)
-\citep{Hao2013, AghaKouchak2014}, which uses SM. Finally, for the proxy
-of productivity, \(zcNDVI\), we used the NDVI. Before using the NDVI, it
-was smoothed using a locally-weighted polynomial regression, following
-the procedure described in \citet{Zambrano2018} and
-\citet{Zambrano2016}.
+\citep{Hao2013, AghaKouchak2014}, which uses soil moisture at 1m depth.
+Finally, for the proxy of productivity, \(zcNDVI\), we used the NDVI.
+Before using the NDVI, it was smoothed using a locally-weighted
+polynomial regression, following the procedure described in
+\citet{Zambrano2018} and \citet{Zambrano2016}.
 
@@
 
-\begin{figure}[!ht]
-
-{\centering \includegraphics{../output/figs/LC_pers80_per_macrozone.png}
-
-}
-
-\caption{\label{fig-LCprop}Proportion of land cover class from the
-persistent land cover for 2001-2022 (\textgreater80\%) per macrozone}
-
-\end{figure}
+\hypertarget{landcover-macroclasess-and-validation}{%
+\subsubsection{Landcover macroclasess and
+validation}\label{landcover-macroclasess-and-validation}}
 
@@
 The IGBP defines 17 classes; from these, we regrouped into ten
-macroclasses for forest, schrublands, savannas, grasslands, wetlands,
-croplands, urban, snow and ice, barren, and water bodies. To validate
-the IGBP MODIS, we compare the macroclasses with the ones of a more
-detailed landcover map made by \citet{Zhao2016} for Chile for the years
-2013--2014 (LCChile). The later has a spatial resolution of 30 m and
-three levels of defined classes; from those, we used level 1, which fits
-with the macroclasses derived from the IGBP MODIS. We chose the years
-2013 and 2014 from MODIS IGBP because the LCChile was made with data
-acquired in 2013 and 2014. We resampled LCChile to the spatial
-resolution of the IGBP MODIS using the nearest neighbor method. Then, we
+macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9
+to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
+13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
+Thus, we have a landcover raster time-series with the ten classes for
+2001 and 2023.
+
+To validate the landcover obtained, we compare the macroclasses with the
+ones of a more detailed landcover map made by \citet{Zhao2016} for Chile
+with samples acquired in the years 2013--2014 (LCChile). The later has a
+spatial resolution of 30 m and three levels of defined classes; from
+those, we used level 1, which fits with the macroclasses landcover. We
+chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from landcover
+macrolcasses to validate with LCChile.
+
+We follow the next procedure:
+
+\begin{enumerate}
+\def\labelenumi{\roman{enumi})}
+\tightlist
+\item
+  resampled LCChile to the spatial resolution (500m) of the landcover
+  macroclasses using the nearest neighbor method,
+\item
   took a random sample of 1000 points within continental Chile and
-extracted the classes that fell within each point for LCCHILE, IGP2013,
-and IGBP2014. Finally, we applied a confusion matrix and estimated the
-accuracy and F1 score as metrics of performance. The validation of
-IGBP2013 and IGBP2014 with LCChile reached similar metrics of
-performance, having an accuracy of 0.82 and a F1 score of 0.66 (see
-SS1).
+  extracted the classes that fell within each point for LCChile,
+  IGBP2013, and IGBP2014; we considered the point extracted from LCChile
+  as the truth and the values as the other two years as prediction
+\item
+  calculate a confusion matrix with the classes extracted in the 1000
+  poitns for LCChile, IGBP2013, and IGBP2014. Calculate the performance
+  metrics of accuracy and F1.
+\end{enumerate}
 
-Later, with the goal of analyzing vegetation areas less affected by
-landcover change, we calculated a raster mask for IGBP MODIS considering
-the classes that remain for more than 80\% of the years (2001--2022),
-which allows us to identify the areas with no landcover change for the
-macroclasses. Figure~\ref{fig-LCprop} shows the summary of the
-proportion of surface per landcover class and macrozone, derived from
-the mask over continental Chile.
+\[Accuracy = \frac{TP+TN}{TP+TN+FP+FN}=\frac{correct\,\, classifications}{all\,\, classifications}\]
+\[F1=\frac{2\cdot TP}{2\cdot TP + FP +FN}\] where \(TP\) and \(FN\)
+refer to true positive and false negative, correctly classified classes;
+\(TN\) and \(FP\) to true negative and false positive, wrongly
+classified classes.
 
-Further, we calculated the surface occupied per landcover class into the
-five macrozones (``Norte Grande'' to ``Austral'') per year. After that,
-we calculated the trend's change in surface; we used the Sen' slope
-\citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}. This way, we
-obtain a matrix of trends of 5 x 5 (macrozones x landcover).
+\hypertarget{landcover-persistence-mask-2001-2022}{%
+\subsubsection{Landcover persistence mask
+2001-2022}\label{landcover-persistence-mask-2001-2022}}
 
-We will explore if the trend in landcover classess is associated to
-trend of the drought indices. For this we will use the regularizations
-techniques of Lasso \citep{Tibshirani2010} and Ridge regressión
-\citep{Hoerl1970}. Also, we will test random forest for this purpose
-\citep{Ho1995}.
+The time series of NDVI is affected by climatic conditions, vegetation
+development, seasonality, and changes in vegetation type. In this study,
+we want to analyze the variation in vegetation productivity in different
+landcover types and how it is affected by water demand, water supply,
+and soil moisture. In order to avoid changes due to a change in the
+landcover type, that will wrongly impact NDVI. We will develop a
+persistence mask for landcover for 2001--2023. Thereby, we reduce an
+important source of variation on a regional scale.
+
+Thus, we calculated a raster mask for IGBP MODIS considering the
+macroclasses that remain without change for more than 80\% of the years
+(2001--2022) per pixel, which allows us to identify the areas with no
+landcover change for the macroclasses.
+
+\hypertarget{landcover-trend-and-drought-indices}{%
+\subsubsection{Landcover trend and drought
+indices}\label{landcover-trend-and-drought-indices}}
+
+We calculated the surface occupied per landcover class into the five
+macrozones (``Norte Grande'' to ``Austral'') per year for 2001--2023.
+After that, we calculated the trend's change in surface; we used the
+Sen' slope \citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}.
+This way, we obtain a matrix of trends of 5 x 5 (macrozones x
+landcover). The aim is to later explore if the trend in landcover
+classes is associated with a trend in the drought indices. For this, we
+will use the techniques of regresion and regularization of Lasso
+\citep{Tibshirani2010} and Ridge \citep{Hoerl1970}. Also, we will test
+random forests for this purpose \citep{Ho1995}. We will choose the trend
+of landcover surface per macroclass and macrozone as the response
+variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI
+for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor
+variables. With this analysis, we expect to gather insights regarding
+whether there is a pattern of climatic influence along Chile or if what
+is happening in Central Chile has to do with more localized climatic
+conditions.
 
@@
 
+\hypertarget{mann-kendall-and-sens-slope}{%
+\subsubsection{Mann-Kendall and Sen's
+slope}\label{mann-kendall-and-sens-slope}}
+
 To estimate if there are significant positive or negative trends for the
@@
 \citep{Kendall1975}. To determine the magnitude of the trend, we used
-Sen's slope \citep{Sen1968}. One of the advantages of applying this
-methodology is that the Sen's slope is not affected by outliers as
-regular regression does. We applied both to the six time scales from
-1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and
-SSI. In the case of zcNDVI (six months) was for 2000 to 2023. Thus, we
-have 31 trends. Also, we extracted the trend aggregated by macrozone and
-landcover class, obtaining a table of 31x5x5 (drought indices trends x
-macrozone x landcover class). We will use this data in
+Sen's slope \citep{Sen1968}. Some of the advantages of applying this
+methodology are that the Sen's slope is not affected by outliers as
+regular regression does, and it is a non-paramteric method that is not
+affected by the distribution of the data. We applied both to the six
+time scales from 1981 to 2023 (monthly frequency) and the indices SPI,
+EDDI, SPEI, and SSI. In the case of zcNDVI (six months) was for 2000 to
+2023. Thus, we have 31 trends. Also, we extracted the trend aggregated
+by macrozone and landcover class, obtaining a table of 31x5x5 (drought
+indices trends x macrozone x landcover class). We will use this data in
 Section~\ref{sec-methods_lulc} to analyze if there is a strong
@@
 
-For zcNDVI, we want to remove the effect of the Andes's mountain and
-mask out trends that were at an elevation above 1500m because there is
-scarce vegetation.
+\hypertarget{trend-in-vegetation-productivity-without-landcover-change}{%
+\subsubsection{Trend in vegetation productivity without landcover
+change}\label{trend-in-vegetation-productivity-without-landcover-change}}
 
@@
 
+\hypertarget{data-1}{%
+\subsection{Data}\label{data-1}}
+
+\hypertarget{validation-of-era5l-variables-1}{%
+\subsubsection{Validation of ERA5L
+variables}\label{validation-of-era5l-variables-1}}
+
 \hypertarget{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}{%
@@
 
-Figure~\ref{fig-zcNDVI_var} and Table \ref{tab-landcoverTrend} show that
-the macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'',
-and ``Austral'' with 36\%, 31\%, and 34\%, respectively. The ``Norte
-Chico'' shows an increase in barrend land of 111 \(km^2 year^{-1}\) and
-a reduction in the class savanna -70 \(km^2 year^{-1}\). In the
-``Centro'' and ``Sur'' there are changes in the Chilean matorral, with
-an important reduction in savanna (-136 to -318 \(km^2 yr^{-1}\)), and
-an increase in shrubland and grassland. Showing a change for more dense
-vegetation types. It appears to be a change from the surface occupied by
-cropland from the ``Centro'' to the ``Sur.'' Also, there is a high
-increase in forest (397 \(km^2 yr^{-1}\)) in the ``Sur,'' replacing the
-savanna lost.
+\hypertarget{landcover-macroclasess-and-validation-1}{%
+\subsubsection{Landcover macroclasess and
+validation}\label{landcover-macroclasess-and-validation-1}}
+
+The validation of IGBP2013 and IGBP2014 with LCChile reached similar
+metrics of performance, having an accuracy of 0.82 and a F1 score of
+0.66 (see SS1).
+
+\hypertarget{landcover-persistence-mask-2001-2022-1}{%
+\subsubsection{Landcover persistence mask
+2001-2022}\label{landcover-persistence-mask-2001-2022-1}}
+
+The macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'',
+and ``Austral'' with 36\%, 31\%, and 34\%, respectively
+(Figure~\ref{fig-studyArea} and Table \ref{tab-landcoverTrend})
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/LC_pers80_per_macrozone.png}
+
+}
+
+\caption{\label{fig-LCprop}Proportion of land cover class from the
+persistent land cover for 2001-2022 (\textgreater80\%) per macrozone}
+
+\end{figure}
+
+Figure~\ref{fig-LCprop} shows the summary of the proportion of surface
+per landcover class and macrozone, derived from the mask over
+continental Chile.
+
+\hypertarget{landcover-trend-and-drought-indices-1}{%
+\subsubsection{Landcover trend and drought
+indices}\label{landcover-trend-and-drought-indices-1}}
 
@@
 \end{table}
+
+The ``Norte Chico'' shows an increase in barrend land of 111
+\(km^2 year^{-1}\) and a reduction in the class savanna -70
+\(km^2 year^{-1}\). In the ``Centro'' and ``Sur'' there are changes in
+the Chilean matorral, with an important reduction in savanna (-136 to
+-318 \(km^2 yr^{-1}\)), and an increase in shrubland and grassland.
+Showing a change for more dense vegetation types. It appears to be a
+change from the surface occupied by cropland from the ``Centro'' to the
+``Sur.'' Also, there is a high increase in forest (397 \(km^2 yr^{-1}\))
+in the ``Sur,'' replacing the savanna lost.
 
@@
 
-\hypertarget{validation-of-era5l-variables-1}{%
-\subsection{Validation of ERA5L
-variables}\label{validation-of-era5l-variables-1}}
-
 \hypertarget{discussion}{%

### 2024-01-19 scripts/28_mapa_koppen-geiger.R

@@ -3,23 +3,25 @@ library(terra)
 library(fs)
 library(tidyverse)
 
-dir <- '/mnt/data_procesada/data/rasters/Procesados/koppen_geiger_tif/1991_2020'
+dir <- '/mnt/data_procesada/data/rasters/Procesados/koppen_geiger_tif/'
 
 #/1991_2020'
 
 #/2041_2070/ssp585'
 
-files <- dir_ls(dir,regexp = '0p01')
+file1 <- paste0(dir,'1991_2020/koppen_geiger_0p01.tif')
+file2 <- paste0(dir,'2071_2099/ssp585/koppen_geiger_0p01.tif')
+
+kp <- c(rast(file1),rast(file2))
 
 chile <- read_sf('data/processed_data/spatial/chile_continental.gpkg')
-kp <- rast(files)
+
 kp_chile <- crop(kp,chile)
 kp_chile <- mask(kp,chile)
 kp_chile <- trim(kp_chile)
-writeRaster(kp_chile,'data/processed_data/spatial/koppen_geiger_1991_2020.tif')
+#writeRaster(kp_chile,'data/processed_data/spatial/koppen_geiger_1991_2020.tif')
 library(tmap)
 
-kp_chile <- rast('data/processed_data/spatial/koppen_geiger_1991_2020.tif')
 rgb <- coltab(kp_chile)[[1]][2:31,]
 paleta_kg <- read_csv2('data/processed_data/legend_koppen-geiger.csv')
 labs <- paleta_kg$type
@@ -27,9 +29,12 @@ names(labs) <- 1:30
 colores <- rgb(rgb[,2],rgb[,3],rgb[,4],rgb[,5],maxColor = 255)
 names(colores) <- 1:30
 
-tm_shape(kp_chile) +
+names(kp_chile) <- c('1991-2020','2071-2099 ssp585')
+
+map <- tm_shape(kp_chile) +
   tm_raster(palette = colores,
             style = 'cat',
             labels = labs,
             title = 'Climate (1991-2020)') +
   tm_layout(legend.outside = TRUE)
+tmap_save(map,'output/figs/mapa_koppen_presente_futuro.png',scale=1.2)

### 2024-01-22 article/paper_drought_Chile2023.qmd
actualizar repo

@@
 ---
-title: Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile
+title: Drought indices of water demand and supply, soil moisture, vegetation, and their impact on LULCC in continental Chile
 #subtitle: A Short Subtitle
@@
 abstract: |
-  Human-induced greenhouse gas emissions have increased the frequency and/or intensity of weather and climate extremes.
-Central Chile has been affected by a persistent drought which is impacting the hydrological system and vegetation development.
-The region has been the focus of research studies due to the diminishing water supply, this persistent period of water scarcity has been defined as a “mega drought".
-Nevertheless, our results evidence that the water deficit has expanded beyond.
-Our goal is to analyze the impact of drought, measured by drought indices of water supply/demand and vegetation status, in the LULCC (land use land cover change) over continental Chile.
-For the analysis, continental Chile was divided into five zones according to a latitudinal gradient: “Norte Grande”,  “Norte Chico”,  “Zona Central”, “Zona Sur”, and “Zona Austral”.
-We used monthly climatic re-analysis variables for precipitation, temperature and soil moisture for 1981-2023 from ERA5-Land (ERA5L); and MODIS (Moderate-Resolution Imaging Spectroradiometer) product MCD12Q1 for land cover for 2001-2021, and the NDVI vegetation index from product MOD13A2 collection 6.1 for 2000-2023, both from collection 6.1.
-We estimated atmospheric evaporative demand (AED) by combining the Hargreaves-Samani equation with the ERA5L temperature.
-We derived the drought indices SPI (Standardized Precipitation Index), SPEI (Standardized Precipitation Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), SSI (standardized anomaly of cumulative soil moisture), and the zcNDVI (standardized anomaly of cumulative NDVI).
-These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1, 3, and 6 months).
-We analyzed the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to have insights into the impact of water supply and demand on vegetation.
-Our results showed that LULCC had an increasing trend of 412 [km2yr−1] of forest expansion in the “Zona Sur”, together with a  decreasing trend of 24 [km2yr−1] of cropland contraction in the “Zona Central” meanwhile the “Zona Sur” showed an increase of 31 [km2yr−1], and a contraction of 80 [km2yr−1] of bare soil in the “Zona Austral”.
-The EDDI was the less correlated index for the five macro zones and the five types of land cover, showing that the temperature in Chile has little impact on vegetation.
-Higher r-squared values, between 0.5 and 0.8, were obtained at “Norte Chico” and “Zona Central” for the land cover types of savanna, shrubland, grassland, and croplands for the indices SPEI and SSI at time scales of 12 and 24 months.
-The forest type reaches a r-squared of ~0.5 for SSI of 12 months.
-The results indicate that the “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits longer than a year, potentially explained by a low capacity of water storage in those zones that should be further investigated.
+ Central Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development.
+This persistent period of water scarcity has been defined as a “Mega Drought".
+Our objective is to examine the effects of drought on LULCC (land use land cover change) over continental Chile using drought indices of water supply and demand, soil moisture, and vegetation productivity.
+For the analysis, continental Chile was divided into five zones according to a latitudinal gradient: “Norte Grande," “Norte Chico," "Centro," "Sur," and "Austral." The monthly ERA5-Land (ERA5L) variables for precipitation, temperature, and soil moisture were used.
+From 2001 to 2022, we used the land cover MODIS product MCD12Q1, and from 2000 to 2023, we used the NDVI (Normalized Difference Vegetation Index) product MOD13A3 collection 6.1.
+We estimated atmospheric evaporative demand (AED) using the Hargreaves-Samani equation with the ERA5L temperature.
+We used the Standardized Precipitation Index (SPI), the Standardized Precipitation Evapotranspiration Index (SPEI), the Evaporative Demand Drought Index (EDDI), the Standardized Soil Moisture Index (SSI), and the Standardized anomaly of cumulative NDVI (zcNDVI) as drought indicators.
+These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was for 6 months.
+We analyze the trend for LULCC, vegetation productivity, and drought indices.
+Also, we analyzed the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain insights into the impact of water supply and demand on vegetation productivity.
+Our results showed that LULCC were highest in "Centro," "Sur," and "Austral,” with 36%, 31%, and 34%, respectively.
+The EDDI shows that water demand has increased for all zones, with a major increase in "Norte Grande." The drought indices of water supply and soil moisture evidence a decreasing trend, which decreases at longer time scales, from "Norte Grande" to "Sur." "Austral" is the only zone that shows an increase in supply.
+Vegetation productivity measures by zcNDVI present a negative trend in "Norte Chico" and "Centro." Showing to be the zones most impacted by climatic conditions, the years 2020 and 2022 suffered the most extreme drought.
+On the other hand, forests seem to be the most resistant to drought.
+The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
+The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are SSI-12, followed by SPEI-24 and SPEI-12 in "Norte Chico" and "Centro." The results indicate that “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits longer than a year, potentially explained by a low capacity of water storage in those zones that should be further investigated.
 keywords: 
@@
             \newcommand{\elandscape}{\end{landscape}}
+  docx:
+    keep-tex: true
+    journal:
+      name: Journal Name
+      formatting: preprint
+      layout: onecolumn
+      model: 3p
+      cite-style: authoryear
+      natbiboptions: numbers
+    include-in-header: 
+      text: \usepackage{lineno}\linenumbers
+            \usepackage{multirow}
+            \usepackage{lscape}
+            \newcommand{\blandscape}{\begin{landscape}}
+            \newcommand{\elandscape}{\end{landscape}}
 ---
@@
 The sixth assessment report (AR6) of the IPCC [@IPCC2023] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes, and the evidence has been strengthened since AR5 [@IPCC2013].
-There is high confidence that increasing global warming can expand the land area affected by increasing drought frequency and severity [@IPCCCH112021].
+There is a high degree of confidence that rising temperatures will increase the land area where droughts will occur more frequently and with greater severity [@IPCCCH112021].
 Furthermore, drought increases tree mortality and triggers changes in land cover and, consequently, land use, thus impacting ecosystems [@Crausbay2017].
@@
 
-Precipitation is the primary driver of drought and is intensified by temperature [@Luo2017].
+The primary cause of drought is precipitation, and temperature makes it worse [@Luo2017].
 Drought impacts soil moisture, hydrological regimes, and vegetation productivity.
@@
 Lately, @Loon2016 and @AghaKouchak2021 have given an updated definition of drought for the Anthropocene, suggesting that it should be considered the feedback of humans' decisions and activities that drives the anthropogenic drought.
-Even though it has been argued that those definitions do not fully address the ecological dimensions of drought.
-@Crausbay2017 proposed the ecological drought definition as *"an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems"*.
+Even though it has been argued that those definitions do not fully address the ecological dimensions of drought, @Crausbay2017 proposed the ecological drought definition as "an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems.".
 Moreover, many ecological studies have misinterpreted how to characterize drought, for example, sometimes considering "dry" conditions as "drought" [@Slette2019].
-On the other hand, the AR6 [@IPCC2023] states that even if global warming is stabilized at 1.5°–2°C, many parts of the world will be impacted by more severe agricultural and ecological droughts.
+On the other hand, the AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C.
 Then, there is a challenge in conducting drought research, especially to evaluate its impact on ecosystems.
@@
 Reanalysis data, as the ERA5-Land (ERA5L) [@MunozSabater2021] provides hourly climatic information (precipitation, temperature, and soil moisture) without gaps since 1950 with global extension.
-ERA5L has already been used for drought assessment using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@Nouri2023] and for flash drought [@Wang2023] by analyzing soil mositure and evapotranspiration.
+ERA5L has already been used for drought assessment using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@Nouri2023] and for flash drought [@Wang2023] by analyzing soil moisture and evapotranspiration.
 On the other hand, satellite remote sensing [@West2019; @AghaKouchak2015] is the primary method to evaluate how drought impacts vegetation dynamics.
-Vegetation drought indices (VDI) are commonly used as proxies of productivity [@Paruelo2016; @Schucknecht2017], which can be derived from the MODIS (Moderate-Resolution Imaging Spectroradiometer).
+The Moderate-Resolution Imaging Spectroradiometer (MODIS) can be used to get vegetation drought indices (VDI), which are often used as proxies for productivity [@Paruelo2016; @Schucknecht2017].
 Besides, land use and land cover (LULC) change can be driven by drought [@Tran2019; @Akinyemi2021].
-To analyze these changes, multiple LULC products exist [@Grekousis2015], one of those that provides time series since 2001 is the MCD12Q1 [@Friedl2019] from MODIS.
+To analyze these changes, multiple LULC products exist [@Grekousis2015].
+One of those that provides time series since 2001 is the MCD12Q1 [@Friedl2019] from MODIS.
 The variation in water supply and demand is finally reflected in the total water storage (TWS).
-The TWS can be retrieved by the Gravity Recovery and Climate Experiment (GRACE), which allows analyzing water availability changes at coarse resolution [@Ahmed2014; @Ma2017].
-We can use climatic reanalysis (ERA5L) and vegetation data (MODIS) to derive drought indices of supply (i.e., precipitation) and demand (i.e., temperature) and thus evaluate the impact of drought on LULC changes.
+The Gravity Recovery and Climate Experiment (GRACE), which allows analyzing changes in water availability at coarse resolution, can retrieve the TWS [@Ahmed2014; @Ma2017].
+We can find drought indices of supply (i.e., precipitation) and demand (i.e., temperature) using climatic reanalysis (ERA5L) and vegetation data (MODIS).
+This lets us figure out how drought changes LULC.
 Further, the TWS can be assessed with regard to the changes in water supply and demand to gain insight into the impact on water storage.
@@
 
-In this research, we present the raster dataset DDS4Chl, which provides climate variables and drought indices of water demand and supply and vegetation productivity for continental Chile since 1981 at a monthly frequency.
-Those were gathered from the earth observation products ERA5L and MODIS.
-Then, we used DDS4Chl to analyze the impact of drought on different types of land cover classes in continental Chile.
-The specific objectives of the study are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of vegetation productivity (zcNDVI) with drought indices of water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.
+In this research, we aim to analyze the impact of drought on different types of ecosystems (land cover classes) in continental Chile.
+Our specific goals are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to assess the LULC change for 2001–2021 and how it relates to drought indices; iii) to evaluate the relationship between zcNDVI, a measure of vegetation productivity, and drought indices for water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.
 
@@
 The Andes Mountains are a main factor in latitudinal variation [@Garreaud2009].
-To describe the climate and ecosystem of Chile, we use the Koppen-Geiger release by @Beck2023 and the landcover type persistance of 80% of times for 2001–2022, from the IGBP classification scheme [@Friedl2019] (see @sec-methods_lulc).
+To describe the climate and ecosystem of Chile, we use the Koppen-Geiger release by @Beck2023 and the land cover type persistance of 80% of times for 2001–2022, from the IGBP classification scheme [@Friedl2019] (see @sec-methods_lulc).
 "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures.
@@
 In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
-In "Centro," there is a major proportion (~50%) of Chilean matorral (shrubland and savanna, [@Fuentes2021]), followed by grassland (~16%), forest (8%), and croplands (5%).
-The south part of "Sur" and the north part of "Austral" are dominated by an Oceanic climate (Cfb).
+There is a significant amount (50%) of Chilean matorral (shrubland and savanna, [@Fuentes2021]), then grassland (16%), forest (8%), and croplands (5%), in "Centro." The south part of "Sur" and the north part of "Austral" are dominated by an oceanic climate (Cfb).
 Those zones are high in forest and grassland.
@@
         \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
-         & MCD12Q1.061 & Landcover IGBP scheme & & 2001-2022 & ~ & Landcover \\ 
+         & MCD12Q1.061 & land cover IGBP scheme & & 2001-2022 & ~ & land cover \\ 
 \hline
@@
 
-### Landcover macroclasess and validation
+### land cover macroclasess and validation
 
@@
 The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
-Thus, we have a landcover raster time-series with the ten classes for 2001 and 2023.
+Thus, we have a land cover raster time-series with the ten classes for 2001 and 2023.
 
-To validate the landcover obtained, we compare the macroclasses with the ones of a more detailed landcover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile).
-The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses landcover.
-We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from landcover macrolcasses to validate with LCChile.
+To validate the land cover obtained, we compare the macroclasses with the ones of a more detailed land cover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile).
+The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses land cover.
+We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from land cover macrolcasses to validate with LCChile.
 
@@
 
-i) resampled LCChile to the spatial resolution (500m) of the landcover macroclasses using the nearest neighbor method,
+i) resampled LCChile to the spatial resolution (500m) of the land cover macroclasses using the nearest neighbor method,
 ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values as the other two years as prediction
@@
 
-### Landcover persistence mask 2001-2022
+### land cover persistence mask 2001-2022
 
 The time series of NDVI is affected by climatic conditions, vegetation development, seasonality, and changes in vegetation type.
-In this study, we want to analyze the variation in vegetation productivity in different landcover types and how it is affected by water demand, water supply, and soil moisture.
-In order to avoid changes due to a change in the landcover type, that will wrongly impact NDVI.
-We will develop a persistence mask for landcover for 2001–2023.
+In this study, we want to analyze the variation in vegetation productivity in different land cover types and how it is affected by water demand, water supply, and soil moisture.
+In order to avoid changes due to a change in the land cover type, that will wrongly impact NDVI.
+We will develop a persistence mask for land cover for 2001–2023.
 Thereby, we reduce an important source of variation on a regional scale.
 
-Thus, we calculated a raster mask for IGBP MODIS considering the macroclasses that remain without change for more than 80% of the years (2001–2022) per pixel, which allows us to identify the areas with no landcover change for the macroclasses.
+Thus, we calculated a raster mask for IGBP MODIS considering the macroclasses that remain without change for more than 80% of the years (2001–2022) per pixel, which allows us to identify the areas with no land cover change for the macroclasses.
 
-### Landcover trend and drought indices
+### land cover trend and drought indices
 
-We calculated the surface occupied per landcover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023.
+We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023.
 After that, we calculated the trend's change in surface; we used the Sen' slope [@Sen1968] based on Mann-Kendall [@Kendall1975].
-This way, we obtain a matrix of trends of 5 x 5 (macrozones x landcover).
-The aim is to later explore if the trend in landcover classes is associated with a trend in the drought indices.
+This way, we obtain a matrix of trends of 5 x 5 (macrozones x land cover).
+The aim is to later explore if the trend in land cover classes is associated with a trend in the drought indices.
 For this, we will use the techniques of regresion and regularization of Lasso [@Tibshirani2010] and Ridge [@Hoerl1970].
 Also, we will test random forests for this purpose [@Ho1995].
-We will choose the trend of landcover surface per macroclass and macrozone as the response variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor variables.
+We will choose the trend of land cover surface per macroclass and macrozone as the response variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor variables.
 With this analysis, we expect to gather insights regarding whether there is a pattern of climatic influence along Chile or if what is happening in Central Chile has to do with more localized climatic conditions.
@@
 Thus, we have 31 trends.
-Also, we extracted the trend aggregated by macrozone and landcover class, obtaining a table of 31x5x5 (drought indices trends x macrozone x landcover class).
+Also, we extracted the trend aggregated by macrozone and land cover class, obtaining a table of 31x5x5 (drought indices trends x macrozone x land cover class).
 We will use this data in @sec-methods_lulc to analyze if there is a strong relationship between the trends of drought indices and land cover surface within continental Chile.
 
-### Trend in vegetation productivity without landcover change
+### Trend in vegetation productivity without land cover change
 
@@
 We will contrast this hypothesis with what is occurring in Chile.
-To achieve this, we will use the landcover class type that remains more than 80% of types for 2001–2022 to evaluate the trend on zcNDVI and use this as a mask where there are low changes.
+To achieve this, we will use the land cover class type that remains more than 80% of types for 2001–2022 to evaluate the trend on zcNDVI and use this as a mask where there are low changes.
 
-##	Impact for water supply and demand, and soil moisture in vegetation productivity within landcover types
+##	Impact for water supply and demand, and soil moisture in vegetation productivity within land cover types
 
-The aim of this section is to analyze the drought indices of water demand and supply and soil moisture against vegetation to address: i) if short- or long-term time scales are most important in impacting vegetation through Chile; and ii) the strength of the correlation for the variable and the time scale.
-Then, we will summarize for each landcover class and macrozone.
-Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses: forest, cropland, grassland, savanna, and shrubland.
+We analyze the drought indices of water demand and supply and soil moisture against vegetation to address: i) if short- or long-term time scales are most important in impacting vegetation through Chile; and ii) the strength of the correlation for the variable and the time scale.
+Then, we will summarize for each land cover class and macrozone.
+Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses having vehetation: forest, cropland, grassland, savanna, and shrubland.
 
@@
 
-### Landcover macroclasess and validation
+### land cover macroclasess and validation
 
-The validation of IGBP2013 and IGBP2014 with LCChile reached similar metrics of performance, having an accuracy of 0.82 and a F1 score of 0.66 (see SS1).
+For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands.
+@fig-studyArea c shows the spatial distribution of the macroclasses through Chile for the year 2022.
+The validation of IGBP2013 and IGBP2014 with LCChile reached near the same metrics of performance, having an accuracy of ~0.82 and a F1 score of ~0.66 (see SS1).
 
-### Landcover persistence mask 2001-2022
+### land cover persistence mask 2001-2022
 
-The macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend})
+@fig-studyArea d, shows the macroclasses of land cover persistance (80%) during 2021-2022, respectively.
+Within continental Chile, forest is the vegetation type with highest surface with 135,00 $km^2$, followed by grassland (73,176 $km^2$), savanna (54,410 $km^2$), shrubland (24,959 $km^2$), and cropland (3,100 $km^2$) (\ref{tab-land coverSurf}).
+The macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively (@fig-studyArea and Table \ref{tab-land coverTrend}); of its surface that changes the type of land cover.
+@fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.
+
+```{=latex}
+\begin{table}[!ht]
+\caption{Surface of the land cover class that persist during 2001-2022}
+\label{tab-land coverSurf}
+\includegraphics[width = .5\textwidth]{../output/figs/table_surface_landcover_macrozone.png}
+\end{table}
+```
 
@@
 
-@fig-LCprop shows the summary of the proportion of surface per landcover class and macrozone, derived from the mask over continental Chile.
-
-### Landcover trend and drought indices
+### land cover trend and drought indices
 
 \begin{table}[!ht]
-\caption{The value of Sen's slope trend next to the time-series plot of surface per landcover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile.
+\caption{The value of Sen's slope trend next to the time-series plot of surface per land cover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile.
 Values of zero indicate that there was not a significant trend.
 Red dots on the plots indicate the maximum and minimum values of surface.}
-\label{tab-landcoverTrend}
+\label{tab-land coverTrend}
 \includegraphics[]{../output/figs/table_var_landcover_macro.png}
@@
 
-The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna -70 $km^2 year^{-1}$.
-In the "Centro" and "Sur" there are changes in the Chilean matorral, with an important reduction in savanna (-136 to -318 $km^2 yr^{-1}$), and an increase in shrubland and grassland.
+The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna of 70 $km^2 year^{-1}$.
+In the "Centro" and "Sur," there are changes in the Chilean matorral, with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$), and an increase in shrubland and grassland.
 Showing a change for more dense vegetation types.
-It appears to be a change from the surface occupied by cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.
+It appears to be a shift in the area of cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.
+
+Further, we want to address whether the trend in land cover change for 2001–2023 is associated with trends in drought indices of water demand and supply and/or soil moisture for macrozone and land cover macroclasses.
+From the three methods tested, Ridge, Lasso, and Random Forest, neither gives significant results regarding whether the trend in a drought index for any time scale explains the trend in land cover change.
+Nevertheless, in "Norte Chico" and "Centro," there is a decrease in croplands and savanna and an increase in barren land, which is associated with the variation in drought indices.
+Mainly for a decrease in water supply (SPI and SSI) and an increase in water demand (EDDI).
+However, due to the high variability from north to south in Chile, the climatic condition (arid, semi-arid, and humid), and the land cover type, we believe that only in those zones could the LULCC be driven to some degree by drought.
 
 ##	Trend of drought indices for water demand and supply, soil moisture, and vegetation productivity
-
-Regarding vegetation productivity aggregated through the macrozones in the five landcover macroclasses.
-In "Norte Grande" there is a increase trend of 0.02 (z-index) per decade, related with types of grassland and shrubland.
-There is a negative trend in "Norte Chico" with -0.04  and "Centro" with -0.02 per decade.
-In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04.
-In "Centro," shrubland reaches -0.06, followed by grassland with -0.05, and croplands and savanna have ~-0.01 per decade.
-This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023].
-Vegetation reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Norte Chico" and Centro" (Mega Drought).
-The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var).
-The area most affected by a negative trend in vegetation seems to be associated with the Chilean matorral [@Fuentes2021], despite the croplands being as severely impacted by drought as native vegetation does in "Norte Chico.".
 
@@
 Greener colors indicate a positive trend; reder colors correspond to a negative trend and a decrease in vegetation productivity.
-Grey colors indicate either no vegetation or a change in landcover type for 2001–2022.
+Grey colors indicate either no vegetation or a change in land cover type for 2001–2022.
 (b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile.
@@
 knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png")
+```
+
+Regarding vegetation productivity aggregated through the macrozones in the five land cover macroclasses, in "Norte Grande," there is an increase trend of 0.02 (z-index) per decade, related to types of grassland and shrubland.
+There is a negative trend in "Norte Chico" with -0.04 and "Centro" with -0.02 per decade.
+In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04.
+In "Centro," shrubland reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per decade.
+This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023].
+Vegetation reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Norte Chico" and Centro" (Mega Drought).
+The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var).
+Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the Chilean matorral appears to be the region most affected by a negative trend in vegetation [@Fuentes2021].
+
+```{r fig-trendDIMacro}
+#| echo: false
+#| fig-pos: "!ht"
+#| fig-cap: "Trend per decade for the drought indices SPI, EDDI, SPEI, and SSI aggregated by macrozone."
+#| out-width: "\\textwidth"
+knitr::include_graphics('../output/figs/trend_macrozone_drought_indices.png')
 ```
@@
 Because of this, the SPEI (which uses AED) reached its lowest value in "Norte Grande," with -0.03 at 36 months.
-Despite the other macrozones, "Austral" showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend (@fig-trendDI).
+Despite the other macrozones, "Austral" showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend (@fig-trendDI and @fig-trendDIMacro).
 
@@
 
-According to what is shown in @fig-corTimeScale, @fig-corPerson, and Table \ref{tab-corLandcover}, forest seems to be the most resistant type to drought.
+According to what is shown in @fig-corTimeScale, @fig-corPerson, and Table \ref{tab-corland cover}, forest seems to be the most resistant type to drought.
 Showing that only "Centro" is slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit (SSI-12).
-Grasslands are shown to be impacted by a SSI-12 with a rsq = 0.45 and by a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34, respectively) in the "Norte Chico" and to a lesser degree in the "Norte Grande." But, in further south macrozones, this type was not affected by water demand, water supply, or soil moisture.
+In the "Norte Chico" and to a lesser extent in the "Norte Grande," it is evident that a SSI-12 with a rsq = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34, respectively) have an impact on grasslands.
+However, this type was unaffected by soil moisture, water supply, or demand in macrozones further south.
 The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
@@
 This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (rsq = 0.45 in "Centro" and 0.2 in "Sur").
-In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45% and 66% of "Norte Chico." The most affected type by climatic variation was shrubland, where in "Norte Chico" and "Centro" soil moisture explained 59% and precipitation, 37%, with SSI-12 being the most relevant variable, followed by SPI-36 in "Norte Chico" and SPI-24 in "Sur."
+In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45% and 66% of "Norte Chico." The type of land most impacted by climatic variation was shrubland, where soil moisture explained 59% and precipitation, 37%, in "Norte Chico" and "Centro," with SSI-12 being the most relevant variable, then SPI-36 in "Norte Chico" and SPI-24 in "Sur."
 
@@
 \begin{table}[!ht]
-\caption{Summarry per landcover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
-The numbers in each cell indicate the time scale that reached the maximum correlation for the landcover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
-\label{tab-corLandcover}
+\caption{Summarry per land cover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
+The numbers in each cell indicate the time scale that reached the maximum correlation for the land cover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
+\label{tab-corland cover}
 \includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
@@
 
-1.-  Respecto a lo que indica @Vicente-Serrano2018, de que el aumento en la tendencia en severidad de la sequía (hidrológica) tiene que ver más con un aumento de la demanda de agua (LULCC) que a una tendencia en las condiciones climáticas (SPI-12).
+## Ecological and agricultural drought
 
-2.- Sobre los tipos de landcover más afectados por los indicadores de sequía. Asociación con el matorral chileno [@Fuentes2021]. Diferencia entre el Norte Chico y Centro y lo que pasa hacía el sur.
+## Drought trend, LULCC, and climate conditions
 
-3. Como podrian servir estos resultados para desarrollar un predictor de sequía basado.
+1.-  Respecto a lo que indica @Vicente-Serrano2018, de que el aumento en la tendencia en severidad de la sequía (hidrológica) tiene que ver más con un aumento de la demanda de agua (ej, LULCC, amazonas) que a una tendencia en las condiciones climáticas (SPI-12).
+¿Qué pasa en Chile?
 
-4.- Qué se podria hacer mejor en futuras investigaciones del tema.
+## land cover types most impacted by drought throughout Chile
 
-5.- Sequía, aridez, water scarcity [@VanLoon2013] Si puede utilizar el mapa y proyección climática de Koppen_Geiger [@Beck2023] para dirigir esta discusión
+2.- Sobre los tipos de land cover más afectados por los indicadores de sequía. Asociación con el matorral chileno [@Fuentes2021]. Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur.
+
+## Drought indices of water demand and supply, soil moisture to predict changes in vegetation productivity
+
+3. Como podrían servir estos resultados para desarrollar o mejorar un predictor de productividad de la vegetación.
+  - Los datos ERA5L están casi en tiempo real, 7 dias; MODIS también.
+  - EL SSI se ve como un poderoso indicador que explica la variabilidad en la productividad de la vegetación.
+
+## Future outlook
+
+4.- Qué se podría hacer mejor en futuras investigaciones del tema.
+  - mejorar la resolución y calidad de los datos climáticos
+  -
 

### 2024-01-22 article/paper_drought_Chile2023.tex
actualizar repo

@@
 \hypersetup{
-  pdftitle={Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile},
+  pdftitle={Drought indices of water demand and supply, soil moisture, vegetation, and their impact on LULCC in continental Chile},
   pdfauthor={Francisco Zambrano},
@@
 \title{Drought indices of water demand and supply, soil moisture,
-vegetation, and its impact on LULCC in continental Chile}
+vegetation, and their impact on LULCC in continental Chile}
 \author[1]{Francisco Zambrano%
@@
 \begin{abstract}
-Human-induced greenhouse gas emissions have increased the frequency
-and/or intensity of weather and climate extremes. Central Chile has been
-affected by a persistent drought which is impacting the hydrological
-system and vegetation development. The region has been the focus of
-research studies due to the diminishing water supply, this persistent
-period of water scarcity has been defined as a ``mega drought''.
-Nevertheless, our results evidence that the water deficit has expanded
-beyond. Our goal is to analyze the impact of drought, measured by
-drought indices of water supply/demand and vegetation status, in the
-LULCC (land use land cover change) over continental Chile. For the
-analysis, continental Chile was divided into five zones according to a
-latitudinal gradient: ``Norte Grande'', ``Norte Chico'', ``Zona
-Central'', ``Zona Sur'', and ``Zona Austral''. We used monthly climatic
-re-analysis variables for precipitation, temperature and soil moisture
-for 1981-2023 from ERA5-Land (ERA5L); and MODIS (Moderate-Resolution
-Imaging Spectroradiometer) product MCD12Q1 for land cover for 2001-2021,
-and the NDVI vegetation index from product MOD13A2 collection 6.1 for
-2000-2023, both from collection 6.1. We estimated atmospheric
-evaporative demand (AED) by combining the Hargreaves-Samani equation
-with the ERA5L temperature. We derived the drought indices SPI
-(Standardized Precipitation Index), SPEI (Standardized Precipitation
-Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), SSI
-(standardized anomaly of cumulative soil moisture), and the zcNDVI
-(standardized anomaly of cumulative NDVI). These indices were calculated
-for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1,
-3, and 6 months). We analyzed the temporal correlation of SPI, SPEI,
-EDDI, and SSI with zcNDVI to have insights into the impact of water
-supply and demand on vegetation. Our results showed that LULCC had an
-increasing trend of 412 {[}km2yr−1{]} of forest expansion in the ``Zona
-Sur'', together with a decreasing trend of 24 {[}km2yr−1{]} of cropland
-contraction in the ``Zona Central'' meanwhile the ``Zona Sur'' showed an
-increase of 31 {[}km2yr−1{]}, and a contraction of 80 {[}km2yr−1{]} of
-bare soil in the ``Zona Austral''. The EDDI was the less correlated
-index for the five macro zones and the five types of land cover, showing
-that the temperature in Chile has little impact on vegetation. Higher
-r-squared values, between 0.5 and 0.8, were obtained at ``Norte Chico''
-and ``Zona Central'' for the land cover types of savanna, shrubland,
-grassland, and croplands for the indices SPEI and SSI at time scales of
-12 and 24 months. The forest type reaches a r-squared of
-\textasciitilde0.5 for SSI of 12 months. The results indicate that the
-``Norte Chico'' and ``Zona Central'' are the most sensitive regions to
-water supply deficits longer than a year, potentially explained by a low
-capacity of water storage in those zones that should be further
-investigated.
+Central Chile has been the focus of research studies due to the
+persistent decrease in water supply, which is impacting the hydrological
+system and vegetation development. This persistent period of water
+scarcity has been defined as a ``Mega Drought''. Our objective is to
+examine the effects of drought on LULCC (land use land cover change)
+over continental Chile using drought indices of water supply and demand,
+soil moisture, and vegetation productivity. For the analysis,
+continental Chile was divided into five zones according to a latitudinal
+gradient: ``Norte Grande,'' ``Norte Chico,'' ``Centro,'' ``Sur,'' and
+``Austral.'' The monthly ERA5-Land (ERA5L) variables for precipitation,
+temperature, and soil moisture were used. From 2001 to 2022, we used the
+land cover MODIS product MCD12Q1, and from 2000 to 2023, we used the
+NDVI (Normalized Difference Vegetation Index) product MOD13A3 collection
+6.1. We estimated atmospheric evaporative demand (AED) using the
+Hargreaves-Samani equation with the ERA5L temperature. We used the
+Standardized Precipitation Index (SPI), the Standardized Precipitation
+Evapotranspiration Index (SPEI), the Evaporative Demand Drought Index
+(EDDI), the Standardized Soil Moisture Index (SSI), and the Standardized
+anomaly of cumulative NDVI (zcNDVI) as drought indicators. These indices
+were calculated for time scales of 1, 3, 6, 12, 24, and 36 months,
+except for zcNDVI, which was for 6 months. We analyze the trend for
+LULCC, vegetation productivity, and drought indices. Also, we analyzed
+the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain
+insights into the impact of water supply and demand on vegetation
+productivity. Our results showed that LULCC were highest in ``Centro,''
+``Sur,'' and ``Austral,'' with 36\%, 31\%, and 34\%, respectively. The
+EDDI shows that water demand has increased for all zones, with a major
+increase in ``Norte Grande.'' The drought indices of water supply and
+soil moisture evidence a decreasing trend, which decreases at longer
+time scales, from ``Norte Grande'' to ``Sur.'' ``Austral'' is the only
+zone that shows an increase in supply. Vegetation productivity measures
+by zcNDVI present a negative trend in ``Norte Chico'' and ``Centro.''
+Showing to be the zones most impacted by climatic conditions, the years
+2020 and 2022 suffered the most extreme drought. On the other hand,
+forests seem to be the most resistant to drought. The types that show to
+be most affected by variation in climate conditions are shrublands,
+savannas, and croplands. The drought indices that have the capability of
+explaining to a major degree the variance in vegetation productivity are
+SSI-12, followed by SPEI-24 and SPEI-12 in ``Norte Chico'' and
+``Centro.'' The results indicate that ``Norte Chico'' and ``Zona
+Central'' are the most sensitive regions to water supply deficits longer
+than a year, potentially explained by a low capacity of water storage in
+those zones that should be further investigated.
 \end{abstract}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners, frame hidden, interior hidden, enhanced]}{\end{tcolorbox}}\fi
 
@@
 and/or intensity of some weather and climate extremes, and the evidence
-has been strengthened since AR5 \citep{IPCC2013}. There is high
-confidence that increasing global warming can expand the land area
-affected by increasing drought frequency and severity
+has been strengthened since AR5 \citep{IPCC2013}. There is a high degree
+of confidence that rising temperatures will increase the land area where
+droughts will occur more frequently and with greater severity
 \citep{IPCCCH112021}. Furthermore, drought increases tree mortality and
@@
 
-Precipitation is the primary driver of drought and is intensified by
-temperature \citep{Luo2017}. Drought impacts soil moisture, hydrological
+The primary cause of drought is precipitation, and temperature makes it
+worse \citep{Luo2017}. Drought impacts soil moisture, hydrological
 regimes, and vegetation productivity. Initially, drought was commonly
@@
 drought. Even though it has been argued that those definitions do not
-fully address the ecological dimensions of drought. \citet{Crausbay2017}
-proposed the ecological drought definition as \emph{``an episodic
-deficit in water availability that drives ecosystems beyond thresholds
-of vulnerability, impacts ecosystem services, and triggers feedback in
-natural and/or human systems''}. Moreover, many ecological studies have
+fully address the ecological dimensions of drought, \citet{Crausbay2017}
+proposed the ecological drought definition as ``an episodic deficit in
+water availability that drives ecosystems beyond thresholds of
+vulnerability, impacts ecosystem services, and triggers feedback in
+natural and/or human systems.''. Moreover, many ecological studies have
 misinterpreted how to characterize drought, for example, sometimes
 considering ``dry'' conditions as ``drought'' \citep{Slette2019}. On the
-other hand, the AR6 \citep{IPCC2023} states that even if global warming
-is stabilized at 1.5°--2°C, many parts of the world will be impacted by
-more severe agricultural and ecological droughts. Then, there is a
+other hand, the AR6 \citep{IPCC2023} predicts that many regions of the
+world will experience more severe agricultural and ecological droughts
+even if global warming stabilizes at 1.5°--2°C. Then, there is a
 challenge in conducting drought research, especially to evaluate its
@@
 (SPEI) \citep{Nouri2023} and for flash drought \citep{Wang2023} by
-analyzing soil mositure and evapotranspiration. On the other hand,
+analyzing soil moisture and evapotranspiration. On the other hand,
 satellite remote sensing \citep{West2019, AghaKouchak2015} is the
-primary method to evaluate how drought impacts vegetation dynamics.
-Vegetation drought indices (VDI) are commonly used as proxies of
-productivity \citep{Paruelo2016, Schucknecht2017}, which can be derived
-from the MODIS (Moderate-Resolution Imaging Spectroradiometer). Besides,
-land use and land cover (LULC) change can be driven by drought
+primary method to evaluate how drought impacts vegetation dynamics. The
+Moderate-Resolution Imaging Spectroradiometer (MODIS) can be used to get
+vegetation drought indices (VDI), which are often used as proxies for
+productivity \citep{Paruelo2016, Schucknecht2017}. Besides, land use and
+land cover (LULC) change can be driven by drought
 \citep{Tran2019, Akinyemi2021}. To analyze these changes, multiple LULC
-products exist \citep{Grekousis2015}, one of those that provides time
+products exist \citep{Grekousis2015}. One of those that provides time
 series since 2001 is the MCD12Q1 \citep{Friedl2019} from MODIS. The
 variation in water supply and demand is finally reflected in the total
-water storage (TWS). The TWS can be retrieved by the Gravity Recovery
-and Climate Experiment (GRACE), which allows analyzing water
-availability changes at coarse resolution \citep{Ahmed2014, Ma2017}. We
-can use climatic reanalysis (ERA5L) and vegetation data (MODIS) to
-derive drought indices of supply (i.e., precipitation) and demand (i.e.,
-temperature) and thus evaluate the impact of drought on LULC changes.
-Further, the TWS can be assessed with regard to the changes in water
-supply and demand to gain insight into the impact on water storage.
+water storage (TWS). The Gravity Recovery and Climate Experiment
+(GRACE), which allows analyzing changes in water availability at coarse
+resolution, can retrieve the TWS \citep{Ahmed2014, Ma2017}. We can find
+drought indices of supply (i.e., precipitation) and demand (i.e.,
+temperature) using climatic reanalysis (ERA5L) and vegetation data
+(MODIS). This lets us figure out how drought changes LULC. Further, the
+TWS can be assessed with regard to the changes in water supply and
+demand to gain insight into the impact on water storage.
 
@@
 
-In this research, we present the raster dataset DDS4Chl, which provides
-climate variables and drought indices of water demand and supply and
-vegetation productivity for continental Chile since 1981 at a monthly
-frequency. Those were gathered from the earth observation products ERA5L
-and MODIS. Then, we used DDS4Chl to analyze the impact of drought on
-different types of land cover classes in continental Chile. The specific
-objectives of the study are: i) to analyze the trend on multi-scalar
-drought indices for water demand and supply, soil moisture, and
-vegetation productivity for 1981--2023/2001--2023; ii) to evaluate LULC
-change for 2001--2021 and its relation to drought indices; iii) to
-analyze the relationship of a proxy of vegetation productivity (zcNDVI)
-with drought indices of water demand and supply and soil moisture; and
+In this research, we aim to analyze the impact of drought on different
+types of ecosystems (land cover classes) in continental Chile. Our
+specific goals are: i) to analyze the trend on multi-scalar drought
+indices for water demand and supply, soil moisture, and vegetation
+productivity for 1981--2023/2001--2023; ii) to assess the LULC change
+for 2001--2021 and how it relates to drought indices; iii) to evaluate
+the relationship between zcNDVI, a measure of vegetation productivity,
+and drought indices for water demand and supply and soil moisture; and
 iv) to assess if the observed changes in the drought indices are linked
@@
 \citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
-use the Koppen-Geiger release by \citet{Beck2023} and the landcover type
-persistance of 80\% of times for 2001--2022, from the IGBP
+use the Koppen-Geiger release by \citet{Beck2023} and the land cover
+type persistance of 80\% of times for 2001--2022, from the IGBP
 classification scheme \citep{Friedl2019} (see
@@
 main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
-In ``Centro,'' there is a major proportion (\textasciitilde50\%) of
-Chilean matorral (shrubland and savanna, \citep{Fuentes2021}), followed
-by grassland (\textasciitilde16\%), forest (8\%), and croplands (5\%).
-The south part of ``Sur'' and the north part of ``Austral'' are
-dominated by an Oceanic climate (Cfb). Those zones are high in forest
-and grassland. The southern part of the country has a tundra climate,
-and in Patagonia, it is a cold semi-arid area with an extended surface
-of grassland, forest, and, to a lesser extent, savanna.
+There is a significant amount (50\%) of Chilean matorral (shrubland and
+savanna, \citep{Fuentes2021}), then grassland (16\%), forest (8\%), and
+croplands (5\%), in ``Centro.'' The south part of ``Sur'' and the north
+part of ``Austral'' are dominated by an oceanic climate (Cfb). Those
+zones are high in forest and grassland. The southern part of the country
+has a tundra climate, and in Patagonia, it is a cold semi-arid area with
+an extended surface of grassland, forest, and, to a lesser extent,
+savanna.
 
@@
 
-\hypertarget{landcover-macroclasess-and-validation}{%
-\subsubsection{Landcover macroclasess and
-validation}\label{landcover-macroclasess-and-validation}}
+\hypertarget{land-cover-macroclasess-and-validation}{%
+\subsubsection{land cover macroclasess and
+validation}\label{land-cover-macroclasess-and-validation}}
 
@@
 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
-Thus, we have a landcover raster time-series with the ten classes for
+Thus, we have a land cover raster time-series with the ten classes for
 2001 and 2023.
 
-To validate the landcover obtained, we compare the macroclasses with the
-ones of a more detailed landcover map made by \citet{Zhao2016} for Chile
-with samples acquired in the years 2013--2014 (LCChile). The later has a
-spatial resolution of 30 m and three levels of defined classes; from
-those, we used level 1, which fits with the macroclasses landcover. We
-chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from landcover
-macrolcasses to validate with LCChile.
+To validate the land cover obtained, we compare the macroclasses with
+the ones of a more detailed land cover map made by \citet{Zhao2016} for
+Chile with samples acquired in the years 2013--2014 (LCChile). The later
+has a spatial resolution of 30 m and three levels of defined classes;
+from those, we used level 1, which fits with the macroclasses land
+cover. We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from land
+cover macrolcasses to validate with LCChile.
 
@@
 \item
-  resampled LCChile to the spatial resolution (500m) of the landcover
+  resampled LCChile to the spatial resolution (500m) of the land cover
   macroclasses using the nearest neighbor method,
@@
 
-\hypertarget{landcover-persistence-mask-2001-2022}{%
-\subsubsection{Landcover persistence mask
-2001-2022}\label{landcover-persistence-mask-2001-2022}}
+\hypertarget{land-cover-persistence-mask-2001-2022}{%
+\subsubsection{land cover persistence mask
+2001-2022}\label{land-cover-persistence-mask-2001-2022}}
 
@@
 we want to analyze the variation in vegetation productivity in different
-landcover types and how it is affected by water demand, water supply,
-and soil moisture. In order to avoid changes due to a change in the
-landcover type, that will wrongly impact NDVI. We will develop a
-persistence mask for landcover for 2001--2023. Thereby, we reduce an
-important source of variation on a regional scale.
+land cover types and how it is affected by water demand, water supply,
+and soil moisture. In order to avoid changes due to a change in the land
+cover type, that will wrongly impact NDVI. We will develop a persistence
+mask for land cover for 2001--2023. Thereby, we reduce an important
+source of variation on a regional scale.
 
@@
 (2001--2022) per pixel, which allows us to identify the areas with no
-landcover change for the macroclasses.
-
-\hypertarget{landcover-trend-and-drought-indices}{%
-\subsubsection{Landcover trend and drought
-indices}\label{landcover-trend-and-drought-indices}}
-
-We calculated the surface occupied per landcover class into the five
+land cover change for the macroclasses.
+
+\hypertarget{land-cover-trend-and-drought-indices}{%
+\subsubsection{land cover trend and drought
+indices}\label{land-cover-trend-and-drought-indices}}
+
+We calculated the surface occupied per land cover class into the five
 macrozones (``Norte Grande'' to ``Austral'') per year for 2001--2023.
@@
 Sen' slope \citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}.
-This way, we obtain a matrix of trends of 5 x 5 (macrozones x
-landcover). The aim is to later explore if the trend in landcover
-classes is associated with a trend in the drought indices. For this, we
-will use the techniques of regresion and regularization of Lasso
+This way, we obtain a matrix of trends of 5 x 5 (macrozones x land
+cover). The aim is to later explore if the trend in land cover classes
+is associated with a trend in the drought indices. For this, we will use
+the techniques of regresion and regularization of Lasso
 \citep{Tibshirani2010} and Ridge \citep{Hoerl1970}. Also, we will test
 random forests for this purpose \citep{Ho1995}. We will choose the trend
-of landcover surface per macroclass and macrozone as the response
+of land cover surface per macroclass and macrozone as the response
 variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI
@@
 2023. Thus, we have 31 trends. Also, we extracted the trend aggregated
-by macrozone and landcover class, obtaining a table of 31x5x5 (drought
-indices trends x macrozone x landcover class). We will use this data in
+by macrozone and land cover class, obtaining a table of 31x5x5 (drought
+indices trends x macrozone x land cover class). We will use this data in
 Section~\ref{sec-methods_lulc} to analyze if there is a strong
@@
 
-\hypertarget{trend-in-vegetation-productivity-without-landcover-change}{%
-\subsubsection{Trend in vegetation productivity without landcover
-change}\label{trend-in-vegetation-productivity-without-landcover-change}}
+\hypertarget{trend-in-vegetation-productivity-without-land-cover-change}{%
+\subsubsection{Trend in vegetation productivity without land cover
+change}\label{trend-in-vegetation-productivity-without-land-cover-change}}
 
@@
 hypothesis with what is occurring in Chile. To achieve this, we will use
-the landcover class type that remains more than 80\% of types for
+the land cover class type that remains more than 80\% of types for
 2001--2022 to evaluate the trend on zcNDVI and use this as a mask where
@@
 
-\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-landcover-types}{%
+\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-land-cover-types}{%
 \subsection{Impact for water supply and demand, and soil moisture in
-vegetation productivity within landcover
-types}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-landcover-types}}
-
-The aim of this section is to analyze the drought indices of water
-demand and supply and soil moisture against vegetation to address: i) if
-short- or long-term time scales are most important in impacting
-vegetation through Chile; and ii) the strength of the correlation for
-the variable and the time scale. Then, we will summarize for each
-landcover class and macrozone. Thus, we will be able to advance in
-understanding how climate is affecting vegetation, considering the
-impact on the five macroclasses: forest, cropland, grassland, savanna,
-and shrubland.
+vegetation productivity within land cover
+types}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-land-cover-types}}
+
+We analyze the drought indices of water demand and supply and soil
+moisture against vegetation to address: i) if short- or long-term time
+scales are most important in impacting vegetation through Chile; and ii)
+the strength of the correlation for the variable and the time scale.
+Then, we will summarize for each land cover class and macrozone. Thus,
+we will be able to advance in understanding how climate is affecting
+vegetation, considering the impact on the five macroclasses having
+vehetation: forest, cropland, grassland, savanna, and shrubland.
 
@@
 
-\hypertarget{landcover-macroclasess-and-validation-1}{%
-\subsubsection{Landcover macroclasess and
-validation}\label{landcover-macroclasess-and-validation-1}}
-
-The validation of IGBP2013 and IGBP2014 with LCChile reached similar
-metrics of performance, having an accuracy of 0.82 and a F1 score of
-0.66 (see SS1).
-
-\hypertarget{landcover-persistence-mask-2001-2022-1}{%
-\subsubsection{Landcover persistence mask
-2001-2022}\label{landcover-persistence-mask-2001-2022-1}}
-
-The macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'',
-and ``Austral'' with 36\%, 31\%, and 34\%, respectively
-(Figure~\ref{fig-studyArea} and Table \ref{tab-landcoverTrend})
+\hypertarget{land-cover-macroclasess-and-validation-1}{%
+\subsubsection{land cover macroclasess and
+validation}\label{land-cover-macroclasess-and-validation-1}}
+
+For vegetation, we obtained and use hereafter five macroclasses of land
+cover from IGBP MODIS: forest, shrubland, savanna, grassland, and
+croplands. Figure~\ref{fig-studyArea} c shows the spatial distribution
+of the macroclasses through Chile for the year 2022. The validation of
+IGBP2013 and IGBP2014 with LCChile reached near the same metrics of
+performance, having an accuracy of \textasciitilde0.82 and a F1 score of
+\textasciitilde0.66 (see SS1).
+
+\hypertarget{land-cover-persistence-mask-2001-2022-1}{%
+\subsubsection{land cover persistence mask
+2001-2022}\label{land-cover-persistence-mask-2001-2022-1}}
+
+Figure~\ref{fig-studyArea} d, shows the macroclasses of land cover
+persistance (80\%) during 2021-2022, respectively. Within continental
+Chile, forest is the vegetation type with highest surface with 135,00
+\(km^2\), followed by grassland (73,176 \(km^2\)), savanna (54,410
+\(km^2\)), shrubland (24,959 \(km^2\)), and cropland (3,100 \(km^2\))
+(\ref{tab-land coverSurf}). The macrozones with major LULCC for
+2001-2022 were ``Centro'', ``Sur'', and ``Austral'' with 36\%, 31\%, and
+34\%, respectively (Figure~\ref{fig-studyArea} and Table
+\ref{tab-land coverTrend}); of its surface that changes the type of land
+cover. Figure~\ref{fig-LCprop} shows the summary of the proportion of
+surface per land cover class and macrozone, derived from the persistance
+mask over continental Chile.
+
+\begin{table}[!ht]
+\caption{Surface of the land cover class that persist during 2001-2022}
+\label{tab-land coverSurf}
+\includegraphics[width = .5\textwidth]{../output/figs/table_surface_landcover_macrozone.png}
+\end{table}
 
@@
 
-Figure~\ref{fig-LCprop} shows the summary of the proportion of surface
-per landcover class and macrozone, derived from the mask over
-continental Chile.
-
-\hypertarget{landcover-trend-and-drought-indices-1}{%
-\subsubsection{Landcover trend and drought
-indices}\label{landcover-trend-and-drought-indices-1}}
+\hypertarget{land-cover-trend-and-drought-indices-1}{%
+\subsubsection{land cover trend and drought
+indices}\label{land-cover-trend-and-drought-indices-1}}
 
 \begin{table}[!ht]
-\caption{The value of Sen's slope trend next to the time-series plot of surface per landcover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile.
+\caption{The value of Sen's slope trend next to the time-series plot of surface per land cover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile.
 Values of zero indicate that there was not a significant trend.
 Red dots on the plots indicate the maximum and minimum values of surface.}
-\label{tab-landcoverTrend}
+\label{tab-land coverTrend}
 \includegraphics[]{../output/figs/table_var_landcover_macro.png}
@@
 The ``Norte Chico'' shows an increase in barrend land of 111
-\(km^2 year^{-1}\) and a reduction in the class savanna -70
-\(km^2 year^{-1}\). In the ``Centro'' and ``Sur'' there are changes in
-the Chilean matorral, with an important reduction in savanna (-136 to
--318 \(km^2 yr^{-1}\)), and an increase in shrubland and grassland.
-Showing a change for more dense vegetation types. It appears to be a
-change from the surface occupied by cropland from the ``Centro'' to the
-``Sur.'' Also, there is a high increase in forest (397 \(km^2 yr^{-1}\))
-in the ``Sur,'' replacing the savanna lost.
+\(km^2 year^{-1}\) and a reduction in the class savanna of 70
+\(km^2 year^{-1}\). In the ``Centro'' and ``Sur,'' there are changes in
+the Chilean matorral, with an important reduction in savanna (136 to 318
+\(km^2 yr^{-1}\)), and an increase in shrubland and grassland. Showing a
+change for more dense vegetation types. It appears to be a shift in the
+area of cropland from the ``Centro'' to the ``Sur.'' Also, there is a
+high increase in forest (397 \(km^2 yr^{-1}\)) in the ``Sur,'' replacing
+the savanna lost.
+
+Further, we want to address whether the trend in land cover change for
+2001--2023 is associated with trends in drought indices of water demand
+and supply and/or soil moisture for macrozone and land cover
+macroclasses. From the three methods tested, Ridge, Lasso, and Random
+Forest, neither gives significant results regarding whether the trend in
+a drought index for any time scale explains the trend in land cover
+change. Nevertheless, in ``Norte Chico'' and ``Centro,'' there is a
+decrease in croplands and savanna and an increase in barren land, which
+is associated with the variation in drought indices. Mainly for a
+decrease in water supply (SPI and SSI) and an increase in water demand
+(EDDI). However, due to the high variability from north to south in
+Chile, the climatic condition (arid, semi-arid, and humid), and the land
+cover type, we believe that only in those zones could the LULCC be
+driven to some degree by drought.
 
@@
 
-Regarding vegetation productivity aggregated through the macrozones in
-the five landcover macroclasses. In ``Norte Grande'' there is a increase
-trend of 0.02 (z-index) per decade, related with types of grassland and
-shrubland. There is a negative trend in ``Norte Chico'' with -0.04 and
-``Centro'' with -0.02 per decade. In the ``Norte Chico,'' savanna
-(-0.05) has the lowest trend, and the rest of the types are around
--0.04. In ``Centro,'' shrubland reaches -0.06, followed by grassland
-with -0.05, and croplands and savanna have \textasciitilde-0.01 per
-decade. This could be associated either with a reduction in vegetation
-surface, a decrease in biomass, or browning \citep{Miranda2023}.
-Vegetation reached its lowest values since the year 2019, reaching an
-extreme condition in early 2020 and 2022 in the ``Norte Chico'' and
-Centro'' (Mega Drought). The ``Sur'' and ``Austral'' show a positive
-trend of around 0.016 per decade (Figure~\ref{fig-zcNDVI_var}). The area
-most affected by a negative trend in vegetation seems to be associated
-with the Chilean matorral \citep{Fuentes2021}, despite the croplands
-being as severely impacted by drought as native vegetation does in
-``Norte Chico.''.
-
 \begin{figure}[!ht]
@@
 productivity. Grey colors indicate either no vegetation or a change in
-landcover type for 2001--2022. (b) Temporal variation of zcNDVI-6
+land cover type for 2001--2022. (b) Temporal variation of zcNDVI-6
 aggregated at macrozone level within continental Chile. Each horizontal
 panel corresponds to a macrozone from `Norte Grande' to `Austral'.}
+
+\end{figure}
+
+Regarding vegetation productivity aggregated through the macrozones in
+the five land cover macroclasses, in ``Norte Grande,'' there is an
+increase trend of 0.02 (z-index) per decade, related to types of
+grassland and shrubland. There is a negative trend in ``Norte Chico''
+with -0.04 and ``Centro'' with -0.02 per decade. In the ``Norte Chico,''
+savanna (-0.05) has the lowest trend, and the rest of the types are
+around -0.04. In ``Centro,'' shrubland reaches -0.06, grassland -0.05,
+and croplands and savanna -0.01 per decade. This could be associated
+either with a reduction in vegetation surface, a decrease in biomass, or
+browning \citep{Miranda2023}. Vegetation reached its lowest values since
+the year 2019, reaching an extreme condition in early 2020 and 2022 in
+the ``Norte Chico'' and Centro'' (Mega Drought). The ``Sur'' and
+``Austral'' show a positive trend of around 0.016 per decade
+(Figure~\ref{fig-zcNDVI_var}). Despite the croplands suffering from
+drought just as badly as the native vegetation in ``Norte Chico,'' the
+Chilean matorral appears to be the region most affected by a negative
+trend in vegetation \citep{Fuentes2021}.
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/trend_macrozone_drought_indices.png}
+
+}
+
+\caption{\label{fig-trendDIMacro}Trend per decade for the drought
+indices SPI, EDDI, SPEI, and SSI aggregated by macrozone.}
 
@@
 and the lowest for SSI, which shows only a minor increase in the trend
-(Figure~\ref{fig-trendDI}).
+(Figure~\ref{fig-trendDI} and Figure~\ref{fig-trendDIMacro}).
 
@@
 According to what is shown in Figure~\ref{fig-corTimeScale},
-Figure~\ref{fig-corPerson}, and Table \ref{tab-corLandcover}, forest
+Figure~\ref{fig-corPerson}, and Table \ref{tab-corland cover}, forest
 seems to be the most resistant type to drought. Showing that only
 ``Centro'' is slightly (rsq = 0.25) impacted by a 12-month soil moisture
-deficit (SSI-12). Grasslands are shown to be impacted by a SSI-12 with a
-rsq = 0.45 and by a decrease in water supply (SPI-36 and SPEI-24 with
-rsq = 0.28 and 0.34, respectively) in the ``Norte Chico'' and to a
-lesser degree in the ``Norte Grande.'' But, in further south macrozones,
-this type was not affected by water demand, water supply, or soil
-moisture. The types that show to be most affected by variation in
+deficit (SSI-12). In the ``Norte Chico'' and to a lesser extent in the
+``Norte Grande,'' it is evident that a SSI-12 with a rsq = 0.45 and a
+decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34,
+respectively) have an impact on grasslands. However, this type was
+unaffected by soil moisture, water supply, or demand in macrozones
+further south. The types that show to be most affected by variation in
 climate conditions are shrublands, savannas, and croplands. For savannas
@@
 of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45\% and
-66\% of ``Norte Chico.'' The most affected type by climatic variation
-was shrubland, where in ``Norte Chico'' and ``Centro'' soil moisture
-explained 59\% and precipitation, 37\%, with SSI-12 being the most
-relevant variable, followed by SPI-36 in ``Norte Chico'' and SPI-24 in
-``Sur.''
+66\% of ``Norte Chico.'' The type of land most impacted by climatic
+variation was shrubland, where soil moisture explained 59\% and
+precipitation, 37\%, in ``Norte Chico'' and ``Centro,'' with SSI-12
+being the most relevant variable, then SPI-36 in ``Norte Chico'' and
+SPI-24 in ``Sur.''
 
@@
 
+\hypertarget{ecological-and-agricultural-drought}{%
+\subsection{Ecological and agricultural
+drought}\label{ecological-and-agricultural-drought}}
+
+\hypertarget{drought-trend-lulcc-and-climate-conditions}{%
+\subsection{Drought trend, LULCC, and climate
+conditions}\label{drought-trend-lulcc-and-climate-conditions}}
+
 1.- Respecto a lo que indica \citet{Vicente-Serrano2018}, de que el
 aumento en la tendencia en severidad de la sequía (hidrológica) tiene
-que ver más con un aumento de la demanda de agua (LULCC) que a una
-tendencia en las condiciones climáticas (SPI-12).
-
-2.- Sobre los tipos de landcover más afectados por los indicadores de
+que ver más con un aumento de la demanda de agua (ej, LULCC, amazonas)
+que a una tendencia en las condiciones climáticas (SPI-12). ¿Qué pasa en
+Chile?
+
+\hypertarget{land-cover-types-most-impacted-by-drought-throughout-chile}{%
+\subsection{land cover types most impacted by drought throughout
+Chile}\label{land-cover-types-most-impacted-by-drought-throughout-chile}}
+
+2.- Sobre los tipos de land cover más afectados por los indicadores de
 sequía. Asociación con el matorral chileno \citep{Fuentes2021}.
-Diferencia entre el Norte Chico y Centro y lo que pasa hacía el sur.
+Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur.
+
+\hypertarget{drought-indices-of-water-demand-and-supply-soil-moisture-to-predict-changes-in-vegetation-productivity}{%
+\subsection{Drought indices of water demand and supply, soil moisture to
+predict changes in vegetation
+productivity}\label{drought-indices-of-water-demand-and-supply-soil-moisture-to-predict-changes-in-vegetation-productivity}}
 
@@
 \item
-  Como podrian servir estos resultados para desarrollar un predictor de
-  sequía basado.
+  Como podrían servir estos resultados para desarrollar o mejorar un
+  predictor de productividad de la vegetación.
 \end{enumerate}
 
-4.- Qué se podria hacer mejor en futuras investigaciones del tema.
-
-5.- Sequía, aridez, water scarcity \citep{VanLoon2013} Si puede utilizar
-el mapa y proyección climática de Koppen\_Geiger \citep{Beck2023} para
-dirigir esta discusión
+\begin{itemize}
+\tightlist
+\item
+  Los datos ERA5L están casi en tiempo real, 7 dias; MODIS también.
+\item
+  EL SSI se ve como un poderoso indicador que explica la variabilidad en
+  la productividad de la vegetación.
+\end{itemize}
+
+\hypertarget{future-outlook}{%
+\subsection{Future outlook}\label{future-outlook}}
+
+4.- Qué se podría hacer mejor en futuras investigaciones del tema. -
+mejorar la resolución y calidad de los datos climáticos -
 

### 2024-01-22 scripts/34_trend_zcNDVI_macrozona_y_landcover.R

@@ -11,20 +11,11 @@ files <- dir_ls(dir,regexp = 'mann_kendall.*zcNDVI-6.*tif$')
 trends <- rast(files)[[2]]
 names(trends) <- 'zcNDVI-6'
 
-#mascara para elevaciones mayores a 1500m
-elev <- geodata::elevation_30s('chile',path = tempdir()) |> 
-  project('EPSG:32719')
-elev2 <- resample(elev,trends)
-mask <- elev2
-mask[mask<1500] <- 1
-mask[mask>=1500] <- NA
+
 
 macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
   mutate(macrozona = c('Norte Chico','Norte Grande','Austral','Centro','Sur')) |> 
   st_transform(32719)
-
-trends <- resample(trends,mask)
-trends <- mask(trends,mask)
 
 # ahora por tipo de landcover
 

### 2024-01-22 article/paper_drought_Chile2023.tex
actualizar repo gema

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, sharp corners, frame hidden, interior hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{ecological-and-agricultural-drought}{%
-\subsection{Ecological and agricultural
-drought}\label{ecological-and-agricultural-drought}}
-
 \hypertarget{drought-trend-lulcc-and-climate-conditions}{%
@@
 sequía. Asociación con el matorral chileno \citep{Fuentes2021}.
-Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur.
+Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur
+(\citet{Miranda2023})
 

### 2024-01-23 article/paper_drought_Chile2023.qmd
actualizar repo gema
actualizar repo depto

@@
 
-###	in-situ data
+###	Weather stations
 
-POR COMPLETAR
+We validated the ERA5L variables for monthly mean temperature, total precipitation, and volumetric soil water content.
+For temperature and precipitation, we used the weather network from the Ministry of Agriculture of Chile (www.agromet.com).
+We used 266 stations located throughout Chile.
+For soil moisture, we select a soil network for water content and temperature, localized in the Aconcagua's river basin ("Norte Chico").
+This network corresponds to 53 sensors of the type Temperature Moisture Sensor (TMS) [@Wild2019].
 
@@
 
-POR COMPLETAR
+To account for the performance of the ERA5L climatic variables regarding the values measured by the weather stations. We selected the following metrics:
+
+$$MAE = \frac{1}{n}\sum |{E-S}|$$
+$$Bias = \frac{\sum E}{\sum S}$$
+$$ubRMSE =\sqrt{\frac{\sum{ \left[ (E_i-\overline{E})-(S_i-\overline{S}) \right ] ^2}}{n}}$$
+
+$$CC = \frac{\sum (S_i-\bar{S})(E_i-\bar{E})}{\sqrt{(Si-\bar{S})^2(E_i-\bar{E})^2}}$$
+
+$MAE$: mean absolute error
+$bias$: bias
+$ubRMSE$: unbiassed root mean squared error
+$CC$: coefficient of correlation
+$S$: value of the variable measure by the weather station
+$E$: value of the variable measure by ERA5L
 
@@
 ### Validation of ERA5L variables
+
+The average metrics of performance of ERA5L over the 266 weather stations were in the case of temperature: $ubRMSE=1.06\,°C$, $MAE=1.131\,°C$, and $CC=0.963$.
+For precipitation, $MAE=28.1\,mm$, $Bias=1.93$, and $CC=0.845$.
 
@@
 
-## Ecological and agricultural drought
-
 ## Drought trend, LULCC, and climate conditions
@@
 
-2.- Sobre los tipos de land cover más afectados por los indicadores de sequía. Asociación con el matorral chileno [@Fuentes2021]. Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur.
+2.- Sobre los tipos de land cover más afectados por los indicadores de sequía.
+Asociación con el matorral chileno [@Fuentes2021].
+Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur (@Miranda2023)
 

### 2024-01-23 scripts/17_extrarer_correlations_macro_landcover.R

@@ -36,7 +36,7 @@ data_ind <- map_df(1:5,function(i){
   cors_m <- mask(cors1,macro[i,])
   lc_m <- mask(lc,macro[i,])
   cors_df <- zonal(cors_m,lc_m,getmode)
-  names(cors_df)[2:5] <- c('eddi','spi','spei','zcsm')
+  names(cors_df)[2:5] <- c('EDDI','SPI','SPEI','SSI')
   cors_df |> 
     pivot_longer(-1) |> 
     rename(indice = name) |> 
@@ -65,7 +65,7 @@ data_r <- map_df(1:5,function(i){
   cors_m <- mask(cors2,macro[i,])
   lc_m <- mask(lc,macro[i,])
   cors_df <- zonal(cors_m,lc_m,na.rm = TRUE)
-  names(cors_df)[2:5] <- c('eddi','spi','spei','zcsm')
+  names(cors_df)[2:5] <- c('EDDI','SPI','SPEI','SSI')
   cors_df |> 
     pivot_longer(-1) |> 
     rename(indice = name) |> 
@@ -86,13 +86,42 @@ data_r_4gt <- data_r |>
   pivot_wider(names_from=c(clase, name), values_from=value) |> 
   rename_with(\(x) str_c(x,'_r'),-macro)
 
-tabla_gt <- full_join(data_ind_4gt,data_r_4gt) 
+tabla_gt <- full_join(data_ind_4gt,data_r_4gt,by = 'macro') 
+tabla_gt <- tabla_gt |> mutate(macrozone = str_to_title(macro)) |> 
+  select(-macro) |> relocate(macrozone)
 
 library(gt)
 
+tabla_gt$Savanna_EDDI[1] <- tabla_gt$Savanna_SPI[1] <- tabla_gt$Savanna_SPEI[1] <- tabla_gt$Savanna_SSI[1] <- NA
+tabla_gt$Savanna_EDDI_r[1] <- tabla_gt$Savanna_SPI_r[1] <- tabla_gt$Savanna_SPEI_r[1] <- tabla_gt$Savanna_SSI_r[1] <- NA
[…]

### 2024-01-23 scripts/18_extraer_correlations_macro_landcover_meses.R

@@ -17,13 +17,10 @@ pal <- read.csv('data/processed_data/paleta_colores_landcover.csv')
 t <- pal$class
 names(t) <- pal$Name
 
-lc <- rast('/mnt/md0/raster_procesada/MODIS_derived/IGBP.MCD12Q1.061/IGBP80_reclassified.tif')
-
 dir <- '/mnt/md0/raster_procesada/analysis/correlations/meses/'
 files <- dir_ls(dir,regexp = '.*EDDI.*tif$')
 
-cors <- rast(files)
-cors <- resample(cors,lc)
+
 
 macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
   st_transform(32719)

### 2024-01-23 scripts/21_landcover_per_macrozone.R

@@ -41,6 +41,28 @@ LC_data |>
   group_by(zone,LC_type)|> 
   summarize(ntot = n()) -> summ_data2
 
+#tabla con superficie
+#
+summ_data2 |> 
+  mutate(surface = 21.19*ntot*0.01) |> 
+  select(-ntot) |> 
+  mutate(surface = case_when(surface <=20~NA,
+                           .default = surface)) |> 
+  pivot_wider(names_from = LC_type,values_from=surface) |> 
+  rename(macrozone = zone) |> 
+  ungroup() |> 
+   select(1,7:6,4:2,5) |> 
+  gt() |> 
+  fmt_number(decimals = 0) |> 
+  sub_missing(
+    columns = everything(),
+    missing_text = ""
+  ) |> 
+  tab_header(
+    title = md("Surface  [km<sup>2]"),
+  ) |> 
+  gt::gtsave('output/figs/table_surface_landcover_macrozone.png')
+
 summ_data2 |> 
   group_by(zone) |>  
   summarize(totZone = sum(ntot)) |>  

### 2024-01-23 scripts/23_analysis_trend_drought_indices_and_landcover_macroXlandcover.R

@@ -15,6 +15,7 @@ data_ana <- data_trend_di |>
 
 data_model <- data_ana[,c(1:6,9:12,15:18,21:24,29,31)]
 data_model <- data_model |> 
+  filter(macro %in% c('Norte Chico','Centro')) |> 
   #select(-starts_with('EDDI')) |> 
   mutate(
     #landcover = cut(landcover,5),

### 2024-01-23 scripts/25_trend_drought_indices.R

@@ -239,8 +239,19 @@ tmap_save(mapa1,'output/figs/trend_raster_zcNDVI6_SPIs.png',asp=.3)
 data_df <- bind_rows(df_spi,df_spei,df_eddi,df_zcsm)
 
 data_df |> 
-  mutate(index =str_remove(name,'-.*')) |> 
-  ggplot(aes(scale,value,color=index)) +
+  mutate(index =str_remove(name,'-.*'),
+         macro = fct_relevel(macro,'Norte Grande','Norte Chico','Centro','Sur','Austral'),
+         index = case_when(index == 'zcSM' ~ 'SSI',
+                         .default = index)) |> 
+  ggplot(aes(scale,value*10,color=index)) +
   geom_line() +
   geom_point() +
-  facet_grid(.~macro,scale = 'free')
+  scale_colour_viridis_d(name = 'Drought index') +
+  scale_y_continuous(breaks= seq(-0.05,0.05,length.out=9)) +
+  scale_x_continuous(breaks = c(1,3,6,12,24,36)) +
+  facet_grid(.~macro,scale = 'free') +
+  labs(y = 'Trend per decade',x= 'Time scales') +
+  theme_bw() +
+  theme(strip.background = element_rect(fill = 'white'),
+        legend.position = 'bottom')
+ggsave('output/figs/trend_macrozone_drought_indices.png',width =10,height=3)

### 2024-01-23 scripts/35_validacion_era5_soil_water.R

@@ -1,0 +1,84 @@
+library(tidyverse)
+library(terra)
+library(sf)
+library(agvAPI)
+library(fs)
+
+data_sm_tb <- read_rds('data/processed/data_sm_2022.rds') |> 
+  #filter(depth == 30) |> 
+  mutate(datetime =ymd_hm(datetime)) |> 
+  group_by(day = floor_date(datetime,'1 day'),serial) |> 
+  summarize(SM = mean(SM,na.rm = TRUE)) |> 
+  mutate(day = as.character(ymd(day)))
+
+dir_era_sm <- '/mnt/md0/raster_procesada/ERA5-Land_tiff/clima/volumetric_soil_water/daily'
+
+files_era <- dir_ls(dir_era_sm,regexp = '2022.*tif$')
+dates_era <- str_extract(files_era,'[0-9]{4}-[0-9]{2}-[0-9]{2}')
+sm_era <- lapply(seq_along(files_era),\(i){
+  out <- app(rast(files_era[i],lyrs=1:3),mean)
+  names(out) <- dates_era[i]
+  out
+})
+sm_era <- rast(sm_era)
+
+est_garces_sf <- estaciones_garces |> 
+  filter(tipo == 'Humedad_Suelo') |> 
+  st_as_sf(coords = c('lon','lat'),crs=4326)
+         
+data_sm <- terra::extract(sm_era,est_garces_sf)
+
+dates <- seq(ymd("20220101"),ymd("20221231"),by='1 day')
+data_sm <- cbind(st_drop_geometry(est_garces_sf[,'serial']),data_sm)
+#names(data_sm)[3:367] <- as.character(dates)
+
+rmse <- function(x,y) {
+  sqrt(mean((x-y)^2))
+  }
+
+data_sm |> 
+  select(-ID) |> 
+  pivot_longer(-serial,names_to = 'day') |> 
+  right_join(data_sm_tb) |> 
+  mutate(SM = SM/100,date = ymd(day)) |> 
+  ggplot() +
+  geom_point(aes(date,SM),colour='blue') + 
+  geom_line(aes(date,SM),colour='blue') + 
+  geom_point(aes(date,value),colour='red') + 
+  geom_line(aes(date,value),colour='red')  +
+  facet_wrap(serial~.)
+
+data_sm |> 
+  select(-ID) |> 
+  pivot_longer(-serial,names_to = 'day') |> 
+  right_join(data_sm_tb) |> 
[…]

### 2024-01-23 scripts/36_estimar_vwc_sensores_toms.R

@@ -1,0 +1,73 @@
+library(terra)
+library(sf)
+
+meta <- read_rds('data/processed_data/metadata_tomsT4.rds')
+
+meta_sf <- meta |> 
+  group_by(Sitio) |> 
+  summarize(lon = mean(lon_wgs84),
+            lat = mean(lat_wgs84)) |> 
+  st_as_sf(coords = c('lon','lat'),crs = 4326)
+
+dir <- '/home/francisco/Descargas/SoilGRID250m'
+
+files_clay <- dir_ls(paste0(dir,'/clay_chl'),regexp = 'tif$')
+
+clay_100 <- rast(files_clay[-2])
+clay_100 <- app(clay_100,mean)
+
+clay_df <- terra::extract(clay_100,meta_sf) |> 
+  cbind(meta_sf |> st_drop_geometry() |> select(Sitio))
+
+clay_df <- clay_df |> 
+  mutate(mean = mean/10) |> 
+  rename(clay = mean)
+
+files_sand <- dir_ls(paste0(dir,'/sand_chl'),regexp = 'tif$')
+
+sand_100 <- rast(files_sand[-2])
+sand_100 <- app(sand_100,mean)
+
+sand_df <- terra::extract(sand_100,meta_sf) |> 
+  cbind(meta_sf |> st_drop_geometry() |> select(Sitio))
+
+sand_df <- sand_df |> 
+  mutate(mean = mean/10) |> 
+  rename(sand = mean)
+
+files_silt <- dir_ls(paste0(dir,'/silt_chl'),regexp = 'tif$')
+
+silt_100 <- rast(files_silt[-2])
+silt_100 <- app(silt_100,mean)
+
+silt_df <- terra::extract(silt_100,meta_sf) |> 
+  cbind(meta_sf |> st_drop_geometry() |> select(Sitio))
+
+silt_df <- silt_df |> 
+  mutate(mean = mean/10) |> 
+  rename(silt = mean)
+
+data_soil <- bind_cols(sand_df,clay_df,silt_df) |> 
+  select(3,5,8,2) |> 
+  rename(sitio = 1) |> 
+  as_tibble()
+
+iden_soil <- function(v){
+  d <- dist(rbind(v,mc_data_vwc_parameters[,6:8]))
+  n <- d |> as.matrix() |> _[-1,1] |> which.min() |> as.numeric()
+  mc_data_vwc_parameters$soiltype[n]
+}
+
[…]

### 2024-01-23 article/paper_drought_Chile2023.qmd
actualizar validacion era5

@@
 
-We validated the ERA5L variables for monthly mean temperature, total precipitation, and volumetric soil water content.
-For temperature and precipitation, we used the weather network from the Ministry of Agriculture of Chile (www.agromet.com).
-We used 266 stations located throughout Chile.
-For soil moisture, we select a soil network for water content and temperature, localized in the Aconcagua's river basin ("Norte Chico").
-This network corresponds to 53 sensors of the type Temperature Moisture Sensor (TMS) [@Wild2019].
+We compared the ERA5L variables for monthly mean temperature, total precipitation, and volumetric soil water content against values retrieved by weather stations.
+For temperature and precipitation, we used the weather network from the Ministry of Agriculture of Chile (www.agromet.com) between 2015 and 2023.
+We used 277 stations located throughout Chile.
+For soil moisture, we select a private soil network that is owned by the agricultural enterprise Garces Fruit, which has 99 stations in Central Chile, located in cherry fruit crops.
+We used daily data for the year 2022 and the months outside the growing season, May to September, to avoid the effect of irrigation on soil moisture, which is hardly captured by ERA5L.
 
@@
 
-The average metrics of performance of ERA5L over the 266 weather stations were in the case of temperature: $ubRMSE=1.06\,°C$, $MAE=1.131\,°C$, and $CC=0.963$.
-For precipitation, $MAE=28.1\,mm$, $Bias=1.93$, and $CC=0.845$.
+The average metrics of performance of ERA5L over the 266 weather stations were in the case of monthly temperature: $ubRMSE=1.06\,°C$, $MAE=1.131\,°C$, and $CC=0.963$, showing a good agreement, low error, and low overestimation.
+For cumulative monthly precipitation, $MAE=28.1\,mm$, $bias=1.93$, and $CC=0.845$, showing a high correlation and a 93% bias and being overestimated by ERA5L.
+In the case of the 97 soil moisture stations, we averaged for the three depths (30, 60, and 90m) and then compared it with volumetric water content at 1m derived from ERA5L.
+For this case, we made a daily comparison, having a $CC=0.71$, $RMSE=0.174\,m^3m^{-3}$, $MAE=0.167\,m^3m^{-3}$, and $bias=1.74$.
+The ERA5 soil moisture overestimate is 74%, but it has a kind of good correlation.
 

### 2024-01-23 article/paper_drought_Chile2023.tex
actualizar validacion era5

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{in-situ-data}{%
-\subsubsection{in-situ data}\label{in-situ-data}}
-
-POR COMPLETAR
+\hypertarget{weather-stations}{%
+\subsubsection{Weather stations}\label{weather-stations}}
+
+We compared the ERA5L variables for monthly mean temperature, total
+precipitation, and volumetric soil water content against values
+retrieved by weather stations. For temperature and precipitation, we
+used the weather network from the Ministry of Agriculture of Chile
+(www.agromet.com) between 2015 and 2023. We used 277 stations located
+throughout Chile. For soil moisture, we select a private soil network
+that is owned by the agricultural enterprise Garces Fruit, which has 99
+stations in Central Chile, located in cherry fruit crops. We used daily
+data for the year 2022 and the months outside the growing season, May to
+September, to avoid the effect of irrigation on soil moisture, which is
+hardly captured by ERA5L.
 
@@
 
-POR COMPLETAR
+To account for the performance of the ERA5L climatic variables regarding
+the values measured by the weather stations. We selected the following
+metrics:
+
+\[MAE = \frac{1}{n}\sum |{E-S}|\] \[Bias = \frac{\sum E}{\sum S}\]
+\[ubRMSE =\sqrt{\frac{\sum{ \left[ (E_i-\overline{E})-(S_i-\overline{S}) \right ] ^2}}{n}}\]
+
+\[CC = \frac{\sum (S_i-\bar{S})(E_i-\bar{E})}{\sqrt{(Si-\bar{S})^2(E_i-\bar{E})^2}}\]
+
+\(MAE\): mean absolute error \(bias\): bias \(ubRMSE\): unbiassed root
+mean squared error \(CC\): coefficient of correlation \(S\): value of
+the variable measure by the weather station \(E\): value of the variable
+measure by ERA5L
 
@@
 variables}\label{validation-of-era5l-variables-1}}
+
+The average metrics of performance of ERA5L over the 266 weather
+stations were in the case of monthly temperature: \(ubRMSE=1.06\,°C\),
+\(MAE=1.131\,°C\), and \(CC=0.963\), showing a good agreement, low
+error, and low overestimation. For cumulative monthly precipitation,
+\(MAE=28.1\,mm\), \(bias=1.93\), and \(CC=0.845\), showing a high
+correlation and a 93\% bias and being overestimated by ERA5L. In the
+case of the 97 soil moisture stations, we averaged for the three depths
+(30, 60, and 90m) and then compared it with volumetric water content at
+1m derived from ERA5L. For this case, we made a daily comparison, having
+a \(CC=0.71\), \(RMSE=0.174\,m^3m^{-3}\), \(MAE=0.167\,m^3m^{-3}\), and
+\(bias=1.74\). The ERA5 soil moisture overestimate is 74\%, but it has a
+kind of good correlation.
 

### 2024-01-23 scripts/35_validacion_era5_soil_water.R

@@ -4,12 +4,12 @@ library(sf)
 library(agvAPI)
 library(fs)
 
-data_sm_tb <- read_rds('data/processed/data_sm_2022.rds') |> 
+data_sm_tb <- read_rds('data/processed_data/data_sm_2022_garces.rds') |> 
   #filter(depth == 30) |> 
   mutate(datetime =ymd_hm(datetime)) |> 
-  group_by(day = floor_date(datetime,'1 day'),serial) |> 
-  summarize(SM = mean(SM,na.rm = TRUE)) |> 
-  mutate(day = as.character(ymd(day)))
+  group_by(date = floor_date(datetime,'1 day'),serial) |> 
+  summarize(SM_g = mean(SM,na.rm = TRUE)) |> 
+  mutate(date = as.character(ymd(date)))
 
 dir_era_sm <- '/mnt/md0/raster_procesada/ERA5-Land_tiff/clima/volumetric_soil_water/daily'
 
@@ -32,53 +32,44 @@ dates <- seq(ymd("20220101"),ymd("20221231"),by='1 day')
 data_sm <- cbind(st_drop_geometry(est_garces_sf[,'serial']),data_sm)
 #names(data_sm)[3:367] <- as.character(dates)
 
-rmse <- function(x,y) {
-  sqrt(mean((x-y)^2))
-  }
-
-data_sm |> 
+data_unida_sm <- data_sm |> 
   select(-ID) |> 
-  pivot_longer(-serial,names_to = 'day') |> 
-  right_join(data_sm_tb) |> 
-  mutate(SM = SM/100,date = ymd(day)) |> 
-  ggplot() +
-  geom_point(aes(date,SM),colour='blue') + 
-  geom_line(aes(date,SM),colour='blue') + 
-  geom_point(aes(date,value),colour='red') + 
-  geom_line(aes(date,value),colour='red')  +
-  facet_wrap(serial~.)
-
-data_sm |> 
-  select(-ID) |> 
-  pivot_longer(-serial,names_to = 'day') |> 
-  right_join(data_sm_tb) |> 
-  mutate(SM = SM/100,date = ymd(day)) |> 
+  pivot_longer(-serial,names_to = 'date',values_to = 'SM_era5') |> 
+  left_join(data_sm_tb) |> 
+  mutate(SM_g = SM_g/100,date = ymd(date)) 
+  
+library(yardstick)
+metrics_sm <- data_unida_sm |> 
+  mutate(num_mes = month(date)) |> 
[…]

### 2024-01-23 scripts/37_validacion_era5_temp_pre.R

@@ -1,0 +1,172 @@
+## Validación datos ERA5-Land
+
+library(RSQLite)
+pool <- dbConnect(RPostgres::Postgres(),
+                  dbname = "shiny",
+                  host = Sys.getenv("HOST"),
+                  user = "shiny",
+                  password = Sys.getenv("SHINY_PSQL_PWD"))
+
+
+
+data <- tbl(pool, "estaciones_datos")
+
+library(tidyverse)
+data_val <- data |> 
+  select(station_id,fecha_hora,temp_promedio_aire,precipitacion_horaria) |> 
+  filter(fecha_hora >= "2015-01-01") |> 
+  collect()
+
+data_sel <- data_val |> 
+  tidyr::complete(station_id,fecha_hora,fill = list(temp_promedio_aire =NA)) |> 
+  # agrupar por station_id y mes-año
+  dplyr::group_by(station_id, 
+                  mes = lubridate::floor_date(fecha_hora,'month')) |>
+  #calcular el promedio de temperatura y la cantidad de NAs por estación y mes-año
+  dplyr::summarize(temp = mean(temp_promedio_aire,na.rm = TRUE),
+                   prop_NAs = sum(is.na(temp_promedio_aire))/n()) |> 
+  #filtrar los meses que tienen menos de 20% de NAs
+  dplyr::filter(prop_NAs < .2)
+
+data_sel |> 
+  ggplot(aes(as.factor(mes),as.factor(station_id),fill=temp)) +
+  geom_tile() +
+  scale_fill_viridis_c() +
+  #scale_x_discrete() +
+  #scale_x_date(date_breaks = "1 year",date_labels = '%Y-%b',expand = c(0,0)) +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 90))
+
+emas_sel <- data_sel |> 
+  ungroup() |> 
+  tidyr::complete(station_id,mes,fill = list(temp = NA)) |> 
+  dplyr::group_by(station_id) |> 
+  dplyr::summarize(prop_NAs = sum(is.na(temp))/n()) |> 
+  dplyr::filter(prop_NAs < .2) |> 
+  dplyr::pull(station_id)
+
+## Funcion de filtro
+
+filt_out <- function(x,c,...){
[…]

### 2024-01-24 article/paper_drought_Chile2023.tex
actualizar

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, interior hidden, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
 

### 2024-01-24 article/paper_drought_Chile2023.qmd
actualizar discusion

@@
       model: 3p
-      cite-style: authoryear
+      cite-style: number
       natbiboptions: numbers
@@
 In this research, we aim to analyze the impact of drought on different types of ecosystems (land cover classes) in continental Chile.
-Our specific goals are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to assess the LULC change for 2001–2021 and how it relates to drought indices; iii) to evaluate the relationship between zcNDVI, a measure of vegetation productivity, and drought indices for water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.
+Our specific goals are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to assess the LULC change for 2001–2022 and how it relates to drought indices; iii) to evaluate the relationship between zcNDVI, a measure of vegetation productivity, and drought indices for water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.
 
@@
 
-### land cover macroclasess and validation
+### Land cover macroclasess and validation
 
@@
 
-### land cover persistence mask 2001-2022
+### Land cover persistence mask 2001-2022
 
@@
 
-### land cover trend and drought indices
+### Land cover trend and drought indices
 
@@
 
-### land cover persistence mask 2001-2022
+### Land cover persistence mask 2001-2022 {#sec-persistence_mask}
 
@@
 The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var).
-Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the Chilean matorral appears to be the region most affected by a negative trend in vegetation [@Fuentes2021].
+Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the Chilean matorral [@Fuentes2021] appears to be the region most affected by a negative trend in vegetation productivity.
 
@@
 
-## Drought trend, LULCC, and climate conditions
+## Drought trend and attribution to LULCC
 
-1.-  Respecto a lo que indica @Vicente-Serrano2018, de que el aumento en la tendencia en severidad de la sequía (hidrológica) tiene que ver más con un aumento de la demanda de agua (ej, LULCC, amazonas) que a una tendencia en las condiciones climáticas (SPI-12).
+@Vicente-Serrano2021, in a study at the global scale of drought trends, indicates that there have not been significant trends in meteorological drought since 1950.
+Also, state that the increase in hidrological trend in some parts of the globe (northeast Brazil and the Mediterranean region) is related to changes in land cover and specifically to the rapidly increasing irrigated area, which consequently increases water extraction.
+@Kogan2020 analyzed the agricultural drought impact globally and in the main grain producer countries, finding that _"since 1980, the Earth warming has not changed the drought area or intensity"_.
+In our study, we considered the variation in vegetation productivity in Chile for areas without changes in land cover macroclasses (see @sec-persistence_mask), to avoid misleading conclusions that could be related to the increase in water demand due to LULCC.
+Our results show a contrasting perspective.
+There has been a significant trend in the decline of vegetation productivity (zcNDVI) since 2000 for "Norte Chico" and "Centro," which has been extreme between 2020 and 2022, seemsly due to an intense hydrological drought due to the persistance of the Mega Drought [@Garreaud2017].
+Despite we used the persistance mask for vegetations's trend analysis, the cropland, which is the most water-demand type, showed a decrease trend in "Norte Chico" and "Centro." Also, there was an increase in barren land for both types.
+These changes are associated with a decrease in water demand from vegetation.
+On the other hand, for "Norte Chico" and "Centro," our results show a decrease in trends of water supply (SPI and SSI), which are higher at larger time scales and consequently impact the hydrological system.
+We claim that what occurred in central Chile defies findings made at the global level [@Vicente-Serrano2021;@Kogan2020], demonstrating that a constant decrease in water supply rather than an increase in water demand (i.e., irrigated crops) is the main cause of the hydrological drought.
+Finally, central Chile has shown a diminishment in vegetation productivity for all macroclasses, mainly attributed to variation in water supply, i.e., precipitation, which could be strengthened by an increase in water demand by, for example, irrigated crops.
+
+1.-  Respecto a lo que indica @Vicente-Serrano2022, de que el aumento en la tendencia en severidad de la sequía (hidrológica) tiene que ver más con un aumento de la demanda de agua (ej, LULCC, amazonas) que a una tendencia en las condiciones climáticas (SPI-12).
 ¿Qué pasa en Chile?
 
-## land cover types most impacted by drought throughout Chile
+@Chen2022: Assessing the impact of drought-land cover change on global vegetation greenness and productivity
+
+@Kogan2020: muestra que la sequía no se ha intensificado ni expandido a nivel global, por 38 años.
+
+## Land cover types and their impact by drought
+
+Extreme drought conditions are an important driver of tree mortality, as shown by @Senf2020 in Europe.
+However, we found that forest is the type of landcover macroclass less affected by variation in drought indices, being the most resistant landcover class to drought.
+Supporting this is @Fathi-Taperasht2022, who asserts that Indian forests are the most drought-resistant and recover rapidly.
+Similarly, the work of @Wu2024 who analyzed vegetation loss and recovery in response to meteorological drought in the humid subtropical Pearl River basin in China, indicates that forests showed higher drought resistance.
+@Xiao2023 using Vegetation Optical Depth (VOD), kNDVI, and EVI, evaluates ecosystem resistance, showing that, compared to croplands, ecosystems with a higher proportion of forests are more resilient to severe droughts.
+They attribute the difference to a deeper rooting depth of trees, a higher water storage capacity, and different water use strategies between them [@Xiao2023].
 
@@
 Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur (@Miranda2023)
+
+@Fathi-Taperasht2022: muestra que bosques son más resilentes a la sequía en India.
+
+@Wu2024: evalua la perdida de vegetación en diferentes tipos de uso de suelo. Muestra que bosques son más resilentes y cultivos y pastizales, los menos.
+
+@Jiang2020: impacto de la sequía en landcover en China.
 

### 2024-01-24 article/paper_drought_Chile2023.tex
actualizar discusion

@@
 \documentclass[
-  authoryear,
+  number,
   preprint,
@@
 \usepackage[]{natbib}
-\bibliographystyle{elsarticle-harv}
+\bibliographystyle{elsarticle-num}
 \IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, interior hidden, enhanced, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi
 
@@
 classified as meteorological, hydrological, and agricultural
-\citep{Wilhite1985}. Lately, \citet{Loon2016} and
-\citet{AghaKouchak2021} have given an updated definition of drought for
+\citep{Wilhite1985}. Lately, \citep{Loon2016} and
+\citep{AghaKouchak2021} have given an updated definition of drought for
 the Anthropocene, suggesting that it should be considered the feedback
@@
 drought. Even though it has been argued that those definitions do not
-fully address the ecological dimensions of drought, \citet{Crausbay2017}
+fully address the ecological dimensions of drought, \citep{Crausbay2017}
 proposed the ecological drought definition as ``an episodic deficit in
@@
 To evaluate meteorological drought (i.e., water supply), the World
-Meteorological Organization (WMO; \citet{WMO2012}) recommends the
-Standardized Precipitation Index (SPI; \citet{Mckee1993}), a multiscalar
+Meteorological Organization (WMO; \citep{WMO2012}) recommends the
+Standardized Precipitation Index (SPI; \citep{Mckee1993}), a multiscalar
 drought index that allows to monitor precipitation deficits from short-
-to long-term. Following the same approach, \citet{Vicente-Serrano2010}
+to long-term. Following the same approach, \citep{Vicente-Serrano2010}
 incorporates into the SPI the effect of temperature through the use of
@@
 Precipitation Evapotranspiration Index). Similarly, to evaluate solely
-the evaporative demand driven by temperature, \citet{Hobbins2016} and
-\citet{McEvoy2016} came up with the Evaporative Demand Drought Index
+the evaporative demand driven by temperature, \citep{Hobbins2016} and
+\citep{McEvoy2016} came up with the Evaporative Demand Drought Index
 (EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI;
-\citet{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
+\citep{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
 cumulative Normalized Difference Vegetation Index (NDVI), which could be
@@
 productivity for 1981--2023/2001--2023; ii) to assess the LULC change
-for 2001--2021 and how it relates to drought indices; iii) to evaluate
+for 2001--2022 and how it relates to drought indices; iii) to evaluate
 the relationship between zcNDVI, a measure of vegetation productivity,
@@
 \citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
-use the Koppen-Geiger release by \citet{Beck2023} and the land cover
+use the Koppen-Geiger release by \citep{Beck2023} and the land cover
 type persistance of 80\% of times for 2001--2022, from the IGBP
@@
 polynomial regression, following the procedure described in
-\citet{Zambrano2018} and \citet{Zambrano2016}.
+\citep{Zambrano2018} and \citep{Zambrano2016}.
 
@@
 for the scale \(s\), we used a nonparametric approach following
-\citet{Hobbins2016} to derive the drought indices. Thus, the empirically
+\citep{Hobbins2016} to derive the drought indices. Thus, the empirically
 derived probabilities are obtained through an inverse normal
@@
 \hypertarget{land-cover-macroclasess-and-validation}{%
-\subsubsection{land cover macroclasess and
+\subsubsection{Land cover macroclasess and
 validation}\label{land-cover-macroclasess-and-validation}}
@@
 To validate the land cover obtained, we compare the macroclasses with
-the ones of a more detailed land cover map made by \citet{Zhao2016} for
+the ones of a more detailed land cover map made by \citep{Zhao2016} for
 Chile with samples acquired in the years 2013--2014 (LCChile). The later
@@
 \hypertarget{land-cover-persistence-mask-2001-2022}{%
-\subsubsection{land cover persistence mask
+\subsubsection{Land cover persistence mask
 2001-2022}\label{land-cover-persistence-mask-2001-2022}}
@@
 \hypertarget{land-cover-trend-and-drought-indices}{%
-\subsubsection{land cover trend and drought
+\subsubsection{Land cover trend and drought
 indices}\label{land-cover-trend-and-drought-indices}}
@@
 
-\citet{Vicente-Serrano2021} made a global analysis of the drought's
+\citep{Vicente-Serrano2021} made a global analysis of the drought's
 severity trend using SPI, SPEI, and the Standardized Evapotranspiration
-Deficit Index (SEDI; \citet{Vicente-Serrano2018}) to evaluate AED. They
+Deficit Index (SEDI; \citep{Vicente-Serrano2018}) to evaluate AED. They
 indicate that the increase in hydrological drought has been due to
@@
 36-month time scales against zcNDVI. We followed a similar approach to
-that used by \citet{Meroni2016} when using the SPI for meteorological
+that used by \citep{Meroni2016} when using the SPI for meteorological
 drought against the cumulative FAPAR (Fraction of Absorbed
@@
 \hypertarget{land-cover-macroclasess-and-validation-1}{%
-\subsubsection{land cover macroclasess and
+\subsubsection{Land cover macroclasess and
 validation}\label{land-cover-macroclasess-and-validation-1}}
@@
 
-\hypertarget{land-cover-persistence-mask-2001-2022-1}{%
-\subsubsection{land cover persistence mask
-2001-2022}\label{land-cover-persistence-mask-2001-2022-1}}
+\hypertarget{sec-persistence_mask}{%
+\subsubsection{Land cover persistence mask
+2001-2022}\label{sec-persistence_mask}}
 
@@
 \hypertarget{land-cover-trend-and-drought-indices-1}{%
-\subsubsection{land cover trend and drought
+\subsubsection{Land cover trend and drought
 indices}\label{land-cover-trend-and-drought-indices-1}}
@@
 drought just as badly as the native vegetation in ``Norte Chico,'' the
-Chilean matorral appears to be the region most affected by a negative
-trend in vegetation \citep{Fuentes2021}.
+Chilean matorral \citep{Fuentes2021} appears to be the region most
+affected by a negative trend in vegetation productivity.
 
@@
 
-\hypertarget{drought-trend-lulcc-and-climate-conditions}{%
-\subsection{Drought trend, LULCC, and climate
-conditions}\label{drought-trend-lulcc-and-climate-conditions}}
-
-1.- Respecto a lo que indica \citet{Vicente-Serrano2018}, de que el
+\hypertarget{drought-trend-and-attribution-to-lulcc}{%
+\subsection{Drought trend and attribution to
+LULCC}\label{drought-trend-and-attribution-to-lulcc}}
+
+\citep{Vicente-Serrano2021}, in a study at the global scale of drought
+trends, indicates that there have not been significant trends in
+meteorological drought since 1950. Also, state that the increase in
+hidrological trend in some parts of the globe (northeast Brazil and the
+Mediterranean region) is related to changes in land cover and
+specifically to the rapidly increasing irrigated area, which
+consequently increases water extraction. \citep{Kogan2020} analyzed the
+agricultural drought impact globally and in the main grain producer
+countries, finding that \emph{``since 1980, the Earth warming has not
+changed the drought area or intensity''}. In our study, we considered
+the variation in vegetation productivity in Chile for areas without
+changes in land cover macroclasses (see
+Section~\ref{sec-persistence_mask}), to avoid misleading conclusions
+that could be related to the increase in water demand due to LULCC. Our
+results show a contrasting perspective. There has been a significant
+trend in the decline of vegetation productivity (zcNDVI) since 2000 for
+``Norte Chico'' and ``Centro,'' which has been extreme between 2020 and
+2022, seemsly due to an intense hydrological drought due to the
+persistance of the Mega Drought \citep{Garreaud2017}. Despite we used
+the persistance mask for vegetations's trend analysis, the cropland,
+which is the most water-demand type, showed a decrease trend in ``Norte
+Chico'' and ``Centro.'' Also, there was an increase in barren land for
+both types. These changes are associated with a decrease in water demand
+from vegetation. On the other hand, for ``Norte Chico'' and ``Centro,''
+our results show a decrease in trends of water supply (SPI and SSI),
+which are higher at larger time scales and consequently impact the
+hydrological system. We claim that what occurred in central Chile defies
+findings made at the global level
+\citep{Vicente-Serrano2021, Kogan2020}, demonstrating that a constant
+decrease in water supply rather than an increase in water demand (i.e.,
+irrigated crops) is the main cause of the hydrological drought. Finally,
+central Chile has shown a diminishment in vegetation productivity for
+all macroclasses, mainly attributed to variation in water supply, i.e.,
+precipitation, which could be strengthened by an increase in water
+demand by, for example, irrigated crops.
+
+1.- Respecto a lo que indica \citep{Vicente-Serrano2022}, de que el
 aumento en la tendencia en severidad de la sequía (hidrológica) tiene
@@
 
-\hypertarget{land-cover-types-most-impacted-by-drought-throughout-chile}{%
-\subsection{land cover types most impacted by drought throughout
-Chile}\label{land-cover-types-most-impacted-by-drought-throughout-chile}}
+\citep{Chen2022}: Assessing the impact of drought-land cover change on
+global vegetation greenness and productivity
+
+\citep{Kogan2020}: muestra que la sequía no se ha intensificado ni
+expandido a nivel global, por 38 años.
+
+\hypertarget{land-cover-types-and-their-impact-by-drought}{%
+\subsection{Land cover types and their impact by
+drought}\label{land-cover-types-and-their-impact-by-drought}}
+
+Extreme drought conditions are an important driver of tree mortality, as
+shown by \citep{Senf2020} in Europe. However, we found that forest is
+the type of landcover macroclass less affected by variation in drought
+indices, being the most resistant landcover class to drought. Supporting
+this is \citep{Fathi-Taperasht2022}, who asserts that Indian forests are
+the most drought-resistant and recover rapidly. Similarly, the work of
+\citep{Wu2024} who analyzed vegetation loss and recovery in response to
+meteorological drought in the humid subtropical Pearl River basin in
+China, indicates that forests showed higher drought resistance.
+\citep{Xiao2023} using Vegetation Optical Depth (VOD), kNDVI, and EVI,
+evaluates ecosystem resistance, showing that, compared to croplands,
+ecosystems with a higher proportion of forests are more resilient to
+severe droughts. They attribute the difference to a deeper rooting depth
+of trees, a higher water storage capacity, and different water use
+strategies between them \citep{Xiao2023}.
 
@@
 Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur
-(\citet{Miranda2023})
+(\citep{Miranda2023})
+
+\citep{Fathi-Taperasht2022}: muestra que bosques son más resilentes a la
+sequía en India.
+
+\citep{Wu2024}: evalua la perdida de vegetación en diferentes tipos de
+uso de suelo. Muestra que bosques son más resilentes y cultivos y
+pastizales, los menos.
+
+\citep{Jiang2020}: impacto de la sequía en landcover en China.
 

### 2024-01-24 article/paper_drought_Chile2023.tex
actualizar discusion

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, frame hidden, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
 

### 2024-01-25 article/paper_drought_Chile2023.qmd
actualizar

@@
 ---
-title: Drought indices of water demand and supply, soil moisture, vegetation, and their impact on LULCC in continental Chile
+title: Drought indices of water demand and supply and their impact on LULCC in Chile from ERA5-Land and MODIS
 #subtitle: A Short Subtitle
@@
 Supporting this is @Fathi-Taperasht2022, who asserts that Indian forests are the most drought-resistant and recover rapidly.
-Similarly, the work of @Wu2024 who analyzed vegetation loss and recovery in response to meteorological drought in the humid subtropical Pearl River basin in China, indicates that forests showed higher drought resistance.
-@Xiao2023 using Vegetation Optical Depth (VOD), kNDVI, and EVI, evaluates ecosystem resistance, showing that, compared to croplands, ecosystems with a higher proportion of forests are more resilient to severe droughts.
-They attribute the difference to a deeper rooting depth of trees, a higher water storage capacity, and different water use strategies between them [@Xiao2023].
+Similarly, the work of @Wu2024, who analyzed vegetation loss and recovery in response to meteorological drought in the humid subtropical Pearl River basin in China, indicates that forests showed higher drought resistance.
+@Xiao2023, using Vegetation Optical Depth (VOD), kNDVI, and EVI, evaluates ecosystem resistance, showing that, compared to croplands, ecosystems with a higher proportion of forests are more resilient to severe droughts.
+They attribute the difference to a deeper rooting depth of trees, a higher water storage capacity, and different water use strategies between forest and cropland [@Xiao2023].
+Despite our insights, @Venegas2022, who studied species of sclerophyllous populations, indicates that the trees experienced a general decline in growth, and the natural dynamics of their forests may have been altered, mainly due to the cumulative effects of the unprecedented drought (i.e., hydrological drought).
+Nonetheless, the species of sclerophyllous forest were not the predominant surface within the forest macroclass used in this study.
+(REVISAR ESTO, LA ultima parte lo que pasó en el bvosque esclerofilo fue el 2020, año extremo)
 

### 2024-01-25 article/paper_drought_Chile2023.tex
actualizar

@@
 \hypersetup{
-  pdftitle={Drought indices of water demand and supply, soil moisture, vegetation, and their impact on LULCC in continental Chile},
+  pdftitle={Drought indices of water demand and supply and their impact on LULCC in Chile from ERA5-Land and MODIS},
   pdfauthor={Francisco Zambrano},
@@
 \begin{frontmatter}
-\title{Drought indices of water demand and supply, soil moisture,
-vegetation, and their impact on LULCC in continental Chile}
+\title{Drought indices of water demand and supply and their impact on
+LULCC in Chile from ERA5-Land and MODIS}
 \author[1]{Francisco Zambrano%
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, frame hidden, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 the most drought-resistant and recover rapidly. Similarly, the work of
-\citep{Wu2024} who analyzed vegetation loss and recovery in response to
+\citep{Wu2024}, who analyzed vegetation loss and recovery in response to
 meteorological drought in the humid subtropical Pearl River basin in
 China, indicates that forests showed higher drought resistance.
-\citep{Xiao2023} using Vegetation Optical Depth (VOD), kNDVI, and EVI,
+\citep{Xiao2023}, using Vegetation Optical Depth (VOD), kNDVI, and EVI,
 evaluates ecosystem resistance, showing that, compared to croplands,
@@
 of trees, a higher water storage capacity, and different water use
-strategies between them \citep{Xiao2023}.
+strategies between forest and cropland \citep{Xiao2023}. Despite our
+insights, \citep{Venegas2022}, who studied sclerophyllous populations,
+indicates that the trees experienced a general decline in growth, and
+the natural dynamics of their forests may have been altered, mainly due
+to the cumulative effects of the unprecedented drought (i.e.,
+hydrological drought). Nonetheless, sclerophyllous forest is not the
+predominant surface within the forest macroclass used in this study.
 

### 2024-01-25 article/paper_drought_Chile2023.qmd
actualizar jueves

@@
 @Kogan2020 analyzed the agricultural drought impact globally and in the main grain producer countries, finding that _"since 1980, the Earth warming has not changed the drought area or intensity"_.
+
 In our study, we considered the variation in vegetation productivity in Chile for areas without changes in land cover macroclasses (see @sec-persistence_mask), to avoid misleading conclusions that could be related to the increase in water demand due to LULCC.
@@
 There has been a significant trend in the decline of vegetation productivity (zcNDVI) since 2000 for "Norte Chico" and "Centro," which has been extreme between 2020 and 2022, seemsly due to an intense hydrological drought due to the persistance of the Mega Drought [@Garreaud2017].
-Despite we used the persistance mask for vegetations's trend analysis, the cropland, which is the most water-demand type, showed a decrease trend in "Norte Chico" and "Centro." Also, there was an increase in barren land for both types.
+Despite using the persistance mask for vegetation's trend analysis, cropland, which is the most water-demand type, showed a decrease trend in "Norte Chico" and "Centro." Also, there was an increase in barren land for both types.
 These changes are associated with a decrease in water demand from vegetation.
+Nonetheless, we used the persistant landcover to ensure that the pixel has the same class; in the case of croplands, it could happen that some areas had changed crops for others with higher water consumption.
+But this effect should be minor compared to the results from ladcover macroclasses.
+
 On the other hand, for "Norte Chico" and "Centro," our results show a decrease in trends of water supply (SPI and SSI), which are higher at larger time scales and consequently impact the hydrological system.
 We claim that what occurred in central Chile defies findings made at the global level [@Vicente-Serrano2021;@Kogan2020], demonstrating that a constant decrease in water supply rather than an increase in water demand (i.e., irrigated crops) is the main cause of the hydrological drought.
-Finally, central Chile has shown a diminishment in vegetation productivity for all macroclasses, mainly attributed to variation in water supply, i.e., precipitation, which could be strengthened by an increase in water demand by, for example, irrigated crops.
-
-1.-  Respecto a lo que indica @Vicente-Serrano2022, de que el aumento en la tendencia en severidad de la sequía (hidrológica) tiene que ver más con un aumento de la demanda de agua (ej, LULCC, amazonas) que a una tendencia en las condiciones climáticas (SPI-12).
-¿Qué pasa en Chile?
-
-@Chen2022: Assessing the impact of drought-land cover change on global vegetation greenness and productivity
-
-@Kogan2020: muestra que la sequía no se ha intensificado ni expandido a nivel global, por 38 años.
+Finally, central Chile has shown a diminishment in vegetation productivity for all macroclasses, mainly attributed to variation in water supply, i.e., precipitation, which could be strengthened by an increase in water demand by, for example, an increase in the surface area of irrigated crops.
 
 ## Land cover types and their impact by drought
+
+We found that shrubland, savannas (Chilean matorral), and croplands are the most sensitive to climate conditions.
+Being most affected by the 12-month soil moisture deficit.
+In a study in the Yangtze River Basin in China, Jiang2020 analyzed the impact of drought on vegetation using the SPEI and the Enahanced Vegetation Index (EVI).
+They found that cropland was more sensitive to drought than cropland, showing that cropland responds strongly to short- and medium-term drought (< SPEI-6).
+In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro." In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016; @Potopova2015; @Dai2020].
+Short-term precipitation deficits impact soil water, and thus less water is available for plant growth.
+However, we found that in "Norte Chico," an SPI-36 and SPEI-12 had a higher impact, which are associated with hydrological drought (long-term), and in "Centro," an SPI-12 and SPEI-12.
+Thus, we attribute this impact to the hydrological drought that has decreased groundwater storage [@Taucare2024], which in turn is impacted by long-term deficits, and consequently, the vegetation is more dependent on groundwater.
+In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared (\ref{tab-corland cover}).
+What can be explained is that, south of "Centro," predominate forest and grassland, the most resistant types.
+Also, drought episodes have been less frequent and intense.
+The drought episodes have had a lower impact on water availability for vegetation.
 
@@
 Similarly, the work of @Wu2024, who analyzed vegetation loss and recovery in response to meteorological drought in the humid subtropical Pearl River basin in China, indicates that forests showed higher drought resistance.
-@Xiao2023, using Vegetation Optical Depth (VOD), kNDVI, and EVI, evaluates ecosystem resistance, showing that, compared to croplands, ecosystems with a higher proportion of forests are more resilient to severe droughts.
+Using Vegetation Optical Depth (VOD), kNDVI, and EVI, @Xiao2023, tests the resistance of ecosystems and finds that ecosystems with more forests are better able to handle severe droughts than croplands.
 They attribute the difference to a deeper rooting depth of trees, a higher water storage capacity, and different water use strategies between forest and cropland [@Xiao2023].
-Despite our insights, @Venegas2022, who studied species of sclerophyllous populations, indicates that the trees experienced a general decline in growth, and the natural dynamics of their forests may have been altered, mainly due to the cumulative effects of the unprecedented drought (i.e., hydrological drought).
-Nonetheless, the species of sclerophyllous forest were not the predominant surface within the forest macroclass used in this study.
-(REVISAR ESTO, LA ultima parte lo que pasó en el bvosque esclerofilo fue el 2020, año extremo)
 
-2.- Sobre los tipos de land cover más afectados por los indicadores de sequía.
-Asociación con el matorral chileno [@Fuentes2021].
-Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur (@Miranda2023)
+In contrast to our observations, @Venegas2022, who investigated Cryptocarya alba and Beilschmiedia miersii (both members of the Lauraceae family) inhabiting sclerophyllous forests in Chile, concluded that the overall development of the trees had decelerated, potentially indicating alterations in the inherent dynamics of their respective forests.
+They attributed it to the cumulative effects of the unprecedented drought (i.e., hydrological drought).
+Nonetheless, the trees within sclerophyllous forests are highly heterogeneous, and the species analyzed by @Venegas2022 are not the predominant ones.
+Because of this, when we used MOD13A3 with a 1km spatial resolution to measure vegetation condition, it took the average condition of 1 square kilometer.
+Thus, we attribute that forest to being the most resistant to drought, due to the fact that most of the species comprising it are highly resilient to water scarcity.
+Then, for a study of the dynamics of forest to drought, e.g., sclerophyllous forest, on a local scale using remote sensing, we should use operational products at higher spatial resolutions as provided by the missions Landsat or Sentinel that could capture the heterogenity of the forest type.
 
-@Fathi-Taperasht2022: muestra que bosques son más resilentes a la sequía en India.
-
-@Wu2024: evalua la perdida de vegetación en diferentes tipos de uso de suelo. Muestra que bosques son más resilentes y cultivos y pastizales, los menos.
-
-@Jiang2020: impacto de la sequía en landcover en China.
+## ERA5-Land derived soil moisture and vegetation productivity
 

### 2024-01-25 article/paper_drought_Chile2023.tex
actualizar jueves

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 countries, finding that \emph{``since 1980, the Earth warming has not
-changed the drought area or intensity''}. In our study, we considered
-the variation in vegetation productivity in Chile for areas without
-changes in land cover macroclasses (see
+changed the drought area or intensity''}.
+
+In our study, we considered the variation in vegetation productivity in
+Chile for areas without changes in land cover macroclasses (see
 Section~\ref{sec-persistence_mask}), to avoid misleading conclusions
@@
 2022, seemsly due to an intense hydrological drought due to the
-persistance of the Mega Drought \citep{Garreaud2017}. Despite we used
-the persistance mask for vegetations's trend analysis, the cropland,
-which is the most water-demand type, showed a decrease trend in ``Norte
-Chico'' and ``Centro.'' Also, there was an increase in barren land for
-both types. These changes are associated with a decrease in water demand
-from vegetation. On the other hand, for ``Norte Chico'' and ``Centro,''
-our results show a decrease in trends of water supply (SPI and SSI),
-which are higher at larger time scales and consequently impact the
-hydrological system. We claim that what occurred in central Chile defies
-findings made at the global level
-\citep{Vicente-Serrano2021, Kogan2020}, demonstrating that a constant
-decrease in water supply rather than an increase in water demand (i.e.,
-irrigated crops) is the main cause of the hydrological drought. Finally,
-central Chile has shown a diminishment in vegetation productivity for
-all macroclasses, mainly attributed to variation in water supply, i.e.,
-precipitation, which could be strengthened by an increase in water
-demand by, for example, irrigated crops.
-
-1.- Respecto a lo que indica \citep{Vicente-Serrano2022}, de que el
-aumento en la tendencia en severidad de la sequía (hidrológica) tiene
-que ver más con un aumento de la demanda de agua (ej, LULCC, amazonas)
-que a una tendencia en las condiciones climáticas (SPI-12). ¿Qué pasa en
-Chile?
-
-\citep{Chen2022}: Assessing the impact of drought-land cover change on
-global vegetation greenness and productivity
-
-\citep{Kogan2020}: muestra que la sequía no se ha intensificado ni
-expandido a nivel global, por 38 años.
+persistance of the Mega Drought \citep{Garreaud2017}. Despite using the
+persistance mask for vegetation's trend analysis, cropland, which is the
+most water-demand type, showed a decrease trend in ``Norte Chico'' and
+``Centro.'' Also, there was an increase in barren land for both types.
+These changes are associated with a decrease in water demand from
+vegetation. Nonetheless, we used the persistant landcover to ensure that
+the pixel has the same class; in the case of croplands, it could happen
+that some areas had changed crops for others with higher water
+consumption. But this effect should be minor compared to the results
+from ladcover macroclasses.
+
+On the other hand, for ``Norte Chico'' and ``Centro,'' our results show
+a decrease in trends of water supply (SPI and SSI), which are higher at
+larger time scales and consequently impact the hydrological system. We
+claim that what occurred in central Chile defies findings made at the
+global level \citep{Vicente-Serrano2021, Kogan2020}, demonstrating that
+a constant decrease in water supply rather than an increase in water
+demand (i.e., irrigated crops) is the main cause of the hydrological
+drought. Finally, central Chile has shown a diminishment in vegetation
+productivity for all macroclasses, mainly attributed to variation in
+water supply, i.e., precipitation, which could be strengthened by an
+increase in water demand by, for example, an increase in the surface
+area of irrigated crops.
 
@@
 drought}\label{land-cover-types-and-their-impact-by-drought}}
+
+We found that shrubland, savannas (Chilean matorral), and croplands are
+the most sensitive to climate conditions. Being most affected by the
+12-month soil moisture deficit. In a study in the Yangtze River Basin in
+China, Jiang2020 analyzed the impact of drought on vegetation using the
+SPEI and the Enahanced Vegetation Index (EVI). They found that cropland
+was more sensitive to drought than cropland, showing that cropland
+responds strongly to short- and medium-term drought (\textless{}
+SPEI-6). In our case, the SPEI-12 was the one that most impacted the
+croplands in ``Norte Chico'' and ``Centro.'' In general, most studies
+show that croplands are most sensitive to short-term drought
+(\textless{} SPI-6) \citep{Zambrano2016, Potopova2015, Dai2020}.
+Short-term precipitation deficits impact soil water, and thus less water
+is available for plant growth. However, we found that in ``Norte
+Chico,'' an SPI-36 and SPEI-12 had a higher impact, which are associated
+with hydrological drought (long-term), and in ``Centro,'' an SPI-12 and
+SPEI-12. Thus, we attribute this impact to the hydrological drought that
+has decreased groundwater storage \citep{Taucare2024}, which in turn is
+impacted by long-term deficits, and consequently, the vegetation is more
+dependent on groundwater. In ``Sur'' and ``Austral,'' the correlations
+between drought indices and vegetation productivity decrease, as do the
+time scales that reach the maximum r-squared (\ref{tab-corland cover}).
+What can be explained is that, south of ``Centro,'' predominate forest
+and grassland, the most resistant types. Also, drought episodes have
+been less frequent and intense. The drought episodes have had a lower
+impact on water availability for vegetation.
 
@@
 meteorological drought in the humid subtropical Pearl River basin in
-China, indicates that forests showed higher drought resistance.
-\citep{Xiao2023}, using Vegetation Optical Depth (VOD), kNDVI, and EVI,
-evaluates ecosystem resistance, showing that, compared to croplands,
-ecosystems with a higher proportion of forests are more resilient to
-severe droughts. They attribute the difference to a deeper rooting depth
-of trees, a higher water storage capacity, and different water use
-strategies between forest and cropland \citep{Xiao2023}. Despite our
-insights, \citep{Venegas2022}, who studied sclerophyllous populations,
-indicates that the trees experienced a general decline in growth, and
-the natural dynamics of their forests may have been altered, mainly due
-to the cumulative effects of the unprecedented drought (i.e.,
-hydrological drought). Nonetheless, sclerophyllous forest is not the
-predominant surface within the forest macroclass used in this study.
-
-2.- Sobre los tipos de land cover más afectados por los indicadores de
-sequía. Asociación con el matorral chileno \citep{Fuentes2021}.
-Diferencia entre el Norte Chico, Centro y lo que pasa hacía el sur
-(\citep{Miranda2023})
-
-\citep{Fathi-Taperasht2022}: muestra que bosques son más resilentes a la
-sequía en India.
-
-\citep{Wu2024}: evalua la perdida de vegetación en diferentes tipos de
-uso de suelo. Muestra que bosques son más resilentes y cultivos y
-pastizales, los menos.
-
-\citep{Jiang2020}: impacto de la sequía en landcover en China.
+China, indicates that forests showed higher drought resistance. Using
+Vegetation Optical Depth (VOD), kNDVI, and EVI, \citep{Xiao2023}, tests
+the resistance of ecosystems and finds that ecosystems with more forests
+are better able to handle severe droughts than croplands. They attribute
+the difference to a deeper rooting depth of trees, a higher water
+storage capacity, and different water use strategies between forest and
+cropland \citep{Xiao2023}.
+
+In contrast to our observations, \citep{Venegas2022}, who investigated
+Cryptocarya alba and Beilschmiedia miersii (both members of the
+Lauraceae family) inhabiting sclerophyllous forests in Chile, concluded
+that the overall development of the trees had decelerated, potentially
+indicating alterations in the inherent dynamics of their respective
+forests. They attributed it to the cumulative effects of the
+unprecedented drought (i.e., hydrological drought). Nonetheless, the
+trees within sclerophyllous forests are highly heterogeneous, and the
+species analyzed by \citep{Venegas2022} are not the predominant ones.
+Because of this, when we used MOD13A3 with a 1km spatial resolution to
+measure vegetation condition, it took the average condition of 1 square
+kilometer. Thus, we attribute that forest to being the most resistant to
+drought, due to the fact that most of the species comprising it are
+highly resilient to water scarcity. Then, for a study of the dynamics of
+forest to drought, e.g., sclerophyllous forest, on a local scale using
+remote sensing, we should use operational products at higher spatial
+resolutions as provided by the missions Landsat or Sentinel that could
+capture the heterogenity of the forest type.
 

### 2024-01-26 article/paper_drought_Chile2023.qmd
actualizar repo

@@
 They attributed it to the cumulative effects of the unprecedented drought (i.e., hydrological drought).
-Nonetheless, the trees within sclerophyllous forests are highly heterogeneous, and the species analyzed by @Venegas2022 are not the predominant ones.
-Because of this, when we used MOD13A3 with a 1km spatial resolution to measure vegetation condition, it took the average condition of 1 square kilometer.
 Thus, we attribute that forest to being the most resistant to drought, due to the fact that most of the species comprising it are highly resilient to water scarcity.
-Then, for a study of the dynamics of forest to drought, e.g., sclerophyllous forest, on a local scale using remote sensing, we should use operational products at higher spatial resolutions as provided by the missions Landsat or Sentinel that could capture the heterogenity of the forest type.
+compared to the other landcover classes.
+Also, when we used MOD13A3 with a 1km spatial resolution to measure vegetation condition, it took the average condition of 1 square kilometer.
+Then, to study how a type of forest (sclerophyllous forest) changes in response to drought on a local level using remote sensing, we should use operational products with higher spatial resolutions, such as those from Landsat or Sentinel, which can show how different the forests are.
 
-## ERA5-Land derived soil moisture and vegetation productivity
+## Soil moisture, vegetation productivity, and agricultural drought.
+
+The main external factors that affect biomass production by vegetation are ET and SM, and the rate of ET in turn depends on the availability of water storage in the root zone.
+Thus, soil moisture plays a key role in land carbon uptake and, consequently, in the production of biomass [@Humphrey2021].
+Moreover, @Zhang2022 indicates there is a bidirectional causality between soil moisture and vegetation productivity.
+For this, lastly, some studies have redefined agricultural drought as soil moisture drought from the hydrological perspective [@Loon2016;@Samaniego2018].
+Even though soil moisture is the external factor most determinant of vegetation biomass, there are multiple internal factors, such as species, physiological characteristics, and plant hydraulics, that would affect vegetation productivity.
+Because of that, we believe that agricultural drought, referring to the drought that impacts vegetation productivity, is the most proper term, as originally defined by @Wilhite1985.
+
+Our results showed that the drought index derived from soil moisture (SSI) was the most effective in explaining vegetation productivity within landcover macroclasses.
+However, there is strong spatial variability throughout Chile and between classes, mainly attributable to climate heterogeneity, hydrological status, or vegetation resistance to water scarcity.
+The semi-arid "Norte Chico" and the Mediterranean "Centro" were where SSI had the best performance.
+Medium-term deficits of 12 months are more relevant in the response of vegetation, which decreases to the south, and in the case of croplands, they seem to react in a shorter time, with six months (SSI-6) in "Centro."
+
+@Chatterjee2022
 
@@
   - EL SSI se ve como un poderoso indicador que explica la variabilidad en la productividad de la vegetación.
+
+## Early drought forecasting
 

### 2024-01-26 article/paper_drought_Chile2023.tex
actualizar repo

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 
+\hypertarget{soil-moisture-vegetation-productivity-and-agricultural-drought.}{%
+\subsection{Soil moisture, vegetation productivity, and agricultural
+drought.}\label{soil-moisture-vegetation-productivity-and-agricultural-drought.}}
+
+The main external factors that affect biomass production by vegetation
+are ET and SM, and the rate of ET in turn depends on the availability of
+water storage in the root zone. Thus, soil moisture plays a key role in
+land carbon uptake and, consequently, in the production of biomass
+\citep{Humphrey2021}. Moreover, \citep{Zhang2022} indicates there is a
+bidirectional causality between soil moisture and vegetation
+productivity. For this, lastly, are some studies that have redefined
+agricultural drought as soil moisture drought from the hydrological
+perspective \citep{Loon2016, Samaniego2018}. Even though soil moisture
+is the external factor most determinant of vegetation biomass, there are
+multiple internal factors, such as species, physiological
+characteristics, and plant hydraulics, that would affect vegetation
+productivity. Because of that, we believe that agricultural drought,
+referring to the drought that impacts vegetation productivity, is the
+most proper term, as originally defined by \citep{Wilhite1985}.
+
+Our results showed that the drought index derived from soil moisture
+(SSI) was the most effective in explaining vegetation productivity
+within landcover macroclasses. However, there is strong spatial
+variability throughout Chile, mainly attributable to the climate
+heterogeneity.
+
+\citep{Chatterjee2022}
+
 \hypertarget{drought-indices-of-water-demand-and-supply-soil-moisture-to-predict-changes-in-vegetation-productivity}{%
@@
 
+\hypertarget{early-drought-forecasting}{%
+\subsection{Early drought forecasting}\label{early-drought-forecasting}}
+
 \hypertarget{future-outlook}{%

### 2024-01-26 article/paper_drought_Chile2023.tex
actualizar repo

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 forests. They attributed it to the cumulative effects of the
-unprecedented drought (i.e., hydrological drought). Nonetheless, the
-trees within sclerophyllous forests are highly heterogeneous, and the
-species analyzed by \citep{Venegas2022} are not the predominant ones.
-Because of this, when we used MOD13A3 with a 1km spatial resolution to
-measure vegetation condition, it took the average condition of 1 square
-kilometer. Thus, we attribute that forest to being the most resistant to
-drought, due to the fact that most of the species comprising it are
-highly resilient to water scarcity. Then, for a study of the dynamics of
-forest to drought, e.g., sclerophyllous forest, on a local scale using
-remote sensing, we should use operational products at higher spatial
-resolutions as provided by the missions Landsat or Sentinel that could
-capture the heterogenity of the forest type.
+unprecedented drought (i.e., hydrological drought). Thus, we attribute
+that forest to being the most resistant to drought, due to the fact that
+most of the species comprising it are highly resilient to water
+scarcity. compared to the other landcover classes. Also, when we used
+MOD13A3 with a 1km spatial resolution to measure vegetation condition,
+it took the average condition of 1 square kilometer. Then, to study how
+a type of forest (sclerophyllous forest) changes in response to drought
+on a local level using remote sensing, we should use operational
+products with higher spatial resolutions, such as those from Landsat or
+Sentinel, which can show how different the forests are.
 
@@
 bidirectional causality between soil moisture and vegetation
-productivity. For this, lastly, are some studies that have redefined
-agricultural drought as soil moisture drought from the hydrological
-perspective \citep{Loon2016, Samaniego2018}. Even though soil moisture
-is the external factor most determinant of vegetation biomass, there are
+productivity. For this, lastly, some studies have redefined agricultural
+drought as soil moisture drought from the hydrological perspective
+\citep{Loon2016, Samaniego2018}. Even though soil moisture is the
+external factor most determinant of vegetation biomass, there are
 multiple internal factors, such as species, physiological
@@
 within landcover macroclasses. However, there is strong spatial
-variability throughout Chile, mainly attributable to the climate
-heterogeneity.
+variability throughout Chile and between classes, mainly attributable to
+climate heterogeneity, hydrological status, or vegetation resistance to
+water scarcity. The semi-arid ``Norte Chico'' and the Mediterranean
+``Centro'' were where SSI had the best performance. Medium-term deficits
+of 12 months are more relevant in the response of vegetation, which
+decreases to the south, and in the case of croplands, they seem to react
+in a shorter time, with six months (SSI-6) in ``Centro.''
 

### 2024-03-12 article/paper_drought_Chile2023.qmd
update paper qm intro

@@
 ---
-title: Drought indices of water demand and supply and their impact on LULCC in Chile from ERA5-Land and MODIS
+title: Drought, vegetation productivity, and land cover change in Chile
 #subtitle: A Short Subtitle
@@
     note: This is the first author footnote.
-  # - name: Bob Security
-  #   email: bob@example.com
-  #   affiliations:
-  #       - id: another-u
-  #         name: Another University
-  #         department: Department Name
-  #         address: Street Address
-  #         city: City
-  #         state: State
-  #         postal-code: Postal Code
+  - name: Anton Vrieling
+    email: a.vrieling@utwente.nl
+    affiliations:
+        - id: itc
+          name: University of Twente
+          department: Faculty of Geo-Information Science and Earth Observation (ITC)
+          #address: Street Address
+          city:  Enschede, The Netherlands
+  - name: Francisco Fernández
+    email: a.vrieling@utwente.nl
+    affiliations:
+        - id: uss
+          name: Universidad San Sebastian
+          department: 
+          #address: Street Address
+          city:  
+  - name: Francisco Meza
+  - name: Alejandro Venegas-González
+  - name: Iongel Duran-Llacer
+  - name: Nicolas Raab
+  - name: Dylan Craven        
   #   note: |
@@
   Central Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development.
-This persistent period of water scarcity has been defined as a “Mega Drought".
-Our objective is to examine the effects of drought on LULCC (land use land cover change) over continental Chile using drought indices of water supply and demand, soil moisture, and vegetation productivity.
-For the analysis, continental Chile was divided into five zones according to a latitudinal gradient: “Norte Grande," “Norte Chico," "Centro," "Sur," and "Austral." The monthly ERA5-Land (ERA5L) variables for precipitation, temperature, and soil moisture were used.
+This persistent period of water scarcity has been defined as a megadrought.
+Our objective is to assess the impact of drought on LULCC (land use land cover change) over continental Chile using drought indices of water supply and demand, soil moisture, and their impact on vegetation productivity.
+The monthly ERA5-Land (ERA5L) variables for precipitation, temperature, and soil moisture were used.
 From 2001 to 2022, we used the land cover MODIS product MCD12Q1, and from 2000 to 2023, we used the NDVI (Normalized Difference Vegetation Index) product MOD13A3 collection 6.1.
-We estimated atmospheric evaporative demand (AED) using the Hargreaves-Samani equation with the ERA5L temperature.
-We used the Standardized Precipitation Index (SPI), the Standardized Precipitation Evapotranspiration Index (SPEI), the Evaporative Demand Drought Index (EDDI), the Standardized Soil Moisture Index (SSI), and the Standardized anomaly of cumulative NDVI (zcNDVI) as drought indicators.
+As drought indices, we compute the standardized anomaly of cumulative NDVI (zcNDVI), the Standardized Precipitation Evapotranspiration Index (SPEI), the Evaporative Demand Drought Index (EDDI), and the Standardized Soil Moisture Index (SSI).
 These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was for 6 months.
@@
 Also, we analyzed the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain insights into the impact of water supply and demand on vegetation productivity.
-Our results showed that LULCC were highest in "Centro," "Sur," and "Austral,” with 36%, 31%, and 34%, respectively.
+Our results showed that LULCC was highest in "Centro," "Sur," and "Austral,” with 36%, 31%, and 34% of change in the surface type, respectively.
 The EDDI shows that water demand has increased for all zones, with a major increase in "Norte Grande." The drought indices of water supply and soil moisture evidence a decreasing trend, which decreases at longer time scales, from "Norte Grande" to "Sur." "Austral" is the only zone that shows an increase in supply.
-Vegetation productivity measures by zcNDVI present a negative trend in "Norte Chico" and "Centro." Showing to be the zones most impacted by climatic conditions, the years 2020 and 2022 suffered the most extreme drought.
-On the other hand, forests seem to be the most resistant to drought.
+Vegetation productivity measures by zcNDVI present a negative trend in "Norte Chico" and "Centro." On the other hand, forests seem to be the most resistant to drought.
 The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
-The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are SSI-12, followed by SPEI-24 and SPEI-12 in "Norte Chico" and "Centro." The results indicate that “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits longer than a year, potentially explained by a low capacity of water storage in those zones that should be further investigated.
+The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are SSI-12, followed by SPEI-24 and SPEI-12 in "Norte Chico" and "Centro." The results indicate that “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits lasting longer than a year.
+Our results can help develop a robust vegetation productivity forecasting model for land cover classes in Chile.
 keywords: 
@@
 
-The sixth assessment report (AR6) of the IPCC [@IPCC2023] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes, and the evidence has been strengthened since AR5 [@IPCC2013].
-There is a high degree of confidence that rising temperatures will increase the land area where droughts will occur more frequently and with greater severity [@IPCCCH112021].
-Furthermore, drought increases tree mortality and triggers changes in land cover and, consequently, land use, thus impacting ecosystems [@Crausbay2017].
-Nevertheless, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.
+Drought is often classified as meteorological when there is a decrease in precipitation below the mean average of several years (more than 30 years), hydrological when these anomalies last for long periods (months to years) and affect water systems, and agricultural when the deficit impacts plant health anomalies and leads to decreased productivity [@Wilhite1985].
+However, it is important to note that drought is also influenced by human activities, which were not considered in the definitions.
+Thus, @Loon2016 and @AghaKouchak2021 have given an updated definition of drought for the Anthropocene, suggesting that it should be considered the feedback of humans’ decisions and activities that drives the anthropogenic drought.
+Simultaneously, drought leads to heightened tree mortality and induces alterations in land cover and land use, ultimately affecting ecosystems [@Crausbay2017].
+Even though many ecological studies have misinterpreted how to characterize drought, for example, sometimes considering "dry" conditions as "drought" [@Slette2019].
+Then, @Crausbay2017 proposed the ecological drought definition as "an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems." In light of current global warming, it is crucial to study the interaction between drought and ecosystems in order to understand their feedback and impact on water security.
+[@Bakker2012]
 
-The primary cause of drought is precipitation, and temperature makes it worse [@Luo2017].
-Drought impacts soil moisture, hydrological regimes, and vegetation productivity.
-Initially, drought was commonly classified as meteorological, hydrological, and agricultural [@Wilhite1985].
-Lately, @Loon2016 and @AghaKouchak2021 have given an updated definition of drought for the Anthropocene, suggesting that it should be considered the feedback of humans' decisions and activities that drives the anthropogenic drought.
-Even though it has been argued that those definitions do not fully address the ecological dimensions of drought, @Crausbay2017 proposed the ecological drought definition as "an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems.".
-Moreover, many ecological studies have misinterpreted how to characterize drought, for example, sometimes considering "dry" conditions as "drought" [@Slette2019].
-On the other hand, the AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C.
-Then, there is a challenge in conducting drought research, especially to evaluate its impact on ecosystems.
+Human-induced greenhouse gas emissions have increased the frequency and/or intensity of drought as a result of global warming, according to the sixth assessment report (AR6) of the Intergovernmental Panel on Climate Change (IPCC) [@IPCC2023].
+The evidence supporting this claim has been strengthened since AR5 [@IPCC2013].
+Recent studies, however, have produced contrasting findings, suggesting that drought has not exhibited a significant trend over the past forty years.
+[@Vicente-Serrano2022; @Kogan2020].
+@Vicente-Serrano2022 analyzed the meteorological drought trend on a global scale, finding that only in a few regions has there been an increase in the severity of drought.
+Moreover, they attribute the increase in droughts over the past forty years solely to an increase in atmospheric evaporative demand (AED), which in turn enhances vegetation water demand, with important implications for agricultural and ecological droughts.
+Also, they state that "the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change".
+Similarly, @Kogan2020 analyzed the drought trend using vegetation health methods, finding that for the globe, hemispheres, and main grain-producing countries, drought has not expanded or intensified for the last 38 years.
+Further, the @IPCC2021 suggests that there is a high degree of confidence that rising temperatures will increase the extent, frequency, and severity of droughts.
+Also, AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C.
+To better evaluate the impact of drought trends on ecosystems, assessments are needed that relate meteorological and soil moisture variables to their effects on vegetation.
 
-Chile has been facing a persistent rainfall deficit for more than a decade [@Garreaud2017], which has impacted vegetation development [@Zambrano2023] and the hydrological system [@Boisier2018].
-Current drought conditions have affected crop productivity [@Zambrano2016; @Zambrano2018], forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], land cover change [@Fuentes2021], water supply in watersheds [@AlvarezGarreton2021], and have had economic impacts [@Fernandez2023].
-In 2019–2020, the drought severity reached an extreme condition in Central Chile (30–34°S) not seen for at least 40 years, and the evidence indicates that the impact is transversal to the land cover classes of forest, grassland, and cropland [@Zambrano2023].
-The prolonged lack of precipitation in Central Chile is producing changes in ecosystem dynamics that must be studied.
+From 1960 to 2019, land use change has impacted around one-third of the Earth's surface, which is four times more than previously thought [@Winkler2021].
+Multiple studies aim to analyze and forecast changes in land cover globally [@Winkler2021; @Song2018] and regionally [@Chamling2020; @Homer2020; @Yang2021].
+Some others seek to analyze the impact of land cover change on climate conditions such as temperature and precipitation [@Luyssaert2014;@Pitman2012].
+There is less research on the interaction between drought and land cover change [@Chen2022;@Akinyemi2021;@Peng2017].
+@Peng2017 conducted a worldwide investigation utilizing net primary production to examine the spatial and temporal variations in vegetation productivity at global level.
+The study aimed to assess the influence of drought by comparing the twelve-month Standardized Precipitation Evapotranspiration Index (SPEI) and land cover change.
+According to their findings, drought is responsible for 37% of the decline in vegetation productivity, while water availability accounts for 55% of the variation.
+@Chen2022 studied the trend of vegetation greenness and productivity and its relation to meteorological drought (SPEI of twelve months in December) and soil moisture at the global level.
+The results showed lower correlations (<0.2) for both variables.
+@Akinyemi2021 evaluates drought trends and land cover change using vegetation indices in Botwsana in a semi-arid climate.
+These studies mostly looked at how changes in land cover and vegetation productivity are related to a single drought index (SPEI) over a single time period of 12 months.
+SPEI takes into account the combined effect of precipitation and AED as a water balance, but it does not allow us to know the contribution of each variable on its own.
+Some things worth investigating in terms of land cover change and vegetation productivity are: i) How do they respond to short- to long-term meteorological droughts?
+ii) How do they behave in humid and arid climatic zones regarding drought?
+And iii) What is the role of soil moisture?
+Likewise, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.
 
-For the spatiotemporal assessment of drought impact (i.e., by water supply and demand) on land cover changes, we need climatic realiable variables such as precipitation, temperature, soil moisture, land cover, and vegetation status.
-For developing countries like Chile, the weather networks present several disadvantages, such as gaps, a short history, and low-quality data.
-Reanalysis data, as the ERA5-Land (ERA5L) [@MunozSabater2021] provides hourly climatic information (precipitation, temperature, and soil moisture) without gaps since 1950 with global extension.
-ERA5L has already been used for drought assessment using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@Nouri2023] and for flash drought [@Wang2023] by analyzing soil moisture and evapotranspiration.
-On the other hand, satellite remote sensing [@West2019; @AghaKouchak2015] is the primary method to evaluate how drought impacts vegetation dynamics.
-The Moderate-Resolution Imaging Spectroradiometer (MODIS) can be used to get vegetation drought indices (VDI), which are often used as proxies for productivity [@Paruelo2016; @Schucknecht2017].
-Besides, land use and land cover (LULC) change can be driven by drought [@Tran2019; @Akinyemi2021].
-To analyze these changes, multiple LULC products exist [@Grekousis2015].
-One of those that provides time series since 2001 is the MCD12Q1 [@Friedl2019] from MODIS.
-The variation in water supply and demand is finally reflected in the total water storage (TWS).
-The Gravity Recovery and Climate Experiment (GRACE), which allows analyzing changes in water availability at coarse resolution, can retrieve the TWS [@Ahmed2014; @Ma2017].
-We can find drought indices of supply (i.e., precipitation) and demand (i.e., temperature) using climatic reanalysis (ERA5L) and vegetation data (MODIS).
-This lets us figure out how drought changes LULC.
-Further, the TWS can be assessed with regard to the changes in water supply and demand to gain insight into the impact on water storage.
+Chile's diverse climatic and ecosystem types (@Beck2023;@Luebert2022) make it an ideal natural laboratory for studying climate and ecosystems.
+Additionally, the country has experienced severe drought conditions that have had significant effects on vegetation and water storage.
+Central Chile faced a persistent precipitation deficit between 2010 and 2022, defined as a megadrought [@Garreaud2017], which has impacted the Chilean ecosystem.
+This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December having values below one standard deviation.
+Some studies have addressed how this drought affects single ecosystems in terms of forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018;@Zambrano2016].
+We found one study regarding land cover and drought in Chile.
+The study by @Fuentes2021  evaluates water scarcity and land cover change in Chile between 29° and 39° of south latitude.
+@Fuentes2021 used the SPEI of one month for evaluating drought, which led to misleading results.
+For example, they did not find a temporal trend in the SPEI but found a decreasing trend in water availability and an increase trend on AED, which in turn should have been capable of being captured with longer time scales of the SPEI.
+The term "megadrought" in Chile is used to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018].
+Hence, it is imperative to assess temporal scales that take into account the cumnulative effect within some years.
+There is little knowledge about the relationship between drought and ecosystem in Chile; thus, it is important to understand in more detail how meteorological and soil moisture droughts influence ecosystem dynamics to inform adaptation options.
 
-To evaluate meteorological drought (i.e., water supply), the World Meteorological Organization (WMO; @WMO2012) recommends the Standardized Precipitation Index (SPI; @Mckee1993), a multiscalar drought index that allows to monitor precipitation deficits from short- to long-term.
-Following the same approach, @Vicente-Serrano2010 incorporates into the SPI the effect of temperature through the use of potential evapotranspiration, thus proposing the SPEI (Standardized Precipitation Evapotranspiration Index).
-Similarly, to evaluate solely the evaporative demand driven by temperature, @Hobbins2016 and @McEvoy2016 came up with the Evaporative Demand Drought Index (EDDI).
-For vegetation, in a similar manner as the SPI, SPEI and EDDI; @Zambrano2018 proposed the zcNDVI, a standardized anomaly of the cumulative Normalized Difference Vegetation Index (NDVI), which could be acumulated over the growing season or any period (e.g., months), resulting in a multiscalar drought index.
-For soil moisture, several drought indices exist, such as the Soil Moisture Deficit Index (SDMI) a normalized index [@Narasimhan2005] and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021] which is a normalized index using vegetation, land surface temperature, and a vegetation condition index (VCI, [@Kogan1995]).
-From TWS, we can estimate the standardized terrestrial water storage index (STI) [@Cui2021], a standardized anomaly that follows the methodology of the SPI, SPEI, EDDI, and zcNDVI.
-Thereby, we have drought indices for water supply, demand, and storage, which can help to make a comprehensive assessment of drought.
-
-In this research, we aim to analyze the impact of drought on different types of ecosystems (land cover classes) in continental Chile.
-Our specific goals are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to assess the LULC change for 2001–2022 and how it relates to drought indices; iii) to evaluate the relationship between zcNDVI, a measure of vegetation productivity, and drought indices for water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.
+A detailed spatiotemporal assessment of the interaction of drought for short- to long-term and land cover change requires information on vegetation as well as weather variables such as precipitation, temperature, and soil moisture.
+Weather networks in Chile present some disadvantages, such as spatio-temporal gaps, a short history, and irregular quality, which make them difficult to represent the whole extent of the country spatially.
+In order to do this, we use reanalysis data from ERA5-Land [@MunozSabater2021] to create drought indices that consider AED, precipitation, and soil moisture over a range of time periods, from the short to the long term.
+Also, we use vegetation spectral information and annual land cover change from the Moderate-Resolution Imaging Spectroradiometer (MODIS).
+We expect to gain insight regarding the temporal evolution of water demand, water supply, and soil moisture, as well as the interaction with land cover change and vegetation productivity.
+Here, we analyze the multi-dimensional impacts of drought across ecosystems in continental Chile.
+More specifically, we aim to assess: i) temporal changes in land-use cover and the direction and magnitude of their relationships with drought indices for water demand and supply, soil moisture, and vegetation productivity; ii) short- to long-term temporal trends in multi-scalar drought indices; and iii) the relationship between vegetation productivity and drought indices for water demand and supply and soil moisture across Chilean ecosystems.
 
@@
 
-Continetal Chile has a diverse climate condition from north to south and east to west [@Aceituno2021] (@fig-studyArea), which determines its great ecosystem diversity (@fig-LCprop).
-The Andes Mountains are a main factor in latitudinal variation [@Garreaud2009].
-To describe the climate and ecosystem of Chile, we use the Koppen-Geiger release by @Beck2023 and the land cover type persistance of 80% of times for 2001–2022, from the IGBP classification scheme [@Friedl2019] (see @sec-methods_lulc).
+Continental Chile has a diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] (@fig-studyArea a), which determines its great ecosystem diversity (@Luebert2022) (@fig-studyArea c).
+The Andes Mountains are a main factor in climate latitudinal variation [@Garreaud2009].
+In order to characterize the climate and ecosystem of Chile, we utilize the Köppen-Geiger classification system developed by @Beck2023 and the land cover data derived from the MODIS product for the period of 2001–2022, based on the International Geosphere-Biosphere Programme (IGBP) classification scheme proposed by @Friedl2019.
 "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures.
 At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk).
-Mainly, the land is barren, with a minor surface of vegetation types such as shrubland and grassland.
-In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
-There is a significant amount (50%) of Chilean matorral (shrubland and savanna, [@Fuentes2021]), then grassland (16%), forest (8%), and croplands (5%), in "Centro." The south part of "Sur" and the north part of "Austral" are dominated by an oceanic climate (Cfb).
-Those zones are high in forest and grassland.
-The southern part of the country has a tundra climate, and in Patagonia, it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.
+In these two northern regions, the land is mostly bare, with a minor surface of vegetation types such as shrubland and grassland.
+In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warm to hot summers (Csa and Csb).
+Land cover in "Centro" comprises a significant amount of shrubland and savanna (50%), grassland (16%), forest (8%), and croplands (5%).
+An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral." Those zones are high in forest and grassland.
+The southern part of the country has a tundra climate, and in "Austral", it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.
 
@@
 (b) Topography reference map.
-(c) Land cover classes for 2022.
+(c) land cover classes for 2022.
 (d) Persistent land cover classes (> 80%) for 2001-2022'
@@
 For the calculation of AED, we used `{SPEI}` [@Bergueria2023].
+For mapping, we use {tmap} [@Tennekes2018].
+For data analysis, the suite {tidyverse} [@Wickham2019] was used.
 
@@
 
-To derive a proxy of vegetation productivity, we used the product MOD13A3 collection 6.1 from MODIS [@Didan2015].
+To derive a proxy for vegetation productivity, we used the product MOD13A3 collection 6.1 from MODIS [@Didan2015].
 It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency.
@@
 For soil moisture, we select a private soil network that is owned by the agricultural enterprise Garces Fruit, which has 99 stations in Central Chile, located in cherry fruit crops.
-We used daily data for the year 2022 and the months outside the growing season, May to September, to avoid the effect of irrigation on soil moisture, which is hardly captured by ERA5L.
+The sensors are installed at 30, 60, and 90m and are the model Teros 12 from MeterGroup.
+To avoid the effect of irrigation on soil moisture, which ERA5L hardly captures, we used daily data for the year 2022 and the months outside the growing season, May to September.
 
@@
 To evaluate water demand, we chose the $EDDI$ [@Hobbins2016; @McEvoy2016] index, which uses the $AED$.
-For supply, we used the index recommended by the World Meteorological Organization (WMO) for monitoring drought, the SPI [@Mckee1993].
+We used the SPI [@Mckee1993], a drought monitoring index that the World Meteorological Organization (WMO) recommends.
 We calculated the SPEI, which used a balance between $P$ and $AED$, in this case, an auxiliary variable $D = P-AED$ is used.
@@
 Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI.
-Before using the NDVI, it was smoothed using a locally-weighted polynomial regression, following the procedure described in @Zambrano2018 and @Zambrano2016.
+Before using the NDVI, it was smoothed using a locally weighted polynomial regression, following the procedure described in @Zambrano2018 and @Zambrano2016.
 
@@
 The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
-Thus, we have a land cover raster time-series with the ten classes for 2001 and 2023.
+Thus, we have a land cover raster time series with the ten classes for 2001 and 2023.
 
@@
 i) resampled LCChile to the spatial resolution (500m) of the land cover macroclasses using the nearest neighbor method,
-ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values as the other two years as prediction
-iii) calculate a confusion matrix with the classes extracted in the 1000 poitns for LCChile, IGBP2013, and IGBP2014. Calculate the performance metrics of accuracy and F1.
+ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values from the other two years as predictions.
+iii) calculate a confusion matrix with the classes extracted from the 1000 points for LCChile, IGBP2013, and IGBP2014. Calculate the performance metrics of accuracy and F1.
 
@@
 
-The time series of NDVI is affected by climatic conditions, vegetation development, seasonality, and changes in vegetation type.
-In this study, we want to analyze the variation in vegetation productivity in different land cover types and how it is affected by water demand, water supply, and soil moisture.
-In order to avoid changes due to a change in the land cover type, that will wrongly impact NDVI.
-We will develop a persistence mask for land cover for 2001–2023.
+Climate, vegetation development, seasonality, and changes in vegetation type all have an impact on the time series of NDVI.
+In this study, we want to examine the variation in vegetation productivity across various land cover types and how water demand, water supply, and soil moisture affect it.
+In order to avoid changes due to a change in the land cover type that will wrongly impact NDVI, we will develop a persistence mask for land cover for 2001–2023.
 Thereby, we reduce an important source of variation on a regional scale.
 
-Thus, we calculated a raster mask for IGBP MODIS considering the macroclasses that remain without change for more than 80% of the years (2001–2022) per pixel, which allows us to identify the areas with no land cover change for the macroclasses.
+Therefore, we generated a raster mask for IGBP MODIS per pixel using macroclasses that remain constant for at least 80% of the years (2001–2022).
+This enables us to identify regions where the land cover of the macroclasses has remained constant.
 
@@
 Then, we will summarize for each land cover class and macrozone.
-Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses having vehetation: forest, cropland, grassland, savanna, and shrubland.
+Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses of vegetation: forest, cropland, grassland, savanna, and shrubland.
 
-To assess how water demand and supply and soil moisture are related to vegetation productivity (zcNDVI), we analyze the linear correlation between the indices SPI, SPEI, EDDI, and SSI for 1, 3, 6, 12, 24, and 36-month time scales against zcNDVI.
-We followed a similar approach to that used by @Meroni2016 when using the SPI for meteorological drought against the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation) as a proxy for vegetation productivity.
-We made a pixel-to-pixel linear correlation analysis per index.
-First, we calculate the Pearson coefficient of correlation for the six time scales and let the time scale that reaches the maximum correlation be significant (p < 0.05).
-Then, we extracted the Pearson correlation value corresponding to the time scales that reached the maximum value.
-Thus, we derived two raster maps per index, the first with the time scales and the second with the correlation value.
-
-<!-- ##	Total water storage and drought indices -->
-
-<!-- Terrestrial Water Storage (TWS) is defined as the total amount of water stored on land that includes any natural or artificial water bodies, such as groundwater, soil moisture, rivers, lakes, snowpack, ice, and biomass water [@Humphrey2023; @Deng2023].
-TWS changes evidence the effects of multiple water fluxes on the hydrological cycle [@Deng2023].
-These are reflected in the temporal variation of observations of the Earth's gravitational field [@Abolafia2021; @Sabzehee2023] and, in recent decades, the twin Gravity Recovery and Climate Experiment (GRACE) satellites and their Follow-On mission (GRACE-FO) have provided valuable results on globally distributed TWS anomalies [@Tapley2019; @Ferreira2023].
-The GRACE mission was launched in March 2002 and was operational until October 2017 [@Ramjeawon2022].
-Then its GRACE Follow-On successor was launched in May 2018 [@Landerer2020; @Yin2022].
-The information provided by these satellites is used to construct monthly maps of the Earth's average gravity field, providing details of the movement of water masses or water mass anomaly estimates relative to the long-term average gravity field [@Humphrey2023; @Wahr2004].
--->
-
-<!-- In this study, RL06.1_V3 GRACE mascon (mass concentration) monthly solutions were used, which are provided by the Jet Propulsion Laboratory (JPL-M, https://grace.jpl.nasa.gov).
-Each GRACE Tellus monthly grid at 0.5 degrees represents the deviation of surface mass for that month relative to a reference time average (2004-2009 baseline), which is subtracted from all other monthly grids to provide terrestrial water storage anomalies (TWSA) [@Ramjeawon2022; @Yin2022].
-This JPL-M version of the data employs a Coastal Resolution Improvement (CRI) filter that reduces signal leakage errors across coastlines [@NASA2019; @Wiese2016].
-Although the mascon solutions greatly reduced leakage errors, a gain factor, which is used to enhance the spatial resolution, was applied to the dataset as recommended for hydrological studies [@Ramjeawon2022; @Yin2022].
-The water storage and height anomalies are given in Equivalent Water Height or Thickness units (EWH, cm) [@Sabzehee2023], and their temporal resolution monthly is from April 2002 to May 2023.
--->
+An analysis is conducted on the linear correlation between the indices SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36 months, and zcNDVI.
+The objective is to determine the impact of soil moisture and water demand and supply on vegetation productivity.
+We implemented a methodology similar to that of @Meroni2016 when comparing the SPI for meteorological drought to the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation), which served as an indicator for vegetation productivity.
+A pixel-to-pixel linear correlation analysis was performed for each index.
+To begin, the Pearson coefficient of correlation is computed for each of the six time scales.
+A significant time scale is identified as the one that attains the highest correlation (p < 0.05).
+Subsequently, the Pearson correlation coefficient corresponding to the time scales at which the value peaked was extracted.
+As a result, for each index, we generated two raster maps: one containing the time scales and the other the correlation value.
 
@@
 The ERA5 soil moisture overestimate is 74%, but it has a kind of good correlation.
+For more detailed information, consult the supplementary material (SSX).
 
@@
 
-@fig-studyArea d, shows the macroclasses of land cover persistance (80%) during 2021-2022, respectively.
-Within continental Chile, forest is the vegetation type with highest surface with 135,00 $km^2$, followed by grassland (73,176 $km^2$), savanna (54,410 $km^2$), shrubland (24,959 $km^2$), and cropland (3,100 $km^2$) (\ref{tab-land coverSurf}).
-The macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively (@fig-studyArea and Table \ref{tab-land coverTrend}); of its surface that changes the type of land cover.
+@fig-studyArea d, shows the macroclasses of land cover persistance (80%) during 2021-2022, respectively (\ref{tab-landcoverSurf}).
+Within continental Chile, forest is the vegetation type with the highest surface area at 135,00 $km^2$, followed by grassland (73,176 $km^2$), savanna (54,410 $km^2$), shrubland (24,959 $km^2$), and cropland (3,100 $km^2$) (Table \ref{tab-landcoverSurf}).
+The macrozones with major LULCC for 2001–2022 were "Centro," "Sur," and "Austral" with 36%, 31%, and 34%, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend}); of its surface that changes the type of land cover.
 @fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.
@@
 \caption{Surface of the land cover class that persist during 2001-2022}
-\label{tab-land coverSurf}
+\label{tab-landcoverSurf}
 \includegraphics[width = .5\textwidth]{../output/figs/table_surface_landcover_macrozone.png}
@@
 Red dots on the plots indicate the maximum and minimum values of surface.}
-\label{tab-land coverTrend}
+\label{tab-landcoverTrend}
 \includegraphics[]{../output/figs/table_var_landcover_macro.png}
@@
 Mainly for a decrease in water supply (SPI and SSI) and an increase in water demand (EDDI).
-However, due to the high variability from north to south in Chile, the climatic condition (arid, semi-arid, and humid), and the land cover type, we believe that only in those zones could the LULCC be driven to some degree by drought.
+However, due to the high variability from north to south in Chile, the climatic conditions (arid, semi-arid, and humid), and the land cover type, we believe that only in those zones could the LULCC be driven to some degree by drought.
 
@@
 
-Regarding vegetation productivity aggregated through the macrozones in the five land cover macroclasses, in "Norte Grande," there is an increase trend of 0.02 (z-index) per decade, related to types of grassland and shrubland.
+In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.02 in the five land cover macroclasses, with respect to grassland and shrubland categories.
 There is a negative trend in "Norte Chico" with -0.04 and "Centro" with -0.02 per decade.
@@
 ```
-
-Analyzing the water supply, the macrozones that have the lowest trend are "Norte Chico" and "Centro," where the SPI, SPEI, and SSI show that it decreases at longer time scales due to the prolonged reduction in precipitation.
-At 36 months, it reaches trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and SSI (@fig-trendDI).
-For "Sur," the behavior is similar, decreasing at longer scales and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI.
-On the other hand, all macrozones show an increase in the trend in all the drought indices, with "Norte Grande" having the highest at 36 months (0.042 per decade).
-Because of this, the SPEI (which uses AED) reached its lowest value in "Norte Grande," with -0.03 at 36 months.
-Despite the other macrozones, "Austral" showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend (@fig-trendDI and @fig-trendDIMacro).
-
-##	Impact for water supply and demand, and soil moisture in vegetation productivity
-
-According to what is shown in @fig-corTimeScale, @fig-corPerson, and Table \ref{tab-corland cover}, forest seems to be the most resistant type to drought.
-Showing that only "Centro" is slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit (SSI-12).
-In the "Norte Chico" and to a lesser extent in the "Norte Grande," it is evident that a SSI-12 with a rsq = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34, respectively) have an impact on grasslands.
-However, this type was unaffected by soil moisture, water supply, or demand in macrozones further south.
-The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
-For savannas in "Norte Chico," the SSI-12 and SPI-24 reached an rsq of 0.74 and 0.58, respectively.
-This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (rsq = 0.45 in "Centro" and 0.2 in "Sur").
-In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45% and 66% of "Norte Chico." The type of land most impacted by climatic variation was shrubland, where soil moisture explained 59% and precipitation, 37%, in "Norte Chico" and "Centro," with SSI-12 being the most relevant variable, then SPI-36 in "Norte Chico" and SPI-24 in "Sur."
-
-\newpage
 
@@
 The numbers in each cell indicate the time scale that reached the maximum correlation for the land cover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
-\label{tab-corland cover}
+\label{tab-corlandcover}
 \includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
@@
 ```
+
+Analyzing the water supply, the macrozones that have the lowest trend are "Norte Chico" and "Centro," where the SPI, SPEI, and SSI show that it decreases at longer time scales due to the prolonged reduction in precipitation.
+At 36 months, it reaches trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and SSI (@fig-trendDI).
+For "Sur," the behavior is similar, decreasing at longer scales and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI.
+On the other hand, all macrozones show an increase in the trend in all the drought indices, with "Norte Grande" having the highest at 36 months (0.042 per decade).
+Because of this, the SPEI (which uses AED) reached its lowest value in "Norte Grande," with -0.03 at 36 months.
+Despite the other macrozones, "Austral" showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend (@fig-trendDI and @fig-trendDIMacro).
+
+##	Impact for water supply and demand, and soil moisture in vegetation productivity
+
+According to what is shown in @fig-corTimeScale, @fig-corPerson, and Table \ref{tab-corlandcover}, forest seems to be the most resistant type to drought.
+Showing that only "Centro" is slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit (SSI-12).
+In the "Norte Chico" and to a lesser extent in the "Norte Grande," it is evident that a SSI-12 with a rsq = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34, respectively) have an impact on grasslands.
+However, this type was unaffected by soil moisture, water supply, or demand in macrozones further south.
+The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
+For savannas in "Norte Chico," the SSI-12 and SPI-24 reached an rsq of 0.74 and 0.58, respectively.
+This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (rsq = 0.45 in "Centro" and 0.2 in "Sur").
+In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45% and 66% of "Norte Chico." The type of land most impacted by climatic variation was shrubland, where soil moisture explained 59% and precipitation, 37%, in "Norte Chico" and "Centro," with SSI-12 being the most relevant variable, then SPI-36 in "Norte Chico" and SPI-24 in "Sur."
 
@@
 These changes are associated with a decrease in water demand from vegetation.
-Nonetheless, we used the persistant landcover to ensure that the pixel has the same class; in the case of croplands, it could happen that some areas had changed crops for others with higher water consumption.
+Nonetheless, we used the persistant land cover to ensure that the pixel has the same class; in the case of croplands, it could happen that some areas had changed crops for others with higher water consumption.
 But this effect should be minor compared to the results from ladcover macroclasses.
@@
 They found that cropland was more sensitive to drought than cropland, showing that cropland responds strongly to short- and medium-term drought (< SPEI-6).
-In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro." In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016; @Potopova2015; @Dai2020].
+In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro." In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016; @Potopova2015; @Dai2020;@Rhee2010].
 Short-term precipitation deficits impact soil water, and thus less water is available for plant growth.
@@
 Thus, we attribute this impact to the hydrological drought that has decreased groundwater storage [@Taucare2024], which in turn is impacted by long-term deficits, and consequently, the vegetation is more dependent on groundwater.
-In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared (\ref{tab-corland cover}).
+In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared (\ref{tab-corlandcover}).
 What can be explained is that, south of "Centro," predominate forest and grassland, the most resistant types.
@@
 Extreme drought conditions are an important driver of tree mortality, as shown by @Senf2020 in Europe.
-However, we found that forest is the type of landcover macroclass less affected by variation in drought indices, being the most resistant landcover class to drought.
+However, we found that forest is the type of land cover macroclass less affected by variation in drought indices, being the most resistant land cover class to drought.
 Supporting this is @Fathi-Taperasht2022, who asserts that Indian forests are the most drought-resistant and recover rapidly.
@@
 
-In contrast to our observations, @Venegas2022, who investigated Cryptocarya alba and Beilschmiedia miersii (both members of the Lauraceae family) inhabiting sclerophyllous forests in Chile, concluded that the overall development of the trees had decelerated, potentially indicating alterations in the inherent dynamics of their respective forests.
+In contrast to what we obtained, @Venegas2022, who studied Cryptocarya alba and Beilschmiedia miersii (both from the Lauraceae family) that live in sclerophyllous forests in Chile, found that the trees' overall growth had slowed down.
+This could mean that the natural dynamics of their forests have changed.
 They attributed it to the cumulative effects of the unprecedented drought (i.e., hydrological drought).
-Thus, we attribute that forest to being the most resistant to drought, due to the fact that most of the species comprising it are highly resilient to water scarcity.
-compared to the other landcover classes.
-Also, when we used MOD13A3 with a 1km spatial resolution to measure vegetation condition, it took the average condition of 1 square kilometer.
-Then, to study how a type of forest (sclerophyllous forest) changes in response to drought on a local level using remote sensing, we should use operational products with higher spatial resolutions, such as those from Landsat or Sentinel, which can show how different the forests are.
+Thus, we attribute that forest to being the most resistant to drought, due to the fact that most of the species comprising it are highly resilient to water scarcity compared to the other land cover classes.
+Nonetheless, if we want to go deep in our analysis, we should use earth observation data that is able to capture a higher level of detail.
+For example, when we used MOD13A3 with a 1km spatial resolution to measure vegetation condition, it took the average condition of 1 square kilometer.
+Then, to study how a type of forest (e.g., sclerophyllous forest) changes in response to drought on a local level using remote sensing, we should use operational products with higher spatial resolutions, such as those from Landsat or Sentinel.
 
@@
 Moreover, @Zhang2022 indicates there is a bidirectional causality between soil moisture and vegetation productivity.
-For this, lastly, some studies have redefined agricultural drought as soil moisture drought from the hydrological perspective [@Loon2016;@Samaniego2018].
+Lastly, some studies have redefined agricultural drought as soil moisture drought from an hydrological perspective [@Loon2016;@Samaniego2018].
 Even though soil moisture is the external factor most determinant of vegetation biomass, there are multiple internal factors, such as species, physiological characteristics, and plant hydraulics, that would affect vegetation productivity.
@@
 
-Our results showed that the drought index derived from soil moisture (SSI) was the most effective in explaining vegetation productivity within landcover macroclasses.
-However, there is strong spatial variability throughout Chile and between classes, mainly attributable to climate heterogeneity, hydrological status, or vegetation resistance to water scarcity.
+The study results showed that the soil moisture-based drought index (SSI) was better at explaining vegetation productivity across land cover macroclasses than meteorological drought indices like SPI, SPEI, and EDDI.
+In the early growing season and especially in irrigated rather than rainfed croplands, soil moisture has better skills than SPI and SPEI for estimating gross primary production (GPP), according to @Chatterjee2022's evaluation of the SPI and SPEI and their correlation with GPP in the CONUS.
+Also, @Zhou2021 indicates that the monthly scaled Standardized Water Deficit Index (SWDI) can accurately show the effects of agricultural drought in most of China.
+Complementary, @Nicolai2017 analyzed the time-lag between SWDI and the Vegetation Condition Index (VCI), and they state that there was no or little time-lag with croplands but a significant time-lag in the case of forests.
+
+In our case, there is strong spatial variability throughout Chile and between classes, mainly attributable to climate heterogeneity, hydrological status, or vegetation resistance to water scarcity.
 The semi-arid "Norte Chico" and the Mediterranean "Centro" were where SSI had the best performance.
-Medium-term deficits of 12 months are more relevant in the response of vegetation, which decreases to the south, and in the case of croplands, they seem to react in a shorter time, with six months (SSI-6) in "Centro."
+In Chile, medium-term deficits of 12 months are more relevant in the response of vegetation, which decreases to the south, and in the case of croplands, they seem to react in a shorter time, with six months (SSI-6) in "Centro." This variation for croplands could be related to the fact that in "Norte Chico," the majority of crops are irrigated, but to the south there is a higher proportion of rainfed agriculture, which is most dependent on the short-term availability of water.
+Rather, in the "Norte Chico," the orchards are more dependent on the storage of water in dams of groundwater reservoirs, which are affected by long-term drought (e.g., SPI-36).
 
-@Chatterjee2022
+## Early drought forecasting for vegetation productivity
 
-## Drought indices of water demand and supply, soil moisture to predict changes in vegetation productivity
+We analyzed the correlation between meteorological and soil moisture drought indices with zcNDVI.
+From our findings, we could further use the drought indices and time scales that have the highest r-squared to develop a combined model to forecast vegetation productivity in Chile.
+Despite the fact that SSI was the best-performing index, the rest of the indices should help to enhance predictability.
+@Zambrano2018 proposed a prediction model for cropland surface in Chile at administrative units with a 1- to 6-month lead time using zcNDVI from MODIS and climate oscilation indices.
+The results given were r-squared, ranging from 0.95 at a 1-month lead time to 0.37 at a 6-month lead time.
+Thus, incorporating the results of this study with those made by @Zambrano2018, we could develop a combined forecasting model at the pixel level for the macroclasses in Chile based on drought indices of water demand and supply and soil moisture.
 
-3. Como podrían servir estos resultados para desarrollar o mejorar un predictor de productividad de la vegetación.
-  - Los datos ERA5L están casi en tiempo real, 7 dias; MODIS también.
-  - EL SSI se ve como un poderoso indicador que explica la variabilidad en la productividad de la vegetación.
-
-## Early drought forecasting
-
-## Future outlook
-
-4.- Qué se podría hacer mejor en futuras investigaciones del tema.
-  - mejorar la resolución y calidad de los datos climáticos
-  -
+## Future outlook (to complete)
 
@@
 
-\newpage
+There is a trend toward decreasing water supply in most parts of Chile, less in the "Austral," which is stronger in the "Centro" and "Norte Chico." The whole country showed an increase in water demand.
+Vegetation productivity only showed a decrease in the "Norte Chico" and "Centro," being highest for shrubland and croplands.
+Forest is the land cover most resistant to drought, and shrubland and cropland are the most sensitive.
+
+A soil moisture deficit of 12 months (SSI-12) is highly correlated with vegetation productivity for the land cover classes of shrubland, savannas, croplands, and forest in "Norte Chico" and "Centro." For the southern part of the country with humid conditions, the correlation with SSI decreases.
+Soil moisture overcomes the capacity to explain vegetation productivity over the supply and demand drought indices in the entire territory.
+
+The variation in vegetation productivity appears to be associated with climate variation rather than anthropogenic factors (e.g., an increase in water demand by irrigated crops).
+Even though switching to more demanding crops on the land could increase the impact of drought on vegetation, this would need to be more thoroughly investigated, for instance at the watershed level.
+
+The results of this study could help in the development of a robust forecasting system for land cover classes in Chile., helping to improve preparedness for climate change impacts on vegetation.
+
+# Supplementary material {-}
 

### 2024-03-12 article/paper_drought_Chile2023.tex
update paper qm intro

--- 2024-01-25 article/paper_drought_Chile2023.tex
@@
 \hypersetup{
-  pdftitle={Drought indices of water demand and supply and their impact on LULCC in Chile from ERA5-Land and MODIS},
-  pdfauthor={Francisco Zambrano},
+  pdftitle={Drought, vegetation productivity, and land cover change in Chile},
+  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Fernández; Francisco Meza; Alejandro Venegas-González; Iongel Duran-Llacer; Nicolas Raab; Dylan Craven},
   pdfkeywords={drought, land cover change, satellite},
@@
 \begin{frontmatter}
-\title{Drought indices of water demand and supply and their impact on
-LULCC in Chile from ERA5-Land and MODIS}
+\title{Drought, vegetation productivity, and land cover change in Chile}
 \author[1]{Francisco Zambrano%
@@
  \ead{francisco.zambrano@umayor.cl} 
+\author[2]{Anton Vrieling%
+%
+}
+ \ead{a.vrieling@utwente.nl} 
+\author[3]{Francisco Fernández%
+%
+}
+ \ead{a.vrieling@utwente.nl} 
+\author[]{Francisco Meza%
+%
+}
+
+\author[]{Alejandro Venegas-González%
+%
+}
+
+\author[]{Iongel Duran-Llacer%
+%
+}
+
+\author[]{Nicolas Raab%
+%
+}
+
+\author[]{Dylan Craven%
+%
+}
 
@@
 Chile},postcode={7500994},postcodesep={}}
+\affiliation[2]{organization={University of Twente, Faculty of
+Geo-Information Science and Earth Observation (ITC)},city={Enschede, The
+Netherlands},postcodesep={}}
+\affiliation[3]{organization={Universidad San
+Sebastian},,postcodesep={}}
 
@@
 system and vegetation development. This persistent period of water
-scarcity has been defined as a ``Mega Drought''. Our objective is to
-examine the effects of drought on LULCC (land use land cover change)
-over continental Chile using drought indices of water supply and demand,
-soil moisture, and vegetation productivity. For the analysis,
-continental Chile was divided into five zones according to a latitudinal
-gradient: ``Norte Grande,'' ``Norte Chico,'' ``Centro,'' ``Sur,'' and
-``Austral.'' The monthly ERA5-Land (ERA5L) variables for precipitation,
-temperature, and soil moisture were used. From 2001 to 2022, we used the
-land cover MODIS product MCD12Q1, and from 2000 to 2023, we used the
-NDVI (Normalized Difference Vegetation Index) product MOD13A3 collection
-6.1. We estimated atmospheric evaporative demand (AED) using the
-Hargreaves-Samani equation with the ERA5L temperature. We used the
-Standardized Precipitation Index (SPI), the Standardized Precipitation
-Evapotranspiration Index (SPEI), the Evaporative Demand Drought Index
-(EDDI), the Standardized Soil Moisture Index (SSI), and the Standardized
-anomaly of cumulative NDVI (zcNDVI) as drought indicators. These indices
-were calculated for time scales of 1, 3, 6, 12, 24, and 36 months,
-except for zcNDVI, which was for 6 months. We analyze the trend for
-LULCC, vegetation productivity, and drought indices. Also, we analyzed
-the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain
-insights into the impact of water supply and demand on vegetation
-productivity. Our results showed that LULCC were highest in ``Centro,''
-``Sur,'' and ``Austral,'' with 36\%, 31\%, and 34\%, respectively. The
-EDDI shows that water demand has increased for all zones, with a major
-increase in ``Norte Grande.'' The drought indices of water supply and
-soil moisture evidence a decreasing trend, which decreases at longer
-time scales, from ``Norte Grande'' to ``Sur.'' ``Austral'' is the only
-zone that shows an increase in supply. Vegetation productivity measures
-by zcNDVI present a negative trend in ``Norte Chico'' and ``Centro.''
-Showing to be the zones most impacted by climatic conditions, the years
-2020 and 2022 suffered the most extreme drought. On the other hand,
-forests seem to be the most resistant to drought. The types that show to
-be most affected by variation in climate conditions are shrublands,
-savannas, and croplands. The drought indices that have the capability of
-explaining to a major degree the variance in vegetation productivity are
-SSI-12, followed by SPEI-24 and SPEI-12 in ``Norte Chico'' and
-``Centro.'' The results indicate that ``Norte Chico'' and ``Zona
-Central'' are the most sensitive regions to water supply deficits longer
-than a year, potentially explained by a low capacity of water storage in
-those zones that should be further investigated.
+scarcity has been defined as a megadrought. Our objective is to assess
+the impact of drought on LULCC (land use land cover change) over
+continental Chile using drought indices of water supply and demand, soil
+moisture, and their impact on vegetation productivity. The monthly
+ERA5-Land (ERA5L) variables for precipitation, temperature, and soil
+moisture were used. From 2001 to 2022, we used the land cover MODIS
+product MCD12Q1, and from 2000 to 2023, we used the NDVI (Normalized
+Difference Vegetation Index) product MOD13A3 collection 6.1. As drought
+indices, we compute the standardized anomaly of cumulative NDVI
+(zcNDVI), the Standardized Precipitation Evapotranspiration Index
+(SPEI), the Evaporative Demand Drought Index (EDDI), and the
+Standardized Soil Moisture Index (SSI). These indices were calculated
+for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI,
+which was for 6 months. We analyze the trend for LULCC, vegetation
+productivity, and drought indices. Also, we analyzed the temporal
+correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain insights
+into the impact of water supply and demand on vegetation productivity.
+Our results showed that LULCC was highest in ``Centro,'' ``Sur,'' and
+``Austral,'' with 36\%, 31\%, and 34\% of change in the surface type,
+respectively. The EDDI shows that water demand has increased for all
+zones, with a major increase in ``Norte Grande.'' The drought indices of
+water supply and soil moisture evidence a decreasing trend, which
+decreases at longer time scales, from ``Norte Grande'' to ``Sur.''
+``Austral'' is the only zone that shows an increase in supply.
+Vegetation productivity measures by zcNDVI present a negative trend in
+``Norte Chico'' and ``Centro.'' On the other hand, forests seem to be
+the most resistant to drought. The types that show to be most affected
+by variation in climate conditions are shrublands, savannas, and
+croplands. The drought indices that have the capability of explaining to
+a major degree the variance in vegetation productivity are SSI-12,
+followed by SPEI-24 and SPEI-12 in ``Norte Chico'' and ``Centro.'' The
+results indicate that ``Norte Chico'' and ``Zona Central'' are the most
+sensitive regions to water supply deficits lasting longer than a year.
+Our results can help develop a robust vegetation productivity
+forecasting model for land cover classes in Chile.
 \end{abstract}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, boxrule=0pt, frame hidden, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 
-The sixth assessment report (AR6) of the IPCC \citep{IPCC2023} indicates
-that human-induced greenhouse gas emissions have increased the frequency
-and/or intensity of some weather and climate extremes, and the evidence
-has been strengthened since AR5 \citep{IPCC2013}. There is a high degree
-of confidence that rising temperatures will increase the land area where
-droughts will occur more frequently and with greater severity
-\citep{IPCCCH112021}. Furthermore, drought increases tree mortality and
-triggers changes in land cover and, consequently, land use, thus
-impacting ecosystems \citep{Crausbay2017}. Nevertheless, there is a lack
-of understanding of how the alteration in water supply and demand is
-affecting land cover transformations.
-
-The primary cause of drought is precipitation, and temperature makes it
-worse \citep{Luo2017}. Drought impacts soil moisture, hydrological
-regimes, and vegetation productivity. Initially, drought was commonly
-classified as meteorological, hydrological, and agricultural
-\citep{Wilhite1985}. Lately, \citep{Loon2016} and
-\citep{AghaKouchak2021} have given an updated definition of drought for
-the Anthropocene, suggesting that it should be considered the feedback
-of humans' decisions and activities that drives the anthropogenic
-drought. Even though it has been argued that those definitions do not
-fully address the ecological dimensions of drought, \citep{Crausbay2017}
-proposed the ecological drought definition as ``an episodic deficit in
-water availability that drives ecosystems beyond thresholds of
-vulnerability, impacts ecosystem services, and triggers feedback in
-natural and/or human systems.''. Moreover, many ecological studies have
+Drought is often classified as meteorological when there is a decrease
+in precipitation below the mean average of several years (more than 30
+years), hydrological when these anomalies last for long periods (months
+to years) and affect water systems, and agricultural when the deficit
+impacts plant health anomalies and leads to decreased productivity
+\citep{Wilhite1985}. However, it is important to note that drought is
+also influenced by human activities, which were not considered in the
+definitions. Thus, \citet{Loon2016} and \citet{AghaKouchak2021} have
+given an updated definition of drought for the Anthropocene, suggesting
+that it should be considered the feedback of humans' decisions and
+activities that drives the anthropogenic drought. Simultaneously,
+drought leads to heightened tree mortality and induces alterations in
+land cover and land use, ultimately affecting ecosystems
+\citep{Crausbay2017}. Even though many ecological studies have
 misinterpreted how to characterize drought, for example, sometimes
-considering ``dry'' conditions as ``drought'' \citep{Slette2019}. On the
-other hand, the AR6 \citep{IPCC2023} predicts that many regions of the
-world will experience more severe agricultural and ecological droughts
-even if global warming stabilizes at 1.5°--2°C. Then, there is a
-challenge in conducting drought research, especially to evaluate its
-impact on ecosystems.
-
-Chile has been facing a persistent rainfall deficit for more than a
-decade \citep{Garreaud2017}, which has impacted vegetation development
-\citep{Zambrano2023} and the hydrological system \citep{Boisier2018}.
-Current drought conditions have affected crop productivity
-\citep{Zambrano2016, Zambrano2018}, forest development
+considering ``dry'' conditions as ``drought'' \citep{Slette2019}. Then,
+\citet{Crausbay2017} proposed the ecological drought definition as ``an
+episodic deficit in water availability that drives ecosystems beyond
+thresholds of vulnerability, impacts ecosystem services, and triggers
+feedback in natural and/or human systems.'' In light of current global
+warming, it is crucial to study the interaction between drought and
+ecosystems in order to understand their feedback and impact on water
+security. \citep{Bakker2012}
+
+Human-induced greenhouse gas emissions have increased the frequency
+and/or intensity of drought as a result of global warming, according to
+the sixth assessment report (AR6) of the Intergovernmental Panel on
+Climate Change (IPCC) \citep{IPCC2023}. The evidence supporting this
+claim has been strengthened since AR5 \citep{IPCC2013}. Recent studies,
+however, have produced contrasting findings, suggesting that drought has
+not exhibited a significant trend over the past forty years.
+\citep{Vicente-Serrano2022, Kogan2020}. \citet{Vicente-Serrano2022}
+analyzed the meteorological drought trend on a global scale, finding
+that only in a few regions has there been an increase in the severity of
+drought. Moreover, they attribute the increase in droughts over the past
+forty years solely to an increase in atmospheric evaporative demand
+(AED), which in turn enhances vegetation water demand, with important
+implications for agricultural and ecological droughts. Also, they state
+that ``the increase in hydrological droughts has been primarily observed
+in regions with high water demand and land cover change''. Similarly,
+\citet{Kogan2020} analyzed the drought trend using vegetation health
+methods, finding that for the globe, hemispheres, and main
+grain-producing countries, drought has not expanded or intensified for
+the last 38 years. Further, the \citet{IPCC2021} suggests that there is
+a high degree of confidence that rising temperatures will increase the
+extent, frequency, and severity of droughts. Also, AR6 \citep{IPCC2023}
+predicts that many regions of the world will experience more severe
+agricultural and ecological droughts even if global warming stabilizes
+at 1.5°--2°C. To better evaluate the impact of drought trends on
+ecosystems, assessments are needed that relate meteorological and soil
+moisture variables to their effects on vegetation.
+
+From 1960 to 2019, land use change has impacted around one-third of the
+Earth's surface, which is four times more than previously thought
+\citep{Winkler2021}. Multiple studies aim to analyze and forecast
+changes in land cover globally \citep{Winkler2021, Song2018} and
+regionally \citep{Chamling2020, Homer2020, Yang2021}. Some others seek
+to analyze the impact of land cover change on climate conditions such as
+temperature and precipitation \citep{Luyssaert2014, Pitman2012}. There
+is less research on the interaction between drought and land cover
+change \citep{Chen2022, Akinyemi2021, Peng2017}. \citet{Peng2017}
+conducted a worldwide investigation utilizing net primary production to
+examine the spatial and temporal variations in vegetation productivity
+at global level. The study aimed to assess the influence of drought by
+comparing the twelve-month Standardized Precipitation Evapotranspiration
+Index (SPEI) and land cover change. According to their findings, drought
+is responsible for 37\% of the decline in vegetation productivity, while
+water availability accounts for 55\% of the variation. \citet{Chen2022}
+studied the trend of vegetation greenness and productivity and its
+relation to meteorological drought (SPEI of twelve months in December)
+and soil moisture at the global level. The results showed lower
+correlations (\textless0.2) for both variables. \citet{Akinyemi2021}
+evaluates drought trends and land cover change using vegetation indices
+in Botwsana in a semi-arid climate. These studies mostly looked at how
+changes in land cover and vegetation productivity are related to a
+single drought index (SPEI) over a single time period of 12 months. SPEI
+takes into account the combined effect of precipitation and AED as a
+water balance, but it does not allow us to know the contribution of each
+variable on its own. Some things worth investigating in terms of land
+cover change and vegetation productivity are: i) How do they respond to
+short- to long-term meteorological droughts? ii) How do they behave in
+humid and arid climatic zones regarding drought? And iii) What is the
+role of soil moisture? Likewise, there is a lack of understanding of how
+the alteration in water supply and demand is affecting land cover
+transformations.
+
+Chile's diverse climatic and ecosystem types
+(\citet{Beck2023};\citet{Luebert2022}) make it an ideal natural
+laboratory for studying climate and ecosystems. Additionally, the
+country has experienced severe drought conditions that have had
+significant effects on vegetation and water storage. Central Chile faced
+a persistent precipitation deficit between 2010 and 2022, defined as a
+megadrought \citep{Garreaud2017}, which has impacted the Chilean
+ecosystem. This megadrought was defined by the Standardized
+Precipitation Index (SPI) of twelve months in December having values
+below one standard deviation. Some studies have addressed how this
+drought affects single ecosystems in terms of forest development
 \citep{Miranda2020, Venegas2018}, forest fire occurrence
-\citep{UrrutiaJalabert2018}, land cover change \citep{Fuentes2021},
-water supply in watersheds \citep{AlvarezGarreton2021}, and have had
-economic impacts \citep{Fernandez2023}. In 2019--2020, the drought
-severity reached an extreme condition in Central Chile (30--34°S) not
-seen for at least 40 years, and the evidence indicates that the impact
-is transversal to the land cover classes of forest, grassland, and
-cropland \citep{Zambrano2023}. The prolonged lack of precipitation in
-Central Chile is producing changes in ecosystem dynamics that must be
-studied.
-
-For the spatiotemporal assessment of drought impact (i.e., by water
-supply and demand) on land cover changes, we need climatic realiable
-variables such as precipitation, temperature, soil moisture, land cover,
-and vegetation status. For developing countries like Chile, the weather
-networks present several disadvantages, such as gaps, a short history,
-and low-quality data. Reanalysis data, as the ERA5-Land (ERA5L)
-\citep{MunozSabater2021} provides hourly climatic information
-(precipitation, temperature, and soil moisture) without gaps since 1950
-with global extension. ERA5L has already been used for drought
-assessment using the Standardized Precipitation-Evapotranspiration Index
-(SPEI) \citep{Nouri2023} and for flash drought \citep{Wang2023} by
-analyzing soil moisture and evapotranspiration. On the other hand,
-satellite remote sensing \citep{West2019, AghaKouchak2015} is the
-primary method to evaluate how drought impacts vegetation dynamics. The
-Moderate-Resolution Imaging Spectroradiometer (MODIS) can be used to get
-vegetation drought indices (VDI), which are often used as proxies for
-productivity \citep{Paruelo2016, Schucknecht2017}. Besides, land use and
-land cover (LULC) change can be driven by drought
-\citep{Tran2019, Akinyemi2021}. To analyze these changes, multiple LULC
-products exist \citep{Grekousis2015}. One of those that provides time
-series since 2001 is the MCD12Q1 \citep{Friedl2019} from MODIS. The
-variation in water supply and demand is finally reflected in the total
-water storage (TWS). The Gravity Recovery and Climate Experiment
-(GRACE), which allows analyzing changes in water availability at coarse
-resolution, can retrieve the TWS \citep{Ahmed2014, Ma2017}. We can find
-drought indices of supply (i.e., precipitation) and demand (i.e.,
-temperature) using climatic reanalysis (ERA5L) and vegetation data
-(MODIS). This lets us figure out how drought changes LULC. Further, the
-TWS can be assessed with regard to the changes in water supply and
-demand to gain insight into the impact on water storage.
-
-To evaluate meteorological drought (i.e., water supply), the World
-Meteorological Organization (WMO; \citep{WMO2012}) recommends the
-Standardized Precipitation Index (SPI; \citep{Mckee1993}), a multiscalar
-drought index that allows to monitor precipitation deficits from short-
-to long-term. Following the same approach, \citep{Vicente-Serrano2010}
-incorporates into the SPI the effect of temperature through the use of
-potential evapotranspiration, thus proposing the SPEI (Standardized
-Precipitation Evapotranspiration Index). Similarly, to evaluate solely
-the evaporative demand driven by temperature, \citep{Hobbins2016} and
-\citep{McEvoy2016} came up with the Evaporative Demand Drought Index
-(EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI;
-\citep{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
-cumulative Normalized Difference Vegetation Index (NDVI), which could be
-acumulated over the growing season or any period (e.g., months),
-resulting in a multiscalar drought index. For soil moisture, several
-drought indices exist, such as the Soil Moisture Deficit Index (SDMI) a
-normalized index \citep{Narasimhan2005} and the Soil Moisture
-Agricultural Drought Index (SMADI) \citep{Souza2021} which is a
-normalized index using vegetation, land surface temperature, and a
-vegetation condition index (VCI, \citep{Kogan1995}). From TWS, we can
-estimate the standardized terrestrial water storage index (STI)
-\citep{Cui2021}, a standardized anomaly that follows the methodology of
-the SPI, SPEI, EDDI, and zcNDVI. Thereby, we have drought indices for
-water supply, demand, and storage, which can help to make a
-comprehensive assessment of drought.
-
-In this research, we aim to analyze the impact of drought on different
-types of ecosystems (land cover classes) in continental Chile. Our
-specific goals are: i) to analyze the trend on multi-scalar drought
-indices for water demand and supply, soil moisture, and vegetation
-productivity for 1981--2023/2001--2023; ii) to assess the LULC change
-for 2001--2022 and how it relates to drought indices; iii) to evaluate
-the relationship between zcNDVI, a measure of vegetation productivity,
-and drought indices for water demand and supply and soil moisture; and
-iv) to assess if the observed changes in the drought indices are linked
-to TWS.
+\citep{UrrutiaJalabert2018}, and crop productivity
+\citep{Zambrano2023, Zambrano2018, Zambrano2016}. We found one study
+regarding land cover and drought in Chile. The study by
+\citet{Fuentes2021} evaluates water scarcity and land cover change in
+Chile between 29° and 39° of south latitude. \citet{Fuentes2021} used
+the SPEI of one month for evaluating drought, which led to misleading
+results. For example, they did not find a temporal trend in the SPEI but
+found a decreasing trend in water availability and an increase trend on
+AED, which in turn should have been capable of being captured with
+longer time scales of the SPEI. The term ``megadrought'' in Chile is
+used to describe a prolonged water shortage that lasts for several
+years, resulting in a permanent deficit that impacts the hydrological
+system \citep{Boisier2018}. Hence, it is imperative to assess temporal
+scales that take into account the cumnulative effect within some years.
+There is little knowledge about the relationship between drought and
+ecosystem in Chile; thus, it is important to understand in more detail
+how meteorological and soil moisture droughts influence ecosystem
+dynamics to inform adaptation options.
+
+A detailed spatiotemporal assessment of the interaction of drought for
+short- to long-term and land cover change requires information on
+vegetation as well as weather variables such as precipitation,
+temperature, and soil moisture. Weather networks in Chile present some
+disadvantages, such as spatio-temporal gaps, a short history, and
+irregular quality, which make them difficult to represent the whole
+extent of the country spatially. In order to do this, we use reanalysis
+data from ERA5-Land \citep{MunozSabater2021} to create drought indices
+that consider AED, precipitation, and soil moisture over a range of time
+periods, from the short to the long term. Also, we use vegetation
+spectral information and annual land cover change from the
+Moderate-Resolution Imaging Spectroradiometer (MODIS). We expect to gain
+insight regarding the temporal evolution of water demand, water supply,
+and soil moisture, as well as the interaction with land cover change and
+vegetation productivity. Here, we analyze the multi-dimensional impacts
+of drought across ecosystems in continental Chile. More specifically, we
+aim to assess: i) temporal changes in land-use cover and the direction
+and magnitude of their relationships with drought indices for water
+demand and supply, soil moisture, and vegetation productivity; ii)
+short- to long-term temporal trends in multi-scalar drought indices; and
+iii) the relationship between vegetation productivity and drought
+indices for water demand and supply and soil moisture across Chilean
+ecosystems.
 
@@
 
-Continetal Chile has a diverse climate condition from north to south and
-east to west \citep{Aceituno2021} (Figure~\ref{fig-studyArea}), which
-determines its great ecosystem diversity (Figure~\ref{fig-LCprop}). The
-Andes Mountains are a main factor in latitudinal variation
-\citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
-use the Koppen-Geiger release by \citep{Beck2023} and the land cover
-type persistance of 80\% of times for 2001--2022, from the IGBP
-classification scheme \citep{Friedl2019} (see
-Section~\ref{sec-methods_lulc}). ``Norte Grande'' and ``Norte Chico''
-predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
-temperatures. At the south of ``Norte Chico,'' the climate changes to an
-arid steppe with cold temperatures (Bsk). Mainly, the land is barren,
-with a minor surface of vegetation types such as shrubland and
-grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
-main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
-There is a significant amount (50\%) of Chilean matorral (shrubland and
-savanna, \citep{Fuentes2021}), then grassland (16\%), forest (8\%), and
-croplands (5\%), in ``Centro.'' The south part of ``Sur'' and the north
-part of ``Austral'' are dominated by an oceanic climate (Cfb). Those
-zones are high in forest and grassland. The southern part of the country
-has a tundra climate, and in Patagonia, it is a cold semi-arid area with
-an extended surface of grassland, forest, and, to a lesser extent,
-savanna.
+Continental Chile has a diverse climate conditions with strong gradients
+from north to south and east to west \citep{Aceituno2021}
+(Figure~\ref{fig-studyArea} a), which determines its great ecosystem
+diversity (\citet{Luebert2022}) (Figure~\ref{fig-studyArea} c). The
+Andes Mountains are a main factor in climate latitudinal variation
+\citep{Garreaud2009}. In order to characterize the climate and ecosystem
+of Chile, we utilize the Köppen-Geiger classification system developed
+by \citet{Beck2023} and the land cover data derived from the MODIS
+product for the period of 2001--2022, based on the International
+Geosphere-Biosphere Programme (IGBP) classification scheme proposed by
+\citet{Friedl2019}. ``Norte Grande'' and ``Norte Chico'' predominate in
+an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At
+the south of ``Norte Chico,'' the climate changes to an arid steppe with
+cold temperatures (Bsk). In these two northern regions, the land is
+mostly bare, with a minor surface of vegetation types such as shrubland
+and grassland. In the zones ``Centro'' and the north half of ``Sur,''
+the main climate is Mediterranean, with warm to hot summers (Csa and
+Csb). Land cover in ``Centro'' comprises a significant amount of
+shrubland and savanna (50\%), grassland (16\%), forest (8\%), and
+croplands (5\%). An oceanic climate (Cfb) predominates in the south of
+``Sur'' and the north of ``Austral.'' Those zones are high in forest and
+grassland. The southern part of the country has a tundra climate, and in
+``Austral'', it is a cold semi-arid area with an extended surface of
+grassland, forest, and, to a lesser extent, savanna.
 
@@
 ``Centro'', ``Sur'', and ``Austral''. (b) Topography reference map. (c)
-Land cover classes for 2022. (d) Persistent land cover classes
+land cover classes for 2022. (d) Persistent land cover classes
 (\textgreater{} 80\%) for 2001-2022}
@@
 used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
-used \texttt{\{SPEI\}} \citep{Bergueria2023}.
+used \texttt{\{SPEI\}} \citep{Bergueria2023}. For mapping, we use
+\{tmap\} \citep{Tennekes2018}. For data analysis, the suite
+\{tidyverse\} \citep{Wickham2019} was used.
 
@@
 
-To derive a proxy of vegetation productivity, we used the product
+To derive a proxy for vegetation productivity, we used the product
 MOD13A3 collection 6.1 from MODIS \citep{Didan2015}. It provides
@@
 that is owned by the agricultural enterprise Garces Fruit, which has 99
-stations in Central Chile, located in cherry fruit crops. We used daily
-data for the year 2022 and the months outside the growing season, May to
-September, to avoid the effect of irrigation on soil moisture, which is
-hardly captured by ERA5L.
+stations in Central Chile, located in cherry fruit crops. The sensors
+are installed at 30, 60, and 90m and are the model Teros 12 from
+MeterGroup. To avoid the effect of irrigation on soil moisture, which
+ERA5L hardly captures, we used daily data for the year 2022 and the
+months outside the growing season, May to September.
 
@@
 To evaluate water demand, we chose the \(EDDI\)
-\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). For
-supply, we used the index recommended by the World Meteorological
-Organization (WMO) for monitoring drought, the SPI \citep{Mckee1993}. We
-calculated the SPEI, which used a balance between \(P\) and \(AED\), in
-this case, an auxiliary variable \(D = P-AED\) is used. In this study,
-we used the \(SSI\) (standardized soil moisture index at 1 m)
+\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). We used
+the SPI \citep{Mckee1993}, a drought monitoring index that the World
+Meteorological Organization (WMO) recommends. We calculated the SPEI,
+which used a balance between \(P\) and \(AED\), in this case, an
+auxiliary variable \(D = P-AED\) is used. In this study, we used the
+\(SSI\) (standardized soil moisture index at 1 m)
 \citep{Hao2013, AghaKouchak2014}, which uses soil moisture at 1m depth.
 Finally, for the proxy of productivity, \(zcNDVI\), we used the NDVI.
-Before using the NDVI, it was smoothed using a locally-weighted
+Before using the NDVI, it was smoothed using a locally weighted
 polynomial regression, following the procedure described in
-\citep{Zambrano2018} and \citep{Zambrano2016}.
+\citet{Zambrano2018} and \citet{Zambrano2016}.
 
@@
 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
-Thus, we have a land cover raster time-series with the ten classes for
+Thus, we have a land cover raster time series with the ten classes for
 2001 and 2023.
@@
 To validate the land cover obtained, we compare the macroclasses with
-the ones of a more detailed land cover map made by \citep{Zhao2016} for
+the ones of a more detailed land cover map made by \citet{Zhao2016} for
 Chile with samples acquired in the years 2013--2014 (LCChile). The later
@@
   IGBP2013, and IGBP2014; we considered the point extracted from LCChile
-  as the truth and the values as the other two years as prediction
+  as the truth and the values from the other two years as predictions.
 \item
-  calculate a confusion matrix with the classes extracted in the 1000
-  poitns for LCChile, IGBP2013, and IGBP2014. Calculate the performance
+  calculate a confusion matrix with the classes extracted from the 1000
+  points for LCChile, IGBP2013, and IGBP2014. Calculate the performance
   metrics of accuracy and F1.
@@
 
-The time series of NDVI is affected by climatic conditions, vegetation
-development, seasonality, and changes in vegetation type. In this study,
-we want to analyze the variation in vegetation productivity in different
-land cover types and how it is affected by water demand, water supply,
-and soil moisture. In order to avoid changes due to a change in the land
-cover type, that will wrongly impact NDVI. We will develop a persistence
-mask for land cover for 2001--2023. Thereby, we reduce an important
-source of variation on a regional scale.
-
-Thus, we calculated a raster mask for IGBP MODIS considering the
-macroclasses that remain without change for more than 80\% of the years
-(2001--2022) per pixel, which allows us to identify the areas with no
-land cover change for the macroclasses.
+Climate, vegetation development, seasonality, and changes in vegetation
+type all have an impact on the time series of NDVI. In this study, we
+want to examine the variation in vegetation productivity across various
+land cover types and how water demand, water supply, and soil moisture
+affect it. In order to avoid changes due to a change in the land cover
+type that will wrongly impact NDVI, we will develop a persistence mask
+for land cover for 2001--2023. Thereby, we reduce an important source of
+variation on a regional scale.
+
+Therefore, we generated a raster mask for IGBP MODIS per pixel using
+macroclasses that remain constant for at least 80\% of the years
+(2001--2022). This enables us to identify regions where the land cover
+of the macroclasses has remained constant.
 
@@
 we will be able to advance in understanding how climate is affecting
-vegetation, considering the impact on the five macroclasses having
-vehetation: forest, cropland, grassland, savanna, and shrubland.
-
-To assess how water demand and supply and soil moisture are related to
-vegetation productivity (zcNDVI), we analyze the linear correlation
-between the indices SPI, SPEI, EDDI, and SSI for 1, 3, 6, 12, 24, and
-36-month time scales against zcNDVI. We followed a similar approach to
-that used by \citep{Meroni2016} when using the SPI for meteorological
-drought against the cumulative FAPAR (Fraction of Absorbed
-Photosynthetically Active Radiation) as a proxy for vegetation
-productivity. We made a pixel-to-pixel linear correlation analysis per
-index. First, we calculate the Pearson coefficient of correlation for
-the six time scales and let the time scale that reaches the maximum
-correlation be significant (p \textless{} 0.05). Then, we extracted the
-Pearson correlation value corresponding to the time scales that reached
-the maximum value. Thus, we derived two raster maps per index, the first
-with the time scales and the second with the correlation value.
+vegetation, considering the impact on the five macroclasses of
+vegetation: forest, cropland, grassland, savanna, and shrubland.
+
+An analysis is conducted on the linear correlation between the indices
+SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36
+months, and zcNDVI. The objective is to determine the impact of soil
+moisture and water demand and supply on vegetation productivity. We
+implemented a methodology similar to that of \citet{Meroni2016} when
+comparing the SPI for meteorological drought to the cumulative FAPAR
+(Fraction of Absorbed Photosynthetically Active Radiation), which served
+as an indicator for vegetation productivity. A pixel-to-pixel linear
+correlation analysis was performed for each index. To begin, the Pearson
+coefficient of correlation is computed for each of the six time scales.
+A significant time scale is identified as the one that attains the
+highest correlation (p \textless{} 0.05). Subsequently, the Pearson
+correlation coefficient corresponding to the time scales at which the
+value peaked was extracted. As a result, for each index, we generated
+two raster maps: one containing the time scales and the other the
+correlation value.
 
@@
 \(bias=1.74\). The ERA5 soil moisture overestimate is 74\%, but it has a
-kind of good correlation.
+kind of good correlation. For more detailed information, consult the
+supplementary material (SSX).
 
@@
 Figure~\ref{fig-studyArea} d, shows the macroclasses of land cover
-persistance (80\%) during 2021-2022, respectively. Within continental
-Chile, forest is the vegetation type with highest surface with 135,00
-\(km^2\), followed by grassland (73,176 \(km^2\)), savanna (54,410
-\(km^2\)), shrubland (24,959 \(km^2\)), and cropland (3,100 \(km^2\))
-(\ref{tab-land coverSurf}). The macrozones with major LULCC for
-2001-2022 were ``Centro'', ``Sur'', and ``Austral'' with 36\%, 31\%, and
-34\%, respectively (Figure~\ref{fig-studyArea} and Table
-\ref{tab-land coverTrend}); of its surface that changes the type of land
+persistance (80\%) during 2021-2022, respectively
+(\ref{tab-landcoverSurf}). Within continental Chile, forest is the
+vegetation type with the highest surface area at 135,00 \(km^2\),
+followed by grassland (73,176 \(km^2\)), savanna (54,410 \(km^2\)),
+shrubland (24,959 \(km^2\)), and cropland (3,100 \(km^2\)) (Table
+\ref{tab-landcoverSurf}). The macrozones with major LULCC for 2001--2022
+were ``Centro,'' ``Sur,'' and ``Austral'' with 36\%, 31\%, and 34\%,
+respectively (Figure~\ref{fig-studyArea} and Table
+\ref{tab-landcoverTrend}); of its surface that changes the type of land
 cover. Figure~\ref{fig-LCprop} shows the summary of the proportion of
@@
 (EDDI). However, due to the high variability from north to south in
-Chile, the climatic condition (arid, semi-arid, and humid), and the land
-cover type, we believe that only in those zones could the LULCC be
+Chile, the climatic conditions (arid, semi-arid, and humid), and the
+land cover type, we believe that only in those zones could the LULCC be
 driven to some degree by drought.
@@
 
-Regarding vegetation productivity aggregated through the macrozones in
-the five land cover macroclasses, in ``Norte Grande,'' there is an
-increase trend of 0.02 (z-index) per decade, related to types of
-grassland and shrubland. There is a negative trend in ``Norte Chico''
-with -0.04 and ``Centro'' with -0.02 per decade. In the ``Norte Chico,''
-savanna (-0.05) has the lowest trend, and the rest of the types are
-around -0.04. In ``Centro,'' shrubland reaches -0.06, grassland -0.05,
-and croplands and savanna -0.01 per decade. This could be associated
-either with a reduction in vegetation surface, a decrease in biomass, or
-browning \citep{Miranda2023}. Vegetation reached its lowest values since
-the year 2019, reaching an extreme condition in early 2020 and 2022 in
-the ``Norte Chico'' and Centro'' (Mega Drought). The ``Sur'' and
-``Austral'' show a positive trend of around 0.016 per decade
-(Figure~\ref{fig-zcNDVI_var}). Despite the croplands suffering from
-drought just as badly as the native vegetation in ``Norte Chico,'' the
-Chilean matorral \citep{Fuentes2021} appears to be the region most
-affected by a negative trend in vegetation productivity.
+In ``Norte Grande,'' vegetation productivity, as per the z-index,
+exhibits a yearly increase of 0.02 in the five land cover macroclasses,
+with respect to grassland and shrubland categories. There is a negative
+trend in ``Norte Chico'' with -0.04 and ``Centro'' with -0.02 per
+decade. In the ``Norte Chico,'' savanna (-0.05) has the lowest trend,
+and the rest of the types are around -0.04. In ``Centro,'' shrubland
+reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per
+decade. This could be associated either with a reduction in vegetation
+surface, a decrease in biomass, or browning \citep{Miranda2023}.
+Vegetation reached its lowest values since the year 2019, reaching an
+extreme condition in early 2020 and 2022 in the ``Norte Chico'' and
+Centro'' (Mega Drought). The ``Sur'' and ``Austral'' show a positive
+trend of around 0.016 per decade (Figure~\ref{fig-zcNDVI_var}). Despite
+the croplands suffering from drought just as badly as the native
+vegetation in ``Norte Chico,'' the Chilean matorral \citep{Fuentes2021}
+appears to be the region most affected by a negative trend in vegetation
+productivity.
 
@@
 \end{figure}
+
+\blandscape
+
+\begin{figure}
+
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendDI-1}SPI (Standardized Precipitation Index)}
+\end{minipage}%
+%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendDI-2}SPEI (Standardized Precipitation
+Evapotranspiration Index)}
+\end{minipage}%
+\newline
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendDI-3}EDDI (Evaporative Demand Drought
+Index)}
+\end{minipage}%
+%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_zcSM_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendDI-4}SSMI (Standardized Soil Moisture
+Index)}
+\end{minipage}%
+
+\caption{\label{fig-trendDI}Linear trend of the drought index (*) at
+time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023}
+
+\end{figure}
+
+\elandscape
+
+\begin{table}[!ht]
+\caption{Summarry per land cover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
+The numbers in each cell indicate the time scale that reached the maximum correlation for the land cover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
+\label{tab-corlandcover}
+\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
+\end{table}
 
@@
 According to what is shown in Figure~\ref{fig-corTimeScale},
-Figure~\ref{fig-corPerson}, and Table \ref{tab-corland cover}, forest
+Figure~\ref{fig-corPerson}, and Table \ref{tab-corlandcover}, forest
 seems to be the most resistant type to drought. Showing that only
@@
 
-\newpage
-
-\blandscape
-
-\begin{figure}
-
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-1}SPI (Standardized Precipitation Index)}
-\end{minipage}%
-%
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-2}SPEI (Standardized Precipitation
-Evapotranspiration Index)}
-\end{minipage}%
-\newline
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-3}EDDI (Evaporative Demand Drought
-Index)}
-\end{minipage}%
-%
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_zcSM_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-4}SSMI (Standardized Soil Moisture
-Index)}
-\end{minipage}%
-
-\caption{\label{fig-trendDI}Linear trend of the drought index (*) at
-time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023}
-
-\end{figure}
-
-\elandscape
-
-\begin{table}[!ht]
-\caption{Summarry per land cover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
-The numbers in each cell indicate the time scale that reached the maximum correlation for the land cover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
-\label{tab-corland cover}
-\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
-\end{table}
-
 \begin{figure}[!ht]
@@
 
-\citep{Vicente-Serrano2021}, in a study at the global scale of drought
+\citet{Vicente-Serrano2021}, in a study at the global scale of drought
 trends, indicates that there have not been significant trends in
@@
 specifically to the rapidly increasing irrigated area, which
-consequently increases water extraction. \citep{Kogan2020} analyzed the
+consequently increases water extraction. \citet{Kogan2020} analyzed the
 agricultural drought impact globally and in the main grain producer
@@
 These changes are associated with a decrease in water demand from
-vegetation. Nonetheless, we used the persistant landcover to ensure that
-the pixel has the same class; in the case of croplands, it could happen
-that some areas had changed crops for others with higher water
+vegetation. Nonetheless, we used the persistant land cover to ensure
+that the pixel has the same class; in the case of croplands, it could
+happen that some areas had changed crops for others with higher water
 consumption. But this effect should be minor compared to the results
@@
 show that croplands are most sensitive to short-term drought
-(\textless{} SPI-6) \citep{Zambrano2016, Potopova2015, Dai2020}.
-Short-term precipitation deficits impact soil water, and thus less water
-is available for plant growth. However, we found that in ``Norte
-Chico,'' an SPI-36 and SPEI-12 had a higher impact, which are associated
-with hydrological drought (long-term), and in ``Centro,'' an SPI-12 and
+(\textless{} SPI-6)
+\citep{Zambrano2016, Potopova2015, Dai2020, Rhee2010}. Short-term
+precipitation deficits impact soil water, and thus less water is
+available for plant growth. However, we found that in ``Norte Chico,''
+an SPI-36 and SPEI-12 had a higher impact, which are associated with
+hydrological drought (long-term), and in ``Centro,'' an SPI-12 and
 SPEI-12. Thus, we attribute this impact to the hydrological drought that
@@
 between drought indices and vegetation productivity decrease, as do the
-time scales that reach the maximum r-squared (\ref{tab-corland cover}).
+time scales that reach the maximum r-squared (\ref{tab-corlandcover}).
 What can be explained is that, south of ``Centro,'' predominate forest
@@
 Extreme drought conditions are an important driver of tree mortality, as
-shown by \citep{Senf2020} in Europe. However, we found that forest is
-the type of landcover macroclass less affected by variation in drought
-indices, being the most resistant landcover class to drought. Supporting
-this is \citep{Fathi-Taperasht2022}, who asserts that Indian forests are
-the most drought-resistant and recover rapidly. Similarly, the work of
-\citep{Wu2024}, who analyzed vegetation loss and recovery in response to
-meteorological drought in the humid subtropical Pearl River basin in
-China, indicates that forests showed higher drought resistance. Using
-Vegetation Optical Depth (VOD), kNDVI, and EVI, \citep{Xiao2023}, tests
-the resistance of ecosystems and finds that ecosystems with more forests
-are better able to handle severe droughts than croplands. They attribute
-the difference to a deeper rooting depth of trees, a higher water
-storage capacity, and different water use strategies between forest and
-cropland \citep{Xiao2023}.
-
-In contrast to our observations, \citep{Venegas2022}, who investigated
-Cryptocarya alba and Beilschmiedia miersii (both members of the
-Lauraceae family) inhabiting sclerophyllous forests in Chile, concluded
-that the overall development of the trees had decelerated, potentially
-indicating alterations in the inherent dynamics of their respective
-forests. They attributed it to the cumulative effects of the
-unprecedented drought (i.e., hydrological drought). Nonetheless, the
-trees within sclerophyllous forests are highly heterogeneous, and the
-species analyzed by \citep{Venegas2022} are not the predominant ones.
-Because of this, when we used MOD13A3 with a 1km spatial resolution to
-measure vegetation condition, it took the average condition of 1 square
-kilometer. Thus, we attribute that forest to being the most resistant to
+shown by \citet{Senf2020} in Europe. However, we found that forest is
+the type of land cover macroclass less affected by variation in drought
+indices, being the most resistant land cover class to drought.
+Supporting this is \citet{Fathi-Taperasht2022}, who asserts that Indian
+forests are the most drought-resistant and recover rapidly. Similarly,
+the work of \citet{Wu2024}, who analyzed vegetation loss and recovery in
+response to meteorological drought in the humid subtropical Pearl River
+basin in China, indicates that forests showed higher drought resistance.
+Using Vegetation Optical Depth (VOD), kNDVI, and EVI, \citet{Xiao2023},
+tests the resistance of ecosystems and finds that ecosystems with more
+forests are better able to handle severe droughts than croplands. They
+attribute the difference to a deeper rooting depth of trees, a higher
+water storage capacity, and different water use strategies between
+forest and cropland \citep{Xiao2023}.
+
+In contrast to what we obtained, \citet{Venegas2022}, who studied
+Cryptocarya alba and Beilschmiedia miersii (both from the Lauraceae
+family) that live in sclerophyllous forests in Chile, found that the
+trees' overall growth had slowed down. This could mean that the natural
+dynamics of their forests have changed. They attributed it to the
+cumulative effects of the unprecedented drought (i.e., hydrological
+drought). Thus, we attribute that forest to being the most resistant to
 drought, due to the fact that most of the species comprising it are
-highly resilient to water scarcity. Then, for a study of the dynamics of
-forest to drought, e.g., sclerophyllous forest, on a local scale using
-remote sensing, we should use operational products at higher spatial
-resolutions as provided by the missions Landsat or Sentinel that could
-capture the heterogenity of the forest type.
-
-\hypertarget{drought-indices-of-water-demand-and-supply-soil-moisture-to-predict-changes-in-vegetation-productivity}{%
-\subsection{Drought indices of water demand and supply, soil moisture to
-predict changes in vegetation
-productivity}\label{drought-indices-of-water-demand-and-supply-soil-moisture-to-predict-changes-in-vegetation-productivity}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  Como podrían servir estos resultados para desarrollar o mejorar un
-  predictor de productividad de la vegetación.
-\end{enumerate}
-
-\begin{itemize}
-\tightlist
-\item
-  Los datos ERA5L están casi en tiempo real, 7 dias; MODIS también.
-\item
-  EL SSI se ve como un poderoso indicador que explica la variabilidad en
-  la productividad de la vegetación.
-\end{itemize}
-
-\hypertarget{future-outlook}{%
-\subsection{Future outlook}\label{future-outlook}}
-
-4.- Qué se podría hacer mejor en futuras investigaciones del tema. -
-mejorar la resolución y calidad de los datos climáticos -
+highly resilient to water scarcity compared to the other land cover
+classes. Nonetheless, if we want to go deep in our analysis, we should
+use earth observation data that is able to capture a higher level of
+detail. For example, when we used MOD13A3 with a 1km spatial resolution
+to measure vegetation condition, it took the average condition of 1
+square kilometer. Then, to study how a type of forest (e.g.,
+sclerophyllous forest) changes in response to drought on a local level
+using remote sensing, we should use operational products with higher
+spatial resolutions, such as those from Landsat or Sentinel.
+
+\hypertarget{soil-moisture-vegetation-productivity-and-agricultural-drought.}{%
+\subsection{Soil moisture, vegetation productivity, and agricultural
+drought.}\label{soil-moisture-vegetation-productivity-and-agricultural-drought.}}
+
+The main external factors that affect biomass production by vegetation
+are ET and SM, and the rate of ET in turn depends on the availability of
+water storage in the root zone. Thus, soil moisture plays a key role in
+land carbon uptake and, consequently, in the production of biomass
+\citep{Humphrey2021}. Moreover, \citet{Zhang2022} indicates there is a
+bidirectional causality between soil moisture and vegetation
+productivity. Lastly, some studies have redefined agricultural drought
+as soil moisture drought from an hydrological perspective
+\citep{Loon2016, Samaniego2018}. Even though soil moisture is the
+external factor most determinant of vegetation biomass, there are
+multiple internal factors, such as species, physiological
+characteristics, and plant hydraulics, that would affect vegetation
+productivity. Because of that, we believe that agricultural drought,
+referring to the drought that impacts vegetation productivity, is the
+most proper term, as originally defined by \citet{Wilhite1985}.
+
+The study results showed that the soil moisture-based drought index
+(SSI) was better at explaining vegetation productivity across land cover
+macroclasses than meteorological drought indices like SPI, SPEI, and
+EDDI. In the early growing season and especially in irrigated rather
+than rainfed croplands, soil moisture has better skills than SPI and
+SPEI for estimating gross primary production (GPP), according to
+\citet{Chatterjee2022}'s evaluation of the SPI and SPEI and their
+correlation with GPP in the CONUS. Also, \citet{Zhou2021} indicates that
+the monthly scaled Standardized Water Deficit Index (SWDI) can
+accurately show the effects of agricultural drought in most of China.
+Complementary, \citet{Nicolai2017} analyzed the time-lag between SWDI
+and the Vegetation Condition Index (VCI), and they state that there was
+no or little time-lag with croplands but a significant time-lag in the
+case of forests.
+
+In our case, there is strong spatial variability throughout Chile and
+between classes, mainly attributable to climate heterogeneity,
+hydrological status, or vegetation resistance to water scarcity. The
+semi-arid ``Norte Chico'' and the Mediterranean ``Centro'' were where
+SSI had the best performance. In Chile, medium-term deficits of 12
+months are more relevant in the response of vegetation, which decreases
+to the south, and in the case of croplands, they seem to react in a
+shorter time, with six months (SSI-6) in ``Centro.'' This variation for
+croplands could be related to the fact that in ``Norte Chico,'' the
+majority of crops are irrigated, but to the south there is a higher
+proportion of rainfed agriculture, which is most dependent on the
+short-term availability of water. Rather, in the ``Norte Chico,'' the
+orchards are more dependent on the storage of water in dams of
+groundwater reservoirs, which are affected by long-term drought (e.g.,
+SPI-36).
+
+\hypertarget{early-drought-forecasting-for-vegetation-productivity}{%
+\subsection{Early drought forecasting for vegetation
+productivity}\label{early-drought-forecasting-for-vegetation-productivity}}
+
+We analyzed the correlation between meteorological and soil moisture
+drought indices with zcNDVI. From our findings, we could further use the
+drought indices and time scales that have the highest r-squared to
+develop a combined model to forecast vegetation productivity in Chile.
+Despite the fact that SSI was the best-performing index, the rest of the
+indices should help to enhance predictability. \citet{Zambrano2018}
+proposed a prediction model for cropland surface in Chile at
+administrative units with a 1- to 6-month lead time using zcNDVI from
+MODIS and climate oscilation indices. The results given were r-squared,
+ranging from 0.95 at a 1-month lead time to 0.37 at a 6-month lead time.
+Thus, incorporating the results of this study with those made by
+\citet{Zambrano2018}, we could develop a combined forecasting model at
+the pixel level for the macroclasses in Chile based on drought indices
+of water demand and supply and soil moisture.
+
+\hypertarget{future-outlook-to-complete}{%
+\subsection{Future outlook (to
+complete)}\label{future-outlook-to-complete}}
 
@@
 
-\newpage
+There is a trend toward decreasing water supply in most parts of Chile,
+less in the ``Austral,'' which is stronger in the ``Centro'' and ``Norte
+Chico.'' The whole country showed an increase in water demand.
+Vegetation productivity only showed a decrease in the ``Norte Chico''
+and ``Centro,'' being highest for shrubland and croplands. Forest is the
+land cover most resistant to drought, and shrubland and cropland are the
+most sensitive.
+
+A soil moisture deficit of 12 months (SSI-12) is highly correlated with
+vegetation productivity for the land cover classes of shrubland,
+savannas, croplands, and forest in ``Norte Chico'' and ``Centro.'' For
+the southern part of the country with humid conditions, the correlation
+with SSI decreases. Soil moisture overcomes the capacity to explain
+vegetation productivity over the supply and demand drought indices in
+the entire territory.
+
+The variation in vegetation productivity appears to be associated with
+climate variation rather than anthropogenic factors (e.g., an increase
+in water demand by irrigated crops). Even though switching to more
+demanding crops on the land could increase the impact of drought on
+vegetation, this would need to be more thoroughly investigated, for
+instance at the watershed level.
+
+The results of this study could help in the development of a robust
+forecasting system for land cover classes in Chile., helping to improve
+preparedness for climate change impacts on vegetation.
+
+\hypertarget{supplementary-material}{%
+\section*{Supplementary material}\label{supplementary-material}}
+\addcontentsline{toc}{section}{Supplementary material}
 

### 2024-03-12 article/paper_drought_Chile2023.qmd
update version1 article

@@
 
-## Software and packages used
-
-For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, `R` [@R2023].
-For downloading ERA5L, we used the `{ecmwfr}` package [@Hufkens2019].
-For processing raster data, we used `{terra}` [@Hijmans2023] and `{stars}` [@Pebesma2023].
-For managing vectorial data, we used `{sf}` [@Pebesma2018].
-For the calculation of AED, we used `{SPEI}` [@Bergueria2023].
-For mapping, we use {tmap} [@Tennekes2018].
-For data analysis, the suite {tidyverse} [@Wickham2019] was used.
-
 ##	Data
 
-###	Earth observation data
+### Gridded meteorological and vegetation data
 
-For water supply and demand variables, we used ERA5L [@MunozSabater2021], a reanalys dataset that provides the evolution of land variables since 1950.
-It has a spatial resolution of 0.1°, hourly frequency, and global coverage.
-We selected the variables for total precipitation, 2 meter temperature maximum and minimum, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).
-The data was downloaded using the Copernicus Climate Data Store (CDS) Application Program Interface (API) implemented in `{ecmfwr}` [@Hufkens2019].
-
-To derive a proxy for vegetation productivity, we used the product MOD13A3 collection 6.1 from MODIS [@Didan2015].
+To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS.
+This product has a yearly frequency from 2001 to 2022.
+The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
+Thus, we have a land cover raster time series with the ten classes for 2001 and 2023.
+Prior to using this class, we compared it with the types of land cover made by @Zhao2016, which is a more detailed land cover map of Chile with a 30 m spatial resolution for 2012–2013.
+It reached a global accuracy of ~0.82.
+The procedure of validation is demonstrated in Section S1 of the supplementary material.To derive a proxy for vegetation productivity, we used the MOD13A3 collection 6.1 product from MODIS [@Didan2015].
 It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency.
-The MOD13A3.061 and MCD12Q1.061 were retrieved from the online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, https://lpdaa.usgs.gov/tools/data-pool/.
+The MOD13A3 and MCD12Q1 were retrieved from the online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, https://lpdaa.usgs.gov/tools/data-pool/.
 
@@
 
-###	Weather stations
+For soil moisture, water supply, and water demand variables, we used ERA5L [@MunozSabater2021], a reanalysis dataset that provides the evolution of the meteorological and soil moisture variables since 1950.
+It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage.
+We selected the variables for total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).
 
-We compared the ERA5L variables for monthly mean temperature, total precipitation, and volumetric soil water content against values retrieved by weather stations.
-For temperature and precipitation, we used the weather network from the Ministry of Agriculture of Chile (www.agromet.com) between 2015 and 2023.
-We used 277 stations located throughout Chile.
-For soil moisture, we select a private soil network that is owned by the agricultural enterprise Garces Fruit, which has 99 stations in Central Chile, located in cherry fruit crops.
-The sensors are installed at 30, 60, and 90m and are the model Teros 12 from MeterGroup.
-To avoid the effect of irrigation on soil moisture, which ERA5L hardly captures, we used daily data for the year 2022 and the months outside the growing season, May to September.
+## Land cover change and trend
 
-### Validation of ERA5L variables
+To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS.
+@Zambrano2018 and @Fuentes2021 have previously used this product for studies of drought and land cover.
+The MCD12Q1 has a yearly frequency from 2001 to 2022.
+The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
+Thus, we have a land cover raster time series with the ten macroclasses for 2001 and 2023.
+We validate the land cover macroclasses regarding a highly detailed (30 m of spatial resolution) land cover map made for Chile by @Zhao2016 for 2013-2014.
+Our results showed a global accuracy of ~0.82 and a F1 score of ~0.66.
+Section S2 in the Supplementary Material shows the procedure for validation.
 
-To account for the performance of the ERA5L climatic variables regarding the values measured by the weather stations. We selected the following metrics:
+Climate, vegetation development, seasonality, and changes in vegetation type all have an impact on the time series of NDVI.
+In this study, we want to examine the variation in vegetation productivity across various land cover types and how water demand, water supply, and soil moisture affect it.
+In order to avoid changes due to a change in the land cover type that will wrongly impact NDVI, we developed a persistence mask for land cover for 2001–2022.
+Thereby, we reduce an important source of variation on a regional scale.
+Therefore, we generated a raster mask for IGBP MODIS per pixel using macroclasses that remained unchanged for at least 80% of the years (2001–2022).
+This enabled us to identify regions where the land cover macroclasses are persistent.
+We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023.
+After that, we calculated the trend’s change in surface per type.
+We used the Sen’ slope [@Sen1968] based on Mann-Kendall [@Kendall1975].
 
-$$MAE = \frac{1}{n}\sum |{E-S}|$$
-$$Bias = \frac{\sum E}{\sum S}$$
-$$ubRMSE =\sqrt{\frac{\sum{ \left[ (E_i-\overline{E})-(S_i-\overline{S}) \right ] ^2}}{n}}$$
-
-$$CC = \frac{\sum (S_i-\bar{S})(E_i-\bar{E})}{\sqrt{(Si-\bar{S})^2(E_i-\bar{E})^2}}$$
-
-$MAE$: mean absolute error
-$bias$: bias
-$ubRMSE$: unbiassed root mean squared error
-$CC$: coefficient of correlation
-$S$: value of the variable measure by the weather station
-$E$: value of the variable measure by ERA5L
-
-## Drought Indices
+## Trend of drought and land cover change
 
@@
 
-For the indices EDDI and SPEI that use water demand, first we have to calculate the AED. For this, we used the method of Hargreaves [@Hargreaves1994;@Hargreaves1985]:
+In order to compute the drought indices that uses water demand, it is necessary to first calculate the AED.
+To do this, we employed the Hargreaves technique [@Hargreaves1994;@Hargreaves1985], by applying the following equation:
 
@@
 We chose the method of Hargreaves to estimate AED because of its simplicity, which only requires temperatures and extrarrestrial radiation.
-Also, it has been recommended over other methods when the use of several climatic variables is limited [@Vicente-Serrano2014].
+Also, it has been recommended over other methods (e.g., Penman-Monteith) when the access to climatic variables is limited [@Vicente-Serrano2014].
 
@@
 
-We derived the drought indices of water supply and demand, soil moisture from the ERA5L dataset, and vegetation from the MODIS product, all at monthly frequency.
+To derive the drought indices of water supply and demand, soil moisture, and vegetation we used the ERA5L dataset and the MODIS product, with a monthly frequency for 1981-2023 and 2000-2023, respectively.
 
-To evaluate water demand, we chose the $EDDI$ [@Hobbins2016; @McEvoy2016] index, which uses the $AED$.
-We used the SPI [@Mckee1993], a drought monitoring index that the World Meteorological Organization (WMO) recommends.
-We calculated the SPEI, which used a balance between $P$ and $AED$, in this case, an auxiliary variable $D = P-AED$ is used.
-In this study, we used the $SSI$ (standardized soil moisture index at 1 m) [@Hao2013;@AghaKouchak2014], which uses soil moisture at 1m depth.
-Finally, for the proxy of productivity, $zcNDVI$, we used the NDVI.
-Before using the NDVI, it was smoothed using a locally weighted polynomial regression, following the procedure described in @Zambrano2018 and @Zambrano2016.
+The dought indices correspond to a historical anomaly with regard to a variable (e.g., meteorological, vegetation, or soil moisture).
+To account for the anomaly, the common practice is to derive it following a statistical parametric methodology in which it is assumed that the statistical distribution of the data is known (@Heim2002).
+A wrong decision is usually the highest source of uncertainty (@Laimighofer2022).
+In the case of Chile, due to its high degree of climatic variability, it is complex to choose a proper distribution without previous research.
+Here, we follow a non-parametric methodology for the calculation of the drought indices, in a similar manner as the framework proposed by @Farahmand2015; @Hobbins2016;@McEvoy2016.
 
-All the indices are multi-scalar and were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was calculated for 6 months.
-The goal is to be able to evaluate short- and long-term droughts in water demand and supply and soil moisture.
-This is particularly important for central Chile because it has suffered from a prolonged decrease in precipitation for more than 12 years [@Garreaud2020;@Boisier2018;@Garreaud2017].
+For the purpose of monitoring water supply drought, we used the well-known Standardized Precipitation Index (SPI), which the World Meteorological Organization (WMO) recommended.
+The SPI solely relies on precipitation data.
+Also, it has been used worldwide for the study of drought, including in Chile (@Garreaud2017; @Zambrano2017).
+The primary cause of drought is precipitation anomalies, and temperature usually makes it worse (Luo et al.
+2017).
+Nowadays, there is an increase in attention toward using water demand separately to monitor droughts.
+(@Vicente-Serrano2020; @Noguera2022).
+Thus, to evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI), developed by Hobbins et al.
+(016) and McEvoy et al.
+(2016), which is based on the AED.
+EDDI is currently used for monitoring drought in the United States (https://psl.noaa.gov/eddi/).
+In our case, we used only temperature for AED, a difference from the original formulation of EDDI, which also considered wind besides temperature.
+To consider the combined effect of water supply and demand, we selected the SPEI, which corresponds to a balance between precipitation and AED.
+@Vicente-Serrano2010 proposed the SPEI, and it has improved the SPI by incorporating temperature for drought monitoring.
+For SPEI, an auxiliary variable D = P-AED is calculated.
+Soil moisture is the main driver of vegetation productivity, particularly in semi-arid regions (@Li2022).
+Hence, for soil water drought, we used the SSI (Standardized Soil Moisture Index) (Hao and AghaKouchak 2013; A.
+AghaKouchak 2014) which is a multi-scale index similar to SPI, SPEI, and EDDI.
+In our case, for the SSI, we used the average soil moisture from ERA5L at 1m depth.
+Finally, for the proxy of productivity, we used the zcNDVI proposed by Zambrano et al.
+(2018) which will be derived from the NDVI retrieved from MOD13A1.
 
-To calculate the drought indices, first we must calculate the accumulation of the variable.
-In this case, for generalization purposes, we will use $V$, referring to $P$, $AED$, $D$, $NDVI$, and $SM$ (Table \ref{tab-desEOD}).
-We cumulated each $V$ over the time series of $n$ values, and for the time scales $s$:
+To derive the drought indices, first we must calculate the sum of the variables with regard to the time scale (s).
+In this case, for generalization purposes, we will use $V$, referring to variables $P$, $AED$, $D$, $NDVI$, and $SM$ (#ref{tab-desEOD}).
+We cumulated each $V$  over the time series of $n$ values (months), and for the time scales $s$:
 
@@
 
-It corresponds to a moving window (convolution) that sums the variable for $s$ starting for the last month $n$ until the month, which could sum for $s$ months (n-s+1).
-Once the variable is cumulated over time for the scale $s$, we used a nonparametric approach following @Hobbins2016 to derive the drought indices.
+The $A_{si}$ corresponds to a moving window (convolution) that sums the variable for time scales $s$ from the last month, month by month, until the first month in which it could sum for $s$ months.
+Once the variable is cumulated over time for the scale $s$.
 Thus, the empirically derived probabilities are obtained through an inverse normal approximation [@Abramowitz1968].
-Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i$ to derive the $P(A_i)$ probabilities across a period of interest:
+Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i$ to derive the $P(a_i)$  probabilities across a period of interest:
 
@@
 
-##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture {#sec-methods_lulc}
+The drought indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months at a monthly frequency for 1981–2023 in order to be used for short- to long-term evaluation of drought.
+In the case of the proxy of vegetation productivity (zcNDVI) it was calculated for a time scale of six months at monthly frequency for 2000–2023.
+For zcNDVI, we test time scales of 1, 3, 6, and 12 months; we choose to use six months because that shows a more robust representation of vegetation productivity due to the seasonality of vegetation in Chile.
 
-### Land cover macroclasess and validation
-
-To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS.
-This product has a yearly frequency from 2001 to 2022.
-The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
-Thus, we have a land cover raster time series with the ten classes for 2001 and 2023.
-
-To validate the land cover obtained, we compare the macroclasses with the ones of a more detailed land cover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile).
-The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses land cover.
-We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from land cover macrolcasses to validate with LCChile.
-
-We follow the next procedure:
-
-i) resampled LCChile to the spatial resolution (500m) of the land cover macroclasses using the nearest neighbor method,
-ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values from the other two years as predictions.
-iii) calculate a confusion matrix with the classes extracted from the 1000 points for LCChile, IGBP2013, and IGBP2014. Calculate the performance metrics of accuracy and F1.
-
-$$Accuracy = \frac{TP+TN}{TP+TN+FP+FN}=\frac{correct\,\, classifications}{all\,\, classifications}$$
-$$F1=\frac{2\cdot TP}{2\cdot TP + FP +FN}$$
-where $TP$ and $FN$ refer to true positive and false negative, correctly classified classes; $TN$ and $FP$ to true negative and false positive, wrongly classified classes.
-
-### Land cover persistence mask 2001-2022
-
-Climate, vegetation development, seasonality, and changes in vegetation type all have an impact on the time series of NDVI.
-In this study, we want to examine the variation in vegetation productivity across various land cover types and how water demand, water supply, and soil moisture affect it.
-In order to avoid changes due to a change in the land cover type that will wrongly impact NDVI, we will develop a persistence mask for land cover for 2001–2023.
-Thereby, we reduce an important source of variation on a regional scale.
-
-Therefore, we generated a raster mask for IGBP MODIS per pixel using macroclasses that remain constant for at least 80% of the years (2001–2022).
-This enables us to identify regions where the land cover of the macroclasses has remained constant.
-
-### Land cover trend and drought indices
-
-We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023.
-After that, we calculated the trend's change in surface; we used the Sen' slope [@Sen1968] based on Mann-Kendall [@Kendall1975].
-This way, we obtain a matrix of trends of 5 x 5 (macrozones x land cover).
-The aim is to later explore if the trend in land cover classes is associated with a trend in the drought indices.
-For this, we will use the techniques of regresion and regularization of Lasso [@Tibshirani2010] and Ridge [@Hoerl1970].
-Also, we will test random forests for this purpose [@Ho1995].
-We will choose the trend of land cover surface per macroclass and macrozone as the response variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor variables.
-With this analysis, we expect to gather insights regarding whether there is a pattern of climatic influence along Chile or if what is happening in Central Chile has to do with more localized climatic conditions.
-
-## Trend of drought indices for water demand and supply, soil moisture, and vegetation productivity
-
-### Mann-Kendall and Sen's slope
+### Trend of drought indices
 
 To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975].
-To determine the magnitude of the trend, we used Sen's slope [@Sen1968].
-Some of the advantages of applying this methodology are that the Sen's slope is not affected by outliers as regular regression does, and it is a non-paramteric method that is not affected by the distribution of the data.
-We applied both to the six time scales from 1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI.
-In the case of zcNDVI (six months) was for 2000 to 2023.
-Thus, we have 31 trends.
-Also, we extracted the trend aggregated by macrozone and land cover class, obtaining a table of 31x5x5 (drought indices trends x macrozone x land cover class).
-We will use this data in @sec-methods_lulc to analyze if there is a strong relationship between the trends of drought indices and land cover surface within continental Chile.
+To determine the magnitude of the trend, we used Sen’s slope [@Sen1968].
+Some of the advantages of applying this methodology are that the Sen’s slope is not affected by outliers as regular regression does, and it is a non-parametric method that is not influenced by the distribution of the data.
+We applied Mann-Kendall to see if the trend was significant and Sen's slope to estimate the magnitude of the trend.
+We did this to the six time scales from 1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI.
+Thus, we have trends per index and time scale (24 in total).
+Then, we extracted the trend aggregated by macrozone and per land cover persitent macroclasses.
 
-### Trend in vegetation productivity without land cover change
+## Drought impacts on vegetation productivity within land cover
 
-@Vicente-Serrano2021 made a global analysis of the drought's severity trend using SPI, SPEI, and the Standardized Evapotranspiration Deficit Index (SEDI; @Vicente-Serrano2018) to evaluate AED.
-They indicate that the increase in hydrological drought has been due to anthropogenic effects rather than climate change.
-This is because the global increase in AED did not explain the change in the spatial pattern of the hydrological drought.
-Also, they state that _"the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change"_.
-We will contrast this hypothesis with what is occurring in Chile.
-To achieve this, we will use the land cover class type that remains more than 80% of types for 2001–2022 to evaluate the trend on zcNDVI and use this as a mask where there are low changes.
+We analyzed the trend of vegetation productivity over the unchanged land cover macroclasses.
+This way, we tried to reduce the noise in the vegetation due to a change in land cover from year to year.
+To achieve this, we will use the persistent mask of land cover macroclasses, which are the types that remain more than 80% of the time for 2001–2022.
+We used this to evaluate the trend in zcNDVI per land cover class and macrozone.
 
-##	Impact for water supply and demand, and soil moisture in vegetation productivity within land cover types
-
-We analyze the drought indices of water demand and supply and soil moisture against vegetation to address: i) if short- or long-term time scales are most important in impacting vegetation through Chile; and ii) the strength of the correlation for the variable and the time scale.
-Then, we will summarize for each land cover class and macrozone.
+We examine the drought indices of water demand, water supply, soil moisture, and their connection with vegetation productivity to investigate two main questions: i) whether short-term or long-term time scales have a greater impact on vegetation across Chile and its specific regions; and ii) the spatial variation in the strength of the correlation between the variables and time scales.  Then, we will summarize for each land cover class and macrozone.
 Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses of vegetation: forest, cropland, grassland, savanna, and shrubland.
 
-An analysis is conducted on the linear correlation between the indices SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36 months, and zcNDVI.
+We conducted an analysis on the linear correlation between the indices SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36 months, and zcNDVI.
 The objective is to determine the impact of soil moisture and water demand and supply on vegetation productivity.
-We implemented a methodology similar to that of @Meroni2016 when comparing the SPI for meteorological drought to the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation), which served as an indicator for vegetation productivity.
-A pixel-to-pixel linear correlation analysis was performed for each index.
+We used a method similar to that used by @Meroni2017 which compared the SPI with the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation).
+A pixel-to-pixel linear correlation analysis was performed for each index within the persistent mask of land cover macroclasses.
 To begin, the Pearson coefficient of correlation is computed for each of the six time scales.
@@
 Subsequently, the Pearson correlation coefficient corresponding to the time scales at which the value peaked was extracted.
-As a result, for each index, we generated two raster maps: one containing the time scales and the other the correlation value.
+As a result, for each index, we generated two raster maps: 1) containing the raster with values of the time scales that reached the maximum correlation, and 2) having the value of the correlation obtained.
+
+## Software and packages used
+
+For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, `R` [@R2023].
+For downloading ERA5L, we used the `{ecmwfr}` package [@Hufkens2019].
+For processing raster data, we used `{terra}` [@Hijmans2023] and `{stars}` [@Pebesma2023].
+For managing vectorial data, we used `{sf}` [@Pebesma2018].
+For the calculation of AED, we used `{SPEI}` [@Bergueria2023].
+For mapping, we use {tmap} [@Tennekes2018].
+For data analysis, the suite {tidyverse} [@Wickham2019] was used.
 
@@
 
-## Data
-
-### Validation of ERA5L variables
-
-The average metrics of performance of ERA5L over the 266 weather stations were in the case of monthly temperature: $ubRMSE=1.06\,°C$, $MAE=1.131\,°C$, and $CC=0.963$, showing a good agreement, low error, and low overestimation.
-For cumulative monthly precipitation, $MAE=28.1\,mm$, $bias=1.93$, and $CC=0.845$, showing a high correlation and a 93% bias and being overestimated by ERA5L.
-In the case of the 97 soil moisture stations, we averaged for the three depths (30, 60, and 90m) and then compared it with volumetric water content at 1m derived from ERA5L.
-For this case, we made a daily comparison, having a $CC=0.71$, $RMSE=0.174\,m^3m^{-3}$, $MAE=0.167\,m^3m^{-3}$, and $bias=1.74$.
-The ERA5 soil moisture overestimate is 74%, but it has a kind of good correlation.
-For more detailed information, consult the supplementary material (SSX).
-
-##	LULC change for 2001-2022 and its relation with water supply and demand, and soil moisture
-
-### Land cover macroclasess and validation
-
-For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands.
-@fig-studyArea c shows the spatial distribution of the macroclasses through Chile for the year 2022.
-The validation of IGBP2013 and IGBP2014 with LCChile reached near the same metrics of performance, having an accuracy of ~0.82 and a F1 score of ~0.66 (see SS1).
-
-### Land cover persistence mask 2001-2022 {#sec-persistence_mask}
-
-@fig-studyArea d, shows the macroclasses of land cover persistance (80%) during 2021-2022, respectively (\ref{tab-landcoverSurf}).
-Within continental Chile, forest is the vegetation type with the highest surface area at 135,00 $km^2$, followed by grassland (73,176 $km^2$), savanna (54,410 $km^2$), shrubland (24,959 $km^2$), and cropland (3,100 $km^2$) (Table \ref{tab-landcoverSurf}).
-The macrozones with major LULCC for 2001–2022 were "Centro," "Sur," and "Austral" with 36%, 31%, and 34%, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend}); of its surface that changes the type of land cover.
-@fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.
+# Land cover change and trend
 
@@
 
-### Land cover trend and drought indices
+For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands.
+[@fig-studyArea]c shows the spatial distribution of the macroclasses through Chile for the year 2022.
+[@fig-studyArea]d shows the macroclasses of land cover persistance (80%) during 2021–2022, respectively (#ref{tab-landcoverSurf}).
+Within continental Chile, barren land is the land cover class with the highest surface area (277,870 km2).
+The largest type of vegetation, with 137,085 km2, is forest.
+Grassland (74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland (3,146 km2) are the other types (Table 2).
+The macrozones with major changes for 2001–2022 were "Centro," "Sur," and "Austral," with 36%, 31%, and 34% of their surface changing the type of land cover, respectively (@fig-studyArea and #ref{tab-landcoverTrend}).
+@fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.
 
+```{=latex}
 \begin{table}[!ht]
@@
 \end{table}
-
-The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna of 70 $km^2 year^{-1}$.
-In the "Centro" and "Sur," there are changes in the Chilean matorral, with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$), and an increase in shrubland and grassland.
-Showing a change for more dense vegetation types.
-It appears to be a shift in the area of cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.
-
-Further, we want to address whether the trend in land cover change for 2001–2023 is associated with trends in drought indices of water demand and supply and/or soil moisture for macrozone and land cover macroclasses.
-From the three methods tested, Ridge, Lasso, and Random Forest, neither gives significant results regarding whether the trend in a drought index for any time scale explains the trend in land cover change.
-Nevertheless, in "Norte Chico" and "Centro," there is a decrease in croplands and savanna and an increase in barren land, which is associated with the variation in drought indices.
-Mainly for a decrease in water supply (SPI and SSI) and an increase in water demand (EDDI).
-However, due to the high variability from north to south in Chile, the climatic conditions (arid, semi-arid, and humid), and the land cover type, we believe that only in those zones could the LULCC be driven to some degree by drought.
-
-##	Trend of drought indices for water demand and supply, soil moisture, and vegetation productivity
-
-```{r fig-zcNDVI_var}
-#| echo: false
-#| fig-pos: "!ht"
-#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023.
-Greener colors indicate a positive trend; reder colors correspond to a negative trend and a decrease in vegetation productivity.
-Grey colors indicate either no vegetation or a change in land cover type for 2001–2022.
-(b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile.
-Each horizontal panel corresponds to a macrozone from 'Norte Grande' to 'Austral'."
-#| out-width: "\\textwidth"
-knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png")
 ```
 
-In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.02 in the five land cover macroclasses, with respect to grassland and shrubland categories.
-There is a negative trend in "Norte Chico" with -0.04 and "Centro" with -0.02 per decade.
-In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04.
-In "Centro," shrubland reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per decade.
-This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023].
-Vegetation reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Norte Chico" and Centro" (Mega Drought).
-The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var).
-Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the Chilean matorral [@Fuentes2021] appears to be the region most affected by a negative trend in vegetation productivity.
+The "Norte Chico" shows an increase in barren land of 111 $km^2 yr^{-1}$  and a reduction in the class savanna of 70 $km^2 yr^{-1}$.
+In the "Centro" and "Sur," there are changes with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$ ), and an increase in shrubland and grassland.
+Showing a change for more dense vegetation types.
+It appears to be a shift in the area cultivated (croplands) from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," replacing the savanna lost (#ref{tab-landcoverTrend}).
 
-```{r fig-trendDIMacro}
-#| echo: false
-#| fig-pos: "!ht"
-#| fig-cap: "Trend per decade for the drought indices SPI, EDDI, SPEI, and SSI aggregated by macrozone."
-#| out-width: "\\textwidth"
-knitr::include_graphics('../output/figs/trend_macrozone_drought_indices.png')
-```
+## Trend on drought and its relation to land cover change
+
+### Trend of drought indices
+
+@fig-trendDI shows the spatial variation of the trend for the drought indices from short- to long-term scales.
+SPI and SPEI have a decreasing trend from "Norte Chico" to "Sur." However, there is an increasing trend in "Austral." The degree of the trend is stronger at higher time scales.
+The SSI indicates that in "Norte Grande," there are surfaces that have increased in the soutwest part and in the northeast have decreased, and is shown for all time scales.
+Similar to SPI and SPEI, SSI decreases at higher time scales.
+EDDI showed a positive trend for the whole of continental Chile, with a higher trend toward the north and a descending gradient toward the south.
+The degree of trend increases at higher time scales.
 
@@
 
-```{=latex}
-\begin{table}[!ht]
-\caption{Summarry per land cover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
-The numbers in each cell indicate the time scale that reached the maximum correlation for the land cover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
-\label{tab-corlandcover}
-\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
-\end{table}
+```{r fig-trendDIMacro}
+#| echo: false
+#| fig-pos: "!ht"
+#| fig-cap: "Trend per decade for the drought indices SPI, EDDI, SPEI, and SSI aggregated by macrozone."
+#| out-width: "\\textwidth"
+knitr::include_graphics('../output/figs/trend_macrozone_drought_indices.png')
 ```
 
-Analyzing the water supply, the macrozones that have the lowest trend are "Norte Chico" and "Centro," where the SPI, SPEI, and SSI show that it decreases at longer time scales due to the prolonged reduction in precipitation.
-At 36 months, it reaches trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and SSI (@fig-trendDI).
+In @fig-trendDIMacro, the averaged aggregation per macrozone, drought index, and time scale are shown.
+The macrozones that have the lowest trend are "Norte Chico" and "Centro," where the SPI, SPEI, and SSI show that it decreases at longer time scales.
+Potentially explained due to the prolonged reduction in precipitation that has affected the hydrological system in Chile.
+At 36 months, it reaches trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and SSI.
 For "Sur," the behavior is similar, decreasing at longer scales and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI.
-On the other hand, all macrozones show an increase in the trend in all the drought indices, with "Norte Grande" having the highest at 36 months (0.042 per decade).
-Because of this, the SPEI (which uses AED) reached its lowest value in "Norte Grande," with -0.03 at 36 months.
-Despite the other macrozones, "Austral" showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend (@fig-trendDI and @fig-trendDIMacro).
+"Norte Grande" has the highest trend at 36 months for EDDI (0.042 per decade), and "Centro" has the lowest for SPI and SPEI.
+In "Norte Grande" and "Norte Chico," which are in a semi-arid climate, it is evident that the EDDI has an effect on the difference between the SPI and SPEI index, which is not seen in the other macrozones.
+Contrary to the other macrozones, “Austral” showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend.
 
-##	Impact for water supply and demand, and soil moisture in vegetation productivity
+### Relationship between drought indices and land cover change
 
-According to what is shown in @fig-corTimeScale, @fig-corPerson, and Table \ref{tab-corlandcover}, forest seems to be the most resistant type to drought.
-Showing that only "Centro" is slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit (SSI-12).
-In the "Norte Chico" and to a lesser extent in the "Norte Grande," it is evident that a SSI-12 with a rsq = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34, respectively) have an impact on grasslands.
-However, this type was unaffected by soil moisture, water supply, or demand in macrozones further south.
-The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
-For savannas in "Norte Chico," the SSI-12 and SPI-24 reached an rsq of 0.74 and 0.58, respectively.
-This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (rsq = 0.45 in "Centro" and 0.2 in "Sur").
-In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45% and 66% of "Norte Chico." The type of land most impacted by climatic variation was shrubland, where soil moisture explained 59% and precipitation, 37%, in "Norte Chico" and "Centro," with SSI-12 being the most relevant variable, then SPI-36 in "Norte Chico" and SPI-24 in "Sur."
+We look at the relationship between drought index and land cover change in Figure 5 by comparing the trends in land cover change (in terms of the total surface area per land cover type and macrozone) and drought indexes.
+Figure 5 shows that the negative trends in cropland (-0.029) and savanna (-0.026) in “Norte Chico” are the highest and are associated with an increase trend in EDDI and a decreasing trend in SPI, SPEI, and SSI.
+On the contrary, the shrubland in “Centro” has an increase (0.04) linked to a decrease in SPI, SPEI, and SSI and to an increase in EDDI.
+In "Austral,” the positive trend in shrubland fits with the positive trend in all the drought indices.
+The rest of the land cover types in the macrozones show weak associations with water supply, water demand, or soil moisture drought indices.
+
+```{r figTrendsLandDrought}
+#| echo: false
+#| fig-pos: "!ht"
+#| fig-cap: "Relationship between the trend in land cover change and the trend in drought indices for the five macrozones.
+Vertical panels correspond to 1, 3, 6, 12, 24, and 36 months of the time scale by drought index.
+Horizontal panels show each drought index"
+#| out-width: "\\textwidth"
+knitr::include_graphics('../output/figs/points_landcover_drought_indices_trend_and_time_scale.png')
+```
+
+## Drought impacts on vegetation productivity within land cover
+
+### Vegetation productivity
+
+```{r fig-zcNDVI_var}
+#| echo: false
+#| fig-pos: "!ht"
+#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023.
+Greener colors indicate a positive trend; reder colors correspond to a negative trend and a decrease in vegetation productivity.
+Grey colors indicate either no vegetation or a change in land cover type for 2001–2022.
+(b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile.
+Each horizontal panel corresponds to a macrozone from 'Norte Grande' to 'Austral'."
+#| out-width: "\\textwidth"
+knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png")
+```
+
+In @fig-zcNDVI_var it is showed the spatial map of trends in zcNDVI ([fig-zcNDVI_var]a) and the temporal variation of zcNDVI within the aggregated macrozones ([fig-zcNDVI_var]b).
+In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.02 with respect to grassland and shrubland categories.
+There is a negative trend in "Norte Chico" with -0.04 and "Centro" with -0.02 per decade.
+In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04.
+In "Centro," shrubland reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per decade.
+This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023].
+Vegetation reached its lowest values since the year 2019, with an extreme condition in early 2020 and 2022 in the "Norte Chico" and "Centro".
+The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var).
+Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the savanna and shurbland appears to be the region most affected by a negative trend in vegetation productivity.
+
+### Correlation between vegetation productivity and drought indices
 
@@
 
+@fig-corTimeScale is a map that shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between different drought indicators and plant productivity over time.
+The spatial variation of time scales reached per index is mostly for time scales above 12 months.
+In the case of SSI, the predominant scales are 6 and 12 months.
+For all indices, to the north, the time scales are higher and diminish toward the south until the south part of "Austral" increases.
+In @fig-corPerson, the map of Pearson correlation values is shown.
+The EDDI reached correlations above 0.5 between "Norte Chico" and "Sur." The correlation changes from negative to positive toward the Andes Mountains and to the sea, just as in the northern part of "Austral." The SPI and SPEI have similar results, with the higher values in "Norte Chico" and "Centro" being higher than 0.6.
+Following a similar spatial pattern as EDDI.
+The SSI showed to be the index that has a major spatial extension with a higher correlation.
+It has a similar correlation to SPI and SPEI for "Norte Chico" and "Sur,"  but has improvements for "Sur."
+
+```{=latex}
+\begin{table}[!ht]
+\caption{Summarry per land cover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
+The numbers in each cell indicate the time scale that reached the maximum correlation for the land cover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
+\label{tab-corlandcover}
+\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
+\end{table}
+```
+
+In #ref{tab-corlandcover}, we aggregate per macrozone and landcover the correlation analysis presented in @fig-corTimeScale and @fig-corPerson.
+According to what is shown, forests seem to be the most resistant to drought.
+Showing that only "Centro" is slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit (SSI-12).
+In the "Norte Chico" and to a lesser extent in the "Norte Grande," it is evident that a SSI-12 with a rsq = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34, respectively) have an impact on grasslands.
+However, this type was unaffected by soil moisture, water supply, or demand in macrozones further south.
+The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
+For savannas in "Norte Chico," the SSI-12 and SPI-24 reached an rsq of 0.74 and 0.58, respectively.
+This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (rsq = 0.45 in "Centro" and 0.2 in "Sur").
+In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45% and 66% of the variability in "Norte Chico." The type of land most impacted by climatic variation was shrubland, where soil moisture explained 59% and precipitation, 37%, in "Norte Chico" and "Centro," with SSI-12 being the most relevant variable, then SPI-36 in "Norte Chico" and SPI-24 in "Sur."
+
 #	Discussion
 
-## Drought trend and attribution to LULCC
+## Drought trend and attribution to land cover
 
-@Vicente-Serrano2021, in a study at the global scale of drought trends, indicates that there have not been significant trends in meteorological drought since 1950.
+@Vicente-Serrano2022, in a study at the global scale of drought trends, indicate that there have not been significant trends in meteorological drought since 1950.
 Also, state that the increase in hidrological trend in some parts of the globe (northeast Brazil and the Mediterranean region) is related to changes in land cover and specifically to the rapidly increasing irrigated area, which consequently increases water extraction.
-@Kogan2020 analyzed the agricultural drought impact globally and in the main grain producer countries, finding that _"since 1980, the Earth warming has not changed the drought area or intensity"_.
-
-In our study, we considered the variation in vegetation productivity in Chile for areas without changes in land cover macroclasses (see @sec-persistence_mask), to avoid misleading conclusions that could be related to the increase in water demand due to LULCC.
+@Kogan2020 analyzed the agricultural drought impact globally and in the main grain producer countries, finding that _"since 1980, the Earth warming has not changed the drought area or intensity."_ In our study, we considered the variation in vegetation productivity in Chile for areas without changes in land cover, to avoid misleading conclusions that could be related to the increase in water demand due to land cover change.
 Our results show a contrasting perspective.
-There has been a significant trend in the decline of vegetation productivity (zcNDVI) since 2000 for "Norte Chico" and "Centro," which has been extreme between 2020 and 2022, seemsly due to an intense hydrological drought due to the persistance of the Mega Drought [@Garreaud2017].
+There has been a significant trend in the decline of vegetation productivity (zcNDVI) since 2000 for "Norte Chico" and "Centro," which has been extreme between 2020 and 2022, seemsly due to an intense hydrological drought due to the persistance of the mega drought [@Garreaud2017].
+However, a rise in irrigated land doesn't seem to have an impact on this hydrological drought.
 Despite using the persistance mask for vegetation's trend analysis, cropland, which is the most water-demand type, showed a decrease trend in "Norte Chico" and "Centro." Also, there was an increase in barren land for both types.
 These changes are associated with a decrease in water demand from vegetation.
-Nonetheless, we used the persistant land cover to ensure that the pixel has the same class; in the case of croplands, it could happen that some areas had changed crops for others with higher water consumption.
-But this effect should be minor compared to the results from ladcover macroclasses.
+Nonetheless, we used the persistant land cover to ensure that the pixel has the same class; in the case of croplands, it could happen that some areas had changed crops for others with higher water consumption and consequently increase water demand.
+But this effect should be minor compared to the results from land cover change at this scale of analysis.
 
-On the other hand, for "Norte Chico" and "Centro," our results show a decrease in trends of water supply (SPI and SSI), which are higher at larger time scales and consequently impact the hydrological system.
-We claim that what occurred in central Chile defies findings made at the global level [@Vicente-Serrano2021;@Kogan2020], demonstrating that a constant decrease in water supply rather than an increase in water demand (i.e., irrigated crops) is the main cause of the hydrological drought.
-Finally, central Chile has shown a diminishment in vegetation productivity for all macroclasses, mainly attributed to variation in water supply, i.e., precipitation, which could be strengthened by an increase in water demand by, for example, an increase in the surface area of irrigated crops.
+On the other hand, for "Norte Chico" and "Centro," our results show a decrease in trends of water supply (SPI and SSI), which are higher at larger time scales, which is evidence of the hydrological drought.
+We say that what happened in central Chile goes against what has been found on a global scale [@Vicente-Serrano2022;@Kogan2020].
+This shows that the main cause of the hydrological drought in Chile was a steady drop in water supply made worse by an increase in AED, but it seems that in zones most affected by drought, the main cause is not an increase in water demand by vegetation like irrigated crops.
+Finally, north-central Chile has experienced a decline in vegetation productivity across all macroclasses, which is primarily attributable to variations in water supply and soil moisture.
+An increase in water demand, such as an increase in the surface area of irrigated crops, could strengthen this trend.
 
@@
 
-We found that shrubland, savannas (Chilean matorral), and croplands are the most sensitive to climate conditions.
-Being most affected by the 12-month soil moisture deficit.
-In a study in the Yangtze River Basin in China, Jiang2020 analyzed the impact of drought on vegetation using the SPEI and the Enahanced Vegetation Index (EVI).
-They found that cropland was more sensitive to drought than cropland, showing that cropland responds strongly to short- and medium-term drought (< SPEI-6).
-In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro." In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016; @Potopova2015; @Dai2020;@Rhee2010].
+We discovered that croplands, savannas, and shrubland are the most susceptible to climatic changes and are most affected by the 12-month soil moisture deficit.
+In a study in the Yangtze River Basin in China, @Jiang2020 analyzed the impact of drought on vegetation using the SPEI and the Enahanced Vegetation Index (EVI).
+They found that cropland was more sensitive to drought than grassland, showing that cropland responds strongly to short- and medium-term drought (< SPEI-6).
+In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro." In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016;@Potopova2015;@Dai2020;@Rhee2010].
 Short-term precipitation deficits impact soil water, and thus less water is available for plant growth.
@@
 Thus, we attribute this impact to the hydrological drought that has decreased groundwater storage [@Taucare2024], which in turn is impacted by long-term deficits, and consequently, the vegetation is more dependent on groundwater.
-In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared (\ref{tab-corlandcover}).
+In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared.
 What can be explained is that, south of "Centro," predominate forest and grassland, the most resistant types.
@@
 
-Extreme drought conditions are an important driver of tree mortality, as shown by @Senf2020 in Europe.
+According to @Senf2020, severe drought conditions in Europe are a significant cause of tree mortality.
 However, we found that forest is the type of land cover macroclass less affected by variation in drought indices, being the most resistant land cover class to drought.
-Supporting this is @Fathi-Taperasht2022, who asserts that Indian forests are the most drought-resistant and recover rapidly.
+Supporting this is @Fathi-Taperasht2022], who assert that Indian forests are the most drought-resistant and recover rapidly.
 Similarly, the work of @Wu2024, who analyzed vegetation loss and recovery in response to meteorological drought in the humid subtropical Pearl River basin in China, indicates that forests showed higher drought resistance.
-Using Vegetation Optical Depth (VOD), kNDVI, and EVI, @Xiao2023, tests the resistance of ecosystems and finds that ecosystems with more forests are better able to handle severe droughts than croplands.
+Using Vegetation Optical Depth (VOD), kNDVI, and EVI, @Xiao2023 test the resistance of ecosystems and find that ecosystems with more forests are better able to handle severe droughts than croplands.
 They attribute the difference to a deeper rooting depth of trees, a higher water storage capacity, and different water use strategies between forest and cropland [@Xiao2023].
@@
 For example, when we used MOD13A3 with a 1km spatial resolution to measure vegetation condition, it took the average condition of 1 square kilometer.
-Then, to study how a type of forest (e.g., sclerophyllous forest) changes in response to drought on a local level using remote sensing, we should use operational products with higher spatial resolutions, such as those from Landsat or Sentinel.
+Then, to use remote sensing to look at how a certain type of forest (like sclerophyllous forest) changes in response to drought on a local level, we should use operational products with higher spatial resolutions, like those from Landsat or Sentinel.
+This will let us do a more thorough analysis.
 
@@
 
-The main external factors that affect biomass production by vegetation are ET and SM, and the rate of ET in turn depends on the availability of water storage in the root zone.
+The main external factors that affect biomass production by vegetation are actual evapotranspiration and soil moisture, and the rate of ET in turn depends on the availability of water storage in the root zone.
 Thus, soil moisture plays a key role in land carbon uptake and, consequently, in the production of biomass [@Humphrey2021].
-Moreover, @Zhang2022 indicates there is a bidirectional causality between soil moisture and vegetation productivity.
-Lastly, some studies have redefined agricultural drought as soil moisture drought from an hydrological perspective [@Loon2016;@Samaniego2018].
+Moreover, @Zhang2022 indicate there is a bidirectional causality between soil moisture and vegetation productivity.
+Lastly, some studies have redefined agricultural drought as soil moisture drought from a hydrological perspective [@Loon2016;@Samaniego2018].
 Even though soil moisture is the external factor most determinant of vegetation biomass, there are multiple internal factors, such as species, physiological characteristics, and plant hydraulics, that would affect vegetation productivity.
@@
 The study results showed that the soil moisture-based drought index (SSI) was better at explaining vegetation productivity across land cover macroclasses than meteorological drought indices like SPI, SPEI, and EDDI.
-In the early growing season and especially in irrigated rather than rainfed croplands, soil moisture has better skills than SPI and SPEI for estimating gross primary production (GPP), according to @Chatterjee2022's evaluation of the SPI and SPEI and their correlation with GPP in the CONUS.
-Also, @Zhou2021 indicates that the monthly scaled Standardized Water Deficit Index (SWDI) can accurately show the effects of agricultural drought in most of China.
-Complementary, @Nicolai2017 analyzed the time-lag between SWDI and the Vegetation Condition Index (VCI), and they state that there was no or little time-lag with croplands but a significant time-lag in the case of forests.
+In the early growing season and especially in irrigated rather than rainfed croplands, soil moisture has better skills than SPI and SPEI for estimating gross primary production (GPP).
+This according to @Chatterjee2022 evaluation of the SPI and SPEI and their correlation with GPP in the CONUS.
+Also, @Zhou2021 indicate that the monthly scaled Standardized Water Deficit Index (SWDI) can accurately show the effects of agricultural drought in most of China.
+@Nicolai2017 also looked at the time-lag between the SWDI and the Vegetation Condition Index (VCI).
+They found that there was little to no time-lag in croplands but a greater time-lag in forests.
 
@@
 
-## Early drought forecasting for vegetation productivity
-
-We analyzed the correlation between meteorological and soil moisture drought indices with zcNDVI.
-From our findings, we could further use the drought indices and time scales that have the highest r-squared to develop a combined model to forecast vegetation productivity in Chile.
-Despite the fact that SSI was the best-performing index, the rest of the indices should help to enhance predictability.
-@Zambrano2018 proposed a prediction model for cropland surface in Chile at administrative units with a 1- to 6-month lead time using zcNDVI from MODIS and climate oscilation indices.
-The results given were r-squared, ranging from 0.95 at a 1-month lead time to 0.37 at a 6-month lead time.
-Thus, incorporating the results of this study with those made by @Zambrano2018, we could develop a combined forecasting model at the pixel level for the macroclasses in Chile based on drought indices of water demand and supply and soil moisture.
-
 ## Future outlook (to complete)
@@
 
-There is a trend toward decreasing water supply in most parts of Chile, less in the "Austral," which is stronger in the "Centro" and "Norte Chico." The whole country showed an increase in water demand.
-Vegetation productivity only showed a decrease in the "Norte Chico" and "Centro," being highest for shrubland and croplands.
-Forest is the land cover most resistant to drought, and shrubland and cropland are the most sensitive.
+There is a trend toward decreasing water supply in most parts of Chile, particularly in the “Centro” and “Norte Chico” regions.
+The whole country showed an increase in AED.
+Vegetation productivity only showed a decrease in the “Norte Chico” and “Centro,” being highest for shrubland and croplands.
+Forest is the land cover most resistant to drought, as shown along Chile, and shrubland and cropland are the most sensitive.
 
-A soil moisture deficit of 12 months (SSI-12) is highly correlated with vegetation productivity for the land cover classes of shrubland, savannas, croplands, and forest in "Norte Chico" and "Centro." For the southern part of the country with humid conditions, the correlation with SSI decreases.
+A soil moisture deficit of 12 months (SSI-12) is highly correlated with vegetation productivity for the land cover classes of shrubland, savannas, croplands, and forest in “Norte Chico” and “Centro.” For the southern part of the country with humid conditions, the correlation with SSI decreases.
 Soil moisture overcomes the capacity to explain vegetation productivity over the supply and demand drought indices in the entire territory.
@@
 
-The results of this study could help in the development of a robust forecasting system for land cover classes in Chile., helping to improve preparedness for climate change impacts on vegetation.
+The results of this study could help in the development of a robust forecasting system for land cover classes in Chile, helping to improve preparedness for climate change impacts on vegetation.
 

### 2024-03-12 article/paper_drought_Chile2023.tex
update version1 article

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, boxrule=0pt, frame hidden, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, breakable, interior hidden, sharp corners, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{software-and-packages-used}{%
-\subsection{Software and packages
-used}\label{software-and-packages-used}}
-
-For the downloading, processing, and analysis of the spatio-temporal
-data, we used the open source software for statistical computing and
-graphics, \texttt{R} \citep{R2023}. For downloading ERA5L, we used the
-\texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing raster
-data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
-\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
-used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
-used \texttt{\{SPEI\}} \citep{Bergueria2023}. For mapping, we use
-\{tmap\} \citep{Tennekes2018}. For data analysis, the suite
-\{tidyverse\} \citep{Wickham2019} was used.
-
 \hypertarget{data}{%
@@
 
-\hypertarget{earth-observation-data}{%
-\subsubsection{Earth observation data}\label{earth-observation-data}}
-
-For water supply and demand variables, we used ERA5L
-\citep{MunozSabater2021}, a reanalys dataset that provides the evolution
-of land variables since 1950. It has a spatial resolution of 0.1°,
-hourly frequency, and global coverage. We selected the variables for
-total precipitation, 2 meter temperature maximum and minimum, and
-volumetric soil water layers between 0 and 100cm of depth (layer 1 to
-layer 3). The data was downloaded using the Copernicus Climate Data
-Store (CDS) Application Program Interface (API) implemented in
-\texttt{\{ecmfwr\}} \citep{Hufkens2019}.
-
-To derive a proxy for vegetation productivity, we used the product
-MOD13A3 collection 6.1 from MODIS \citep{Didan2015}. It provides
+\hypertarget{gridded-meteorological-and-vegetation-data}{%
+\subsubsection{Gridded meteorological and vegetation
+data}\label{gridded-meteorological-and-vegetation-data}}
+
+To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
+6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
+The IGBP defines 17 classes; from these, we regrouped into ten
+macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9
+to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
+13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
+Thus, we have a land cover raster time series with the ten classes for
+2001 and 2023. Prior to using this class, we compared it with the types
+of land cover made by \citet{Zhao2016}, which is a more detailed land
+cover map of Chile with a 30 m spatial resolution for 2012--2013. It
+reached a global accuracy of \textasciitilde0.82. The procedure of
+validation is demonstrated in Section S1 of the supplementary
+material.To derive a proxy for vegetation productivity, we used the
+MOD13A3 collection 6.1 product from MODIS \citep{Didan2015}. It provides
 vegetation indices (NDVI and EVI) at 1km of spatial resolution and
-monthly frequency. The MOD13A3.061 and MCD12Q1.061 were retrieved from
-the online Data Pool, courtesy of the NASA EOSDIS Land Processes
-Distributed Active Archive Center (LP DAAC), USGS Earth Resources
-Observation and Science (EROS) Center, Sioux Falls, South Dakota,
+monthly frequency. The MOD13A3 and MCD12Q1 were retrieved from the
+online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed
+Active Archive Center (LP DAAC), USGS Earth Resources Observation and
+Science (EROS) Center, Sioux Falls, South Dakota,
 https://lpdaa.usgs.gov/tools/data-pool/.
@@
 
-\hypertarget{weather-stations}{%
-\subsubsection{Weather stations}\label{weather-stations}}
-
-We compared the ERA5L variables for monthly mean temperature, total
-precipitation, and volumetric soil water content against values
-retrieved by weather stations. For temperature and precipitation, we
-used the weather network from the Ministry of Agriculture of Chile
-(www.agromet.com) between 2015 and 2023. We used 277 stations located
-throughout Chile. For soil moisture, we select a private soil network
-that is owned by the agricultural enterprise Garces Fruit, which has 99
-stations in Central Chile, located in cherry fruit crops. The sensors
-are installed at 30, 60, and 90m and are the model Teros 12 from
-MeterGroup. To avoid the effect of irrigation on soil moisture, which
-ERA5L hardly captures, we used daily data for the year 2022 and the
-months outside the growing season, May to September.
-
-\hypertarget{validation-of-era5l-variables}{%
-\subsubsection{Validation of ERA5L
-variables}\label{validation-of-era5l-variables}}
-
-To account for the performance of the ERA5L climatic variables regarding
-the values measured by the weather stations. We selected the following
-metrics:
-
-\[MAE = \frac{1}{n}\sum |{E-S}|\] \[Bias = \frac{\sum E}{\sum S}\]
-\[ubRMSE =\sqrt{\frac{\sum{ \left[ (E_i-\overline{E})-(S_i-\overline{S}) \right ] ^2}}{n}}\]
-
-\[CC = \frac{\sum (S_i-\bar{S})(E_i-\bar{E})}{\sqrt{(Si-\bar{S})^2(E_i-\bar{E})^2}}\]
-
-\(MAE\): mean absolute error \(bias\): bias \(ubRMSE\): unbiassed root
-mean squared error \(CC\): coefficient of correlation \(S\): value of
-the variable measure by the weather station \(E\): value of the variable
-measure by ERA5L
-
-\hypertarget{drought-indices}{%
-\subsection{Drought Indices}\label{drought-indices}}
+For soil moisture, water supply, and water demand variables, we used
+ERA5L \citep{MunozSabater2021}, a reanalysis dataset that provides the
+evolution of the meteorological and soil moisture variables since 1950.
+It has a spatial resolution of 0.1° (9 km), hourly frequency, and global
+coverage. We selected the variables for total precipitation, maximum and
+minimum temperature at 2 meters, and volumetric soil water layers
+between 0 and 100cm of depth (layer 1 to layer 3).
+
+\hypertarget{land-cover-change-and-trend}{%
+\subsection{Land cover change and
+trend}\label{land-cover-change-and-trend}}
+
+To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
+6.1 from MODIS. \citet{Zambrano2018} and \citet{Fuentes2021} have
+previously used this product for studies of drought and land cover. The
+MCD12Q1 has a yearly frequency from 2001 to 2022. The IGBP defines 17
+classes; from these, we regrouped into ten macroclasses, as follows:
+classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as
+grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as
+snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land
+cover raster time series with the ten macroclasses for 2001 and 2023. We
+validate the land cover macroclasses regarding a highly detailed (30 m
+of spatial resolution) land cover map made for Chile by \citet{Zhao2016}
+for 2013-2014. Our results showed a global accuracy of
+\textasciitilde0.82 and a F1 score of \textasciitilde0.66. Section S2 in
+the Supplementary Material shows the procedure for validation.
+
+Climate, vegetation development, seasonality, and changes in vegetation
+type all have an impact on the time series of NDVI. In this study, we
+want to examine the variation in vegetation productivity across various
+land cover types and how water demand, water supply, and soil moisture
+affect it. In order to avoid changes due to a change in the land cover
+type that will wrongly impact NDVI, we developed a persistence mask for
+land cover for 2001--2022. Thereby, we reduce an important source of
+variation on a regional scale. Therefore, we generated a raster mask for
+IGBP MODIS per pixel using macroclasses that remained unchanged for at
+least 80\% of the years (2001--2022). This enabled us to identify
+regions where the land cover macroclasses are persistent. We calculated
+the surface occupied per land cover class into the five macrozones
+(``Norte Grande'' to ``Austral'') per year for 2001--2023. After that,
+we calculated the trend's change in surface per type. We used the Sen'
+slope \citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}.
+
+\hypertarget{trend-of-drought-and-land-cover-change}{%
+\subsection{Trend of drought and land cover
+change}\label{trend-of-drought-and-land-cover-change}}
 
@@
 
-For the indices EDDI and SPEI that use water demand, first we have to
-calculate the AED. For this, we used the method of Hargreaves
-\citep{Hargreaves1994, Hargreaves1985}:
+In order to compute the drought indices that uses water demand, it is
+necessary to first calculate the AED. To do this, we employed the
+Hargreaves technique \citep{Hargreaves1994, Hargreaves1985}, by applying
+the following equation:
 
@@
 simplicity, which only requires temperatures and extrarrestrial
-radiation. Also, it has been recommended over other methods when the use
-of several climatic variables is limited \citep{Vicente-Serrano2014}.
+radiation. Also, it has been recommended over other methods (e.g.,
+Penman-Monteith) when the access to climatic variables is limited
+\citep{Vicente-Serrano2014}.
 
@@
 
-We derived the drought indices of water supply and demand, soil moisture
-from the ERA5L dataset, and vegetation from the MODIS product, all at
-monthly frequency.
-
-To evaluate water demand, we chose the \(EDDI\)
-\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). We used
-the SPI \citep{Mckee1993}, a drought monitoring index that the World
-Meteorological Organization (WMO) recommends. We calculated the SPEI,
-which used a balance between \(P\) and \(AED\), in this case, an
-auxiliary variable \(D = P-AED\) is used. In this study, we used the
-\(SSI\) (standardized soil moisture index at 1 m)
-\citep{Hao2013, AghaKouchak2014}, which uses soil moisture at 1m depth.
-Finally, for the proxy of productivity, \(zcNDVI\), we used the NDVI.
-Before using the NDVI, it was smoothed using a locally weighted
-polynomial regression, following the procedure described in
-\citet{Zambrano2018} and \citet{Zambrano2016}.
-
-All the indices are multi-scalar and were calculated for time scales of
-1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was calculated
-for 6 months. The goal is to be able to evaluate short- and long-term
-droughts in water demand and supply and soil moisture. This is
-particularly important for central Chile because it has suffered from a
-prolonged decrease in precipitation for more than 12 years
-\citep{Garreaud2020, Boisier2018, Garreaud2017}.
-
-To calculate the drought indices, first we must calculate the
-accumulation of the variable. In this case, for generalization purposes,
-we will use \(V\), referring to \(P\), \(AED\), \(D\), \(NDVI\), and
-\(SM\) (Table \ref{tab-desEOD}). We cumulated each \(V\) over the time
-series of \(n\) values, and for the time scales \(s\):
+To derive the drought indices of water supply and demand, soil moisture,
+and vegetation we used the ERA5L dataset and the MODIS product, with a
+monthly frequency for 1981-2023 and 2000-2023, respectively.
+
+The dought indices correspond to a historical anomaly with regard to a
+variable (e.g., meteorological, vegetation, or soil moisture). To
+account for the anomaly, the common practice is to derive it following a
+statistical parametric methodology in which it is assumed that the
+statistical distribution of the data is known (\citet{Heim2002}). A
+wrong decision is usually the highest source of uncertainty
+(\citet{Laimighofer2022}). In the case of Chile, due to its high degree
+of climatic variability, it is complex to choose a proper distribution
+without previous research. Here, we follow a non-parametric methodology
+for the calculation of the drought indices, in a similar manner as the
+framework proposed by \citet{Farahmand2015};
+\citet{Hobbins2016};\citet{McEvoy2016}.
+
+For the purpose of monitoring water supply drought, we used the
+well-known Standardized Precipitation Index (SPI), which the World
+Meteorological Organization (WMO) recommended. The SPI solely relies on
+precipitation data. Also, it has been used worldwide for the study of
+drought, including in Chile (\citet{Garreaud2017};
+\citet{Zambrano2017}). The primary cause of drought is precipitation
+anomalies, and temperature usually makes it worse (Luo et al.~2017).
+Nowadays, there is an increase in attention toward using water demand
+separately to monitor droughts. (\citet{Vicente-Serrano2020};
+\citet{Noguera2022}). Thus, to evaluate water demand, we chose the
+Evaporative Demand Drought Index (EDDI), developed by Hobbins et
+al.~(016) and McEvoy et al.~(2016), which is based on the AED. EDDI is
+currently used for monitoring drought in the United States
+(https://psl.noaa.gov/eddi/). In our case, we used only temperature for
+AED, a difference from the original formulation of EDDI, which also
+considered wind besides temperature. To consider the combined effect of
+water supply and demand, we selected the SPEI, which corresponds to a
+balance between precipitation and AED. \citet{Vicente-Serrano2010}
+proposed the SPEI, and it has improved the SPI by incorporating
+temperature for drought monitoring. For SPEI, an auxiliary variable D =
+P-AED is calculated. Soil moisture is the main driver of vegetation
+productivity, particularly in semi-arid regions (\citet{Li2022}). Hence,
+for soil water drought, we used the SSI (Standardized Soil Moisture
+Index) (Hao and AghaKouchak 2013; A. AghaKouchak 2014) which is a
+multi-scale index similar to SPI, SPEI, and EDDI. In our case, for the
+SSI, we used the average soil moisture from ERA5L at 1m depth. Finally,
+for the proxy of productivity, we used the zcNDVI proposed by Zambrano
+et al.~(2018) which will be derived from the NDVI retrieved from
+MOD13A1.
+
+To derive the drought indices, first we must calculate the sum of the
+variables with regard to the time scale (s). In this case, for
+generalization purposes, we will use \(V\), referring to variables
+\(P\), \(AED\), \(D\), \(NDVI\), and \(SM\) (\#ref\{tab-desEOD\}). We
+cumulated each \(V\) over the time series of \(n\) values (months), and
+for the time scales \(s\):
 
@@
 
-It corresponds to a moving window (convolution) that sums the variable
-for \(s\) starting for the last month \(n\) until the month, which could
-sum for \(s\) months (n-s+1). Once the variable is cumulated over time
-for the scale \(s\), we used a nonparametric approach following
-\citet{Hobbins2016} to derive the drought indices. Thus, the empirically
-derived probabilities are obtained through an inverse normal
+The \(A_{si}\) corresponds to a moving window (convolution) that sums
+the variable for time scales \(s\) from the last month, month by month,
+until the first month in which it could sum for \(s\) months. Once the
+variable is cumulated over time for the scale \(s\). Thus, the
+empirically derived probabilities are obtained through an inverse normal
 approximation \citep{Abramowitz1968}. Then, we used the empirical Tukey
 plotting position \citep{Wilks2011} over \(A_i\) to derive the
-\(P(A_i)\) probabilities across a period of interest:
+\(P(a_i)\) probabilities across a period of interest:
 
@@
 
-\hypertarget{sec-methods_lulc}{%
-\subsection{LULC change for 2001-2022 and its relation with water supply
-and demand, and soil moisture}\label{sec-methods_lulc}}
-
-\hypertarget{land-cover-macroclasess-and-validation}{%
-\subsubsection{Land cover macroclasess and
-validation}\label{land-cover-macroclasess-and-validation}}
-
-To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
-6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
-The IGBP defines 17 classes; from these, we regrouped into ten
-macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9
-to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
-13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
-Thus, we have a land cover raster time series with the ten classes for
-2001 and 2023.
-
-To validate the land cover obtained, we compare the macroclasses with
-the ones of a more detailed land cover map made by \citet{Zhao2016} for
-Chile with samples acquired in the years 2013--2014 (LCChile). The later
-has a spatial resolution of 30 m and three levels of defined classes;
-from those, we used level 1, which fits with the macroclasses land
-cover. We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from land
-cover macrolcasses to validate with LCChile.
-
-We follow the next procedure:
-
-\begin{enumerate}
-\def\labelenumi{\roman{enumi})}
-\tightlist
-\item
-  resampled LCChile to the spatial resolution (500m) of the land cover
-  macroclasses using the nearest neighbor method,
-\item
-  took a random sample of 1000 points within continental Chile and
-  extracted the classes that fell within each point for LCChile,
-  IGBP2013, and IGBP2014; we considered the point extracted from LCChile
-  as the truth and the values from the other two years as predictions.
-\item
-  calculate a confusion matrix with the classes extracted from the 1000
-  points for LCChile, IGBP2013, and IGBP2014. Calculate the performance
-  metrics of accuracy and F1.
-\end{enumerate}
-
-\[Accuracy = \frac{TP+TN}{TP+TN+FP+FN}=\frac{correct\,\, classifications}{all\,\, classifications}\]
-\[F1=\frac{2\cdot TP}{2\cdot TP + FP +FN}\] where \(TP\) and \(FN\)
-refer to true positive and false negative, correctly classified classes;
-\(TN\) and \(FP\) to true negative and false positive, wrongly
-classified classes.
-
-\hypertarget{land-cover-persistence-mask-2001-2022}{%
-\subsubsection{Land cover persistence mask
-2001-2022}\label{land-cover-persistence-mask-2001-2022}}
-
-Climate, vegetation development, seasonality, and changes in vegetation
-type all have an impact on the time series of NDVI. In this study, we
-want to examine the variation in vegetation productivity across various
-land cover types and how water demand, water supply, and soil moisture
-affect it. In order to avoid changes due to a change in the land cover
-type that will wrongly impact NDVI, we will develop a persistence mask
-for land cover for 2001--2023. Thereby, we reduce an important source of
-variation on a regional scale.
-
-Therefore, we generated a raster mask for IGBP MODIS per pixel using
-macroclasses that remain constant for at least 80\% of the years
-(2001--2022). This enables us to identify regions where the land cover
-of the macroclasses has remained constant.
-
-\hypertarget{land-cover-trend-and-drought-indices}{%
-\subsubsection{Land cover trend and drought
-indices}\label{land-cover-trend-and-drought-indices}}
-
-We calculated the surface occupied per land cover class into the five
-macrozones (``Norte Grande'' to ``Austral'') per year for 2001--2023.
-After that, we calculated the trend's change in surface; we used the
-Sen' slope \citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}.
-This way, we obtain a matrix of trends of 5 x 5 (macrozones x land
-cover). The aim is to later explore if the trend in land cover classes
-is associated with a trend in the drought indices. For this, we will use
-the techniques of regresion and regularization of Lasso
-\citep{Tibshirani2010} and Ridge \citep{Hoerl1970}. Also, we will test
-random forests for this purpose \citep{Ho1995}. We will choose the trend
-of land cover surface per macroclass and macrozone as the response
-variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI
-for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor
-variables. With this analysis, we expect to gather insights regarding
-whether there is a pattern of climatic influence along Chile or if what
-is happening in Central Chile has to do with more localized climatic
-conditions.
-
-\hypertarget{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}{%
-\subsection{Trend of drought indices for water demand and supply, soil
-moisture, and vegetation
-productivity}\label{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity}}
-
-\hypertarget{mann-kendall-and-sens-slope}{%
-\subsubsection{Mann-Kendall and Sen's
-slope}\label{mann-kendall-and-sens-slope}}
+The drought indices were calculated for time scales of 1, 3, 6, 12, 24,
+and 36 months at a monthly frequency for 1981--2023 in order to be used
+for short- to long-term evaluation of drought. In the case of the proxy
+of vegetation productivity (zcNDVI) it was calculated for a time scale
+of six months at monthly frequency for 2000--2023. For zcNDVI, we test
+time scales of 1, 3, 6, and 12 months; we choose to use six months
+because that shows a more robust representation of vegetation
+productivity due to the seasonality of vegetation in Chile.
+
+\hypertarget{trend-of-drought-indices}{%
+\subsubsection{Trend of drought
+indices}\label{trend-of-drought-indices}}
 
@@
 methodology are that the Sen's slope is not affected by outliers as
-regular regression does, and it is a non-paramteric method that is not
-affected by the distribution of the data. We applied both to the six
-time scales from 1981 to 2023 (monthly frequency) and the indices SPI,
-EDDI, SPEI, and SSI. In the case of zcNDVI (six months) was for 2000 to
-2023. Thus, we have 31 trends. Also, we extracted the trend aggregated
-by macrozone and land cover class, obtaining a table of 31x5x5 (drought
-indices trends x macrozone x land cover class). We will use this data in
-Section~\ref{sec-methods_lulc} to analyze if there is a strong
-relationship between the trends of drought indices and land cover
-surface within continental Chile.
-
-\hypertarget{trend-in-vegetation-productivity-without-land-cover-change}{%
-\subsubsection{Trend in vegetation productivity without land cover
-change}\label{trend-in-vegetation-productivity-without-land-cover-change}}
-
-\citet{Vicente-Serrano2021} made a global analysis of the drought's
-severity trend using SPI, SPEI, and the Standardized Evapotranspiration
-Deficit Index (SEDI; \citet{Vicente-Serrano2018}) to evaluate AED. They
-indicate that the increase in hydrological drought has been due to
-anthropogenic effects rather than climate change. This is because the
-global increase in AED did not explain the change in the spatial pattern
-of the hydrological drought. Also, they state that \emph{``the increase
-in hydrological droughts has been primarily observed in regions with
-high water demand and land cover change''}. We will contrast this
-hypothesis with what is occurring in Chile. To achieve this, we will use
-the land cover class type that remains more than 80\% of types for
-2001--2022 to evaluate the trend on zcNDVI and use this as a mask where
-there are low changes.
-
-\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-land-cover-types}{%
-\subsection{Impact for water supply and demand, and soil moisture in
-vegetation productivity within land cover
-types}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-within-land-cover-types}}
-
-We analyze the drought indices of water demand and supply and soil
-moisture against vegetation to address: i) if short- or long-term time
-scales are most important in impacting vegetation through Chile; and ii)
-the strength of the correlation for the variable and the time scale.
-Then, we will summarize for each land cover class and macrozone. Thus,
-we will be able to advance in understanding how climate is affecting
-vegetation, considering the impact on the five macroclasses of
-vegetation: forest, cropland, grassland, savanna, and shrubland.
-
-An analysis is conducted on the linear correlation between the indices
+regular regression does, and it is a non-parametric method that is not
+influenced by the distribution of the data. We applied Mann-Kendall to
+see if the trend was significant and Sen's slope to estimate the
+magnitude of the trend. We did this to the six time scales from 1981 to
+2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI. Thus,
+we have trends per index and time scale (24 in total). Then, we
+extracted the trend aggregated by macrozone and per land cover persitent
+macroclasses.
+
+\hypertarget{drought-impacts-on-vegetation-productivity-within-land-cover}{%
+\subsection{Drought impacts on vegetation productivity within land
+cover}\label{drought-impacts-on-vegetation-productivity-within-land-cover}}
+
+We analyzed the trend of vegetation productivity over the unchanged land
+cover macroclasses. This way, we tried to reduce the noise in the
+vegetation due to a change in land cover from year to year. To achieve
+this, we will use the persistent mask of land cover macroclasses, which
+are the types that remain more than 80\% of the time for 2001--2022. We
+used this to evaluate the trend in zcNDVI per land cover class and
+macrozone.
+
+We examine the drought indices of water demand, water supply, soil
+moisture, and their connection with vegetation productivity to
+investigate two main questions: i) whether short-term or long-term time
+scales have a greater impact on vegetation across Chile and its specific
+regions; and ii) the spatial variation in the strength of the
+correlation between the variables and time scales.~ Then, we will
+summarize for each land cover class and macrozone. Thus, we will be able
+to advance in understanding how climate is affecting vegetation,
+considering the impact on the five macroclasses of vegetation: forest,
+cropland, grassland, savanna, and shrubland.
+
+We conducted an analysis on the linear correlation between the indices
 SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36
 months, and zcNDVI. The objective is to determine the impact of soil
-moisture and water demand and supply on vegetation productivity. We
-implemented a methodology similar to that of \citet{Meroni2016} when
-comparing the SPI for meteorological drought to the cumulative FAPAR
-(Fraction of Absorbed Photosynthetically Active Radiation), which served
-as an indicator for vegetation productivity. A pixel-to-pixel linear
-correlation analysis was performed for each index. To begin, the Pearson
-coefficient of correlation is computed for each of the six time scales.
-A significant time scale is identified as the one that attains the
-highest correlation (p \textless{} 0.05). Subsequently, the Pearson
-correlation coefficient corresponding to the time scales at which the
-value peaked was extracted. As a result, for each index, we generated
-two raster maps: one containing the time scales and the other the
-correlation value.
+moisture and water demand and supply on vegetation productivity. We used
+a method similar to that used by \citet{Meroni2017} which compared the
+SPI with the cumulative FAPAR (Fraction of Absorbed Photosynthetically
+Active Radiation). A pixel-to-pixel linear correlation analysis was
+performed for each index within the persistent mask of land cover
+macroclasses. To begin, the Pearson coefficient of correlation is
+computed for each of the six time scales. A significant time scale is
+identified as the one that attains the highest correlation (p
+\textless{} 0.05). Subsequently, the Pearson correlation coefficient
+corresponding to the time scales at which the value peaked was
+extracted. As a result, for each index, we generated two raster maps: 1)
+containing the raster with values of the time scales that reached the
+maximum correlation, and 2) having the value of the correlation
+obtained.
+
+\hypertarget{software-and-packages-used}{%
+\subsection{Software and packages
+used}\label{software-and-packages-used}}
+
+For the downloading, processing, and analysis of the spatio-temporal
+data, we used the open source software for statistical computing and
+graphics, \texttt{R} \citep{R2023}. For downloading ERA5L, we used the
+\texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing raster
+data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
+\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
+used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
+used \texttt{\{SPEI\}} \citep{Bergueria2023}. For mapping, we use
+\{tmap\} \citep{Tennekes2018}. For data analysis, the suite
+\{tidyverse\} \citep{Wickham2019} was used.
 
@@
 
-\hypertarget{data-1}{%
-\subsection{Data}\label{data-1}}
-
-\hypertarget{validation-of-era5l-variables-1}{%
-\subsubsection{Validation of ERA5L
-variables}\label{validation-of-era5l-variables-1}}
-
-The average metrics of performance of ERA5L over the 266 weather
-stations were in the case of monthly temperature: \(ubRMSE=1.06\,°C\),
-\(MAE=1.131\,°C\), and \(CC=0.963\), showing a good agreement, low
-error, and low overestimation. For cumulative monthly precipitation,
-\(MAE=28.1\,mm\), \(bias=1.93\), and \(CC=0.845\), showing a high
-correlation and a 93\% bias and being overestimated by ERA5L. In the
-case of the 97 soil moisture stations, we averaged for the three depths
-(30, 60, and 90m) and then compared it with volumetric water content at
-1m derived from ERA5L. For this case, we made a daily comparison, having
-a \(CC=0.71\), \(RMSE=0.174\,m^3m^{-3}\), \(MAE=0.167\,m^3m^{-3}\), and
-\(bias=1.74\). The ERA5 soil moisture overestimate is 74\%, but it has a
-kind of good correlation. For more detailed information, consult the
-supplementary material (SSX).
-
-\hypertarget{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}{%
-\subsection{LULC change for 2001-2022 and its relation with water supply
-and demand, and soil
-moisture}\label{lulc-change-for-2001-2022-and-its-relation-with-water-supply-and-demand-and-soil-moisture}}
-
-\hypertarget{land-cover-macroclasess-and-validation-1}{%
-\subsubsection{Land cover macroclasess and
-validation}\label{land-cover-macroclasess-and-validation-1}}
-
-For vegetation, we obtained and use hereafter five macroclasses of land
-cover from IGBP MODIS: forest, shrubland, savanna, grassland, and
-croplands. Figure~\ref{fig-studyArea} c shows the spatial distribution
-of the macroclasses through Chile for the year 2022. The validation of
-IGBP2013 and IGBP2014 with LCChile reached near the same metrics of
-performance, having an accuracy of \textasciitilde0.82 and a F1 score of
-\textasciitilde0.66 (see SS1).
-
-\hypertarget{sec-persistence_mask}{%
-\subsubsection{Land cover persistence mask
-2001-2022}\label{sec-persistence_mask}}
-
-Figure~\ref{fig-studyArea} d, shows the macroclasses of land cover
-persistance (80\%) during 2021-2022, respectively
-(\ref{tab-landcoverSurf}). Within continental Chile, forest is the
-vegetation type with the highest surface area at 135,00 \(km^2\),
-followed by grassland (73,176 \(km^2\)), savanna (54,410 \(km^2\)),
-shrubland (24,959 \(km^2\)), and cropland (3,100 \(km^2\)) (Table
-\ref{tab-landcoverSurf}). The macrozones with major LULCC for 2001--2022
-were ``Centro,'' ``Sur,'' and ``Austral'' with 36\%, 31\%, and 34\%,
-respectively (Figure~\ref{fig-studyArea} and Table
-\ref{tab-landcoverTrend}); of its surface that changes the type of land
-cover. Figure~\ref{fig-LCprop} shows the summary of the proportion of
-surface per land cover class and macrozone, derived from the persistance
-mask over continental Chile.
+\hypertarget{land-cover-change-and-trend-1}{%
+\section{Land cover change and
+trend}\label{land-cover-change-and-trend-1}}
 
@@
 
-\hypertarget{land-cover-trend-and-drought-indices-1}{%
-\subsubsection{Land cover trend and drought
-indices}\label{land-cover-trend-and-drought-indices-1}}
+For vegetation, we obtained and use hereafter five macroclasses of land
+cover from IGBP MODIS: forest, shrubland, savanna, grassland, and
+croplands. Figure~\ref{fig-studyArea}c shows the spatial distribution of
+the macroclasses through Chile for the year 2022.
+Figure~\ref{fig-studyArea}d shows the macroclasses of land cover
+persistance (80\%) during 2021--2022, respectively
+(\#ref\{tab-landcoverSurf\}). Within continental Chile, barren land is
+the land cover class with the highest surface area (277,870 km2). The
+largest type of vegetation, with 137,085 km2, is forest. Grassland
+(74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland
+(3,146 km2) are the other types (Table 2). The macrozones with major
+changes for 2001--2022 were ``Centro,'' ``Sur,'' and ``Austral,'' with
+36\%, 31\%, and 34\% of their surface changing the type of land cover,
+respectively (Figure~\ref{fig-studyArea} and
+\#ref\{tab-landcoverTrend\}). Figure~\ref{fig-LCprop} shows the summary
+of the proportion of surface per land cover class and macrozone, derived
+from the persistance mask over continental Chile.
 
@@
 
-The ``Norte Chico'' shows an increase in barrend land of 111
-\(km^2 year^{-1}\) and a reduction in the class savanna of 70
-\(km^2 year^{-1}\). In the ``Centro'' and ``Sur,'' there are changes in
-the Chilean matorral, with an important reduction in savanna (136 to 318
-\(km^2 yr^{-1}\)), and an increase in shrubland and grassland. Showing a
-change for more dense vegetation types. It appears to be a shift in the
-area of cropland from the ``Centro'' to the ``Sur.'' Also, there is a
-high increase in forest (397 \(km^2 yr^{-1}\)) in the ``Sur,'' replacing
-the savanna lost.
-
-Further, we want to address whether the trend in land cover change for
-2001--2023 is associated with trends in drought indices of water demand
-and supply and/or soil moisture for macrozone and land cover
-macroclasses. From the three methods tested, Ridge, Lasso, and Random
-Forest, neither gives significant results regarding whether the trend in
-a drought index for any time scale explains the trend in land cover
-change. Nevertheless, in ``Norte Chico'' and ``Centro,'' there is a
-decrease in croplands and savanna and an increase in barren land, which
-is associated with the variation in drought indices. Mainly for a
-decrease in water supply (SPI and SSI) and an increase in water demand
-(EDDI). However, due to the high variability from north to south in
-Chile, the climatic conditions (arid, semi-arid, and humid), and the
-land cover type, we believe that only in those zones could the LULCC be
-driven to some degree by drought.
-
-\hypertarget{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity-1}{%
-\subsection{Trend of drought indices for water demand and supply, soil
-moisture, and vegetation
-productivity}\label{trend-of-drought-indices-for-water-demand-and-supply-soil-moisture-and-vegetation-productivity-1}}
+The ``Norte Chico'' shows an increase in barren land of 111
+\(km^2 yr^{-1}\) and a reduction in the class savanna of 70
+\(km^2 yr^{-1}\). In the ``Centro'' and ``Sur,'' there are changes with
+an important reduction in savanna (136 to 318 \(km^2 yr^{-1}\) ), and an
+increase in shrubland and grassland. Showing a change for more dense
+vegetation types. It appears to be a shift in the area cultivated
+(croplands) from the ``Centro'' to the ``Sur.'' Also, there is a high
+increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' replacing the
+savanna lost (\#ref\{tab-landcoverTrend\}).
+
+\hypertarget{trend-on-drought-and-its-relation-to-land-cover-change}{%
+\subsection{Trend on drought and its relation to land cover
+change}\label{trend-on-drought-and-its-relation-to-land-cover-change}}
+
+\hypertarget{trend-of-drought-indices-1}{%
+\subsubsection{Trend of drought
+indices}\label{trend-of-drought-indices-1}}
+
+Figure~\ref{fig-trendDI} shows the spatial variation of the trend for
+the drought indices from short- to long-term scales. SPI and SPEI have a
+decreasing trend from ``Norte Chico'' to ``Sur.'' However, there is an
+increasing trend in ``Austral.'' The degree of the trend is stronger at
+higher time scales. The SSI indicates that in ``Norte Grande,'' there
+are surfaces that have increased in the soutwest part and in the
+northeast have decreased, and is shown for all time scales. Similar to
+SPI and SPEI, SSI decreases at higher time scales. EDDI showed a
+positive trend for the whole of continental Chile, with a higher trend
+toward the north and a descending gradient toward the south. The degree
+of trend increases at higher time scales.
+
+\blandscape
+
+\begin{figure}
+
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendDI-1}SPI (Standardized Precipitation Index)}
+\end{minipage}%
+%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendDI-2}SPEI (Standardized Precipitation
+Evapotranspiration Index)}
+\end{minipage}%
+\newline
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendDI-3}EDDI (Evaporative Demand Drought
+Index)}
+\end{minipage}%
+%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_zcSM_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendDI-4}SSMI (Standardized Soil Moisture
+Index)}
+\end{minipage}%
+
+\caption{\label{fig-trendDI}Linear trend of the drought index (*) at
+time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023}
+
+\end{figure}
+
+\elandscape
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/trend_macrozone_drought_indices.png}
+
+}
+
+\caption{\label{fig-trendDIMacro}Trend per decade for the drought
+indices SPI, EDDI, SPEI, and SSI aggregated by macrozone.}
+
+\end{figure}
+
+In Figure~\ref{fig-trendDIMacro}, the averaged aggregation per
+macrozone, drought index, and time scale are shown. The macrozones that
+have the lowest trend are ``Norte Chico'' and ``Centro,'' where the SPI,
+SPEI, and SSI show that it decreases at longer time scales. Potentially
+explained due to the prolonged reduction in precipitation that has
+affected the hydrological system in Chile. At 36 months, it reaches
+trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and
+SSI. For ``Sur,'' the behavior is similar, decreasing at longer scales
+and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI.
+``Norte Grande'' has the highest trend at 36 months for EDDI (0.042 per
+decade), and ``Centro'' has the lowest for SPI and SPEI. In ``Norte
+Grande'' and ``Norte Chico,'' which are in a semi-arid climate, it is
+evident that the EDDI has an effect on the difference between the SPI
+and SPEI index, which is not seen in the other macrozones. Contrary to
+the other macrozones, ``Austral'' showed an increase in all indices,
+being the highest for EDDI at 36 months (0.025) and the lowest for SSI,
+which shows only a minor increase in the trend.
+
+\hypertarget{relationship-between-drought-indices-and-land-cover-change}{%
+\subsubsection{Relationship between drought indices and land cover
+change}\label{relationship-between-drought-indices-and-land-cover-change}}
+
+We look at the relationship between drought index and land cover change
+in Figure 5 by comparing the trends in land cover change (in terms of
+the total surface area per land cover type and macrozone) and drought
+indexes. Figure 5 shows that the negative trends in cropland (-0.029)
+and savanna (-0.026) in ``Norte Chico'' are the highest and are
+associated with an increase trend in EDDI and a decreasing trend in SPI,
+SPEI, and SSI. On the contrary, the shrubland in ``Centro'' has an
+increase (0.04) linked to a decrease in SPI, SPEI, and SSI and to an
+increase in EDDI. In ``Austral,'' the positive trend in shrubland fits
+with the positive trend in all the drought indices. The rest of the land
+cover types in the macrozones show weak associations with water supply,
+water demand, or soil moisture drought indices.
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/points_landcover_drought_indices_trend_and_time_scale.png}
+
+}
+
+\caption{Relationship between the trend in land cover change and the
+trend in drought indices for the five macrozones. Vertical panels
+correspond to 1, 3, 6, 12, 24, and 36 months of the time scale by
+drought index. Horizontal panels show each drought index}
+
+\end{figure}
+
+\hypertarget{drought-impacts-on-vegetation-productivity-within-land-cover-1}{%
+\subsection{Drought impacts on vegetation productivity within land
+cover}\label{drought-impacts-on-vegetation-productivity-within-land-cover-1}}
+
+\hypertarget{vegetation-productivity}{%
+\subsubsection{Vegetation productivity}\label{vegetation-productivity}}
 
@@
 
-In ``Norte Grande,'' vegetation productivity, as per the z-index,
-exhibits a yearly increase of 0.02 in the five land cover macroclasses,
-with respect to grassland and shrubland categories. There is a negative
-trend in ``Norte Chico'' with -0.04 and ``Centro'' with -0.02 per
-decade. In the ``Norte Chico,'' savanna (-0.05) has the lowest trend,
-and the rest of the types are around -0.04. In ``Centro,'' shrubland
-reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per
-decade. This could be associated either with a reduction in vegetation
-surface, a decrease in biomass, or browning \citep{Miranda2023}.
-Vegetation reached its lowest values since the year 2019, reaching an
-extreme condition in early 2020 and 2022 in the ``Norte Chico'' and
-Centro'' (Mega Drought). The ``Sur'' and ``Austral'' show a positive
+In Figure~\ref{fig-zcNDVI_var} it is showed the spatial map of trends in
+zcNDVI ({[}fig-zcNDVI\_var{]}a) and the temporal variation of zcNDVI
+within the aggregated macrozones ({[}fig-zcNDVI\_var{]}b). In ``Norte
+Grande,'' vegetation productivity, as per the z-index, exhibits a yearly
+increase of 0.02 with respect to grassland and shrubland categories.
+There is a negative trend in ``Norte Chico'' with -0.04 and ``Centro''
+with -0.02 per decade. In the ``Norte Chico,'' savanna (-0.05) has the
+lowest trend, and the rest of the types are around -0.04. In ``Centro,''
+shrubland reaches -0.06, grassland -0.05, and croplands and savanna
+-0.01 per decade. This could be associated either with a reduction in
+vegetation surface, a decrease in biomass, or browning
+\citep{Miranda2023}. Vegetation reached its lowest values since the year
+2019, with an extreme condition in early 2020 and 2022 in the ``Norte
+Chico'' and ``Centro''. The ``Sur'' and ``Austral'' show a positive
 trend of around 0.016 per decade (Figure~\ref{fig-zcNDVI_var}). Despite
 the croplands suffering from drought just as badly as the native
-vegetation in ``Norte Chico,'' the Chilean matorral \citep{Fuentes2021}
-appears to be the region most affected by a negative trend in vegetation
-productivity.
+vegetation in ``Norte Chico,'' the savanna and shurbland appears to be
+the region most affected by a negative trend in vegetation productivity.
+
+\hypertarget{correlation-between-vegetation-productivity-and-drought-indices}{%
+\subsubsection{Correlation between vegetation productivity and drought
+indices}\label{correlation-between-vegetation-productivity-and-drought-indices}}
 
@@
 
-{\centering \includegraphics{../output/figs/trend_macrozone_drought_indices.png}
-
-}
-
-\caption{\label{fig-trendDIMacro}Trend per decade for the drought
-indices SPI, EDDI, SPEI, and SSI aggregated by macrozone.}
+{\centering \includegraphics{../output/figs/mapa_cor_selec_indices_zcNDVI6.png}
+
+}
+
+\caption{\label{fig-corTimeScale}Time scales per drought index that
+reach the maximum coefficient of determination}
 
@@
 
-\blandscape
-
-\begin{figure}
-
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-1}SPI (Standardized Precipitation Index)}
-\end{minipage}%
-%
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-2}SPEI (Standardized Precipitation
-Evapotranspiration Index)}
-\end{minipage}%
-\newline
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-3}EDDI (Evaporative Demand Drought
-Index)}
-\end{minipage}%
-%
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_zcSM_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-4}SSMI (Standardized Soil Moisture
-Index)}
-\end{minipage}%
-
-\caption{\label{fig-trendDI}Linear trend of the drought index (*) at
-time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023}
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/mapa_cor_r_indices_zcNDVI6.png}
+
+}
+
+\caption{\label{fig-corPerson}Pearson correlation value for the time
+scales and drought index that reach the maximum coefficient of
+determination}
 
@@
 
-\elandscape
+Figure~\ref{fig-corTimeScale} is a map that shows the highest
+coefficient of determination (r-squared, or rsq) found in the regression
+analysis between different drought indicators and plant productivity
+over time. The spatial variation of time scales reached per index is
+mostly for time scales above 12 months. In the case of SSI, the
+predominant scales are 6 and 12 months. For all indices, to the north,
+the time scales are higher and diminish toward the south until the south
+part of ``Austral'' increases. In Figure~\ref{fig-corPerson}, the map of
+Pearson correlation values is shown. The EDDI reached correlations above
+0.5 between ``Norte Chico'' and ``Sur.'' The correlation changes from
+negative to positive toward the Andes Mountains and to the sea, just as
+in the northern part of ``Austral.'' The SPI and SPEI have similar
+results, with the higher values in ``Norte Chico'' and ``Centro'' being
+higher than 0.6. Following a similar spatial pattern as EDDI. The SSI
+showed to be the index that has a major spatial extension with a higher
+correlation. It has a similar correlation to SPI and SPEI for ``Norte
+Chico'' and ``Sur,''~ but has improvements for ``Sur.''
 
@@
 
-Analyzing the water supply, the macrozones that have the lowest trend
-are ``Norte Chico'' and ``Centro,'' where the SPI, SPEI, and SSI show
-that it decreases at longer time scales due to the prolonged reduction
-in precipitation. At 36 months, it reaches trends between -0.03 and
--0.04 (z-score) per decade for SPI, SPEI, and SSI
-(Figure~\ref{fig-trendDI}). For ``Sur,'' the behavior is similar,
-decreasing at longer scales and having between -0.016 and -0.025 per
-decade for SPI, SPEI, and SSI. On the other hand, all macrozones show an
-increase in the trend in all the drought indices, with ``Norte Grande''
-having the highest at 36 months (0.042 per decade). Because of this, the
-SPEI (which uses AED) reached its lowest value in ``Norte Grande,'' with
--0.03 at 36 months. Despite the other macrozones, ``Austral'' showed an
-increase in all indices, being the highest for EDDI at 36 months (0.025)
-and the lowest for SSI, which shows only a minor increase in the trend
-(Figure~\ref{fig-trendDI} and Figure~\ref{fig-trendDIMacro}).
-
-\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}{%
-\subsection{Impact for water supply and demand, and soil moisture in
-vegetation
-productivity}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}}
-
-According to what is shown in Figure~\ref{fig-corTimeScale},
-Figure~\ref{fig-corPerson}, and Table \ref{tab-corlandcover}, forest
-seems to be the most resistant type to drought. Showing that only
-``Centro'' is slightly (rsq = 0.25) impacted by a 12-month soil moisture
-deficit (SSI-12). In the ``Norte Chico'' and to a lesser extent in the
-``Norte Grande,'' it is evident that a SSI-12 with a rsq = 0.45 and a
-decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34,
+In \#ref\{tab-corlandcover\}, we aggregate per macrozone and landcover
+the correlation analysis presented in Figure~\ref{fig-corTimeScale} and
+Figure~\ref{fig-corPerson}. According to what is shown, forests seem to
+be the most resistant to drought. Showing that only ``Centro'' is
+slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit
+(SSI-12). In the ``Norte Chico'' and to a lesser extent in the ``Norte
+Grande,'' it is evident that a SSI-12 with a rsq = 0.45 and a decrease
+in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34,
 respectively) have an impact on grasslands. However, this type was
@@
 of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45\% and
-66\% of ``Norte Chico.'' The type of land most impacted by climatic
-variation was shrubland, where soil moisture explained 59\% and
-precipitation, 37\%, in ``Norte Chico'' and ``Centro,'' with SSI-12
-being the most relevant variable, then SPI-36 in ``Norte Chico'' and
-SPI-24 in ``Sur.''
-
-\begin{figure}[!ht]
-
-{\centering \includegraphics{../output/figs/mapa_cor_selec_indices_zcNDVI6.png}
-
-}
-
-\caption{\label{fig-corTimeScale}Time scales per drought index that
-reach the maximum coefficient of determination}
-
-\end{figure}
-
-\begin{figure}[!ht]
-
-{\centering \includegraphics{../output/figs/mapa_cor_r_indices_zcNDVI6.png}
-
-}
-
-\caption{\label{fig-corPerson}Pearson correlation value for the time
-scales and drought index that reach the maximum coefficient of
-determination}
-
-\end{figure}
+66\% of the variability in ``Norte Chico.'' The type of land most
+impacted by climatic variation was shrubland, where soil moisture
+explained 59\% and precipitation, 37\%, in ``Norte Chico'' and
+``Centro,'' with SSI-12 being the most relevant variable, then SPI-36 in
+``Norte Chico'' and SPI-24 in ``Sur.''
 
@@
 
-\hypertarget{drought-trend-and-attribution-to-lulcc}{%
-\subsection{Drought trend and attribution to
-LULCC}\label{drought-trend-and-attribution-to-lulcc}}
-
-\citet{Vicente-Serrano2021}, in a study at the global scale of drought
-trends, indicates that there have not been significant trends in
+\hypertarget{drought-trend-and-attribution-to-land-cover}{%
+\subsection{Drought trend and attribution to land
+cover}\label{drought-trend-and-attribution-to-land-cover}}
+
+\citet{Vicente-Serrano2022}, in a study at the global scale of drought
+trends, indicate that there have not been significant trends in
 meteorological drought since 1950. Also, state that the increase in
@@
 countries, finding that \emph{``since 1980, the Earth warming has not
-changed the drought area or intensity''}.
-
-In our study, we considered the variation in vegetation productivity in
-Chile for areas without changes in land cover macroclasses (see
-Section~\ref{sec-persistence_mask}), to avoid misleading conclusions
-that could be related to the increase in water demand due to LULCC. Our
+changed the drought area or intensity.''} In our study, we considered
+the variation in vegetation productivity in Chile for areas without
+changes in land cover, to avoid misleading conclusions that could be
+related to the increase in water demand due to land cover change. Our
 results show a contrasting perspective. There has been a significant
@@
 2022, seemsly due to an intense hydrological drought due to the
-persistance of the Mega Drought \citep{Garreaud2017}. Despite using the
-persistance mask for vegetation's trend analysis, cropland, which is the
-most water-demand type, showed a decrease trend in ``Norte Chico'' and
-``Centro.'' Also, there was an increase in barren land for both types.
-These changes are associated with a decrease in water demand from
-vegetation. Nonetheless, we used the persistant land cover to ensure
-that the pixel has the same class; in the case of croplands, it could
-happen that some areas had changed crops for others with higher water
-consumption. But this effect should be minor compared to the results
-from ladcover macroclasses.
+persistance of the mega drought \citep{Garreaud2017}. However, a rise in
+irrigated land doesn't seem to have an impact on this hydrological
+drought. Despite using the persistance mask for vegetation's trend
+analysis, cropland, which is the most water-demand type, showed a
+decrease trend in ``Norte Chico'' and ``Centro.'' Also, there was an
+increase in barren land for both types. These changes are associated
+with a decrease in water demand from vegetation. Nonetheless, we used
+the persistant land cover to ensure that the pixel has the same class;
+in the case of croplands, it could happen that some areas had changed
+crops for others with higher water consumption and consequently increase
+water demand. But this effect should be minor compared to the results
+from land cover change at this scale of analysis.
 
@@
 a decrease in trends of water supply (SPI and SSI), which are higher at
-larger time scales and consequently impact the hydrological system. We
-claim that what occurred in central Chile defies findings made at the
-global level \citep{Vicente-Serrano2021, Kogan2020}, demonstrating that
-a constant decrease in water supply rather than an increase in water
-demand (i.e., irrigated crops) is the main cause of the hydrological
-drought. Finally, central Chile has shown a diminishment in vegetation
-productivity for all macroclasses, mainly attributed to variation in
-water supply, i.e., precipitation, which could be strengthened by an
-increase in water demand by, for example, an increase in the surface
-area of irrigated crops.
+larger time scales, which is evidence of the hydrological drought. We
+say that what happened in central Chile goes against what has been found
+on a global scale \citep{Vicente-Serrano2022, Kogan2020}. This shows
+that the main cause of the hydrological drought in Chile was a steady
+drop in water supply made worse by an increase in AED, but it seems that
+in zones most affected by drought, the main cause is not an increase in
+water demand by vegetation like irrigated crops. Finally, north-central
+Chile has experienced a decline in vegetation productivity across all
+macroclasses, which is primarily attributable to variations in water
+supply and soil moisture. An increase in water demand, such as an
+increase in the surface area of irrigated crops, could strengthen this
+trend.
 
@@
 
-We found that shrubland, savannas (Chilean matorral), and croplands are
-the most sensitive to climate conditions. Being most affected by the
-12-month soil moisture deficit. In a study in the Yangtze River Basin in
-China, Jiang2020 analyzed the impact of drought on vegetation using the
+We discovered that croplands, savannas, and shrubland are the most
+susceptible to climatic changes and are most affected by the 12-month
+soil moisture deficit. In a study in the Yangtze River Basin in China,
+\citet{Jiang2020} analyzed the impact of drought on vegetation using the
 SPEI and the Enahanced Vegetation Index (EVI). They found that cropland
-was more sensitive to drought than cropland, showing that cropland
+was more sensitive to drought than grassland, showing that cropland
 responds strongly to short- and medium-term drought (\textless{}
@@
 between drought indices and vegetation productivity decrease, as do the
-time scales that reach the maximum r-squared (\ref{tab-corlandcover}).
-What can be explained is that, south of ``Centro,'' predominate forest
-and grassland, the most resistant types. Also, drought episodes have
-been less frequent and intense. The drought episodes have had a lower
-impact on water availability for vegetation.
-
-Extreme drought conditions are an important driver of tree mortality, as
-shown by \citet{Senf2020} in Europe. However, we found that forest is
+time scales that reach the maximum r-squared. What can be explained is
+that, south of ``Centro,'' predominate forest and grassland, the most
+resistant types. Also, drought episodes have been less frequent and
+intense. The drought episodes have had a lower impact on water
+availability for vegetation.
+
+According to \citet{Senf2020}, severe drought conditions in Europe are a
+significant cause of tree mortality. However, we found that forest is
 the type of land cover macroclass less affected by variation in drought
 indices, being the most resistant land cover class to drought.
-Supporting this is \citet{Fathi-Taperasht2022}, who asserts that Indian
-forests are the most drought-resistant and recover rapidly. Similarly,
-the work of \citet{Wu2024}, who analyzed vegetation loss and recovery in
-response to meteorological drought in the humid subtropical Pearl River
-basin in China, indicates that forests showed higher drought resistance.
-Using Vegetation Optical Depth (VOD), kNDVI, and EVI, \citet{Xiao2023},
-tests the resistance of ecosystems and finds that ecosystems with more
-forests are better able to handle severe droughts than croplands. They
-attribute the difference to a deeper rooting depth of trees, a higher
-water storage capacity, and different water use strategies between
-forest and cropland \citep{Xiao2023}.
+Supporting this is \citet{Fathi-Taperasht2022}{]}, who assert that
+Indian forests are the most drought-resistant and recover rapidly.
+Similarly, the work of \citet{Wu2024}, who analyzed vegetation loss and
+recovery in response to meteorological drought in the humid subtropical
+Pearl River basin in China, indicates that forests showed higher drought
+resistance. Using Vegetation Optical Depth (VOD), kNDVI, and EVI,
+\citet{Xiao2023} test the resistance of ecosystems and find that
+ecosystems with more forests are better able to handle severe droughts
+than croplands. They attribute the difference to a deeper rooting depth
+of trees, a higher water storage capacity, and different water use
+strategies between forest and cropland \citep{Xiao2023}.
 
@@
 to measure vegetation condition, it took the average condition of 1
-square kilometer. Then, to study how a type of forest (e.g.,
-sclerophyllous forest) changes in response to drought on a local level
-using remote sensing, we should use operational products with higher
-spatial resolutions, such as those from Landsat or Sentinel.
+square kilometer. Then, to use remote sensing to look at how a certain
+type of forest (like sclerophyllous forest) changes in response to
+drought on a local level, we should use operational products with higher
+spatial resolutions, like those from Landsat or Sentinel. This will let
+us do a more thorough analysis.
 
@@
 The main external factors that affect biomass production by vegetation
-are ET and SM, and the rate of ET in turn depends on the availability of
-water storage in the root zone. Thus, soil moisture plays a key role in
-land carbon uptake and, consequently, in the production of biomass
-\citep{Humphrey2021}. Moreover, \citet{Zhang2022} indicates there is a
-bidirectional causality between soil moisture and vegetation
-productivity. Lastly, some studies have redefined agricultural drought
-as soil moisture drought from an hydrological perspective
-\citep{Loon2016, Samaniego2018}. Even though soil moisture is the
-external factor most determinant of vegetation biomass, there are
-multiple internal factors, such as species, physiological
-characteristics, and plant hydraulics, that would affect vegetation
-productivity. Because of that, we believe that agricultural drought,
-referring to the drought that impacts vegetation productivity, is the
-most proper term, as originally defined by \citet{Wilhite1985}.
+are actual evapotranspiration and soil moisture, and the rate of ET in
+turn depends on the availability of water storage in the root zone.
+Thus, soil moisture plays a key role in land carbon uptake and,
+consequently, in the production of biomass \citep{Humphrey2021}.
+Moreover, \citet{Zhang2022} indicate there is a bidirectional causality
+between soil moisture and vegetation productivity. Lastly, some studies
+have redefined agricultural drought as soil moisture drought from a
+hydrological perspective \citep{Loon2016, Samaniego2018}. Even though
+soil moisture is the external factor most determinant of vegetation
+biomass, there are multiple internal factors, such as species,
+physiological characteristics, and plant hydraulics, that would affect
+vegetation productivity. Because of that, we believe that agricultural
+drought, referring to the drought that impacts vegetation productivity,
+is the most proper term, as originally defined by \citet{Wilhite1985}.
 
@@
 than rainfed croplands, soil moisture has better skills than SPI and
-SPEI for estimating gross primary production (GPP), according to
-\citet{Chatterjee2022}'s evaluation of the SPI and SPEI and their
-correlation with GPP in the CONUS. Also, \citet{Zhou2021} indicates that
+SPEI for estimating gross primary production (GPP). This according to
+\citet{Chatterjee2022} evaluation of the SPI and SPEI and their
+correlation with GPP in the CONUS. Also, \citet{Zhou2021} indicate that
 the monthly scaled Standardized Water Deficit Index (SWDI) can
 accurately show the effects of agricultural drought in most of China.
-Complementary, \citet{Nicolai2017} analyzed the time-lag between SWDI
-and the Vegetation Condition Index (VCI), and they state that there was
-no or little time-lag with croplands but a significant time-lag in the
-case of forests.
+\citet{Nicolai2017} also looked at the time-lag between the SWDI and the
+Vegetation Condition Index (VCI). They found that there was little to no
+time-lag in croplands but a greater time-lag in forests.
 
@@
 
-\hypertarget{early-drought-forecasting-for-vegetation-productivity}{%
-\subsection{Early drought forecasting for vegetation
-productivity}\label{early-drought-forecasting-for-vegetation-productivity}}
-
-We analyzed the correlation between meteorological and soil moisture
-drought indices with zcNDVI. From our findings, we could further use the
-drought indices and time scales that have the highest r-squared to
-develop a combined model to forecast vegetation productivity in Chile.
-Despite the fact that SSI was the best-performing index, the rest of the
-indices should help to enhance predictability. \citet{Zambrano2018}
-proposed a prediction model for cropland surface in Chile at
-administrative units with a 1- to 6-month lead time using zcNDVI from
-MODIS and climate oscilation indices. The results given were r-squared,
-ranging from 0.95 at a 1-month lead time to 0.37 at a 6-month lead time.
-Thus, incorporating the results of this study with those made by
-\citet{Zambrano2018}, we could develop a combined forecasting model at
-the pixel level for the macroclasses in Chile based on drought indices
-of water demand and supply and soil moisture.
-
 \hypertarget{future-outlook-to-complete}{%
@@
 There is a trend toward decreasing water supply in most parts of Chile,
-less in the ``Austral,'' which is stronger in the ``Centro'' and ``Norte
-Chico.'' The whole country showed an increase in water demand.
-Vegetation productivity only showed a decrease in the ``Norte Chico''
-and ``Centro,'' being highest for shrubland and croplands. Forest is the
-land cover most resistant to drought, and shrubland and cropland are the
-most sensitive.
+particularly in the ``Centro'' and ``Norte Chico'' regions. The whole
+country showed an increase in AED. Vegetation productivity only showed a
+decrease in the ``Norte Chico'' and ``Centro,'' being highest for
+shrubland and croplands. Forest is the land cover most resistant to
+drought, as shown along Chile, and shrubland and cropland are the most
+sensitive.
 
@@
 The results of this study could help in the development of a robust
-forecasting system for land cover classes in Chile., helping to improve
+forecasting system for land cover classes in Chile, helping to improve
 preparedness for climate change impacts on vegetation.

### 2024-03-12 article/supplementary.qmd
update version1 article

---
title: "Supplementary Material"
format: pdf
editor: visual
---

# Validation of ERA5L variables

## Methods

We compared the ERA5L variables for monthly mean temperature, total precipitation, and volumetric soil water content against values retrieved by weather stations. For temperature and precipitation, we used the weather network from the Ministry of Agriculture of Chile (www.agromet.com) between 2015 and 2023. We used 277 stations located throughout Chile. For soil moisture, we select a private soil network that is owned by the agricultural enterprise Garces Fruit, which has 99 stations in Central Chile, located in cherry fruit crops. The sensors are installed at 30, 60, and 90m and are the model Teros 12 from MeterGroup. To avoid the effect of irrigation on soil moisture, which ERA5L hardly captures, we used daily data for the year 2022 and the months outside the growing season, May to September.

To account for the performance of the ERA5L climatic variables regarding the values measured by the weather stations. We selected the following metrics:

$$MAE = \frac{1}{n}\sum |{E-S}|$$
$$Bias = \frac{\sum E}{\sum S}$$
$$ubRMSE =\sqrt{\frac{\sum{ \left[ (E_i-\overline{E})-(S_i-\overline{S}) \right ] ^2}}{n}}$$

$$CC = \frac{\sum (S_i-\bar{S})(E_i-\bar{E})}{\sqrt{(Si-\bar{S})^2(E_i-\bar{E})^2}}$$

$MAE$: mean absolute error
$bias$: bias
$ubRMSE$: unbiassed root mean squared error
$CC$: coefficient of correlation
$S$: value of the variable measure by the weather station
$E$: value of the variable measure by ERA5L

## Results

The average metrics of performance of ERA5L over the 266 weather stations were in the case of monthly temperature: $ubRMSE=1.06\,°C$, $MAE=1.131\,°C$, and $CC=0.963$, showing a good agreement, low error, and low overestimation. For cumulative monthly precipitation, $MAE=28.1\,mm$, $bias=1.93$, and $CC=0.845$, showing a high correlation and a 93% bias and being overestimated by ERA5L. In the case of the 97 soil moisture stations, we averaged for the three depths (30, 60, and 90m) and then compared it with volumetric water content at 1m derived from ERA5L. For this case, we made a daily comparison, having a $CC=0.71$, $RMSE=0.174\,m^3m^{-3}$, $MAE=0.167\,m^3m^{-3}$, and $bias=1.74$. The ERA5 soil moisture overestimate is 74%, but it has a kind of good correlation.

# Land cover macroclasess and validation

## Methods

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land cover raster time series with the ten classes for 2001 and 2023.

To validate the land cover obtained, we compare the macroclasses with the ones of a more detailed land cover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile). The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses land cover. We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from land cover macrolcasses to validate with LCChile.

We follow the next procedure:

i) resampled LCChile to the spatial resolution (500m) of the land cover macroclasses using the nearest neighbor method,
ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values from the other two years as predictions.
iii) calculate a confusion matrix with the classes extracted from the 1000 points for LCChile, IGBP2013, and IGBP2014. Calculate the performance metrics of accuracy and F1.

$$Accuracy = \frac{TP+TN}{TP+TN+FP+FN}=\frac{correct\,\, classifications}{all\,\, classifications}$$
$$F1=\frac{2\cdot TP}{2\cdot TP + FP +FN}$$

where $TP$ and $FN$ refer to true positive and false negative, correctly classified classes; $TN$ and $FP$ to true negative and false positive, wrongly classified classes.

## Results

For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands. Figure 1 c shows the spatial distribution of the macroclasses through Chile for the year 2022. The validation of IGBP2013 and IGBP2014 with LCChile reached near the same metrics of performance, having an accuracy of ~0.82 and a F1 score of ~0.66.

### 2024-03-13 article/paper_drought_Chile2023.qmd
first draft

@@
 ---
-title: Drought, vegetation productivity, and land cover change in Chile
+title: Multi-temporal drought and its interaction with land cover change and vegetation productivity in continental Chile
 #subtitle: A Short Subtitle
@@
   - name: Francisco Fernández
-    email: a.vrieling@utwente.nl
+    email: francisco.fernandez@uss.cl
     affiliations:
@@
   - name: Francisco Meza
+  - name: Iongel Duran-Llacer
   - name: Alejandro Venegas-González
-  - name: Iongel Duran-Llacer
   - name: Nicolas Raab
@@
 abstract: |
-  Central Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development.
+ The north-central region of Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development.
 This persistent period of water scarcity has been defined as a megadrought.
-Our objective is to assess the impact of drought on LULCC (land use land cover change) over continental Chile using drought indices of water supply and demand, soil moisture, and their impact on vegetation productivity.
-The monthly ERA5-Land (ERA5L) variables for precipitation, temperature, and soil moisture were used.
-From 2001 to 2022, we used the land cover MODIS product MCD12Q1, and from 2000 to 2023, we used the NDVI (Normalized Difference Vegetation Index) product MOD13A3 collection 6.1.
-As drought indices, we compute the standardized anomaly of cumulative NDVI (zcNDVI), the Standardized Precipitation Evapotranspiration Index (SPEI), the Evaporative Demand Drought Index (EDDI), and the Standardized Soil Moisture Index (SSI).
-These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was for 6 months.
-We analyze the trend for LULCC, vegetation productivity, and drought indices.
-Also, we analyzed the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain insights into the impact of water supply and demand on vegetation productivity.
-Our results showed that LULCC was highest in "Centro," "Sur," and "Austral,” with 36%, 31%, and 34% of change in the surface type, respectively.
-The EDDI shows that water demand has increased for all zones, with a major increase in "Norte Grande." The drought indices of water supply and soil moisture evidence a decreasing trend, which decreases at longer time scales, from "Norte Grande" to "Sur." "Austral" is the only zone that shows an increase in supply.
-Vegetation productivity measures by zcNDVI present a negative trend in "Norte Chico" and "Centro." On the other hand, forests seem to be the most resistant to drought.
+The aim of our study is to evaluate the land cover change over continental Chile and to examine how this is connected to drought indices of water supply, atmospheric evaporative demand (AED), soil moisture, and their effects on vegetation productivity.
+The drought indices were derived using monthly ERA5-Land reanalysis data spanning from 1981 to 2023.
+The Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets were utilized to obtain information on annual land cover and monthly vegetation productivity.
+We analyzed short- (1, 3, 6 months) to long-term (12, 24, 36 months) time scales of drought.
+Our results showed that land cover change was highest in the south-central part of the country, reaching changes as high as 36% in the surface type.
+The water demand has increased for the whole country, with a major increase in the north.
+The AED and soil moisture evidence a decreasing trend, which decreases at longer time scales and from north to south.
+The extreme south part of the country shows an increase in supply.
+Vegetation productivity has a negative trend in the north-central region for all land cover types.
+On the other hand, forests seem to be the most resistant type to drought.
 The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
-The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are SSI-12, followed by SPEI-24 and SPEI-12 in "Norte Chico" and "Centro." The results indicate that “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits lasting longer than a year.
-Our results can help develop a robust vegetation productivity forecasting model for land cover classes in Chile.
+The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are the ones that consider soil moisture for twelve months and the combined effect of precipitation and AED for 24 and 12 months.
+The results indicate that the north-central region is the most sensitive to water supply deficits lasting longer than a year.
 keywords: 
@@
 
-## Trend of drought and land cover change
+## Trend of drought and interaction to land cover change
 
@@
 To derive the drought indices, first we must calculate the sum of the variables with regard to the time scale (s).
-In this case, for generalization purposes, we will use $V$, referring to variables $P$, $AED$, $D$, $NDVI$, and $SM$ (#ref{tab-desEOD}).
+In this case, for generalization purposes, we will use $V$, referring to variables $P$, $AED$, $D$, $NDVI$, and $SM$ (Table \ref{tab-desEOD}).
 We cumulated each $V$  over the time series of $n$ values (months), and for the time scales $s$:
@@
 Then, we extracted the trend aggregated by macrozone and per land cover persitent macroclasses.
+
+### Intercation drought indices and land cover change
+
+We wanted to explore the relationship between the trend in land cover classes and the trend in the drought indices.
+Regarding the trends of land cover obtained (see Section 3.4.5), we divide each trend (km2) per land cover class and macrozone by the total surface.
+That way, we relativize the trend regarding the total land cover surface type per zone.
+To the trends of the drought indices over Chile, we applied the mask of persistent land cover macroclass and aggregated the trend for land cover class and macrozone.
+We then explore visually the relationship between the trend of drought indices for the short- to long-term and the trend of land cover.
 
@@
 [@fig-studyArea]c shows the spatial distribution of the macroclasses through Chile for the year 2022.
-[@fig-studyArea]d shows the macroclasses of land cover persistance (80%) during 2021–2022, respectively (#ref{tab-landcoverSurf}).
+[@fig-studyArea]d shows the macroclasses of land cover persistance (80%) during 2021–2022, respectively (Table \ref{tab-landcoverSurf}).
 Within continental Chile, barren land is the land cover class with the highest surface area (277,870 km2).
 The largest type of vegetation, with 137,085 km2, is forest.
-Grassland (74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland (3,146 km2) are the other types (Table 2).
-The macrozones with major changes for 2001–2022 were "Centro," "Sur," and "Austral," with 36%, 31%, and 34% of their surface changing the type of land cover, respectively (@fig-studyArea and #ref{tab-landcoverTrend}).
+Grassland (74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland (3,146 km2) are the other types (Table \ref{tab-landcoverSurf}).
+The macrozones with major changes for 2001–2022 were "Centro," "Sur," and "Austral," with 36%, 31%, and 34% of their surface changing the type of land cover, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend}).
 @fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.
@@
 Showing a change for more dense vegetation types.
-It appears to be a shift in the area cultivated (croplands) from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," replacing the savanna lost (#ref{tab-landcoverTrend}).
+It appears to be a shift in the area cultivated (croplands) from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," replacing the savanna lost (Table \ref{tab-landcoverTrend}).
 
-## Trend on drought and its relation to land cover change
+## Trend of drought and interaction to land cover change
 
@@
 
-In #ref{tab-corlandcover}, we aggregate per macrozone and landcover the correlation analysis presented in @fig-corTimeScale and @fig-corPerson.
+In Table \ref{tab-corlandcover}, we aggregate per macrozone and landcover the correlation analysis presented in @fig-corTimeScale and @fig-corPerson.
 According to what is shown, forests seem to be the most resistant to drought.

### 2024-03-13 article/paper_drought_Chile2023.tex
first draft

@@
 \hypersetup{
-  pdftitle={Drought, vegetation productivity, and land cover change in Chile},
-  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Fernández; Francisco Meza; Alejandro Venegas-González; Iongel Duran-Llacer; Nicolas Raab; Dylan Craven},
+  pdftitle={Multi-temporal drought and its effects on land cover change and vegetation productivity in continental Chile},
+  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Fernández; Francisco Meza; Iongel Duran-Llacer; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
   pdfkeywords={drought, land cover change, satellite},
@@
 \begin{frontmatter}
-\title{Drought, vegetation productivity, and land cover change in Chile}
+\title{Multi-temporal drought and its effects on land cover change and
+vegetation productivity in continental Chile}
 \author[1]{Francisco Zambrano%
@@
 }
- \ead{a.vrieling@utwente.nl} 
+ \ead{francisco.fernandez@uss.cl} 
 \author[]{Francisco Meza%
@@
 
+\author[]{Iongel Duran-Llacer%
+%
+}
+
 \author[]{Alejandro Venegas-González%
-%
-}
-
-\author[]{Iongel Duran-Llacer%
 %
@@
 \begin{abstract}
-Central Chile has been the focus of research studies due to the
-persistent decrease in water supply, which is impacting the hydrological
-system and vegetation development. This persistent period of water
-scarcity has been defined as a megadrought. Our objective is to assess
-the impact of drought on LULCC (land use land cover change) over
-continental Chile using drought indices of water supply and demand, soil
-moisture, and their impact on vegetation productivity. The monthly
-ERA5-Land (ERA5L) variables for precipitation, temperature, and soil
-moisture were used. From 2001 to 2022, we used the land cover MODIS
-product MCD12Q1, and from 2000 to 2023, we used the NDVI (Normalized
-Difference Vegetation Index) product MOD13A3 collection 6.1. As drought
-indices, we compute the standardized anomaly of cumulative NDVI
-(zcNDVI), the Standardized Precipitation Evapotranspiration Index
-(SPEI), the Evaporative Demand Drought Index (EDDI), and the
-Standardized Soil Moisture Index (SSI). These indices were calculated
-for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI,
-which was for 6 months. We analyze the trend for LULCC, vegetation
-productivity, and drought indices. Also, we analyzed the temporal
-correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain insights
-into the impact of water supply and demand on vegetation productivity.
-Our results showed that LULCC was highest in ``Centro,'' ``Sur,'' and
-``Austral,'' with 36\%, 31\%, and 34\% of change in the surface type,
-respectively. The EDDI shows that water demand has increased for all
-zones, with a major increase in ``Norte Grande.'' The drought indices of
-water supply and soil moisture evidence a decreasing trend, which
-decreases at longer time scales, from ``Norte Grande'' to ``Sur.''
-``Austral'' is the only zone that shows an increase in supply.
-Vegetation productivity measures by zcNDVI present a negative trend in
-``Norte Chico'' and ``Centro.'' On the other hand, forests seem to be
-the most resistant to drought. The types that show to be most affected
-by variation in climate conditions are shrublands, savannas, and
-croplands. The drought indices that have the capability of explaining to
-a major degree the variance in vegetation productivity are SSI-12,
-followed by SPEI-24 and SPEI-12 in ``Norte Chico'' and ``Centro.'' The
-results indicate that ``Norte Chico'' and ``Zona Central'' are the most
-sensitive regions to water supply deficits lasting longer than a year.
-Our results can help develop a robust vegetation productivity
-forecasting model for land cover classes in Chile.
+The north-central region of Chile has been the focus of research studies
+due to the persistent decrease in water supply, which is impacting the
+hydrological system and vegetation development. This persistent period
+of water scarcity has been defined as a megadrought. The aim of our
+study is to evaluate the land cover change over continental Chile and to
+examine how this is connected to drought indices of water supply,
+atmospheric evaporative demand (AED), soil moisture, and their effects
+on vegetation productivity. The drought indices were derived using
+monthly ERA5-Land reanalysis data spanning from 1981 to 2023. The
+Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets were
+utilized to obtain information on annual land cover and monthly
+vegetation productivity. We analyzed short- (1, 3, 6 months) to
+long-term (12, 24, 36 months) time scales of drought. Our results showed
+that land cover change was highest in the south-central part of the
+country, reaching changes as high as 36\% in the surface type. The water
+demand has increased for the whole country, with a major increase in the
+north. The AED and soil moisture evidence a decreasing trend, which
+decreases at longer time scales and from north to south. The extreme
+south part of the country shows an increase in supply. Vegetation
+productivity has a negative trend in the north-central region for all
+land cover types. On the other hand, forests seem to be the most
+resistant type to drought. The types that show to be most affected by
+variation in climate conditions are shrublands, savannas, and croplands.
+The drought indices that have the capability of explaining to a major
+degree the variance in vegetation productivity are the ones that
+consider soil moisture for twelve months and the combined effect of
+precipitation and AED for 24 and 12 months. The results indicate that
+the north-central region is the most sensitive to water supply deficits
+lasting longer than a year.
 \end{abstract}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, breakable, interior hidden, sharp corners, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{trend-of-drought-and-land-cover-change}{%
-\subsection{Trend of drought and land cover
-change}\label{trend-of-drought-and-land-cover-change}}
+\hypertarget{trend-of-drought-and-interaction-to-land-cover-change}{%
+\subsection{Trend of drought and interaction to land cover
+change}\label{trend-of-drought-and-interaction-to-land-cover-change}}
 
@@
 generalization purposes, we will use \(V\), referring to variables
-\(P\), \(AED\), \(D\), \(NDVI\), and \(SM\) (\#ref\{tab-desEOD\}). We
+\(P\), \(AED\), \(D\), \(NDVI\), and \(SM\) (Table \ref{tab-desEOD}). We
 cumulated each \(V\) over the time series of \(n\) values (months), and
@@
 macroclasses.
+
+\hypertarget{intercation-drought-indices-and-land-cover-change}{%
+\subsubsection{Intercation drought indices and land cover
+change}\label{intercation-drought-indices-and-land-cover-change}}
+
+We wanted to explore the relationship between the trend in land cover
+classes and the trend in the drought indices. Regarding the trends of
+land cover obtained (see Section 3.4.5), we divide each trend (km2) per
+land cover class and macrozone by the total surface. That way, we
+relativize the trend regarding the total land cover surface type per
+zone. To the trends of the drought indices over Chile, we applied the
+mask of persistent land cover macroclass and aggregated the trend for
+land cover class and macrozone. We then explore visually the
+relationship between the trend of drought indices for the short- to
+long-term and the trend of land cover.
 
@@
 Figure~\ref{fig-studyArea}d shows the macroclasses of land cover
-persistance (80\%) during 2021--2022, respectively
-(\#ref\{tab-landcoverSurf\}). Within continental Chile, barren land is
-the land cover class with the highest surface area (277,870 km2). The
+persistance (80\%) during 2021--2022, respectively (Table
+\ref{tab-landcoverSurf}). Within continental Chile, barren land is the
+land cover class with the highest surface area (277,870 km2). The
 largest type of vegetation, with 137,085 km2, is forest. Grassland
 (74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland
-(3,146 km2) are the other types (Table 2). The macrozones with major
-changes for 2001--2022 were ``Centro,'' ``Sur,'' and ``Austral,'' with
-36\%, 31\%, and 34\% of their surface changing the type of land cover,
-respectively (Figure~\ref{fig-studyArea} and
-\#ref\{tab-landcoverTrend\}). Figure~\ref{fig-LCprop} shows the summary
-of the proportion of surface per land cover class and macrozone, derived
+(3,146 km2) are the other types (Table \ref{tab-landcoverSurf}). The
+macrozones with major changes for 2001--2022 were ``Centro,'' ``Sur,''
+and ``Austral,'' with 36\%, 31\%, and 34\% of their surface changing the
+type of land cover, respectively (Figure~\ref{fig-studyArea} and Table
+\ref{tab-landcoverTrend}). Figure~\ref{fig-LCprop} shows the summary of
+the proportion of surface per land cover class and macrozone, derived
 from the persistance mask over continental Chile.
@@
 increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' replacing the
-savanna lost (\#ref\{tab-landcoverTrend\}).
-
-\hypertarget{trend-on-drought-and-its-relation-to-land-cover-change}{%
-\subsection{Trend on drought and its relation to land cover
-change}\label{trend-on-drought-and-its-relation-to-land-cover-change}}
+savanna lost (Table \ref{tab-landcoverTrend}).
+
+\hypertarget{trend-of-drought-and-interaction-to-land-cover-change-1}{%
+\subsection{Trend of drought and interaction to land cover
+change}\label{trend-of-drought-and-interaction-to-land-cover-change-1}}
 
@@
 
-In \#ref\{tab-corlandcover\}, we aggregate per macrozone and landcover
-the correlation analysis presented in Figure~\ref{fig-corTimeScale} and
-Figure~\ref{fig-corPerson}. According to what is shown, forests seem to
-be the most resistant to drought. Showing that only ``Centro'' is
-slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit
-(SSI-12). In the ``Norte Chico'' and to a lesser extent in the ``Norte
-Grande,'' it is evident that a SSI-12 with a rsq = 0.45 and a decrease
-in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34,
-respectively) have an impact on grasslands. However, this type was
-unaffected by soil moisture, water supply, or demand in macrozones
-further south. The types that show to be most affected by variation in
-climate conditions are shrublands, savannas, and croplands. For savannas
-in ``Norte Chico,'' the SSI-12 and SPI-24 reached an rsq of 0.74 and
-0.58, respectively. This value decreases to the south, but the SSI-12 is
-still the variable explaining more of the variation in vegetation
-productivity (rsq = 0.45 in ``Centro'' and 0.2 in ``Sur''). In the case
-of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45\% and
-66\% of the variability in ``Norte Chico.'' The type of land most
-impacted by climatic variation was shrubland, where soil moisture
-explained 59\% and precipitation, 37\%, in ``Norte Chico'' and
+In Table \ref{tab-corlandcover}, we aggregate per macrozone and
+landcover the correlation analysis presented in
+Figure~\ref{fig-corTimeScale} and Figure~\ref{fig-corPerson}. According
+to what is shown, forests seem to be the most resistant to drought.
+Showing that only ``Centro'' is slightly (rsq = 0.25) impacted by a
+12-month soil moisture deficit (SSI-12). In the ``Norte Chico'' and to a
+lesser extent in the ``Norte Grande,'' it is evident that a SSI-12 with
+a rsq = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with rsq
+= 0.28 and 0.34, respectively) have an impact on grasslands. However,
+this type was unaffected by soil moisture, water supply, or demand in
+macrozones further south. The types that show to be most affected by
+variation in climate conditions are shrublands, savannas, and croplands.
+For savannas in ``Norte Chico,'' the SSI-12 and SPI-24 reached an rsq of
+0.74 and 0.58, respectively. This value decreases to the south, but the
+SSI-12 is still the variable explaining more of the variation in
+vegetation productivity (rsq = 0.45 in ``Centro'' and 0.2 in ``Sur'').
+In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain
+between 45\% and 66\% of the variability in ``Norte Chico.'' The type of
+land most impacted by climatic variation was shrubland, where soil
+moisture explained 59\% and precipitation, 37\%, in ``Norte Chico'' and
 ``Centro,'' with SSI-12 being the most relevant variable, then SPI-36 in

### 2024-03-19 article/paper_drought_Chile2023.qmd
update paper random forest landcover vs drought

@@
 ---
-title: Multi-temporal drought and its interaction with land cover change and vegetation productivity in continental Chile
+title: The effects of multidimensional drought on land cover change and vegetation productivity in continental Chile
 #subtitle: A Short Subtitle
@@
           city:  Enschede, The Netherlands
+  - name: Francisco Meza
+  - name: Iongel Duran-Llacer
   - name: Francisco Fernández
@@
           city:  
-  - name: Francisco Meza
-  - name: Iongel Duran-Llacer
   - name: Alejandro Venegas-González
@@
 We wanted to explore the relationship between the trend in land cover classes and the trend in the drought indices.
-Regarding the trends of land cover obtained (see Section 3.4.5), we divide each trend (km2) per land cover class and macrozone by the total surface.
-That way, we relativize the trend regarding the total land cover surface type per zone.
-To the trends of the drought indices over Chile, we applied the mask of persistent land cover macroclass and aggregated the trend for land cover class and macrozone.
-We then explore visually the relationship between the trend of drought indices for the short- to long-term and the trend of land cover.
+For this purpose, in order to have more representative results, we conducted the analysis over sub-basins within continental Chile.
+We use 469 basins, which have a surface area between 0.0746 and  24,000 ($km^2$), and a median area of 1,249 ($km^2$).
+For each basin, we calculate the relative trend per land cover type, considering the proportion of the type relative to the total surface of the basin.
+Then, we extracted per basin the average trend of the drought indices SPI, SPEI, EDDI, SSI, and all their time scales 1, 3, 6, 12, 24, and 36.
+Also, we extracted the average trend in the proxy of vegetation productivity (zcNDVI).
+We wanted to analyze which drought indices and time scales have a major impact on changes in land cover type.
+
+We have 25 predictors, which are drought indices and vegetation productivity.
+We analyzed the 25 predictors per type of landcover.
+For the analysis, we selected the method of random forest (@Ho1995).
+Because it allows to find no linear relationship, it reduces overfitting and can derive the feature importance, which helps for a better understanding of the relationships.
+The importance of the variable is calculated by permuting out-of-bag (OOB) data per tree and calculating the mean standard error in the OOB.
+Then the same is done after permuting each predictor variable.
+Random forest uses multiple decision trees and allows for classification and regression.
+
+We analyzed the 25 predictors per type of landcover, thus running six models.
+We used random forests for regression and trained 1000 forests.
+For more reliable results for the important variables, we resampled by creating ten folds, running a random forest per fold, and calculating the r-squared (rsq), root mean square error (RMSE), and variable importance ten times.
 
@@
 
-We look at the relationship between drought index and land cover change in Figure 5 by comparing the trends in land cover change (in terms of the total surface area per land cover type and macrozone) and drought indexes.
-Figure 5 shows that the negative trends in cropland (-0.029) and savanna (-0.026) in “Norte Chico” are the highest and are associated with an increase trend in EDDI and a decreasing trend in SPI, SPEI, and SSI.
-On the contrary, the shrubland in “Centro” has an increase (0.04) linked to a decrease in SPI, SPEI, and SSI and to an increase in EDDI.
-In "Austral,” the positive trend in shrubland fits with the positive trend in all the drought indices.
-The rest of the land cover types in the macrozones show weak associations with water supply, water demand, or soil moisture drought indices.
+According to Table \ref{tab-landcoverTrendRF}, barren land is the type of land cover that is most affected by the trend of drought indices that reach a rsq = 0.4.
+This is attributed to the fact that it is linked to short-term deficits in precipitation (SPI-6 and SPEI-6) and long-term AED (EDDI-12 and EDDI-24).
+The trend in drought indices accounts for about a quarter of the change in grassland, savanna, and shrubland types.
+These types are most affected by the increase in short-term water demand (EDDI-6).
+The changes in forests are strongly impacted by the increase in long-term water demand (EDDI-36).
+The trend in croplands cannot be attributed to variations in water demand, water supply, soil moisture, or vegetation productivity.
 
-```{r figTrendsLandDrought}
+```{=latex}
+\begin{table}[!ht]
+\caption{The five most important trends of drought indices in estimating the landcover trend per land cover type and the r-squared (rsq) reached by each random forest model.}
+\label{tab-landcoverTrendRF}
+\includegraphics[]{../output/figs/table_importance_trends_landcover_vs_drought.png}
+\end{table}
+```
+
+```{r fig-TrendsLandDrought}
 #| echo: false
@@
 ```
+
+We look at the relationship between drought index and land cover change in @fig-TrendsLandDrought by comparing the trends in land cover change (in terms of the total surface area per land cover type and macrozone) and drought indexes.
+@fig-TrendsLandDrought shows that the negative trends in cropland (-0.029) and savanna (-0.026) in “Norte Chico” are the highest and are associated with an increase trend in EDDI and a decreasing trend in SPI, SPEI, and SSI.
+On the contrary, the shrubland in “Centro” has an increase (0.04) linked to a decrease in SPI, SPEI, and SSI and to an increase in EDDI.
+In "Austral,” the positive trend in shrubland fits with the positive trend in all the drought indices.
+The rest of the land cover types in the macrozones show weak associations with water supply, water demand, or soil moisture drought indices.
 

### 2024-03-19 article/paper_drought_Chile2023.tex
update paper random forest landcover vs drought

@@
 \hypersetup{
-  pdftitle={Multi-temporal drought and its effects on land cover change and vegetation productivity in continental Chile},
-  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Fernández; Francisco Meza; Iongel Duran-Llacer; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
+  pdftitle={The effects of multidimensional drought on land cover change and vegetation productivity in continental Chile},
+  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
   pdfkeywords={drought, land cover change, satellite},
@@
 \begin{frontmatter}
-\title{Multi-temporal drought and its effects on land cover change and
+\title{The effects of multidimensional drought on land cover change and
 vegetation productivity in continental Chile}
@@
  \ead{a.vrieling@utwente.nl} 
+\author[]{Francisco Meza%
+%
+}
+
+\author[]{Iongel Duran-Llacer%
+%
+}
+
 \author[3]{Francisco Fernández%
@@
  \ead{francisco.fernandez@uss.cl} 
-\author[]{Francisco Meza%
-%
-}
-
-\author[]{Iongel Duran-Llacer%
-%
-}
-
 \author[]{Alejandro Venegas-González%
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, enhanced, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, enhanced]}{\end{tcolorbox}}\fi
 
@@
 We wanted to explore the relationship between the trend in land cover
-classes and the trend in the drought indices. Regarding the trends of
-land cover obtained (see Section 3.4.5), we divide each trend (km2) per
-land cover class and macrozone by the total surface. That way, we
-relativize the trend regarding the total land cover surface type per
-zone. To the trends of the drought indices over Chile, we applied the
-mask of persistent land cover macroclass and aggregated the trend for
-land cover class and macrozone. We then explore visually the
-relationship between the trend of drought indices for the short- to
-long-term and the trend of land cover.
+classes and the trend in the drought indices. For this purpose, in order
+to have more representative results, we conducted the analysis over
+sub-basins within continental Chile. We use 469 basins, which have a
+surface area between 0.0746 and 24,000 (\(km^2\)), and a median area of
+1,249 (\(km^2\)). For each basin, we calculate the relative trend per
+land cover type, considering the proportion of the type relative to the
+total surface of the basin. Then, we extracted per basin the average
+trend of the drought indices SPI, SPEI, EDDI, SSI, and all their time
+scales 1, 3, 6, 12, 24, and 36. Also, we extracted the average trend in
+the proxy of vegetation productivity (zcNDVI). We wanted to analyze
+which drought indices and time scales have a major impact on changes in
+land cover type.
+
+We have 25 predictors, which are drought indices and vegetation
+productivity. We analyzed the 25 predictors per type of landcover. For
+the analysis, we selected the method of random forest (\citet{Ho1995}).
+Because it allows to find no linear relationship, it reduces overfitting
+and can derive the feature importance, which helps for a better
+understanding of the relationships. The importance of the variable is
+calculated by permuting out-of-bag (OOB) data per tree and calculating
+the mean standard error in the OOB. Then the same is done after
+permuting each predictor variable. Random forest uses multiple decision
+trees and allows for classification and regression.
+
+We analyzed the 25 predictors per type of landcover, thus running six
+models. We used random forests for regression and trained 1000 forests.
+For more reliable results for the important variables, we resampled by
+creating ten folds, running a random forest per fold, and calculating
+the r-squared (rsq), root mean square error (RMSE), and variable
+importance ten times.
 
@@
 
+According to Table \ref{tab-landcoverTrendRF}, barren land is the type
+of land cover that is most affected by the trend of drought indices that
+reach a rsq = 0.4. This is attributed to the fact that it is linked to
+short-term deficits in precipitation (SPI-6 and SPEI-6) and long-term
+AED (EDDI-12 and EDDI-24). The trend in drought indices accounts for
+about a quarter of the change in grassland, savanna, and shrubland
+types. These types are most affected by the increase in short-term water
+demand (EDDI-6). The changes in forests are strongly impacted by the
+increase in long-term water demand (EDDI-36). The trend in croplands
+cannot be attributed to variations in water demand, water supply, soil
+moisture, or vegetation productivity.
+
+\begin{table}[!ht]
+\caption{The five most important trends of drought indices in estimating the landcover trend per land cover type and the r-squared (rsq) reached by each random forest model.}
+\label{tab-landcoverTrendRF}
+\includegraphics[]{../output/figs/table_importance_trends_landcover_vs_drought.png}
+\end{table}
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/points_landcover_drought_indices_trend_and_time_scale.png}
+
+}
+
+\caption{\label{fig-TrendsLandDrought}Relationship between the trend in
+land cover change and the trend in drought indices for the five
+macrozones. Vertical panels correspond to 1, 3, 6, 12, 24, and 36 months
+of the time scale by drought index. Horizontal panels show each drought
+index}
+
+\end{figure}
+
 We look at the relationship between drought index and land cover change
-in Figure 5 by comparing the trends in land cover change (in terms of
-the total surface area per land cover type and macrozone) and drought
-indexes. Figure 5 shows that the negative trends in cropland (-0.029)
-and savanna (-0.026) in ``Norte Chico'' are the highest and are
-associated with an increase trend in EDDI and a decreasing trend in SPI,
-SPEI, and SSI. On the contrary, the shrubland in ``Centro'' has an
-increase (0.04) linked to a decrease in SPI, SPEI, and SSI and to an
-increase in EDDI. In ``Austral,'' the positive trend in shrubland fits
-with the positive trend in all the drought indices. The rest of the land
-cover types in the macrozones show weak associations with water supply,
-water demand, or soil moisture drought indices.
-
-\begin{figure}[!ht]
-
-{\centering \includegraphics{../output/figs/points_landcover_drought_indices_trend_and_time_scale.png}
-
-}
-
-\caption{Relationship between the trend in land cover change and the
-trend in drought indices for the five macrozones. Vertical panels
-correspond to 1, 3, 6, 12, 24, and 36 months of the time scale by
-drought index. Horizontal panels show each drought index}
-
-\end{figure}
+in Figure~\ref{fig-TrendsLandDrought} by comparing the trends in land
+cover change (in terms of the total surface area per land cover type and
+macrozone) and drought indexes. Figure~\ref{fig-TrendsLandDrought} shows
+that the negative trends in cropland (-0.029) and savanna (-0.026) in
+``Norte Chico'' are the highest and are associated with an increase
+trend in EDDI and a decreasing trend in SPI, SPEI, and SSI. On the
+contrary, the shrubland in ``Centro'' has an increase (0.04) linked to a
+decrease in SPI, SPEI, and SSI and to an increase in EDDI. In
+``Austral,'' the positive trend in shrubland fits with the positive
+trend in all the drought indices. The rest of the land cover types in
+the macrozones show weak associations with water supply, water demand,
+or soil moisture drought indices.
 

### 2024-03-19 article/supplementary.qmd
update paper random forest landcover vs drought

@@
 title: "Supplementary Material"
-format: pdf
-editor: visual
+author: "Francisco Zambrano"
+format: 
+  pdf:
+    number-sections: true
+editor: source
 ---
@@
 
-$$MAE = \frac{1}{n}\sum |{E-S}|$$
-$$Bias = \frac{\sum E}{\sum S}$$
-$$ubRMSE =\sqrt{\frac{\sum{ \left[ (E_i-\overline{E})-(S_i-\overline{S}) \right ] ^2}}{n}}$$
+$$MAE = \frac{1}{n}\sum |{E-S}|$$ $$Bias = \frac{\sum E}{\sum S}$$ $$ubRMSE =\sqrt{\frac{\sum{ \left[ (E_i-\overline{E})-(S_i-\overline{S}) \right ] ^2}}{n}}$$
 
@@
 
-$MAE$: mean absolute error
-$bias$: bias
-$ubRMSE$: unbiassed root mean squared error
-$CC$: coefficient of correlation
-$S$: value of the variable measure by the weather station
-$E$: value of the variable measure by ERA5L
+$MAE$: mean absolute error $bias$: bias $ubRMSE$: unbiassed root mean squared error $CC$: coefficient of correlation $S$: value of the variable measure by the weather station $E$: value of the variable measure by ERA5L
 
@@
 
-$$Accuracy = \frac{TP+TN}{TP+TN+FP+FN}=\frac{correct\,\, classifications}{all\,\, classifications}$$
-$$F1=\frac{2\cdot TP}{2\cdot TP + FP +FN}$$
+$$Accuracy = \frac{TP+TN}{TP+TN+FP+FN}=\frac{correct\,\, classifications}{all\,\, classifications}$$ $$F1=\frac{2\cdot TP}{2\cdot TP + FP +FN}$$
 
@@
 Figure 1 c shows the spatial distribution of the macroclasses through Chile for the year 2022.
-The validation of IGBP2013 and IGBP2014 with LCChile reached near the same metrics of performance, having an accuracy of ~0.82 and a F1 score of ~0.66.
+The validation of IGBP2013 and IGBP2014 with LCChile reached near the same metrics of performance, having an accuracy of \~0.82 and a F1 score of \~0.66.

### 2024-03-19 scripts/21_landcover_per_macrozone.R

@@ -40,11 +40,11 @@ LC_data |>
   na.omit() |>  
   group_by(zone,LC_type)|> 
   summarize(ntot = n()) -> summ_data2
-
+write_rds(summ_data2,'data/processed_data/surface_landcover_macrozone.rds')
 #tabla con superficie
 #
 summ_data2 |> 
-  mutate(surface = 21.19*ntot*0.01) |> 
+  mutate(surface = 21.5*ntot*0.01) |> 
   select(-ntot) |> 
   mutate(surface = case_when(surface <=20~NA,
                            .default = surface)) |> 
@@ -53,6 +53,12 @@ summ_data2 |>
   ungroup() |> 
    select(1,7:6,4:2,5) |> 
   gt() |> 
+  grand_summary_rows(
+    columns = 2:7,
+    fmt = ~ fmt_number(., decimal = 0, use_seps = TRUE),
+    fns = list(
+      label = 'Total',fn ="sum"
+    )) |> 
   fmt_number(decimals = 0) |> 
   sub_missing(
     columns = everything(),

### 2024-03-19 scripts/23_analysis_trend_drought_indices_and_landcover_macroXlandcover.R

@@ -13,16 +13,74 @@ data_ana <- data_trend_di |>
   full_join(data_trend_lc) |> 
   drop_na()
 
-data_model <- data_ana[,c(1:6,9:12,15:18,21:24,29,31)]
+#cargar la superficie de landcover ṕor macro class para relativizar las superficies
+
+lc_surf <- read_rds('data/processed_data/surface_landcover_macrozone.rds') |> 
+  set_names(c('macro','class','surf_total'))
+
+#buscando una relación con visualización
+#
+data_ana |> 
+  left_join(lc_surf) |> 
+  mutate(landcover_r = landcover) |> 
+  select(-surf_total) |> 
+  pivot_longer(-c(1:2,landcover,landcover_r)) |> 
+  mutate(
+    index = str_extract(name,'[[:alpha:]]+'),
+    scale = str_extract(name,'[0-9]+'),
+    landcover = landcover_r,
+    scale_t =fct(scale,levels =as.character(c(1,3,6,12,24,36))),
+    macro = fct(macro,levels = c('Norte Grande','Norte Chico','Centro','Sur','Austral')),
+    class = fct(class,levels = c('Forest','Cropland','Grassland','Savanna','Shrubland','Barren land')),
+    landcover = case_when(
+      macro == 'Norte Grande' & class %in% c('Forest', 'Cropland','Savanna') ~ NA,
+      macro == 'Norte Chico' & class == 'Forest' ~ NA,
+      macro == 'Sur' & class %in% c('Shrubland') ~ NA,
+      macro == 'Austral' & class == 'Cropland' ~ NA,
+      .default = landcover
+    ),
+    index = case_when(
+      index == 'zcSM' ~ 'SSI',
+      .default = index)) |>  
+  filter(index != 'zcNDVI' & 
+           name %in% c('EDDI-6','EDDI-36',
+                        'SPI-6','SPI-36',
+                       'zcSM-6','zcSM-36')) |> 
+  mutate(name = factor(name,
+                     levels = c('SPI-6','EDDI-6','zcSM-6',
+                                'SPI-36','EDDI-36','zcSM-36'),
+                    labels = c('SPI-6','EDDI-6','SSI-6',
+                               'SPI-36','EDDI-36','SSI-36'))) |> 
+  ggplot(aes(value*10,landcover,color = macro,shape = class)) +
+  geom_point() + 
[…]

### 2024-03-19 scripts/30_validacion_landcover_IGBP.R

@@ -20,7 +20,7 @@ lc2<- c(lc_modis_r,lc_chile)
 
 chl <- ne_countries(country = 'chile',scale = 'small',returnclass = 'sf') |> 
   st_transform(32719)
-samples <- st_sample(chl,1000)
+samples <- st_sample(chl,10000)
 
 df <- terra::extract(lc2,vect(samples))
 names(df) <- c('ID','IGBP2013','IGBP2014','LC_CHILE')
@@ -40,6 +40,7 @@ data <- df |>
   ungroup() |> 
   mutate(LC_CHILE_2 = ifelse(LC_CHILE_2 == 0, 6,LC_CHILE_2),
          LC_CHILE_2 = ifelse(LC_CHILE_2 == 80, 8,LC_CHILE_2)) |> 
+  filter(LC_CHILE_2 != 5) |> 
   mutate(across(everything(),\(x) fct(sort(as.character(x)))))
 
 library(yardstick)

### 2024-03-19 scripts/32_extraer_tendencias_indices_cuencas_landcover.R

@@ -52,3 +52,4 @@ df_scales |>
   rename(index = name,trend = value) |> 
   unnest(trend) |> 
   write_rds('data/processed_data/df_trends_indices_cuencas.rds')
+

### 2024-03-19 scripts/33_analysis_trend_drought_indices_and_landcover_cuencas.R

@@ -3,64 +3,44 @@ library(tidymodels)
 
 data_trend_lc <- read_rds('data/processed_data/trends_landcover_cuencas_2001-2022.rds') |> 
   pivot_longer(-cuenca) |> 
-  rename(landcover = value,class = name)
+  rename(landcover = value,class = name) |> 
+  replace_na(list(landcover = 0)) |> 
+  pivot_wider(names_from = 'class',values_from = 'landcover')
 
-data_trend_di <- read_rds('data/processed_data/df_trends_indices_cuencas.rds') |> 
-  rename(class = clase)
+data_trend_di <- read_rds('data/processed_data/df_trends_indices_cuencas.rds') |>
+  mutate(index = case_when(
+    index == 'zcSM-1' ~ 'SSI-1',
+    index == 'zcSM-3' ~ 'SSI-3',
+    index == 'zcSM-6' ~ 'SSI-6',
+    index == 'zcSM-12' ~ 'SSI-12',
+    index == 'zcSM-24' ~ 'SSI-24',
+    index == 'zcSM-36' ~ 'SSI-36',
+    .default = index)
+  ) |> 
+  replace_na(list(trend = 0)) |> 
+  pivot_wider(names_from = 'index',values_from = 'trend') |> 
+  rename(cuenca = COD_SUBC )
 
 data_ana <- data_trend_di |> 
   full_join(data_trend_lc) |> 
   drop_na()
 
-data_model <- data_ana[,c(1:6,9:12,15:18,21:24,29,31)]
-data_model <- data_model |> 
-  #select(-starts_with('EDDI')) |> 
-  mutate(
-    #landcover = cut(landcover,5),
-    landcover = cut(landcover,quantile(landcover,c(0,1/3,2/3,1)),
-                    include.lowest = TRUE),
-         class = factor(class),
-         macro = factor(macro))
+#data_model <- data_ana[,c(1:6,9:12,15:18,21:24,29,31)]
+data_model <- data_ana 
+   # mutate(
+   #   `zcNDVI-6` = cut(`zcNDVI-6`,5)
+   # )
+#     # landcover = cut(landcover,quantile(landcover,c(0,1/3,2/3,1)),
+#     #                 include.lowest = TRUE),
+#          class = factor(class),
+#          macro = factor(macro))
 
-df_split <- initial_split(data_model,prop =.9)
-df_train <- training(df_split)
[…]

### 2024-03-19 article/paper_drought_Chile2023.qmd
update article

@@
 ---
-title: The effects of multidimensional drought on land cover change and vegetation productivity in continental Chile
+title: The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile
 #subtitle: A Short Subtitle
@@
 SPEI takes into account the combined effect of precipitation and AED as a water balance, but it does not allow us to know the contribution of each variable on its own.
-Some things worth investigating in terms of land cover change and vegetation productivity are: i) How do they respond to short- to long-term meteorological droughts?
-ii) How do they behave in humid and arid climatic zones regarding drought?
-And iii) What is the role of soil moisture?
+Some things worth investigating in terms of land cover change and vegetation productivity are: i) How do they respond to short- to long-term meteorological and soil moisture droughts?
+ii) How is the drought impacting land cover changes?
+And iii) How do they behave in humid and arid climatic zones regarding drought?
 Likewise, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.
 
-Chile's diverse climatic and ecosystem types (@Beck2023;@Luebert2022) make it an ideal natural laboratory for studying climate and ecosystems.
+For monitoring drought, the World Meteorological Organization recommends the SPI (Standardized Precipitation Index) [@WMO2012].
+The SPI is a multi-scalar drought index that only uses precipitation to assess short- to long-term droughts.
+The primary cause of drought is precipitation anomalies, and temperature usually makes it worse [@Luo2017].
+Nowadays, there is an increase in attention toward using AED separately to monitor droughts [@Vicente-Serrano2020].
+One reason is due to its attribution to increasing flash droughts in water-limited regions [@Noguera2022].
+@Vicente-Serrano2010 proposed the Standardized Precipitation Evapotranspiration Index (SPEI), which incorporated the temperature effect by subtracting AED from precipitation.
+SPEI allows for analysis of the combined effect of precipitation and AED.
+Since its formulation, it has been used worldwide for the study and monitoring of drought [@Gebrechorkos2023;@Liu2021].
+@Hobbins2016 and @McEvoy2016 developed the Evaporative Demand Drought Index (EDDI) to monitor droughts solely using the AED, and it has proven effective in monitoring flash droughts [@Li2024;@Ford2023].
+For soil moisture, several drought indices exist, such as the Soil Moisture Deficit Index (SDMI) [@Narasimhan2005] and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021].
+@Hao2013 and @AghaKouchak2014 proposed the Standardized Soil Moisture Index (SSI), which has a similar formulation as the SPI, SPEI, and EDDI.
+Thus, there are plenty of drought indices that allow for a comprehensive assessment of drought from short- to long-term scales and consider variables from the earth's water balance on their own (e.g., precipitation, AED, soil moisture).
+Using this information, we can advance our understanding of the impact of drought on ecosystems.
+
+Chile's diverse climatic and ecosystem types [@Beck2023;@Luebert2022] make it an ideal natural laboratory for studying climate and ecosystems.
 Additionally, the country has experienced severe drought conditions that have had significant effects on vegetation and water storage.
-Central Chile faced a persistent precipitation deficit between 2010 and 2022, defined as a megadrought [@Garreaud2017], which has impacted the Chilean ecosystem.
+North-central Chile has faced a persistent precipitation deficit since 2010, defined as a mega drought.
+[@Garreaud2017], which has impacted the Chilean ecosystem.
 This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December having values below one standard deviation.
@@
 The term "megadrought" in Chile is used to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018].
-Hence, it is imperative to assess temporal scales that take into account the cumnulative effect within some years.
-There is little knowledge about the relationship between drought and ecosystem in Chile; thus, it is important to understand in more detail how meteorological and soil moisture droughts influence ecosystem dynamics to inform adaptation options.
+Therefore, it is crucial to evaluate temporal scales that consider the cumulative impact over a period of several years.
+There is limited understanding of the correlation between drought and the environment in Chile.
+Therefore, it is crucial to have a deeper understanding of how meteorological and soil moisture droughts impact the dynamics of the environment in order to make informed decisions on how to adapt.
 
-A detailed spatiotemporal assessment of the interaction of drought for short- to long-term and land cover change requires information on vegetation as well as weather variables such as precipitation, temperature, and soil moisture.
+<!-- A detailed spatiotemporal assessment of the interaction of drought for short- to long-term and land cover change requires information on vegetation as well as weather variables such as precipitation, temperature, and soil moisture.
 Weather networks in Chile present some disadvantages, such as spatio-temporal gaps, a short history, and irregular quality, which make them difficult to represent the whole extent of the country spatially.
@@
 We expect to gain insight regarding the temporal evolution of water demand, water supply, and soil moisture, as well as the interaction with land cover change and vegetation productivity.
+-->
+
 Here, we analyze the multi-dimensional impacts of drought across ecosystems in continental Chile.
-More specifically, we aim to assess: i) temporal changes in land-use cover and the direction and magnitude of their relationships with drought indices for water demand and supply, soil moisture, and vegetation productivity; ii) short- to long-term temporal trends in multi-scalar drought indices; and iii) the relationship between vegetation productivity and drought indices for water demand and supply and soil moisture across Chilean ecosystems.
+More specifically, we aim to assess: i) short- to long-term temporal trends in multi-scalar drought indices; ii) temporal changes in land-use cover and the direction and magnitude of their relationships with trends in drought indices; and iii) the trend in vegetation productivity and its relationship with drought indices across Chilean ecosystems.
 
@@
 
-Continental Chile has a diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] (@fig-studyArea a), which determines its great ecosystem diversity (@Luebert2022) (@fig-studyArea c).
+Continental Chile has diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] (@fig-studyArea a), which determines its great ecosystem diversity [@Luebert2022] (@fig-studyArea c).
 The Andes Mountains are a main factor in climate latitudinal variation [@Garreaud2009].
-In order to characterize the climate and ecosystem of Chile, we utilize the Köppen-Geiger classification system developed by @Beck2023 and the land cover data derived from the MODIS product for the period of 2001–2022, based on the International Geosphere-Biosphere Programme (IGBP) classification scheme proposed by @Friedl2019.
 "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures.
 At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk).
-In these two northern regions, the land is mostly bare, with a minor surface of vegetation types such as shrubland and grassland.
+In these two northern regions, the land is mostly bare, with a small surface of vegetation types such as shrubland and grassland.
 In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warm to hot summers (Csa and Csb).
@@
 An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral." Those zones are high in forest and grassland.
-The southern part of the country has a tundra climate, and in "Austral", it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.
+The southern part of the country has a tundra climate, and in "Austral," it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.
 
@@
 
-To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS.
-This product has a yearly frequency from 2001 to 2022.
-The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
-Thus, we have a land cover raster time series with the ten classes for 2001 and 2023.
-Prior to using this class, we compared it with the types of land cover made by @Zhao2016, which is a more detailed land cover map of Chile with a 30 m spatial resolution for 2012–2013.
-It reached a global accuracy of ~0.82.
-The procedure of validation is demonstrated in Section S1 of the supplementary material.To derive a proxy for vegetation productivity, we used the MOD13A3 collection 6.1 product from MODIS [@Didan2015].
-It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency.
-The MOD13A3 and MCD12Q1 were retrieved from the online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, https://lpdaa.usgs.gov/tools/data-pool/.
+To analyze land cover change, we use the classification scheme by the IGBP (International Geosphere-Biosphere Programme) from the product MCD12Q1 collection 6.1 from MODIS.
+To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the product MOD13A3 collection 6.1 from MODIS [@Didan2015].
+MOD13A3 provides vegetation indices at 1km of spatial resolution and monthly frequency.
+The NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and MCD12Q1 from the online Data Pool, accessible at https://lpdaa.usgs.gov/tools/data-pool/.
 
@@
 \begin{table}[!ht]
-\caption{Description of the earth observation data used }
+\caption{Description of the satellite and reanalysis data used}
 \label{tab-desEOD}
@@
 \end{tabular}
-{\raggedright *Derived from ERA5L with Eq. \ref{eq-AED}. \par}
+{\raggedright *Calculated from maximum and minimum temperatures derived from ERA5L with Eq. \ref{eq-AED}. \par}
 \end{table}
@@
 
-For soil moisture, water supply, and water demand variables, we used ERA5L [@MunozSabater2021], a reanalysis dataset that provides the evolution of the meteorological and soil moisture variables since 1950.
+For soil moisture, water supply, and water demand variables, we used ERA5L (ECMWF Reanalysis version 5 over land) [@MunozSabater2021], a reanalysis dataset that provides the evolution of atmospheric and land variables since 1950.
 It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage.
 We selected the variables for total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).
+Table \ref{tab-desEOD} shows a summary of the data and its main characteristics.
 
-## Land cover change and trend
-
-To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS.
-@Zambrano2018 and @Fuentes2021 have previously used this product for studies of drought and land cover.
-The MCD12Q1 has a yearly frequency from 2001 to 2022.
-The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
-Thus, we have a land cover raster time series with the ten macroclasses for 2001 and 2023.
-We validate the land cover macroclasses regarding a highly detailed (30 m of spatial resolution) land cover map made for Chile by @Zhao2016 for 2013-2014.
-Our results showed a global accuracy of ~0.82 and a F1 score of ~0.66.
-Section S2 in the Supplementary Material shows the procedure for validation.
-
-Climate, vegetation development, seasonality, and changes in vegetation type all have an impact on the time series of NDVI.
-In this study, we want to examine the variation in vegetation productivity across various land cover types and how water demand, water supply, and soil moisture affect it.
-In order to avoid changes due to a change in the land cover type that will wrongly impact NDVI, we developed a persistence mask for land cover for 2001–2022.
-Thereby, we reduce an important source of variation on a regional scale.
-Therefore, we generated a raster mask for IGBP MODIS per pixel using macroclasses that remained unchanged for at least 80% of the years (2001–2022).
-This enabled us to identify regions where the land cover macroclasses are persistent.
-We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023.
-After that, we calculated the trend’s change in surface per type.
-We used the Sen’ slope [@Sen1968] based on Mann-Kendall [@Kendall1975].
-
-## Trend of drought and interaction to land cover change
+## Trend of short- to long-term drought
 
@@
 
-In order to compute the drought indices that uses water demand, it is necessary to first calculate the AED.
-To do this, we employed the Hargreaves technique [@Hargreaves1994;@Hargreaves1985], by applying the following equation:
+In order to compute the drought indices that use water demand, it is necessary to first calculate the AED.
+To do this, we employed the Hargreaves method [@Hargreaves1994;@Hargreaves1985] by applying the following equation:
 
@@
 
-where $Ra$ $(MJ\,m^2\, day^{-1})$ is extraterrestrial radiation; $T$, $T_{max}$, and $T_{min}$ are mean, maximum, and minimum temperature $(°C)$.
-We calculate the centroid coordinates per pixel and use the latitude to estimate $Ra$.
-
+where $Ra$ $(MJ\,m^2\, day^{-1})$ is extraterrestrial radiation; $T$, $T_{max}$, and $T_{min}$ are mean, maximum, and minimum temperature $(°C)$ at 2m.
+For calculating $Ra$ we used the coordinate of the latitud of the centroid of each pixel.
 We chose the method of Hargreaves to estimate AED because of its simplicity, which only requires temperatures and extrarrestrial radiation.
@@
 
-To derive the drought indices of water supply and demand, soil moisture, and vegetation we used the ERA5L dataset and the MODIS product, with a monthly frequency for 1981-2023 and 2000-2023, respectively.
+To derive the drought indices of water supply and demand, soil moisture, and vegetation (i.e., the proxy of productivity), we used the ERA5L dataset and the MODIS product, with a monthly frequency for 1981–2023 and 2000–2023, respectively.
 
@@
 Also, it has been used worldwide for the study of drought, including in Chile (@Garreaud2017; @Zambrano2017).
-The primary cause of drought is precipitation anomalies, and temperature usually makes it worse (Luo et al.
-2017).
+The primary cause of drought is precipitation anomalies, and temperature usually makes it worse [@Luo2017].
 Nowadays, there is an increase in attention toward using water demand separately to monitor droughts.
 (@Vicente-Serrano2020; @Noguera2022).
-Thus, to evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI), developed by Hobbins et al.
-(016) and McEvoy et al.
-(2016), which is based on the AED.
+Thus, to evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI), developed by @Hobbins2016 and @McEvoy2016, which is based on the AED.
 EDDI is currently used for monitoring drought in the United States (https://psl.noaa.gov/eddi/).
@@
 Soil moisture is the main driver of vegetation productivity, particularly in semi-arid regions (@Li2022).
-Hence, for soil water drought, we used the SSI (Standardized Soil Moisture Index) (Hao and AghaKouchak 2013; A.
-AghaKouchak 2014) which is a multi-scale index similar to SPI, SPEI, and EDDI.
+Hence, for soil water drought, we used the SSI (Standardized Soil Moisture Index) [@Hao2013;@AghaKouchak2014] which is a multi-scale index similar to SPI, SPEI, and EDDI.
 In our case, for the SSI, we used the average soil moisture from ERA5L at 1m depth.
-Finally, for the proxy of productivity, we used the zcNDVI proposed by Zambrano et al.
-(2018) which will be derived from the NDVI retrieved from MOD13A1.
+Finally, for the proxy of productivity, we used the zcNDVI proposed by @Zambrano2018 which will be derived from the NDVI retrieved from MOD13A1.
 
@@
 
-### Intercation drought indices and land cover change
+## Interaction of land cover and drought
+
+### Land cover change
+
+To analyze the land cover change, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS.
+@Zambrano2018 and @Fuentes2021 have previously used this product for studies of drought and land cover in Chile.
+The MCD12Q1 has a yearly frequency from 2001 to 2022.
+The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
+Thus, we have a land cover raster time series with the ten macroclasses for 2001 and 2023.
+We validate the land cover macroclasses regarding a highly detailed (30 m of spatial resolution) land cover map made for Chile by @Zhao2016 for 2013-2014.
+Our results showed a global accuracy of ~0.82 and a F1 score of ~0.66.
+Section S2 in the Supplementary Material shows the procedure for validation.
+
+Climate, vegetation development, seasonality, and changes in vegetation type all have an impact on the time series of NDVI.
+In this study, we want to examine the variation in vegetation productivity across various land cover types and how water demand, water supply, and soil moisture affect it.
+In order to avoid changes due to a change in the land cover type that will wrongly impact NDVI, we developed a persistence mask for land cover for 2001–2022.
+Thereby, we reduce an important source of variation on a regional scale.
+Therefore, we generated a raster mask for IGBP MODIS per pixel using macroclasses that remained unchanged for at least 80% of the years (2001–2022).
+This enabled us to identify regions where the land cover macroclasses are persistent.
+We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023.
+After that, we calculated the trend’s change in surface per type.
+We used the Sen’ slope [@Sen1968] based on Mann-Kendall [@Kendall1975].
+
+### Relationship between land cover and drought trends
 
@@
 
-## Drought impacts on vegetation productivity within land cover
+## Drought impacts on vegetation productivity
 
@@
 
-# Land cover change and trend
-
-```{=latex}
-\begin{table}[!ht]
-\caption{Surface of the land cover class that persist during 2001-2022}
-\label{tab-landcoverSurf}
-\includegraphics[width = .5\textwidth]{../output/figs/table_surface_landcover_macrozone.png}
-\end{table}
-```
-
-```{r fig-LCprop}
-#| echo: false
-#| fig-pos: "!ht"
-#| fig-cap: "Proportion of land cover class from the persistent land cover for 2001-2022 (>80%) per macrozone"
-#| out-width: "\\textwidth"
-knitr::include_graphics("../output/figs/LC_pers80_per_macrozone.png")
-```
-
-For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands.
-[@fig-studyArea]c shows the spatial distribution of the macroclasses through Chile for the year 2022.
-[@fig-studyArea]d shows the macroclasses of land cover persistance (80%) during 2021–2022, respectively (Table \ref{tab-landcoverSurf}).
-Within continental Chile, barren land is the land cover class with the highest surface area (277,870 km2).
-The largest type of vegetation, with 137,085 km2, is forest.
-Grassland (74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland (3,146 km2) are the other types (Table \ref{tab-landcoverSurf}).
-The macrozones with major changes for 2001–2022 were "Centro," "Sur," and "Austral," with 36%, 31%, and 34% of their surface changing the type of land cover, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend}).
-@fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.
-
-```{=latex}
-\begin{table}[!ht]
-\caption{The value of Sen's slope trend next to the time-series plot of surface per land cover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile.
-Values of zero indicate that there was not a significant trend.
-Red dots on the plots indicate the maximum and minimum values of surface.}
-\label{tab-landcoverTrend}
-\includegraphics[]{../output/figs/table_var_landcover_macro.png}
-\end{table}
-```
-
-The "Norte Chico" shows an increase in barren land of 111 $km^2 yr^{-1}$  and a reduction in the class savanna of 70 $km^2 yr^{-1}$.
-In the "Centro" and "Sur," there are changes with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$ ), and an increase in shrubland and grassland.
-Showing a change for more dense vegetation types.
-It appears to be a shift in the area cultivated (croplands) from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," replacing the savanna lost (Table \ref{tab-landcoverTrend}).
-
-## Trend of drought and interaction to land cover change
-
-### Trend of drought indices
+## Trend of short- to long-term drought
 
@@
 
+### Interaction of land cover and drought
+
+## Land cover change
+
+```{=latex}
+\begin{table}[!ht]
+\caption{Surface of the land cover class that persist during 2001-2022}
+\label{tab-landcoverSurf}
+\includegraphics[width = .5\textwidth]{../output/figs/table_surface_landcover_macrozone.png}
+\end{table}
+```
+
+```{r fig-LCprop}
+#| echo: false
+#| fig-pos: "!ht"
+#| fig-cap: "Proportion of land cover class from the persistent land cover for 2001-2022 (>80%) per macrozone"
+#| out-width: "\\textwidth"
+knitr::include_graphics("../output/figs/LC_pers80_per_macrozone.png")
+```
+
+For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands.
+[@fig-studyArea]c shows the spatial distribution of the macroclasses through Chile for the year 2022.
+[@fig-studyArea]d shows the macroclasses of land cover persistance (80%) during 2021–2022, respectively (Table \ref{tab-landcoverSurf}).
+Within continental Chile, barren land is the land cover class with the highest surface area (277,870 km2).
+The largest type of vegetation, with 137,085 km2, is forest.
+Grassland (74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland (3,146 km2) are the other types (Table \ref{tab-landcoverSurf}).
+The macrozones with major changes for 2001–2022 were "Centro," "Sur," and "Austral," with 36%, 31%, and 34% of their surface changing the type of land cover, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend}).
+@fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.
+
+```{=latex}
+\begin{table}[!ht]
+\caption{The value of Sen's slope trend next to the time-series plot of surface per land cover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile.
+Values of zero indicate that there was not a significant trend.
+Red dots on the plots indicate the maximum and minimum values of surface.}
+\label{tab-landcoverTrend}
+\includegraphics[]{../output/figs/table_var_landcover_macro.png}
+\end{table}
+```
+
+The "Norte Chico" shows an increase in barren land of 111 $km^2 yr^{-1}$  and a reduction in the class savanna of 70 $km^2 yr^{-1}$.
+In the "Centro" and "Sur," there are changes with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$ ), and an increase in shrubland and grassland.
+Showing a change for more dense vegetation types.
+It appears to be a shift in the area cultivated (croplands) from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," replacing the savanna lost (Table \ref{tab-landcoverTrend}).
+
 ### Relationship between drought indices and land cover change
 
-According to Table \ref{tab-landcoverTrendRF}, barren land is the type of land cover that is most affected by the trend of drought indices that reach a rsq = 0.4.
-This is attributed to the fact that it is linked to short-term deficits in precipitation (SPI-6 and SPEI-6) and long-term AED (EDDI-12 and EDDI-24).
-The trend in drought indices accounts for about a quarter of the change in grassland, savanna, and shrubland types.
-These types are most affected by the increase in short-term water demand (EDDI-6).
-The changes in forests are strongly impacted by the increase in long-term water demand (EDDI-36).
-The trend in croplands cannot be attributed to variations in water demand, water supply, soil moisture, or vegetation productivity.
+According to Table \ref{tab-landcoverTrendRF}, the trends in drought indices reach an r-squared between 0.32 and 0.39 for the types of forest, grassland, savanna, shrubland, and barren land.
+It is more likely that short- and medium-term increases in AED (EDDI-6 and EDDI-12) and short-term precipitation deficits (SPI-6 and SPEI-6) contributed to changes in grassland and bare land.
+The short-term increase of AED (EDDI-3 and EDDI-6) and the longer duration of the precipitation deficit (SPI-24, SPI-36, and SPEI-36) most likely contribute to the changes in shrubland.
+The changes in savanna are associated with a short- and long-term increase in AED and a three-year precipitation deficit (SPI-36).
+The increase in cumulative AED from 12 to 36 months is the strongest associated variable that contributes to changes in forests, followed by the reduction of soil moisture over six and 36 months.
+More details about the results of the variable importance can be found in the supplementary material in Section S3.
 
@@
 #| fig-pos: "!ht"
-#| fig-cap: "Relationship between the trend in land cover change and the trend in drought indices for the five macrozones.
+#| fig-cap: "Relationship between the trend in land cover change (y-axis) and the trend in drought indices (x-axis) for the five macrozones.
 Vertical panels correspond to 1, 3, 6, 12, 24, and 36 months of the time scale by drought index.
@@
 
-We look at the relationship between drought index and land cover change in @fig-TrendsLandDrought by comparing the trends in land cover change (in terms of the total surface area per land cover type and macrozone) and drought indexes.
-@fig-TrendsLandDrought shows that the negative trends in cropland (-0.029) and savanna (-0.026) in “Norte Chico” are the highest and are associated with an increase trend in EDDI and a decreasing trend in SPI, SPEI, and SSI.
-On the contrary, the shrubland in “Centro” has an increase (0.04) linked to a decrease in SPI, SPEI, and SSI and to an increase in EDDI.
-In "Austral,” the positive trend in shrubland fits with the positive trend in all the drought indices.
-The rest of the land cover types in the macrozones show weak associations with water supply, water demand, or soil moisture drought indices.
+We study the connection between the SPI, EDDI, and SSI drought indices and changes in land cover in @fig-TrendsLandDrought.
+To do this, we compare the relative changes in land cover (in terms of the total surface area per land cover type and macrozone) over six and thirty-six months.
+@fig-TrendsLandDrought shows that the forest in the "Sur," shrubland and grassland in "Centro," barren land in "Norte Chico," and savanna in "Austral" showed an increase in the surface of landcover associated with an increase in EDDI.
+Savanna in "Centro," "Sur," and "Norte Chico" decreases with the increase in EDDI.
+The SPI and SSI showed similar behavior regarding the trend in land cover type.
+A decrease in SPI and SSI is associated with an increase in the surface in shurubland and grassland in "Centro," forest in "Sur," and barren land in "Norte Chico," as well as a decrease trend in savanna in "Norte Chico," "Centro," and "Sur."
 
@@
 
-In @fig-zcNDVI_var it is showed the spatial map of trends in zcNDVI ([fig-zcNDVI_var]a) and the temporal variation of zcNDVI within the aggregated macrozones ([fig-zcNDVI_var]b).
+In @fig-zcNDVI_var it is showed the spatial map of trends in zcNDVI ([@fig-zcNDVI_var]a) and the temporal variation of zcNDVI within the aggregated macrozones ([@fig-zcNDVI_var]b).
 In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.02 with respect to grassland and shrubland categories.

### 2024-03-19 article/paper_drought_Chile2023.tex
update article

@@
 \hypersetup{
-  pdftitle={The effects of multidimensional drought on land cover change and vegetation productivity in continental Chile},
+  pdftitle={The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile},
   pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
@@
 \begin{frontmatter}
-\title{The effects of multidimensional drought on land cover change and
+\title{The effects of multi-dimensional drought on land cover change and
 vegetation productivity in continental Chile}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, interior hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, frame hidden, enhanced, interior hidden, breakable]}{\end{tcolorbox}}\fi
 
@@
 cover change and vegetation productivity are: i) How do they respond to
-short- to long-term meteorological droughts? ii) How do they behave in
-humid and arid climatic zones regarding drought? And iii) What is the
-role of soil moisture? Likewise, there is a lack of understanding of how
-the alteration in water supply and demand is affecting land cover
-transformations.
+short- to long-term meteorological and soil moisture droughts? ii) How
+is the drought impacting land cover changes? And iii) How do they behave
+in humid and arid climatic zones regarding drought? Likewise, there is a
+lack of understanding of how the alteration in water supply and demand
+is affecting land cover transformations.
+
+For monitoring drought, the World Meteorological Organization recommends
+the SPI (Standardized Precipitation Index) \citep{WMO2012}. The SPI is a
+multi-scalar drought index that only uses precipitation to assess short-
+to long-term droughts. The primary cause of drought is precipitation
+anomalies, and temperature usually makes it worse \citep{Luo2017}.
+Nowadays, there is an increase in attention toward using AED separately
+to monitor droughts \citep{Vicente-Serrano2020}. One reason is due to
+its attribution to increasing flash droughts in water-limited regions
+\citep{Noguera2022}. \citet{Vicente-Serrano2010} proposed the
+Standardized Precipitation Evapotranspiration Index (SPEI), which
+incorporated the temperature effect by subtracting AED from
+precipitation. SPEI allows for analysis of the combined effect of
+precipitation and AED. Since its formulation, it has been used worldwide
+for the study and monitoring of drought
+\citep{Gebrechorkos2023, Liu2021}. \citet{Hobbins2016} and
+\citet{McEvoy2016} developed the Evaporative Demand Drought Index (EDDI)
+to monitor droughts solely using the AED, and it has proven effective in
+monitoring flash droughts \citep{Li2024, Ford2023}. For soil moisture,
+several drought indices exist, such as the Soil Moisture Deficit Index
+(SDMI) \citep{Narasimhan2005} and the Soil Moisture Agricultural Drought
+Index (SMADI) \citep{Souza2021}. \citet{Hao2013} and
+\citet{AghaKouchak2014} proposed the Standardized Soil Moisture Index
+(SSI), which has a similar formulation as the SPI, SPEI, and EDDI. Thus,
+there are plenty of drought indices that allow for a comprehensive
+assessment of drought from short- to long-term scales and consider
+variables from the earth's water balance on their own (e.g.,
+precipitation, AED, soil moisture). Using this information, we can
+advance our understanding of the impact of drought on ecosystems.
 
 Chile's diverse climatic and ecosystem types
-(\citet{Beck2023};\citet{Luebert2022}) make it an ideal natural
-laboratory for studying climate and ecosystems. Additionally, the
-country has experienced severe drought conditions that have had
-significant effects on vegetation and water storage. Central Chile faced
-a persistent precipitation deficit between 2010 and 2022, defined as a
-megadrought \citep{Garreaud2017}, which has impacted the Chilean
-ecosystem. This megadrought was defined by the Standardized
-Precipitation Index (SPI) of twelve months in December having values
-below one standard deviation. Some studies have addressed how this
-drought affects single ecosystems in terms of forest development
-\citep{Miranda2020, Venegas2018}, forest fire occurrence
-\citep{UrrutiaJalabert2018}, and crop productivity
+\citep{Beck2023, Luebert2022} make it an ideal natural laboratory for
+studying climate and ecosystems. Additionally, the country has
+experienced severe drought conditions that have had significant effects
+on vegetation and water storage. North-central Chile has faced a
+persistent precipitation deficit since 2010, defined as a mega drought.
+\citep{Garreaud2017}, which has impacted the Chilean ecosystem. This
+megadrought was defined by the Standardized Precipitation Index (SPI) of
+twelve months in December having values below one standard deviation.
+Some studies have addressed how this drought affects single ecosystems
+in terms of forest development \citep{Miranda2020, Venegas2018}, forest
+fire occurrence \citep{UrrutiaJalabert2018}, and crop productivity
 \citep{Zambrano2023, Zambrano2018, Zambrano2016}. We found one study
@@
 years, resulting in a permanent deficit that impacts the hydrological
-system \citep{Boisier2018}. Hence, it is imperative to assess temporal
-scales that take into account the cumnulative effect within some years.
-There is little knowledge about the relationship between drought and
-ecosystem in Chile; thus, it is important to understand in more detail
-how meteorological and soil moisture droughts influence ecosystem
-dynamics to inform adaptation options.
-
-A detailed spatiotemporal assessment of the interaction of drought for
-short- to long-term and land cover change requires information on
-vegetation as well as weather variables such as precipitation,
-temperature, and soil moisture. Weather networks in Chile present some
-disadvantages, such as spatio-temporal gaps, a short history, and
-irregular quality, which make them difficult to represent the whole
-extent of the country spatially. In order to do this, we use reanalysis
-data from ERA5-Land \citep{MunozSabater2021} to create drought indices
-that consider AED, precipitation, and soil moisture over a range of time
-periods, from the short to the long term. Also, we use vegetation
-spectral information and annual land cover change from the
-Moderate-Resolution Imaging Spectroradiometer (MODIS). We expect to gain
-insight regarding the temporal evolution of water demand, water supply,
-and soil moisture, as well as the interaction with land cover change and
-vegetation productivity. Here, we analyze the multi-dimensional impacts
-of drought across ecosystems in continental Chile. More specifically, we
-aim to assess: i) temporal changes in land-use cover and the direction
-and magnitude of their relationships with drought indices for water
-demand and supply, soil moisture, and vegetation productivity; ii)
-short- to long-term temporal trends in multi-scalar drought indices; and
-iii) the relationship between vegetation productivity and drought
-indices for water demand and supply and soil moisture across Chilean
-ecosystems.
+system \citep{Boisier2018}. Therefore, it is crucial to evaluate
+temporal scales that consider the cumulative impact over a period of
+several years. There is limited understanding of the correlation between
+drought and the environment in Chile. Therefore, it is crucial to have a
+deeper understanding of how meteorological and soil moisture droughts
+impact the dynamics of the environment in order to make informed
+decisions on how to adapt.
+
+Here, we analyze the multi-dimensional impacts of drought across
+ecosystems in continental Chile. More specifically, we aim to assess: i)
+short- to long-term temporal trends in multi-scalar drought indices; ii)
+temporal changes in land-use cover and the direction and magnitude of
+their relationships with trends in drought indices; and iii) the trend
+in vegetation productivity and its relationship with drought indices
+across Chilean ecosystems.
 
@@
 
-Continental Chile has a diverse climate conditions with strong gradients
+Continental Chile has diverse climate conditions with strong gradients
 from north to south and east to west \citep{Aceituno2021}
 (Figure~\ref{fig-studyArea} a), which determines its great ecosystem
-diversity (\citet{Luebert2022}) (Figure~\ref{fig-studyArea} c). The
-Andes Mountains are a main factor in climate latitudinal variation
-\citep{Garreaud2009}. In order to characterize the climate and ecosystem
-of Chile, we utilize the Köppen-Geiger classification system developed
-by \citet{Beck2023} and the land cover data derived from the MODIS
-product for the period of 2001--2022, based on the International
-Geosphere-Biosphere Programme (IGBP) classification scheme proposed by
-\citet{Friedl2019}. ``Norte Grande'' and ``Norte Chico'' predominate in
-an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At
+diversity \citep{Luebert2022} (Figure~\ref{fig-studyArea} c). The Andes
+Mountains are a main factor in climate latitudinal variation
+\citep{Garreaud2009}. ``Norte Grande'' and ``Norte Chico'' predominate
+in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At
 the south of ``Norte Chico,'' the climate changes to an arid steppe with
 cold temperatures (Bsk). In these two northern regions, the land is
-mostly bare, with a minor surface of vegetation types such as shrubland
+mostly bare, with a small surface of vegetation types such as shrubland
 and grassland. In the zones ``Centro'' and the north half of ``Sur,''
@@
 grassland. The southern part of the country has a tundra climate, and in
-``Austral'', it is a cold semi-arid area with an extended surface of
+``Austral,'' it is a cold semi-arid area with an extended surface of
 grassland, forest, and, to a lesser extent, savanna.
@@
 
-To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
-6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
-The IGBP defines 17 classes; from these, we regrouped into ten
-macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9
-to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
-13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
-Thus, we have a land cover raster time series with the ten classes for
-2001 and 2023. Prior to using this class, we compared it with the types
-of land cover made by \citet{Zhao2016}, which is a more detailed land
-cover map of Chile with a 30 m spatial resolution for 2012--2013. It
-reached a global accuracy of \textasciitilde0.82. The procedure of
-validation is demonstrated in Section S1 of the supplementary
-material.To derive a proxy for vegetation productivity, we used the
-MOD13A3 collection 6.1 product from MODIS \citep{Didan2015}. It provides
-vegetation indices (NDVI and EVI) at 1km of spatial resolution and
-monthly frequency. The MOD13A3 and MCD12Q1 were retrieved from the
-online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed
-Active Archive Center (LP DAAC), USGS Earth Resources Observation and
-Science (EROS) Center, Sioux Falls, South Dakota,
+To analyze land cover change, we use the classification scheme by the
+IGBP (International Geosphere-Biosphere Programme) from the product
+MCD12Q1 collection 6.1 from MODIS. To derive a proxy for vegetation
+productivity, we used the Normalized Difference Vegetation Index (NDVI)
+from the product MOD13A3 collection 6.1 from MODIS \citep{Didan2015}.
+MOD13A3 provides vegetation indices at 1km of spatial resolution and
+monthly frequency. The NASA EOSDIS Land Processes Distributed Active
+Archive Center (LP DAAC), USGS Earth Resources Observation and Science
+(EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and
+MCD12Q1 from the online Data Pool, accessible at
 https://lpdaa.usgs.gov/tools/data-pool/.
@@
 For soil moisture, water supply, and water demand variables, we used
-ERA5L \citep{MunozSabater2021}, a reanalysis dataset that provides the
-evolution of the meteorological and soil moisture variables since 1950.
-It has a spatial resolution of 0.1° (9 km), hourly frequency, and global
-coverage. We selected the variables for total precipitation, maximum and
-minimum temperature at 2 meters, and volumetric soil water layers
-between 0 and 100cm of depth (layer 1 to layer 3).
-
-\hypertarget{land-cover-change-and-trend}{%
-\subsection{Land cover change and
-trend}\label{land-cover-change-and-trend}}
-
-To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
-6.1 from MODIS. \citet{Zambrano2018} and \citet{Fuentes2021} have
-previously used this product for studies of drought and land cover. The
-MCD12Q1 has a yearly frequency from 2001 to 2022. The IGBP defines 17
-classes; from these, we regrouped into ten macroclasses, as follows:
-classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as
-grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as
-snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land
-cover raster time series with the ten macroclasses for 2001 and 2023. We
-validate the land cover macroclasses regarding a highly detailed (30 m
-of spatial resolution) land cover map made for Chile by \citet{Zhao2016}
-for 2013-2014. Our results showed a global accuracy of
-\textasciitilde0.82 and a F1 score of \textasciitilde0.66. Section S2 in
-the Supplementary Material shows the procedure for validation.
+ERA5L (ECMWF Reanalysis version 5 over land) \citep{MunozSabater2021}, a
+reanalysis dataset that provides the evolution of atmospheric and land
+variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly
+frequency, and global coverage. We selected the variables for total
+precipitation, maximum and minimum temperature at 2 meters, and
+volumetric soil water layers between 0 and 100cm of depth (layer 1 to
+layer 3). Table \ref{tab-desEOD} shows a summary of the data and its
+main characteristics.
+
+\hypertarget{trend-of-short--to-long-term-drought}{%
+\subsection{Trend of short- to long-term
+drought}\label{trend-of-short--to-long-term-drought}}
+
+\hypertarget{atmospheric-evaporative-demand-aed}{%
+\subsubsection{Atmospheric Evaporative Demand
+(AED)}\label{atmospheric-evaporative-demand-aed}}
+
+In order to compute the drought indices that use water demand, it is
+necessary to first calculate the AED. To do this, we employed the
+Hargreaves method \citep{Hargreaves1994, Hargreaves1985} by applying the
+following equation:
+
+\begin{equation}\protect\hypertarget{eq-AED}{}{AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}}\label{eq-AED}\end{equation}
+
+where \(Ra\) \((MJ\,m^2\, day^{-1})\) is extraterrestrial radiation;
+\(T\), \(T_{max}\), and \(T_{min}\) are mean, maximum, and minimum
+temperature \((°C)\) at 2m. For calculating \(Ra\) we used the
+coordinate of the latitud of the centroid of each pixel. We chose the
+method of Hargreaves to estimate AED because of its simplicity, which
+only requires temperatures and extrarrestrial radiation. Also, it has
+been recommended over other methods (e.g., Penman-Monteith) when the
+access to climatic variables is limited \citep{Vicente-Serrano2014}.
+
+\hypertarget{non-parametric-calculation-of-drought-indices}{%
+\subsubsection{Non-parametric calculation of drought
+indices}\label{non-parametric-calculation-of-drought-indices}}
+
+To derive the drought indices of water supply and demand, soil moisture,
+and vegetation (i.e., the proxy of productivity), we used the ERA5L
+dataset and the MODIS product, with a monthly frequency for 1981--2023
+and 2000--2023, respectively.
+
+The dought indices correspond to a historical anomaly with regard to a
+variable (e.g., meteorological, vegetation, or soil moisture). To
+account for the anomaly, the common practice is to derive it following a
+statistical parametric methodology in which it is assumed that the
+statistical distribution of the data is known (\citet{Heim2002}). A
+wrong decision is usually the highest source of uncertainty
+(\citet{Laimighofer2022}). In the case of Chile, due to its high degree
+of climatic variability, it is complex to choose a proper distribution
+without previous research. Here, we follow a non-parametric methodology
+for the calculation of the drought indices, in a similar manner as the
+framework proposed by \citet{Farahmand2015};
+\citet{Hobbins2016};\citet{McEvoy2016}.
+
+For the purpose of monitoring water supply drought, we used the
+well-known Standardized Precipitation Index (SPI), which the World
+Meteorological Organization (WMO) recommended. The SPI solely relies on
+precipitation data. Also, it has been used worldwide for the study of
+drought, including in Chile (\citet{Garreaud2017};
+\citet{Zambrano2017}). The primary cause of drought is precipitation
+anomalies, and temperature usually makes it worse \citep{Luo2017}.
+Nowadays, there is an increase in attention toward using water demand
+separately to monitor droughts. (\citet{Vicente-Serrano2020};
+\citet{Noguera2022}). Thus, to evaluate water demand, we chose the
+Evaporative Demand Drought Index (EDDI), developed by
+\citet{Hobbins2016} and \citet{McEvoy2016}, which is based on the AED.
+EDDI is currently used for monitoring drought in the United States
+(https://psl.noaa.gov/eddi/). In our case, we used only temperature for
+AED, a difference from the original formulation of EDDI, which also
+considered wind besides temperature. To consider the combined effect of
+water supply and demand, we selected the SPEI, which corresponds to a
+balance between precipitation and AED. \citet{Vicente-Serrano2010}
+proposed the SPEI, and it has improved the SPI by incorporating
+temperature for drought monitoring. For SPEI, an auxiliary variable D =
+P-AED is calculated. Soil moisture is the main driver of vegetation
+productivity, particularly in semi-arid regions (\citet{Li2022}). Hence,
+for soil water drought, we used the SSI (Standardized Soil Moisture
+Index) \citep{Hao2013, AghaKouchak2014} which is a multi-scale index
+similar to SPI, SPEI, and EDDI. In our case, for the SSI, we used the
+average soil moisture from ERA5L at 1m depth. Finally, for the proxy of
+productivity, we used the zcNDVI proposed by \citet{Zambrano2018} which
+will be derived from the NDVI retrieved from MOD13A1.
+
+To derive the drought indices, first we must calculate the sum of the
+variables with regard to the time scale (s). In this case, for
+generalization purposes, we will use \(V\), referring to variables
+\(P\), \(AED\), \(D\), \(NDVI\), and \(SM\) (Table \ref{tab-desEOD}). We
+cumulated each \(V\) over the time series of \(n\) values (months), and
+for the time scales \(s\):
+
+\begin{equation}\protect\hypertarget{eq-sumvar}{}{A_{si} = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  }\label{eq-sumvar}\end{equation}
+
+The \(A_{si}\) corresponds to a moving window (convolution) that sums
+the variable for time scales \(s\) from the last month, month by month,
+until the first month in which it could sum for \(s\) months. Once the
+variable is cumulated over time for the scale \(s\). Thus, the
+empirically derived probabilities are obtained through an inverse normal
+approximation \citep{Abramowitz1968}. Then, we used the empirical Tukey
+plotting position \citep{Wilks2011} over \(A_i\) to derive the
+\(P(a_i)\) probabilities across a period of interest:
+
+\begin{equation}\protect\hypertarget{eq-probPai}{}{P(A_i) = \frac{i-0.33}{n+0.33'}}\label{eq-probPai}\end{equation}
+
+The drought indices \(SPI\), \(SPEI\), \(EDDI\), \(SSI\), and \(zcNDVI\)
+are obtained following the inverse normal approximation:
+
+\begin{equation}\protect\hypertarget{eq-DI}{}{DI(A_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}}\label{eq-DI}\end{equation}
+
+\(DI\) is referring to the drought index calculated for the variable
+\(V\). The values for the constats are: \(C_0 = 2.515517\),
+\(C_1 = 0.802853\), \(C_2 = 0.010328\), \(d_1 = 1.432788\),
+\(d_2 = 0.189269\), and \(d3 = 0.001308\). For \(P(A) \leq 0.5\),
+W=\(\sqrt{-2\cdot ln(P(A_i))}\) , and for \(P(A_i) > 0.5\), replace
+\(P(A_i)\) with \(1-P(A_i)\) and reverse the sign of \(DI(A_i)\).
+
+The drought indices were calculated for time scales of 1, 3, 6, 12, 24,
+and 36 months at a monthly frequency for 1981--2023 in order to be used
+for short- to long-term evaluation of drought. In the case of the proxy
+of vegetation productivity (zcNDVI) it was calculated for a time scale
+of six months at monthly frequency for 2000--2023. For zcNDVI, we test
+time scales of 1, 3, 6, and 12 months; we choose to use six months
+because that shows a more robust representation of vegetation
+productivity due to the seasonality of vegetation in Chile.
+
+\hypertarget{trend-of-drought-indices}{%
+\subsubsection{Trend of drought
+indices}\label{trend-of-drought-indices}}
+
+To estimate if there are significant positive or negative trends for the
+drought indices, we used the non-parametric test of Mann-Kendall
+\citep{Kendall1975}. To determine the magnitude of the trend, we used
+Sen's slope \citep{Sen1968}. Some of the advantages of applying this
+methodology are that the Sen's slope is not affected by outliers as
+regular regression does, and it is a non-parametric method that is not
+influenced by the distribution of the data. We applied Mann-Kendall to
+see if the trend was significant and Sen's slope to estimate the
+magnitude of the trend. We did this to the six time scales from 1981 to
+2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI. Thus,
+we have trends per index and time scale (24 in total). Then, we
+extracted the trend aggregated by macrozone and per land cover persitent
+macroclasses.
+
+\hypertarget{interaction-of-land-cover-and-drought}{%
+\subsection{Interaction of land cover and
+drought}\label{interaction-of-land-cover-and-drought}}
+
+\hypertarget{land-cover-change}{%
+\subsubsection{Land cover change}\label{land-cover-change}}
+
+To analyze the land cover change, we use the IGBP scheme from the
+MCD12Q1 collection 6.1 from MODIS. \citet{Zambrano2018} and
+\citet{Fuentes2021} have previously used this product for studies of
+drought and land cover in Chile. The MCD12Q1 has a yearly frequency from
+2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into
+ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands,
+8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to
+croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to
+water bodies. Thus, we have a land cover raster time series with the ten
+macroclasses for 2001 and 2023. We validate the land cover macroclasses
+regarding a highly detailed (30 m of spatial resolution) land cover map
+made for Chile by \citet{Zhao2016} for 2013-2014. Our results showed a
+global accuracy of \textasciitilde0.82 and a F1 score of
+\textasciitilde0.66. Section S2 in the Supplementary Material shows the
+procedure for validation.
 
@@
 
-\hypertarget{trend-of-drought-and-interaction-to-land-cover-change}{%
-\subsection{Trend of drought and interaction to land cover
-change}\label{trend-of-drought-and-interaction-to-land-cover-change}}
-
-\hypertarget{atmospheric-evaporative-demand-aed}{%
-\subsubsection{Atmospheric Evaporative Demand
-(AED)}\label{atmospheric-evaporative-demand-aed}}
-
-In order to compute the drought indices that uses water demand, it is
-necessary to first calculate the AED. To do this, we employed the
-Hargreaves technique \citep{Hargreaves1994, Hargreaves1985}, by applying
-the following equation:
-
-\begin{equation}\protect\hypertarget{eq-AED}{}{AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}}\label{eq-AED}\end{equation}
-
-where \(Ra\) \((MJ\,m^2\, day^{-1})\) is extraterrestrial radiation;
-\(T\), \(T_{max}\), and \(T_{min}\) are mean, maximum, and minimum
-temperature \((°C)\). We calculate the centroid coordinates per pixel
-and use the latitude to estimate \(Ra\).
-
-We chose the method of Hargreaves to estimate AED because of its
-simplicity, which only requires temperatures and extrarrestrial
-radiation. Also, it has been recommended over other methods (e.g.,
-Penman-Monteith) when the access to climatic variables is limited
-\citep{Vicente-Serrano2014}.
-
-\hypertarget{non-parametric-calculation-of-drought-indices}{%
-\subsubsection{Non-parametric calculation of drought
-indices}\label{non-parametric-calculation-of-drought-indices}}
-
-To derive the drought indices of water supply and demand, soil moisture,
-and vegetation we used the ERA5L dataset and the MODIS product, with a
-monthly frequency for 1981-2023 and 2000-2023, respectively.
-
-The dought indices correspond to a historical anomaly with regard to a
-variable (e.g., meteorological, vegetation, or soil moisture). To
-account for the anomaly, the common practice is to derive it following a
-statistical parametric methodology in which it is assumed that the
-statistical distribution of the data is known (\citet{Heim2002}). A
-wrong decision is usually the highest source of uncertainty
-(\citet{Laimighofer2022}). In the case of Chile, due to its high degree
-of climatic variability, it is complex to choose a proper distribution
-without previous research. Here, we follow a non-parametric methodology
-for the calculation of the drought indices, in a similar manner as the
-framework proposed by \citet{Farahmand2015};
-\citet{Hobbins2016};\citet{McEvoy2016}.
-
-For the purpose of monitoring water supply drought, we used the
-well-known Standardized Precipitation Index (SPI), which the World
-Meteorological Organization (WMO) recommended. The SPI solely relies on
-precipitation data. Also, it has been used worldwide for the study of
-drought, including in Chile (\citet{Garreaud2017};
-\citet{Zambrano2017}). The primary cause of drought is precipitation
-anomalies, and temperature usually makes it worse (Luo et al.~2017).
-Nowadays, there is an increase in attention toward using water demand
-separately to monitor droughts. (\citet{Vicente-Serrano2020};
-\citet{Noguera2022}). Thus, to evaluate water demand, we chose the
-Evaporative Demand Drought Index (EDDI), developed by Hobbins et
-al.~(016) and McEvoy et al.~(2016), which is based on the AED. EDDI is
-currently used for monitoring drought in the United States
-(https://psl.noaa.gov/eddi/). In our case, we used only temperature for
-AED, a difference from the original formulation of EDDI, which also
-considered wind besides temperature. To consider the combined effect of
-water supply and demand, we selected the SPEI, which corresponds to a
-balance between precipitation and AED. \citet{Vicente-Serrano2010}
-proposed the SPEI, and it has improved the SPI by incorporating
-temperature for drought monitoring. For SPEI, an auxiliary variable D =
-P-AED is calculated. Soil moisture is the main driver of vegetation
-productivity, particularly in semi-arid regions (\citet{Li2022}). Hence,
-for soil water drought, we used the SSI (Standardized Soil Moisture
-Index) (Hao and AghaKouchak 2013; A. AghaKouchak 2014) which is a
-multi-scale index similar to SPI, SPEI, and EDDI. In our case, for the
-SSI, we used the average soil moisture from ERA5L at 1m depth. Finally,
-for the proxy of productivity, we used the zcNDVI proposed by Zambrano
-et al.~(2018) which will be derived from the NDVI retrieved from
-MOD13A1.
-
-To derive the drought indices, first we must calculate the sum of the
-variables with regard to the time scale (s). In this case, for
-generalization purposes, we will use \(V\), referring to variables
-\(P\), \(AED\), \(D\), \(NDVI\), and \(SM\) (Table \ref{tab-desEOD}). We
-cumulated each \(V\) over the time series of \(n\) values (months), and
-for the time scales \(s\):
-
-\begin{equation}\protect\hypertarget{eq-sumvar}{}{A_{si} = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  }\label{eq-sumvar}\end{equation}
-
-The \(A_{si}\) corresponds to a moving window (convolution) that sums
-the variable for time scales \(s\) from the last month, month by month,
-until the first month in which it could sum for \(s\) months. Once the
-variable is cumulated over time for the scale \(s\). Thus, the
-empirically derived probabilities are obtained through an inverse normal
-approximation \citep{Abramowitz1968}. Then, we used the empirical Tukey
-plotting position \citep{Wilks2011} over \(A_i\) to derive the
-\(P(a_i)\) probabilities across a period of interest:
-
-\begin{equation}\protect\hypertarget{eq-probPai}{}{P(A_i) = \frac{i-0.33}{n+0.33'}}\label{eq-probPai}\end{equation}
-
-The drought indices \(SPI\), \(SPEI\), \(EDDI\), \(SSI\), and \(zcNDVI\)
-are obtained following the inverse normal approximation:
-
-\begin{equation}\protect\hypertarget{eq-DI}{}{DI(A_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}}\label{eq-DI}\end{equation}
-
-\(DI\) is referring to the drought index calculated for the variable
-\(V\). The values for the constats are: \(C_0 = 2.515517\),
-\(C_1 = 0.802853\), \(C_2 = 0.010328\), \(d_1 = 1.432788\),
-\(d_2 = 0.189269\), and \(d3 = 0.001308\). For \(P(A) \leq 0.5\),
-W=\(\sqrt{-2\cdot ln(P(A_i))}\) , and for \(P(A_i) > 0.5\), replace
-\(P(A_i)\) with \(1-P(A_i)\) and reverse the sign of \(DI(A_i)\).
-
-The drought indices were calculated for time scales of 1, 3, 6, 12, 24,
-and 36 months at a monthly frequency for 1981--2023 in order to be used
-for short- to long-term evaluation of drought. In the case of the proxy
-of vegetation productivity (zcNDVI) it was calculated for a time scale
-of six months at monthly frequency for 2000--2023. For zcNDVI, we test
-time scales of 1, 3, 6, and 12 months; we choose to use six months
-because that shows a more robust representation of vegetation
-productivity due to the seasonality of vegetation in Chile.
-
-\hypertarget{trend-of-drought-indices}{%
-\subsubsection{Trend of drought
-indices}\label{trend-of-drought-indices}}
-
-To estimate if there are significant positive or negative trends for the
-drought indices, we used the non-parametric test of Mann-Kendall
-\citep{Kendall1975}. To determine the magnitude of the trend, we used
-Sen's slope \citep{Sen1968}. Some of the advantages of applying this
-methodology are that the Sen's slope is not affected by outliers as
-regular regression does, and it is a non-parametric method that is not
-influenced by the distribution of the data. We applied Mann-Kendall to
-see if the trend was significant and Sen's slope to estimate the
-magnitude of the trend. We did this to the six time scales from 1981 to
-2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI. Thus,
-we have trends per index and time scale (24 in total). Then, we
-extracted the trend aggregated by macrozone and per land cover persitent
-macroclasses.
-
-\hypertarget{intercation-drought-indices-and-land-cover-change}{%
-\subsubsection{Intercation drought indices and land cover
-change}\label{intercation-drought-indices-and-land-cover-change}}
+\hypertarget{relationship-between-land-cover-and-drought-trends}{%
+\subsubsection{Relationship between land cover and drought
+trends}\label{relationship-between-land-cover-and-drought-trends}}
 
@@
 
-\hypertarget{drought-impacts-on-vegetation-productivity-within-land-cover}{%
-\subsection{Drought impacts on vegetation productivity within land
-cover}\label{drought-impacts-on-vegetation-productivity-within-land-cover}}
+\hypertarget{drought-impacts-on-vegetation-productivity}{%
+\subsection{Drought impacts on vegetation
+productivity}\label{drought-impacts-on-vegetation-productivity}}
 
@@
 
-\hypertarget{land-cover-change-and-trend-1}{%
-\section{Land cover change and
-trend}\label{land-cover-change-and-trend-1}}
+\hypertarget{trend-of-short--to-long-term-drought-1}{%
+\subsection{Trend of short- to long-term
+drought}\label{trend-of-short--to-long-term-drought-1}}
+
+Figure~\ref{fig-trendDI} shows the spatial variation of the trend for
+the drought indices from short- to long-term scales. SPI and SPEI have a
+decreasing trend from ``Norte Chico'' to ``Sur.'' However, there is an
+increasing trend in ``Austral.'' The degree of the trend is stronger at
+higher time scales. The SSI indicates that in ``Norte Grande,'' there
+are surfaces that have increased in the soutwest part and in the
+northeast have decreased, and is shown for all time scales. Similar to
+SPI and SPEI, SSI decreases at higher time scales. EDDI showed a
+positive trend for the whole of continental Chile, with a higher trend
+toward the north and a descending gradient toward the south. The degree
+of trend increases at higher time scales.
+
+\blandscape
+
+\begin{figure}
+
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendDI-1}SPI (Standardized Precipitation Index)}
+\end{minipage}%
+%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendDI-2}SPEI (Standardized Precipitation
+Evapotranspiration Index)}
+\end{minipage}%
+\newline
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendDI-3}EDDI (Evaporative Demand Drought
+Index)}
+\end{minipage}%
+%
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
+
+\includegraphics{../output/figs/trend_raster_zcSM_1981-2023.png}
+
+}
+
+}
+
+\subcaption{\label{fig-trendDI-4}SSMI (Standardized Soil Moisture
+Index)}
+\end{minipage}%
+
+\caption{\label{fig-trendDI}Linear trend of the drought index (*) at
+time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023}
+
+\end{figure}
+
+\elandscape
+
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/trend_macrozone_drought_indices.png}
+
+}
+
+\caption{\label{fig-trendDIMacro}Trend per decade for the drought
+indices SPI, EDDI, SPEI, and SSI aggregated by macrozone.}
+
+\end{figure}
+
+In Figure~\ref{fig-trendDIMacro}, the averaged aggregation per
+macrozone, drought index, and time scale are shown. The macrozones that
+have the lowest trend are ``Norte Chico'' and ``Centro,'' where the SPI,
+SPEI, and SSI show that it decreases at longer time scales. Potentially
+explained due to the prolonged reduction in precipitation that has
+affected the hydrological system in Chile. At 36 months, it reaches
+trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and
+SSI. For ``Sur,'' the behavior is similar, decreasing at longer scales
+and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI.
+``Norte Grande'' has the highest trend at 36 months for EDDI (0.042 per
+decade), and ``Centro'' has the lowest for SPI and SPEI. In ``Norte
+Grande'' and ``Norte Chico,'' which are in a semi-arid climate, it is
+evident that the EDDI has an effect on the difference between the SPI
+and SPEI index, which is not seen in the other macrozones. Contrary to
+the other macrozones, ``Austral'' showed an increase in all indices,
+being the highest for EDDI at 36 months (0.025) and the lowest for SSI,
+which shows only a minor increase in the trend.
+
+\hypertarget{interaction-of-land-cover-and-drought-1}{%
+\subsubsection{Interaction of land cover and
+drought}\label{interaction-of-land-cover-and-drought-1}}
+
+\hypertarget{land-cover-change-1}{%
+\subsection{Land cover change}\label{land-cover-change-1}}
 
@@
 
-\hypertarget{trend-of-drought-and-interaction-to-land-cover-change-1}{%
-\subsection{Trend of drought and interaction to land cover
-change}\label{trend-of-drought-and-interaction-to-land-cover-change-1}}
-
-\hypertarget{trend-of-drought-indices-1}{%
-\subsubsection{Trend of drought
-indices}\label{trend-of-drought-indices-1}}
-
-Figure~\ref{fig-trendDI} shows the spatial variation of the trend for
-the drought indices from short- to long-term scales. SPI and SPEI have a
-decreasing trend from ``Norte Chico'' to ``Sur.'' However, there is an
-increasing trend in ``Austral.'' The degree of the trend is stronger at
-higher time scales. The SSI indicates that in ``Norte Grande,'' there
-are surfaces that have increased in the soutwest part and in the
-northeast have decreased, and is shown for all time scales. Similar to
-SPI and SPEI, SSI decreases at higher time scales. EDDI showed a
-positive trend for the whole of continental Chile, with a higher trend
-toward the north and a descending gradient toward the south. The degree
-of trend increases at higher time scales.
-
-\blandscape
-
-\begin{figure}
-
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-1}SPI (Standardized Precipitation Index)}
-\end{minipage}%
-%
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-2}SPEI (Standardized Precipitation
-Evapotranspiration Index)}
-\end{minipage}%
-\newline
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-3}EDDI (Evaporative Demand Drought
-Index)}
-\end{minipage}%
-%
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_zcSM_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-4}SSMI (Standardized Soil Moisture
-Index)}
-\end{minipage}%
-
-\caption{\label{fig-trendDI}Linear trend of the drought index (*) at
-time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023}
-
-\end{figure}
-
-\elandscape
-
-\begin{figure}[!ht]
-
-{\centering \includegraphics{../output/figs/trend_macrozone_drought_indices.png}
-
-}
-
-\caption{\label{fig-trendDIMacro}Trend per decade for the drought
-indices SPI, EDDI, SPEI, and SSI aggregated by macrozone.}
-
-\end{figure}
-
-In Figure~\ref{fig-trendDIMacro}, the averaged aggregation per
-macrozone, drought index, and time scale are shown. The macrozones that
-have the lowest trend are ``Norte Chico'' and ``Centro,'' where the SPI,
-SPEI, and SSI show that it decreases at longer time scales. Potentially
-explained due to the prolonged reduction in precipitation that has
-affected the hydrological system in Chile. At 36 months, it reaches
-trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and
-SSI. For ``Sur,'' the behavior is similar, decreasing at longer scales
-and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI.
-``Norte Grande'' has the highest trend at 36 months for EDDI (0.042 per
-decade), and ``Centro'' has the lowest for SPI and SPEI. In ``Norte
-Grande'' and ``Norte Chico,'' which are in a semi-arid climate, it is
-evident that the EDDI has an effect on the difference between the SPI
-and SPEI index, which is not seen in the other macrozones. Contrary to
-the other macrozones, ``Austral'' showed an increase in all indices,
-being the highest for EDDI at 36 months (0.025) and the lowest for SSI,
-which shows only a minor increase in the trend.
-
 \hypertarget{relationship-between-drought-indices-and-land-cover-change}{%
@@
 
-According to Table \ref{tab-landcoverTrendRF}, barren land is the type
-of land cover that is most affected by the trend of drought indices that
-reach a rsq = 0.4. This is attributed to the fact that it is linked to
-short-term deficits in precipitation (SPI-6 and SPEI-6) and long-term
-AED (EDDI-12 and EDDI-24). The trend in drought indices accounts for
-about a quarter of the change in grassland, savanna, and shrubland
-types. These types are most affected by the increase in short-term water
-demand (EDDI-6). The changes in forests are strongly impacted by the
-increase in long-term water demand (EDDI-36). The trend in croplands
-cannot be attributed to variations in water demand, water supply, soil
-moisture, or vegetation productivity.
+According to Table \ref{tab-landcoverTrendRF}, the trends in drought
+indices reach an r-squared between 0.32 and 0.39 for the types of
+forest, grassland, savanna, shrubland, and barren land. It is more
+likely that short- and medium-term increases in AED (EDDI-6 and EDDI-12)
+and short-term precipitation deficits (SPI-6 and SPEI-6) contributed to
+changes in grassland and bare land. The short-term increase of AED
+(EDDI-3 and EDDI-6) and the longer duration of the precipitation deficit
+(SPI-24, SPI-36, and SPEI-36) most likely contribute to the changes in
+shrubland. The changes in savanna are associated with a short- and
+long-term increase in AED and a three-year precipitation deficit
+(SPI-36). The increase in cumulative AED from 12 to 36 months is the
+strongest associated variable that contributes to changes in forests,
+followed by the reduction of soil moisture over six and 36 months. More
+details about the results of the variable importance can be found in the
+supplementary material in Section S3.
 
@@
 \caption{\label{fig-TrendsLandDrought}Relationship between the trend in
-land cover change and the trend in drought indices for the five
-macrozones. Vertical panels correspond to 1, 3, 6, 12, 24, and 36 months
-of the time scale by drought index. Horizontal panels show each drought
-index}
+land cover change (y-axis) and the trend in drought indices (x-axis) for
+the five macrozones. Vertical panels correspond to 1, 3, 6, 12, 24, and
+36 months of the time scale by drought index. Horizontal panels show
+each drought index}
 
@@
 
-We look at the relationship between drought index and land cover change
-in Figure~\ref{fig-TrendsLandDrought} by comparing the trends in land
-cover change (in terms of the total surface area per land cover type and
-macrozone) and drought indexes. Figure~\ref{fig-TrendsLandDrought} shows
-that the negative trends in cropland (-0.029) and savanna (-0.026) in
-``Norte Chico'' are the highest and are associated with an increase
-trend in EDDI and a decreasing trend in SPI, SPEI, and SSI. On the
-contrary, the shrubland in ``Centro'' has an increase (0.04) linked to a
-decrease in SPI, SPEI, and SSI and to an increase in EDDI. In
-``Austral,'' the positive trend in shrubland fits with the positive
-trend in all the drought indices. The rest of the land cover types in
-the macrozones show weak associations with water supply, water demand,
-or soil moisture drought indices.
-
-\hypertarget{drought-impacts-on-vegetation-productivity-within-land-cover-1}{%
+We study the connection between the SPI, EDDI, and SSI drought indices
+and changes in land cover in Figure~\ref{fig-TrendsLandDrought}. To do
+this, we compare the relative changes in land cover (in terms of the
+total surface area per land cover type and macrozone) over six and
+thirty-six months. Figure~\ref{fig-TrendsLandDrought} shows that the
+forest in the ``Sur,'' shrubland and grassland in ``Centro,'' barren
+land in ``Norte Chico,'' and savanna in ``Austral'' showed an increase
+in the surface of landcover associated with an increase in EDDI. Savanna
+in ``Centro,'' ``Sur,'' and ``Norte Chico'' decreases with the increase
+in EDDI. The SPI and SSI showed similar behavior regarding the trend in
+land cover type. A decrease in SPI and SSI is associated with an
+increase in the surface in shurubland and grassland in ``Centro,''
+forest in ``Sur,'' and barren land in ``Norte Chico,'' as well as a
+decrease trend in savanna in ``Norte Chico,'' ``Centro,'' and ``Sur.''
+
+\hypertarget{drought-impacts-on-vegetation-productivity-within-land-cover}{%
 \subsection{Drought impacts on vegetation productivity within land
-cover}\label{drought-impacts-on-vegetation-productivity-within-land-cover-1}}
+cover}\label{drought-impacts-on-vegetation-productivity-within-land-cover}}
 
@@
 In Figure~\ref{fig-zcNDVI_var} it is showed the spatial map of trends in
-zcNDVI ({[}fig-zcNDVI\_var{]}a) and the temporal variation of zcNDVI
-within the aggregated macrozones ({[}fig-zcNDVI\_var{]}b). In ``Norte
-Grande,'' vegetation productivity, as per the z-index, exhibits a yearly
-increase of 0.02 with respect to grassland and shrubland categories.
-There is a negative trend in ``Norte Chico'' with -0.04 and ``Centro''
-with -0.02 per decade. In the ``Norte Chico,'' savanna (-0.05) has the
-lowest trend, and the rest of the types are around -0.04. In ``Centro,''
-shrubland reaches -0.06, grassland -0.05, and croplands and savanna
--0.01 per decade. This could be associated either with a reduction in
-vegetation surface, a decrease in biomass, or browning
-\citep{Miranda2023}. Vegetation reached its lowest values since the year
-2019, with an extreme condition in early 2020 and 2022 in the ``Norte
-Chico'' and ``Centro''. The ``Sur'' and ``Austral'' show a positive
-trend of around 0.016 per decade (Figure~\ref{fig-zcNDVI_var}). Despite
-the croplands suffering from drought just as badly as the native
+zcNDVI (Figure~\ref{fig-zcNDVI_var}a) and the temporal variation of
+zcNDVI within the aggregated macrozones (Figure~\ref{fig-zcNDVI_var}b).
+In ``Norte Grande,'' vegetation productivity, as per the z-index,
+exhibits a yearly increase of 0.02 with respect to grassland and
+shrubland categories. There is a negative trend in ``Norte Chico'' with
+-0.04 and ``Centro'' with -0.02 per decade. In the ``Norte Chico,''
+savanna (-0.05) has the lowest trend, and the rest of the types are
+around -0.04. In ``Centro,'' shrubland reaches -0.06, grassland -0.05,
+and croplands and savanna -0.01 per decade. This could be associated
+either with a reduction in vegetation surface, a decrease in biomass, or
+browning \citep{Miranda2023}. Vegetation reached its lowest values since
+the year 2019, with an extreme condition in early 2020 and 2022 in the
+``Norte Chico'' and ``Centro''. The ``Sur'' and ``Austral'' show a
+positive trend of around 0.016 per decade (Figure~\ref{fig-zcNDVI_var}).
+Despite the croplands suffering from drought just as badly as the native
 vegetation in ``Norte Chico,'' the savanna and shurbland appears to be

### 2024-03-19 article/supplementary.qmd
update article

@@
 The validation of IGBP2013 and IGBP2014 with LCChile reached near the same metrics of performance, having an accuracy of \~0.82 and a F1 score of \~0.66.
+
+# Relationship between drought indices and land cover change
+
+::: {#fig-varImportance layout-ncol=3}
+
+![Forest](../output/figs/fig_errorbar_resample_random_forest_trends_Forest.png){#fig-varimp_forest}
+
+![Cropland](../output/figs/fig_errorbar_resample_random_forest_trends_Cropland.png){#fig-varimp_crops}
+
+![Grassland](../output/figs/fig_errorbar_resample_random_forest_trends_Grassland.png){#fig-varimp_grass}
+
+![Savanna](../output/figs/fig_errorbar_resample_random_forest_trends_Savanna.png){#fig-varimp_savanna}
+
+![Shrubland](../output/figs/fig_errorbar_resample_random_forest_trends_Shrubland.png){#fig-varimp_shrubs}
+
+![Barren land](../output/figs/fig_errorbar_resample_random_forest_trends_Barren_land.png){#fig-varimp_barren}
+
+Error bar for the variables importance obtained from the 10 resamples folded by the Random Forest.
+
+:::

### 2024-03-19 scripts/33_analysis_trend_drought_indices_and_landcover_cuencas.R

@@ -52,7 +52,7 @@ get_rf_imp <- function(x) {
  
 df <- map(1:6,\(i){
   t <- which(vars == vars[[i]])
-  df_split <- initial_split(data_model |> select(-vars[-t]),prop =.8)
+  df_split <- initial_split(data_model |> select(-vars[-t]),prop =0.99)
   df_train <- training(df_split)
   df_test <- testing(df_split)
   
@@ -110,7 +110,7 @@ df <- map(1:6,\(i){
     labs(x = "Variable importance", y = NULL) +
     annotate("text",x=max(df_var_imp$Mean),y=1,label =paste0('rsq=',round(df_met$mean[2],2))) +
     theme_bw()
-  ggsave(paste0('output/figs/fig_errorbar_resample_random_forest_trends_',vars[i],'.png'))
+  ggsave(paste0('output/figs/fig_errorbar_resample_random_forest_trends_',vars[i],'.png'),scale =1.5)
   df_var_imp <- df_var_imp |> 
     slice_max(Mean, n = 5) |> 
     mutate(type = vars[i])
@@ -134,36 +134,20 @@ tabla2 <- tabla2 |>
   pivot_wider(names_from = type,values_from = mean) |> 
   select(.metric,Forest,Cropland,Grassland, Savanna,Shrubland,Barren_land )
 
-head <- paste0(names(tabla2[2:7]),'\n rsq=',round(tabla2[1,2:7],2))
+head <- c('Position',paste0(names(tabla2[2:7]),'\n (rsq=',round(tabla2[1,2:7],2),')'))
 
+library(gt)
+library(broom.helpers)
 tabla |> 
   select(-Mean,-Variance) |> 
-  mutate(pos = rep(1:5,6)) |> 
+  mutate(pos = rep(1:5,6),
+         Variable = .clean_backticks(Variable)) |>
   pivot_wider(names_from = type,values_from = Variable) |> 
-  select(Forest,Cropland,Grassland, Savanna,Shrubland,Barren_land )
+  select(pos,Forest,Cropland,Grassland, Savanna,Shrubland,Barren_land) |> 
+  set_names(head) |> 
+  gt() |> 
+  cols_align(align = 'center') |> 
+  gtsave('output/figs/table_importance_trends_landcover_vs_drought.png')
+  
+  
 
-# para zcNDVI
-
-df_split <- initial_split(data_model,prop =.8)
-df_train <- training(df_split)
[…]

### 2024-03-19 article/paper_drought_Chile2023.tex
update article

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, frame hidden, enhanced, interior hidden, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, interior hidden, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
 
@@
 
-\hypertarget{software-and-packages-used}{%
-\subsection{Software and packages
-used}\label{software-and-packages-used}}
+\hypertarget{software}{%
+\subsection{Software}\label{software}}
 
@@
 \{tmap\} \citep{Tennekes2018}. For data analysis, the suite
-\{tidyverse\} \citep{Wickham2019} was used.
+\{tidyverse\} \citep{Wickham2019} was used. For the random forest
+modeling we used the \{tidymodels\}\citep{Kuhn2020} and
+\{ranger\}\citep{Wright2017} packages.
 

### 2024-03-20 article/paper_drought_Chile2023.qmd
update article

@@
 To analyze land cover change, we use the classification scheme by the IGBP (International Geosphere-Biosphere Programme) from the product MCD12Q1 collection 6.1 from MODIS.
+The MCD12Q1 has a yearly frequency from 2001 to 2022 and defines 17 classes.
 To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the product MOD13A3 collection 6.1 from MODIS [@Didan2015].
@@
 To derive the drought indices of water supply and demand, soil moisture, and vegetation (i.e., the proxy of productivity), we used the ERA5L dataset and the MODIS product, with a monthly frequency for 1981–2023 and 2000–2023, respectively.
-
 The dought indices correspond to a historical anomaly with regard to a variable (e.g., meteorological, vegetation, or soil moisture).
-To account for the anomaly, the common practice is to derive it following a statistical parametric methodology in which it is assumed that the statistical distribution of the data is known (@Heim2002).
-A wrong decision is usually the highest source of uncertainty (@Laimighofer2022).
+To account for the anomaly, the common practice is to derive it following a statistical parametric methodology in which it is assumed that the statistical distribution of the data is known [@Heim2002].
+A wrong decision is usually the highest source of uncertainty [@Laimighofer2022].
 In the case of Chile, due to its high degree of climatic variability, it is complex to choose a proper distribution without previous research.
@@
 
-For the purpose of monitoring water supply drought, we used the well-known Standardized Precipitation Index (SPI), which the World Meteorological Organization (WMO) recommended.
-The SPI solely relies on precipitation data.
-Also, it has been used worldwide for the study of drought, including in Chile (@Garreaud2017; @Zambrano2017).
-The primary cause of drought is precipitation anomalies, and temperature usually makes it worse [@Luo2017].
-Nowadays, there is an increase in attention toward using water demand separately to monitor droughts.
-(@Vicente-Serrano2020; @Noguera2022).
-Thus, to evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI), developed by @Hobbins2016 and @McEvoy2016, which is based on the AED.
-EDDI is currently used for monitoring drought in the United States (https://psl.noaa.gov/eddi/).
-In our case, we used only temperature for AED, a difference from the original formulation of EDDI, which also considered wind besides temperature.
-To consider the combined effect of water supply and demand, we selected the SPEI, which corresponds to a balance between precipitation and AED.
-@Vicente-Serrano2010 proposed the SPEI, and it has improved the SPI by incorporating temperature for drought monitoring.
-For SPEI, an auxiliary variable D = P-AED is calculated.
-Soil moisture is the main driver of vegetation productivity, particularly in semi-arid regions (@Li2022).
-Hence, for soil water drought, we used the SSI (Standardized Soil Moisture Index) [@Hao2013;@AghaKouchak2014] which is a multi-scale index similar to SPI, SPEI, and EDDI.
+For the purpose of monitoring water supply drought, we used the well-known Standardized Precipitation Index (SPI), which relies on precipitation data.
+To evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI), developed by @Hobbins2016 and @McEvoy2016, which is based on the AED.
+The United States currently monitors drought using the EDDI (https://psl.noaa.gov/eddi/) as an experimental index.
+To consider the combined effect of water supply and demand, we selected the SPEI [@Vicente-Serrano2010].
+For SPEI, an auxiliary variable $D = P-AED$ is calculated.
+Soil moisture is the main driver of vegetation productivity, particularly in semi-arid regions [@Li2022].
+Hence, for soil water drought, we used the SSI (Standardized Soil Moisture Index) [@Hao2013;@AghaKouchak2014].
 In our case, for the SSI, we used the average soil moisture from ERA5L at 1m depth.
-Finally, for the proxy of productivity, we used the zcNDVI proposed by @Zambrano2018 which will be derived from the NDVI retrieved from MOD13A1.
+Finally, for the proxy of productivity, we used the zcNDVI proposed by @Zambrano2018, which was derived from the monthly time series of NDVI retrieved from MOD13A1.
+All the indices are multi-scalar and can be used for the analysis of short- to long-term droughts.
 
@@
 The $A_{si}$ corresponds to a moving window (convolution) that sums the variable for time scales $s$ from the last month, month by month, until the first month in which it could sum for $s$ months.
-Once the variable is cumulated over time for the scale $s$.
-Thus, the empirically derived probabilities are obtained through an inverse normal approximation [@Abramowitz1968].
+An inverse normal approximation [@Abramowitz1968] obtains the empirically derived probabilities once the variable cumulates over time for the scale $s$.
 Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i$ to derive the $P(a_i)$ probabilities across a period of interest:
@@
 
-$DI$ is referring to the drought index calculated for the variable $V$.
+$DI$ is referring to the drought index calculated for the variable $V$ (i.e., SPI, SPEI, EDDI, SSI, and zcNDVI).
 The values for the constats are: $C_0 = 2.515517$, $C_1 = 0.802853$, $C_2 = 0.010328$, $d_1 = 1.432788$, $d_2 = 0.189269$, and $d3 = 0.001308$.
@@
 In the case of the proxy of vegetation productivity (zcNDVI) it was calculated for a time scale of six months at monthly frequency for 2000–2023.
-For zcNDVI, we test time scales of 1, 3, 6, and 12 months; we choose to use six months because that shows a more robust representation of vegetation productivity due to the seasonality of vegetation in Chile.
+For zcNDVI, we test time scales of 1, 3, 6, and 12 months; we choose to use six months because that shows a more stable representation of vegetation productivity.
 
@@
 To analyze the land cover change, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS.
-@Zambrano2018 and @Fuentes2021 have previously used this product for studies of drought and land cover in Chile.
-The MCD12Q1 has a yearly frequency from 2001 to 2022.
-The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
+This product has been previously used for studies of drought and land cover in Chile [@Fuentes2021;@Zambrano2018].
+From the 17 classes, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
 Thus, we have a land cover raster time series with the ten macroclasses for 2001 and 2023.
@@
 
-Climate, vegetation development, seasonality, and changes in vegetation type all have an impact on the time series of NDVI.
-In this study, we want to examine the variation in vegetation productivity across various land cover types and how water demand, water supply, and soil moisture affect it.
-In order to avoid changes due to a change in the land cover type that will wrongly impact NDVI, we developed a persistence mask for land cover for 2001–2022.
+We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023.
+After that, we calculated the trend's change in surface per land cover type and macroclass.
+We used Mann-Kendall for the significance of the trend [@Kendall1975] and Sen's slope to calculate the magnitude [@Sen1968].
+
+Later in this study, we will examine the variation in vegetation productivity across various land cover types and how water demand and supply, and soil moisture affect it.
+In order to avoid variations due to a change in the land cover type from year-to-year that will wrongly impact NDVI, we developed a persistence mask for land cover for 2001–2022.
 Thereby, we reduce an important source of variation on a regional scale.
@@
 This enabled us to identify regions where the land cover macroclasses are persistent.
-We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023.
-After that, we calculated the trend’s change in surface per type.
-We used the Sen’ slope [@Sen1968] based on Mann-Kendall [@Kendall1975].
 
@@
 
-We have 25 predictors, which are drought indices and vegetation productivity.
-We analyzed the 25 predictors per type of landcover.
-For the analysis, we selected the method of random forest (@Ho1995).
-Because it allows to find no linear relationship, it reduces overfitting and can derive the feature importance, which helps for a better understanding of the relationships.
-The importance of the variable is calculated by permuting out-of-bag (OOB) data per tree and calculating the mean standard error in the OOB.
-Then the same is done after permuting each predictor variable.
+We have 25 predictors, including drought indices and vegetation productivity.
+We analyzed the 25 predictors per type of landcover, thus running six random forest models.
 Random forest uses multiple decision trees and allows for classification and regression.
+Some advantages are that it allows to find no linear relationship, reduces overfitting, and allows to derive the variable importance.
+We used random forests for regression and trained 1000 forests.
+To obtain more reliable results, we resampled by creating ten folds, running a random forest per fold, and calculating the r-squared (rsq), root mean square error (RMSE), and variable importance.
 
-We analyzed the 25 predictors per type of landcover, thus running six models.
-We used random forests for regression and trained 1000 forests.
-For more reliable results for the important variables, we resampled by creating ten folds, running a random forest per fold, and calculating the r-squared (rsq), root mean square error (RMSE), and variable importance ten times.
+The variable importance helps for a better understanding of the relationships by finding which variable has a higher contribution to the model.
+We calculate the variable's importance by permuting out-of-bag (OOB) data per tree and computing the mean standard error in the OOB.
+After permuting each predictor variable, we repeat the process for the resting variable.
+We repeated this process ten times (per fold) to obtain the performance metrics (rsq, RMSE, and variable importance).
 
@@
 We analyzed the trend of vegetation productivity over the unchanged land cover macroclasses.
+To achieve this, we used the persistent mask of land cover macroclasses.
 This way, we tried to reduce the noise in the vegetation due to a change in land cover from year to year.
-To achieve this, we will use the persistent mask of land cover macroclasses, which are the types that remain more than 80% of the time for 2001–2022.
-We used this to evaluate the trend in zcNDVI per land cover class and macrozone.
 
-We examine the drought indices of water demand, water supply, soil moisture, and their connection with vegetation productivity to investigate two main questions: i) whether short-term or long-term time scales have a greater impact on vegetation across Chile and its specific regions; and ii) the spatial variation in the strength of the correlation between the variables and time scales.  Then, we will summarize for each land cover class and macrozone.
-Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses of vegetation: forest, cropland, grassland, savanna, and shrubland.
+We examine the drought indices of water demand, water supply, and soil moisture and their correlation with vegetation productivity.
+The objective is to determine the impact of soil moisture and water demand and supply on vegetation productivity.
+We want to address three main questions: Which of the drought variables—supply, demand, or soil moisture—most helps to explain the changes in plant productivity?
+Which of the short-term or long-term time scales of the drought variable has a bigger effect on vegetation productivity in Chile?
+And finally, how strong is the relationship between the variables and the drought index?
+Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five land cover types: forest, cropland, grassland, savanna, and shrubland.
 
-We conducted an analysis on the linear correlation between the indices SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36 months, and zcNDVI.
-The objective is to determine the impact of soil moisture and water demand and supply on vegetation productivity.
+We conducted an analysis on the linear correlation between the indices SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36 months with zcNDVI.
 We used a method similar to that used by @Meroni2017 which compared the SPI with the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation).
@@
 To begin, the Pearson coefficient of correlation is computed for each of the six time scales.
-A significant time scale is identified as the one that attains the highest correlation (p < 0.05).
+A time scale is identified as the one that attains the highest correlation (p < 0.05).
 Subsequently, the Pearson correlation coefficient corresponding to the time scales at which the value peaked was extracted.
-As a result, for each index, we generated two raster maps: 1) containing the raster with values of the time scales that reached the maximum correlation, and 2) having the value of the correlation obtained.
+As a result, for each index, we generated two raster maps: 1) containing the raster with values of the time scales and drought index that reached the maximum correlation, and 2) having the magnitude of the correlation obtained by the drought index at the time scales.
 
-## Software and packages used
+## Software
 
@@
 For data analysis, the suite {tidyverse} [@Wickham2019] was used.
+For the random forest modeling, we used the {tidymodels}[@Kuhn2020] and {ranger}[@Wright2017] packages.
 
@@
 [@fig-studyArea]d shows the macroclasses of land cover persistance (80%) during 2021–2022, respectively (Table \ref{tab-landcoverSurf}).
-Within continental Chile, barren land is the land cover class with the highest surface area (277,870 km2).
-The largest type of vegetation, with 137,085 km2, is forest.
-Grassland (74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland (3,146 km2) are the other types (Table \ref{tab-landcoverSurf}).
+Within continental Chile, barren land is the land cover class with the highest surface area (277,870 $km^2$).
+The largest type of vegetation, with 137,085 $km^2$, is forest.
+Grassland has 74,247 $km^2$, savanna 55,206 $km^2$, shrubland 25,341 $km^2$, and cropland 3,146 $km^2$ (Table \ref{tab-landcoverSurf}).
 The macrozones with major changes for 2001–2022 were "Centro," "Sur," and "Austral," with 36%, 31%, and 34% of their surface changing the type of land cover, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend}).
@@
 
-The "Norte Chico" shows an increase in barren land of 111 $km^2 yr^{-1}$  and a reduction in the class savanna of 70 $km^2 yr^{-1}$.
+From the trend analysis in Table \ref{tab-landcoverTrend}, we can indicate that the "Norte Chico" shows an increase in barren land of 111 $km^2 yr^{-1}$ and a reduction in the class savanna of 70 $km^2 yr^{-1}$.
 In the "Centro" and "Sur," there are changes with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$ ), and an increase in shrubland and grassland.
 Showing a change for more dense vegetation types.
-It appears to be a shift in the area cultivated (croplands) from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," replacing the savanna lost (Table \ref{tab-landcoverTrend}).
+The area under cultivation (croplands) appears to be shifting from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," replacing the savanna lost (Table \ref{tab-landcoverTrend}).
 
@@
 
-According to Table \ref{tab-landcoverTrendRF}, the trends in drought indices reach an r-squared between 0.32 and 0.39 for the types of forest, grassland, savanna, shrubland, and barren land.
+According to Table \ref{tab-landcoverTrendRF}, the random forest models for estimating the landcover trend from the trends in drought indices reach an r-squared between 0.32 and 0.39 for the types of forest, grassland, savanna, shrubland, and barren land.
 It is more likely that short- and medium-term increases in AED (EDDI-6 and EDDI-12) and short-term precipitation deficits (SPI-6 and SPEI-6) contributed to changes in grassland and bare land.
@@
 The increase in cumulative AED from 12 to 36 months is the strongest associated variable that contributes to changes in forests, followed by the reduction of soil moisture over six and 36 months.
-More details about the results of the variable importance can be found in the supplementary material in Section S3.
+The supplementary material in Section S3 provides further details about the variable's importance.
 
@@
 
-### Vegetation productivity
+### Trends in vegetation productivity
 
@@
 #| fig-pos: "!ht"
-#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023.
+#| fig-cap: "(a) Map of the linear trend of the index zcNDVI for 2000–2023.
 Greener colors indicate a positive trend; reder colors correspond to a negative trend and a decrease in vegetation productivity.
@@
 
-In @fig-zcNDVI_var it is showed the spatial map of trends in zcNDVI ([@fig-zcNDVI_var]a) and the temporal variation of zcNDVI within the aggregated macrozones ([@fig-zcNDVI_var]b).
-In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.02 with respect to grassland and shrubland categories.
-There is a negative trend in "Norte Chico" with -0.04 and "Centro" with -0.02 per decade.
-In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04.
-In "Centro," shrubland reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per decade.
-This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023].
+In @fig-zcNDVI_var it is showed the spatial map of trends in zcNDVI ([@fig-zcNDVI_var]a).
+In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.027 for grassland and 0.032 for shrubland.
+In the "Norte Chico," savanna has the lowest trend of -0.062, cropland -0.047, shrubland -0.042, and grassland -0.037.
+In "Centro," shrubland reaches -0.07, savanna -0.031, cropland -0.024, forest -0.017, and grassland -0.005 per decade.
+This decrease in productivity could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023].
+
+The temporal variation within the macrozones is showed in [@fig-zcNDVI_var]b).
+There is a negative trend in "Norte Chico" with -0.035 and "Centro" with -0.02 per decade.
 Vegetation reached its lowest values since the year 2019, with an extreme condition in early 2020 and 2022 in the "Norte Chico" and "Centro".
-The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var).
-Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the savanna and shurbland appears to be the region most affected by a negative trend in vegetation productivity.
+The "Sur" and "Austral" show a positive trend of around 0.012 and 0.016, respectively per decade (@fig-zcNDVI_var).
+Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the savanna and shurbland appears to be the region most affected by a negative trend in vegetation productivity across Chile.
 

### 2024-03-20 article/paper_drought_Chile2023.tex
update article

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, interior hidden, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, breakable, sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
 
@@
 IGBP (International Geosphere-Biosphere Programme) from the product
-MCD12Q1 collection 6.1 from MODIS. To derive a proxy for vegetation
-productivity, we used the Normalized Difference Vegetation Index (NDVI)
-from the product MOD13A3 collection 6.1 from MODIS \citep{Didan2015}.
-MOD13A3 provides vegetation indices at 1km of spatial resolution and
-monthly frequency. The NASA EOSDIS Land Processes Distributed Active
-Archive Center (LP DAAC), USGS Earth Resources Observation and Science
-(EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and
-MCD12Q1 from the online Data Pool, accessible at
-https://lpdaa.usgs.gov/tools/data-pool/.
+MCD12Q1 collection 6.1 from MODIS. The MCD12Q1 has a yearly frequency
+from 2001 to 2022 and defines 17 classes. To derive a proxy for
+vegetation productivity, we used the Normalized Difference Vegetation
+Index (NDVI) from the product MOD13A3 collection 6.1 from MODIS
+\citep{Didan2015}. MOD13A3 provides vegetation indices at 1km of spatial
+resolution and monthly frequency. The NASA EOSDIS Land Processes
+Distributed Active Archive Center (LP DAAC), USGS Earth Resources
+Observation and Science (EROS) Center, Sioux Falls, South Dakota,
+provided the MOD13A3 and MCD12Q1 from the online Data Pool, accessible
+at https://lpdaa.usgs.gov/tools/data-pool/.
 
@@
 dataset and the MODIS product, with a monthly frequency for 1981--2023
-and 2000--2023, respectively.
-
-The dought indices correspond to a historical anomaly with regard to a
-variable (e.g., meteorological, vegetation, or soil moisture). To
-account for the anomaly, the common practice is to derive it following a
-statistical parametric methodology in which it is assumed that the
-statistical distribution of the data is known (\citet{Heim2002}). A
-wrong decision is usually the highest source of uncertainty
-(\citet{Laimighofer2022}). In the case of Chile, due to its high degree
-of climatic variability, it is complex to choose a proper distribution
-without previous research. Here, we follow a non-parametric methodology
-for the calculation of the drought indices, in a similar manner as the
-framework proposed by \citet{Farahmand2015};
+and 2000--2023, respectively. The dought indices correspond to a
+historical anomaly with regard to a variable (e.g., meteorological,
+vegetation, or soil moisture). To account for the anomaly, the common
+practice is to derive it following a statistical parametric methodology
+in which it is assumed that the statistical distribution of the data is
+known \citep{Heim2002}. A wrong decision is usually the highest source
+of uncertainty \citep{Laimighofer2022}. In the case of Chile, due to its
+high degree of climatic variability, it is complex to choose a proper
+distribution without previous research. Here, we follow a non-parametric
+methodology for the calculation of the drought indices, in a similar
+manner as the framework proposed by \citet{Farahmand2015};
 \citet{Hobbins2016};\citet{McEvoy2016}.
@@
 For the purpose of monitoring water supply drought, we used the
-well-known Standardized Precipitation Index (SPI), which the World
-Meteorological Organization (WMO) recommended. The SPI solely relies on
-precipitation data. Also, it has been used worldwide for the study of
-drought, including in Chile (\citet{Garreaud2017};
-\citet{Zambrano2017}). The primary cause of drought is precipitation
-anomalies, and temperature usually makes it worse \citep{Luo2017}.
-Nowadays, there is an increase in attention toward using water demand
-separately to monitor droughts. (\citet{Vicente-Serrano2020};
-\citet{Noguera2022}). Thus, to evaluate water demand, we chose the
-Evaporative Demand Drought Index (EDDI), developed by
-\citet{Hobbins2016} and \citet{McEvoy2016}, which is based on the AED.
-EDDI is currently used for monitoring drought in the United States
-(https://psl.noaa.gov/eddi/). In our case, we used only temperature for
-AED, a difference from the original formulation of EDDI, which also
-considered wind besides temperature. To consider the combined effect of
-water supply and demand, we selected the SPEI, which corresponds to a
-balance between precipitation and AED. \citet{Vicente-Serrano2010}
-proposed the SPEI, and it has improved the SPI by incorporating
-temperature for drought monitoring. For SPEI, an auxiliary variable D =
-P-AED is calculated. Soil moisture is the main driver of vegetation
-productivity, particularly in semi-arid regions (\citet{Li2022}). Hence,
-for soil water drought, we used the SSI (Standardized Soil Moisture
-Index) \citep{Hao2013, AghaKouchak2014} which is a multi-scale index
-similar to SPI, SPEI, and EDDI. In our case, for the SSI, we used the
-average soil moisture from ERA5L at 1m depth. Finally, for the proxy of
-productivity, we used the zcNDVI proposed by \citet{Zambrano2018} which
-will be derived from the NDVI retrieved from MOD13A1.
+well-known Standardized Precipitation Index (SPI), which relies on
+precipitation data. To evaluate water demand, we chose the Evaporative
+Demand Drought Index (EDDI), developed by \citet{Hobbins2016} and
+\citet{McEvoy2016}, which is based on the AED. The United States
+currently monitors drought using the EDDI (https://psl.noaa.gov/eddi/)
+as an experimental index. To consider the combined effect of water
+supply and demand, we selected the SPEI \citep{Vicente-Serrano2010}. For
+SPEI, an auxiliary variable \(D = P-AED\) is calculated. Soil moisture
+is the main driver of vegetation productivity, particularly in semi-arid
+regions \citep{Li2022}. Hence, for soil water drought, we used the SSI
+(Standardized Soil Moisture Index) \citep{Hao2013, AghaKouchak2014}. In
+our case, for the SSI, we used the average soil moisture from ERA5L at
+1m depth. Finally, for the proxy of productivity, we used the zcNDVI
+proposed by \citet{Zambrano2018}, which was derived from the monthly
+time series of NDVI retrieved from MOD13A1. All the indices are
+multi-scalar and can be used for the analysis of short- to long-term
+droughts.
 
@@
 the variable for time scales \(s\) from the last month, month by month,
-until the first month in which it could sum for \(s\) months. Once the
-variable is cumulated over time for the scale \(s\). Thus, the
-empirically derived probabilities are obtained through an inverse normal
-approximation \citep{Abramowitz1968}. Then, we used the empirical Tukey
-plotting position \citep{Wilks2011} over \(A_i\) to derive the
-\(P(a_i)\) probabilities across a period of interest:
+until the first month in which it could sum for \(s\) months. An inverse
+normal approximation \citep{Abramowitz1968} obtains the empirically
+derived probabilities once the variable cumulates over time for the
+scale \(s\). Then, we used the empirical Tukey plotting position
+\citep{Wilks2011} over \(A_i\) to derive the \(P(a_i)\) probabilities
+across a period of interest:
 
@@
 \(DI\) is referring to the drought index calculated for the variable
-\(V\). The values for the constats are: \(C_0 = 2.515517\),
-\(C_1 = 0.802853\), \(C_2 = 0.010328\), \(d_1 = 1.432788\),
-\(d_2 = 0.189269\), and \(d3 = 0.001308\). For \(P(A) \leq 0.5\),
+\(V\) (i.e., SPI, SPEI, EDDI, SSI, and zcNDVI). The values for the
+constats are: \(C_0 = 2.515517\), \(C_1 = 0.802853\),
+\(C_2 = 0.010328\), \(d_1 = 1.432788\), \(d_2 = 0.189269\), and
+\(d3 = 0.001308\). For \(P(A) \leq 0.5\),
 W=\(\sqrt{-2\cdot ln(P(A_i))}\) , and for \(P(A_i) > 0.5\), replace
@@
 time scales of 1, 3, 6, and 12 months; we choose to use six months
-because that shows a more robust representation of vegetation
-productivity due to the seasonality of vegetation in Chile.
+because that shows a more stable representation of vegetation
+productivity.
 
@@
 To analyze the land cover change, we use the IGBP scheme from the
-MCD12Q1 collection 6.1 from MODIS. \citet{Zambrano2018} and
-\citet{Fuentes2021} have previously used this product for studies of
-drought and land cover in Chile. The MCD12Q1 has a yearly frequency from
-2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into
-ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands,
-8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to
-croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to
+MCD12Q1 collection 6.1 from MODIS. This product has been previously used
+for studies of drought and land cover in Chile
+\citep{Fuentes2021, Zambrano2018}. From the 17 classes, we regrouped
+into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to
+shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14
+to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to
 water bodies. Thus, we have a land cover raster time series with the ten
@@
 
-Climate, vegetation development, seasonality, and changes in vegetation
-type all have an impact on the time series of NDVI. In this study, we
-want to examine the variation in vegetation productivity across various
-land cover types and how water demand, water supply, and soil moisture
-affect it. In order to avoid changes due to a change in the land cover
-type that will wrongly impact NDVI, we developed a persistence mask for
-land cover for 2001--2022. Thereby, we reduce an important source of
-variation on a regional scale. Therefore, we generated a raster mask for
-IGBP MODIS per pixel using macroclasses that remained unchanged for at
-least 80\% of the years (2001--2022). This enabled us to identify
-regions where the land cover macroclasses are persistent. We calculated
-the surface occupied per land cover class into the five macrozones
-(``Norte Grande'' to ``Austral'') per year for 2001--2023. After that,
-we calculated the trend's change in surface per type. We used the Sen'
-slope \citep{Sen1968} based on Mann-Kendall \citep{Kendall1975}.
+We calculated the surface occupied per land cover class into the five
+macrozones (``Norte Grande'' to ``Austral'') per year for 2001--2023.
+After that, we calculated the trend's change in surface per land cover
+type and macroclass. We used Mann-Kendall for the significance of the
+trend \citep{Kendall1975} and Sen's slope to calculate the magnitude
+\citep{Sen1968}.
+
+Later in this study, we will examine the variation in vegetation
+productivity across various land cover types and how water demand and
+supply, and soil moisture affect it. In order to avoid variations due to
+a change in the land cover type from year-to-year that will wrongly
+impact NDVI, we developed a persistence mask for land cover for
+2001--2022. Thereby, we reduce an important source of variation on a
+regional scale. Therefore, we generated a raster mask for IGBP MODIS per
+pixel using macroclasses that remained unchanged for at least 80\% of
+the years (2001--2022). This enabled us to identify regions where the
+land cover macroclasses are persistent.
 
@@
 sub-basins within continental Chile. We use 469 basins, which have a
-surface area between 0.0746 and 24,000 (\(km^2\)), and a median area of
+surface area between 0.0746 and ~24,000 (\(km^2\)), and a median area of
 1,249 (\(km^2\)). For each basin, we calculate the relative trend per
@@
 
-We have 25 predictors, which are drought indices and vegetation
-productivity. We analyzed the 25 predictors per type of landcover. For
-the analysis, we selected the method of random forest (\citet{Ho1995}).
-Because it allows to find no linear relationship, it reduces overfitting
-and can derive the feature importance, which helps for a better
-understanding of the relationships. The importance of the variable is
-calculated by permuting out-of-bag (OOB) data per tree and calculating
-the mean standard error in the OOB. Then the same is done after
-permuting each predictor variable. Random forest uses multiple decision
-trees and allows for classification and regression.
-
-We analyzed the 25 predictors per type of landcover, thus running six
-models. We used random forests for regression and trained 1000 forests.
-For more reliable results for the important variables, we resampled by
-creating ten folds, running a random forest per fold, and calculating
-the r-squared (rsq), root mean square error (RMSE), and variable
-importance ten times.
+We have 25 predictors, including drought indices and vegetation
+productivity. We analyzed the 25 predictors per type of landcover, thus
+running six random forest models. Random forest uses multiple decision
+trees and allows for classification and regression. Some advantages are
+that it allows to find no linear relationship, reduces overfitting, and
+allows to derive the variable importance. We used random forests for
+regression and trained 1000 forests. To obtain more reliable results, we
+resampled by creating ten folds, running a random forest per fold, and
+calculating the r-squared (rsq), root mean square error (RMSE), and
+variable importance.
+
+The variable importance helps for a better understanding of the
+relationships by finding which variable has a higher contribution to the
+model. We calculate the variable's importance by permuting out-of-bag
+(OOB) data per tree and computing the mean standard error in the OOB.
+After permuting each predictor variable, we repeat the process for the
+resting variable. We repeated this process ten times (per fold) to
+obtain the performance metrics (rsq, RMSE, and variable importance).
 
@@
 We analyzed the trend of vegetation productivity over the unchanged land
+cover macroclasses. To achieve this, we used the persistent mask of land
 cover macroclasses. This way, we tried to reduce the noise in the
-vegetation due to a change in land cover from year to year. To achieve
-this, we will use the persistent mask of land cover macroclasses, which
-are the types that remain more than 80\% of the time for 2001--2022. We
-used this to evaluate the trend in zcNDVI per land cover class and
-macrozone.
-
-We examine the drought indices of water demand, water supply, soil
-moisture, and their connection with vegetation productivity to
-investigate two main questions: i) whether short-term or long-term time
-scales have a greater impact on vegetation across Chile and its specific
-regions; and ii) the spatial variation in the strength of the
-correlation between the variables and time scales.~ Then, we will
-summarize for each land cover class and macrozone. Thus, we will be able
-to advance in understanding how climate is affecting vegetation,
-considering the impact on the five macroclasses of vegetation: forest,
-cropland, grassland, savanna, and shrubland.
+vegetation due to a change in land cover from year to year.
+
+We examine the drought indices of water demand, water supply, and soil
+moisture and their correlation with vegetation productivity. The
+objective is to determine the impact of soil moisture and water demand
+and supply on vegetation productivity. We want to address three main
+questions: Which of the drought variables---supply, demand, or soil
+moisture---most helps to explain the changes in plant productivity?
+Which of the short-term or long-term time scales of the drought variable
+has a bigger effect on vegetation productivity in Chile? And finally,
+how strong is the relationship between the variables and the drought
+index? Thus, we will be able to advance in understanding how climate is
+affecting vegetation, considering the impact on the five land cover
+types: forest, cropland, grassland, savanna, and shrubland.
 
@@
 SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36
-months, and zcNDVI. The objective is to determine the impact of soil
-moisture and water demand and supply on vegetation productivity. We used
-a method similar to that used by \citet{Meroni2017} which compared the
-SPI with the cumulative FAPAR (Fraction of Absorbed Photosynthetically
-Active Radiation). A pixel-to-pixel linear correlation analysis was
-performed for each index within the persistent mask of land cover
-macroclasses. To begin, the Pearson coefficient of correlation is
-computed for each of the six time scales. A significant time scale is
-identified as the one that attains the highest correlation (p
-\textless{} 0.05). Subsequently, the Pearson correlation coefficient
-corresponding to the time scales at which the value peaked was
-extracted. As a result, for each index, we generated two raster maps: 1)
-containing the raster with values of the time scales that reached the
-maximum correlation, and 2) having the value of the correlation
-obtained.
+months with zcNDVI. We used a method similar to that used by
+\citet{Meroni2017} which compared the SPI with the cumulative FAPAR
+(Fraction of Absorbed Photosynthetically Active Radiation). A
+pixel-to-pixel linear correlation analysis was performed for each index
+within the persistent mask of land cover macroclasses. To begin, the
+Pearson coefficient of correlation is computed for each of the six time
+scales. A time scale is identified as the one that attains the highest
+correlation (p \textless{} 0.05). Subsequently, the Pearson correlation
+coefficient corresponding to the time scales at which the value peaked
+was extracted. As a result, for each index, we generated two raster
+maps: 1) containing the raster with values of the time scales and
+drought index that reached the maximum correlation, and 2) having the
+magnitude of the correlation obtained by the drought index at the time
+scales.
 
@@
 \{tidyverse\} \citep{Wickham2019} was used. For the random forest
-modeling we used the \{tidymodels\}\citep{Kuhn2020} and
+modeling, we used the \{tidymodels\}\citep{Kuhn2020} and
 \{ranger\}\citep{Wright2017} packages.
@@
 \ref{tab-landcoverSurf}). Within continental Chile, barren land is the
-land cover class with the highest surface area (277,870 km2). The
-largest type of vegetation, with 137,085 km2, is forest. Grassland
-(74,247 km2), savanna (55,206 km2), shrubland (25,341 km2), and cropland
-(3,146 km2) are the other types (Table \ref{tab-landcoverSurf}). The
+land cover class with the highest surface area (277,870 \(km^2\)). The
+largest type of vegetation, with 137,085 \(km^2\), is forest. Grassland
+has 74,247 \(km^2\), savanna 55,206 \(km^2\), shrubland 25,341 \(km^2\),
+and cropland 3,146 \(km^2\) (Table \ref{tab-landcoverSurf}). The
 macrozones with major changes for 2001--2022 were ``Centro,'' ``Sur,''
@@
 
-The ``Norte Chico'' shows an increase in barren land of 111
-\(km^2 yr^{-1}\) and a reduction in the class savanna of 70
+From the trend analysis in Table \ref{tab-landcoverTrend}, we can
+indicate that the ``Norte Chico'' shows an increase in barren land of
+111 \(km^2 yr^{-1}\) and a reduction in the class savanna of 70
 \(km^2 yr^{-1}\). In the ``Centro'' and ``Sur,'' there are changes with
@@
 increase in shrubland and grassland. Showing a change for more dense
-vegetation types. It appears to be a shift in the area cultivated
-(croplands) from the ``Centro'' to the ``Sur.'' Also, there is a high
+vegetation types. The area under cultivation (croplands) appears to be
+shifting from the ``Centro'' to the ``Sur.'' Also, there is a high
 increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' replacing the
@@
 
-According to Table \ref{tab-landcoverTrendRF}, the trends in drought
-indices reach an r-squared between 0.32 and 0.39 for the types of
-forest, grassland, savanna, shrubland, and barren land. It is more
-likely that short- and medium-term increases in AED (EDDI-6 and EDDI-12)
-and short-term precipitation deficits (SPI-6 and SPEI-6) contributed to
+According to Table \ref{tab-landcoverTrendRF}, the random forest models
+for estimating the landcover trend from the trends in drought indices
+reach an r-squared between 0.32 and 0.39 for the types of forest,
+grassland, savanna, shrubland, and barren land. It is more likely that
+short- and medium-term increases in AED (EDDI-6 and EDDI-12) and
+short-term precipitation deficits (SPI-6 and SPEI-6) contributed to
 changes in grassland and bare land. The short-term increase of AED
@@
 strongest associated variable that contributes to changes in forests,
-followed by the reduction of soil moisture over six and 36 months. More
-details about the results of the variable importance can be found in the
-supplementary material in Section S3.
+followed by the reduction of soil moisture over six and 36 months. The
+supplementary material in Section S3 provides further details about the
+variable's importance.
 
@@
 
-\hypertarget{vegetation-productivity}{%
-\subsubsection{Vegetation productivity}\label{vegetation-productivity}}
+\hypertarget{trends-in-vegetation-productivity}{%
+\subsubsection{Trends in vegetation
+productivity}\label{trends-in-vegetation-productivity}}
 
@@
 \caption{\label{fig-zcNDVI_var}(a) Map of the linear trend of the index
-zcNDVI-6 for 2001--2023. Greener colors indicate a positive trend; reder
+zcNDVI for 2000--2023. Greener colors indicate a positive trend; reder
 colors correspond to a negative trend and a decrease in vegetation
@@
 In Figure~\ref{fig-zcNDVI_var} it is showed the spatial map of trends in
-zcNDVI (Figure~\ref{fig-zcNDVI_var}a) and the temporal variation of
-zcNDVI within the aggregated macrozones (Figure~\ref{fig-zcNDVI_var}b).
-In ``Norte Grande,'' vegetation productivity, as per the z-index,
-exhibits a yearly increase of 0.02 with respect to grassland and
-shrubland categories. There is a negative trend in ``Norte Chico'' with
--0.04 and ``Centro'' with -0.02 per decade. In the ``Norte Chico,''
-savanna (-0.05) has the lowest trend, and the rest of the types are
-around -0.04. In ``Centro,'' shrubland reaches -0.06, grassland -0.05,
-and croplands and savanna -0.01 per decade. This could be associated
-either with a reduction in vegetation surface, a decrease in biomass, or
-browning \citep{Miranda2023}. Vegetation reached its lowest values since
-the year 2019, with an extreme condition in early 2020 and 2022 in the
-``Norte Chico'' and ``Centro''. The ``Sur'' and ``Austral'' show a
-positive trend of around 0.016 per decade (Figure~\ref{fig-zcNDVI_var}).
-Despite the croplands suffering from drought just as badly as the native
-vegetation in ``Norte Chico,'' the savanna and shurbland appears to be
-the region most affected by a negative trend in vegetation productivity.
+zcNDVI (Figure~\ref{fig-zcNDVI_var}a). In ``Norte Grande,'' vegetation
+productivity, as per the z-index, exhibits a yearly increase of 0.027
+for grassland and 0.032 for shrubland. In the ``Norte Chico,'' savanna
+has the lowest trend of -0.062, cropland -0.047, shrubland -0.042, and
+grassland -0.037. In ``Centro,'' shrubland reaches -0.07, savanna
+-0.031, cropland -0.024, forest -0.017, and grassland -0.005 per decade.
+This decrease in productivity could be associated either with a
+reduction in vegetation surface, a decrease in biomass, or browning
+\citep{Miranda2023}.
+
+The temporal variation within the macrozones is showed in
+Figure~\ref{fig-zcNDVI_var}b). There is a negative trend in ``Norte
+Chico'' with -0.035 and ``Centro'' with -0.02 per decade. Vegetation
+reached its lowest values since the year 2019, with an extreme condition
+in early 2020 and 2022 in the ``Norte Chico'' and ``Centro''. The
+``Sur'' and ``Austral'' show a positive trend of around 0.012 and 0.016,
+respectively per decade (Figure~\ref{fig-zcNDVI_var}). Despite the
+croplands suffering from drought just as badly as the native vegetation
+in ``Norte Chico,'' the savanna and shurbland appears to be the region
+most affected by a negative trend in vegetation productivity across
+Chile.
 

### 2024-03-20 article/supplementary.qmd
update article

@@
 author: "Francisco Zambrano"
+date: "`r Sys.Date()`"
 format: 
@@
 :::
+
+# Trend of vegetation productivity
+
+```{r}
+#| echo: false
+#| out-width: 75%
+#| fig-cap: "Heatmap of trends in zcNDVI for 2000 to 2023 per macrozone and landcover macroclass."
+knitr::include_graphics('../output/figs/heatmap_trends_zcNDVI_macro_landcover.png')
+```

### 2024-03-20 article/paper_drought_Chile2023.qmd
update article

@@
   - land cover change
+  - vegetation productivity
   - satellite
@@
 @Hao2013 and @AghaKouchak2014 proposed the Standardized Soil Moisture Index (SSI), which has a similar formulation as the SPI, SPEI, and EDDI.
-Thus, there are plenty of drought indices that allow for a comprehensive assessment of drought from short- to long-term scales and consider variables from the earth's water balance on their own (e.g., precipitation, AED, soil moisture).
+Thus, there are plenty of drought indices that allow for a comprehensive assessment of drought on short- to long-term scales and that allow for the use of single variables from the earth's water balance (e.g., precipitation, AED, soil moisture).
+The variation in climate variables impacts vegetation development, and unfavorable conditions such as low precipitation and high temperatures usually generate a decrease in vegetation productivity.
+To monitor the response of vegetation, the common practice is to use satellite data.
+The Normalized Difference Vegetation Index (NDVI) has been widely used as a proxy for biomass production [@Camps-Valls2021;@Paruelo2016;@Helman2014].
+For Chile's cultivated land, @Zambrano2018 introduced the zcNDVI for assessing seasonal biomass production in response to drought.
 Using this information, we can advance our understanding of the impact of drought on ecosystems.
@@
 Therefore, it is crucial to evaluate temporal scales that consider the cumulative impact over a period of several years.
-There is limited understanding of the correlation between drought and the environment in Chile.
-Therefore, it is crucial to have a deeper understanding of how meteorological and soil moisture droughts impact the dynamics of the environment in order to make informed decisions on how to adapt.
+The association between drought and the environment in Chile is not well comprehended.
+Hence, it is imperative to acquire a more profound comprehension of the manner in which climatic and soil moisture droughts influence environmental dynamics, in order to make well-informed decisions on adaptation strategies.
 
@@
 
-## Trend of short- to long-term drought
+## Short- to long-term drought
 
@@
 This way, we tried to reduce the noise in the vegetation due to a change in land cover from year to year.
+We used the zcNDVI as a proxy of vegetation productivity.
+In Chile's cultivated land, @Zambrano2018 introduced the zcNDVI for assessing seasonal biomass production in relation to climate.
 
@@
 We want to address three main questions: Which of the drought variables—supply, demand, or soil moisture—most helps to explain the changes in plant productivity?
-Which of the short-term or long-term time scales of the drought variable has a bigger effect on vegetation productivity in Chile?
+How do the short- to long-term time scales of the drought variable affect vegetation productivity in Chile?
 And finally, how strong is the relationship between the variables and the drought index?
@@
 We conducted an analysis on the linear correlation between the indices SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36 months with zcNDVI.
-We used a method similar to that used by @Meroni2017 which compared the SPI with the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation).
-A pixel-to-pixel linear correlation analysis was performed for each index within the persistent mask of land cover macroclasses.
-To begin, the Pearson coefficient of correlation is computed for each of the six time scales.
+We used a method similar to that used by @Meroni2017 which compared the SPI time-scales with the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation).
+We performed a pixel-to-pixel linear correlation analysis for each index within the persistent mask of land cover macroclasses.
+We first compute the Pearson coefficient of correlation for each of the six time scales.
 A time scale is identified as the one that attains the highest correlation (p < 0.05).
-Subsequently, the Pearson correlation coefficient corresponding to the time scales at which the value peaked was extracted.
+We then extracted the Pearson correlation coefficient corresponding to the time scales where the value peaked.
 As a result, for each index, we generated two raster maps: 1) containing the raster with values of the time scales and drought index that reached the maximum correlation, and 2) having the magnitude of the correlation obtained by the drought index at the time scales.
@@
 For mapping, we use {tmap} [@Tennekes2018].
-For data analysis, the suite {tidyverse} [@Wickham2019] was used.
+For data analysis and visualization, the suite {tidyverse} [@Wickham2019] was used.
 For the random forest modeling, we used the {tidymodels}[@Kuhn2020] and {ranger}[@Wright2017] packages.
@@
 
-In @fig-trendDIMacro, the averaged aggregation per macrozone, drought index, and time scale are shown.
-The macrozones that have the lowest trend are "Norte Chico" and "Centro," where the SPI, SPEI, and SSI show that it decreases at longer time scales.
+The @fig-trendDIMacro displays the averaged aggregation per macrozone, the drought index, and the timescale.
+The macrozones that reached the lowest trend for SPI, SPEI, and SSI are "Norte Chico" and "Centro," where the indices also decrease at longer time scales.
 Potentially explained due to the prolonged reduction in precipitation that has affected the hydrological system in Chile.
@@
 
-### Interaction of land cover and drought
+## Interaction of land cover and drought
 
-## Land cover change
+### Land cover change
 
@@
 Showing a change for more dense vegetation types.
-The area under cultivation (croplands) appears to be shifting from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," replacing the savanna lost (Table \ref{tab-landcoverTrend}).
+The area under cultivation (croplands) appears to be shifting from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," seemingly replacing the savanna lost (Table \ref{tab-landcoverTrend}).
 
 ### Relationship between drought indices and land cover change
-
-According to Table \ref{tab-landcoverTrendRF}, the random forest models for estimating the landcover trend from the trends in drought indices reach an r-squared between 0.32 and 0.39 for the types of forest, grassland, savanna, shrubland, and barren land.
-It is more likely that short- and medium-term increases in AED (EDDI-6 and EDDI-12) and short-term precipitation deficits (SPI-6 and SPEI-6) contributed to changes in grassland and bare land.
-The short-term increase of AED (EDDI-3 and EDDI-6) and the longer duration of the precipitation deficit (SPI-24, SPI-36, and SPEI-36) most likely contribute to the changes in shrubland.
-The changes in savanna are associated with a short- and long-term increase in AED and a three-year precipitation deficit (SPI-36).
-The increase in cumulative AED from 12 to 36 months is the strongest associated variable that contributes to changes in forests, followed by the reduction of soil moisture over six and 36 months.
-The supplementary material in Section S3 provides further details about the variable's importance.
 
@@
 ```
+
+According to Table \ref{tab-landcoverTrendRF}, the random forest models for estimating the landcover trend from the trends in drought indices reach an r-squared between 0.32 and 0.39 for the types of forest, grassland, savanna, shrubland, and barren land.
+It is more likely that short- and medium-term increases in AED (EDDI-6 and EDDI-12) and short-term precipitation deficits (SPI-6 and SPEI-6) contributed to changes in grassland and bare land.
+The short-term increase of AED (EDDI-3 and EDDI-6) and the longer duration of the precipitation deficit (SPI-24, SPI-36, and SPEI-36) most likely contribute to the changes in shrubland.
+The changes in savanna are associated with a short- and long-term increase in AED and a three-year precipitation deficit (SPI-36).
+The increase in cumulative AED from 12 to 36 months is the strongest associated variable that contributes to changes in forests, followed by the reduction of soil moisture over six and 36 months.
+The supplementary material in Section S3 provides further details about the variable's importance.
 
@@
 
+The temporal variation within the macrozones is shown in [@fig-zcNDVI_var]b).
+There is a negative trend in "Norte Chico" with -0.035 and "Centro" with -0.02 per decade.
+Vegetation reached its lowest values for 2019-2022, with an extreme condition in early 2020 and 2022 in the "Norte Chico" and "Centro".
+The "Sur" and "Austral" show a positive trend of around 0.012 and 0.016, respectively, per decade (@fig-zcNDVI_var).
+
 In @fig-zcNDVI_var it is showed the spatial map of trends in zcNDVI ([@fig-zcNDVI_var]a).
@@
 In "Centro," shrubland reaches -0.07, savanna -0.031, cropland -0.024, forest -0.017, and grassland -0.005 per decade.
-This decrease in productivity could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023].
+This decrease in productivity could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning.
 
-The temporal variation within the macrozones is showed in [@fig-zcNDVI_var]b).
-There is a negative trend in "Norte Chico" with -0.035 and "Centro" with -0.02 per decade.
-Vegetation reached its lowest values since the year 2019, with an extreme condition in early 2020 and 2022 in the "Norte Chico" and "Centro".
-The "Sur" and "Austral" show a positive trend of around 0.012 and 0.016, respectively per decade (@fig-zcNDVI_var).
-Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the savanna and shurbland appears to be the region most affected by a negative trend in vegetation productivity across Chile.
+<!-- Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the savanna and shurbland appears to be the region most affected by a negative trend in vegetation productivity across Chile.
+-->
 
@@
 
-@fig-corTimeScale is a map that shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between different drought indicators and plant productivity over time.
+@fig-corTimeScale shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between different drought indicators and zcNDVI over time scales (1, 3, 6, 12, 24, and 36 months).
 The spatial variation of time scales reached per index is mostly for time scales above 12 months.
 In the case of SSI, the predominant scales are 6 and 12 months.
-For all indices, to the north, the time scales are higher and diminish toward the south until the south part of "Austral" increases.
-In @fig-corPerson, the map of Pearson correlation values is shown.
+For all indices, to the north, the time scales are higher and diminish toward the south until the south part of "Austral," where they increase.
+In @fig-corPerson, the map of Pearson correlation values (r) is shown.
 The EDDI reached correlations above 0.5 between "Norte Chico" and "Sur." The correlation changes from negative to positive toward the Andes Mountains and to the sea, just as in the northern part of "Austral." The SPI and SPEI have similar results, with the higher values in "Norte Chico" and "Centro" being higher than 0.6.
-Following a similar spatial pattern as EDDI.
+Following a similar spatial pattern as EDDI but with an opposite sign.
 The SSI showed to be the index that has a major spatial extension with a higher correlation.
@@
 
-## Drought trend and attribution to land cover
+## Drought trend and attribution to land cover change
 
-@Vicente-Serrano2022, in a study at the global scale of drought trends, indicate that there have not been significant trends in meteorological drought since 1950.
+@Vicente-Serrano2022, in a study at the global scale of drought trends, indicates that there have not been significant trends in meteorological drought since 1950.
 Also, state that the increase in hidrological trend in some parts of the globe (northeast Brazil and the Mediterranean region) is related to changes in land cover and specifically to the rapidly increasing irrigated area, which consequently increases water extraction.
-@Kogan2020 analyzed the agricultural drought impact globally and in the main grain producer countries, finding that _"since 1980, the Earth warming has not changed the drought area or intensity."_ In our study, we considered the variation in vegetation productivity in Chile for areas without changes in land cover, to avoid misleading conclusions that could be related to the increase in water demand due to land cover change.
+@Kogan2020 analyzed the agricultural drought impact globally and in the main grain producer countries, finding that "since 1980, the Earth warming has not changed the drought area or intensity." In our study, we took into account the variation in vegetation productivity in Chile, specifically in areas without any changes in land cover, to prevent any misleading conclusions about the increase in water demand due to land cover change.
 Our results show a contrasting perspective.
-There has been a significant trend in the decline of vegetation productivity (zcNDVI) since 2000 for "Norte Chico" and "Centro," which has been extreme between 2020 and 2022, seemsly due to an intense hydrological drought due to the persistance of the mega drought [@Garreaud2017].
-However, a rise in irrigated land doesn't seem to have an impact on this hydrological drought.
-Despite using the persistance mask for vegetation's trend analysis, cropland, which is the most water-demand type, showed a decrease trend in "Norte Chico" and "Centro." Also, there was an increase in barren land for both types.
+The SPI, SPEI, and SSI (water supply) showed a decrease in trends, except for the southern part, and an increase in EDDI (water demand).
+The trend, positive or negative, was stronger as the time scales increased.
+Trends in the long term (e.g., 36 months) are evidence of how human-induced climate change is affecting Chile, which seems to be due to an intense hydrological drought resulting from the persistence of the precipitation deficit.
+We found that there has been a significant trend in the decline of vegetation productivity (zcNDVI) since 2000 for the north-central part of the country, which has been extreme between 2020 and 2022 and has impacted natural and cultivated land.
+Additionally, we demonstrated that the drought, primarily due to an increase in AED, accounts for about 30% of the changes in land cover types (excluding croplands).
 These changes are associated with a decrease in water demand from vegetation.
-Nonetheless, we used the persistant land cover to ensure that the pixel has the same class; in the case of croplands, it could happen that some areas had changed crops for others with higher water consumption and consequently increase water demand.
-But this effect should be minor compared to the results from land cover change at this scale of analysis.
+Moreover, the most water-demanding type, cropland, showed a decrease in the north-central region, while barren land showed an increase.
+The north-central part of the country primarily experienced these changes due to a higher increase in AED.
+Thus, we have evidence of a significant decline in water supply and an increase in AED for the north-central part of Chile, which show to be the most relevant variables for drought conditions.
+Some questions arise regarding what is occurring with the cultivated land.
+We used the unchanged land cover to ensure that an increase in surface area is not considered in the trend analysis.
+For croplands, it could happen that some areas have changed the types of crops for others with higher water demand, which consequently increases water demand.
+However, this effect should be minor compared to the decrease in water supply and increase in water demand at this scale of analysis.
 
-On the other hand, for "Norte Chico" and "Centro," our results show a decrease in trends of water supply (SPI and SSI), which are higher at larger time scales, which is evidence of the hydrological drought.
-We say that what happened in central Chile goes against what has been found on a global scale [@Vicente-Serrano2022;@Kogan2020].
-This shows that the main cause of the hydrological drought in Chile was a steady drop in water supply made worse by an increase in AED, but it seems that in zones most affected by drought, the main cause is not an increase in water demand by vegetation like irrigated crops.
-Finally, north-central Chile has experienced a decline in vegetation productivity across all macroclasses, which is primarily attributable to variations in water supply and soil moisture.
+This shows that the main cause of the hydrological drought in Chile was a steady drop in water supply made worse by an increase in AED, but it seems that in zones most affected by drought, the main cause is not an increase in water demand due to an intensification of cultivated land (e.g., an increase in irrigated crops).
+North-central Chile has experienced a decline in vegetation productivity across land cover types, which is primarily attributable to variations in water supply and soil moisture.
 An increase in water demand, such as an increase in the surface area of irrigated crops, could strengthen this trend.
 
-## Land cover types and their impact by drought
+## Impact of short- to long-term droughts on land cover types
 
-We discovered that croplands, savannas, and shrubland are the most susceptible to climatic changes and are most affected by the 12-month soil moisture deficit.
+We found that the 12-month soil moisture deficit most affects the productivity of vegetation in all land cover types along Chile.
 In a study in the Yangtze River Basin in China, @Jiang2020 analyzed the impact of drought on vegetation using the SPEI and the Enahanced Vegetation Index (EVI).
@@
 In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared.
-What can be explained is that, south of "Centro," predominate forest and grassland, the most resistant types.
-Also, drought episodes have been less frequent and intense.
+The possible reason for this is that the most resistant types, forest and grassland, predominate south of "Centro." Also, drought episodes have been less frequent and intense.
 The drought episodes have had a lower impact on water availability for vegetation.
@@
 According to @Senf2020, severe drought conditions in Europe are a significant cause of tree mortality.
-However, we found that forest is the type of land cover macroclass less affected by variation in drought indices, being the most resistant land cover class to drought.
-Supporting this is @Fathi-Taperasht2022], who assert that Indian forests are the most drought-resistant and recover rapidly.
+However, we discovered that forests, as the most resilient land cover class to drought, experience less variation in drought indices.
+Supporting this is @Fathi-Taperasht2022, who asserts that Indian forests are the most drought-resistant and recover rapidly.
 Similarly, the work of @Wu2024, who analyzed vegetation loss and recovery in response to meteorological drought in the humid subtropical Pearl River basin in China, indicates that forests showed higher drought resistance.
-Using Vegetation Optical Depth (VOD), kNDVI, and EVI, @Xiao2023 test the resistance of ecosystems and find that ecosystems with more forests are better able to handle severe droughts than croplands.
-They attribute the difference to a deeper rooting depth of trees, a higher water storage capacity, and different water use strategies between forest and cropland [@Xiao2023].
-
-In contrast to what we obtained, @Venegas2022, who studied Cryptocarya alba and Beilschmiedia miersii (both from the Lauraceae family) that live in sclerophyllous forests in Chile, found that the trees' overall growth had slowed down.
+Using Vegetation Optical Depth (VOD), kNDVI, and EVI, @Xiao2023 tests the resistance of ecosystems and finds that ecosystems with more forests are better able to handle severe droughts than croplands.
+They attribute the difference to a deeper rooting depth for trees, a higher water storage capacity, and different water use strategies between forest and cropland [@Xiao2023].
+In contrast, @Venegas2022, who studied Cryptocarya alba and Beilschmiedia miersii (both from the Lauraceae family) that live in sclerophyllous forests in Chile, found that the trees' overall growth had slowed down.
 This could mean that the natural dynamics of their forests have changed.
 They attributed it to the cumulative effects of the unprecedented drought (i.e., hydrological drought).
+
 Thus, we attribute that forest to being the most resistant to drought, due to the fact that most of the species comprising it are highly resilient to water scarcity compared to the other land cover classes.
@@
 
-## Soil moisture, vegetation productivity, and agricultural drought.
+## Vegetation productivity and drought.
 

### 2024-03-20 article/paper_drought_Chile2023.tex
update article

@@
   pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
-  pdfkeywords={drought, land cover change, satellite},
+  pdfkeywords={drought, land cover change, vegetation
+productivity, satellite},
   colorlinks=true,
@@
 \begin{keyword}
-    drought \sep land cover change \sep 
+    drought \sep land cover change \sep vegetation productivity \sep 
     satellite
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, breakable, sharp corners, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, interior hidden, enhanced]}{\end{tcolorbox}}\fi
 
@@
 there are plenty of drought indices that allow for a comprehensive
-assessment of drought from short- to long-term scales and consider
-variables from the earth's water balance on their own (e.g.,
-precipitation, AED, soil moisture). Using this information, we can
-advance our understanding of the impact of drought on ecosystems.
+assessment of drought on short- to long-term scales and that allow for
+the use of single variables from the earth's water balance (e.g.,
+precipitation, AED, soil moisture). The variation in climate variables
+impacts vegetation development, and unfavorable conditions such as low
+precipitation and high temperatures usually generate a decrease in
+vegetation productivity. To monitor the response of vegetation, the
+common practice is to use satellite data. The Normalized Difference
+Vegetation Index (NDVI) has been widely used as a proxy for biomass
+production \citep{Camps-Valls2021, Paruelo2016, Helman2014}. For Chile's
+cultivated land, \citet{Zambrano2018} introduced the zcNDVI for
+assessing seasonal biomass production in response to drought. Using this
+information, we can advance our understanding of the impact of drought
+on ecosystems.
 
@@
 temporal scales that consider the cumulative impact over a period of
-several years. There is limited understanding of the correlation between
-drought and the environment in Chile. Therefore, it is crucial to have a
-deeper understanding of how meteorological and soil moisture droughts
-impact the dynamics of the environment in order to make informed
-decisions on how to adapt.
+several years. The association between drought and the environment in
+Chile is not well comprehended. Hence, it is imperative to acquire a
+more profound comprehension of the manner in which climatic and soil
+moisture droughts influence environmental dynamics, in order to make
+well-informed decisions on adaptation strategies.
 
@@
 
-\hypertarget{trend-of-short--to-long-term-drought}{%
-\subsection{Trend of short- to long-term
-drought}\label{trend-of-short--to-long-term-drought}}
+\hypertarget{short--to-long-term-drought}{%
+\subsection{Short- to long-term
+drought}\label{short--to-long-term-drought}}
 
@@
 cover macroclasses. This way, we tried to reduce the noise in the
-vegetation due to a change in land cover from year to year.
+vegetation due to a change in land cover from year to year. We used the
+zcNDVI as a proxy of vegetation productivity. In Chile's cultivated
+land, \citet{Zambrano2018} introduced the zcNDVI for assessing seasonal
+biomass production in relation to climate.
 
@@
 questions: Which of the drought variables---supply, demand, or soil
-moisture---most helps to explain the changes in plant productivity?
-Which of the short-term or long-term time scales of the drought variable
-has a bigger effect on vegetation productivity in Chile? And finally,
-how strong is the relationship between the variables and the drought
-index? Thus, we will be able to advance in understanding how climate is
-affecting vegetation, considering the impact on the five land cover
-types: forest, cropland, grassland, savanna, and shrubland.
+moisture---most helps to explain the changes in plant productivity? How
+do the short- to long-term time scales of the drought variable affect
+vegetation productivity in Chile? And finally, how strong is the
+relationship between the variables and the drought index? Thus, we will
+be able to advance in understanding how climate is affecting vegetation,
+considering the impact on the five land cover types: forest, cropland,
+grassland, savanna, and shrubland.
 
@@
 months with zcNDVI. We used a method similar to that used by
-\citet{Meroni2017} which compared the SPI with the cumulative FAPAR
-(Fraction of Absorbed Photosynthetically Active Radiation). A
-pixel-to-pixel linear correlation analysis was performed for each index
-within the persistent mask of land cover macroclasses. To begin, the
-Pearson coefficient of correlation is computed for each of the six time
-scales. A time scale is identified as the one that attains the highest
-correlation (p \textless{} 0.05). Subsequently, the Pearson correlation
-coefficient corresponding to the time scales at which the value peaked
-was extracted. As a result, for each index, we generated two raster
-maps: 1) containing the raster with values of the time scales and
-drought index that reached the maximum correlation, and 2) having the
-magnitude of the correlation obtained by the drought index at the time
-scales.
+\citet{Meroni2017} which compared the SPI time-scales with the
+cumulative FAPAR (Fraction of Absorbed Photosynthetically Active
+Radiation). We performed a pixel-to-pixel linear correlation analysis
+for each index within the persistent mask of land cover macroclasses. We
+first compute the Pearson coefficient of correlation for each of the six
+time scales. A time scale is identified as the one that attains the
+highest correlation (p \textless{} 0.05). We then extracted the Pearson
+correlation coefficient corresponding to the time scales where the value
+peaked. As a result, for each index, we generated two raster maps: 1)
+containing the raster with values of the time scales and drought index
+that reached the maximum correlation, and 2) having the magnitude of the
+correlation obtained by the drought index at the time scales.
 
@@
 used \texttt{\{SPEI\}} \citep{Bergueria2023}. For mapping, we use
-\{tmap\} \citep{Tennekes2018}. For data analysis, the suite
-\{tidyverse\} \citep{Wickham2019} was used. For the random forest
+\{tmap\} \citep{Tennekes2018}. For data analysis and visualization, the
+suite \{tidyverse\} \citep{Wickham2019} was used. For the random forest
 modeling, we used the \{tidymodels\}\citep{Kuhn2020} and
@@
 
-\hypertarget{trend-of-short--to-long-term-drought-1}{%
-\subsection{Trend of short- to long-term
-drought}\label{trend-of-short--to-long-term-drought-1}}
+\hypertarget{short--to-long-term-drought-1}{%
+\subsection{Short- to long-term
+drought}\label{short--to-long-term-drought-1}}
 
@@
 
-In Figure~\ref{fig-trendDIMacro}, the averaged aggregation per
-macrozone, drought index, and time scale are shown. The macrozones that
-have the lowest trend are ``Norte Chico'' and ``Centro,'' where the SPI,
-SPEI, and SSI show that it decreases at longer time scales. Potentially
-explained due to the prolonged reduction in precipitation that has
-affected the hydrological system in Chile. At 36 months, it reaches
-trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and
-SSI. For ``Sur,'' the behavior is similar, decreasing at longer scales
-and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI.
-``Norte Grande'' has the highest trend at 36 months for EDDI (0.042 per
-decade), and ``Centro'' has the lowest for SPI and SPEI. In ``Norte
-Grande'' and ``Norte Chico,'' which are in a semi-arid climate, it is
-evident that the EDDI has an effect on the difference between the SPI
-and SPEI index, which is not seen in the other macrozones. Contrary to
-the other macrozones, ``Austral'' showed an increase in all indices,
-being the highest for EDDI at 36 months (0.025) and the lowest for SSI,
-which shows only a minor increase in the trend.
+The Figure~\ref{fig-trendDIMacro} displays the averaged aggregation per
+macrozone, the drought index, and the timescale. The macrozones that
+reached the lowest trend for SPI, SPEI, and SSI are ``Norte Chico'' and
+``Centro,'' where the indices also decrease at longer time scales.
+Potentially explained due to the prolonged reduction in precipitation
+that has affected the hydrological system in Chile. At 36 months, it
+reaches trends between -0.03 and -0.04 (z-score) per decade for SPI,
+SPEI, and SSI. For ``Sur,'' the behavior is similar, decreasing at
+longer scales and having between -0.016 and -0.025 per decade for SPI,
+SPEI, and SSI. ``Norte Grande'' has the highest trend at 36 months for
+EDDI (0.042 per decade), and ``Centro'' has the lowest for SPI and SPEI.
+In ``Norte Grande'' and ``Norte Chico,'' which are in a semi-arid
+climate, it is evident that the EDDI has an effect on the difference
+between the SPI and SPEI index, which is not seen in the other
+macrozones. Contrary to the other macrozones, ``Austral'' showed an
+increase in all indices, being the highest for EDDI at 36 months (0.025)
+and the lowest for SSI, which shows only a minor increase in the trend.
 
 \hypertarget{interaction-of-land-cover-and-drought-1}{%
-\subsubsection{Interaction of land cover and
+\subsection{Interaction of land cover and
 drought}\label{interaction-of-land-cover-and-drought-1}}
@@
 \hypertarget{land-cover-change-1}{%
-\subsection{Land cover change}\label{land-cover-change-1}}
+\subsubsection{Land cover change}\label{land-cover-change-1}}
 
@@
 shifting from the ``Centro'' to the ``Sur.'' Also, there is a high
-increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' replacing the
-savanna lost (Table \ref{tab-landcoverTrend}).
+increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' seemingly
+replacing the savanna lost (Table \ref{tab-landcoverTrend}).
 
@@
 change}\label{relationship-between-drought-indices-and-land-cover-change}}
+
+\begin{table}[!ht]
+\caption{The five most important trends of drought indices in estimating the landcover trend per land cover type and the r-squared (rsq) reached by each random forest model.}
+\label{tab-landcoverTrendRF}
+\includegraphics[]{../output/figs/table_importance_trends_landcover_vs_drought.png}
+\end{table}
 
@@
 
-\begin{table}[!ht]
-\caption{The five most important trends of drought indices in estimating the landcover trend per land cover type and the r-squared (rsq) reached by each random forest model.}
-\label{tab-landcoverTrendRF}
-\includegraphics[]{../output/figs/table_importance_trends_landcover_vs_drought.png}
-\end{table}
-
 \begin{figure}[!ht]
@@
 
+The temporal variation within the macrozones is shown in
+Figure~\ref{fig-zcNDVI_var}b). There is a negative trend in ``Norte
+Chico'' with -0.035 and ``Centro'' with -0.02 per decade. Vegetation
+reached its lowest values for 2019-2022, with an extreme condition in
+early 2020 and 2022 in the ``Norte Chico'' and ``Centro''. The ``Sur''
+and ``Austral'' show a positive trend of around 0.012 and 0.016,
+respectively, per decade (Figure~\ref{fig-zcNDVI_var}).
+
 In Figure~\ref{fig-zcNDVI_var} it is showed the spatial map of trends in
@@
 This decrease in productivity could be associated either with a
-reduction in vegetation surface, a decrease in biomass, or browning
-\citep{Miranda2023}.
-
-The temporal variation within the macrozones is showed in
-Figure~\ref{fig-zcNDVI_var}b). There is a negative trend in ``Norte
-Chico'' with -0.035 and ``Centro'' with -0.02 per decade. Vegetation
-reached its lowest values since the year 2019, with an extreme condition
-in early 2020 and 2022 in the ``Norte Chico'' and ``Centro''. The
-``Sur'' and ``Austral'' show a positive trend of around 0.012 and 0.016,
-respectively per decade (Figure~\ref{fig-zcNDVI_var}). Despite the
-croplands suffering from drought just as badly as the native vegetation
-in ``Norte Chico,'' the savanna and shurbland appears to be the region
-most affected by a negative trend in vegetation productivity across
-Chile.
+reduction in vegetation surface, a decrease in biomass, or browning.
 
@@
 
-Figure~\ref{fig-corTimeScale} is a map that shows the highest
-coefficient of determination (r-squared, or rsq) found in the regression
-analysis between different drought indicators and plant productivity
-over time. The spatial variation of time scales reached per index is
-mostly for time scales above 12 months. In the case of SSI, the
-predominant scales are 6 and 12 months. For all indices, to the north,
-the time scales are higher and diminish toward the south until the south
-part of ``Austral'' increases. In Figure~\ref{fig-corPerson}, the map of
-Pearson correlation values is shown. The EDDI reached correlations above
-0.5 between ``Norte Chico'' and ``Sur.'' The correlation changes from
-negative to positive toward the Andes Mountains and to the sea, just as
-in the northern part of ``Austral.'' The SPI and SPEI have similar
-results, with the higher values in ``Norte Chico'' and ``Centro'' being
-higher than 0.6. Following a similar spatial pattern as EDDI. The SSI
-showed to be the index that has a major spatial extension with a higher
-correlation. It has a similar correlation to SPI and SPEI for ``Norte
-Chico'' and ``Sur,''~ but has improvements for ``Sur.''
+Figure~\ref{fig-corTimeScale} shows the highest coefficient of
+determination (r-squared, or rsq) found in the regression analysis
+between different drought indicators and zcNDVI over time scales (1, 3,
+6, 12, 24, and 36 months). The spatial variation of time scales reached
+per index is mostly for time scales above 12 months. In the case of SSI,
+the predominant scales are 6 and 12 months. For all indices, to the
+north, the time scales are higher and diminish toward the south until
+the south part of ``Austral,'' where they increase. In
+Figure~\ref{fig-corPerson}, the map of Pearson correlation values (r) is
+shown. The EDDI reached correlations above 0.5 between ``Norte Chico''
+and ``Sur.'' The correlation changes from negative to positive toward
+the Andes Mountains and to the sea, just as in the northern part of
+``Austral.'' The SPI and SPEI have similar results, with the higher
+values in ``Norte Chico'' and ``Centro'' being higher than 0.6.
+Following a similar spatial pattern as EDDI but with an opposite sign.
+The SSI showed to be the index that has a major spatial extension with a
+higher correlation. It has a similar correlation to SPI and SPEI for
+``Norte Chico'' and ``Sur,''~ but has improvements for ``Sur.''
 
@@
 
-\hypertarget{drought-trend-and-attribution-to-land-cover}{%
-\subsection{Drought trend and attribution to land
-cover}\label{drought-trend-and-attribution-to-land-cover}}
+\hypertarget{drought-trend-and-attribution-to-land-cover-change}{%
+\subsection{Drought trend and attribution to land cover
+change}\label{drought-trend-and-attribution-to-land-cover-change}}
 
 \citet{Vicente-Serrano2022}, in a study at the global scale of drought
-trends, indicate that there have not been significant trends in
+trends, indicates that there have not been significant trends in
 meteorological drought since 1950. Also, state that the increase in
@@
 agricultural drought impact globally and in the main grain producer
-countries, finding that \emph{``since 1980, the Earth warming has not
-changed the drought area or intensity.''} In our study, we considered
-the variation in vegetation productivity in Chile for areas without
-changes in land cover, to avoid misleading conclusions that could be
-related to the increase in water demand due to land cover change. Our
-results show a contrasting perspective. There has been a significant
-trend in the decline of vegetation productivity (zcNDVI) since 2000 for
-``Norte Chico'' and ``Centro,'' which has been extreme between 2020 and
-2022, seemsly due to an intense hydrological drought due to the
-persistance of the mega drought \citep{Garreaud2017}. However, a rise in
-irrigated land doesn't seem to have an impact on this hydrological
-drought. Despite using the persistance mask for vegetation's trend
-analysis, cropland, which is the most water-demand type, showed a
-decrease trend in ``Norte Chico'' and ``Centro.'' Also, there was an
-increase in barren land for both types. These changes are associated
-with a decrease in water demand from vegetation. Nonetheless, we used
-the persistant land cover to ensure that the pixel has the same class;
-in the case of croplands, it could happen that some areas had changed
-crops for others with higher water consumption and consequently increase
-water demand. But this effect should be minor compared to the results
-from land cover change at this scale of analysis.
-
-On the other hand, for ``Norte Chico'' and ``Centro,'' our results show
-a decrease in trends of water supply (SPI and SSI), which are higher at
-larger time scales, which is evidence of the hydrological drought. We
-say that what happened in central Chile goes against what has been found
-on a global scale \citep{Vicente-Serrano2022, Kogan2020}. This shows
-that the main cause of the hydrological drought in Chile was a steady
-drop in water supply made worse by an increase in AED, but it seems that
-in zones most affected by drought, the main cause is not an increase in
-water demand by vegetation like irrigated crops. Finally, north-central
-Chile has experienced a decline in vegetation productivity across all
-macroclasses, which is primarily attributable to variations in water
-supply and soil moisture. An increase in water demand, such as an
-increase in the surface area of irrigated crops, could strengthen this
-trend.
-
-\hypertarget{land-cover-types-and-their-impact-by-drought}{%
-\subsection{Land cover types and their impact by
-drought}\label{land-cover-types-and-their-impact-by-drought}}
-
-We discovered that croplands, savannas, and shrubland are the most
-susceptible to climatic changes and are most affected by the 12-month
-soil moisture deficit. In a study in the Yangtze River Basin in China,
-\citet{Jiang2020} analyzed the impact of drought on vegetation using the
-SPEI and the Enahanced Vegetation Index (EVI). They found that cropland
-was more sensitive to drought than grassland, showing that cropland
-responds strongly to short- and medium-term drought (\textless{}
-SPEI-6). In our case, the SPEI-12 was the one that most impacted the
-croplands in ``Norte Chico'' and ``Centro.'' In general, most studies
-show that croplands are most sensitive to short-term drought
-(\textless{} SPI-6)
+countries, finding that ``since 1980, the Earth warming has not changed
+the drought area or intensity.'' In our study, we took into account the
+variation in vegetation productivity in Chile, specifically in areas
+without any changes in land cover, to prevent any misleading conclusions
+about the increase in water demand due to land cover change. Our results
+show a contrasting perspective. The SPI, SPEI, and SSI (water supply)
+showed a decrease in trends, except for the southern part, and an
+increase in EDDI (water demand). The trend, positive or negative, was
+stronger as the time scales increased. Trends in the long term (e.g., 36
+months) are evidence of how human-induced climate change is affecting
+Chile, which seems to be due to an intense hydrological drought
+resulting from the persistence of the precipitation deficit. We found
+that there has been a significant trend in the decline of vegetation
+productivity (zcNDVI) since 2000 for the north-central part of the
+country, which has been extreme between 2020 and 2022 and has impacted
+natural and cultivated land. Additionally, we demonstrated that the
+drought, primarily due to an increase in AED, accounts for about 30\% of
+the changes in land cover types (excluding croplands). These changes are
+associated with a decrease in water demand from vegetation. Moreover,
+the most water-demanding type, cropland, showed a decrease in the
+north-central region, while barren land showed an increase. The
+north-central part of the country primarily experienced these changes
+due to a higher increase in AED. Thus, we have evidence of a significant
+decline in water supply and an increase in AED for the north-central
+part of Chile, which show to be the most relevant variables for drought
+conditions. Some questions arise regarding what is occurring with the
+cultivated land. We used the unchanged land cover to ensure that an
+increase in surface area is not considered in the trend analysis. For
+croplands, it could happen that some areas have changed the types of
+crops for others with higher water demand, which consequently increases
+water demand. However, this effect should be minor compared to the
+decrease in water supply and increase in water demand at this scale of
+analysis.
+
+This shows that the main cause of the hydrological drought in Chile was
+a steady drop in water supply made worse by an increase in AED, but it
+seems that in zones most affected by drought, the main cause is not an
+increase in water demand due to an intensification of cultivated land
+(e.g., an increase in irrigated crops). North-central Chile has
+experienced a decline in vegetation productivity across land cover
+types, which is primarily attributable to variations in water supply and
+soil moisture. An increase in water demand, such as an increase in the
+surface area of irrigated crops, could strengthen this trend.
+
+\hypertarget{impact-of-short--to-long-term-droughts-on-land-cover-types}{%
+\subsection{Impact of short- to long-term droughts on land cover
+types}\label{impact-of-short--to-long-term-droughts-on-land-cover-types}}
+
+We found that the 12-month soil moisture deficit most affects the
+productivity of vegetation in all land cover types along Chile. In a
+study in the Yangtze River Basin in China, \citet{Jiang2020} analyzed
+the impact of drought on vegetation using the SPEI and the Enahanced
+Vegetation Index (EVI). They found that cropland was more sensitive to
+drought than grassland, showing that cropland responds strongly to
+short- and medium-term drought (\textless{} SPEI-6). In our case, the
+SPEI-12 was the one that most impacted the croplands in ``Norte Chico''
+and ``Centro.'' In general, most studies show that croplands are most
+sensitive to short-term drought (\textless{} SPI-6)
 \citep{Zambrano2016, Potopova2015, Dai2020, Rhee2010}. Short-term
@@
 between drought indices and vegetation productivity decrease, as do the
-time scales that reach the maximum r-squared. What can be explained is
-that, south of ``Centro,'' predominate forest and grassland, the most
-resistant types. Also, drought episodes have been less frequent and
+time scales that reach the maximum r-squared. The possible reason for
+this is that the most resistant types, forest and grassland, predominate
+south of ``Centro.'' Also, drought episodes have been less frequent and
 intense. The drought episodes have had a lower impact on water
@@
 According to \citet{Senf2020}, severe drought conditions in Europe are a
-significant cause of tree mortality. However, we found that forest is
-the type of land cover macroclass less affected by variation in drought
-indices, being the most resistant land cover class to drought.
-Supporting this is \citet{Fathi-Taperasht2022}{]}, who assert that
-Indian forests are the most drought-resistant and recover rapidly.
-Similarly, the work of \citet{Wu2024}, who analyzed vegetation loss and
-recovery in response to meteorological drought in the humid subtropical
-Pearl River basin in China, indicates that forests showed higher drought
-resistance. Using Vegetation Optical Depth (VOD), kNDVI, and EVI,
-\citet{Xiao2023} test the resistance of ecosystems and find that
-ecosystems with more forests are better able to handle severe droughts
-than croplands. They attribute the difference to a deeper rooting depth
-of trees, a higher water storage capacity, and different water use
-strategies between forest and cropland \citep{Xiao2023}.
-
-In contrast to what we obtained, \citet{Venegas2022}, who studied
+significant cause of tree mortality. However, we discovered that
+forests, as the most resilient land cover class to drought, experience
+less variation in drought indices. Supporting this is
+\citet{Fathi-Taperasht2022}, who asserts that Indian forests are the
+most drought-resistant and recover rapidly. Similarly, the work of
+\citet{Wu2024}, who analyzed vegetation loss and recovery in response to
+meteorological drought in the humid subtropical Pearl River basin in
+China, indicates that forests showed higher drought resistance. Using
+Vegetation Optical Depth (VOD), kNDVI, and EVI, \citet{Xiao2023} tests
+the resistance of ecosystems and finds that ecosystems with more forests
+are better able to handle severe droughts than croplands. They attribute
+the difference to a deeper rooting depth for trees, a higher water
+storage capacity, and different water use strategies between forest and
+cropland \citep{Xiao2023}. In contrast, \citet{Venegas2022}, who studied
 Cryptocarya alba and Beilschmiedia miersii (both from the Lauraceae
@@
 cumulative effects of the unprecedented drought (i.e., hydrological
-drought). Thus, we attribute that forest to being the most resistant to
-drought, due to the fact that most of the species comprising it are
-highly resilient to water scarcity compared to the other land cover
-classes. Nonetheless, if we want to go deep in our analysis, we should
-use earth observation data that is able to capture a higher level of
-detail. For example, when we used MOD13A3 with a 1km spatial resolution
-to measure vegetation condition, it took the average condition of 1
-square kilometer. Then, to use remote sensing to look at how a certain
-type of forest (like sclerophyllous forest) changes in response to
-drought on a local level, we should use operational products with higher
-spatial resolutions, like those from Landsat or Sentinel. This will let
-us do a more thorough analysis.
-
-\hypertarget{soil-moisture-vegetation-productivity-and-agricultural-drought.}{%
-\subsection{Soil moisture, vegetation productivity, and agricultural
-drought.}\label{soil-moisture-vegetation-productivity-and-agricultural-drought.}}
+drought).
+
+Thus, we attribute that forest to being the most resistant to drought,
+due to the fact that most of the species comprising it are highly
+resilient to water scarcity compared to the other land cover classes.
+Nonetheless, if we want to go deep in our analysis, we should use earth
+observation data that is able to capture a higher level of detail. For
+example, when we used MOD13A3 with a 1km spatial resolution to measure
+vegetation condition, it took the average condition of 1 square
+kilometer. Then, to use remote sensing to look at how a certain type of
+forest (like sclerophyllous forest) changes in response to drought on a
+local level, we should use operational products with higher spatial
+resolutions, like those from Landsat or Sentinel. This will let us do a
+more thorough analysis.
+
+\hypertarget{vegetation-productivity-and-drought.}{%
+\subsection{Vegetation productivity and
+drought.}\label{vegetation-productivity-and-drought.}}
 

### 2024-03-21 article/paper_drought_Chile2023.qmd
update article

@@
             \newcommand{\elandscape}{\end{landscape}}
-  docx:
+  elsevier-docx:
     keep-tex: true
+    number-sections: true
     journal:
@@
 Similarly, @Kogan2020 analyzed the drought trend using vegetation health methods, finding that for the globe, hemispheres, and main grain-producing countries, drought has not expanded or intensified for the last 38 years.
-Further, the @IPCC2021 suggests that there is a high degree of confidence that rising temperatures will increase the extent, frequency, and severity of droughts.
+Further, @IPCC2021 suggests that there is a high degree of confidence that rising temperatures will increase the extent, frequency, and severity of droughts.
 Also, AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C.
@@
 
-## Short- to long-term drought
+## Short- to long-term drought trends
 
@@
 The drought indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months at a monthly frequency for 1981–2023 in order to be used for short- to long-term evaluation of drought.
-In the case of the proxy of vegetation productivity (zcNDVI) it was calculated for a time scale of six months at monthly frequency for 2000–2023.
-For zcNDVI, we test time scales of 1, 3, 6, and 12 months; we choose to use six months because that shows a more stable representation of vegetation productivity.
+In the case of the proxy of vegetation productivity (zcNDVI), it was calculated for a time scale of six months at monthly frequency for 2000–2023.
+For zcNDVI, we test time scales of 1, 3, 6, and 12 months in December and its correlation with net primary production (NPP) obtained from the MOD17A3HGF product from MODIS.
+We choose to use six months because r-squared with NPP increases from one to six months and from six to 12 months has little improvement (see supplementary material in Section S5).
 
@@
 Grey colors indicate either no vegetation or a change in land cover type for 2001–2022.
-(b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile.
+(b) Temporal variation of zcNDVI aggregated at macrozone level within continental Chile.
 Each horizontal panel corresponds to a macrozone from 'Norte Grande' to 'Austral'."
@@
 
-@fig-corTimeScale shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between different drought indicators and zcNDVI over time scales (1, 3, 6, 12, 24, and 36 months).
+@fig-corTimeScale shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between zcNDVI and  different drought indicators over time scales of 1, 3, 6, 12, 24, and 36 months.
 The spatial variation of time scales reached per index is mostly for time scales above 12 months.
@@
 
-## Drought trend and attribution to land cover change
+## The main drivers of drought in Chile
 
@@
 
-This shows that the main cause of the hydrological drought in Chile was a steady drop in water supply made worse by an increase in AED, but it seems that in zones most affected by drought, the main cause is not an increase in water demand due to an intensification of cultivated land (e.g., an increase in irrigated crops).
+This shows that the main cause of the hydrological drought in Chile was a steady drop in water supply made worse by an increase in AED, but it seems that in zones most affected by drought, the main cause is not an increase in vegetation water demand due to an intensification of cultivated land (e.g., an increase in irrigated crops).
 North-central Chile has experienced a decline in vegetation productivity across land cover types, which is primarily attributable to variations in water supply and soil moisture.
 An increase in water demand, such as an increase in the surface area of irrigated crops, could strengthen this trend.
+But it is out of the scope of this study.
+Future work should focus on the regions where the drought has been more severe and has a high proportion of irrigated crops to get insight on the real impact of irrigation on those zones.
 
-## Impact of short- to long-term droughts on land cover types
+## Land cover sensitivity to drought
 
-We found that the 12-month soil moisture deficit most affects the productivity of vegetation in all land cover types along Chile.
+We analyzed two main impacts of drought on land cover.
+First, the attribution of drought to the change in surface area per land cover type.
+Drought accounts for about 30% of the surface change per land cover type, with the exception of croplands.
+The main variables associated with these changes are the increase in AED and, in second place, the decrease in precipitation.
+Second, we analyzed the time series of drought indices and vegetation productivity per land cover type.
+In this case, the most important variables that had an impact on zcNDVI were the soil moisture deficit, followed by the precipitation deficit, and in third place, AED.
+
 In a study in the Yangtze River Basin in China, @Jiang2020 analyzed the impact of drought on vegetation using the SPEI and the Enahanced Vegetation Index (EVI).
@@
 
+We found that the 12-month soil moisture deficit most affects the productivity of vegetation in all land cover types along Chile.
 The main external factors that affect biomass production by vegetation are actual evapotranspiration and soil moisture, and the rate of ET in turn depends on the availability of water storage in the root zone.
@@
 
-# Supplementary material {-}
-
 # References {-}

### 2024-03-21 article/paper_drought_Chile2023.tex
update article

@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, interior hidden, enhanced]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi
 
@@
 grain-producing countries, drought has not expanded or intensified for
-the last 38 years. Further, the \citet{IPCC2021} suggests that there is
-a high degree of confidence that rising temperatures will increase the
+the last 38 years. Further, \citet{IPCC2021} suggests that there is a
+high degree of confidence that rising temperatures will increase the
 extent, frequency, and severity of droughts. Also, AR6 \citep{IPCC2023}
@@
 
-\hypertarget{short--to-long-term-drought}{%
-\subsection{Short- to long-term
-drought}\label{short--to-long-term-drought}}
+\hypertarget{short--to-long-term-drought-trends}{%
+\subsection{Short- to long-term drought
+trends}\label{short--to-long-term-drought-trends}}
 
@@
 for short- to long-term evaluation of drought. In the case of the proxy
-of vegetation productivity (zcNDVI) it was calculated for a time scale
+of vegetation productivity (zcNDVI), it was calculated for a time scale
 of six months at monthly frequency for 2000--2023. For zcNDVI, we test
-time scales of 1, 3, 6, and 12 months; we choose to use six months
-because that shows a more stable representation of vegetation
-productivity.
+time scales of 1, 3, 6, and 12 months in December and its correlation
+with net primary production (NPP) obtained from the MOD17A3HGF product
+from MODIS. We choose to use six months because r-squared with NPP
+increases from one to six months and from six to 12 months has little
+improvement (see supplementary material in Section S5).
 
@@
 
-\hypertarget{short--to-long-term-drought-1}{%
-\subsection{Short- to long-term
-drought}\label{short--to-long-term-drought-1}}
+\hypertarget{short--to-long-term-drought-trends-1}{%
+\subsection{Short- to long-term drought
+trends}\label{short--to-long-term-drought-trends-1}}
 
@@
 productivity. Grey colors indicate either no vegetation or a change in
-land cover type for 2001--2022. (b) Temporal variation of zcNDVI-6
+land cover type for 2001--2022. (b) Temporal variation of zcNDVI
 aggregated at macrozone level within continental Chile. Each horizontal
@@
 determination (r-squared, or rsq) found in the regression analysis
-between different drought indicators and zcNDVI over time scales (1, 3,
-6, 12, 24, and 36 months). The spatial variation of time scales reached
-per index is mostly for time scales above 12 months. In the case of SSI,
-the predominant scales are 6 and 12 months. For all indices, to the
-north, the time scales are higher and diminish toward the south until
-the south part of ``Austral,'' where they increase. In
+between zcNDVI and different drought indicators over time scales of 1,
+3, 6, 12, 24, and 36 months. The spatial variation of time scales
+reached per index is mostly for time scales above 12 months. In the case
+of SSI, the predominant scales are 6 and 12 months. For all indices, to
+the north, the time scales are higher and diminish toward the south
+until the south part of ``Austral,'' where they increase. In
 Figure~\ref{fig-corPerson}, the map of Pearson correlation values (r) is
@@
 
-\hypertarget{drought-trend-and-attribution-to-land-cover-change}{%
-\subsection{Drought trend and attribution to land cover
-change}\label{drought-trend-and-attribution-to-land-cover-change}}
+\hypertarget{the-main-drivers-of-drought-in-chile}{%
+\subsection{The main drivers of drought in
+Chile}\label{the-main-drivers-of-drought-in-chile}}
 
@@
 seems that in zones most affected by drought, the main cause is not an
-increase in water demand due to an intensification of cultivated land
-(e.g., an increase in irrigated crops). North-central Chile has
-experienced a decline in vegetation productivity across land cover
-types, which is primarily attributable to variations in water supply and
-soil moisture. An increase in water demand, such as an increase in the
-surface area of irrigated crops, could strengthen this trend.
-
-\hypertarget{impact-of-short--to-long-term-droughts-on-land-cover-types}{%
-\subsection{Impact of short- to long-term droughts on land cover
-types}\label{impact-of-short--to-long-term-droughts-on-land-cover-types}}
-
-We found that the 12-month soil moisture deficit most affects the
-productivity of vegetation in all land cover types along Chile. In a
-study in the Yangtze River Basin in China, \citet{Jiang2020} analyzed
-the impact of drought on vegetation using the SPEI and the Enahanced
-Vegetation Index (EVI). They found that cropland was more sensitive to
-drought than grassland, showing that cropland responds strongly to
-short- and medium-term drought (\textless{} SPEI-6). In our case, the
-SPEI-12 was the one that most impacted the croplands in ``Norte Chico''
-and ``Centro.'' In general, most studies show that croplands are most
-sensitive to short-term drought (\textless{} SPI-6)
+increase in vegetation water demand due to an intensification of
+cultivated land (e.g., an increase in irrigated crops). North-central
+Chile has experienced a decline in vegetation productivity across land
+cover types, which is primarily attributable to variations in water
+supply and soil moisture. An increase in water demand, such as an
+increase in the surface area of irrigated crops, could strengthen this
+trend. But it is out of the scope of this study. Future work should
+focus on the regions where the drought has been more severe and has a
+high proportion of irrigated crops to get insight on the real impact of
+irrigation on those zones.
+
+\hypertarget{land-cover-sensitivity-to-drought}{%
+\subsection{Land cover sensitivity to
+drought}\label{land-cover-sensitivity-to-drought}}
+
+We analyzed two main impacts of drought on land cover. First, the
+attribution of drought to the change in surface area per land cover
+type. Drought accounts for about 30\% of the surface change per land
+cover type, with the exception of croplands. The main variables
+associated with these changes are the increase in AED and, in second
+place, the decrease in precipitation. Second, we analyzed the time
+series of drought indices and vegetation productivity per land cover
+type. In this case, the most important variables that had an impact on
+zcNDVI were the soil moisture deficit, followed by the precipitation
+deficit, and in third place, AED.
+
+In a study in the Yangtze River Basin in China, \citet{Jiang2020}
+analyzed the impact of drought on vegetation using the SPEI and the
+Enahanced Vegetation Index (EVI). They found that cropland was more
+sensitive to drought than grassland, showing that cropland responds
+strongly to short- and medium-term drought (\textless{} SPEI-6). In our
+case, the SPEI-12 was the one that most impacted the croplands in
+``Norte Chico'' and ``Centro.'' In general, most studies show that
+croplands are most sensitive to short-term drought (\textless{} SPI-6)
 \citep{Zambrano2016, Potopova2015, Dai2020, Rhee2010}. Short-term
@@
 
-The main external factors that affect biomass production by vegetation
-are actual evapotranspiration and soil moisture, and the rate of ET in
-turn depends on the availability of water storage in the root zone.
-Thus, soil moisture plays a key role in land carbon uptake and,
-consequently, in the production of biomass \citep{Humphrey2021}.
-Moreover, \citet{Zhang2022} indicate there is a bidirectional causality
-between soil moisture and vegetation productivity. Lastly, some studies
-have redefined agricultural drought as soil moisture drought from a
+We found that the 12-month soil moisture deficit most affects the
+productivity of vegetation in all land cover types along Chile. The main
+external factors that affect biomass production by vegetation are actual
+evapotranspiration and soil moisture, and the rate of ET in turn depends
+on the availability of water storage in the root zone. Thus, soil
+moisture plays a key role in land carbon uptake and, consequently, in
+the production of biomass \citep{Humphrey2021}. Moreover,
+\citet{Zhang2022} indicate there is a bidirectional causality between
+soil moisture and vegetation productivity. Lastly, some studies have
+redefined agricultural drought as soil moisture drought from a
 hydrological perspective \citep{Loon2016, Samaniego2018}. Even though
@@
 
-\hypertarget{supplementary-material}{%
-\section*{Supplementary material}\label{supplementary-material}}
-\addcontentsline{toc}{section}{Supplementary material}
-
 \renewcommand\refname{References}

### 2024-03-21 article/supplementary.qmd
update article

@@
 #| out-width: 75%
+#| fig-pos: '!ht'
 #| fig-cap: "Heatmap of trends in zcNDVI for 2000 to 2023 per macrozone and landcover macroclass."
@@
 ```
+
+# Vegetation productivity
+
+We analyzed the correlation of zcNDVI for time scales of 1, 3, 6, and 12 months versus net primary production (NPP).
+Both were obtained from MODIS products.
+zcNDVI from MOD13A3.061 and NPP from MOD17A3HGF.061.
+
+```{r}
+#| echo: false
+#| out-width: 75%
+#| fig-pos: '!ht'
+#| fig-cap: "Spatial variation of the r-squared values obtained from the yearly correlation of zcNDVI of 1, 3, 6, and 12 months with the net primary productivity (NPP) for continental Chile."
+knitr::include_graphics('../output/figs/map_r2_NPP_vs_zcNDVI.png')
+```
+
+```{r}
+#| echo: false
+#| out-width: 75%
+#| fig-pos: '!ht'
+#| fig-cap: "A heatmap showing the r-squared values obtained from the yearly correlation of zcNDVI of 1, 3, 6, and 12 months with the net primary productivity (NPP) for continental Chile."
+knitr::include_graphics('../output/figs/heatmap_r2_npp_vs_zcndvi_per_landcover.png')
+```

### 2024-04-04 article/paper_drought_Chile2023.qmd
table landcover

@@
           postal-code: 7500994
+        - id: odes
+          name: Universidad Mayor
+          department: Observatorio de Sequía para la Agricultura y la Biodiversidad de Chile (ODES)
     attributes:
@@
  The north-central region of Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development.
-This persistent period of water scarcity has been defined as a megadrought.
-The aim of our study is to evaluate the land cover change over continental Chile and to examine how this is connected to drought indices of water supply, atmospheric evaporative demand (AED), soil moisture, and their effects on vegetation productivity.
-The drought indices were derived using monthly ERA5-Land reanalysis data spanning from 1981 to 2023.
-The Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets were utilized to obtain information on annual land cover and monthly vegetation productivity.
-We analyzed short- (1, 3, 6 months) to long-term (12, 24, 36 months) time scales of drought.
-Our results showed that land cover change was highest in the south-central part of the country, reaching changes as high as 36% in the surface type.
-The water demand has increased for the whole country, with a major increase in the north.
-The AED and soil moisture evidence a decreasing trend, which decreases at longer time scales and from north to south.
-The extreme south part of the country shows an increase in supply.
-Vegetation productivity has a negative trend in the north-central region for all land cover types.
-On the other hand, forests seem to be the most resistant type to drought.
-The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
-The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are the ones that consider soil moisture for twelve months and the combined effect of precipitation and AED for 24 and 12 months.
-The results indicate that the north-central region is the most sensitive to water supply deficits lasting longer than a year.
+This persistent period of water scarcity has been defined as a mega-drought.
+The aim of our study is to evaluate the interaction of drought, land cover change, and vegetation productivity over continental Chile.
+We used drought indices for atmospheric evaporative demand (AED), water supply, and soil moisture from short-term (1, 3, 6 months) to long-term (12, 24, 36 months) time scales.
+We derived the drought indices using monthly ERA5-Land reanalysis data spanning from 1981 to 2023.
+We utilized the Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets to derive information on annual land cover and monthly vegetation productivity.
+Our results showed that from south to north, Chile has a declining trend in water supply, and across the whole country, there is an increase in AED.
+These trends are stronger at longer time scales.
+The trend in vegetation productivity in the north-central area is affecting, to a higher degree, shrubland and savanna, followed by croplands and forests.
+The drought explains about 30% of the change in land cover type across Chile for forest, grassland, shrubland, and savanna.
+The increase in AED is the main driver of the change in land cover, followed by a reduction in precipitation and soil moisture.
+The change in vegetation productivity has been severe in the north-central part of the country for all land cover types, particularly savanna, shrubland, and croplands.
+The anomaly in soil moisture over the past 12 months (SSI-12) is the main variable explaining these changes, followed by anomalies in cumulated precipitation over one to two years (SPI-12 and SPI-24).
+Our results provide insightful information that would help in developing adaptation measures for ecosystems in Chile to cope with climate change and drought.
+
 keywords: 
@@
 In our case, for the SSI, we used the average soil moisture from ERA5L at 1m depth.
-Finally, for the proxy of productivity, we used the zcNDVI proposed by @Zambrano2018, which was derived from the monthly time series of NDVI retrieved from MOD13A1.
+Finally, for the proxy of productivity, we used the zcNDVI proposed by @Zambrano2018, which was derived from the monthly time series of NDVI derived from MOD13A1.
 All the indices are multi-scalar and can be used for the analysis of short- to long-term droughts.
@@
 In the case of the proxy of vegetation productivity (zcNDVI), it was calculated for a time scale of six months at monthly frequency for 2000–2023.
-For zcNDVI, we test time scales of 1, 3, 6, and 12 months in December and its correlation with net primary production (NPP) obtained from the MOD17A3HGF product from MODIS.
-We choose to use six months because r-squared with NPP increases from one to six months and from six to 12 months has little improvement (see supplementary material in Section S5).
+For zcNDVI, we test time scales of 1, 3, 6, and 12 months in December and their correlation with net primary production (NPP) obtained from the MOD17A3HGF product from MODIS.
+We choose to use six months because r-squared with NPP increases from one to six months and from six to 12 months has little improvement.
+The r-squared were between 0.31 for forest and 0.72 for shrubland (see supplementary material in Section S5).
 
@@
 
-## Future outlook (to complete)
+## Drought information to aid in adaptation
+
+Different climate components, such as evaporative water demand, water supply, soil moisture, and their impact on vegetation, should be considered when evaluating the multi-dimensional nature of drought.
+For a better understanding of the propagation of drought [@VanLoon2012] from meteorological to hydrological drought, we should consider the climatic response at different time scales, ranging from short to long.
+Furthermore, we must make this information publicly available and update it frequently to aid in the development of adaptation policies.
+The drought observatory for agriculture and biodiversity of Chile (ODES) shares this information aggregated for multiple administrative and hydrological units across Chile ([https://odes-chile.org/app/unidades](https://odes-chile.org/app/unidades)), with the goal of helping to prepare for future climatic conditions.
 
@@
 
-There is a trend toward decreasing water supply in most parts of Chile, particularly in the “Centro” and “Norte Chico” regions.
-The whole country showed an increase in AED.
-Vegetation productivity only showed a decrease in the “Norte Chico” and “Centro,” being highest for shrubland and croplands.
-Forest is the land cover most resistant to drought, as shown along Chile, and shrubland and cropland are the most sensitive.
+There is a trend toward decreasing water supply (SPI, SPEI, and SSI) in most of Chile; just in the southern part, there is an increase.
+The trend is most strong in the north-central zone.
+The whole country showed an increase in water demand (AED).
+The magnitude of the trend increases over longer time scales, which is evidence that the deficit is impacting the hydrological system in Chile.
+The trend in vegetation productivity in the north-central area is affecting, to a higher degree, shrubland and savanna, followed by croplands and forests.
+The most important changes in land cover are the increase of forest in "Sur" and shrubland and grassland in "Centro;" and the increase of savanna in "Centro" and "Sur.".
 
-A soil moisture deficit of 12 months (SSI-12) is highly correlated with vegetation productivity for the land cover classes of shrubland, savannas, croplands, and forest in “Norte Chico” and “Centro.” For the southern part of the country with humid conditions, the correlation with SSI decreases.
-Soil moisture overcomes the capacity to explain vegetation productivity over the supply and demand drought indices in the entire territory.
+The drought explains about 30% of the change in land cover type across Chile for forest, grassland, shrubland, and savanna.
+There is no evidence of an effect of drought on the change in cropland surface area.
+The increase in AED is the main driver of the change in land cover, followed by a reduction in precipitation and soil moisture.
+The drought time scales vary regarding the land cover type.
 
-The variation in vegetation productivity appears to be associated with climate variation rather than anthropogenic factors (e.g., an increase in water demand by irrigated crops).
-Even though switching to more demanding crops on the land could increase the impact of drought on vegetation, this would need to be more thoroughly investigated, for instance at the watershed level.
+The change in vegetation productivity has been severe in the north-central part of the country for all land cover types, particularly savanna, shrubland, and croplands.
+The anomaly in soil moisture over the past 12 months is the main variable explaining these changes, followed by anomalies in cumulated precipitation over one to two years.
+The variation in AED seems to intensify the drought impact on vegetation productivity.
 
-The results of this study could help in the development of a robust forecasting system for land cover classes in Chile, helping to improve preparedness for climate change impacts on vegetation.
+The results of this study provide insightful information that would help in developing adaptation measures for ecosystems in Chile to cope with climate change and drought.
+
+# Acknowledgment
+
+The National Research and Development Agency of Chile (ANID) funded this study through the drought emergency project FSEQ210022, Fondecyt Iniciación N°11190360, and Fondecyt Regular N°1210526.
 

### 2024-04-04 article/paper_drought_Chile2023.tex
table landcover

@@
 vegetation productivity in continental Chile}
-\author[1]{Francisco Zambrano%
+\author[1,2]{Francisco Zambrano%
 \corref{cor1}%
@@
  \ead{francisco.zambrano@umayor.cl} 
-\author[2]{Anton Vrieling%
+\author[3]{Anton Vrieling%
 %
@@
 
-\author[3]{Francisco Fernández%
+\author[4]{Francisco Fernández%
 %
@@
 Chile},postcode={7500994},postcodesep={}}
-\affiliation[2]{organization={University of Twente, Faculty of
+\affiliation[2]{organization={Universidad Mayor, Observatorio de Sequía
+para la Agricultura y la Biodiversidad de Chile (ODES)},,postcodesep={}}
+\affiliation[3]{organization={University of Twente, Faculty of
 Geo-Information Science and Earth Observation (ITC)},city={Enschede, The
 Netherlands},postcodesep={}}
-\affiliation[3]{organization={Universidad San
+\affiliation[4]{organization={Universidad San
 Sebastian},,postcodesep={}}
@@
 hydrological system and vegetation development. This persistent period
-of water scarcity has been defined as a megadrought. The aim of our
-study is to evaluate the land cover change over continental Chile and to
-examine how this is connected to drought indices of water supply,
-atmospheric evaporative demand (AED), soil moisture, and their effects
-on vegetation productivity. The drought indices were derived using
-monthly ERA5-Land reanalysis data spanning from 1981 to 2023. The
-Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets were
-utilized to obtain information on annual land cover and monthly
-vegetation productivity. We analyzed short- (1, 3, 6 months) to
-long-term (12, 24, 36 months) time scales of drought. Our results showed
-that land cover change was highest in the south-central part of the
-country, reaching changes as high as 36\% in the surface type. The water
-demand has increased for the whole country, with a major increase in the
-north. The AED and soil moisture evidence a decreasing trend, which
-decreases at longer time scales and from north to south. The extreme
-south part of the country shows an increase in supply. Vegetation
-productivity has a negative trend in the north-central region for all
-land cover types. On the other hand, forests seem to be the most
-resistant type to drought. The types that show to be most affected by
-variation in climate conditions are shrublands, savannas, and croplands.
-The drought indices that have the capability of explaining to a major
-degree the variance in vegetation productivity are the ones that
-consider soil moisture for twelve months and the combined effect of
-precipitation and AED for 24 and 12 months. The results indicate that
-the north-central region is the most sensitive to water supply deficits
-lasting longer than a year.
+of water scarcity has been defined as a mega-drought. The aim of our
+study is to evaluate the interaction of drought, land cover change, and
+vegetation productivity over continental Chile. We used drought indices
+for atmospheric evaporative demand (AED), water supply, and soil
+moisture from short-term (1, 3, 6 months) to long-term (12, 24, 36
+months) time scales. We derived the drought indices using monthly
+ERA5-Land reanalysis data spanning from 1981 to 2023. We utilized the
+Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets to derive
+information on annual land cover and monthly vegetation productivity.
+Our results showed that from south to north, Chile has a declining trend
+in water supply, and across the whole country, there is an increase in
+AED. These trends are stronger at longer time scales. The trend in
+vegetation productivity in the north-central area is affecting, to a
+higher degree, shrubland and savanna, followed by croplands and forests.
+The drought explains about 30\% of the change in land cover type across
+Chile for forest, grassland, shrubland, and savanna. The increase in AED
+is the main driver of the change in land cover, followed by a reduction
+in precipitation and soil moisture. The change in vegetation
+productivity has been severe in the north-central part of the country
+for all land cover types, particularly savanna, shrubland, and
+croplands. The anomaly in soil moisture over the past 12 months (SSI-12)
+is the main variable explaining these changes, followed by anomalies in
+cumulated precipitation over one to two years (SPI-12 and SPI-24). Our
+results provide insightful information that would help in developing
+adaptation measures for ecosystems in Chile to cope with climate change
+and drought.
 \end{abstract}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, breakable, frame hidden]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 proposed by \citet{Zambrano2018}, which was derived from the monthly
-time series of NDVI retrieved from MOD13A1. All the indices are
+time series of NDVI derived from MOD13A1. All the indices are
 multi-scalar and can be used for the analysis of short- to long-term
@@
 of six months at monthly frequency for 2000--2023. For zcNDVI, we test
-time scales of 1, 3, 6, and 12 months in December and its correlation
+time scales of 1, 3, 6, and 12 months in December and their correlation
 with net primary production (NPP) obtained from the MOD17A3HGF product
@@
 increases from one to six months and from six to 12 months has little
-improvement (see supplementary material in Section S5).
+improvement. The r-squared were between 0.31 for forest and 0.72 for
+shrubland (see supplementary material in Section S5).
 
@@
 
-\hypertarget{future-outlook-to-complete}{%
-\subsection{Future outlook (to
-complete)}\label{future-outlook-to-complete}}
+\hypertarget{drought-information-to-aid-in-adaptation}{%
+\subsection{Drought information to aid in
+adaptation}\label{drought-information-to-aid-in-adaptation}}
+
+Different climate components, such as evaporative water demand, water
+supply, soil moisture, and their impact on vegetation, should be
+considered when evaluating the multi-dimensional nature of drought. For
+a better understanding of the propagation of drought \citep{VanLoon2012}
+from meteorological to hydrological drought, we should consider the
+climatic response at different time scales, ranging from short to long.
+Furthermore, we must make this information publicly available and update
+it frequently to aid in the development of adaptation policies. The
+drought observatory for agriculture and biodiversity of Chile (ODES)
+shares this information aggregated for multiple administrative and
+hydrological units across Chile
+(\url{https://odes-chile.org/app/unidades}), with the goal of helping to
+prepare for future climatic conditions.
 
@@
 
-There is a trend toward decreasing water supply in most parts of Chile,
-particularly in the ``Centro'' and ``Norte Chico'' regions. The whole
-country showed an increase in AED. Vegetation productivity only showed a
-decrease in the ``Norte Chico'' and ``Centro,'' being highest for
-shrubland and croplands. Forest is the land cover most resistant to
-drought, as shown along Chile, and shrubland and cropland are the most
-sensitive.
-
-A soil moisture deficit of 12 months (SSI-12) is highly correlated with
-vegetation productivity for the land cover classes of shrubland,
-savannas, croplands, and forest in ``Norte Chico'' and ``Centro.'' For
-the southern part of the country with humid conditions, the correlation
-with SSI decreases. Soil moisture overcomes the capacity to explain
-vegetation productivity over the supply and demand drought indices in
-the entire territory.
-
-The variation in vegetation productivity appears to be associated with
-climate variation rather than anthropogenic factors (e.g., an increase
-in water demand by irrigated crops). Even though switching to more
-demanding crops on the land could increase the impact of drought on
-vegetation, this would need to be more thoroughly investigated, for
-instance at the watershed level.
-
-The results of this study could help in the development of a robust
-forecasting system for land cover classes in Chile, helping to improve
-preparedness for climate change impacts on vegetation.
+There is a trend toward decreasing water supply (SPI, SPEI, and SSI) in
+most of Chile; just in the southern part, there is an increase. The
+trend is most strong in the north-central zone. The whole country showed
+an increase in water demand (AED). The magnitude of the trend increases
+over longer time scales, which is evidence that the deficit is impacting
+the hydrological system in Chile. The trend in vegetation productivity
+in the north-central area is affecting, to a higher degree, shrubland
+and savanna, followed by croplands and forests. The most important
+changes in land cover are the increase of forest in ``Sur'' and
+shrubland and grassland in ``Centro;'' and the increase of savanna in
+``Centro'' and ``Sur.''.
+
+The drought explains about 30\% of the change in land cover type across
+Chile for forest, grassland, shrubland, and savanna. There is no
+evidence of an effect of drought on the change in cropland surface area.
+The increase in AED is the main driver of the change in land cover,
+followed by a reduction in precipitation and soil moisture. The drought
+time scales vary regarding the land cover type.
+
+The change in vegetation productivity has been severe in the
+north-central part of the country for all land cover types, particularly
+savanna, shrubland, and croplands. The anomaly in soil moisture over the
+past 12 months is the main variable explaining these changes, followed
+by anomalies in cumulated precipitation over one to two years. The
+variation in AED seems to intensify the drought impact on vegetation
+productivity.
+
+The results of this study provide insightful information that would help
+in developing adaptation measures for ecosystems in Chile to cope with
+climate change and drought.
+
+\hypertarget{acknowledgment}{%
+\section{Acknowledgment}\label{acknowledgment}}
+
+The National Research and Development Agency of Chile (ANID) funded this
+study through the drought emergency project FSEQ210022, Fondecyt
+Iniciación N°11190360, and Fondecyt Regular N°1210526.
 

### 2024-04-04 scripts/23_analysis_trend_drought_indices_and_landcover_macroXlandcover.R

@@ -13,6 +13,11 @@ data_ana <- data_trend_di |>
   full_join(data_trend_lc) |> 
   drop_na()
 
+paleta <- read_csv('data/processed_data/paleta_colores_landcover.csv')
+colors <-  rgb(paleta$R,paleta$G,paleta$B,maxColorValue = 255)
+attr(colors,'names') <- paleta$Name
+colors[3] <- 'red'
+
 #cargar la superficie de landcover ṕor macro class para relativizar las superficies
 
 lc_surf <- read_rds('data/processed_data/surface_landcover_macrozone.rds') |> 
@@ -51,13 +56,13 @@ data_ana |>
                                 'SPI-36','EDDI-36','zcSM-36'),
                     labels = c('SPI-6','EDDI-6','SSI-6',
                                'SPI-36','EDDI-36','SSI-36'))) |> 
-  ggplot(aes(value*10,landcover,color = macro,shape = class)) +
+  ggplot(aes(value*10,landcover,color = class,shape = macro)) +
   geom_point() + 
   geom_hline(yintercept = 0,color = 'red',linetype = 'dashed',alpha = .6) +
   geom_vline(xintercept = 0,color = 'red',linetype = 'dashed',alpha = .6) +
   labs(x = 'Trend of drought index (per decade)',
        y = 'Relative trend of land cover change') +
-  scale_color_viridis_d('Macrozone') +
+  scale_color_manual('class',values = colors,name = 'Land cover') +
   #scale_x_sqrt() +
   scale_shape('Land cover') +
   facet_wrap(.~name,ncol=3,nrow=2) +

### 2024-04-04 scripts/39_crear_gt_table_landcover_IGBP_macroclass.R

@@ -1,0 +1,10 @@
+library(gt)
+library(readr)
+
+tabla <- read_csv('data/raw_data/tabla_landcover.csv')
+
+tabla |> 
+  gt() |> 
+  cols_width(Name~100,
+             Description~px(400)) |> 
+  gtsave('output/figs/tabla_landcover_macroclass.png')

### 2024-04-11 article/paper_drought_Chile2023.qmd
last update antes egu2024

--- 2024-03-20 article/paper_drought_Chile2023.qmd
@@
 ---
-title: The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile
+title: The effects of drought on land cover change and vegetation productivity in continental Chile
 #subtitle: A Short Subtitle
@@
         - id: umayor
-          name: Universidad Mayor
-          department: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad
-          #address: Street Address
-          city: Santiago, Chile
-          #state: State
-          postal-code: 7500994
+          name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad, Universidad Mayor
+          city: Santiago
+          country: 7500994, Chile
+        - id: odes
+          name: Observatorio de Sequía para la Agricultura y la Biodiversidad de Chile (ODES), Universidad Mayor
+          city: Santiago
+          country: 7500994, Chile
     attributes:
         corresponding: true
-    note: This is the first author footnote.
+    #note: This is the first author footnote.
   - name: Anton Vrieling
@@
         - id: itc
-          name: University of Twente
-          department: Faculty of Geo-Information Science and Earth Observation (ITC)
-          #address: Street Address
+          name: Faculty of Geo-Information Science and Earth, University of Twente
           city:  Enschede, The Netherlands
   - name: Francisco Meza
+    affiliations:
+        - id: agrouc
+          name: Facultad de Agronomía y Sistemas Naturales, Pontificia Universidad Católica de Chile. 
+          city: Santiago
+          country: Chile
+        - id: instuc
+          name: Instituto para el Desarrollo Sustentable. Pontificia Universidad Católica de Chile
+          city: Santiago
+          country: Chile
+        - id: cambioglobal
+          name: Centro Interdisciplinario de Cambio Global, Pontificia Universidad Católica de Chile, Santiago, Chile.
   - name: Iongel Duran-Llacer
+    affiliations:
+        - id: umayor-io
+          name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Universidad Mayor,
+          city: Santiago
+          country: 7500994, Chile
   - name: Francisco Fernández
@@
         - id: uss
-          name: Universidad San Sebastian
-          department: 
-          #address: Street Address
-          city:  
+          name: Center of Economics for Sustainable Development (CEDES), Faculty of Economics and Government, Universidad San Sebastian
+          city: Santiago
+          country: Chile
+        - id: capes
+          name: Center of Applied Ecology and Sustainability (CAPES)
+          city: Santiago
+          country: Chile
   - name: Alejandro Venegas-González
   - name: Nicolas Raab
+    affiliations:
+        - id: agrouc
   - name: Dylan Craven
+    affiliations:
+        - id: gema
+          department: GEMA Center for Genomics, Ecology & Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750, Santiago, Chile
+        - id: do
+          name: Data Observatory Foundation, Santiago, Chile
   #   note: |
@@
 abstract: |
- The north-central region of Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development.
-This persistent period of water scarcity has been defined as a megadrought.
-The aim of our study is to evaluate the land cover change over continental Chile and to examine how this is connected to drought indices of water supply, atmospheric evaporative demand (AED), soil moisture, and their effects on vegetation productivity.
-The drought indices were derived using monthly ERA5-Land reanalysis data spanning from 1981 to 2023.
-The Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets were utilized to obtain information on annual land cover and monthly vegetation productivity.
-We analyzed short- (1, 3, 6 months) to long-term (12, 24, 36 months) time scales of drought.
-Our results showed that land cover change was highest in the south-central part of the country, reaching changes as high as 36% in the surface type.
-The water demand has increased for the whole country, with a major increase in the north.
-The AED and soil moisture evidence a decreasing trend, which decreases at longer time scales and from north to south.
-The extreme south part of the country shows an increase in supply.
-Vegetation productivity has a negative trend in the north-central region for all land cover types.
-On the other hand, forests seem to be the most resistant type to drought.
-The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
-The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are the ones that consider soil moisture for twelve months and the combined effect of precipitation and AED for 24 and 12 months.
-The results indicate that the north-central region is the most sensitive to water supply deficits lasting longer than a year.
+ Chile has experienced a persistent decrease in water supply, which impacts the hydrological system and vegetation development.
+This persistent period of water scarcity has been defined as a mega-drought.
+There are few studies about the relationship between drought and ecosystem changes that can help with a better understanding of ecological drought.
+The aim of our study is to evaluate the interaction of drought, land cover change, and vegetation productivity over continental Chile.
+To assess drought, we used drought indices for atmospheric evaporative demand (AED), water supply, and soil moisture from short- (1, 3, 6 months) to long-term (12, 24, 36 months) time scales.
+We derived the drought indices using monthly ERA5-Land reanalysis data from 1981 to 2023.
+We used the Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets to derive information on annual land cover and monthly vegetation productivity.
+Our results showed that except for the Austral part, Chile has a temporal decreasing trend in water supply, and across the whole country, there is an increase in AED.
+These trends become stronger over longer time scales.
+We found a negative trend in vegetation productivity in the north-central area, which is more prominent for shrubland and savanna as compared to croplands and forests.
+The anomaly in soil moisture over the past 12 months (SSI-12) is the most important variable explaining these changes, followed by anomalies in accumulated precipitation over one to two years (SPI-12 and SPI-24).
+The variable importance obtained by random forest models indicates that drought is explaining about 20–30% of the change in land cover surface across Chile for forest, grassland, shrubland, and savanna but has no relation to the changes in croplands.
+The increase in AED is the main variable associated with the change in land cover, followed by a reduction in precipitation and soil moisture.
+Our findings provide insightful information that could assist in developing adaptation measures for Chilean ecosystems to cope with climate change and drought.
+Also, this study could contribute to a better comprehension of ecological drought.
+
 keywords: 
@@
   - vegetation productivity
-  - satellite
+  - ecosystem
 date: last-modified
@@
 
-Drought is often classified as meteorological when there is a decrease in precipitation below the mean average of several years (more than 30 years), hydrological when these anomalies last for long periods (months to years) and affect water systems, and agricultural when the deficit impacts plant health anomalies and leads to decreased productivity [@Wilhite1985].
-However, it is important to note that drought is also influenced by human activities, which were not considered in the definitions.
-Thus, @Loon2016 and @AghaKouchak2021 have given an updated definition of drought for the Anthropocene, suggesting that it should be considered the feedback of humans’ decisions and activities that drives the anthropogenic drought.
-Simultaneously, drought leads to heightened tree mortality and induces alterations in land cover and land use, ultimately affecting ecosystems [@Crausbay2017].
-Even though many ecological studies have misinterpreted how to characterize drought, for example, sometimes considering "dry" conditions as "drought" [@Slette2019].
-Then, @Crausbay2017 proposed the ecological drought definition as "an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems." In light of current global warming, it is crucial to study the interaction between drought and ecosystems in order to understand their feedback and impact on water security.
-[@Bakker2012]
+Drought is often classified as 1) meteorological when precipitation in a specific period remains below the mean precipitation experienced in the same period during multiple years (more than 30 years usually), 2) hydrological when these anomalies last for long periods (months to years) and affect water systems, and 3) agricultural when the deficit negatively impacts plant health and leads to decreased productivity of crops or pastures [@Wilhite1985].
+However, because drought is also influenced by human activities,  @Loon2016 and @AghaKouchak2021 expanded the drought definition for the Anthropocene, indicating that the feedback of human decisions and activities should also be considered (i.e.,  anthropogenic drought).
+Droughts lead to increased tree mortality [@Cheng2024] and induces alterations in land cover and land use, ultimately affecting ecosystems [@Crausbay2017].
+Even though many ecological studies have at times mistakenly considered "dry" conditions as "drought" [@Slette2019].
+Ecological drought can be defined as _"an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems"_ [@Crausbay2017].
+In light of current global warming, it is crucial to study the interaction between drought and ecosystems in order to understand their feedback and impact on future water security [@Bakker2012].
 
-Human-induced greenhouse gas emissions have increased the frequency and/or intensity of drought as a result of global warming, according to the sixth assessment report (AR6) of the Intergovernmental Panel on Climate Change (IPCC) [@IPCC2023].
+Global warming, as a result of human-induced greenhouse gas emissions, has increased the frequency and intensity of drought, according to the sixth assessment report (AR6) of the Intergovernmental Panel on Climate Change (IPCC) [@IPCC2023].
 The evidence supporting this claim has been strengthened since AR5 [@IPCC2013].
-Recent studies, however, have produced contrasting findings, suggesting that drought has not exhibited a significant trend over the past forty years.
-[@Vicente-Serrano2022; @Kogan2020].
-@Vicente-Serrano2022 analyzed the meteorological drought trend on a global scale, finding that only in a few regions has there been an increase in the severity of drought.
-Moreover, they attribute the increase in droughts over the past forty years solely to an increase in atmospheric evaporative demand (AED), which in turn enhances vegetation water demand, with important implications for agricultural and ecological droughts.
-Also, they state that "the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change".
-Similarly, @Kogan2020 analyzed the drought trend using vegetation health methods, finding that for the globe, hemispheres, and main grain-producing countries, drought has not expanded or intensified for the last 38 years.
-Further, the @IPCC2021 suggests that there is a high degree of confidence that rising temperatures will increase the extent, frequency, and severity of droughts.
+Recent studies, however, have produced contrasting findings, with some suggesting that drought has not exhibited a significant trend over the past forty years [@Vicente-Serrano2022; @Kogan2020].
+@Vicente-Serrano2022 analyzed the trend in meteorological drought on a global scale, finding that only in a few regions an increase in the severity of drought was observed.
+Moreover, they attributed this increase solely to an increase in atmospheric evaporative demand (AED) due to higher temperatures, which in turn enhances vegetation water demand, with important implications for agricultural and ecological droughts.
+Also, they state that _"the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change, led by an increase in agricultural land"_.
+Similarly, @Kogan2020 analyzed the drought trend using remotely-sensed vegetation health indicators, finding that for the globe and main grain-producing countries, drought has not expanded or intensified during the past 38 years.
+Nonetheless, @IPCC2021 suggests that there is a medium to high degree of confidence that rising temperatures will increase the extent, frequency, and severity of agricultural and ecological droughts.
 Also, AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C.
-To better evaluate the impact of drought trends on ecosystems, assessments are needed that relate meteorological and soil moisture variables to their effects on vegetation.
+To better evaluate the impact of drought trends on ecosystems, assessments that correlate meteorological and soil moisture variables to their effects on vegetation are much needed.
 
-From 1960 to 2019, land use change has impacted around one-third of the Earth's surface, which is four times more than previously thought [@Winkler2021].
-Multiple studies aim to analyze and forecast changes in land cover globally [@Winkler2021; @Song2018] and regionally [@Chamling2020; @Homer2020; @Yang2021].
-Some others seek to analyze the impact of land cover change on climate conditions such as temperature and precipitation [@Luyssaert2014;@Pitman2012].
-There is less research on the interaction between drought and land cover change [@Chen2022;@Akinyemi2021;@Peng2017].
-@Peng2017 conducted a worldwide investigation utilizing net primary production to examine the spatial and temporal variations in vegetation productivity at global level.
-The study aimed to assess the influence of drought by comparing the twelve-month Standardized Precipitation Evapotranspiration Index (SPEI) and land cover change.
-According to their findings, drought is responsible for 37% of the decline in vegetation productivity, while water availability accounts for 55% of the variation.
-@Chen2022 studied the trend of vegetation greenness and productivity and its relation to meteorological drought (SPEI of twelve months in December) and soil moisture at the global level.
-The results showed lower correlations (<0.2) for both variables.
-@Akinyemi2021 evaluates drought trends and land cover change using vegetation indices in Botwsana in a semi-arid climate.
-These studies mostly looked at how changes in land cover and vegetation productivity are related to a single drought index (SPEI) over a single time period of 12 months.
-SPEI takes into account the combined effect of precipitation and AED as a water balance, but it does not allow us to know the contribution of each variable on its own.
-Some things worth investigating in terms of land cover change and vegetation productivity are: i) How do they respond to short- to long-term meteorological and soil moisture droughts?
-ii) How is the drought impacting land cover changes?
-And iii) How do they behave in humid and arid climatic zones regarding drought?
+From 1960 to 2019, land use change has impacted around one-third of the Earth’s surface, which is four times more than previously thought [@Winkler2021].
+Multiple studies aim to analyze and forecast changes in land cover globally [@Winkler2021;@Song2018] and regionally [@Chamling2020;@Homer2020;@Yang2021;@Schulz2010;@Echeverria2012].
+Some seek to analyze the impact of land cover change on climate conditions such as temperature and precipitation [@Luyssaert2014;@Pitman2012].
+There is less research on drought and its relation to land cover change and vegetation productivity [@Chen2022;@Akinyemi2021;@Peng2017].
+@Peng2017 utilized net primary productivity to examine the spatial and temporal variations in vegetation productivity at global level and assess to what extent drought influenced this variability by comparing the twelve-month Standardized Precipitation Evapotranspiration Index (SPEI) and land cover change.
+According to their findings, drought is responsible for 37% of the decline and accounts for 55% of the variability in vegetation productivity.
+@Chen2022 instead found poor correlations (r<0.2) between the vegetation productivity trends against meteorological drought (SPEI of twelve months in December) and soil moisture at the global level.
+These studies mostly looked at how changes in land cover and vegetation productivity are related to a single drought index (SPEI) obtained for 12 month periods.
+SPEI takes into account the combined effect of precipitation and AED as a water balance, but it does not allow to know the contribution of each variable on its own.
+To better understand these contributions on land cover change and vegetation productivity the following questions may be asked: i) how do land cover and vegetation productivity respond to short- to long-term meteorological and soil moisture droughts?
+And ii) How is this response different between humid and arid climatic zones?
 Likewise, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.
 
+To address the previous questions over extensive regions, we can utilize gridded data on water availability, vegetation conditions, and the respective drought indices.
 For monitoring drought, the World Meteorological Organization recommends the SPI (Standardized Precipitation Index) [@WMO2012].
 The SPI is a multi-scalar drought index that only uses precipitation to assess short- to long-term droughts.
-The primary cause of drought is precipitation anomalies, and temperature usually makes it worse [@Luo2017].
-Nowadays, there is an increase in attention toward using AED separately to monitor droughts [@Vicente-Serrano2020].
-One reason is due to its attribution to increasing flash droughts in water-limited regions [@Noguera2022].
-@Vicente-Serrano2010 proposed the Standardized Precipitation Evapotranspiration Index (SPEI), which incorporated the temperature effect by subtracting AED from precipitation.
-SPEI allows for analysis of the combined effect of precipitation and AED.
-Since its formulation, it has been used worldwide for the study and monitoring of drought [@Gebrechorkos2023;@Liu2021].
+@Vicente-Serrano2010 proposed the Standardized Precipitation Evapotranspiration Index (SPEI), which incorporates the temperature effect by subtracting AED from precipitation.
+SPEI allows for analyzing the combined effect of precipitation and AED.
+Since its formulation, it has been used worldwide for the study and monitoring of drought [@Gebrechorkos2023;@Liu2024].
+Recently, there has been more interest in using AED to track droughts separately to better disengage the effects of precipitation from temperature-dependent effects [@Vicente-Serrano2020].
+One of the reasons is that AED is linked to more flash droughts in limited water regions [@Noguera2022].
 @Hobbins2016 and @McEvoy2016 developed the Evaporative Demand Drought Index (EDDI) to monitor droughts solely using the AED, and it has proven effective in monitoring flash droughts [@Li2024;@Ford2023].
@@
 @Hao2013 and @AghaKouchak2014 proposed the Standardized Soil Moisture Index (SSI), which has a similar formulation as the SPI, SPEI, and EDDI.
-Thus, there are plenty of drought indices that allow for a comprehensive assessment of drought on short- to long-term scales and that allow for the use of single variables from the earth's water balance (e.g., precipitation, AED, soil moisture).
-The variation in climate variables impacts vegetation development, and unfavorable conditions such as low precipitation and high temperatures usually generate a decrease in vegetation productivity.
-To monitor the response of vegetation, the common practice is to use satellite data.
-The Normalized Difference Vegetation Index (NDVI) has been widely used as a proxy for biomass production [@Camps-Valls2021;@Paruelo2016;@Helman2014].
-For Chile's cultivated land, @Zambrano2018 introduced the zcNDVI for assessing seasonal biomass production in response to drought.
-Using this information, we can advance our understanding of the impact of drought on ecosystems.
+Thus, many drought indices exist that allow for a comprehensive assessment of drought on short- to long-term scales and that allow for the use of single variables from Earth’s water balance (e.g., precipitation, AED, soil moisture).
+Climatic variability impacts vegetation development, with unfavorable conditions such as low precipitation and high temperatures usually promote a decrease in plant productivity.
+To monitor the response of vegetation for large areas, the common practice is to use satellite data.
+For example, the Normalized Difference Vegetation Index (NDVI) derived from frequent satellite observations of red and near infrared spectral reflectance, has been widely used as a proxy for biomass production [@Camps-Valls2021; @Paruelo2016;@Helman2014].
+For Chile's cultivated land, @Zambrano2018 used the zcNDVI for assessing seasonal biomass production in response to drought.
+Comparing the various meteo-related and vegetation-based drought indices, we can further our understanding of the impact of drought on ecosystems.
 
@@
 Additionally, the country has experienced severe drought conditions that have had significant effects on vegetation and water storage.
-North-central Chile has faced a persistent precipitation deficit since 2010, defined as a mega drought.
-[@Garreaud2017], which has impacted the Chilean ecosystem.
-This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December having values below one standard deviation.
-Some studies have addressed how this drought affects single ecosystems in terms of forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018;@Zambrano2016].
-We found one study regarding land cover and drought in Chile.
-The study by @Fuentes2021  evaluates water scarcity and land cover change in Chile between 29° and 39° of south latitude.
-@Fuentes2021 used the SPEI of one month for evaluating drought, which led to misleading results.
-For example, they did not find a temporal trend in the SPEI but found a decreasing trend in water availability and an increase trend on AED, which in turn should have been capable of being captured with longer time scales of the SPEI.
-The term "megadrought" in Chile is used to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018].
+North-central Chile has faced a persistent precipitation deficit since 2010, defined as a mega-drought [@Garreaud2017], which has impacted the Chilean ecosystem and consequently makes it highly vulnerable to climate change [@Barria2021;@AlvarezGarreton2021].
+This mega-drought was defined by the annual time series of the Standardized Precipitation Index (SPI) at a time scale of twelve months at the end of each year (December) when having values below one standard deviation.
+Some studies have addressed how this drought affects single ecosystems in terms of forest growth [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018;@Zambrano2016].
+The term "mega-drought" is used in Chile to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018].
 Therefore, it is crucial to evaluate temporal scales that consider the cumulative impact over a period of several years.
-The association between drought and the environment in Chile is not well comprehended.
-Hence, it is imperative to acquire a more profound comprehension of the manner in which climatic and soil moisture droughts influence environmental dynamics, in order to make well-informed decisions on adaptation strategies.
-
-<!-- A detailed spatiotemporal assessment of the interaction of drought for short- to long-term and land cover change requires information on vegetation as well as weather variables such as precipitation, temperature, and soil moisture.
-Weather networks in Chile present some disadvantages, such as spatio-temporal gaps, a short history, and irregular quality, which make them difficult to represent the whole extent of the country spatially.
-In order to do this, we use reanalysis data from ERA5-Land [@MunozSabater2021] to create drought indices that consider AED, precipitation, and soil moisture over a range of time periods, from the short to the long term.
-Also, we use vegetation spectral information and annual land cover change from the Moderate-Resolution Imaging Spectroradiometer (MODIS).
-We expect to gain insight regarding the temporal evolution of water demand, water supply, and soil moisture, as well as the interaction with land cover change and vegetation productivity.
--->
+In Chile, the relationship between drought and the environment remains poorly understood.
+Hence, we aim to contribute to understanding how climatic and soil moisture droughts influence ecosystem dynamics in order to provide useful information that helps for a better understanding of ecological droughts and, at the same time, helps to make well-informed decisions on adaptation strategies.
 
@@
 
-Continental Chile has diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] (@fig-studyArea a), which determines its great ecosystem diversity [@Luebert2022] (@fig-studyArea c).
-The Andes Mountains are a main factor in climate latitudinal variation [@Garreaud2009].
-"Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures.
+Continental Chile has diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] ([@fig-studyArea]a), which determines its great ecosystem diversity [@Luebert2022] ([@fig-studyArea]c).
+The Andes Mountains are a main factor in climate variation [@Garreaud2009].
+For an aggregated overview of the results of the study, we used the five Chilean macrozones: "Norte Grande," "Norte Chico," "Centro," "Sur," and "Asutral".
+"Norte Grande" (17°34’–25°42’S) and "Norte Chico" (25°42'-32°8'S) predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures.
 At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk).
 In these two northern regions, the land is mostly bare, with a small surface of vegetation types such as shrubland and grassland.
-In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warm to hot summers (Csa and Csb).
+In the macrozones "Centro" (32°08'-36°12'S) and the northern half of "Sur," (36°12'-43°48'S) the main climate is Mediterranean, with warm to hot summers (Csa and Csb).
 Land cover in "Centro" comprises a significant amount of shrubland and savanna (50%), grassland (16%), forest (8%), and croplands (5%).
-An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral." Those zones are high in forest and grassland.
+An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral" (43°48'-56°00'S) Those zones have a large areal extent of forest and grassland.
 The southern part of the country has a tundra climate, and in "Austral," it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.
@@
 
-To analyze land cover change, we use the classification scheme by the IGBP (International Geosphere-Biosphere Programme) from the product MCD12Q1 collection 6.1 from MODIS.
-The MCD12Q1 has a yearly frequency from 2001 to 2022 and defines 17 classes.
-To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the product MOD13A3 collection 6.1 from MODIS [@Didan2015].
-MOD13A3 provides vegetation indices at 1km of spatial resolution and monthly frequency.
+To analyze land cover change, we used the classification scheme by the IGBP (International Geosphere-Biosphere Programme) from the product MCD12Q1 Collection 6.1 from MODIS.
+The MCD12Q1 product is produced for each year from 2001 to 2022 and defines 17 classes (see Table S1).
+To maintain our focus on a large scale and follow the FAO classification [@FAO2022], we considered native and planted forests as "forests", which represent ecosystems dominated by larger trees.
+To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the product MOD13A3 Collection 6.1 from MODIS [@Didan2015].
+MOD13A3 provides vegetation indices with 1km spatial resolution and monthly frequency.
 The NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and MCD12Q1 from the online Data Pool, accessible at https://lpdaa.usgs.gov/tools/data-pool/.
@@
 \label{tab-desEOD}
-\small
+\scriptsize
 \centering
-\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
+\begin{tabular}{p{0.1\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
 \hline
@@
 It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage.
-We selected the variables for total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).
-Table \ref{tab-desEOD} shows a summary of the data and its main characteristics.
+We selected the variables for total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100 cm of depth (layer 1 to layer 3).
+\ref{tab-desEOD} shows a summary of the data and its main characteristics.
 
-## Short- to long-term drought
+## Short- to long-term drought trends
 
@@
 
-In order to compute the drought indices that use water demand, it is necessary to first calculate the AED.
+To compute the drought indices that use water demand, it is necessary to first calculate the AED.
 To do this, we employed the Hargreaves method [@Hargreaves1994;@Hargreaves1985] by applying the following equation:
@@
 where $Ra$ $(MJ\,m^2\, day^{-1})$ is extraterrestrial radiation; $T$, $T_{max}$, and $T_{min}$ are mean, maximum, and minimum temperature $(°C)$ at 2m.
-For calculating $Ra$ we used the coordinate of the latitud of the centroid of each pixel.
+For calculating $Ra$ we used the coordinate of the latitud of the centroid of each pixel as follow:
+
+$$R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]$$ {#eq-Ra}
+where
+
+$Ra$: extraterrestrial radiation $[MJ\, m^{-2} day-1]$,  
+$G_{sc}$: solar constant = 0.0820 $[MJ\,m^{-2} min^{-1}]$,    
+$d_r$: inverse relative distance Earth-Sun,  
+$\omega_s$ sunset hour angle  $[rad]$,  
+$\phi$: latitude $[rad]$,  
+$\delta$: solar declination $[rad]$.
+
 We chose the method of Hargreaves to estimate AED because of its simplicity, which only requires temperatures and extrarrestrial radiation.
@@
 To derive the drought indices of water supply and demand, soil moisture, and vegetation (i.e., the proxy of productivity), we used the ERA5L dataset and the MODIS product, with a monthly frequency for 1981–2023 and 2000–2023, respectively.
-The dought indices correspond to a historical anomaly with regard to a variable (e.g., meteorological, vegetation, or soil moisture).
-To account for the anomaly, the common practice is to derive it following a statistical parametric methodology in which it is assumed that the statistical distribution of the data is known [@Heim2002].
-A wrong decision is usually the highest source of uncertainty [@Laimighofer2022].
-In the case of Chile, due to its high degree of climatic variability, it is complex to choose a proper distribution without previous research.
-Here, we follow a non-parametric methodology for the calculation of the drought indices, in a similar manner as the framework proposed by @Farahmand2015; @Hobbins2016;@McEvoy2016.
+The drought indices correspond to a historical anomaly of a variable (e.g., meteorological, vegetation, or soil moisture).
+To account for the anomaly, the common practice is to derive it following a statistical parametric method in which it is assumed that the statistical distribution of the data is known [@Heim2002].
+A wrong decision in the statistical distribution is usually the highest source of uncertainty [@Laimighofer2022].
+In the case of Chile, due to its high degree of climatic variability, it is difficult to choose a proper distribution without previous research that could be applicable throughout Chile.
+Here, we follow a non-parametric method for the calculation of the drought indices, in a similar manner as the framework proposed by @Farahmand2015.
 
@@
 To consider the combined effect of water supply and demand, we selected the SPEI [@Vicente-Serrano2010].
-For SPEI, an auxiliary variable $D = P-AED$ is calculated.
+For SPEI, an auxiliary variable $D=P-AED$  is calculated.
 Soil moisture is the main driver of vegetation productivity, particularly in semi-arid regions [@Li2022].
-Hence, for soil water drought, we used the SSI (Standardized Soil Moisture Index) [@Hao2013;@AghaKouchak2014].
-In our case, for the SSI, we used the average soil moisture from ERA5L at 1m depth.
-Finally, for the proxy of productivity, we used the zcNDVI proposed by @Zambrano2018, which was derived from the monthly time series of NDVI retrieved from MOD13A1.
+Hence, for soil water drought, we used the SSI (Standardized Soil Moisture Index) [@Hao2013].
+For the SSI, we used the average soil moisture from ERA5L at 1m depth.
+Finally, for the proxy of productivity, we used the zcNDVI [@Zambrano2018], which was derived from the monthly time series of NDVI derived from MOD13A1.
 All the indices are multi-scalar and can be used for the analysis of short- to long-term droughts.
 
-To derive the drought indices, first we must calculate the sum of the variables with regard to the time scale (s).
+To derive the drought indices, we first calculate the sum of the variables with regard to the time scale(s).
 In this case, for generalization purposes, we will use $V$, referring to variables $P$, $AED$, $D$, $NDVI$, and $SM$ (Table \ref{tab-desEOD}).
-We cumulated each $V$  over the time series of $n$ values (months), and for the time scales $s$:
+We accumulated each  over the time series of  values (months), and for the time scales $s$:
 
-$$A_{si} = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  $$ {#eq-sumvar}
+$$A^s_i = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  $$ {#eq-sumvar}
 
-The $A_{si}$ corresponds to a moving window (convolution) that sums the variable for time scales $s$ from the last month, month by month, until the first month in which it could sum for $s$ months.
-An inverse normal approximation [@Abramowitz1968] obtains the empirically derived probabilities once the variable cumulates over time for the scale $s$.
+The $A^s_i$ corresponds to a moving window (convolution) that sums the variable for time scales $s$.
+Start from the last month (n) and sum the variable for s months, then follow month by month until the first month in which it could sum for s months (n-s+1).
+For example, using as a variable the precipitation, a period of twelve months (n), and a time scale of three months (s), it will be:
+
+$$
+\begin{split}
+A^3_1 &= P_{oct} +P_{nov} +P_{dic} \\
+\vdots\,\,\, &= \,\,\,\vdots + \,\,\,\vdots + \,\,\,\vdots \\
+A^3_{10} &= P_{jan}+P_{feb} +P_{mar}
+\end{split}
+$$
+
+Then, an inverse normal approximation [@Abramowitz1968] obtains the empirically derived probabilities once the variable cumulates over time for the scale $s$.
 Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i$ to derive the $P(a_i)$ probabilities across a period of interest:
 
-$$P(A_i) = \frac{i-0.33}{n+0.33'}$$ {#eq-probPai}
+$$P(A^s_i) = \frac{i-0.33}{n+0.33'}$$ {#eq-probPai}
 
@@
 
-$$DI(A_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}$$ {#eq-DI}
+$$DI(A^s_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}$$ {#eq-DI}
 
@@
 The values for the constats are: $C_0 = 2.515517$, $C_1 = 0.802853$, $C_2 = 0.010328$, $d_1 = 1.432788$, $d_2 = 0.189269$, and $d3 = 0.001308$.
-For $P(A) \leq 0.5$, W=$\sqrt{-2\cdot ln(P(A_i))}$ , and for $P(A_i) > 0.5$, replace $P(A_i)$ with $1-P(A_i)$ and reverse the sign of $DI(A_i)$.
+For $P(A^s_i) \leq 0.5$, W=$\sqrt{-2\cdot ln(P(A^s_i))}$ , and for $P(A^s_i) > 0.5$, replace $P(A^s_i)$ with $1-P(A^s_i)$ and reverse the sign of $DI(A^s_i)$.
 
 The drought indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months at a monthly frequency for 1981–2023 in order to be used for short- to long-term evaluation of drought.
-In the case of the proxy of vegetation productivity (zcNDVI) it was calculated for a time scale of six months at monthly frequency for 2000–2023.
-For zcNDVI, we test time scales of 1, 3, 6, and 12 months; we choose to use six months because that shows a more stable representation of vegetation productivity.
+
+For the proxy of vegetation productivity, we chose the time scale that best correlates with annual net primary productivity (NPP) across continental Chile.
+For this purpose, we calculated the zcNDVI for time scales of 1, 3, 6, and 12 months in December and compared it with the annual NPP.
+We used the NPP from the MOD17A3HGF [@Running2019] dataset (MODIS).
+We choose to use six months because the r-squared of zcNDVI with NPP highly increases from one to six months, but from six to 12 months has a lower improvement.
+We obtained an r-squared of 0.31 for forest and 0.72 for shrubland (refer to the supplementary material in Section S5).
+Then, we chose the proxy of vegetation productivity for six months, which we will name zcNDVI hereafter.
+It was calculated at a monthly frequency for 2000–2023.
 
@@
 To determine the magnitude of the trend, we used Sen’s slope [@Sen1968].
-Some of the advantages of applying this methodology are that the Sen’s slope is not affected by outliers as regular regression does, and it is a non-parametric method that is not influenced by the distribution of the data.
-We applied Mann-Kendall to see if the trend was significant and Sen's slope to estimate the magnitude of the trend.
-We did this to the six time scales from 1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI.
-Thus, we have trends per index and time scale (24 in total).
-Then, we extracted the trend aggregated by macrozone and per land cover persitent macroclasses.
+Some of the advantages of applying this methodology are that the Sen's slope is not affected by outliers unlike regular regressions does, and it is a non-parametric method that is not influenced by the distribution of the data.
+We applied the Mann-Kendall test to see if the trend was significant and Sen's slope to estimate the magnitude of the trend.
+We did this for the indices SPI, EDDI, SPEI, and SSI using the six time scales with data from 1981 to 2023 (monthly frequency), resulting in 24 trends (per index and time scale).
+Then, we extracted the trend aggregated by each of the five macrozones: "Norte Grande" to "Asutral," and per land cover type: grassland, forest, cropland, shrubland, savanna, and barren land ([@fig-studyArea]d).
 
@@
 This product has been previously used for studies of drought and land cover in Chile [@Fuentes2021;@Zambrano2018].
-From the 17 classes, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
+We regrouped the 17 classes into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
 Thus, we have a land cover raster time series with the ten macroclasses for 2001 and 2023.
@@
 
-Later in this study, we will examine the variation in vegetation productivity across various land cover types and how water demand and supply, and soil moisture affect it.
-In order to avoid variations due to a change in the land cover type from year-to-year that will wrongly impact NDVI, we developed a persistence mask for land cover for 2001–2022.
-Thereby, we reduce an important source of variation on a regional scale.
-Therefore, we generated a raster mask for IGBP MODIS per pixel using macroclasses that remained unchanged for at least 80% of the years (2001–2022).
-This enabled us to identify regions where the land cover macroclasses are persistent.
+To assess how water demand and supply, and soil moisture affect the variation in vegetation productivity across various land cover types, we avoid analyzing areas that have major land cover changes for 2021–2022.
+To assess how zcNDVI varied irrespective of land cover change, we developed a persistence mask for land cover, which only retains pixels for which the macroclass remained the same for at least 80% of the years (2001–2022) ([@fig-studyArea]d).
 
@@
 
-We wanted to explore the relationship between the trend in land cover classes and the trend in the drought indices.
-For this purpose, in order to have more representative results, we conducted the analysis over sub-basins within continental Chile.
-We use 469 basins, which have a surface area between 0.0746 and  24,000 ($km^2$), and a median area of 1,249 ($km^2$).
-For each basin, we calculate the relative trend per land cover type, considering the proportion of the type relative to the total surface of the basin.
-Then, we extracted per basin the average trend of the drought indices SPI, SPEI, EDDI, SSI, and all their time scales 1, 3, 6, 12, 24, and 36.
-Also, we extracted the average trend in the proxy of vegetation productivity (zcNDVI).
-We wanted to analyze which drought indices and time scales have a major impact on changes in land cover type.
+The goal of this section is to identify which drought indices and time scales have a major impact on changes in land cover type.
+We examined the relationship between the trend in land cover classes and the trend in drought indices.
+To have more representative results, we conducted the analysis over sub-basins within continental Chile.
+We used 469 basins, which have a surface area between 0.0746 and 24,000 km2 and a median area of 1,249 km2.
+For each basin, we calculated the trend per land cover type, considering the proportion of the type relative to the total surface of the basin.
+Then, we extracted per basin the average trend (Sen's slope) of the drought indices SPI, SPEI, EDDI, SSI, and all their time scales 1, 3, 6, 12, 24, and 36.
+Also, we extracted the average trend in the proxy of vegetation productivity (zcNDVI)
 
-We have 25 predictors, including drought indices and vegetation productivity.
-We analyzed the 25 predictors per type of landcover, thus running six random forest models.
-Random forest uses multiple decision trees and allows for classification and regression.
-Some advantages are that it allows to find no linear relationship, reduces overfitting, and allows to derive the variable importance.
-We used random forests for regression and trained 1000 forests.
-To obtain more reliable results, we resampled by creating ten folds, running a random forest per fold, and calculating the r-squared (rsq), root mean square error (RMSE), and variable importance.
+Random forest [@Ho1995] employs multiple decision trees, allowing for classification and regression.
+Some advantages include the ability to find non-linear relationships, reduce overfitting, and derive variable importance.
+We used the regression random forest to model the trends in land cover per macroclass, using drought indices as predictors.
+This included the four drought indices per six time scales and the zcNDVI, totaling 25 predictors.
+As a result, we created six random forest models, one per trend in land cover macroclass.
+We trained 1000 forests in a resampled scheme to obtain more reliable results regarding variable importance.
+We resampled by creating ten folds, running a random forest per fold, and calculating the r-squared (rsq), root mean square error (RMSE), and variable importance.
+The variable importance helps for a better understanding of the relationships by finding which variable has a higher contribution to the model.
+Thus, we calculate the variable's importance by permuting out-of-bag (OOB) data per tree and computing the mean standard error in the OOB.
+After permuting each predictor variable, we repeated the process for the remaining variable.
+We repeated this process ten times (per fold) to obtain the performance metrics (rsq, RMSE, and variable importance).
 
-The variable importance helps for a better understanding of the relationships by finding which variable has a higher contribution to the model.
-We calculate the variable's importance by permuting out-of-bag (OOB) data per tree and computing the mean standard error in the OOB.
-After permuting each predictor variable, we repeat the process for the resting variable.
-We repeated this process ten times (per fold) to obtain the performance metrics (rsq, RMSE, and variable importance).
+Finally, we visually explore the connection between the SPI, EDDI, and SSI drought indices for short- and long-term changes in land cover.
+To do this, we compare the relative changes in land cover surface (in terms of the total surface area per land cover type and macrozone) with the drought indices of six (short-term) and thirty-six months (long-term).
+We created a scatterplot in which the x-axis shows drought indices trends and the y-axis shows land cover change trends.
 
@@
 
-We analyzed the trend of vegetation productivity over the unchanged land cover macroclasses.
-To achieve this, we used the persistent mask of land cover macroclasses.
-This way, we tried to reduce the noise in the vegetation due to a change in land cover from year to year.
+For each land cover macroclass, we analyzed the trend of vegetation productivity over the unchanged land cover macroclasses.
+To achieve this, we used the persistent mask of land cover macroclasses , thus reducing the possibility of evaluating productivity trends that are due to year-to-year variation in land cover.
 We used the zcNDVI as a proxy of vegetation productivity.
-In Chile's cultivated land, @Zambrano2018 introduced the zcNDVI for assessing seasonal biomass production in relation to climate.
+To assess productivity in Chile's cultivated land, Zambrano et al.
+(2018) used the zcNDVI for assessing seasonal biomass production in relation to climate.
 
 We examine the drought indices of water demand, water supply, and soil moisture and their correlation with vegetation productivity.
-The objective is to determine the impact of soil moisture and water demand and supply on vegetation productivity.
-We want to address three main questions: Which of the drought variables—supply, demand, or soil moisture—most helps to explain the changes in plant productivity?
+The objective is to determine to what extent soil moisture and water demand and supply affect vegetation productivity, thus addressing three main questions: 1) Which of the drought variables—supply, demand, or soil moisture—helps most in explaining the changes in vegetation productivity?
 How do the short- to long-term time scales of the drought variable affect vegetation productivity in Chile?
-And finally, how strong is the relationship between the variables and the drought index?
-Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five land cover types: forest, cropland, grassland, savanna, and shrubland.
+2) How strong is the relationship between the variables and the drought index?
+And finally, 3) how does the correlation vary per-land cover type?
+Answering these questions should advance our understanding of how climate is affecting vegetation, considering the impact on the five land cover types: forest, cropland, grassland, savanna, and shrubland.
 
 We conducted an analysis on the linear correlation between the indices SPI, SPEI, EDDI, and SSI over time periods of 1, 3, 6, 12, 24, and 36 months with zcNDVI.
-We used a method similar to that used by @Meroni2017 which compared the SPI time-scales with the cumulative FAPAR (Fraction of Absorbed Photosynthetically Active Radiation).
+We used a method similar to that used by @Meroni2017 which compared the SPI time-scales with the cumulative fAPAR (fraction of Absorbed Photosynthetically Active Radiation).
 We performed a pixel-to-pixel linear correlation analysis for each index within the persistent mask of land cover macroclasses.
@@
 
-## Short- to long-term drought
+## Short- to long-term drought trends
 
 @fig-trendDI shows the spatial variation of the trend for the drought indices from short- to long-term scales.
+The white space on the maps indicates a not significant trend.
 SPI and SPEI have a decreasing trend from "Norte Chico" to "Sur." However, there is an increasing trend in "Austral." The degree of the trend is stronger at higher time scales.
-The SSI indicates that in "Norte Grande," there are surfaces that have increased in the soutwest part and in the northeast have decreased, and is shown for all time scales.
+The SSI indicates that in "Norte Grande," there are surfaces that have increased in the southwest and in the northeast have decreased, and are shown for all time scales.
 Similar to SPI and SPEI, SSI decreases at higher time scales.
@@
 
-knitr::include_graphics("../output/figs/trend_raster_SPI_1981-2023.png")
 knitr::include_graphics("../output/figs/trend_raster_SPEI_1981-2023.png")
@@
 The macrozones that reached the lowest trend for SPI, SPEI, and SSI are "Norte Chico" and "Centro," where the indices also decrease at longer time scales.
-Potentially explained due to the prolonged reduction in precipitation that has affected the hydrological system in Chile.
+This may potentially be explained by the prolonged reduction in precipitation that has affected the hydrological system in Chile.
 At 36 months, it reaches trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and SSI.
@@
 In "Norte Grande" and "Norte Chico," which are in a semi-arid climate, it is evident that the EDDI has an effect on the difference between the SPI and SPEI index, which is not seen in the other macrozones.
-Contrary to the other macrozones, “Austral” showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend.
+Contrary to the other macrozones, "Austral" showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend.
 
@@
 \begin{table}[!ht]
-\caption{Surface of the land cover class that persist during 2001-2022}
+\caption{Surface per land cover class that persists during 2001–2022.}
 \label{tab-landcoverSurf}
@@
 #| fig-pos: "!ht"
-#| fig-cap: "Proportion of land cover class from the persistent land cover for 2001-2022 (>80%) per macrozone"
+#| fig-cap: "Proportion of land cover class from the persistent land cover for 2001-2022 (>80%) per macrozone and land cover macroclass."
 #| out-width: "\\textwidth"
-knitr::include_graphics("../output/figs/LC_pers80_per_macrozone.png")
+knitr::include_graphics("../output/figs/proportion_landcover_macroclass_2001-2022.png")
 ```
 
-For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands.
+For vegetation, we obtained and used hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands.
 [@fig-studyArea]c  shows the spatial distribution of the macroclasses through Chile for the year 2022.
-[@fig-studyArea]d shows the macroclasses of land cover persistance (80%) during 2021–2022, respectively (Table \ref{tab-landcoverSurf}).
+[@fig-studyArea]d shows the macroclasses of land cover persistence (80%) during 2021–2022, respectively (Table \ref{tab-landcoverSurf}).
 Within continental Chile, barren land is the land cover class with the highest surface area (277,870 $km^2$).
@@
 The macrozones with major changes for 2001–2022 were "Centro," "Sur," and "Austral," with 36%, 31%, and 34% of their surface changing the type of land cover, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend}).
-@fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.
+@fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistence mask over continental Chile.
 
@@
 Values of zero indicate that there was not a significant trend.
-Red dots on the plots indicate the maximum and minimum values of surface.}
+The red dots on the plots indicate the maximum and minimum values of the surface.
+Cells without values indicate that the land cover type was not significant in that macrozone.}
 \label{tab-landcoverTrend}
@@
 From the trend analysis in Table \ref{tab-landcoverTrend}, we can indicate that the "Norte Chico" shows an increase in barren land of 111 $km^2 yr^{-1}$ and a reduction in the class savanna of 70 $km^2 yr^{-1}$.
-In the "Centro" and "Sur," there are changes with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$ ), and an increase in shrubland and grassland.
-Showing a change for more dense vegetation types.
-The area under cultivation (croplands) appears to be shifting from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$ ) in the "Sur," seemingly replacing the savanna lost (Table \ref{tab-landcoverTrend}).
+In the "Centro" and "Sur," there are changes with an important reduction in savanna with 136 and 319 $km^2 yr^{-1}$, respectively, and an increase in shrubland and grassland, showing a change for more dense vegetation types.
+The area under cultivation (croplands) appears to be shifting from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," seemingly replacing the savanna lost (Table \ref{tab-landcoverTrend}).
 
 ### Relationship between drought indices and land cover change
+
+```{r fig-varImportance}
+#| echo: false
+#| fig-pos: "!ht"
+#| fig-cap: "cdcdcdc"
+#| fig-width: 3
+#| out-width: ".01\\textwidth"
+knitr::include_graphics("../output/figs/bars_relative_importance_RF.png")
+```
 
@@
 \label{tab-landcoverTrendRF}
-\includegraphics[]{../output/figs/table_importance_trends_landcover_vs_drought.png}
+\includegraphics[]{../output/figs/table_r2_most_important_variables.png}
 \end{table}
@@
 
-According to Table \ref{tab-landcoverTrendRF}, the random forest models for estimating the landcover trend from the trends in drought indices reach an r-squared between 0.32 and 0.39 for the types of forest, grassland, savanna, shrubland, and barren land.
-It is more likely that short- and medium-term increases in AED (EDDI-6 and EDDI-12) and short-term precipitation deficits (SPI-6 and SPEI-6) contributed to changes in grassland and bare land.
-The short-term increase of AED (EDDI-3 and EDDI-6) and the longer duration of the precipitation deficit (SPI-24, SPI-36, and SPEI-36) most likely contribute to the changes in shrubland.
+The random forest models for estimating the land cover trend based on the trends in drought indices reach an $R^2$ between 0.32 and 0.39 for the land cover types (Table \ref{tab-landcoverTrendRF}) excluding cropland.
+It is more likely that short- and medium-term increases in AED (EDDI-6 and EDDI-12) and short-term precipitation deficits (SPI-6 and SPEI-6) are associated to changes in grassland and bare land.
+The short-term increase of AED (EDDI-3 and EDDI-6) and the longer duration of the precipitation deficit (SPI-24, SPI-36, and SPEI-36) were the most important variables that correlated with changes in shrubland.
 The changes in savanna are associated with a short- and long-term increase in AED and a three-year precipitation deficit (SPI-36).
@@
 
-We study the connection between the SPI, EDDI, and SSI drought indices and changes in land cover in @fig-TrendsLandDrought.
-To do this, we compare the relative changes in land cover (in terms of the total surface area per land cover type and macrozone) over six and thirty-six months.
-@fig-TrendsLandDrought shows that the forest in the "Sur," shrubland and grassland in "Centro," barren land in "Norte Chico," and savanna in "Austral" showed an increase in the surface of landcover associated with an increase in EDDI.
+@fig-TrendsLandDrought shows the connection between the SPI, EDDI, and SSI drought indices and changes in land cover.
+Forest in the "Sur," shrubland and grassland in "Centro," barren land in "Norte Chico," and savanna in "Austral" showed an increase in land cover extent, which was associated with an increase in EDDI.
 Savanna in "Centro," "Sur," and "Norte Chico" decreases with the increase in EDDI.
 The SPI and SSI showed similar behavior regarding the trend in land cover type.
-A decrease in SPI and SSI is associated with an increase in the surface in shurubland and grassland in "Centro," forest in "Sur," and barren land in "Norte Chico," as well as a decrease trend in savanna in "Norte Chico," "Centro," and "Sur."
+A decrease in SPI and SSI is associated with an increase in the surface in shrubland and grassland in "Centro," forest in "Sur," and barren land in "Norte Chico," as well as a decrease trend in savanna in "Norte Chico," "Centro," and "Sur."
 
@@
 
-The temporal variation within the macrozones is shown in [@fig-zcNDVI_var]b).
+The temporal variation within the macrozones is shown in [@fig-zcNDVI_var]b.
 There is a negative trend in "Norte Chico" with -0.035 and "Centro" with -0.02 per decade.
@@
 #| out-width: "\\textwidth"
-#| fig-cap: "Time scales per drought index that reach the maximum coefficient of determination"
+#| fig-cap: "Time scales per drought index that reach the maximum coefficient of determination. White spaces indicate no significant correlation."
 knitr::include_graphics("../output/figs/mapa_cor_selec_indices_zcNDVI6.png")
@@
 #| out-width: "\\textwidth"
-#| fig-cap: "Pearson correlation value for the time scales and drought index that reach the maximum coefficient of determination"
+#| fig-cap: "Pearson correlation value for the time scales and drought index that reach the maximum coefficient of determination. White spaces indicate no significant correlation."
 knitr::include_graphics("../output/figs/mapa_cor_r_indices_zcNDVI6.png")
@@
 
-@fig-corTimeScale shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between different drought indicators and zcNDVI over time scales (1, 3, 6, 12, 24, and 36 months).
+@fig-corTimeScale shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between zcNDVI and different drought indicators over time scales of 1, 3, 6, 12, 24, and 36 months.
 The spatial variation of time scales reached per index is mostly for time scales above 12 months.
@@
 \caption{Summarry per land cover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
-The numbers in each cell indicate the time scale that reached the maximum correlation for the land cover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
+The numbers in each cell indicate the time scale that reached the maximum correlation for the land cover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.
+Cells without values indicate that the land cover type was not significant in that macrozone.}
 \label{tab-corlandcover}
@@
 
-In Table \ref{tab-corlandcover}, we aggregate per macrozone and landcover the correlation analysis presented in @fig-corTimeScale and @fig-corPerson.
+In Table \ref{tab-corlandcover}, we aggregate per macrozone and land cover the correlation analysis presented in @fig-corTimeScale and @fig-corPerson.
 According to what is shown, forests seem to be the most resistant to drought.
@@
 
-## Drought trend and attribution to land cover change
+## Vegetation water demand and its relation to drought
 
-@Vicente-Serrano2022, in a study at the global scale of drought trends, indicates that there have not been significant trends in meteorological drought since 1950.
-Also, state that the increase in hidrological trend in some parts of the globe (northeast Brazil and the Mediterranean region) is related to changes in land cover and specifically to the rapidly increasing irrigated area, which consequently increases water extraction.
-@Kogan2020 analyzed the agricultural drought impact globally and in the main grain producer countries, finding that "since 1980, the Earth warming has not changed the drought area or intensity." In our study, we took into account the variation in vegetation productivity in Chile, specifically in areas without any changes in land cover, to prevent any misleading conclusions about the increase in water demand due to land cover change.
-Our results show a contrasting perspective.
-The SPI, SPEI, and SSI (water supply) showed a decrease in trends, except for the southern part, and an increase in EDDI (water demand).
-The trend, positive or negative, was stronger as the time scales increased.
-Trends in the long term (e.g., 36 months) are evidence of how human-induced climate change is affecting Chile, which seems to be due to an intense hydrological drought resulting from the persistence of the precipitation deficit.
-We found that there has been a significant trend in the decline of vegetation productivity (zcNDVI) since 2000 for the north-central part of the country, which has been extreme between 2020 and 2022 and has impacted natural and cultivated land.
-Additionally, we demonstrated that the drought, primarily due to an increase in AED, accounts for about 30% of the changes in land cover types (excluding croplands).
-These changes are associated with a decrease in water demand from vegetation.
-Moreover, the most water-demanding type, cropland, showed a decrease in the north-central region, while barren land showed an increase.
-The north-central part of the country primarily experienced these changes due to a higher increase in AED.
-Thus, we have evidence of a significant decline in water supply and an increase in AED for the north-central part of Chile, which show to be the most relevant variables for drought conditions.
-Some questions arise regarding what is occurring with the cultivated land.
-We used the unchanged land cover to ensure that an increase in surface area is not considered in the trend analysis.
-For croplands, it could happen that some areas have changed the types of crops for others with higher water demand, which consequently increases water demand.
-However, this effect should be minor compared to the decrease in water supply and increase in water demand at this scale of analysis.
+In our study, we took into account the variation in vegetation productivity in Chile, specifically in areas without any changes in land cover, to prevent any misleading conclusions about the increase in water demand due to land cover change.
+Our results show a contrasting perspective regarding the evidence provided by @Vicente-Serrano2022 on a global scale, who indicates that the increase in drought is led by an increase in agricultural land, which in turn increases water demand.
 
-This shows that the main cause of the hydrological drought in Chile was a steady drop in water supply made worse by an increase in AED, but it seems that in zones most affected by drought, the main cause is not an increase in water demand due to an intensification of cultivated land (e.g., an increase in irrigated crops).
+Our results indicate that except for the southern part of the country, the SPI, SPEI, and SSI (water supply) showed declining trends, while the EDDI (water demand) increased across continental Chile.
+The trends in water demand and supply were stronger as the time scales increased, indicating a long-term reduction in water supply (except for the southern part) and an increase in water demand by the atmosphere.
+Also, we found that there has been a significant declining trend in vegetation productivity (zcNDVI) since 2000 for the north-central part of the country, which reached its lowest level between 2020 and 2022 and has impacted natural and cultivated land.
+Further, croplands showed a decrease in surface area for the north-central region, while barren land increased.
+We link these changes to a decrease in the water demand from vegetation because, despite the increase in AED, the surface area for water-demanding vegetation is declining as well as the biomass production.
+However, some questions arise regarding what is occurring with the cultivated land.
+Evidence suggests that higher-water-demanding crops have replaced croplands in the Petorca basin (central Chile), leading to an increase in water abstraction [@Munoz2020;@Duran2020].
+Nonetheless, at this scale of analysis, the effect of higher crop water demand on drought is minor compared to the decrease in water supply and increase in AED over all land cover types.
+
+The long-scale trends (e.g., 36 months) demonstrate the impact of human-induced climate change on water availability in Chile, potentially due to an intense hydrological drought stemming from the ongoing precipitation deficit and rising AED.
+But it is likely that in zones most affected by drought, the main cause is not an increase in vegetation water demand due to an intensification of cultivated land (e.g., an increase in irrigated crops) like in other parts of the globe [@Vicente-Serrano2020].
 North-central Chile has experienced a decline in vegetation productivity across land cover types, which is primarily attributable to variations in water supply and soil moisture.
-An increase in water demand, such as an increase in the surface area of irrigated crops, could strengthen this trend.
+An increase in water demand, led by an increase in the surface area of irrigated crops or the change to more water-demanding crops, could strengthen this trend, however, it escapes the scope of this study.
+Future work should focus on the regions where the drought has been more severe and has a high proportion of irrigated crops to get insight on the real impact of irrigation on ecosystems in those zones.
 
-## Impact of short- to long-term droughts on land cover types
+## Sensitivity of land cover vegetation to short- and long-term drought
 
-We found that the 12-month soil moisture deficit most affects the productivity of vegetation in all land cover types along Chile.
-In a study in the Yangtze River Basin in China, @Jiang2020 analyzed the impact of drought on vegetation using the SPEI and the Enahanced Vegetation Index (EVI).
+We analyzed the time series of drought indices and vegetation productivity per land cover type.
+Our results indicate that forest is the type most resistant to drought, and shrublands, savannas, and croplands have higher sensitivity.
+
+In their study in the Yangtze River Basin in China, @Jiang2020 analyzed the impact of drought on vegetation using the SPEI and the Enhanced Vegetation Index (EVI).
 They found that cropland was more sensitive to drought than grassland, showing that cropland responds strongly to short- and medium-term drought (< SPEI-6).
 In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro." In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016;@Potopova2015;@Dai2020;@Rhee2010].
-Short-term precipitation deficits impact soil water, and thus less water is available for plant growth.
-However, we found that in "Norte Chico," an SPI-36 and SPEI-12 had a higher impact, which are associated with hydrological drought (long-term), and in "Centro," an SPI-12 and SPEI-12.
-Thus, we attribute this impact to the hydrological drought that has decreased groundwater storage [@Taucare2024], which in turn is impacted by long-term deficits, and consequently, the vegetation is more dependent on groundwater.
+Short-term precipitation deficits have an impact on soil water, so less water is available for plant growth.
+However, we found that in "Norte Chico," an SPI-36 and SPEI-12 had a higher impact, which are associated with long-term water deficit, and in "Centro," an SPI-12 and SPEI-12.
+Thus, we hypothesize that this impact could be attributed to the hydrological drought that has decreased groundwater storage [@Taucare2024], which in turn is impacted by long-term deficits, and consequently, the vegetation is more dependent on groundwater.
 In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared.
-The possible reason for this is that the most resistant types, forest and grassland, predominate south of "Centro." Also, drought episodes have been less frequent and intense.
-The drought episodes have had a lower impact on water availability for vegetation.
+The possible reason for this is that the most resistant types, forest and grassland, predominate south of "Centro." Also, drought episodes have been less frequent and intense and have had a lower impact on water availability for vegetation.
 
 According to @Senf2020, severe drought conditions in Europe are a significant cause of tree mortality.
-However, we discovered that forests, as the most resilient land cover class to drought, experience less variation in drought indices.
-Supporting this is @Fathi-Taperasht2022, who asserts that Indian forests are the most drought-resistant and recover rapidly.
+However, we discovered that forests are the most resilient land cover class to drought, with less variation in drought indices.
+Supporting this is @Fathi-Taperasht2022, who assert that Indian forests are the most drought-resistant and recover rapidly.
 Similarly, the work of @Wu2024, who analyzed vegetation loss and recovery in response to meteorological drought in the humid subtropical Pearl River basin in China, indicates that forests showed higher drought resistance.
-Using Vegetation Optical Depth (VOD), kNDVI, and EVI, @Xiao2023 tests the resistance of ecosystems and finds that ecosystems with more forests are better able to handle severe droughts than croplands.
+Using Vegetation Optical Depth (VOD), kNDVI, and EVI, @Xiao2023 tested the resistance of ecosystems and found that ecosystems with more forests are better able to handle severe droughts than croplands.
 They attribute the difference to a deeper rooting depth for trees, a higher water storage capacity, and different water use strategies between forest and cropland [@Xiao2023].
-In contrast, @Venegas2022, who studied Cryptocarya alba and Beilschmiedia miersii (both from the Lauraceae family) that live in sclerophyllous forests in Chile, found that the trees' overall growth had slowed down.
-This could mean that the natural dynamics of their forests have changed.
-They attributed it to the cumulative effects of the unprecedented drought (i.e., hydrological drought).
 
-Thus, we attribute that forest to being the most resistant to drought, due to the fact that most of the species comprising it are highly resilient to water scarcity compared to the other land cover classes.
-Nonetheless, if we want to go deep in our analysis, we should use earth observation data that is able to capture a higher level of detail.
-For example, when we used MOD13A3 with a 1km spatial resolution to measure vegetation condition, it took the average condition of 1 square kilometer.
-Then, to use remote sensing to look at how a certain type of forest (like sclerophyllous forest) changes in response to drought on a local level, we should use operational products with higher spatial resolutions, like those from Landsat or Sentinel.
-This will let us do a more thorough analysis.
+In central Chile, @Venegas2022 observed a significant decline in the overall growth of sclerophyllous moist forests (mediterranean forests), which they attributed to increased drought conditions.
+In another study, @Fuentes2021 evaluated water scarcity and land cover change in Chile between 29° and 39° south latitude.
+They used the one-month SPEI for drought evaluation, which resulted in misleading results.
+For instance, they failed to identify a temporal trend in the SPEI, but they still observed a decline in water availability and a rise in AED, trends that should have been detectable if they were using longer SPEI time scales.
+Thus, according to the results presented in this study, for the assessment of drought, it is necessary to consider drought indices on a short- to long-scale basis.
 
-## Vegetation productivity and drought.
+## Vegetation productivity and drought
 
+We found that the 12-month soil moisture deficit affected plant productivity in all land cover types in Chile.
 The main external factors that affect biomass production by vegetation are actual evapotranspiration and soil moisture, and the rate of ET in turn depends on the availability of water storage in the root zone.
 Thus, soil moisture plays a key role in land carbon uptake and, consequently, in the production of biomass [@Humphrey2021].
-Moreover, @Zhang2022 indicate there is a bidirectional causality between soil moisture and vegetation productivity.
-Lastly, some studies have redefined agricultural drought as soil moisture drought from a hydrological perspective [@Loon2016;@Samaniego2018].
-Even though soil moisture is the external factor most determinant of vegetation biomass, there are multiple internal factors, such as species, physiological characteristics, and plant hydraulics, that would affect vegetation productivity.
-Because of that, we believe that agricultural drought, referring to the drought that impacts vegetation productivity, is the most proper term, as originally defined by @Wilhite1985.
-
 The study results showed that the soil moisture-based drought index (SSI) was better at explaining vegetation productivity across land cover macroclasses than meteorological drought indices like SPI, SPEI, and EDDI.
-In the early growing season and especially in irrigated rather than rainfed croplands, soil moisture has better skills than SPI and SPEI for estimating gross primary production (GPP).
-This according to @Chatterjee2022 evaluation of the SPI and SPEI and their correlation with GPP in the CONUS.
+According to [@Chatterjee2022] in the early growing season and especially in irrigated rather than rainfed croplands, soil moisture has better skills than SPI and SPEI for estimating gross primary production (GPP) .
 Also, @Zhou2021 indicate that the monthly scaled Standardized Water Deficit Index (SWDI) can accurately show the effects of agricultural drought in most of China.
@@
 The semi-arid "Norte Chico" and the Mediterranean "Centro" were where SSI had the best performance.
-In Chile, medium-term deficits of 12 months are more relevant in the response of vegetation, which decreases to the south, and in the case of croplands, they seem to react in a shorter time, with six months (SSI-6) in "Centro." This variation for croplands could be related to the fact that in "Norte Chico," the majority of crops are irrigated, but to the south there is a higher proportion of rainfed agriculture, which is most dependent on the short-term availability of water.
-Rather, in the "Norte Chico," the orchards are more dependent on the storage of water in dams of groundwater reservoirs, which are affected by long-term drought (e.g., SPI-36).
+In Chile, medium-term deficits of 12 months are more relevant in the response of vegetation for all land cover types, which decreases to the south, and in the case of croplands, they seem to react in a shorter time, with six months (SSI-6) in "Centro." This variation for croplands could be related to the fact that in "Norte Chico," the majority of crops are irrigated, but to the south there is a higher proportion of rainfed agriculture, which is most dependent on the short-term availability of water.
+Rather, in "Norte Chico," the orchards are more dependent on irrigation, which in turn depends on the availability of storage water in dams of groundwater reservoirs, which are affected by long-term drought (e.g., SPI-36).
 
-## Future outlook (to complete)
+## Drought information to aid in adaptation
+
+Our findings present valuable information for policymakers in developing adaptation strategies for droughts.
+Our results show that the different climate components, such as AED, water supply, soil moisture, and their impact on vegetation, should be considered when evaluating the multi-dimensional nature of drought.
+Also, for a better understanding of drought propagation [@VanLoon2012] from meteorological to agricultural and ecological drought, we should consider the climatic response at different time scales, ranging from short to long.
+Additionally, the spatiotemporal characteristics of our results allow us to distinguish distinct geographical contexts, recognizing the diversity in climate, but also shedding light on agricultural practices (ranging from irrigated to dryland farming), technological advancements in irrigation efficiency, and the region-specific capabilities for drought adaptation, including groundwater management and reservoir water storage.
+This information, combined with agricultural information and statistics, could provide a strong foundation for the development of science-based adaptation policies.
+
+In a commitment to fostering informed and dynamic adaptation efforts, our results are disseminated publicly and continually updated via the Drought Observatory for Agriculture and Biodiversity of Chile (ODES) ([https://odes-chile.org/app/unidades](https://odes-chile.org/app/unidades)) [@Zambrano2023b;@Kunst2023].
+This initiative ensures the availability of extensive climate data, facilitating the development of adaptive strategies that are both responsive to the realities of different regions and grounded in the latest scientific understanding.
+The proactive sharing and updating of such data underscores its key role in enabling policymakers to craft adaptive measures that are finely tuned to the diverse and evolving landscapes of drought impacts.
+Furthermore, the recently promulgated law about climate change in Chile (Law 21.455, https://www.bcn.cl/leychile/navegar?idNorma=1177286), which aims to implement sectoral adaptation plans for agriculture, forests, and biodiversity, could benefit from this information.
 
@@
 
-There is a trend toward decreasing water supply in most parts of Chile, particularly in the “Centro” and “Norte Chico” regions.
-The whole country showed an increase in AED.
-Vegetation productivity only showed a decrease in the “Norte Chico” and “Centro,” being highest for shrubland and croplands.
-Forest is the land cover most resistant to drought, as shown along Chile, and shrubland and cropland are the most sensitive.
+We found a significant trend toward decreasing water supply (SPI, SPEI, and SSI) in most of the Chilean territory, with the exception of the southern region.
+The trend is the strongest in the north-central zone.
+The whole country showed an increase in water demand (AED) due to increasing temperatures.
+The magnitude of the trends is stronger for longer time scales, which is evidence that there is a prolonged precipitation shortage and a prolonged increase in AED.
+The trend in vegetation productivity in the north-central area is affecting shrubland and savanna to a greater degree, followed by croplands and forests.
 
-A soil moisture deficit of 12 months (SSI-12) is highly correlated with vegetation productivity for the land cover classes of shrubland, savannas, croplands, and forest in “Norte Chico” and “Centro.” For the southern part of the country with humid conditions, the correlation with SSI decreases.
-Soil moisture overcomes the capacity to explain vegetation productivity over the supply and demand drought indices in the entire territory.
+We  model about 20–30% of the trends in land cover types, such as forest, grassland, shrubland, and savanna, based on drought indicators across Chile.
+There is no evidence that drought alters cropland surface area.
+The increase in AED is the most important variable explaining the variability in the change in land cover, followed by a reduction in precipitation and soil moisture.
 
-The variation in vegetation productivity appears to be associated with climate variation rather than anthropogenic factors (e.g., an increase in water demand by irrigated crops).
-Even though switching to more demanding crops on the land could increase the impact of drought on vegetation, this would need to be more thoroughly investigated, for instance at the watershed level.
+The trends in drought indices are accompanied by multiple land cover changes in the country, most notably an increase of forest in "Sur," of shrubland and grassland in "Centro,” and of savanna in "Centro" and "Sur." In "Norte Chico" and "Centro,” the croplands have been declining in surface, whereas in "Sur,” there is an increase in cultivated land.
 
-The results of this study could help in the development of a robust forecasting system for land cover classes in Chile, helping to improve preparedness for climate change impacts on vegetation.
+The change in vegetation productivity has been severe in the north-central part of the country for all land cover types, particularly savanna, shrubland, and croplands.
+The anomaly in soil moisture over the past 12 months is the main variable explaining these changes, followed by anomalies in accumulated precipitation over one to two years.
+The variation in AED seems to intensify the drought impact on vegetation productivity.
 
-# Supplementary material {-}
+The results of this study provide insightful information that would assist in developing adaptation measures for Chilean ecosystems to cope with climate change and drought.
+Information that could be used in the scope of the national law on climate change, which seeks to implement adaptation strategies for agriculture, forests, and biodiversity.
+
+# Acknowledgment
+
+The National Research and Development Agency of Chile (ANID) funded this study through the drought emergency project FSEQ210022, Fondecyt Iniciación N°11190360,  Fondecyt Postdoctorado N°3230678, and Fondecyt Regular N°1210526.
 

### 2024-04-11 article/paper_drought_Chile2023.tex
last update antes egu2024

@@
 \hypersetup{
-  pdftitle={The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile},
+  pdftitle={The effects of drought on land cover change and vegetation productivity in continental Chile},
   pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
   pdfkeywords={drought, land cover change, vegetation
-productivity, satellite},
+productivity, ecosystem},
   colorlinks=true,
@@
 \begin{frontmatter}
-\title{The effects of multi-dimensional drought on land cover change and
-vegetation productivity in continental Chile}
+\title{The effects of drought on land cover change and vegetation
+productivity in continental Chile}
 \author[1,2]{Francisco Zambrano%
 \corref{cor1}%
-\fnref{fn1}}
+}
  \ead{francisco.zambrano@umayor.cl} 
@@
  \ead{a.vrieling@utwente.nl} 
-\author[]{Francisco Meza%
+\author[4,5,6]{Francisco Meza%
 %
@@
 
-\author[]{Iongel Duran-Llacer%
+\author[7]{Iongel Duran-Llacer%
 %
@@
 
-\author[4]{Francisco Fernández%
+\author[8,9]{Francisco Fernández%
 %
@@
 
-\author[]{Nicolas Raab%
+\author[4]{Nicolas Raab%
 %
@@
 
-\author[]{Dylan Craven%
+\author[11,12]{Dylan Craven%
 %
@@
 
-\affiliation[1]{organization={Universidad Mayor, Hémera Centro de
-Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en
-Medio Ambiente y Sustentabilidad},city={Santiago,
-Chile},postcode={7500994},postcodesep={}}
-\affiliation[2]{organization={Universidad Mayor, Observatorio de Sequía
-para la Agricultura y la Biodiversidad de Chile (ODES)},,postcodesep={}}
-\affiliation[3]{organization={University of Twente, Faculty of
-Geo-Information Science and Earth Observation (ITC)},city={Enschede, The
+\affiliation[1]{organization={Hémera Centro de Observación de la Tierra,
+Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y
+Sustentabilidad, Universidad Mayor},city={Santiago},country={7500994,
+Chile},countrysep={,},postcodesep={}}
+\affiliation[2]{organization={Observatorio de Sequía para la Agricultura
+y la Biodiversidad de Chile (ODES), Universidad
+Mayor},city={Santiago},country={7500994,
+Chile},countrysep={,},postcodesep={}}
+\affiliation[3]{organization={Faculty of Geo-Information Science and
+Earth, University of Twente},city={Enschede, The
 Netherlands},postcodesep={}}
-\affiliation[4]{organization={Universidad San
-Sebastian},,postcodesep={}}
+\affiliation[4]{organization={Facultad de Agronomía y Sistemas
+Naturales, Pontificia Universidad Católica de
+Chile.},city={Santiago},country={Chile},countrysep={,},postcodesep={}}
+\affiliation[5]{organization={Instituto para el Desarrollo Sustentable.
+Pontificia Universidad Católica de
+Chile},city={Santiago},country={Chile},countrysep={,},postcodesep={}}
+\affiliation[6]{organization={Centro Interdisciplinario de Cambio
+Global, Pontificia Universidad Católica de Chile, Santiago,
+Chile.},,postcodesep={}}
+\affiliation[7]{organization={Hémera Centro de Observación de la Tierra,
+Facultad de Ciencias, Universidad
+Mayor,},city={Santiago},country={7500994,
+Chile},countrysep={,},postcodesep={}}
+\affiliation[8]{organization={Center of Economics for Sustainable
+Development (CEDES), Faculty of Economics and Government, Universidad
+San
+Sebastian},city={Santiago},country={Chile},countrysep={,},postcodesep={}}
+\affiliation[9]{organization={Center of Applied Ecology and
+Sustainability
+(CAPES)},city={Santiago},country={Chile},countrysep={,},postcodesep={}}
+\affiliation[10]{organization={},,postcodesep={}}
+\affiliation[11]{organization={, GEMA Center for Genomics, Ecology \&
+Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750,
+Santiago, Chile},,postcodesep={}}
+\affiliation[12]{organization={Data Observatory Foundation, Santiago,
+Chile},,postcodesep={}}
 
 \cortext[cor1]{Corresponding author}
-\fntext[fn1]{This is the first author footnote.}
 
 \begin{abstract}
-The north-central region of Chile has been the focus of research studies
-due to the persistent decrease in water supply, which is impacting the
-hydrological system and vegetation development. This persistent period
-of water scarcity has been defined as a mega-drought. The aim of our
-study is to evaluate the interaction of drought, land cover change, and
-vegetation productivity over continental Chile. We used drought indices
-for atmospheric evaporative demand (AED), water supply, and soil
-moisture from short-term (1, 3, 6 months) to long-term (12, 24, 36
-months) time scales. We derived the drought indices using monthly
-ERA5-Land reanalysis data spanning from 1981 to 2023. We utilized the
-Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets to derive
-information on annual land cover and monthly vegetation productivity.
-Our results showed that from south to north, Chile has a declining trend
-in water supply, and across the whole country, there is an increase in
-AED. These trends are stronger at longer time scales. The trend in
-vegetation productivity in the north-central area is affecting, to a
-higher degree, shrubland and savanna, followed by croplands and forests.
-The drought explains about 30\% of the change in land cover type across
-Chile for forest, grassland, shrubland, and savanna. The increase in AED
-is the main driver of the change in land cover, followed by a reduction
-in precipitation and soil moisture. The change in vegetation
-productivity has been severe in the north-central part of the country
-for all land cover types, particularly savanna, shrubland, and
-croplands. The anomaly in soil moisture over the past 12 months (SSI-12)
-is the main variable explaining these changes, followed by anomalies in
-cumulated precipitation over one to two years (SPI-12 and SPI-24). Our
-results provide insightful information that would help in developing
-adaptation measures for ecosystems in Chile to cope with climate change
-and drought.
+Chile has experienced a persistent decrease in water supply, which
+impacts the hydrological system and vegetation development. This
+persistent period of water scarcity has been defined as a mega-drought.
+There are few studies about the relationship between drought and
+ecosystem changes that can help with a better understanding of
+ecological drought. The aim of our study is to evaluate the interaction
+of drought, land cover change, and vegetation productivity over
+continental Chile. To assess drought, we used drought indices for
+atmospheric evaporative demand (AED), water supply, and soil moisture
+from short- (1, 3, 6 months) to long-term (12, 24, 36 months) time
+scales. We derived the drought indices using monthly ERA5-Land
+reanalysis data from 1981 to 2023. We used the Moderate-Resolution
+Imaging Spectroradiometer (MODIS) datasets to derive information on
+annual land cover and monthly vegetation productivity. Our results
+showed that except for the Austral part, Chile has a temporal decreasing
+trend in water supply, and across the whole country, there is an
+increase in AED. These trends become stronger over longer time scales.
+We found a negative trend in vegetation productivity in the
+north-central area, which is more prominent for shrubland and savanna as
+compared to croplands and forests. The anomaly in soil moisture over the
+past 12 months (SSI-12) is the most important variable explaining these
+changes, followed by anomalies in accumulated precipitation over one to
+two years (SPI-12 and SPI-24). The variable importance obtained by
+random forest models indicates that drought is explaining about 20--30\%
+of the change in land cover surface across Chile for forest, grassland,
+shrubland, and savanna but has no relation to the changes in croplands.
+The increase in AED is the main variable associated with the change in
+land cover, followed by a reduction in precipitation and soil moisture.
+Our findings provide insightful information that could assist in
+developing adaptation measures for Chilean ecosystems to cope with
+climate change and drought. Also, this study could contribute to a
+better comprehension of ecological drought.
 \end{abstract}
@@
     drought \sep land cover change \sep vegetation productivity \sep 
-    satellite
+    ecosystem
 \end{keyword}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, breakable]}{\end{tcolorbox}}\fi
 
@@
 
-Drought is often classified as meteorological when there is a decrease
-in precipitation below the mean average of several years (more than 30
-years), hydrological when these anomalies last for long periods (months
-to years) and affect water systems, and agricultural when the deficit
-impacts plant health anomalies and leads to decreased productivity
-\citep{Wilhite1985}. However, it is important to note that drought is
-also influenced by human activities, which were not considered in the
-definitions. Thus, \citet{Loon2016} and \citet{AghaKouchak2021} have
-given an updated definition of drought for the Anthropocene, suggesting
-that it should be considered the feedback of humans' decisions and
-activities that drives the anthropogenic drought. Simultaneously,
-drought leads to heightened tree mortality and induces alterations in
-land cover and land use, ultimately affecting ecosystems
-\citep{Crausbay2017}. Even though many ecological studies have
-misinterpreted how to characterize drought, for example, sometimes
-considering ``dry'' conditions as ``drought'' \citep{Slette2019}. Then,
-\citet{Crausbay2017} proposed the ecological drought definition as ``an
+Drought is often classified as 1) meteorological when precipitation in a
+specific period remains below the mean precipitation experienced in the
+same period during multiple years (more than 30 years usually), 2)
+hydrological when these anomalies last for long periods (months to
+years) and affect water systems, and 3) agricultural when the deficit
+negatively impacts plant health and leads to decreased productivity of
+crops or pastures \citep{Wilhite1985}. However, because drought is also
+influenced by human activities, \citet{Loon2016} and
+\citet{AghaKouchak2021} expanded the drought definition for the
+Anthropocene, indicating that the feedback of human decisions and
+activities should also be considered (i.e., anthropogenic drought).
+Droughts lead to increased tree mortality \citep{Cheng2024} and induces
+alterations in land cover and land use, ultimately affecting ecosystems
+\citep{Crausbay2017}. Even though many ecological studies have at times
+mistakenly considered ``dry'' conditions as ``drought''
+\citep{Slette2019}. Ecological drought can be defined as \emph{``an
 episodic deficit in water availability that drives ecosystems beyond
 thresholds of vulnerability, impacts ecosystem services, and triggers
-feedback in natural and/or human systems.'' In light of current global
-warming, it is crucial to study the interaction between drought and
-ecosystems in order to understand their feedback and impact on water
-security. \citep{Bakker2012}
-
-Human-induced greenhouse gas emissions have increased the frequency
-and/or intensity of drought as a result of global warming, according to
-the sixth assessment report (AR6) of the Intergovernmental Panel on
-Climate Change (IPCC) \citep{IPCC2023}. The evidence supporting this
-claim has been strengthened since AR5 \citep{IPCC2013}. Recent studies,
-however, have produced contrasting findings, suggesting that drought has
-not exhibited a significant trend over the past forty years.
+feedback in natural and/or human systems''} \citep{Crausbay2017}. In
+light of current global warming, it is crucial to study the interaction
+between drought and ecosystems in order to understand their feedback and
+impact on future water security \citep{Bakker2012}.
+
+Global warming, as a result of human-induced greenhouse gas emissions,
+has increased the frequency and intensity of drought, according to the
+sixth assessment report (AR6) of the Intergovernmental Panel on Climate
+Change (IPCC) \citep{IPCC2023}. The evidence supporting this claim has
+been strengthened since AR5 \citep{IPCC2013}. Recent studies, however,
+have produced contrasting findings, with some suggesting that drought
+has not exhibited a significant trend over the past forty years
 \citep{Vicente-Serrano2022, Kogan2020}. \citet{Vicente-Serrano2022}
-analyzed the meteorological drought trend on a global scale, finding
-that only in a few regions has there been an increase in the severity of
-drought. Moreover, they attribute the increase in droughts over the past
-forty years solely to an increase in atmospheric evaporative demand
-(AED), which in turn enhances vegetation water demand, with important
+analyzed the trend in meteorological drought on a global scale, finding
+that only in a few regions an increase in the severity of drought was
+observed. Moreover, they attributed this increase solely to an increase
+in atmospheric evaporative demand (AED) due to higher temperatures,
+which in turn enhances vegetation water demand, with important
 implications for agricultural and ecological droughts. Also, they state
-that ``the increase in hydrological droughts has been primarily observed
-in regions with high water demand and land cover change''. Similarly,
-\citet{Kogan2020} analyzed the drought trend using vegetation health
-methods, finding that for the globe, hemispheres, and main
-grain-producing countries, drought has not expanded or intensified for
-the last 38 years. Further, \citet{IPCC2021} suggests that there is a
+that \emph{``the increase in hydrological droughts has been primarily
+observed in regions with high water demand and land cover change, led by
+an increase in agricultural land''}. Similarly, \citet{Kogan2020}
+analyzed the drought trend using remotely-sensed vegetation health
+indicators, finding that for the globe and main grain-producing
+countries, drought has not expanded or intensified during the past 38
+years. Nonetheless, \citet{IPCC2021} suggests that there is a medium to
 high degree of confidence that rising temperatures will increase the
-extent, frequency, and severity of droughts. Also, AR6 \citep{IPCC2023}
-predicts that many regions of the world will experience more severe
-agricultural and ecological droughts even if global warming stabilizes
-at 1.5°--2°C. To better evaluate the impact of drought trends on
-ecosystems, assessments are needed that relate meteorological and soil
-moisture variables to their effects on vegetation.
+extent, frequency, and severity of agricultural and ecological droughts.
+Also, AR6 \citep{IPCC2023} predicts that many regions of the world will
+experience more severe agricultural and ecological droughts even if
+global warming stabilizes at 1.5°--2°C. To better evaluate the impact of
+drought trends on ecosystems, assessments that correlate meteorological
+and soil moisture variables to their effects on vegetation are much
+needed.
 
@@
 changes in land cover globally \citep{Winkler2021, Song2018} and
-regionally \citep{Chamling2020, Homer2020, Yang2021}. Some others seek
-to analyze the impact of land cover change on climate conditions such as
-temperature and precipitation \citep{Luyssaert2014, Pitman2012}. There
-is less research on the interaction between drought and land cover
-change \citep{Chen2022, Akinyemi2021, Peng2017}. \citet{Peng2017}
-conducted a worldwide investigation utilizing net primary production to
-examine the spatial and temporal variations in vegetation productivity
-at global level. The study aimed to assess the influence of drought by
-comparing the twelve-month Standardized Precipitation Evapotranspiration
-Index (SPEI) and land cover change. According to their findings, drought
-is responsible for 37\% of the decline in vegetation productivity, while
-water availability accounts for 55\% of the variation. \citet{Chen2022}
-studied the trend of vegetation greenness and productivity and its
-relation to meteorological drought (SPEI of twelve months in December)
-and soil moisture at the global level. The results showed lower
-correlations (\textless0.2) for both variables. \citet{Akinyemi2021}
-evaluates drought trends and land cover change using vegetation indices
-in Botwsana in a semi-arid climate. These studies mostly looked at how
+regionally
+\citep{Chamling2020, Homer2020, Yang2021, Schulz2010, Echeverria2012}.
+Some seek to analyze the impact of land cover change on climate
+conditions such as temperature and precipitation
+\citep{Luyssaert2014, Pitman2012}. There is less research on drought and
+its relation to land cover change and vegetation productivity
+\citep{Chen2022, Akinyemi2021, Peng2017}. \citet{Peng2017} utilized net
+primary productivity to examine the spatial and temporal variations in
+vegetation productivity at global level and assess to what extent
+drought influenced this variability by comparing the twelve-month
+Standardized Precipitation Evapotranspiration Index (SPEI) and land
+cover change. According to their findings, drought is responsible for
+37\% of the decline and accounts for 55\% of the variability in
+vegetation productivity. \citet{Chen2022} instead found poor
+correlations (r\textless0.2) between the vegetation productivity trends
+against meteorological drought (SPEI of twelve months in December) and
+soil moisture at the global level. These studies mostly looked at how
 changes in land cover and vegetation productivity are related to a
-single drought index (SPEI) over a single time period of 12 months. SPEI
-takes into account the combined effect of precipitation and AED as a
-water balance, but it does not allow us to know the contribution of each
-variable on its own. Some things worth investigating in terms of land
-cover change and vegetation productivity are: i) How do they respond to
-short- to long-term meteorological and soil moisture droughts? ii) How
-is the drought impacting land cover changes? And iii) How do they behave
-in humid and arid climatic zones regarding drought? Likewise, there is a
-lack of understanding of how the alteration in water supply and demand
-is affecting land cover transformations.
-
-For monitoring drought, the World Meteorological Organization recommends
-the SPI (Standardized Precipitation Index) \citep{WMO2012}. The SPI is a
-multi-scalar drought index that only uses precipitation to assess short-
-to long-term droughts. The primary cause of drought is precipitation
-anomalies, and temperature usually makes it worse \citep{Luo2017}.
-Nowadays, there is an increase in attention toward using AED separately
-to monitor droughts \citep{Vicente-Serrano2020}. One reason is due to
-its attribution to increasing flash droughts in water-limited regions
-\citep{Noguera2022}. \citet{Vicente-Serrano2010} proposed the
-Standardized Precipitation Evapotranspiration Index (SPEI), which
-incorporated the temperature effect by subtracting AED from
-precipitation. SPEI allows for analysis of the combined effect of
-precipitation and AED. Since its formulation, it has been used worldwide
-for the study and monitoring of drought
-\citep{Gebrechorkos2023, Liu2021}. \citet{Hobbins2016} and
-\citet{McEvoy2016} developed the Evaporative Demand Drought Index (EDDI)
-to monitor droughts solely using the AED, and it has proven effective in
-monitoring flash droughts \citep{Li2024, Ford2023}. For soil moisture,
-several drought indices exist, such as the Soil Moisture Deficit Index
-(SDMI) \citep{Narasimhan2005} and the Soil Moisture Agricultural Drought
-Index (SMADI) \citep{Souza2021}. \citet{Hao2013} and
-\citet{AghaKouchak2014} proposed the Standardized Soil Moisture Index
-(SSI), which has a similar formulation as the SPI, SPEI, and EDDI. Thus,
-there are plenty of drought indices that allow for a comprehensive
-assessment of drought on short- to long-term scales and that allow for
-the use of single variables from the earth's water balance (e.g.,
-precipitation, AED, soil moisture). The variation in climate variables
-impacts vegetation development, and unfavorable conditions such as low
-precipitation and high temperatures usually generate a decrease in
-vegetation productivity. To monitor the response of vegetation, the
-common practice is to use satellite data. The Normalized Difference
-Vegetation Index (NDVI) has been widely used as a proxy for biomass
+single drought index (SPEI) obtained for 12 month periods. SPEI takes
+into account the combined effect of precipitation and AED as a water
+balance, but it does not allow to know the contribution of each variable
+on its own. To better understand these contributions on land cover
+change and vegetation productivity the following questions may be asked:
+i) how do land cover and vegetation productivity respond to short- to
+long-term meteorological and soil moisture droughts? And ii) How is this
+response different between humid and arid climatic zones? Likewise,
+there is a lack of understanding of how the alteration in water supply
+and demand is affecting land cover transformations.
+
+To address the previous questions over extensive regions, we can utilize
+gridded data on water availability, vegetation conditions, and the
+respective drought indices. For monitoring drought, the World
+Meteorological Organization recommends the SPI (Standardized
+Precipitation Index) \citep{WMO2012}. The SPI is a multi-scalar drought
+index that only uses precipitation to assess short- to long-term
+droughts. \citet{Vicente-Serrano2010} proposed the Standardized
+Precipitation Evapotranspiration Index (SPEI), which incorporates the
+temperature effect by subtracting AED from precipitation. SPEI allows
+for analyzing the combined effect of precipitation and AED. Since its
+formulation, it has been used worldwide for the study and monitoring of
+drought \citep{Gebrechorkos2023, Liu2024}. Recently, there has been more
+interest in using AED to track droughts separately to better disengage
+the effects of precipitation from temperature-dependent effects
+\citep{Vicente-Serrano2020}. One of the reasons is that AED is linked to
+more flash droughts in limited water regions \citep{Noguera2022}.
+\citet{Hobbins2016} and \citet{McEvoy2016} developed the Evaporative
+Demand Drought Index (EDDI) to monitor droughts solely using the AED,
+and it has proven effective in monitoring flash droughts
+\citep{Li2024, Ford2023}. For soil moisture, several drought indices
+exist, such as the Soil Moisture Deficit Index (SDMI)
+\citep{Narasimhan2005} and the Soil Moisture Agricultural Drought Index
+(SMADI) \citep{Souza2021}. \citet{Hao2013} and \citet{AghaKouchak2014}
+proposed the Standardized Soil Moisture Index (SSI), which has a similar
+formulation as the SPI, SPEI, and EDDI. Thus, many drought indices exist
+that allow for a comprehensive assessment of drought on short- to
+long-term scales and that allow for the use of single variables from
+Earth's water balance (e.g., precipitation, AED, soil moisture).
+Climatic variability impacts vegetation development, with unfavorable
+conditions such as low precipitation and high temperatures usually
+promote a decrease in plant productivity. To monitor the response of
+vegetation for large areas, the common practice is to use satellite
+data. For example, the Normalized Difference Vegetation Index (NDVI)
+derived from frequent satellite observations of red and near infrared
+spectral reflectance, has been widely used as a proxy for biomass
 production \citep{Camps-Valls2021, Paruelo2016, Helman2014}. For Chile's
-cultivated land, \citet{Zambrano2018} introduced the zcNDVI for
-assessing seasonal biomass production in response to drought. Using this
-information, we can advance our understanding of the impact of drought
-on ecosystems.
+cultivated land, \citet{Zambrano2018} used the zcNDVI for assessing
+seasonal biomass production in response to drought. Comparing the
+various meteo-related and vegetation-based drought indices, we can
+further our understanding of the impact of drought on ecosystems.
 
@@
 on vegetation and water storage. North-central Chile has faced a
-persistent precipitation deficit since 2010, defined as a mega drought.
-\citep{Garreaud2017}, which has impacted the Chilean ecosystem. This
-megadrought was defined by the Standardized Precipitation Index (SPI) of
-twelve months in December having values below one standard deviation.
-Some studies have addressed how this drought affects single ecosystems
-in terms of forest development \citep{Miranda2020, Venegas2018}, forest
-fire occurrence \citep{UrrutiaJalabert2018}, and crop productivity
-\citep{Zambrano2023, Zambrano2018, Zambrano2016}. We found one study
-regarding land cover and drought in Chile. The study by
-\citet{Fuentes2021} evaluates water scarcity and land cover change in
-Chile between 29° and 39° of south latitude. \citet{Fuentes2021} used
-the SPEI of one month for evaluating drought, which led to misleading
-results. For example, they did not find a temporal trend in the SPEI but
-found a decreasing trend in water availability and an increase trend on
-AED, which in turn should have been capable of being captured with
-longer time scales of the SPEI. The term ``megadrought'' in Chile is
-used to describe a prolonged water shortage that lasts for several
-years, resulting in a permanent deficit that impacts the hydrological
-system \citep{Boisier2018}. Therefore, it is crucial to evaluate
-temporal scales that consider the cumulative impact over a period of
-several years. The association between drought and the environment in
-Chile is not well comprehended. Hence, it is imperative to acquire a
-more profound comprehension of the manner in which climatic and soil
-moisture droughts influence environmental dynamics, in order to make
-well-informed decisions on adaptation strategies.
+persistent precipitation deficit since 2010, defined as a mega-drought
+\citep{Garreaud2017}, which has impacted the Chilean ecosystem and
+consequently makes it highly vulnerable to climate change
+\citep{Barria2021, AlvarezGarreton2021}. This mega-drought was defined
+by the annual time series of the Standardized Precipitation Index (SPI)
+at a time scale of twelve months at the end of each year (December) when
+having values below one standard deviation. Some studies have addressed
+how this drought affects single ecosystems in terms of forest growth
+\citep{Miranda2020, Venegas2018}, forest fire occurrence
+\citep{UrrutiaJalabert2018}, and crop productivity
+\citep{Zambrano2023, Zambrano2018, Zambrano2016}. The term
+``mega-drought'' is used in Chile to describe a prolonged water shortage
+that lasts for several years, resulting in a permanent deficit that
+impacts the hydrological system \citep{Boisier2018}. Therefore, it is
+crucial to evaluate temporal scales that consider the cumulative impact
+over a period of several years. In Chile, the relationship between
+drought and the environment remains poorly understood. Hence, we aim to
+contribute to understanding how climatic and soil moisture droughts
+influence ecosystem dynamics in order to provide useful information that
+helps for a better understanding of ecological droughts and, at the same
+time, helps to make well-informed decisions on adaptation strategies.
 
@@
 from north to south and east to west \citep{Aceituno2021}
-(Figure~\ref{fig-studyArea} a), which determines its great ecosystem
-diversity \citep{Luebert2022} (Figure~\ref{fig-studyArea} c). The Andes
-Mountains are a main factor in climate latitudinal variation
-\citep{Garreaud2009}. ``Norte Grande'' and ``Norte Chico'' predominate
-in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At
-the south of ``Norte Chico,'' the climate changes to an arid steppe with
-cold temperatures (Bsk). In these two northern regions, the land is
-mostly bare, with a small surface of vegetation types such as shrubland
-and grassland. In the zones ``Centro'' and the north half of ``Sur,''
-the main climate is Mediterranean, with warm to hot summers (Csa and
-Csb). Land cover in ``Centro'' comprises a significant amount of
-shrubland and savanna (50\%), grassland (16\%), forest (8\%), and
-croplands (5\%). An oceanic climate (Cfb) predominates in the south of
-``Sur'' and the north of ``Austral.'' Those zones are high in forest and
-grassland. The southern part of the country has a tundra climate, and in
-``Austral,'' it is a cold semi-arid area with an extended surface of
-grassland, forest, and, to a lesser extent, savanna.
+(Figure~\ref{fig-studyArea}a), which determines its great ecosystem
+diversity \citep{Luebert2022} (Figure~\ref{fig-studyArea}c). The Andes
+Mountains are a main factor in climate variation \citep{Garreaud2009}.
+For an aggregated overview of the results of the study, we used the five
+Chilean macrozones: ``Norte Grande,'' ``Norte Chico,'' ``Centro,''
+``Sur,'' and ``Asutral''. ``Norte Grande'' (17°34'--25°42'S) and ``Norte
+Chico'' (25°42'-32°8'S) predominate in an arid desert climate with hot
+(Bwh) and cold (Bwk) temperatures. At the south of ``Norte Chico,'' the
+climate changes to an arid steppe with cold temperatures (Bsk). In these
+two northern regions, the land is mostly bare, with a small surface of
+vegetation types such as shrubland and grassland. In the macrozones
+``Centro'' (32°08'-36°12'S) and the northern half of ``Sur,''
+(36°12'-43°48'S) the main climate is Mediterranean, with warm to hot
+summers (Csa and Csb). Land cover in ``Centro'' comprises a significant
+amount of shrubland and savanna (50\%), grassland (16\%), forest (8\%),
+and croplands (5\%). An oceanic climate (Cfb) predominates in the south
+of ``Sur'' and the north of ``Austral'' (43°48'-56°00'S) Those zones
+have a large areal extent of forest and grassland. The southern part of
+the country has a tundra climate, and in ``Austral,'' it is a cold
+semi-arid area with an extended surface of grassland, forest, and, to a
+lesser extent, savanna.
 
@@
 
-To analyze land cover change, we use the classification scheme by the
+To analyze land cover change, we used the classification scheme by the
 IGBP (International Geosphere-Biosphere Programme) from the product
-MCD12Q1 collection 6.1 from MODIS. The MCD12Q1 has a yearly frequency
-from 2001 to 2022 and defines 17 classes. To derive a proxy for
-vegetation productivity, we used the Normalized Difference Vegetation
-Index (NDVI) from the product MOD13A3 collection 6.1 from MODIS
-\citep{Didan2015}. MOD13A3 provides vegetation indices at 1km of spatial
-resolution and monthly frequency. The NASA EOSDIS Land Processes
-Distributed Active Archive Center (LP DAAC), USGS Earth Resources
-Observation and Science (EROS) Center, Sioux Falls, South Dakota,
-provided the MOD13A3 and MCD12Q1 from the online Data Pool, accessible
-at https://lpdaa.usgs.gov/tools/data-pool/.
+MCD12Q1 Collection 6.1 from MODIS. The MCD12Q1 product is produced for
+each year from 2001 to 2022 and defines 17 classes (see Table S1). To
+maintain our focus on a large scale and follow the FAO classification
+\citep{FAO2022}, we considered native and planted forests as
+``forests'', which represent ecosystems dominated by larger trees. To
+derive a proxy for vegetation productivity, we used the Normalized
+Difference Vegetation Index (NDVI) from the product MOD13A3 Collection
+6.1 from MODIS \citep{Didan2015}. MOD13A3 provides vegetation indices
+with 1km spatial resolution and monthly frequency. The NASA EOSDIS Land
+Processes Distributed Active Archive Center (LP DAAC), USGS Earth
+Resources Observation and Science (EROS) Center, Sioux Falls, South
+Dakota, provided the MOD13A3 and MCD12Q1 from the online Data Pool,
+accessible at https://lpdaa.usgs.gov/tools/data-pool/.
 
@@
 precipitation, maximum and minimum temperature at 2 meters, and
-volumetric soil water layers between 0 and 100cm of depth (layer 1 to
-layer 3). Table \ref{tab-desEOD} shows a summary of the data and its
-main characteristics.
+volumetric soil water layers between 0 and 100 cm of depth (layer 1 to
+layer 3). \ref{tab-desEOD} shows a summary of the data and its main
+characteristics.
 
@@
 
-In order to compute the drought indices that use water demand, it is
-necessary to first calculate the AED. To do this, we employed the
-Hargreaves method \citep{Hargreaves1994, Hargreaves1985} by applying the
-following equation:
+To compute the drought indices that use water demand, it is necessary to
+first calculate the AED. To do this, we employed the Hargreaves method
+\citep{Hargreaves1994, Hargreaves1985} by applying the following
+equation:
 
@@
 temperature \((°C)\) at 2m. For calculating \(Ra\) we used the
-coordinate of the latitud of the centroid of each pixel. We chose the
-method of Hargreaves to estimate AED because of its simplicity, which
-only requires temperatures and extrarrestrial radiation. Also, it has
-been recommended over other methods (e.g., Penman-Monteith) when the
-access to climatic variables is limited \citep{Vicente-Serrano2014}.
+coordinate of the latitud of the centroid of each pixel as follow:
+
+\begin{equation}\protect\hypertarget{eq-Ra}{}{R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]}\label{eq-Ra}\end{equation}
+where
+
+\(Ra\): extraterrestrial radiation \([MJ\, m^{-2} day-1]\),\\
+\(G_{sc}\): solar constant = 0.0820 \([MJ\,m^{-2} min^{-1}]\),\\
+\(d_r\): inverse relative distance Earth-Sun,\\
+\(\omega_s\) sunset hour angle \([rad]\),\\
+\(\phi\): latitude \([rad]\),\\
+\(\delta\): solar declination \([rad]\).
+
+We chose the method of Hargreaves to estimate AED because of its
+simplicity, which only requires temperatures and extrarrestrial
+radiation. Also, it has been recommended over other methods (e.g.,
+Penman-Monteith) when the access to climatic variables is limited
+\citep{Vicente-Serrano2014}.
 
@@
 dataset and the MODIS product, with a monthly frequency for 1981--2023
-and 2000--2023, respectively. The dought indices correspond to a
-historical anomaly with regard to a variable (e.g., meteorological,
-vegetation, or soil moisture). To account for the anomaly, the common
-practice is to derive it following a statistical parametric methodology
-in which it is assumed that the statistical distribution of the data is
-known \citep{Heim2002}. A wrong decision is usually the highest source
-of uncertainty \citep{Laimighofer2022}. In the case of Chile, due to its
-high degree of climatic variability, it is complex to choose a proper
-distribution without previous research. Here, we follow a non-parametric
-methodology for the calculation of the drought indices, in a similar
-manner as the framework proposed by \citet{Farahmand2015};
-\citet{Hobbins2016};\citet{McEvoy2016}.
+and 2000--2023, respectively. The drought indices correspond to a
+historical anomaly of a variable (e.g., meteorological, vegetation, or
+soil moisture). To account for the anomaly, the common practice is to
+derive it following a statistical parametric method in which it is
+assumed that the statistical distribution of the data is known
+\citep{Heim2002}. A wrong decision in the statistical distribution is
+usually the highest source of uncertainty \citep{Laimighofer2022}. In
+the case of Chile, due to its high degree of climatic variability, it is
+difficult to choose a proper distribution without previous research that
+could be applicable throughout Chile. Here, we follow a non-parametric
+method for the calculation of the drought indices, in a similar manner
+as the framework proposed by \citet{Farahmand2015}.
 
@@
 supply and demand, we selected the SPEI \citep{Vicente-Serrano2010}. For
-SPEI, an auxiliary variable \(D = P-AED\) is calculated. Soil moisture
-is the main driver of vegetation productivity, particularly in semi-arid
+SPEI, an auxiliary variable \(D=P-AED\) is calculated. Soil moisture is
+the main driver of vegetation productivity, particularly in semi-arid
 regions \citep{Li2022}. Hence, for soil water drought, we used the SSI
-(Standardized Soil Moisture Index) \citep{Hao2013, AghaKouchak2014}. In
-our case, for the SSI, we used the average soil moisture from ERA5L at
-1m depth. Finally, for the proxy of productivity, we used the zcNDVI
-proposed by \citet{Zambrano2018}, which was derived from the monthly
-time series of NDVI derived from MOD13A1. All the indices are
-multi-scalar and can be used for the analysis of short- to long-term
-droughts.
-
-To derive the drought indices, first we must calculate the sum of the
-variables with regard to the time scale (s). In this case, for
+(Standardized Soil Moisture Index) \citep{Hao2013}. For the SSI, we used
+the average soil moisture from ERA5L at 1m depth. Finally, for the proxy
+of productivity, we used the zcNDVI \citep{Zambrano2018}, which was
+derived from the monthly time series of NDVI derived from MOD13A1. All
+the indices are multi-scalar and can be used for the analysis of short-
+to long-term droughts.
+
+To derive the drought indices, we first calculate the sum of the
+variables with regard to the time scale(s). In this case, for
 generalization purposes, we will use \(V\), referring to variables
 \(P\), \(AED\), \(D\), \(NDVI\), and \(SM\) (Table \ref{tab-desEOD}). We
-cumulated each \(V\) over the time series of \(n\) values (months), and
-for the time scales \(s\):
-
-\begin{equation}\protect\hypertarget{eq-sumvar}{}{A_{si} = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  }\label{eq-sumvar}\end{equation}
-
-The \(A_{si}\) corresponds to a moving window (convolution) that sums
-the variable for time scales \(s\) from the last month, month by month,
-until the first month in which it could sum for \(s\) months. An inverse
-normal approximation \citep{Abramowitz1968} obtains the empirically
-derived probabilities once the variable cumulates over time for the
-scale \(s\). Then, we used the empirical Tukey plotting position
+accumulated each over the time series of values (months), and for the
+time scales \(s\):
+
+\begin{equation}\protect\hypertarget{eq-sumvar}{}{A^s_i = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  }\label{eq-sumvar}\end{equation}
+
+The \(A^s_i\) corresponds to a moving window (convolution) that sums the
+variable for time scales \(s\). Start from the last month (n) and sum
+the variable for s months, then follow month by month until the first
+month in which it could sum for s months (n-s+1). For example, using as
+a variable the precipitation, a period of twelve months (n), and a time
+scale of three months (s), it will be:
+
+\[
+\begin{split}
+A^3_1 &= P_{oct} +P_{nov} +P_{dic} \\
+\vdots\,\,\, &= \,\,\,\vdots + \,\,\,\vdots + \,\,\,\vdots \\
+A^3_{10} &= P_{jan}+P_{feb} +P_{mar}
+\end{split}
+\]
+
+Then, an inverse normal approximation \citep{Abramowitz1968} obtains the
+empirically derived probabilities once the variable cumulates over time
+for the scale \(s\). Then, we used the empirical Tukey plotting position
 \citep{Wilks2011} over \(A_i\) to derive the \(P(a_i)\) probabilities
@@
 
-\begin{equation}\protect\hypertarget{eq-probPai}{}{P(A_i) = \frac{i-0.33}{n+0.33'}}\label{eq-probPai}\end{equation}
+\begin{equation}\protect\hypertarget{eq-probPai}{}{P(A^s_i) = \frac{i-0.33}{n+0.33'}}\label{eq-probPai}\end{equation}
 
@@
 
-\begin{equation}\protect\hypertarget{eq-DI}{}{DI(A_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}}\label{eq-DI}\end{equation}
+\begin{equation}\protect\hypertarget{eq-DI}{}{DI(A^s_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}}\label{eq-DI}\end{equation}
 
@@
 \(C_2 = 0.010328\), \(d_1 = 1.432788\), \(d_2 = 0.189269\), and
-\(d3 = 0.001308\). For \(P(A) \leq 0.5\),
-W=\(\sqrt{-2\cdot ln(P(A_i))}\) , and for \(P(A_i) > 0.5\), replace
-\(P(A_i)\) with \(1-P(A_i)\) and reverse the sign of \(DI(A_i)\).
+\(d3 = 0.001308\). For \(P(A^s_i) \leq 0.5\),
+W=\(\sqrt{-2\cdot ln(P(A^s_i))}\) , and for \(P(A^s_i) > 0.5\), replace
+\(P(A^s_i)\) with \(1-P(A^s_i)\) and reverse the sign of \(DI(A^s_i)\).
 
@@
 and 36 months at a monthly frequency for 1981--2023 in order to be used
-for short- to long-term evaluation of drought. In the case of the proxy
-of vegetation productivity (zcNDVI), it was calculated for a time scale
-of six months at monthly frequency for 2000--2023. For zcNDVI, we test
-time scales of 1, 3, 6, and 12 months in December and their correlation
-with net primary production (NPP) obtained from the MOD17A3HGF product
-from MODIS. We choose to use six months because r-squared with NPP
-increases from one to six months and from six to 12 months has little
-improvement. The r-squared were between 0.31 for forest and 0.72 for
-shrubland (see supplementary material in Section S5).
+for short- to long-term evaluation of drought.
+
+For the proxy of vegetation productivity, we chose the time scale that
+best correlates with annual net primary productivity (NPP) across
+continental Chile. For this purpose, we calculated the zcNDVI for time
+scales of 1, 3, 6, and 12 months in December and compared it with the
+annual NPP. We used the NPP from the MOD17A3HGF \citep{Running2019}
+dataset (MODIS). We choose to use six months because the r-squared of
+zcNDVI with NPP highly increases from one to six months, but from six to
+12 months has a lower improvement. We obtained an r-squared of 0.31 for
+forest and 0.72 for shrubland (refer to the supplementary material in
+Section S5). Then, we chose the proxy of vegetation productivity for six
+months, which we will name zcNDVI hereafter. It was calculated at a
+monthly frequency for 2000--2023.
 
@@
 Sen's slope \citep{Sen1968}. Some of the advantages of applying this
-methodology are that the Sen's slope is not affected by outliers as
-regular regression does, and it is a non-parametric method that is not
-influenced by the distribution of the data. We applied Mann-Kendall to
-see if the trend was significant and Sen's slope to estimate the
-magnitude of the trend. We did this to the six time scales from 1981 to
-2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI. Thus,
-we have trends per index and time scale (24 in total). Then, we
-extracted the trend aggregated by macrozone and per land cover persitent
-macroclasses.
+methodology are that the Sen's slope is not affected by outliers unlike
+regular regressions does, and it is a non-parametric method that is not
+influenced by the distribution of the data. We applied the Mann-Kendall
+test to see if the trend was significant and Sen's slope to estimate the
+magnitude of the trend. We did this for the indices SPI, EDDI, SPEI, and
+SSI using the six time scales with data from 1981 to 2023 (monthly
+frequency), resulting in 24 trends (per index and time scale). Then, we
+extracted the trend aggregated by each of the five macrozones: ``Norte
+Grande'' to ``Asutral,'' and per land cover type: grassland, forest,
+cropland, shrubland, savanna, and barren land
+(Figure~\ref{fig-studyArea}d).
 
@@
 for studies of drought and land cover in Chile
-\citep{Fuentes2021, Zambrano2018}. From the 17 classes, we regrouped
-into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to
-shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14
-to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to
-water bodies. Thus, we have a land cover raster time series with the ten
-macroclasses for 2001 and 2023. We validate the land cover macroclasses
-regarding a highly detailed (30 m of spatial resolution) land cover map
-made for Chile by \citet{Zhao2016} for 2013-2014. Our results showed a
-global accuracy of \textasciitilde0.82 and a F1 score of
-\textasciitilde0.66. Section S2 in the Supplementary Material shows the
-procedure for validation.
+\citep{Fuentes2021, Zambrano2018}. We regrouped the 17 classes into ten
+macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9
+to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
+13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
+Thus, we have a land cover raster time series with the ten macroclasses
+for 2001 and 2023. We validate the land cover macroclasses regarding a
+highly detailed (30 m of spatial resolution) land cover map made for
+Chile by \citet{Zhao2016} for 2013-2014. Our results showed a global
+accuracy of \textasciitilde0.82 and a F1 score of \textasciitilde0.66.
+Section S2 in the Supplementary Material shows the procedure for
+validation.
 
@@
 
-Later in this study, we will examine the variation in vegetation
-productivity across various land cover types and how water demand and
-supply, and soil moisture affect it. In order to avoid variations due to
-a change in the land cover type from year-to-year that will wrongly
-impact NDVI, we developed a persistence mask for land cover for
-2001--2022. Thereby, we reduce an important source of variation on a
-regional scale. Therefore, we generated a raster mask for IGBP MODIS per
-pixel using macroclasses that remained unchanged for at least 80\% of
-the years (2001--2022). This enabled us to identify regions where the
-land cover macroclasses are persistent.
+To assess how water demand and supply, and soil moisture affect the
+variation in vegetation productivity across various land cover types, we
+avoid analyzing areas that have major land cover changes for 2021--2022.
+To assess how zcNDVI varied irrespective of land cover change, we
+developed a persistence mask for land cover, which only retains pixels
+for which the macroclass remained the same for at least 80\% of the
+years (2001--2022) (Figure~\ref{fig-studyArea}d).
 
@@
 
-We wanted to explore the relationship between the trend in land cover
-classes and the trend in the drought indices. For this purpose, in order
-to have more representative results, we conducted the analysis over
-sub-basins within continental Chile. We use 469 basins, which have a
-surface area between 0.0746 and ~24,000 (\(km^2\)), and a median area of
-1,249 (\(km^2\)). For each basin, we calculate the relative trend per
+The goal of this section is to identify which drought indices and time
+scales have a major impact on changes in land cover type. We examined
+the relationship between the trend in land cover classes and the trend
+in drought indices. To have more representative results, we conducted
+the analysis over sub-basins within continental Chile. We used 469
+basins, which have a surface area between 0.0746 and 24,000 km2 and a
+median area of 1,249 km2. For each basin, we calculated the trend per
 land cover type, considering the proportion of the type relative to the
 total surface of the basin. Then, we extracted per basin the average
-trend of the drought indices SPI, SPEI, EDDI, SSI, and all their time
-scales 1, 3, 6, 12, 24, and 36. Also, we extracted the average trend in
-the proxy of vegetation productivity (zcNDVI). We wanted to analyze
-which drought indices and time scales have a major impact on changes in
-land cover type.
-
-We have 25 predictors, including drought indices and vegetation
-productivity. We analyzed the 25 predictors per type of landcover, thus
-running six random forest models. Random forest uses multiple decision
-trees and allows for classification and regression. Some advantages are
-that it allows to find no linear relationship, reduces overfitting, and
-allows to derive the variable importance. We used random forests for
-regression and trained 1000 forests. To obtain more reliable results, we
-resampled by creating ten folds, running a random forest per fold, and
-calculating the r-squared (rsq), root mean square error (RMSE), and
-variable importance.
-
-The variable importance helps for a better understanding of the
-relationships by finding which variable has a higher contribution to the
-model. We calculate the variable's importance by permuting out-of-bag
-(OOB) data per tree and computing the mean standard error in the OOB.
-After permuting each predictor variable, we repeat the process for the
-resting variable. We repeated this process ten times (per fold) to
+trend (Sen's slope) of the drought indices SPI, SPEI, EDDI, SSI, and all
+their time scales 1, 3, 6, 12, 24, and 36. Also, we extracted the
+average trend in the proxy of vegetation productivity (zcNDVI)
+
+Random forest \citep{Ho1995} employs multiple decision trees, allowing
+for classification and regression. Some advantages include the ability
+to find non-linear relationships, reduce overfitting, and derive
+variable importance. We used the regression random forest to model the
+trends in land cover per macroclass, using drought indices as
+predictors. This included the four drought indices per six time scales
+and the zcNDVI, totaling 25 predictors. As a result, we created six
+random forest models, one per trend in land cover macroclass. We trained
+1000 forests in a resampled scheme to obtain more reliable results
+regarding variable importance. We resampled by creating ten folds,
+running a random forest per fold, and calculating the r-squared (rsq),
+root mean square error (RMSE), and variable importance. The variable
+importance helps for a better understanding of the relationships by
+finding which variable has a higher contribution to the model. Thus, we
+calculate the variable's importance by permuting out-of-bag (OOB) data
+per tree and computing the mean standard error in the OOB. After
+permuting each predictor variable, we repeated the process for the
+remaining variable. We repeated this process ten times (per fold) to
 obtain the performance metrics (rsq, RMSE, and variable importance).
+
+Finally, we visually explore the connection between the SPI, EDDI, and
+SSI drought indices for short- and long-term changes in land cover. To
+do this, we compare the relative changes in land cover surface (in terms
+of the total surface area per land cover type and macrozone) with the
+drought indices of six (short-term) and thirty-six months (long-term).
+We created a scatterplot in which the x-axis shows drought indices
+trends and the y-axis shows land cover change trends.
 
@@
 
-We analyzed the trend of vegetation productivity over the unchanged land
-cover macroclasses. To achieve this, we used the persistent mask of land
-cover macroclasses. This way, we tried to reduce the noise in the
-vegetation due to a change in land cover from year to year. We used the
-zcNDVI as a proxy of vegetation productivity. In Chile's cultivated
-land, \citet{Zambrano2018} introduced the zcNDVI for assessing seasonal
+For each land cover macroclass, we analyzed the trend of vegetation
+productivity over the unchanged land cover macroclasses. To achieve
+this, we used the persistent mask of land cover macroclasses , thus
+reducing the possibility of evaluating productivity trends that are due
+to year-to-year variation in land cover. We used the zcNDVI as a proxy
+of vegetation productivity. To assess productivity in Chile's cultivated
+land, Zambrano et al.~(2018) used the zcNDVI for assessing seasonal
 biomass production in relation to climate.
@@
 moisture and their correlation with vegetation productivity. The
-objective is to determine the impact of soil moisture and water demand
-and supply on vegetation productivity. We want to address three main
-questions: Which of the drought variables---supply, demand, or soil
-moisture---most helps to explain the changes in plant productivity? How
-do the short- to long-term time scales of the drought variable affect
-vegetation productivity in Chile? And finally, how strong is the
-relationship between the variables and the drought index? Thus, we will
-be able to advance in understanding how climate is affecting vegetation,
-considering the impact on the five land cover types: forest, cropland,
-grassland, savanna, and shrubland.
+objective is to determine to what extent soil moisture and water demand
+and supply affect vegetation productivity, thus addressing three main
+questions: 1) Which of the drought variables---supply, demand, or soil
+moisture---helps most in explaining the changes in vegetation
+productivity? How do the short- to long-term time scales of the drought
+variable affect vegetation productivity in Chile? 2) How strong is the
+relationship between the variables and the drought index? And finally,
+3) how does the correlation vary per-land cover type? Answering these
+questions should advance our understanding of how climate is affecting
+vegetation, considering the impact on the five land cover types: forest,
+cropland, grassland, savanna, and shrubland.
 
@@
 \citet{Meroni2017} which compared the SPI time-scales with the
-cumulative FAPAR (Fraction of Absorbed Photosynthetically Active
+cumulative fAPAR (fraction of Absorbed Photosynthetically Active
 Radiation). We performed a pixel-to-pixel linear correlation analysis
@@
 Figure~\ref{fig-trendDI} shows the spatial variation of the trend for
-the drought indices from short- to long-term scales. SPI and SPEI have a
+the drought indices from short- to long-term scales. The white space on
+the maps indicates a not significant trend. SPI and SPEI have a
 decreasing trend from ``Norte Chico'' to ``Sur.'' However, there is an
@@
 higher time scales. The SSI indicates that in ``Norte Grande,'' there
-are surfaces that have increased in the soutwest part and in the
-northeast have decreased, and is shown for all time scales. Similar to
-SPI and SPEI, SSI decreases at higher time scales. EDDI showed a
-positive trend for the whole of continental Chile, with a higher trend
-toward the north and a descending gradient toward the south. The degree
-of trend increases at higher time scales.
+are surfaces that have increased in the southwest and in the northeast
+have decreased, and are shown for all time scales. Similar to SPI and
+SPEI, SSI decreases at higher time scales. EDDI showed a positive trend
+for the whole of continental Chile, with a higher trend toward the north
+and a descending gradient toward the south. The degree of trend
+increases at higher time scales.
 
@@
 
-\includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}
+\includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}
 
@@
 
-\includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}
+\includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}
 
@@
 
-\includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}
+\includegraphics{../output/figs/trend_raster_zcSM_1981-2023.png}
 
@@
 \end{minipage}%
-%
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_zcSM_1981-2023.png}
-
-}
-
-}
-
-\subcaption{\label{fig-trendDI-4}SSMI (Standardized Soil Moisture
-Index)}
-\end{minipage}%
 
@@
 reached the lowest trend for SPI, SPEI, and SSI are ``Norte Chico'' and
-``Centro,'' where the indices also decrease at longer time scales.
-Potentially explained due to the prolonged reduction in precipitation
+``Centro,'' where the indices also decrease at longer time scales. This
+may potentially be explained by the prolonged reduction in precipitation
 that has affected the hydrological system in Chile. At 36 months, it
@@
 
-{\centering \includegraphics{../output/figs/LC_pers80_per_macrozone.png}
+{\centering \includegraphics{../output/figs/proportion_landcover_macroclass_2001-2022.png}
 
@@
 \caption{\label{fig-LCprop}Proportion of land cover class from the
-persistent land cover for 2001-2022 (\textgreater80\%) per macrozone}
+persistent land cover for 2001-2022 (\textgreater80\%) per macrozone and
+land cover macroclass.}
 
@@
 
-For vegetation, we obtained and use hereafter five macroclasses of land
+For vegetation, we obtained and used hereafter five macroclasses of land
 cover from IGBP MODIS: forest, shrubland, savanna, grassland, and
@@
 Figure~\ref{fig-studyArea}d shows the macroclasses of land cover
-persistance (80\%) during 2021--2022, respectively (Table
+persistence (80\%) during 2021--2022, respectively (Table
 \ref{tab-landcoverSurf}). Within continental Chile, barren land is the
@@
 the proportion of surface per land cover class and macrozone, derived
-from the persistance mask over continental Chile.
+from the persistence mask over continental Chile.
 
@@
 \(km^2 yr^{-1}\). In the ``Centro'' and ``Sur,'' there are changes with
-an important reduction in savanna (136 to 318 \(km^2 yr^{-1}\) ), and an
-increase in shrubland and grassland. Showing a change for more dense
-vegetation types. The area under cultivation (croplands) appears to be
-shifting from the ``Centro'' to the ``Sur.'' Also, there is a high
-increase in forest (397 \(km^2 yr^{-1}\) ) in the ``Sur,'' seemingly
-replacing the savanna lost (Table \ref{tab-landcoverTrend}).
+an important reduction in savanna with 136 and 319 \(km^2 yr^{-1}\),
+respectively, and an increase in shrubland and grassland, showing a
+change for more dense vegetation types. The area under cultivation
+(croplands) appears to be shifting from the ``Centro'' to the ``Sur.''
+Also, there is a high increase in forest (397 \(km^2 yr^{-1}\)) in the
+``Sur,'' seemingly replacing the savanna lost (Table
+\ref{tab-landcoverTrend}).
 
@@
 
+\begin{figure}[!ht]
+
+{\centering \includegraphics{../output/figs/bars_relative_importance_RF.png}
+
+}
+
+\caption{\label{fig-varImportance}cdcdcdc}
+
+\end{figure}
+
 \begin{table}[!ht]
@@
 \label{tab-landcoverTrendRF}
-\includegraphics[]{../output/figs/table_importance_trends_landcover_vs_drought.png}
+\includegraphics[]{../output/figs/table_r2_most_important_variables.png}
 \end{table}
 
-According to Table \ref{tab-landcoverTrendRF}, the random forest models
-for estimating the landcover trend from the trends in drought indices
-reach an r-squared between 0.32 and 0.39 for the types of forest,
-grassland, savanna, shrubland, and barren land. It is more likely that
-short- and medium-term increases in AED (EDDI-6 and EDDI-12) and
-short-term precipitation deficits (SPI-6 and SPEI-6) contributed to
-changes in grassland and bare land. The short-term increase of AED
-(EDDI-3 and EDDI-6) and the longer duration of the precipitation deficit
-(SPI-24, SPI-36, and SPEI-36) most likely contribute to the changes in
-shrubland. The changes in savanna are associated with a short- and
-long-term increase in AED and a three-year precipitation deficit
-(SPI-36). The increase in cumulative AED from 12 to 36 months is the
-strongest associated variable that contributes to changes in forests,
-followed by the reduction of soil moisture over six and 36 months. The
-supplementary material in Section S3 provides further details about the
-variable's importance.
+The random forest models for estimating the land cover trend based on
+the trends in drought indices reach an \(R^2\) between 0.32 and 0.39 for
+the land cover types (Table \ref{tab-landcoverTrendRF}) excluding
+cropland. It is more likely that short- and medium-term increases in AED
+(EDDI-6 and EDDI-12) and short-term precipitation deficits (SPI-6 and
+SPEI-6) are associated to changes in grassland and bare land. The
+short-term increase of AED (EDDI-3 and EDDI-6) and the longer duration
+of the precipitation deficit (SPI-24, SPI-36, and SPEI-36) were the most
+important variables that correlated with changes in shrubland. The
+changes in savanna are associated with a short- and long-term increase
+in AED and a three-year precipitation deficit (SPI-36). The increase in
+cumulative AED from 12 to 36 months is the strongest associated variable
+that contributes to changes in forests, followed by the reduction of
+soil moisture over six and 36 months. The supplementary material in
+Section S3 provides further details about the variable's importance.
 
@@
 
-We study the connection between the SPI, EDDI, and SSI drought indices
-and changes in land cover in Figure~\ref{fig-TrendsLandDrought}. To do
-this, we compare the relative changes in land cover (in terms of the
-total surface area per land cover type and macrozone) over six and
-thirty-six months. Figure~\ref{fig-TrendsLandDrought} shows that the
-forest in the ``Sur,'' shrubland and grassland in ``Centro,'' barren
-land in ``Norte Chico,'' and savanna in ``Austral'' showed an increase
-in the surface of landcover associated with an increase in EDDI. Savanna
-in ``Centro,'' ``Sur,'' and ``Norte Chico'' decreases with the increase
-in EDDI. The SPI and SSI showed similar behavior regarding the trend in
+Figure~\ref{fig-TrendsLandDrought} shows the connection between the SPI,
+EDDI, and SSI drought indices and changes in land cover. Forest in the
+``Sur,'' shrubland and grassland in ``Centro,'' barren land in ``Norte
+Chico,'' and savanna in ``Austral'' showed an increase in land cover
+extent, which was associated with an increase in EDDI. Savanna in
+``Centro,'' ``Sur,'' and ``Norte Chico'' decreases with the increase in
+EDDI. The SPI and SSI showed similar behavior regarding the trend in
 land cover type. A decrease in SPI and SSI is associated with an
-increase in the surface in shurubland and grassland in ``Centro,''
-forest in ``Sur,'' and barren land in ``Norte Chico,'' as well as a
-decrease trend in savanna in ``Norte Chico,'' ``Centro,'' and ``Sur.''
+increase in the surface in shrubland and grassland in ``Centro,'' forest
+in ``Sur,'' and barren land in ``Norte Chico,'' as well as a decrease
+trend in savanna in ``Norte Chico,'' ``Centro,'' and ``Sur.''
 
@@
 The temporal variation within the macrozones is shown in
-Figure~\ref{fig-zcNDVI_var}b). There is a negative trend in ``Norte
+Figure~\ref{fig-zcNDVI_var}b. There is a negative trend in ``Norte
 Chico'' with -0.035 and ``Centro'' with -0.02 per decade. Vegetation
@@
 \caption{\label{fig-corTimeScale}Time scales per drought index that
-reach the maximum coefficient of determination}
+reach the maximum coefficient of determination. White spaces indicate no
+significant correlation.}
 
@@
 scales and drought index that reach the maximum coefficient of
-determination}
+determination. White spaces indicate no significant correlation.}
 
@@
 higher correlation. It has a similar correlation to SPI and SPEI for
-``Norte Chico'' and ``Sur,''~ but has improvements for ``Sur.''
+``Norte Chico'' and ``Sur,'' but has improvements for ``Sur.''
 
@@
 
-In Table \ref{tab-corlandcover}, we aggregate per macrozone and
-landcover the correlation analysis presented in
+In Table \ref{tab-corlandcover}, we aggregate per macrozone and land
+cover the correlation analysis presented in
 Figure~\ref{fig-corTimeScale} and Figure~\ref{fig-corPerson}. According
@@
 
-\hypertarget{the-main-drivers-of-drought-in-chile}{%
-\subsection{The main drivers of drought in
-Chile}\label{the-main-drivers-of-drought-in-chile}}
-
-\citet{Vicente-Serrano2022}, in a study at the global scale of drought
-trends, indicates that there have not been significant trends in
-meteorological drought since 1950. Also, state that the increase in
-hidrological trend in some parts of the globe (northeast Brazil and the
-Mediterranean region) is related to changes in land cover and
-specifically to the rapidly increasing irrigated area, which
-consequently increases water extraction. \citet{Kogan2020} analyzed the
-agricultural drought impact globally and in the main grain producer
-countries, finding that ``since 1980, the Earth warming has not changed
-the drought area or intensity.'' In our study, we took into account the
-variation in vegetation productivity in Chile, specifically in areas
-without any changes in land cover, to prevent any misleading conclusions
-about the increase in water demand due to land cover change. Our results
-show a contrasting perspective. The SPI, SPEI, and SSI (water supply)
-showed a decrease in trends, except for the southern part, and an
-increase in EDDI (water demand). The trend, positive or negative, was
-stronger as the time scales increased. Trends in the long term (e.g., 36
-months) are evidence of how human-induced climate change is affecting
-Chile, which seems to be due to an intense hydrological drought
-resulting from the persistence of the precipitation deficit. We found
-that there has been a significant trend in the decline of vegetation
+\hypertarget{vegetation-water-demand-and-its-relation-to-drought}{%
+\subsection{Vegetation water demand and its relation to
+drought}\label{vegetation-water-demand-and-its-relation-to-drought}}
+
+In our study, we took into account the variation in vegetation
+productivity in Chile, specifically in areas without any changes in land
+cover, to prevent any misleading conclusions about the increase in water
+demand due to land cover change. Our results show a contrasting
+perspective regarding the evidence provided by
+\citet{Vicente-Serrano2022} on a global scale, who indicates that the
+increase in drought is led by an increase in agricultural land, which in
+turn increases water demand.
+
+Our results indicate that except for the southern part of the country,
+the SPI, SPEI, and SSI (water supply) showed declining trends, while the
+EDDI (water demand) increased across continental Chile. The trends in
+water demand and supply were stronger as the time scales increased,
+indicating a long-term reduction in water supply (except for the
+southern part) and an increase in water demand by the atmosphere. Also,
+we found that there has been a significant declining trend in vegetation
 productivity (zcNDVI) since 2000 for the north-central part of the
-country, which has been extreme between 2020 and 2022 and has impacted
-natural and cultivated land. Additionally, we demonstrated that the
-drought, primarily due to an increase in AED, accounts for about 30\% of
-the changes in land cover types (excluding croplands). These changes are
-associated with a decrease in water demand from vegetation. Moreover,
-the most water-demanding type, cropland, showed a decrease in the
-north-central region, while barren land showed an increase. The
-north-central part of the country primarily experienced these changes
-due to a higher increase in AED. Thus, we have evidence of a significant
-decline in water supply and an increase in AED for the north-central
-part of Chile, which show to be the most relevant variables for drought
-conditions. Some questions arise regarding what is occurring with the
-cultivated land. We used the unchanged land cover to ensure that an
-increase in surface area is not considered in the trend analysis. For
-croplands, it could happen that some areas have changed the types of
-crops for others with higher water demand, which consequently increases
-water demand. However, this effect should be minor compared to the
-decrease in water supply and increase in water demand at this scale of
-analysis.
-
-This shows that the main cause of the hydrological drought in Chile was
-a steady drop in water supply made worse by an increase in AED, but it
-seems that in zones most affected by drought, the main cause is not an
-increase in vegetation water demand due to an intensification of
-cultivated land (e.g., an increase in irrigated crops). North-central
-Chile has experienced a decline in vegetation productivity across land
-cover types, which is primarily attributable to variations in water
-supply and soil moisture. An increase in water demand, such as an
-increase in the surface area of irrigated crops, could strengthen this
-trend. But it is out of the scope of this study. Future work should
-focus on the regions where the drought has been more severe and has a
-high proportion of irrigated crops to get insight on the real impact of
-irrigation on those zones.
-
-\hypertarget{land-cover-sensitivity-to-drought}{%
-\subsection{Land cover sensitivity to
-drought}\label{land-cover-sensitivity-to-drought}}
-
-We analyzed two main impacts of drought on land cover. First, the
-attribution of drought to the change in surface area per land cover
-type. Drought accounts for about 30\% of the surface change per land
-cover type, with the exception of croplands. The main variables
-associated with these changes are the increase in AED and, in second
-place, the decrease in precipitation. Second, we analyzed the time
-series of drought indices and vegetation productivity per land cover
-type. In this case, the most important variables that had an impact on
-zcNDVI were the soil moisture deficit, followed by the precipitation
-deficit, and in third place, AED.
-
-In a study in the Yangtze River Basin in China, \citet{Jiang2020}
+country, which reached its lowest level between 2020 and 2022 and has
+impacted natural and cultivated land. Further, croplands showed a
+decrease in surface area for the north-central region, while barren land
+increased. We link these changes to a decrease in the water demand from
+vegetation because, despite the increase in AED, the surface area for
+water-demanding vegetation is declining as well as the biomass
+production. However, some questions arise regarding what is occurring
+with the cultivated land. Evidence suggests that higher-water-demanding
+crops have replaced croplands in the Petorca basin (central Chile),
+leading to an increase in water abstraction
+\citep{Munoz2020, Duran2020}. Nonetheless, at this scale of analysis,
+the effect of higher crop water demand on drought is minor compared to
+the decrease in water supply and increase in AED over all land cover
+types.
+
+The long-scale trends (e.g., 36 months) demonstrate the impact of
+human-induced climate change on water availability in Chile, potentially
+due to an intense hydrological drought stemming from the ongoing
+precipitation deficit and rising AED. But it is likely that in zones
+most affected by drought, the main cause is not an increase in
+vegetation water demand due to an intensification of cultivated land
+(e.g., an increase in irrigated crops) like in other parts of the globe
+\citep{Vicente-Serrano2020}. North-central Chile has experienced a
+decline in vegetation productivity across land cover types, which is
+primarily attributable to variations in water supply and soil moisture.
+An increase in water demand, led by an increase in the surface area of
+irrigated crops or the change to more water-demanding crops, could
+strengthen this trend, however, it escapes the scope of this study.
+Future work should focus on the regions where the drought has been more
+severe and has a high proportion of irrigated crops to get insight on
+the real impact of irrigation on ecosystems in those zones.
+
+\hypertarget{sensitivity-of-land-cover-vegetation-to-short--and-long-term-drought}{%
+\subsection{Sensitivity of land cover vegetation to short- and long-term
+drought}\label{sensitivity-of-land-cover-vegetation-to-short--and-long-term-drought}}
+
+We analyzed the time series of drought indices and vegetation
+productivity per land cover type. Our results indicate that forest is
+the type most resistant to drought, and shrublands, savannas, and
+croplands have higher sensitivity.
+
+In their study in the Yangtze River Basin in China, \citet{Jiang2020}
 analyzed the impact of drought on vegetation using the SPEI and the
-Enahanced Vegetation Index (EVI). They found that cropland was more
+Enhanced Vegetation Index (EVI). They found that cropland was more
 sensitive to drought than grassland, showing that cropland responds
@@
 \citep{Zambrano2016, Potopova2015, Dai2020, Rhee2010}. Short-term
-precipitation deficits impact soil water, and thus less water is
+precipitation deficits have an impact on soil water, so less water is
 available for plant growth. However, we found that in ``Norte Chico,''
 an SPI-36 and SPEI-12 had a higher impact, which are associated with
-hydrological drought (long-term), and in ``Centro,'' an SPI-12 and
-SPEI-12. Thus, we attribute this impact to the hydrological drought that
-has decreased groundwater storage \citep{Taucare2024}, which in turn is
-impacted by long-term deficits, and consequently, the vegetation is more
-dependent on groundwater. In ``Sur'' and ``Austral,'' the correlations
-between drought indices and vegetation productivity decrease, as do the
-time scales that reach the maximum r-squared. The possible reason for
-this is that the most resistant types, forest and grassland, predominate
-south of ``Centro.'' Also, drought episodes have been less frequent and
-intense. The drought episodes have had a lower impact on water
+long-term water deficit, and in ``Centro,'' an SPI-12 and SPEI-12. Thus,
+we hypothesize that this impact could be attributed to the hydrological
+drought that has decreased groundwater storage \citep{Taucare2024},
+which in turn is impacted by long-term deficits, and consequently, the
+vegetation is more dependent on groundwater. In ``Sur'' and ``Austral,''
+the correlations between drought indices and vegetation productivity
+decrease, as do the time scales that reach the maximum r-squared. The
+possible reason for this is that the most resistant types, forest and
+grassland, predominate south of ``Centro.'' Also, drought episodes have
+been less frequent and intense and have had a lower impact on water
 availability for vegetation.
@@
 According to \citet{Senf2020}, severe drought conditions in Europe are a
-significant cause of tree mortality. However, we discovered that
-forests, as the most resilient land cover class to drought, experience
-less variation in drought indices. Supporting this is
-\citet{Fathi-Taperasht2022}, who asserts that Indian forests are the
-most drought-resistant and recover rapidly. Similarly, the work of
-\citet{Wu2024}, who analyzed vegetation loss and recovery in response to
-meteorological drought in the humid subtropical Pearl River basin in
-China, indicates that forests showed higher drought resistance. Using
-Vegetation Optical Depth (VOD), kNDVI, and EVI, \citet{Xiao2023} tests
-the resistance of ecosystems and finds that ecosystems with more forests
-are better able to handle severe droughts than croplands. They attribute
-the difference to a deeper rooting depth for trees, a higher water
-storage capacity, and different water use strategies between forest and
-cropland \citep{Xiao2023}. In contrast, \citet{Venegas2022}, who studied
-Cryptocarya alba and Beilschmiedia miersii (both from the Lauraceae
-family) that live in sclerophyllous forests in Chile, found that the
-trees' overall growth had slowed down. This could mean that the natural
-dynamics of their forests have changed. They attributed it to the
-cumulative effects of the unprecedented drought (i.e., hydrological
-drought).
-
-Thus, we attribute that forest to being the most resistant to drought,
-due to the fact that most of the species comprising it are highly
-resilient to water scarcity compared to the other land cover classes.
-Nonetheless, if we want to go deep in our analysis, we should use earth
-observation data that is able to capture a higher level of detail. For
-example, when we used MOD13A3 with a 1km spatial resolution to measure
-vegetation condition, it took the average condition of 1 square
-kilometer. Then, to use remote sensing to look at how a certain type of
-forest (like sclerophyllous forest) changes in response to drought on a
-local level, we should use operational products with higher spatial
-resolutions, like those from Landsat or Sentinel. This will let us do a
-more thorough analysis.
-
-\hypertarget{vegetation-productivity-and-drought.}{%
+significant cause of tree mortality. However, we discovered that forests
+are the most resilient land cover class to drought, with less variation
+in drought indices. Supporting this is \citet{Fathi-Taperasht2022}, who
+assert that Indian forests are the most drought-resistant and recover
+rapidly. Similarly, the work of \citet{Wu2024}, who analyzed vegetation
+loss and recovery in response to meteorological drought in the humid
+subtropical Pearl River basin in China, indicates that forests showed
+higher drought resistance. Using Vegetation Optical Depth (VOD), kNDVI,
+and EVI, \citet{Xiao2023} tested the resistance of ecosystems and found
+that ecosystems with more forests are better able to handle severe
+droughts than croplands. They attribute the difference to a deeper
+rooting depth for trees, a higher water storage capacity, and different
+water use strategies between forest and cropland \citep{Xiao2023}.
+
+In central Chile, \citet{Venegas2022} observed a significant decline in
+the overall growth of sclerophyllous moist forests (mediterranean
+forests), which they attributed to increased drought conditions. In
+another study, \citet{Fuentes2021} evaluated water scarcity and land
+cover change in Chile between 29° and 39° south latitude. They used the
+one-month SPEI for drought evaluation, which resulted in misleading
+results. For instance, they failed to identify a temporal trend in the
+SPEI, but they still observed a decline in water availability and a rise
+in AED, trends that should have been detectable if they were using
+longer SPEI time scales. Thus, according to the results presented in
+this study, for the assessment of drought, it is necessary to consider
+drought indices on a short- to long-scale basis.
+
+\hypertarget{vegetation-productivity-and-drought}{%
 \subsection{Vegetation productivity and
-drought.}\label{vegetation-productivity-and-drought.}}
-
-We found that the 12-month soil moisture deficit most affects the
-productivity of vegetation in all land cover types along Chile. The main
-external factors that affect biomass production by vegetation are actual
+drought}\label{vegetation-productivity-and-drought}}
+
+We found that the 12-month soil moisture deficit affected plant
+productivity in all land cover types in Chile. The main external factors
+that affect biomass production by vegetation are actual
 evapotranspiration and soil moisture, and the rate of ET in turn depends
@@
 moisture plays a key role in land carbon uptake and, consequently, in
-the production of biomass \citep{Humphrey2021}. Moreover,
-\citet{Zhang2022} indicate there is a bidirectional causality between
-soil moisture and vegetation productivity. Lastly, some studies have
-redefined agricultural drought as soil moisture drought from a
-hydrological perspective \citep{Loon2016, Samaniego2018}. Even though
-soil moisture is the external factor most determinant of vegetation
-biomass, there are multiple internal factors, such as species,
-physiological characteristics, and plant hydraulics, that would affect
-vegetation productivity. Because of that, we believe that agricultural
-drought, referring to the drought that impacts vegetation productivity,
-is the most proper term, as originally defined by \citet{Wilhite1985}.
-
-The study results showed that the soil moisture-based drought index
-(SSI) was better at explaining vegetation productivity across land cover
-macroclasses than meteorological drought indices like SPI, SPEI, and
-EDDI. In the early growing season and especially in irrigated rather
-than rainfed croplands, soil moisture has better skills than SPI and
-SPEI for estimating gross primary production (GPP). This according to
-\citet{Chatterjee2022} evaluation of the SPI and SPEI and their
-correlation with GPP in the CONUS. Also, \citet{Zhou2021} indicate that
-the monthly scaled Standardized Water Deficit Index (SWDI) can
-accurately show the effects of agricultural drought in most of China.
-\citet{Nicolai2017} also looked at the time-lag between the SWDI and the
-Vegetation Condition Index (VCI). They found that there was little to no
-time-lag in croplands but a greater time-lag in forests.
+the production of biomass \citep{Humphrey2021}. The study results showed
+that the soil moisture-based drought index (SSI) was better at
+explaining vegetation productivity across land cover macroclasses than
+meteorological drought indices like SPI, SPEI, and EDDI. According to
+\citep{Chatterjee2022} in the early growing season and especially in
+irrigated rather than rainfed croplands, soil moisture has better skills
+than SPI and SPEI for estimating gross primary production (GPP) . Also,
+\citet{Zhou2021} indicate that the monthly scaled Standardized Water
+Deficit Index (SWDI) can accurately show the effects of agricultural
+drought in most of China. \citet{Nicolai2017} also looked at the
+time-lag between the SWDI and the Vegetation Condition Index (VCI). They
+found that there was little to no time-lag in croplands but a greater
+time-lag in forests.
 
@@
 SSI had the best performance. In Chile, medium-term deficits of 12
-months are more relevant in the response of vegetation, which decreases
-to the south, and in the case of croplands, they seem to react in a
-shorter time, with six months (SSI-6) in ``Centro.'' This variation for
-croplands could be related to the fact that in ``Norte Chico,'' the
-majority of crops are irrigated, but to the south there is a higher
-proportion of rainfed agriculture, which is most dependent on the
-short-term availability of water. Rather, in the ``Norte Chico,'' the
-orchards are more dependent on the storage of water in dams of
-groundwater reservoirs, which are affected by long-term drought (e.g.,
-SPI-36).
+months are more relevant in the response of vegetation for all land
+cover types, which decreases to the south, and in the case of croplands,
+they seem to react in a shorter time, with six months (SSI-6) in
+``Centro.'' This variation for croplands could be related to the fact
+that in ``Norte Chico,'' the majority of crops are irrigated, but to the
+south there is a higher proportion of rainfed agriculture, which is most
+dependent on the short-term availability of water. Rather, in ``Norte
+Chico,'' the orchards are more dependent on irrigation, which in turn
+depends on the availability of storage water in dams of groundwater
+reservoirs, which are affected by long-term drought (e.g., SPI-36).
 
@@
 
-Different climate components, such as evaporative water demand, water
-supply, soil moisture, and their impact on vegetation, should be
-considered when evaluating the multi-dimensional nature of drought. For
-a better understanding of the propagation of drought \citep{VanLoon2012}
-from meteorological to hydrological drought, we should consider the
-climatic response at different time scales, ranging from short to long.
-Furthermore, we must make this information publicly available and update
-it frequently to aid in the development of adaptation policies. The
-drought observatory for agriculture and biodiversity of Chile (ODES)
-shares this information aggregated for multiple administrative and
-hydrological units across Chile
-(\url{https://odes-chile.org/app/unidades}), with the goal of helping to
-prepare for future climatic conditions.
+Our findings present valuable information for policymakers in developing
+adaptation strategies for droughts. Our results show that the different
+climate components, such as AED, water supply, soil moisture, and their
+impact on vegetation, should be considered when evaluating the
+multi-dimensional nature of drought. Also, for a better understanding of
+drought propagation \citep{VanLoon2012} from meteorological to
+agricultural and ecological drought, we should consider the climatic
+response at different time scales, ranging from short to long.
+Additionally, the spatiotemporal characteristics of our results allow us
+to distinguish distinct geographical contexts, recognizing the diversity
+in climate, but also shedding light on agricultural practices (ranging
+from irrigated to dryland farming), technological advancements in
+irrigation efficiency, and the region-specific capabilities for drought
+adaptation, including groundwater management and reservoir water
+storage. This information, combined with agricultural information and
+statistics, could provide a strong foundation for the development of
+science-based adaptation policies.
+
+In a commitment to fostering informed and dynamic adaptation efforts,
+our results are disseminated publicly and continually updated via the
+Drought Observatory for Agriculture and Biodiversity of Chile (ODES)
+(\url{https://odes-chile.org/app/unidades})
+\citep{Zambrano2023b, Kunst2023}. This initiative ensures the
+availability of extensive climate data, facilitating the development of
+adaptive strategies that are both responsive to the realities of
+different regions and grounded in the latest scientific understanding.
+The proactive sharing and updating of such data underscores its key role
+in enabling policymakers to craft adaptive measures that are finely
+tuned to the diverse and evolving landscapes of drought impacts.
+Furthermore, the recently promulgated law about climate change in Chile
+(Law 21.455, https://www.bcn.cl/leychile/navegar?idNorma=1177286), which
+aims to implement sectoral adaptation plans for agriculture, forests,
+and biodiversity, could benefit from this information.
 
@@
 
-There is a trend toward decreasing water supply (SPI, SPEI, and SSI) in
-most of Chile; just in the southern part, there is an increase. The
-trend is most strong in the north-central zone. The whole country showed
-an increase in water demand (AED). The magnitude of the trend increases
-over longer time scales, which is evidence that the deficit is impacting
-the hydrological system in Chile. The trend in vegetation productivity
-in the north-central area is affecting, to a higher degree, shrubland
-and savanna, followed by croplands and forests. The most important
-changes in land cover are the increase of forest in ``Sur'' and
-shrubland and grassland in ``Centro;'' and the increase of savanna in
-``Centro'' and ``Sur.''.
-
-The drought explains about 30\% of the change in land cover type across
-Chile for forest, grassland, shrubland, and savanna. There is no
-evidence of an effect of drought on the change in cropland surface area.
-The increase in AED is the main driver of the change in land cover,
-followed by a reduction in precipitation and soil moisture. The drought
-time scales vary regarding the land cover type.
+We found a significant trend toward decreasing water supply (SPI, SPEI,
+and SSI) in most of the Chilean territory, with the exception of the
+southern region. The trend is the strongest in the north-central zone.
+The whole country showed an increase in water demand (AED) due to
+increasing temperatures. The magnitude of the trends is stronger for
+longer time scales, which is evidence that there is a prolonged
+precipitation shortage and a prolonged increase in AED. The trend in
+vegetation productivity in the north-central area is affecting shrubland
+and savanna to a greater degree, followed by croplands and forests.
+
+We model about 20--30\% of the trends in land cover types, such as
+forest, grassland, shrubland, and savanna, based on drought indicators
+across Chile. There is no evidence that drought alters cropland surface
+area. The increase in AED is the most important variable explaining the
+variability in the change in land cover, followed by a reduction in
+precipitation and soil moisture.
+
+The trends in drought indices are accompanied by multiple land cover
+changes in the country, most notably an increase of forest in ``Sur,''
+of shrubland and grassland in ``Centro,'' and of savanna in ``Centro''
+and ``Sur.'' In ``Norte Chico'' and ``Centro,'' the croplands have been
+declining in surface, whereas in ``Sur,'' there is an increase in
+cultivated land.
 
@@
 past 12 months is the main variable explaining these changes, followed
-by anomalies in cumulated precipitation over one to two years. The
+by anomalies in accumulated precipitation over one to two years. The
 variation in AED seems to intensify the drought impact on vegetation
@@
 
-The results of this study provide insightful information that would help
-in developing adaptation measures for ecosystems in Chile to cope with
-climate change and drought.
+The results of this study provide insightful information that would
+assist in developing adaptation measures for Chilean ecosystems to cope
+with climate change and drought. Information that could be used in the
+scope of the national law on climate change, which seeks to implement
+adaptation strategies for agriculture, forests, and biodiversity.
 
@@
 study through the drought emergency project FSEQ210022, Fondecyt
-Iniciación N°11190360, and Fondecyt Regular N°1210526.
+Iniciación N°11190360, Fondecyt Postdoctorado N°3230678, and Fondecyt
+Regular N°1210526.
 

### 2024-04-11 article/supplementary.qmd
table landcover
last update antes egu2024

--- 2024-03-19 article/supplementary.qmd
@@
 title: "Supplementary Material"
-author: "Francisco Zambrano"
+author: "Francisco Zambrano et al."
+date: "`r Sys.Date()`"
+bibliography: references.bib
 format: 
@@
 
-We compared the ERA5L variables for monthly mean temperature, total precipitation, and volumetric soil water content against values retrieved by weather stations.
-For temperature and precipitation, we used the weather network from the Ministry of Agriculture of Chile (www.agromet.com) between 2015 and 2023.
+We compared the ERA5L [@MunozSabater2021] variables for monthly mean temperature, total precipitation, and volumetric soil water content against values retrieved by weather stations.
+For temperature and precipitation, we used data from the weather network from the Ministry of Agriculture of Chile (www.agromet.com) between 2015 and 2023.
 We used 277 stations located throughout Chile.
-For soil moisture, we select a private soil network that is owned by the agricultural enterprise Garces Fruit, which has 99 stations in Central Chile, located in cherry fruit crops.
-The sensors are installed at 30, 60, and 90m and are the model Teros 12 from MeterGroup.
-To avoid the effect of irrigation on soil moisture, which ERA5L hardly captures, we used daily data for the year 2022 and the months outside the growing season, May to September.
+For in-situ soil moisture, we used a private soil network that is owned by the agricultural enterprise Garces Fruit, which has 99 stations in Central Chile, located in fields with cherry fruit crops.
+The sensors are installed at 30, 60, and 90cm and are a Advanced Soil Moisture Sensor model Teros 12 from METER Group ([https://metergroup.com/products/teros-12/](https://metergroup.com/products/teros-12/)).
+To avoid comaring ERA5L with in-situ soil moisture levels caused by irrigation, which are not captured by ERA5L, we used daily data for the year 2022 and the months outside the growing season, May to September.
 
-To account for the performance of the ERA5L climatic variables regarding the values measured by the weather stations. We selected the following metrics:
+We selected the following metrics:
 
@@
 
-$MAE$: mean absolute error $bias$: bias $ubRMSE$: unbiassed root mean squared error $CC$: coefficient of correlation $S$: value of the variable measure by the weather station $E$: value of the variable measure by ERA5L
+$MAE$: mean absolute error   
+$bias$: bias  
+$ubRMSE$: unbiassed root mean squared error  
+$CC$: coefficient of correlation  
+$S$: value of the variable measure by the weather station  
+$E$: value of the variable measure by ERA5L
 
@@
 
-The average metrics of performance of ERA5L over the 266 weather stations were in the case of monthly temperature: $ubRMSE=1.06\,°C$, $MAE=1.131\,°C$, and $CC=0.963$, showing a good agreement, low error, and low overestimation.
+The average performance metrics of ERA5L over the 266 weather stations were in the case of monthly temperature: $ubRMSE=1.06\,°C$, $MAE=1.131\,°C$, and $CC=0.963$, showing a good agreement, low error, and low overestimation.
 For cumulative monthly precipitation, $MAE=28.1\,mm$, $bias=1.93$, and $CC=0.845$, showing a high correlation and a 93% bias and being overestimated by ERA5L.
-In the case of the 97 soil moisture stations, we averaged for the three depths (30, 60, and 90m) and then compared it with volumetric water content at 1m derived from ERA5L.
-For this case, we made a daily comparison, having a $CC=0.71$, $RMSE=0.174\,m^3m^{-3}$, $MAE=0.167\,m^3m^{-3}$, and $bias=1.74$.
-The ERA5 soil moisture overestimate is 74%, but it has a kind of good correlation.
+In the case of the 97 soil moisture stations, we averaged for the three depths (30, 60, and 90cm) and then compared it with the volumetric water content in the top 100cm of the soil derived from ERA5L.
+For this case, we made a daily comparison to determine the performance metrics per station, then we averaged over all stations, having a $CC=0.71$, $RMSE=0.174$, $MAE=0.167$, and $bias=1.74$.
+The ERA5 soil moisture overestimate is 74%, but there is some good correlation.
 
-# Land cover macroclasess and validation
+# Land cover macroclasses and validation
 
@@
 
-To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS.
+To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 Collection 6.1 from MODIS.
 This product has a yearly frequency from 2001 to 2022.
-The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
-Thus, we have a land cover raster time series with the ten classes for 2001 and 2023.
+The IGBP defines 17 classes; we regrouped these into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
 
-To validate the land cover obtained, we compare the macroclasses with the ones of a more detailed land cover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile).
-The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses land cover.
-We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from land cover macrolcasses to validate with LCChile.
+To validate the land cover obtained, we compare the macroclasses with those of a more detailed land cover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile).
+The latter has a spatial resolution of 30 m and three hierarchy levels of defined classes; from those, we used level 1, which fits with the macroclass land cover.
+We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from the land cover macrolcasses to compare with LCChile.
+For this comparison, we used the following procedure:
 
-We follow the next procedure:
-
-i)  resampled LCChile to the spatial resolution (500m) of the land cover macroclasses using the nearest neighbor method,
-ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values from the other two years as predictions.
-iii) calculate a confusion matrix with the classes extracted from the 1000 points for LCChile, IGBP2013, and IGBP2014. Calculate the performance metrics of accuracy and F1.
+i)  we resampled LCChile to the spatial resolution (500m) of the land cover macroclasses using the majority method,
+ii) we took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values from the other two years as predictions.
+iii) we derivede a confusion matrix with the classes extracted from the 1000 points for LCChile, IGBP2013, and IGBP2014; and calculate the performance metrics of accuracy and F1.
 
@@
 
-For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands.
-Figure 1 c shows the spatial distribution of the macroclasses through Chile for the year 2022.
-The validation of IGBP2013 and IGBP2014 with LCChile reached near the same metrics of performance, having an accuracy of \~0.82 and a F1 score of \~0.66.
+For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands (Table \ref{tab-landIGBPmacro}).
+
+```{=latex}
+\begin{table}[!ht]
+\caption{Landcover clases from IGBP MODIS and the corresponding macroclasses.}
+\label{tab-landIGBPmacro}
+\centering
+\includegraphics[width = .5\textwidth]{../output/figs/tabla_landcover_macroclass.png}
+\end{table}
+```
 
 # Relationship between drought indices and land cover change
+
+@fig-varImportance shows the ranking of variable importance obtained with the ten folds used in the resampling of the Random Forest model per land cover type.
 
@@
 
-Error bar for the variables importance obtained from the 10 resamples folded by the Random Forest.
+Error bar for the variables importance obtained from the 10 resamples folded by the Random Forest used to model land cover trend using as predictors the trend in drought indices of water supply, water demand, soil moisture, and vegetation productivity.
 
 :::
+
+# Trend of vegetation productivity
+
+@fig-hetmaptrendzcNDVI shows the average trend for zcNDVI for 2000-2023 per macrozone and landcover macroclass.
+
+```{r fig-hetmaptrendzcNDVI}
+#| echo: false
+#| out-width: 75%
+#| fig-pos: '!ht'
+#| fig-cap: "Heatmap of trends in zcNDVI for 2000 to 2023 per macrozone and landcover macroclass."
+knitr::include_graphics('../output/figs/heatmap_trends_zcNDVI_macro_landcover.png')
+```
+
+# Vegetation productivity
+
+We analyzed the correlation of zcNDVI for time scales of 1, 3, 6, and 12 months versus net primary production (NPP).
+We obtained both the zcNDVI from MOD13A3.061 and the NPP from MOD17A3HGF.061, using MODIS products.
+We used the zcNDVI in December to correlate with the annual NPP.
+@fig-r2_npp_zcNDVI shows a map of the r-squared correlation between zcNDVI and NPP, and @fig-hetmap_npp_zcndvi shows the aggregated values per macrozone.
+
+```{r fig-r2_npp_zcNDVI}
+#| echo: false
+#| out-width: 75%
+#| fig-pos: '!ht'
+#| fig-cap: "Spatial variation of the r-squared values obtained from the yearly correlation of zcNDVI of 1, 3, 6, and 12 months with the net primary productivity (NPP) for continental Chile."
+knitr::include_graphics('../output/figs/map_r2_NPP_vs_zcNDVI.png')
+```
+
+```{r fig-hetmap_npp_zcndvi}
+#| echo: false
+#| out-width: 75%
+#| fig-pos: '!ht'
+#| fig-cap: "A heatmap showing the r-squared values obtained from the yearly correlation of zcNDVI of 1, 3, 6, and 12 months with the net primary productivity (NPP) for continental Chile."
+knitr::include_graphics('../output/figs/heatmap_r2_npp_vs_zcndvi_per_landcover.png')
+```
+
+# References
+
+::: {#refs}
+:::

### 2024-04-11 scripts/01_update_MODIS_products.R

@@ -8,7 +8,7 @@ library(httr)
 library(curl)
 
 type <- c('MOLA','MOLT','MOTA')[2]
-product <- 'MOD13A3.061'
+product <- 'MOD17A3HGF.061'
 dir_out <- '/mnt/md0/raster_raw/MODIS/'
 
 updt_modis(type,product,dir_out)

### 2024-04-11 scripts/15_IGBP_variacion_tendencia_2001-2021.R

@@ -80,3 +80,25 @@ data3 |>
   summarize(trend = trend_func(sup_km2,start = c(2001,1),end =c(2022,1),frecuency=1)[2]) |> 
   pivot_wider(names_from = 'LC_type',values_from=3) |> 
   saveRDS('data/processed_data/trends_landcover_2001-2022.rds')  
+
+data3 |> 
+  ggplot(aes(year,sup_km2,fill=LC_type)) + 
+  geom_col(position = 'fill') +
+  scale_fill_manual(values = colors, name = 'Macroclass')+
+  scale_x_continuous(expand = c(0,0),breaks = 2001:2022) + 
+  scale_y_continuous(expand = c(0,0),breaks = seq(0,1,.10), 
+                     labels = scales::percent_format(accuracy = 1)) +
+  facet_grid(.~zone) +
+  theme_bw() + 
+  theme(axis.title = element_blank(),
+        axis.text.x = element_text(angle = 90),
+        strip.background = element_rect(fill = 'white'),
+        legend.position = 'bottom') +
+  guides(fill = guide_legend(nrow = 1))
+ggsave('output/figs/proportion_landcover_macroclass_2001-2022.png',width=10,height=2.5,scale =1.5)
+
+fill = guide_legend(nrow = 1)
+  scale_y_continuous(breaks = seq(0,1,.1), 
+                     labels = scales::percent_format(accuracy = 1),
+                     expand = c(0.01,0)) 
+  

### 2024-04-11 scripts/16_map_study_landcover.R

@@ -117,7 +117,7 @@ map_chl <- tm_shape(bm2) +
   tm_raster(palette = colores,
             style = 'cat',
             labels = labs,
-            title = 'Climate (1990-2020)') +
+            title = 'Climate (1991-2020)') +
   tm_shape(zones) +
   tm_borders(lwd=2,col='black',lty='solid',alpha=0.8) +
   tm_text('macrozona',just='top',xmod=c(5,-7,3,-4,-4),ymod=c(0,-3.5,1,0,0),size=1.5) +

### 2024-04-11 scripts/17_extrarer_correlations_macro_landcover.R

@@ -121,6 +121,7 @@ tabla_gt$Shrubland_SSI_r[5] <- NA
 
 #tabla_gt[1,2] <- NA
 tabla_gt |> 
+  rename(Macrozone = macrozone) |> 
   dplyr::select(1,22:25,18:21,10:13,6:9,2:5,
          46:49,38:41,34:37,30:33,26:29
          ) |> 
@@ -129,7 +130,7 @@ tabla_gt |>
   data_color(
     columns = 22:41,
     target_columns =2:21,
-    palette = viridis::inferno(20),
+    palette = rev(viridis::inferno(20)),
     na_color = 'white',
     alpha = .8,
     domain = c(0,.8)  
@@ -152,7 +153,10 @@ plot <- data_r_4gt |>
   pivot_longer(-macro) |> 
   ggplot(aes(name,value,color=value)) + 
   geom_point() + 
-  scale_color_viridis_c(option = 'inferno',name = 'r-squared',alpha = .8) + theme(legend.position = 'bottom')
+  scale_color_viridis_c(option = 'inferno',
+                        name = 'r-squared',
+                        direction = -1,
+                        alpha = .8) + theme(legend.position = 'bottom')
 
 ggpubr::get_legend(plot) |> 
   ggpubr::as_ggplot() |> 

### 2024-04-11 scripts/21_landcover_per_macrozone.R

@@ -49,7 +49,7 @@ summ_data2 |>
   mutate(surface = case_when(surface <=20~NA,
                            .default = surface)) |> 
   pivot_wider(names_from = LC_type,values_from=surface) |> 
-  rename(macrozone = zone) |> 
+  rename(Macrozone = zone) |> 
   ungroup() |> 
    select(1,7:6,4:2,5) |> 
   gt() |> 
@@ -65,7 +65,7 @@ summ_data2 |>
     missing_text = ""
   ) |> 
   tab_header(
-    title = md("Surface  [km<sup>2]"),
+    title = md("Surface  [km<sup>2</sup>]"),
   ) |> 
   gt::gtsave('output/figs/table_surface_landcover_macrozone.png')
 

### 2024-04-11 scripts/24_extraer_valores_indices_sequia_macrozona_landcover.R

@@ -58,6 +58,7 @@ data_index <- map_df(indices,\(index){
   })
   return(data_trend)
 })
+})
 
 # data_t |> 
 #   pivot_wider(names_from = scale)

### 2024-04-11 scripts/25_trend_drought_indices.R

@@ -20,7 +20,7 @@ lc_pers[lc_pers %in% c(5,7:10)] <- NA
 lc_pers <- resample(lc_pers,zcndvi)
 
 #aplicar mascara de landcover permanenete 
-lc_pers <- resample(lc_pers,zcndvi,method = 'near')
+lc_pers <- resample(lc_pers,zcndvi,method = 'mode')
 zcndvi <- mask(zcndvi,lc_pers)
 
 macro_mask <- rasterize(st_transform(macro,32719),zcndvi[[1]],field ='macrozona')

### 2024-04-11 scripts/26_calculate_man_kendall_trend_on_rasters.R

@@ -24,7 +24,7 @@ map(indices,\(index){
       index_ts <- rast(files)
       cl <- snow::makeCluster(80,"SOCK")
       snow::clusterExport(cl, list = c('trend_func','mk.test','sens.slope'))
-      trend <- app(index_ts,trend_func,cores = cl,start = c(2002,2),end =c(2023,10))
+      trend <- app(index_ts,trend_func,cores = cl,start = c(2022,2),end =c(2024,2))
       snow::stopCluster(cl)
       writeRaster(trend,glue('/mnt/md0/raster_procesada/analysis/trends/trend_mann_kendall_{index}-{scale}.tif'),overwrite = TRUE)
       

### 2024-04-11 scripts/29_mapas_correlaciones_indices_zcNDVI.R

@@ -54,9 +54,11 @@ names(cors_r) <- 1:4
 map_r <- tm_shape(bm) + 
   tm_rgb() +
   tm_shape(cors_r) +
-  tm_raster(style='equal',palette = viridis::inferno(20),
+  tm_raster(style='equal',palette = rev(viridis::inferno(20)),
             midpoint = 0,
-            title = 'r',legend.hist = TRUE) +
+            title = 'r',
+            legend.hist = TRUE,
+            legend.reverse = TRUE) +
   tm_facets(free.scales	= FALSE) + 
   tm_layout(
     panel.labels = c('EDDI','SPI','SPEI','SSI'),

### 2024-04-11 scripts/32_extraer_tendencias_indices_cuencas_landcover.R

@@ -1,14 +1,15 @@
 ## Extraer tabla de tendencias para macrozonas}
 library(sf)
 library(terra)
+library(fs)
 
-cuencas <- read_sf('data/processed_data/spatial/Cuencas_BNA') 
+cuencas <- read_sf('data/processed_data/spatial/subcuencas_bna.gpkg') 
 
 dir <- '/mnt/md0/raster_procesada/analysis/trends'
 files <- dir_ls(dir,regexp = 'mann_kendall.*tif$')
 indices <- str_remove(str_remove(str_extract(files,'l_.*\\.'),'l_'),'\\.')
-indices <- indices[-19]
-trends <- rast(names(files)[-19])
+indices <- indices[-(19:22)]
+trends <- rast(names(files)[-(19:22)])
 trends <- trends[[seq(2,48,2)]]
 names(trends) <- indices
 
@@ -23,7 +24,7 @@ df_scales_2 <- terra::extract(trends[[1:6]],cuencas,fun =\(x){
 })
 
 #zcNDVI
-trends <- rast(names(files)[19])
+trends <- rast(names(files)[22])
 trends <- trends[[2]]
 names(trends) <- 'zcNDVI-6'
 cuencas <- st_transform(cuencas,32719)
@@ -36,19 +37,19 @@ df_scales <- cbind(st_drop_geometry(cuencas),df_scales_1,df_scales_2,df_scales_3
   select(-starts_with('ID')) 
 
 sel_cuen <- df_scales |> 
-  group_by(COD_CUEN) |> 
+  group_by(COD_SUBC) |> 
   summarize(n =n()) |> 
   filter(n == 1 ) |> 
-  pull(COD_CUEN)
+  pull(COD_SUBC)
 
 df_scales <- df_scales |> 
-  filter(COD_CUEN %in% sel_cuen) |> 
-  select(-c(2:3)) |> 
-  mutate(COD_CUEN = fct(COD_CUEN)) 
+  filter(COD_SUBC %in% sel_cuen) |> 
+  select(-c(1,3:5)) |> 
+  mutate(COD_SUBC = fct(COD_SUBC)) 
 
 #guardar data.frame con los valores promedio de las tendencias (mann-kendall) para cada indice en las macrozonas de Chile
 df_scales |> 
-  pivot_longer(-c(COD_CUEN)) |> 
+  pivot_longer(-c(COD_SUBC)) |> 
   rename(index = name,trend = value) |> 
[…]

### 2024-04-11 scripts/33_analysis_trend_drought_indices_and_landcover_cuencas.R

@@ -1,5 +1,8 @@
 library(tidyverse)
 library(tidymodels)
+library(probably)
+
+cuenca_macro <- read_rds('data/processed_data/cuencas_macrozona.rds')
 
 data_trend_lc <- read_rds('data/processed_data/trends_landcover_cuencas_2001-2022.rds') |> 
   pivot_longer(-cuenca) |> 
@@ -23,6 +26,9 @@ data_trend_di <- read_rds('data/processed_data/df_trends_indices_cuencas.rds') |
 
 data_ana <- data_trend_di |> 
   full_join(data_trend_lc) |> 
+  full_join(cuenca_macro,by = c('cuenca' = 'COD_SUBC')) |> 
+  relocate(macrozona,.after=cuenca) |> 
+  relocate(Latitude,.after=macrozona) |> 
   drop_na()
 
 #data_model <- data_ana[,c(1:6,9:12,15:18,21:24,29,31)]
@@ -36,11 +42,11 @@ data_model <- data_ana
 #          macro = factor(macro))
 
 data_model <- data_model |> 
-  rename('Barren_land' = `Barren land`) |> 
-  select(-cuenca)
+  rename('Barren_land' = `Barren land`)
 
 #random forest
 vars <- c("Shrubland",   "Savanna",     "Grassland",   "Barren_land", "Cropland",    "Forest")
+macrozones <- data_model |> distinct(macrozona) |> pull(macrozona)
 
 get_rf_imp <- function(x) {
   x |> 
@@ -48,106 +54,200 @@ get_rf_imp <- function(x) {
     vip::vi()
 }
 
+#probando la tendencia como categoria
+#
 # para tendencia en usos de suelo
  
 df <- map(1:6,\(i){
-  t <- which(vars == vars[[i]])
-  df_split <- initial_split(data_model |> select(-vars[-t]),prop =0.99)
-  df_train <- training(df_split)
-  df_test <- testing(df_split)
+  map(macrozones,\(macro){
+    t <- which(vars == vars[[i]])
+    data_m <- data_model |> filter(macrozona == {{ macro }}) |> select(-cuenca,-macrozona,-Latitude) 
+    df_split <- initial_split(data_m |> select(-vars[-t]),prop =0.85)
+    df_train <- training(df_split)
+    df_test <- testing(df_split)
+    
+    rf_spec <- rand_forest(trees = 1000, mode = "regression") |>
+      set_args(importance = 'impurity')
+    
[…]

### 2024-04-11 scripts/40_join_subcuencas_macrozonas.R

@@ -1,0 +1,18 @@
+library(sf)
+
+cuencas <- read_sf('data/processed_data/spatial/subcuencas_bna.gpkg') 
+
+cuencas <- cuencas |> 
+  st_transform(4326) |> 
+  bind_cols(Latitude = st_coordinates(st_centroid(cuencas))[,2])
+
+macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg')
+
+cuencas_macro <- st_union(cuencas,macro)
+
+cuencas_macro <- cuencas_macro |> 
+  st_drop_geometry() |> 
+  select(COD_SUBC,macrozona,Latitude)
+
+library(readr)
+write_rds(cuencas_macro,'data/processed_data/cuencas_macrozona.rds')

### 2024-05-08 article/paper_drought_Chile2023.qmd
actualizar local servidor

@@
 #| fig-cap: "cdcdcdc"
-#| fig-width: 3
-#| out-width: ".01\\textwidth"
+#| out-width: "\\textwidth"
 knitr::include_graphics("../output/figs/bars_relative_importance_RF.png")

### 2024-05-08 R/updt_modis.R

@@ -40,7 +40,7 @@ updt_modis <- function(type,product,dir_out){
       tryCatch(
       httr::GET(url = paste0(URL,dir,'/',name),
           config = httr::config(connecttimeout = 120),
-          authenticate("frzambra@gmail.com","Traplozx398#"),
+          httr::authenticate("frzambra@gmail.com","Traplozx398#"),
           write_disk(paste0(dir_out,product,'/',dir,'/',name),
                      overwrite=TRUE),
           progress()),

### 2024-05-08 scripts/02_mosaic_chile_modis_products.R

@@ -4,10 +4,10 @@
 
 require(purrr)
 
-dir_in <- '/mnt/md0/raster_raw/MODIS/MOD13A3.061/'
-dir_out <- '/mnt/md0/raster_procesada/MODIS/NDVI.MOD13A3.061/'
+dir_in <- '/mnt/md0/raster_raw/MODIS/MOD17A3HGF.061/'
+dir_out <- '/mnt/md0/raster_procesada/MODIS/NPP.MOD17A3HGF.061/'
 
-mosaic_modisFromHDF(dir_in,dir_out,lyr=1)
+mosaic_modisFromHDF(dir_in,dir_out,lyr=2)
 
 datesUpdate <- dates2Update(dir.out,dirs)
 

### 2024-05-08 scripts/15_IGBP_variacion_tendencia_2001-2021.R

@@ -77,28 +77,6 @@ data3 |>
   ungroup() |>  
   group_by(zone,LC_type) |>  
   #summarize(trend  = as.numeric(tidy(lm(sup_km2~year,data))[2,2])) |> 
-  summarize(trend = trend_func(sup_km2,start = c(2001,1),end =c(2022,1),frecuency=1)[2]) |> 
+  summarize(trend = trend_func(prop,start = c(2001,1),end =c(2022,1),frecuency=1)[2]) |> 
   pivot_wider(names_from = 'LC_type',values_from=3) |> 
   saveRDS('data/processed_data/trends_landcover_2001-2022.rds')  
-
-data3 |> 
-  ggplot(aes(year,sup_km2,fill=LC_type)) + 
-  geom_col(position = 'fill') +
-  scale_fill_manual(values = colors, name = 'Macroclass')+
-  scale_x_continuous(expand = c(0,0),breaks = 2001:2022) + 
-  scale_y_continuous(expand = c(0,0),breaks = seq(0,1,.10), 
-                     labels = scales::percent_format(accuracy = 1)) +
-  facet_grid(.~zone) +
-  theme_bw() + 
-  theme(axis.title = element_blank(),
-        axis.text.x = element_text(angle = 90),
-        strip.background = element_rect(fill = 'white'),
-        legend.position = 'bottom') +
-  guides(fill = guide_legend(nrow = 1))
-ggsave('output/figs/proportion_landcover_macroclass_2001-2022.png',width=10,height=2.5,scale =1.5)
-
-fill = guide_legend(nrow = 1)
-  scale_y_continuous(breaks = seq(0,1,.1), 
-                     labels = scales::percent_format(accuracy = 1),
-                     expand = c(0.01,0)) 
-  

### 2024-05-08 scripts/23_analysis_trend_drought_indices_and_landcover_macroXlandcover.R

@@ -13,11 +13,6 @@ data_ana <- data_trend_di |>
   full_join(data_trend_lc) |> 
   drop_na()
 
-paleta <- read_csv('data/processed_data/paleta_colores_landcover.csv')
-colors <-  rgb(paleta$R,paleta$G,paleta$B,maxColorValue = 255)
-attr(colors,'names') <- paleta$Name
-colors[3] <- 'red'
-
 #cargar la superficie de landcover ṕor macro class para relativizar las superficies
 
 lc_surf <- read_rds('data/processed_data/surface_landcover_macrozone.rds') |> 
@@ -56,13 +51,13 @@ data_ana |>
                                 'SPI-36','EDDI-36','zcSM-36'),
                     labels = c('SPI-6','EDDI-6','SSI-6',
                                'SPI-36','EDDI-36','SSI-36'))) |> 
-  ggplot(aes(value*10,landcover,color = class,shape = macro)) +
+  ggplot(aes(value*10,landcover,color = macro,shape = class)) +
   geom_point() + 
   geom_hline(yintercept = 0,color = 'red',linetype = 'dashed',alpha = .6) +
   geom_vline(xintercept = 0,color = 'red',linetype = 'dashed',alpha = .6) +
   labs(x = 'Trend of drought index (per decade)',
        y = 'Relative trend of land cover change') +
-  scale_color_manual('class',values = colors,name = 'Land cover') +
+  scale_color_viridis_d('Macrozone') +
   #scale_x_sqrt() +
   scale_shape('Land cover') +
   facet_wrap(.~name,ncol=3,nrow=2) +
@@ -70,6 +65,35 @@ data_ana |>
   theme(strip.background = element_rect(fill = 'white'))
 ggsave('output/figs/points_landcover_drought_indices_trend_and_time_scale.png',scale =1.2,width=10,height = 6)
 
+#Tendencia por land cover type and macrozone
+#
+data_trend_di |> 
+  select(1:2,`zcNDVI-6`) |>   
+  dplyr::filter(class %in% c('Forest','Shrubland',
+                             'Savanna','Grassland',
+                             'Cropland','Barren land')
+                ) |> 
+  mutate(
+    class = as.character(class),
+    macro = fct(macro,levels = c('Norte Grande','Norte Chico','Centro','Sur','Austral')),
[…]

### 2024-05-08 scripts/31_landcover_per_basin.R

@@ -34,9 +34,9 @@ cuencas <- read_sf('data/processed_data/spatial/subcuencas_bna.gpkg')
 data <- terra::extract(igbpRecl,cuencas)
 data <- data |> 
   repair_names() |> 
-  cbind(cuencas[data$ID,'COD_CUEN']) |> 
+  cbind(cuencas[data$ID,'COD_SUBC']) |> 
   select(-geom,-ID) |> 
-  relocate("COD_CUEN",.before = "Land_Cover_Type_1") 
+  relocate("COD_SUBC",.before = "Land_Cover_Type_1") 
 names(data) <- c('cuenca',2001:2022)
 
 library(dplyr)

### 2024-05-08 scripts/32_extraer_tendencias_indices_cuencas_landcover.R

@@ -2,14 +2,15 @@
 library(sf)
 library(terra)
 library(fs)
+library(tidyverse)
 
 cuencas <- read_sf('data/processed_data/spatial/subcuencas_bna.gpkg') 
 
 dir <- '/mnt/md0/raster_procesada/analysis/trends'
 files <- dir_ls(dir,regexp = 'mann_kendall.*tif$')
 indices <- str_remove(str_remove(str_extract(files,'l_.*\\.'),'l_'),'\\.')
-indices <- indices[-(19:22)]
-trends <- rast(names(files)[-(19:22)])
+indices <- indices[-19:-22]
+trends <- rast(names(files[-19:-22]))
 trends <- trends[[seq(2,48,2)]]
 names(trends) <- indices
 
@@ -24,10 +25,10 @@ df_scales_2 <- terra::extract(trends[[1:6]],cuencas,fun =\(x){
 })
 
 #zcNDVI
-trends <- rast(names(files)[22])
+trends <- rast(names(files[22]))
 trends <- trends[[2]]
 names(trends) <- 'zcNDVI-6'
-cuencas <- st_transform(cuencas,32719)
+cuencas <- st_transform(cuencas,crs(trends))
 df_scales_3 <- terra::extract(trends,cuencas,fun =\(x){
   table(x) |> sort(decreasing = TRUE) -> x
   as.numeric(names(x[1]))
@@ -37,14 +38,16 @@ df_scales <- cbind(st_drop_geometry(cuencas),df_scales_1,df_scales_2,df_scales_3
   select(-starts_with('ID')) 
 
 sel_cuen <- df_scales |> 
+  select(-1,-3,-4,-5) |> 
   group_by(COD_SUBC) |> 
   summarize(n =n()) |> 
   filter(n == 1 ) |> 
   pull(COD_SUBC)
 
-df_scales <- df_scales |> 
+df_scales <- df_scales |>
+  select(-1,-3,-4,-5) |> 
   filter(COD_SUBC %in% sel_cuen) |> 
-  select(-c(1,3:5)) |> 
+  select(-c(2:3)) |> 
   mutate(COD_SUBC = fct(COD_SUBC)) 
 
 #guardar data.frame con los valores promedio de las tendencias (mann-kendall) para cada indice en las macrozonas de Chile

### 2024-05-08 scripts/33_analysis_trend_drought_indices_and_landcover_cuencas.R

@@ -1,8 +1,5 @@
 library(tidyverse)
 library(tidymodels)
-library(probably)
-
-cuenca_macro <- read_rds('data/processed_data/cuencas_macrozona.rds')
 
 data_trend_lc <- read_rds('data/processed_data/trends_landcover_cuencas_2001-2022.rds') |> 
   pivot_longer(-cuenca) |> 
@@ -26,9 +23,6 @@ data_trend_di <- read_rds('data/processed_data/df_trends_indices_cuencas.rds') |
 
 data_ana <- data_trend_di |> 
   full_join(data_trend_lc) |> 
-  full_join(cuenca_macro,by = c('cuenca' = 'COD_SUBC')) |> 
-  relocate(macrozona,.after=cuenca) |> 
-  relocate(Latitude,.after=macrozona) |> 
   drop_na()
 
 #data_model <- data_ana[,c(1:6,9:12,15:18,21:24,29,31)]
@@ -42,11 +36,11 @@ data_model <- data_ana
 #          macro = factor(macro))
 
 data_model <- data_model |> 
-  rename('Barren_land' = `Barren land`)
+  rename('Barren_land' = `Barren land`) |> 
+  select(-cuenca)
 
 #random forest
 vars <- c("Shrubland",   "Savanna",     "Grassland",   "Barren_land", "Cropland",    "Forest")
-macrozones <- data_model |> distinct(macrozona) |> pull(macrozona)
 
 get_rf_imp <- function(x) {
   x |> 
@@ -54,200 +48,130 @@ get_rf_imp <- function(x) {
     vip::vi()
 }
 
-#probando la tendencia como categoria
-#
 # para tendencia en usos de suelo
  
 df <- map(1:6,\(i){
-  map(macrozones,\(macro){
-    t <- which(vars == vars[[i]])
-    data_m <- data_model |> filter(macrozona == {{ macro }}) |> select(-cuenca,-macrozona,-Latitude) 
-    df_split <- initial_split(data_m |> select(-vars[-t]),prop =0.85)
-    df_train <- training(df_split)
-    df_test <- testing(df_split)
-    
-    rf_spec <- rand_forest(trees = 1000, mode = "regression") |>
-      set_args(importance = 'impurity')
-    
-    rf_wflow <- workflow(as.formula(paste0(vars[i],'~ .')), rf_spec)
-    rf_fit <- fit(rf_wflow, df_train)
+  t <- which(vars == vars[[i]])
[…]

### 2024-05-08 scripts/38_analysis_NPP_zcNDVI.R

@@ -1,0 +1,120 @@
+library(terra)
+library(tidyverse)
+library(fs)
+library(rnaturalearth)
+
+dir <- '/mnt/md0/raster_procesada/MODIS/'
+dir2 <- '/mnt/md0/raster_procesada/MODIS_derived/zcNDVI/'
+
+chl <- ne_countries(country = 'chile',scale = 'small',returnclass = 'sf') |> 
+  st_transform(32719)
+
+files_npp <- dir_ls(paste0(dir,'NPP.MOD17A3HGF.061/'),regexp = 'tif$')
+
+npp <- rast(files_npp)
+npp <- crop(npp,chl)
+npp[npp >30000] <- NA
+
+#detrend NPP
+
+# npp_det <- app(npp,\(val){
+#   df <- data.frame(x=1:length(val),y=val)
+#   df <- na.omit(df)
+#   
+#   if(nrow(df) > 10){
+#     mod <- lm(y~x,df)
+#     out <- df$y-mod$fitted.values
+#   } else out <- rep(NA,length(val))
+#   out
+# },cores=80)
+
+#remove year 2014
+npp <- npp[[-14]]
+
+out <- lapply(c(1,3,6,12),\(scale){
+  files_zcndvi <- dir_ls(paste0(dir2,glue::glue('zcNDVI-{scale}/')),regexp = 'tif$')
+    ind <- basename(files_zcndvi) |> 
+    str_extract('[0-9]{4}-[0-9]{2}-[0-9]{2}') |> 
+    ymd() |> 
+    month() |> 
+    (\(x) which(x == 12))()
+  
+  zcndvi <- rast(files_zcndvi[ind][-1])
+  zcndvi <- zcndvi[[-14]] 
+  zcndvi <- crop(zcndvi,chl)
+  
+  npp <- resample(npp,zcndvi)
+  
+  x <- c(zcndvi,npp)
+  r <-  app(x,fun =\(x){
+    df <- data.frame(X = x[1:22], Y=x[23:length(x)])
+    df <- na.omit(df)
+    if(nrow(df) > 10){
+      mod <- lm(Y~X,df)
+      r <- cor(df$X,df$Y)
+      r2 <- summary(mod)$r.squared
+      pvalue <- summary(mod)$coefficients[2,4]
+      sign <- ifelse(pvalue < 0.05,1,0)
+      out <- c(r,r2,pvalue,sign)
+    } else out <- c(NA,NA,NA,NA)
+    
+    return(out)
[…]

### 2024-05-09 article/paper_drought_Chile2023.qmd
update version may 2024

@@
           city: Santiago
-          country: 7500994, Chile
+          country: 7500994, Chile.
         - id: odes
@@
           city: Santiago
-          country: 7500994, Chile
+          country: 7500994, Chile.
     attributes:
@@
           name: Faculty of Geo-Information Science and Earth, University of Twente
-          city:  Enschede, The Netherlands
+          city:  Enschede
+          country: The Netherlands.
   - name: Francisco Meza
@@
           city: Santiago
-          country: Chile
+          country: Chile.
         - id: instuc
@@
           city: Santiago
-          country: Chile
+          country: Chile.
         - id: cambioglobal
-          name: Centro Interdisciplinario de Cambio Global, Pontificia Universidad Católica de Chile, Santiago, Chile.
+          name: Centro Interdisciplinario de Cambio Global, Pontificia Universidad Católica de Chile
+          city: Santiago
+          country: Chile.
   - name: Iongel Duran-Llacer
@@
           city: Santiago
-          country: 7500994, Chile
+          country: 7500994, Chile.
   - name: Francisco Fernández
@@
           city: Santiago
-          country: Chile
+          country: Chile.
         - id: capes
@@
           city: Santiago
-          country: Chile
+          country: Chile.
   - name: Alejandro Venegas-González
+    affiliations:
+        - id: ica3
+          name: Instituto de Ciencias Agroalimentarias, Animales y Ambientales (ICA3), Universidad de O'Higgins
+          city: San Fernando
+          country: Chile.
   - name: Nicolas Raab
@@
         - id: gema
-          department: GEMA Center for Genomics, Ecology & Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750, Santiago, Chile
+          department: GEMA Center for Genomics, Ecology & Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750
+          city: Santiago
+          country: Chile.
         - id: do
-          name: Data Observatory Foundation, Santiago, Chile
+          name: Data Observatory Foundation
+          city: Santiago
+          country: Chile.
   #   note: |
@@
 This persistent period of water scarcity has been defined as a mega-drought.
-There are few studies about the relationship between drought and ecosystem changes that can help with a better understanding of ecological drought.
+There is yet insufficient understanding of ecological drought in Chile due to the limited studies on the relationship between drought and ecosystem changes.
 The aim of our study is to evaluate the interaction of drought, land cover change, and vegetation productivity over continental Chile.
@@
 We derived the drought indices using monthly ERA5-Land reanalysis data from 1981 to 2023.
-We used the Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets to derive information on annual land cover and monthly vegetation productivity.
-Our results showed that except for the Austral part, Chile has a temporal decreasing trend in water supply, and across the whole country, there is an increase in AED.
+We used Moderate-Resolution Imaging Spectroradiometer (MODIS) datasets to derive information on annual land cover and monthly vegetation productivity.
+Our results showed that, except for the Austral part, Chile has a temporal decreasing trend in water supply, and across the whole country, there is an increase in AED.
 These trends become stronger over longer time scales.
@@
 The anomaly in soil moisture over the past 12 months (SSI-12) is the most important variable explaining these changes, followed by anomalies in accumulated precipitation over one to two years (SPI-12 and SPI-24).
-The variable importance obtained by random forest models indicates that drought is explaining about 20–30% of the change in land cover surface across Chile for forest, grassland, shrubland, and savanna but has no relation to the changes in croplands.
+The variable importance obtained by random forest models indicates that drought explains about 12–41% of the change in land cover surface across Chile for forest, grassland, shrubland, and savanna but has little relation to the changes in croplands.
 The increase in AED is the main variable associated with the change in land cover, followed by a reduction in precipitation and soil moisture.
 Our findings provide insightful information that could assist in developing adaptation measures for Chilean ecosystems to cope with climate change and drought.
-Also, this study could contribute to a better comprehension of ecological drought.
 
@@
 # Introduction
-
-Drought is often classified as 1) meteorological when precipitation in a specific period remains below the mean precipitation experienced in the same period during multiple years (more than 30 years usually), 2) hydrological when these anomalies last for long periods (months to years) and affect water systems, and 3) agricultural when the deficit negatively impacts plant health and leads to decreased productivity of crops or pastures [@Wilhite1985].
+Drought can be classified as 1) meteorological, when precipitation in a specific period remains below the mean precipitation experienced in the same period during multiple years (more than 30 years usually); 2) hydrological, when these anomalies last for long periods (months to years) and affect water systems; and 3) agricultural, when the deficit negatively impacts plant health and leads to decreased productivity of crops or pastures [@Wilhite1985].
 However, because drought is also influenced by human activities, @Loon2016 and @AghaKouchak2021 expanded the drought definition for the Anthropocene, indicating that the feedback of human decisions and activities should also be considered (i.e., anthropogenic drought).
-Droughts lead to increased tree mortality [@Cheng2024] and induces alterations in land cover and land use, ultimately affecting ecosystems [@Crausbay2017].
-Even though many ecological studies have at times mistakenly considered "dry" conditions as "drought" [@Slette2019].
-Ecological drought can be defined as _"an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems"_ [@Crausbay2017].
+Droughts can lead to increased tree mortality [@Cheng2024] and induce alterations in land cover and land use, ultimately affecting ecosystems [@Crausbay2017].
+Even though many ecological studies have at times mistakenly considered "dry" conditions as "drought" [@Slette2019], ecological drought can be defined as _"an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems"_ [@Crausbay2017].
 In light of current global warming, it is crucial to study the interaction between drought and ecosystems in order to understand their feedback and impact on future water security [@Bakker2012].
@@
 Also, AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C.
-To better evaluate the impact of drought trends on ecosystems, assessments that correlate meteorological and soil moisture variables to their effects on vegetation are much needed.
+To better evaluate the impact of drought trends on ecosystems, assessments of the relationship between meteorological and soil moisture variables and their effects on vegetation are much needed.
 
@@
 To better understand these contributions on land cover change and vegetation productivity the following questions may be asked: i) how do land cover and vegetation productivity respond to short- to long-term meteorological and soil moisture droughts?
-And ii) How is this response different between humid and arid climatic zones?
+And ii) how is this response different between humid and arid climatic zones?
 Likewise, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.
@@
 Since its formulation, it has been used worldwide for the study and monitoring of drought [@Gebrechorkos2023;@Liu2024].
-Recently, there has been more interest in using AED to track droughts separately to better disengage the effects of precipitation from temperature-dependent effects [@Vicente-Serrano2020].
-One of the reasons is that AED is linked to more flash droughts in limited water regions [@Noguera2022].
+Recently, there has been more interest in using AED to track droughts separately to better disentangle precipitation from temperature-dependent effects [@Vicente-Serrano2020].
+One of the reasons is that AED is more linked to flash droughts in water-limited regions [@Noguera2022].
 @Hobbins2016 and @McEvoy2016 developed the Evaporative Demand Drought Index (EDDI) to monitor droughts solely using the AED, and it has proven effective in monitoring flash droughts [@Li2024;@Ford2023].
@@
 @Hao2013 and @AghaKouchak2014 proposed the Standardized Soil Moisture Index (SSI), which has a similar formulation as the SPI, SPEI, and EDDI.
-Thus, many drought indices exist that allow for a comprehensive assessment of drought on short- to long-term scales and that allow for the use of single variables from Earth’s water balance (e.g., precipitation, AED, soil moisture).
-Climatic variability impacts vegetation development, with unfavorable conditions such as low precipitation and high temperatures usually promote a decrease in plant productivity.
+Thus, many drought indices exist that allow for a comprehensive assessment of drought on short- to long-term scales and that allow for the use of single variables from Earth's water balance (e.g., precipitation, AED, soil moisture).
+Climatic variability impacts vegetation development, with unfavorable conditions such as low precipitation and high temperatures usually promoting a decrease in plant productivity.
 To monitor the response of vegetation for large areas, the common practice is to use satellite data.
-For example, the Normalized Difference Vegetation Index (NDVI) derived from frequent satellite observations of red and near infrared spectral reflectance, has been widely used as a proxy for biomass production [@Camps-Valls2021; @Paruelo2016;@Helman2014].
+For example, the Normalized Difference Vegetation Index (NDVI) derived from frequent satellite observations of red and near infrared spectral reflectance, has been widely used as a proxy for biomass production @Camps-Valls2021;@Paruelo2016;@Helman2014].
 For Chile's cultivated land, @Zambrano2018 used the zcNDVI for assessing seasonal biomass production in response to drought.
@@
 This mega-drought was defined by the annual time series of the Standardized Precipitation Index (SPI) at a time scale of twelve months at the end of each year (December) when having values below one standard deviation.
-Some studies have addressed how this drought affects single ecosystems in terms of forest growth [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018;@Zambrano2016].
+Some studies have addressed how this drought affects single ecosystems in terms of forest growth [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018,@Zambrano2016].
 The term "mega-drought" is used in Chile to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018].
@@
 
-Continental Chile has diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] ([@fig-studyArea]a), which determines its great ecosystem diversity [@Luebert2022] ([@fig-studyArea]c).
+Continental Chile has diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] [@fig-studyArea]a, which determines its great ecosystem diversity [@Luebert2022] ([@fig-studyArea]c).
 The Andes Mountains are a main factor in climate variation [@Garreaud2009].
-For an aggregated overview of the results of the study, we used the five Chilean macrozones: "Norte Grande," "Norte Chico," "Centro," "Sur," and "Asutral".
-"Norte Grande" (17°34’–25°42’S) and "Norte Chico" (25°42'-32°8'S) predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures.
-At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk).
+For an aggregated overview of the results of the study, we used the five Chilean macrozones: "Norte Grande" (17°34'–25°42'S), "Norte Chico" (25°42'-32°8'S), "Centro" (32°08'-36°12'S), "Sur" (36°12'-43°48'S), and "Austral" (43°48'-56°00'S).
+"Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures.
+At the south of "Norte Chico", the climate changes to an arid steppe with cold temperatures (Bsk).
 In these two northern regions, the land is mostly bare, with a small surface of vegetation types such as shrubland and grassland.
-In the macrozones "Centro" (32°08'-36°12'S) and the northern half of "Sur," (36°12'-43°48'S) the main climate is Mediterranean, with warm to hot summers (Csa and Csb).
+In the macrozones "Centro" and the northern half of "Sur",  the main climate is Mediterranean, with warm to hot summers (Csa and Csb).
 Land cover in "Centro" comprises a significant amount of shrubland and savanna (50%), grassland (16%), forest (8%), and croplands (5%).
-An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral" (43°48'-56°00'S) Those zones have a large areal extent of forest and grassland.
-The southern part of the country has a tundra climate, and in "Austral," it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.
+An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral".
+Those zones have a large area of forest and grassland.
+The southern part of the country has a tundra climate, while "Austral" is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.
 
@@
 MOD13A3 provides vegetation indices with 1km spatial resolution and monthly frequency.
-The NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and MCD12Q1 from the online Data Pool, accessible at https://lpdaa.usgs.gov/tools/data-pool/.
+The NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and MCD12Q1 from the online Data Pool, accessible at [https://lpdaa.usgs.gov/tools/data-pool/](https://lpdaa.usgs.gov/tools/data-pool/).
 
@@
 
-For soil moisture, water supply, and water demand variables, we used ERA5L (ECMWF Reanalysis version 5 over land) [@MunozSabater2021], a reanalysis dataset that provides the evolution of atmospheric and land variables since 1950.
+For soil moisture, water supply, and water demand variables, we used ERA5-Land (ERA5L) (ECMWF Reanalysis version 5 over land) [@MunozSabater2021], a reanalysis dataset that provides the evolution of atmospheric and land variables since 1950.
 It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage.
 We selected the variables for total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100 cm of depth (layer 1 to layer 3).
-\ref{tab-desEOD} shows a summary of the data and its main characteristics.
+Table \ref{tab-desEOD} shows a summary of the data and its main characteristics.
 
@@
 $$R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]$$ {#eq-Ra}
-where
+
+where:
 
@@
 To account for the anomaly, the common practice is to derive it following a statistical parametric method in which it is assumed that the statistical distribution of the data is known [@Heim2002].
-A wrong decision in the statistical distribution is usually the highest source of uncertainty [@Laimighofer2022].
+The use of an erroneous statistical distribution that does not fit the data is usually the highest source of uncertainty [@Laimighofer2022].
 In the case of Chile, due to its high degree of climatic variability, it is difficult to choose a proper distribution without previous research that could be applicable throughout Chile.
@@
 The $A^s_i$ corresponds to a moving window (convolution) that sums the variable for time scales $s$.
-Start from the last month (n) and sum the variable for s months, then follow month by month until the first month in which it could sum for s months (n-s+1).
-For example, using as a variable the precipitation, a period of twelve months (n), and a time scale of three months (s), it will be:
+This summation is done over $s$ months, starting from the most recent month ($n$) back in time until month $n-s+1$.
+For example, using as a variable the precipitation, a period of twelve months ($n$), and a time scale of three months ($s$), it will be:
 
@@
 
-Then, an inverse normal approximation [@Abramowitz1968] obtains the empirically derived probabilities once the variable cumulates over time for the scale $s$.
-Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i$ to derive the $P(a_i)$ probabilities across a period of interest:
+Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i^s$ to derive the $P(A_i^s)$ probabilities across a period of interest:
 
@@
 
-The drought indices $SPI$, $SPEI$, $EDDI$, $SSI$, and $zcNDVI$ are obtained following the inverse normal approximation:
+An inverse normal approximation [@Abramowitz1968] obtains the empirically derived probabilities once the variable cumulates over time for the scale $s$.
+Thus, the drought indices $SPI$, $SPEI$, $EDDI$, $SSI$, and $zcNDVI$ and  obtained following the equation:
 
@@
 We used the NPP from the MOD17A3HGF [@Running2019] dataset (MODIS).
-We choose to use six months because the r-squared of zcNDVI with NPP highly increases from one to six months, but from six to 12 months has a lower improvement.
-We obtained an r-squared of 0.31 for forest and 0.72 for shrubland (refer to the supplementary material in Section S5).
+We chose to use six months because the $R^2$ between zcNDVI and NPP reaches its highest value at six months.
+We obtained an $R^2$ of 0.31 for forest and 0.72 for shrubland (refer to the supplementary material in Section S5).
 Then, we chose the proxy of vegetation productivity for six months, which we will name zcNDVI hereafter.
@@
 
-To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975].
-To determine the magnitude of the trend, we used Sen’s slope [@Sen1968].
-Some of the advantages of applying this methodology are that the Sen's slope is not affected by outliers unlike regular regressions does, and it is a non-parametric method that is not influenced by the distribution of the data.
+To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric Mann-Kendall test [@Kendall1975].
+To determine the magnitude of the trend, we used Sen's slope [@Sen1968].
+Sen's slope has the advantage over normal regression that it is less affected by outliers, and as a non-parametric method it is not influenced by the distribution of the data.
 We applied the Mann-Kendall test to see if the trend was significant and Sen's slope to estimate the magnitude of the trend.
 We did this for the indices SPI, EDDI, SPEI, and SSI using the six time scales with data from 1981 to 2023 (monthly frequency), resulting in 24 trends (per index and time scale).
-Then, we extracted the trend aggregated by each of the five macrozones: "Norte Grande" to "Asutral," and per land cover type: grassland, forest, cropland, shrubland, savanna, and barren land ([@fig-studyArea]d).
+Then, we extracted the trend aggregated by each of the five macrozones: "Norte Grande" to "Austral", and per land cover type: grassland, forest, cropland, shrubland, savanna, and barren land ([@fig-studyArea]d).
 
@@
 
-To analyze the land cover change, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS.
+To analyze the land cover change, we use the IGBP scheme from the MCD12Q1 Collection 6.1 from MODIS.
 This product has been previously used for studies of drought and land cover in Chile [@Fuentes2021;@Zambrano2018].
-We regrouped the 17 classes into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
+We regrouped the 17 classes into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies (Table S1).
 Thus, we have a land cover raster time series with the ten macroclasses for 2001 and 2023.
@@
 
-We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023.
+We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2022.
 After that, we calculated the trend's change in surface per land cover type and macroclass.
@@
 
-To assess how water demand and supply, and soil moisture affect the variation in vegetation productivity across various land cover types, we avoid analyzing areas that have major land cover changes for 2021–2022.
-To assess how zcNDVI varied irrespective of land cover change, we developed a persistence mask for land cover, which only retains pixels for which the macroclass remained the same for at least 80% of the years (2001–2022) ([@fig-studyArea]d).
+To assess how water demand and supply, and soil moisture affect the variation in vegetation productivity across various land cover types, we avoid analyzing areas that experienced major land cover changes in the 2021–2022 period.
+To assess how zcNDVI varied irrespective of land cover change, we developed a persistence mask for land cover, which only retains pixels for which the macroclass remained the same for at least 80% of the 22 years ([@fig-studyArea]d).
 
@@
 
-The goal of this section is to identify which drought indices and time scales have a major impact on changes in land cover type.
-We examined the relationship between the trend in land cover classes and the trend in drought indices.
+To identify which drought indices and time scales have a major impact on changes in land cover type, we examined the relationship between the trend in land cover classes and the trend in drought indices.
 To have more representative results, we conducted the analysis over sub-basins within continental Chile.
-We used 469 basins, which have a surface area between 0.0746 and 24,000 km2 and a median area of 1,249 km2.
+We used 469 basins, which have a surface area between 0.0746 and 24,000 $km^2$ and a median area of 1,249 $km^2$.
 For each basin, we calculated the trend per land cover type, considering the proportion of the type relative to the total surface of the basin.
 Then, we extracted per basin the average trend (Sen's slope) of the drought indices SPI, SPEI, EDDI, SSI, and all their time scales 1, 3, 6, 12, 24, and 36.
-Also, we extracted the average trend in the proxy of vegetation productivity (zcNDVI)
+Also, we extracted the average trend in the proxy of vegetation productivity (zcNDVI).
 
-Random forest [@Ho1995] employs multiple decision trees, allowing for classification and regression.
+We model the trends in land cover per macroclass with the aim of assessing how land cover trends relate to drought indices.
+We used the random forest method [@Ho1995], which employs multiple decision trees, allowing for classification and regression.
 Some advantages include the ability to find non-linear relationships, reduce overfitting, and derive variable importance.
-We used the regression random forest to model the trends in land cover per macroclass, using drought indices as predictors.
-This included the four drought indices per six time scales and the zcNDVI, totaling 25 predictors.
-As a result, we created six random forest models, one per trend in land cover macroclass.
-We trained 1000 forests in a resampled scheme to obtain more reliable results regarding variable importance.
-We resampled by creating ten folds, running a random forest per fold, and calculating the r-squared (rsq), root mean square error (RMSE), and variable importance.
+We included the four drought indices per six time scales and the zcNDVI, totaling 25 predictors.
+As a result, we created thirty random forest models, one for each land cover macroclass trend and per macrozone.
+Each model was trained using 1000 forests in a resampling scheme to obtain more reliable results regarding variable importance.
+We resampled by creating ten folds, running a random forest per fold, and calculating the $R^2$, root mean square error (RMSE), and variable importance.
 The variable importance helps for a better understanding of the relationships by finding which variable has a higher contribution to the model.
-Thus, we calculate the variable's importance by permuting out-of-bag (OOB) data per tree and computing the mean standard error in the OOB.
+Thus, we calculated the variable's importance by permuting out-of-bag (OOB) data per tree and computing the mean standard error in the OOB.
 After permuting each predictor variable, we repeated the process for the remaining variable.
-We repeated this process ten times (per fold) to obtain the performance metrics (rsq, RMSE, and variable importance).
+We repeated this process ten times (per fold) to obtain the performance metrics ($R^2$, RMSE, and variable importance).
 
-Finally, we visually explore the connection between the SPI, EDDI, and SSI drought indices for short- and long-term changes in land cover.
-To do this, we compare the relative changes in land cover surface (in terms of the total surface area per land cover type and macrozone) with the drought indices of six (short-term) and thirty-six months (long-term).
-We created a scatterplot in which the x-axis shows drought indices trends and the y-axis shows land cover change trends.
+Finally, we visually explored the connection between the SPI, EDDI, and SSI drought indices for short- and long-term changes in land cover.
+To do this, we compared the relative changes in land cover surface (in terms of the total surface area per land cover type and macrozone) with the drought indices of six (short-term) and thirty-six months (long-term).
 
@@
 For each land cover macroclass, we analyzed the trend of vegetation productivity over the unchanged land cover macroclasses.
-To achieve this, we used the persistent mask of land cover macroclasses , thus reducing the possibility of evaluating productivity trends that are due to year-to-year variation in land cover.
+To achieve this, we used the persistent mask of land cover macroclasses, thus reducing the possibility of evaluating productivity trends that are due to year-to-year variation in land cover.
 We used the zcNDVI as a proxy of vegetation productivity.
-To assess productivity in Chile's cultivated land, Zambrano et al.
-(2018) used the zcNDVI for assessing seasonal biomass production in relation to climate.
+To assess productivity in Chile's cultivated land, @Zambrano2018 used the zcNDVI for assessing seasonal biomass production in relation to climate.
 
-We examine the drought indices of water demand, water supply, and soil moisture and their correlation with vegetation productivity.
+We examined the drought indices of water demand, water supply, and soil moisture and their correlation with vegetation productivity.
 The objective is to determine to what extent soil moisture and water demand and supply affect vegetation productivity, thus addressing three main questions: 1) Which of the drought variables—supply, demand, or soil moisture—helps most in explaining the changes in vegetation productivity?
-How do the short- to long-term time scales of the drought variable affect vegetation productivity in Chile?
-2) How strong is the relationship between the variables and the drought index?
+2) How do the short- to long-term time scales of the drought variable affect vegetation productivity in Chile, and how strong is the relationship?
 And finally, 3) how does the correlation vary per-land cover type?
@@
 
-For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, `R` [@R2023].
-For downloading ERA5L, we used the `{ecmwfr}` package [@Hufkens2019].
-For processing raster data, we used `{terra}` [@Hijmans2023] and `{stars}` [@Pebesma2023].
-For managing vectorial data, we used `{sf}` [@Pebesma2018].
-For the calculation of AED, we used `{SPEI}` [@Bergueria2023].
-For mapping, we use {tmap} [@Tennekes2018].
+For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, R [@R2023].
+For downloading ERA5L, we used the {ecmwfr} package [@Hufkens2019].
+For processing raster data, we used {terra} [@Hijmans2023] and {stars} [@Pebesma2023].
+For managing vectorial data, we used {sf} [@Pebesma2018].
+For the calculation of AED, we used {SPEI} [@Bergueria2023].
+For mapping, we used {tmap} [@Tennekes2018].
 For data analysis and visualization, the suite {tidyverse} [@Wickham2019] was used.
-For the random forest modeling, we used the {tidymodels}[@Kuhn2020] and {ranger}[@Wright2017] packages.
+For the random forest modeling, we used the {tidymodels} [@Kuhn2020] and {ranger} [@Wright2017] packages.
 
@@
 @fig-trendDI shows the spatial variation of the trend for the drought indices from short- to long-term scales.
-The white space on the maps indicates a not significant trend.
-SPI and SPEI have a decreasing trend from "Norte Chico" to "Sur." However, there is an increasing trend in "Austral." The degree of the trend is stronger at higher time scales.
-The SSI indicates that in "Norte Grande," there are surfaces that have increased in the southwest and in the northeast have decreased, and are shown for all time scales.
+SPI and SPEI have a decreasing trend from "Norte Chico" to "Sur", but an increasing trend in "Austral".
+The degree of the trend is larger at higher time scales.
+In "Norte Grande", the SSI increased in the southwest and decreased in the northeast, for all time scales.
 Similar to SPI and SPEI, SSI decreases at higher time scales.
-EDDI showed a positive trend for the whole of continental Chile, with a higher trend toward the north and a descending gradient toward the south.
-The degree of trend increases at higher time scales.
+EDDI showed a positive trend for the whole of continental Chile, with a higher slope toward the north and a descending gradient toward the south.
+The slope of trend increases at higher time scales.
 
@@
 #| - "SPI (Standardized Precipitation Index)"
+#| - "EDDI (Evaporative Demand Drought Index)"
 #| - "SPEI (Standardized Precipitation Evapotranspiration Index)" 
-#| - "EDDI (Evaporative Demand Drought Index)"
-#| - "SSMI (Standardized Soil Moisture Index)"
+#| - "SSI (Standardized Soil Moisture Index)"
 #| out-width: "\\textwidth"
 
+knitr::include_graphics("../output/figs/trend_raster_SPI_1981-2023.png")
+knitr::include_graphics("../output/figs/trend_raster_EDDI_1981-2023.png")
 knitr::include_graphics("../output/figs/trend_raster_SPEI_1981-2023.png")
-knitr::include_graphics("../output/figs/trend_raster_EDDI_1981-2023.png")
 knitr::include_graphics("../output/figs/trend_raster_zcSM_1981-2023.png")
@@
 
-The @fig-trendDIMacro displays the averaged aggregation per macrozone, the drought index, and the timescale.
-The macrozones that reached the lowest trend for SPI, SPEI, and SSI are "Norte Chico" and "Centro," where the indices also decrease at longer time scales.
+The @fig-trendDIMacro displays the aggregated trend per macrozone, drought index, and timescale.
+The macrozones that reached the lowest trend for SPI, SPEI, and SSI are "Norte Chico" and "Centro", where the indices also decrease at longer time scales.
 This may potentially be explained by the prolonged reduction in precipitation that has affected the hydrological system in Chile.
 At 36 months, it reaches trends between -0.03 and -0.04 (z-score) per decade for SPI, SPEI, and SSI.
-For "Sur," the behavior is similar, decreasing at longer scales and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI.
+For "Sur", the behavior is similar, decreasing at longer scales and having between -0.016 and -0.025 per decade for SPI, SPEI, and SSI.
 "Norte Grande" has the highest trend at 36 months for EDDI (0.042 per decade), and "Centro" has the lowest for SPI and SPEI.
-In "Norte Grande" and "Norte Chico," which are in a semi-arid climate, it is evident that the EDDI has an effect on the difference between the SPI and SPEI index, which is not seen in the other macrozones.
+In "Norte Grande" and "Norte Chico", which are in a semi-arid climate, it is evident that the EDDI has an effect on the difference between the SPI and SPEI index, which is not seen in the other macrozones.
 Contrary to the other macrozones, "Austral" showed an increase in all indices, being the highest for EDDI at 36 months (0.025) and the lowest for SSI, which shows only a minor increase in the trend.
@@
 The largest type of vegetation, with 137,085 $km^2$, is forest.
-Grassland has 74,247 $km^2$, savanna 55,206 $km^2$, shrubland 25,341 $km^2$, and cropland 3,146 $km^2$ (Table \ref{tab-landcoverSurf}).
-The macrozones with major changes for 2001–2022 were "Centro," "Sur," and "Austral," with 36%, 31%, and 34% of their surface changing the type of land cover, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend}).
-@fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistence mask over continental Chile.
+Grassland has 74,247 $km^2$ , savanna 55,206 $km^2$, shrubland 25,341 $km^2$, and cropland 3,146 $km^2$  (Table \ref{tab-landcoverSurf}).
+The macrozones with major changes for 2001–2022 were "Centro", "Sur", and "Austral", with 36%, 31%, and 34% of their surface changing the type of land cover, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend}).
+@fig-LCprop shows the variation for 2001–2022 in the proportion of surface per land cover class and macrozone, derived from the persistence mask over continental Chile.
 
@@
 The red dots on the plots indicate the maximum and minimum values of the surface.
-Cells without values indicate that the land cover type was not significant in that macrozone.}
+The white cells indicate that the landcover class is not significant in terms of surface area.}
 \label{tab-landcoverTrend}
@@
 From the trend analysis in Table \ref{tab-landcoverTrend}, we can indicate that the "Norte Chico" shows an increase in barren land of 111 $km^2 yr^{-1}$  and a reduction in the class savanna of 70 $km^2 yr^{-1}$.
-In the "Centro" and "Sur," there are changes with an important reduction in savanna with 136 and 319 $km^2 yr^{-1}$, respectively, and an increase in shrubland and grassland, showing a change for more dense vegetation types.
-The area under cultivation (croplands) appears to be shifting from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," seemingly replacing the savanna lost (Table \ref{tab-landcoverTrend}).
+In the "Centro" and "Sur", there are changes with an important reduction in savanna with 136 and 319 $km^2 yr^{-1}$, respectively, and an increase in shrubland and grassland, showing a change for more dense vegetation types.
+The area under cultivation (croplands) appears to be shifting from the "Centro" to the "Sur".
+Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur", seemingly replacing the savanna lost (Table  \ref{tab-landcoverTrend}).
 
 ### Relationship between drought indices and land cover change
+
+```{=latex}
+\begin{table}[!ht]
+\caption{The five most important trends of drought indices in estimating the landcover trend per land cover type and the $R^2$ reached by each random forest model.
+The white cells indicate that the landcover class is not significant in terms of surface area.}
+\label{tab-landcoverTrendRF}
+\includegraphics[width=0.5\textwidth]{../output/figs/table_r2_most_important_variables.png}
+\end{table}
+```
 
@@
 #| fig-pos: "!ht"
-#| fig-cap: "cdcdcdc"
-#| out-width: "\\textwidth"
+#| fig-cap: "Relative importance of drought indices for explaining the trend in landcover change across the five macrozones in Chile.
+SPI, Standardized Precipitation Index; EDDI, Evaporative Demand Drought Index; SPEI, Standardized Precipitation Evapotranspiration Index; SSI, Standardized Soil Moisture Index.
+The numbers beside the drough index correspond to the time scales."
+#| out-width: "100%"
 knitr::include_graphics("../output/figs/bars_relative_importance_RF.png")
@@
 
-```{=latex}
-\begin{table}[!ht]
-\caption{The five most important trends of drought indices in estimating the landcover trend per land cover type and the r-squared (rsq) reached by each random forest model.}
-\label{tab-landcoverTrendRF}
-\includegraphics[]{../output/figs/table_r2_most_important_variables.png}
-\end{table}
-```
+Table \ref{tab-landcoverTrendRF} shows the drought indices that are the most important variables in the random forest models, together with the $R^2$ reached.
+The random forest models reached an $R^2$ between 0.12 and 0.41 for the land cover types and macrozones.
+The model shows the highest $R^2$ for shrublands (0.28 to 0.42) and the lowest $R^2$ for croplands (0.16 to 0.20) across all macrozones.
 
-The random forest models for estimating the land cover trend based on the trends in drought indices reach an $R^2$ between 0.32 and 0.39 for the land cover types (Table \ref{tab-landcoverTrendRF}) excluding cropland.
-It is more likely that short- and medium-term increases in AED (EDDI-6 and EDDI-12) and short-term precipitation deficits (SPI-6 and SPEI-6) are associated to changes in grassland and bare land.
-The short-term increase of AED (EDDI-3 and EDDI-6) and the longer duration of the precipitation deficit (SPI-24, SPI-36, and SPEI-36) were the most important variables that correlated with changes in shrubland.
-The changes in savanna are associated with a short- and long-term increase in AED and a three-year precipitation deficit (SPI-36).
-The increase in cumulative AED from 12 to 36 months is the strongest associated variable that contributes to changes in forests, followed by the reduction of soil moisture over six and 36 months.
+@fig-varImportance shows the three most important variables for the models across the five macrozones and per landcover type.
+For shurublands, the SPI of long-term and short-term SSI were the most relevant drought indices within the five macrozones.
+We showed that the trend in short-term EDDI (1–6 months) and long-term SPI (24- and 36-months) affected grasslands and forest changes.
+We showed that trends in SPI-36 and long-term trends in EDDI (12 to 36 months) were associated with changes in savannas.
+The changes in barren land are shown to be limited to the changes in the short-term AED (1 to 6 months).
+Changes in croplands have not been linked to drought across all macrozones.
 The supplementary material in Section S3 provides further details about the variable's importance.
@@
 #| fig-cap: "Relationship between the trend in land cover change (y-axis) and the trend in drought indices (x-axis) for the five macrozones.
-Vertical panels correspond to 1, 3, 6, 12, 24, and 36 months of the time scale by drought index.
-Horizontal panels show each drought index"
+Vertical panels correspond to short (6 months) and long (36 months) time scales.
+Horizontal panels show the drought indices SPI, EDDI, and SSI."
 #| out-width: "\\textwidth"
@@
 
-@fig-TrendsLandDrought shows the connection between the SPI, EDDI, and SSI drought indices and changes in land cover.
-Forest in the "Sur," shrubland and grassland in "Centro," barren land in "Norte Chico," and savanna in "Austral" showed an increase in land cover extent, which was associated with an increase in EDDI.
-Savanna in "Centro," "Sur," and "Norte Chico" decreases with the increase in EDDI.
+@fig-TrendsLandDrought shows the connection between the SPI, EDDI, and SSI drought indices for time scales of 6 and 36 months against changes in land cover.
+Forest in the "Sur", shrubland and grassland in "Centro", barren land in "Norte Chico", and savanna in "Austral" showed an increase in land cover extent, which was associated with an increase in EDDI.
+Savanna in "Centro", "Sur", and "Norte Chico" decreases with the increase in EDDI.
 The SPI and SSI showed similar behavior regarding the trend in land cover type.
-A decrease in SPI and SSI is associated with an increase in the surface in shrubland and grassland in "Centro," forest in "Sur," and barren land in "Norte Chico," as well as a decrease trend in savanna in "Norte Chico," "Centro," and "Sur."
+A decrease in SPI and SSI is associated with an increase in the surface in shrubland and grassland in "Centro", forest in "Sur", and barren land in "Norte Chico", as well as a decrease trend in savanna in "Norte Chico", "Centro", and "Sur".
 
@@
 
+@fig-zcNDVI_var shows a spatial map of trends in zcNDVI ([@fig-zcNDVI_var]a).
+In "Norte Grande", vegetation productivity, as per the z-index, exhibits a yearly increase of 0.027 for grassland and 0.032 for shrubland.
+In "Norte Chico", savanna has the lowest trend slope of -0.062, cropland -0.047, shrubland -0.042, and grassland -0.037.
+In "Centro", shrubland reaches -0.07, savanna -0.031, cropland -0.024, forest -0.017, and grassland -0.005 per decade.
+This decrease in productivity could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning.
+
 The temporal variation within the macrozones is shown in [@fig-zcNDVI_var]b.
@@
 The "Sur" and "Austral" show a positive trend of around 0.012 and 0.016, respectively, per decade (@fig-zcNDVI_var).
-
-In @fig-zcNDVI_var it is showed the spatial map of trends in zcNDVI ([@fig-zcNDVI_var]a).
-In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.027 for grassland and 0.032 for shrubland.
-In the "Norte Chico," savanna has the lowest trend of -0.062, cropland -0.047, shrubland -0.042, and grassland -0.037.
-In "Centro," shrubland reaches -0.07, savanna -0.031, cropland -0.024, forest -0.017, and grassland -0.005 per decade.
-This decrease in productivity could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning.
 
@@
 
-@fig-corTimeScale shows the highest coefficient of determination (r-squared, or rsq) found in the regression analysis between zcNDVI and different drought indicators over time scales of 1, 3, 6, 12, 24, and 36 months.
+@fig-corTimeScale shows the time scales that reached the highest r-squared  in the regression analysis between zcNDVI and different drought indicators over time scales of 1, 3, 6, 12, 24, and 36 months.
 The spatial variation of time scales reached per index is mostly for time scales above 12 months.
 In the case of SSI, the predominant scales are 6 and 12 months.
-For all indices, to the north, the time scales are higher and diminish toward the south until the south part of "Austral," where they increase.
+For all indices, to the north, the time scales are higher and diminish toward the south until the south part of "Austral", where they increase.
 In @fig-corPerson, the map of Pearson correlation values (r) is shown.
-The EDDI reached correlations above 0.5 between "Norte Chico" and "Sur." The correlation changes from negative to positive toward the Andes Mountains and to the sea, just as in the northern part of "Austral." The SPI and SPEI have similar results, with the higher values in "Norte Chico" and "Centro" being higher than 0.6.
+The EDDI reached correlations above 0.5 between “Norte Chico” and "Sur".
+The correlation changes from negative to positive toward the Andes Mountains and to the sea, just as in the northern part of "Austral".
+The SPI and SPEI have similar results, with the higher values in "Norte Chico" and "Centro" being higher than 0.6.
 Following a similar spatial pattern as EDDI but with an opposite sign.
 The SSI showed to be the index that has a major spatial extension with a higher correlation.
-It has a similar correlation to SPI and SPEI for "Norte Chico" and "Sur," but has improvements for "Sur."
+It has a similar correlation to SPI and SPEI for "Norte Chico" and "Centro", but for "Sur" the correlation is higher with SSI.
 
@@
 In Table  \ref{tab-corlandcover}, we aggregate per macrozone and land cover the correlation analysis presented in @fig-corTimeScale and @fig-corPerson.
-According to what is shown, forests seem to be the most resistant to drought.
-Showing that only "Centro" is slightly (rsq = 0.25) impacted by a 12-month soil moisture deficit (SSI-12).
-In the "Norte Chico" and to a lesser extent in the "Norte Grande," it is evident that a SSI-12 with a rsq = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.28 and 0.34, respectively) have an impact on grasslands.
+According to what is shown, forests is likely to be the most resistant to drought.
+Showing that only "Centro" is slightly ($R^2$ = 0.25) impacted by a 12-month soil moisture deficit (SSI-12).
+In the "Norte Chico" and to a lesser extent in the "Norte Grande", it is evident that a SSI-12 with a $R^2$ = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with $R^2$ = 0.28 and 0.34, respectively) have an impact on grasslands.
 However, this type was unaffected by soil moisture, water supply, or demand in macrozones further south.
 The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands.
-For savannas in "Norte Chico," the SSI-12 and SPI-24 reached an rsq of 0.74 and 0.58, respectively.
-This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (rsq = 0.45 in "Centro" and 0.2 in "Sur").
-In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45% and 66% of the variability in "Norte Chico." The type of land most impacted by climatic variation was shrubland, where soil moisture explained 59% and precipitation, 37%, in "Norte Chico" and "Centro," with SSI-12 being the most relevant variable, then SPI-36 in "Norte Chico" and SPI-24 in "Sur."
+For savannas in "Norte Chico", the SSI-12 and SPI-24 reached an $R^2" of 0.74 and 0.58, respectively.
+This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity ($R^2$ = 0.45 in "Centro" and 0.2 in "Sur").
+In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45% and 66% of the variability in "Norte Chico".
+The type of land most impacted by climatic variation was shrubland, where soil moisture explained 59% and precipitation, 37%, in "Norte Chico" and "Centro", with SSI-12 being the most relevant variable, then SPI-36 in "Norte Chico" and SPI-24 in "Sur".
 
@@
 
-In our study, we took into account the variation in vegetation productivity in Chile, specifically in areas without any changes in land cover, to prevent any misleading conclusions about the increase in water demand due to land cover change.
+In our study, we considered the variation in vegetation productivity in Chile, specifically in areas without any changes in land cover, to prevent any misleading conclusions about the increase in water demand due to land cover change.
 Our results show a contrasting perspective regarding the evidence provided by @Vicente-Serrano2022 on a global scale, who indicates that the increase in drought is led by an increase in agricultural land, which in turn increases water demand.
@@
 However, some questions arise regarding what is occurring with the cultivated land.
-Evidence suggests that higher-water-demanding crops have replaced croplands in the Petorca basin (central Chile), leading to an increase in water abstraction [@Munoz2020;@Duran2020].
+Evidence suggests that higher-water-demanding crops have replaced less demanding crops in the Petorca basin (central Chile), leading to an increase in water abstraction [@Munoz2020;@Duran2020].
 Nonetheless, at this scale of analysis, the effect of higher crop water demand on drought is minor compared to the decrease in water supply and increase in AED over all land cover types.
 
-The long-scale trends (e.g., 36 months) demonstrate the impact of human-induced climate change on water availability in Chile, potentially due to an intense hydrological drought stemming from the ongoing precipitation deficit and rising AED.
+The long-scale trends (e.g., 36 months) demonstrate the impact of climate change on water availability in Chile, potentially due to an intense hydrological drought stemming from the ongoing precipitation deficit and rising AED.
 But it is likely that in zones most affected by drought, the main cause is not an increase in vegetation water demand due to an intensification of cultivated land (e.g., an increase in irrigated crops) like in other parts of the globe [@Vicente-Serrano2020].
@@
 We analyzed the time series of drought indices and vegetation productivity per land cover type.
-Our results indicate that forest is the type most resistant to drought, and shrublands, savannas, and croplands have higher sensitivity.
+Our results indicate that forest is the type most resistant to drought, and shrublands, savannas, and croplands have higher sensitivities.
 
@@
 They found that cropland was more sensitive to drought than grassland, showing that cropland responds strongly to short- and medium-term drought (< SPEI-6).
-In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro." In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016;@Potopova2015;@Dai2020;@Rhee2010].
+In our case, the SPEI-12 was the one that most impacted the croplands in "Norte Chico" and "Centro".
+In general, most studies show that croplands are most sensitive to short-term drought (< SPI-6) [@Zambrano2016;@Potopova2015;@Dai2020;@Rhee2010].
 Short-term precipitation deficits have an impact on soil water, so less water is available for plant growth.
-However, we found that in "Norte Chico," an SPI-36 and SPEI-12 had a higher impact, which are associated with long-term water deficit, and in "Centro," an SPI-12 and SPEI-12.
+However, we found that in "Norte Chico", an SPI-36 and SPEI-12 had a higher impact, which are associated with long-term water deficit, and in "Centro", an SPI-12 and SPEI-12.
 Thus, we hypothesize that this impact could be attributed to the hydrological drought that has decreased groundwater storage [@Taucare2024], which in turn is impacted by long-term deficits, and consequently, the vegetation is more dependent on groundwater.
-In "Sur" and "Austral," the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared.
-The possible reason for this is that the most resistant types, forest and grassland, predominate south of "Centro." Also, drought episodes have been less frequent and intense and have had a lower impact on water availability for vegetation.
+In "Sur" and "Austral", the correlations between drought indices and vegetation productivity decrease, as do the time scales that reach the maximum r-squared.
+The possible reason for this is that the most resistant types, forest and grassland, predominate south of "Centro".
+Also, drought episodes have been less frequent and intense and have had a lower impact on water availability for vegetation.
 
-According to @Senf2020, severe drought conditions in Europe are a significant cause of tree mortality.
-However, we discovered that forests are the most resilient land cover class to drought, with less variation in drought indices.
+In central Chile, @Venegas2022 observed a significant decline in the overall growth of sclerophyllous moist forests (mediterranean forests), which they attributed to increased drought conditions.
+However, we found that forests are the most resilient land cover class to drought, with less variation in drought indices.
+In the "Sur", there is a large domain of planted forests that have replaced native vegetation since the 1970s [@Heilmayr2016;@Heilmayr2020;@Miranda2017], impacting biodiversity and ecosystem services [@Rodriguez2018].
+It has recently been shown that these planted forests are responding positively to climate change, and it is expected that they will benefit from future climate scenarios [@Carrasco2022].
+Further, the forests of the “Austral” region correspond to Patagonian ecosystems, mainly native forests dominated by tree species of wide niche breadth .
+Overall, these forests have been more affected by the increase in temperature than by the reduction in moisture [@Fajardo2023;@Holz2018].
+These responses have caused stabilized tree growth, linked to more frequent warm autumns [@Gibson2022].
+It has also been observed that these forests have shown resistance to drought episodes [@Fajardo2023], which might be attributable to their relatively low growth rate.
 Supporting this is @Fathi-Taperasht2022, who assert that Indian forests are the most drought-resistant and recover rapidly.
@@
 
-In central Chile, @Venegas2022 observed a significant decline in the overall growth of sclerophyllous moist forests (mediterranean forests), which they attributed to increased drought conditions.
 In another study, @Fuentes2021 evaluated water scarcity and land cover change in Chile between 29° and 39° south latitude.
@@
 The study results showed that the soil moisture-based drought index (SSI) was better at explaining vegetation productivity across land cover macroclasses than meteorological drought indices like SPI, SPEI, and EDDI.
-According to [@Chatterjee2022] in the early growing season and especially in irrigated rather than rainfed croplands, soil moisture has better skills than SPI and SPEI for estimating gross primary production (GPP) .
+According to [@Chatterjee2022] in the early growing season and especially in irrigated rather than rainfed croplands, soil moisture has better skills than SPI and SPEI for estimating gross primary production (GPP).
 Also, @Zhou2021 indicate that the monthly scaled Standardized Water Deficit Index (SWDI) can accurately show the effects of agricultural drought in most of China.
@@
 The semi-arid "Norte Chico" and the Mediterranean "Centro" were where SSI had the best performance.
-In Chile, medium-term deficits of 12 months are more relevant in the response of vegetation for all land cover types, which decreases to the south, and in the case of croplands, they seem to react in a shorter time, with six months (SSI-6) in "Centro." This variation for croplands could be related to the fact that in "Norte Chico," the majority of crops are irrigated, but to the south there is a higher proportion of rainfed agriculture, which is most dependent on the short-term availability of water.
-Rather, in "Norte Chico," the orchards are more dependent on irrigation, which in turn depends on the availability of storage water in dams of groundwater reservoirs, which are affected by long-term drought (e.g., SPI-36).
+In Chile, medium-term deficits of 12 months are more relevant in the response of vegetation for all land cover types, which decreases to the south, and in the case of croplands, they seem to react in a shorter time, with six months (SSI-6) in "Centro".
+This variation for croplands could be related to the fact that in "Norte Chico", the majority of crops are irrigated, but to the south there is a higher proportion of rainfed agriculture, which is most dependent on the short-term availability of water.
+Rather, in "Norte Chico", the orchards are more dependent on irrigation, which in turn depends on the availability of storage water in dams or groundwater reservoirs, which are affected by long-term drought (e.g., SPI-36).
 
@@
 
-In a commitment to fostering informed and dynamic adaptation efforts, our results are disseminated publicly and continually updated via the Drought Observatory for Agriculture and Biodiversity of Chile (ODES) ([https://odes-chile.org/app/unidades](https://odes-chile.org/app/unidades)) [@Zambrano2023b;@Kunst2023].
-This initiative ensures the availability of extensive climate data, facilitating the development of adaptive strategies that are both responsive to the realities of different regions and grounded in the latest scientific understanding.
+In a commitment to fostering informed and dynamic adaptation efforts, our results are disseminated publicly and continually updated via the Drought Observatory for Agriculture and Biodiversity of Chile (ODES) [https://odes-chile.org/app/unidades](https://odes-chile.org/app/unidades) [@Zambrano2023b;@Kunst2023].
+This initiative ensures the availability and easy accessibility of extensive climate data, facilitating the development of adaptive strategies that are both responsive to the realities of different regions and grounded in the latest scientific understanding.
 The proactive sharing and updating of such data underscores its key role in enabling policymakers to craft adaptive measures that are finely tuned to the diverse and evolving landscapes of drought impacts.
-Furthermore, the recently promulgated law about climate change in Chile (Law 21.455, https://www.bcn.cl/leychile/navegar?idNorma=1177286), which aims to implement sectoral adaptation plans for agriculture, forests, and biodiversity, could benefit from this information.
+Furthermore, the recently promulgated law about climate change in Chile (Law 21.455, [https://www.bcn.cl/leychile/navegar?idNorma=1177286](https://www.bcn.cl/leychile/navegar?idNorma=1177286)), which aims to implement sectoral adaptation plans for agriculture, forests, and biodiversity, could benefit from this information.
 
@@
 
-We  model about 20–30% of the trends in land cover types, such as forest, grassland, shrubland, and savanna, based on drought indicators across Chile.
-There is no evidence that drought alters cropland surface area.
-The increase in AED is the most important variable explaining the variability in the change in land cover, followed by a reduction in precipitation and soil moisture.
+Using Random Forest and variable importance metrics, we assessed how land cover trends relate to drought indices and found that drought indices across Chile could explain about 12–41% of the trends in land cover types.
+Drought explains a higher variance in changes in shrublands, followed by savanna, grasslands, and forest.
+The changes in croplands showed the lowest association with drought.
+We found that the short- to long-term trend in AED was the most important variable that could partially explain the observed land cover trend, followed by long-term trends in precipitation and short-term trends in soil moisture.
 
-The trends in drought indices are accompanied by multiple land cover changes in the country, most notably an increase of forest in "Sur," of shrubland and grassland in "Centro,” and of savanna in "Centro" and "Sur." In "Norte Chico" and "Centro,” the croplands have been declining in surface, whereas in "Sur,” there is an increase in cultivated land.
+The trends in drought indices are accompanied by multiple land cover changes in the country, most notably an increase of forest in "Sur", of shrubland and grassland in "Centro", and of savanna in "Centro" and "Sur".
+In "Norte Chico" and "Centro", the croplands have been declining in surface, whereas in "Sur", there is an increase in cultivated land.
 
@@
 The anomaly in soil moisture over the past 12 months is the main variable explaining these changes, followed by anomalies in accumulated precipitation over one to two years.
-The variation in AED seems to intensify the drought impact on vegetation productivity.
+The variation in AED is likely to exacerbate the drought's impact on vegetation productivity.
 
-The results of this study provide insightful information that would assist in developing adaptation measures for Chilean ecosystems to cope with climate change and drought.
-Information that could be used in the scope of the national law on climate change, which seeks to implement adaptation strategies for agriculture, forests, and biodiversity.
+The results of this study provide insightful information that can assist in developing adaptation measures for Chilean ecosystems to cope with climate change and drought.
+This information could be used in the scope of the national law on climate change, which seeks to implement adaptation strategies for agriculture, forests, and biodiversity.
 

### 2024-05-09 article/paper_drought_Chile2023.tex
update version may 2024

@@
 \makeatletter
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
 \@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \makeatother
@@
 \bibliographystyle{elsarticle-harv}
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 }
- \ead{francisco.fernandez@uss.cl} 
-\author[]{Alejandro Venegas-González%
+\author[10]{Alejandro Venegas-González%
 %
@@
 
-\author[11,12]{Dylan Craven%
+\author[12,13]{Dylan Craven%
 %
@@
 Sustentabilidad, Universidad Mayor},city={Santiago},country={7500994,
-Chile},countrysep={,},postcodesep={}}
+Chile.}}
 \affiliation[2]{organization={Observatorio de Sequía para la Agricultura
@@
 Mayor},city={Santiago},country={7500994,
-Chile},countrysep={,},postcodesep={}}
+Chile.}}
 \affiliation[3]{organization={Faculty of Geo-Information Science and
-Earth, University of Twente},city={Enschede, The
-Netherlands},postcodesep={}}
+Earth, University of Twente},city={Enschede},country={The
+Netherlands.}}
 \affiliation[4]{organization={Facultad de Agronomía y Sistemas
 Naturales, Pontificia Universidad Católica de
-Chile.},city={Santiago},country={Chile},countrysep={,},postcodesep={}}
+Chile.},city={Santiago},country={Chile.}}
 \affiliation[5]{organization={Instituto para el Desarrollo Sustentable.
 Pontificia Universidad Católica de
-Chile},city={Santiago},country={Chile},countrysep={,},postcodesep={}}
+Chile},city={Santiago},country={Chile.}}
 \affiliation[6]{organization={Centro Interdisciplinario de Cambio
-Global, Pontificia Universidad Católica de Chile, Santiago,
-Chile.},,postcodesep={}}
+Global, Pontificia Universidad Católica de
+Chile},city={Santiago},country={Chile.}}
 \affiliation[7]{organization={Hémera Centro de Observación de la Tierra,
@@
 Mayor,},city={Santiago},country={7500994,
-Chile},countrysep={,},postcodesep={}}
+Chile.}}
 \affiliation[8]{organization={Center of Economics for Sustainable
@@
 San
-Sebastian},city={Santiago},country={Chile},countrysep={,},postcodesep={}}
+Sebastian},city={Santiago},country={Chile.}}
 \affiliation[9]{organization={Center of Applied Ecology and
 Sustainability
-(CAPES)},city={Santiago},country={Chile},countrysep={,},postcodesep={}}
-\affiliation[10]{organization={},,postcodesep={}}
+(CAPES)},city={Santiago},country={Chile.}}
+\affiliation[10]{organization={Instituto de Ciencias Agroalimentarias,
+Animales y Ambientales (ICA3), Universidad de O'Higgins},city={San
+Fernando},country={Chile.}}
 \affiliation[11]{organization={, GEMA Center for Genomics, Ecology \&
-Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750,
-Santiago, Chile},,postcodesep={}}
-\affiliation[12]{organization={Data Observatory Foundation, Santiago,
-Chile},,postcodesep={}}
+Environment, Universidad Mayor, Camino La Pirámide Huechuraba
+5750},city={Santiago},country={Chile.}}
+\affiliation[12]{organization={Data Observatory
+Foundation},city={Santiago},country={Chile.}}
 
@@
 persistent period of water scarcity has been defined as a mega-drought.
-There are few studies about the relationship between drought and
-ecosystem changes that can help with a better understanding of
-ecological drought. The aim of our study is to evaluate the interaction
+There is yet insufficient understanding of ecological drought in Chile
+due to the limited studies on the relationship between drought and
+ecosystem changes. The aim of our study is to evaluate the interaction
 of drought, land cover change, and vegetation productivity over
@@
 scales. We derived the drought indices using monthly ERA5-Land
-reanalysis data from 1981 to 2023. We used the Moderate-Resolution
-Imaging Spectroradiometer (MODIS) datasets to derive information on
-annual land cover and monthly vegetation productivity. Our results
-showed that except for the Austral part, Chile has a temporal decreasing
-trend in water supply, and across the whole country, there is an
-increase in AED. These trends become stronger over longer time scales.
-We found a negative trend in vegetation productivity in the
-north-central area, which is more prominent for shrubland and savanna as
-compared to croplands and forests. The anomaly in soil moisture over the
-past 12 months (SSI-12) is the most important variable explaining these
-changes, followed by anomalies in accumulated precipitation over one to
-two years (SPI-12 and SPI-24). The variable importance obtained by
-random forest models indicates that drought is explaining about 20--30\%
-of the change in land cover surface across Chile for forest, grassland,
-shrubland, and savanna but has no relation to the changes in croplands.
-The increase in AED is the main variable associated with the change in
-land cover, followed by a reduction in precipitation and soil moisture.
-Our findings provide insightful information that could assist in
-developing adaptation measures for Chilean ecosystems to cope with
-climate change and drought. Also, this study could contribute to a
-better comprehension of ecological drought.
+reanalysis data from 1981 to 2023. We used Moderate-Resolution Imaging
+Spectroradiometer (MODIS) datasets to derive information on annual land
+cover and monthly vegetation productivity. Our results showed that,
+except for the Austral part, Chile has a temporal decreasing trend in
+water supply, and across the whole country, there is an increase in AED.
+These trends become stronger over longer time scales. We found a
+negative trend in vegetation productivity in the north-central area,
+which is more prominent for shrubland and savanna as compared to
+croplands and forests. The anomaly in soil moisture over the past 12
+months (SSI-12) is the most important variable explaining these changes,
+followed by anomalies in accumulated precipitation over one to two years
+(SPI-12 and SPI-24). The variable importance obtained by random forest
+models indicates that drought explains about 12--41\% of the change in
+land cover surface across Chile for forest, grassland, shrubland, and
+savanna but has little relation to the changes in croplands. The
+increase in AED is the main variable associated with the change in land
+cover, followed by a reduction in precipitation and soil moisture. Our
+findings provide insightful information that could assist in developing
+adaptation measures for Chilean ecosystems to cope with climate change
+and drought.
 \end{abstract}
@@
 \end{frontmatter}
+
 \captionsetup{justification=raggedright,singlelinecheck=false}
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, boxrule=0pt, sharp corners, breakable]}{\end{tcolorbox}}\fi
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Drought is often classified as 1) meteorological when precipitation in a
+\section{Introduction}\label{introduction}
+
+Drought can be classified as 1) meteorological, when precipitation in a
 specific period remains below the mean precipitation experienced in the
-same period during multiple years (more than 30 years usually), 2)
-hydrological when these anomalies last for long periods (months to
-years) and affect water systems, and 3) agricultural when the deficit
+same period during multiple years (more than 30 years usually); 2)
+hydrological, when these anomalies last for long periods (months to
+years) and affect water systems; and 3) agricultural, when the deficit
 negatively impacts plant health and leads to decreased productivity of
@@
 activities should also be considered (i.e., anthropogenic drought).
-Droughts lead to increased tree mortality \citep{Cheng2024} and induces
-alterations in land cover and land use, ultimately affecting ecosystems
-\citep{Crausbay2017}. Even though many ecological studies have at times
-mistakenly considered ``dry'' conditions as ``drought''
-\citep{Slette2019}. Ecological drought can be defined as \emph{``an
+Droughts can lead to increased tree mortality \citep{Cheng2024} and
+induce alterations in land cover and land use, ultimately affecting
+ecosystems \citep{Crausbay2017}. Even though many ecological studies
+have at times mistakenly considered ``dry'' conditions as ``drought''
+\citep{Slette2019}, ecological drought can be defined as \emph{``an
 episodic deficit in water availability that drives ecosystems beyond
@@
 global warming stabilizes at 1.5°--2°C. To better evaluate the impact of
-drought trends on ecosystems, assessments that correlate meteorological
-and soil moisture variables to their effects on vegetation are much
-needed.
+drought trends on ecosystems, assessments of the relationship between
+meteorological and soil moisture variables and their effects on
+vegetation are much needed.
 
@@
 i) how do land cover and vegetation productivity respond to short- to
-long-term meteorological and soil moisture droughts? And ii) How is this
+long-term meteorological and soil moisture droughts? And ii) how is this
 response different between humid and arid climatic zones? Likewise,
@@
 drought \citep{Gebrechorkos2023, Liu2024}. Recently, there has been more
-interest in using AED to track droughts separately to better disengage
-the effects of precipitation from temperature-dependent effects
-\citep{Vicente-Serrano2020}. One of the reasons is that AED is linked to
-more flash droughts in limited water regions \citep{Noguera2022}.
+interest in using AED to track droughts separately to better disentangle
+precipitation from temperature-dependent effects
+\citep{Vicente-Serrano2020}. One of the reasons is that AED is more
+linked to flash droughts in water-limited regions \citep{Noguera2022}.
 \citet{Hobbins2016} and \citet{McEvoy2016} developed the Evaporative
@@
 conditions such as low precipitation and high temperatures usually
-promote a decrease in plant productivity. To monitor the response of
+promoting a decrease in plant productivity. To monitor the response of
 vegetation for large areas, the common practice is to use satellite
@@
 spectral reflectance, has been widely used as a proxy for biomass
-production \citep{Camps-Valls2021, Paruelo2016, Helman2014}. For Chile's
-cultivated land, \citet{Zambrano2018} used the zcNDVI for assessing
-seasonal biomass production in response to drought. Comparing the
-various meteo-related and vegetation-based drought indices, we can
+production
+\citet{Camps-Valls2021};\citet{Paruelo2016};\citet{Helman2014}{]}. For
+Chile's cultivated land, \citet{Zambrano2018} used the zcNDVI for
+assessing seasonal biomass production in response to drought. Comparing
+the various meteo-related and vegetation-based drought indices, we can
 further our understanding of the impact of drought on ecosystems.
@@
 \citep{UrrutiaJalabert2018}, and crop productivity
-\citep{Zambrano2023, Zambrano2018, Zambrano2016}. The term
+\citep[\citet{Zambrano2016}]{Zambrano2023, Zambrano2018}. The term
 ``mega-drought'' is used in Chile to describe a prolonged water shortage
@@
 
-\hypertarget{study-area}{%
-\section{Study area}\label{study-area}}
+\section{Study area}\label{study-area}
 
@@
 from north to south and east to west \citep{Aceituno2021}
-(Figure~\ref{fig-studyArea}a), which determines its great ecosystem
+Figure~\ref{fig-studyArea}a, which determines its great ecosystem
 diversity \citep{Luebert2022} (Figure~\ref{fig-studyArea}c). The Andes
@@
 For an aggregated overview of the results of the study, we used the five
-Chilean macrozones: ``Norte Grande,'' ``Norte Chico,'' ``Centro,''
-``Sur,'' and ``Asutral''. ``Norte Grande'' (17°34'--25°42'S) and ``Norte
-Chico'' (25°42'-32°8'S) predominate in an arid desert climate with hot
-(Bwh) and cold (Bwk) temperatures. At the south of ``Norte Chico,'' the
-climate changes to an arid steppe with cold temperatures (Bsk). In these
-two northern regions, the land is mostly bare, with a small surface of
-vegetation types such as shrubland and grassland. In the macrozones
-``Centro'' (32°08'-36°12'S) and the northern half of ``Sur,''
-(36°12'-43°48'S) the main climate is Mediterranean, with warm to hot
-summers (Csa and Csb). Land cover in ``Centro'' comprises a significant
-amount of shrubland and savanna (50\%), grassland (16\%), forest (8\%),
-and croplands (5\%). An oceanic climate (Cfb) predominates in the south
-of ``Sur'' and the north of ``Austral'' (43°48'-56°00'S) Those zones
-have a large areal extent of forest and grassland. The southern part of
-the country has a tundra climate, and in ``Austral,'' it is a cold
-semi-arid area with an extended surface of grassland, forest, and, to a
-lesser extent, savanna.
+Chilean macrozones: ``Norte Grande'' (17°34'--25°42'S), ``Norte Chico''
+(25°42'-32°8'S), ``Centro'' (32°08'-36°12'S), ``Sur'' (36°12'-43°48'S),
+and ``Austral'' (43°48'-56°00'S). ``Norte Grande'' and ``Norte Chico''
+predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
+temperatures. At the south of ``Norte Chico'', the climate changes to an
+arid steppe with cold temperatures (Bsk). In these two northern regions,
+the land is mostly bare, with a small surface of vegetation types such
+as shrubland and grassland. In the macrozones ``Centro'' and the
+northern half of ``Sur'', the main climate is Mediterranean, with warm
+to hot summers (Csa and Csb). Land cover in ``Centro'' comprises a
+significant amount of shrubland and savanna (50\%), grassland (16\%),
+forest (8\%), and croplands (5\%). An oceanic climate (Cfb) predominates
+in the south of ``Sur'' and the north of ``Austral''. Those zones have a
+large area of forest and grassland. The southern part of the country has
+a tundra climate, while ``Austral'' is a cold semi-arid area with an
+extended surface of grassland, forest, and, to a lesser extent, savanna.
 
@@
 
-{\centering \includegraphics{../output/figs/map_study_con_landcover.png}
+\centering{
+
+\includegraphics{../output/figs/map_study_con_landcover.png}
 
@@
 
-\end{figure*}
-
-\hypertarget{materials-and-methods}{%
-\section{Materials and Methods}\label{materials-and-methods}}
-
-\hypertarget{data}{%
-\subsection{Data}\label{data}}
-
-\hypertarget{gridded-meteorological-and-vegetation-data}{%
+\end{figure*}%
+
+\section{Materials and Methods}\label{materials-and-methods}
+
+\subsection{Data}\label{data}
+
 \subsubsection{Gridded meteorological and vegetation
-data}\label{gridded-meteorological-and-vegetation-data}}
+data}\label{gridded-meteorological-and-vegetation-data}
 
@@
 Dakota, provided the MOD13A3 and MCD12Q1 from the online Data Pool,
-accessible at https://lpdaa.usgs.gov/tools/data-pool/.
-
-\begin{table}[!ht]
-\caption{Description of the satellite and reanalysis data used}
-\label{tab-desEOD}
+accessible at \url{https://lpdaa.usgs.gov/tools/data-pool/}.
+
+\begin{table}
+
+\centering{
+
+[!ht]
+
 \scriptsize
@@
 {\raggedright *Calculated from maximum and minimum temperatures derived from ERA5L with Eq. \ref{eq-AED}. \par}
-\end{table}
+
+}
+
+\caption{\label{tab-desEOD}Description of the satellite and reanalysis data used}
+
+\end{table}%
 
 For soil moisture, water supply, and water demand variables, we used
-ERA5L (ECMWF Reanalysis version 5 over land) \citep{MunozSabater2021}, a
-reanalysis dataset that provides the evolution of atmospheric and land
-variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly
-frequency, and global coverage. We selected the variables for total
-precipitation, maximum and minimum temperature at 2 meters, and
-volumetric soil water layers between 0 and 100 cm of depth (layer 1 to
-layer 3). \ref{tab-desEOD} shows a summary of the data and its main
-characteristics.
-
-\hypertarget{short--to-long-term-drought-trends}{%
+ERA5-Land (ERA5L) (ECMWF Reanalysis version 5 over land)
+\citep{MunozSabater2021}, a reanalysis dataset that provides the
+evolution of atmospheric and land variables since 1950. It has a spatial
+resolution of 0.1° (9 km), hourly frequency, and global coverage. We
+selected the variables for total precipitation, maximum and minimum
+temperature at 2 meters, and volumetric soil water layers between 0 and
+100 cm of depth (layer 1 to layer 3). Table \ref{tab-desEOD} shows a
+summary of the data and its main characteristics.
+
 \subsection{Short- to long-term drought
-trends}\label{short--to-long-term-drought-trends}}
-
-\hypertarget{atmospheric-evaporative-demand-aed}{%
+trends}\label{short--to-long-term-drought-trends}
+
 \subsubsection{Atmospheric Evaporative Demand
-(AED)}\label{atmospheric-evaporative-demand-aed}}
+(AED)}\label{atmospheric-evaporative-demand-aed}
 
@@
 
-\begin{equation}\protect\hypertarget{eq-AED}{}{AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}}\label{eq-AED}\end{equation}
+\begin{equation}\phantomsection\label{eq-AED}{AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}}\end{equation}
 
@@
 
-\begin{equation}\protect\hypertarget{eq-Ra}{}{R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]}\label{eq-Ra}\end{equation}
-where
+\begin{equation}\phantomsection\label{eq-Ra}{R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]}\end{equation}
+
+where:
 
@@
 
-\hypertarget{non-parametric-calculation-of-drought-indices}{%
 \subsubsection{Non-parametric calculation of drought
-indices}\label{non-parametric-calculation-of-drought-indices}}
+indices}\label{non-parametric-calculation-of-drought-indices}
 
@@
 assumed that the statistical distribution of the data is known
-\citep{Heim2002}. A wrong decision in the statistical distribution is
-usually the highest source of uncertainty \citep{Laimighofer2022}. In
-the case of Chile, due to its high degree of climatic variability, it is
-difficult to choose a proper distribution without previous research that
-could be applicable throughout Chile. Here, we follow a non-parametric
-method for the calculation of the drought indices, in a similar manner
-as the framework proposed by \citet{Farahmand2015}.
+\citep{Heim2002}. The use of an erroneous statistical distribution that
+does not fit the data is usually the highest source of uncertainty
+\citep{Laimighofer2022}. In the case of Chile, due to its high degree of
+climatic variability, it is difficult to choose a proper distribution
+without previous research that could be applicable throughout Chile.
+Here, we follow a non-parametric method for the calculation of the
+drought indices, in a similar manner as the framework proposed by
+\citet{Farahmand2015}.
 
@@
 
-\begin{equation}\protect\hypertarget{eq-sumvar}{}{A^s_i = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  }\label{eq-sumvar}\end{equation}
+\begin{equation}\phantomsection\label{eq-sumvar}{A^s_i = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  }\end{equation}
 
 The \(A^s_i\) corresponds to a moving window (convolution) that sums the
-variable for time scales \(s\). Start from the last month (n) and sum
-the variable for s months, then follow month by month until the first
-month in which it could sum for s months (n-s+1). For example, using as
-a variable the precipitation, a period of twelve months (n), and a time
-scale of three months (s), it will be:
+variable for time scales \(s\). This summation is done over \(s\)
+months, starting from the most recent month (\(n\)) back in time until
+month \(n-s+1\). For example, using as a variable the precipitation, a
+period of twelve months (\(n\)), and a time scale of three months
+(\(s\)), it will be:
 
@@
 
-Then, an inverse normal approximation \citep{Abramowitz1968} obtains the
+Then, we used the empirical Tukey plotting position \citep{Wilks2011}
+over \(A_i^s\) to derive the \(P(A_i^s)\) probabilities across a period
+of interest:
+
+\begin{equation}\phantomsection\label{eq-probPai}{P(A^s_i) = \frac{i-0.33}{n+0.33'}}\end{equation}
+
+An inverse normal approximation \citep{Abramowitz1968} obtains the
 empirically derived probabilities once the variable cumulates over time
-for the scale \(s\). Then, we used the empirical Tukey plotting position
-\citep{Wilks2011} over \(A_i\) to derive the \(P(a_i)\) probabilities
-across a period of interest:
-
-\begin{equation}\protect\hypertarget{eq-probPai}{}{P(A^s_i) = \frac{i-0.33}{n+0.33'}}\label{eq-probPai}\end{equation}
-
-The drought indices \(SPI\), \(SPEI\), \(EDDI\), \(SSI\), and \(zcNDVI\)
-are obtained following the inverse normal approximation:
-
-\begin{equation}\protect\hypertarget{eq-DI}{}{DI(A^s_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}}\label{eq-DI}\end{equation}
+for the scale \(s\). Thus, the drought indices \(SPI\), \(SPEI\),
+\(EDDI\), \(SSI\), and \(zcNDVI\) and obtained following the equation:
+
+\begin{equation}\phantomsection\label{eq-DI}{DI(A^s_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}}\end{equation}
 
@@
 annual NPP. We used the NPP from the MOD17A3HGF \citep{Running2019}
-dataset (MODIS). We choose to use six months because the r-squared of
-zcNDVI with NPP highly increases from one to six months, but from six to
-12 months has a lower improvement. We obtained an r-squared of 0.31 for
-forest and 0.72 for shrubland (refer to the supplementary material in
-Section S5). Then, we chose the proxy of vegetation productivity for six
-months, which we will name zcNDVI hereafter. It was calculated at a
-monthly frequency for 2000--2023.
-
-\hypertarget{trend-of-drought-indices}{%
-\subsubsection{Trend of drought
-indices}\label{trend-of-drought-indices}}
+dataset (MODIS). We chose to use six months because the \(R^2\) between
+zcNDVI and NPP reaches its highest value at six months. We obtained an
+\(R^2\) of 0.31 for forest and 0.72 for shrubland (refer to the
+supplementary material in Section S5). Then, we chose the proxy of
+vegetation productivity for six months, which we will name zcNDVI
+hereafter. It was calculated at a monthly frequency for 2000--2023.
+
+\subsubsection{Trend of drought indices}\label{trend-of-drought-indices}
 
 To estimate if there are significant positive or negative trends for the
-drought indices, we used the non-parametric test of Mann-Kendall
+drought indices, we used the non-parametric Mann-Kendall test
 \citep{Kendall1975}. To determine the magnitude of the trend, we used
-Sen's slope \citep{Sen1968}. Some of the advantages of applying this
-methodology are that the Sen's slope is not affected by outliers unlike
-regular regressions does, and it is a non-parametric method that is not
-influenced by the distribution of the data. We applied the Mann-Kendall
-test to see if the trend was significant and Sen's slope to estimate the
-magnitude of the trend. We did this for the indices SPI, EDDI, SPEI, and
-SSI using the six time scales with data from 1981 to 2023 (monthly
-frequency), resulting in 24 trends (per index and time scale). Then, we
-extracted the trend aggregated by each of the five macrozones: ``Norte
-Grande'' to ``Asutral,'' and per land cover type: grassland, forest,
-cropland, shrubland, savanna, and barren land
+Sen's slope \citep{Sen1968}. Sen's slope has the advantage over normal
+regression that it is less affected by outliers, and as a non-parametric
+method it is not influenced by the distribution of the data. We applied
+the Mann-Kendall test to see if the trend was significant and Sen's
+slope to estimate the magnitude of the trend. We did this for the
+indices SPI, EDDI, SPEI, and SSI using the six time scales with data
+from 1981 to 2023 (monthly frequency), resulting in 24 trends (per index
+and time scale). Then, we extracted the trend aggregated by each of the
+five macrozones: ``Norte Grande'' to ``Austral'', and per land cover
+type: grassland, forest, cropland, shrubland, savanna, and barren land
 (Figure~\ref{fig-studyArea}d).
 
-\hypertarget{interaction-of-land-cover-and-drought}{%
 \subsection{Interaction of land cover and
-drought}\label{interaction-of-land-cover-and-drought}}
-
-\hypertarget{land-cover-change}{%
-\subsubsection{Land cover change}\label{land-cover-change}}
+drought}\label{interaction-of-land-cover-and-drought}
+
+\subsubsection{Land cover change}\label{land-cover-change}
 
 To analyze the land cover change, we use the IGBP scheme from the
-MCD12Q1 collection 6.1 from MODIS. This product has been previously used
+MCD12Q1 Collection 6.1 from MODIS. This product has been previously used
 for studies of drought and land cover in Chile
@@
 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
-13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.
-Thus, we have a land cover raster time series with the ten macroclasses
-for 2001 and 2023. We validate the land cover macroclasses regarding a
-highly detailed (30 m of spatial resolution) land cover map made for
-Chile by \citet{Zhao2016} for 2013-2014. Our results showed a global
-accuracy of \textasciitilde0.82 and a F1 score of \textasciitilde0.66.
-Section S2 in the Supplementary Material shows the procedure for
-validation.
+13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies
+(Table S1). Thus, we have a land cover raster time series with the ten
+macroclasses for 2001 and 2023. We validate the land cover macroclasses
+regarding a highly detailed (30 m of spatial resolution) land cover map
+made for Chile by \citet{Zhao2016} for 2013-2014. Our results showed a
+global accuracy of \textasciitilde0.82 and a F1 score of
+\textasciitilde0.66. Section S2 in the Supplementary Material shows the
+procedure for validation.
 
 We calculated the surface occupied per land cover class into the five
-macrozones (``Norte Grande'' to ``Austral'') per year for 2001--2023.
+macrozones (``Norte Grande'' to ``Austral'') per year for 2001--2022.
 After that, we calculated the trend's change in surface per land cover
@@
 variation in vegetation productivity across various land cover types, we
-avoid analyzing areas that have major land cover changes for 2021--2022.
-To assess how zcNDVI varied irrespective of land cover change, we
-developed a persistence mask for land cover, which only retains pixels
-for which the macroclass remained the same for at least 80\% of the
-years (2001--2022) (Figure~\ref{fig-studyArea}d).
-
-\hypertarget{relationship-between-land-cover-and-drought-trends}{%
+avoid analyzing areas that experienced major land cover changes in the
+2021--2022 period. To assess how zcNDVI varied irrespective of land
+cover change, we developed a persistence mask for land cover, which only
+retains pixels for which the macroclass remained the same for at least
+80\% of the 22 years (Figure~\ref{fig-studyArea}d).
+
 \subsubsection{Relationship between land cover and drought
-trends}\label{relationship-between-land-cover-and-drought-trends}}
-
-The goal of this section is to identify which drought indices and time
-scales have a major impact on changes in land cover type. We examined
-the relationship between the trend in land cover classes and the trend
-in drought indices. To have more representative results, we conducted
-the analysis over sub-basins within continental Chile. We used 469
-basins, which have a surface area between 0.0746 and 24,000 km2 and a
-median area of 1,249 km2. For each basin, we calculated the trend per
-land cover type, considering the proportion of the type relative to the
-total surface of the basin. Then, we extracted per basin the average
-trend (Sen's slope) of the drought indices SPI, SPEI, EDDI, SSI, and all
-their time scales 1, 3, 6, 12, 24, and 36. Also, we extracted the
-average trend in the proxy of vegetation productivity (zcNDVI)
-
-Random forest \citep{Ho1995} employs multiple decision trees, allowing
-for classification and regression. Some advantages include the ability
-to find non-linear relationships, reduce overfitting, and derive
-variable importance. We used the regression random forest to model the
-trends in land cover per macroclass, using drought indices as
-predictors. This included the four drought indices per six time scales
-and the zcNDVI, totaling 25 predictors. As a result, we created six
-random forest models, one per trend in land cover macroclass. We trained
-1000 forests in a resampled scheme to obtain more reliable results
-regarding variable importance. We resampled by creating ten folds,
-running a random forest per fold, and calculating the r-squared (rsq),
-root mean square error (RMSE), and variable importance. The variable
-importance helps for a better understanding of the relationships by
-finding which variable has a higher contribution to the model. Thus, we
-calculate the variable's importance by permuting out-of-bag (OOB) data
-per tree and computing the mean standard error in the OOB. After
-permuting each predictor variable, we repeated the process for the
-remaining variable. We repeated this process ten times (per fold) to
-obtain the performance metrics (rsq, RMSE, and variable importance).
-
-Finally, we visually explore the connection between the SPI, EDDI, and
+trends}\label{relationship-between-land-cover-and-drought-trends}
+
+To identify which drought indices and time scales have a major impact on
+changes in land cover type, we examined the relationship between the
+trend in land cover classes and the trend in drought indices. To have
+more representative results, we conducted the analysis over sub-basins
+within continental Chile. We used 469 basins, which have a surface area
+between 0.0746 and 24,000 \(km^2\) and a median area of 1,249 \(km^2\).
+For each basin, we calculated the trend per land cover type, considering
+the proportion of the type relative to the total surface of the basin.
+Then, we extracted per basin the average trend (Sen's slope) of the
+drought indices SPI, SPEI, EDDI, SSI, and all their time scales 1, 3, 6,
+12, 24, and 36. Also, we extracted the average trend in the proxy of
+vegetation productivity (zcNDVI).
+
+We model the trends in land cover per macroclass with the aim of
+assessing how land cover trends relate to drought indices. We used the
+random forest method \citep{Ho1995}, which employs multiple decision
+trees, allowing for classification and regression. Some advantages
+include the ability to find non-linear relationships, reduce
+overfitting, and derive variable importance. We included the four
+drought indices per six time scales and the zcNDVI, totaling 25
+predictors. As a result, we created thirty random forest models, one for
+each land cover macroclass trend and per macrozone. Each model was
+trained using 1000 forests in a resampling scheme to obtain more
+reliable results regarding variable importance. We resampled by creating
+ten folds, running a random forest per fold, and calculating the
+\(R^2\), root mean square error (RMSE), and variable importance. The
+variable importance helps for a better understanding of the
+relationships by finding which variable has a higher contribution to the
+model. Thus, we calculated the variable's importance by permuting
+out-of-bag (OOB) data per tree and computing the mean standard error in
+the OOB. After permuting each predictor variable, we repeated the
+process for the remaining variable. We repeated this process ten times
+(per fold) to obtain the performance metrics (\(R^2\), RMSE, and
+variable importance).
+
+Finally, we visually explored the connection between the SPI, EDDI, and
 SSI drought indices for short- and long-term changes in land cover. To
-do this, we compare the relative changes in land cover surface (in terms
-of the total surface area per land cover type and macrozone) with the
-drought indices of six (short-term) and thirty-six months (long-term).
-We created a scatterplot in which the x-axis shows drought indices
-trends and the y-axis shows land cover change trends.
-
-\hypertarget{drought-impacts-on-vegetation-productivity}{%
+do this, we compared the relative changes in land cover surface (in
+terms of the total surface area per land cover type and macrozone) with
+the drought indices of six (short-term) and thirty-six months
+(long-term).
+
 \subsection{Drought impacts on vegetation
-productivity}\label{drought-impacts-on-vegetation-productivity}}
+productivity}\label{drought-impacts-on-vegetation-productivity}
 
@@
 productivity over the unchanged land cover macroclasses. To achieve
-this, we used the persistent mask of land cover macroclasses , thus
+this, we used the persistent mask of land cover macroclasses, thus
 reducing the possibility of evaluating productivity trends that are due
@@
 of vegetation productivity. To assess productivity in Chile's cultivated
-land, Zambrano et al.~(2018) used the zcNDVI for assessing seasonal
+land, \citet{Zambrano2018} used the zcNDVI for assessing seasonal
 biomass production in relation to climate.
 
-We examine the drought indices of water demand, water supply, and soil
+We examined the drought indices of water demand, water supply, and soil
 moisture and their correlation with vegetation productivity. The
@@
 moisture---helps most in explaining the changes in vegetation
-productivity? How do the short- to long-term time scales of the drought
-variable affect vegetation productivity in Chile? 2) How strong is the
-relationship between the variables and the drought index? And finally,
-3) how does the correlation vary per-land cover type? Answering these
-questions should advance our understanding of how climate is affecting
-vegetation, considering the impact on the five land cover types: forest,
-cropland, grassland, savanna, and shrubland.
+productivity? 2) How do the short- to long-term time scales of the
+drought variable affect vegetation productivity in Chile, and how strong
+is the relationship? And finally, 3) how does the correlation vary
+per-land cover type? Answering these questions should advance our
+understanding of how climate is affecting vegetation, considering the
+impact on the five land cover types: forest, cropland, grassland,
+savanna, and shrubland.
 
@@
 
-\hypertarget{software}{%
-\subsection{Software}\label{software}}
+\subsection{Software}\label{software}
 
@@
 data, we used the open source software for statistical computing and
-graphics, \texttt{R} \citep{R2023}. For downloading ERA5L, we used the
-\texttt{\{ecmwfr\}} package \citep{Hufkens2019}. For processing raster
-data, we used \texttt{\{terra\}} \citep{Hijmans2023} and
-\texttt{\{stars\}} \citep{Pebesma2023}. For managing vectorial data, we
-used \texttt{\{sf\}} \citep{Pebesma2018}. For the calculation of AED, we
-used \texttt{\{SPEI\}} \citep{Bergueria2023}. For mapping, we use
-\{tmap\} \citep{Tennekes2018}. For data analysis and visualization, the
-suite \{tidyverse\} \citep{Wickham2019} was used. For the random forest
-modeling, we used the \{tidymodels\}\citep{Kuhn2020} and
-\{ranger\}\citep{Wright2017} packages.
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{short--to-long-term-drought-trends-1}{%
+graphics, R \citep{R2023}. For downloading ERA5L, we used the \{ecmwfr\}
+package \citep{Hufkens2019}. For processing raster data, we used
+\{terra\} \citep{Hijmans2023} and \{stars\} \citep{Pebesma2023}. For
+managing vectorial data, we used \{sf\} \citep{Pebesma2018}. For the
+calculation of AED, we used \{SPEI\} \citep{Bergueria2023}. For mapping,
+we used \{tmap\} \citep{Tennekes2018}. For data analysis and
+visualization, the suite \{tidyverse\} \citep{Wickham2019} was used. For
+the random forest modeling, we used the \{tidymodels\} \citep{Kuhn2020}
+and \{ranger\} \citep{Wright2017} packages.
+
+\section{Results}\label{results}
+
 \subsection{Short- to long-term drought
-trends}\label{short--to-long-term-drought-trends-1}}
+trends}\label{short--to-long-term-drought-trends-1}
 
 Figure~\ref{fig-trendDI} shows the spatial variation of the trend for
-the drought indices from short- to long-term scales. The white space on
-the maps indicates a not significant trend. SPI and SPEI have a
-decreasing trend from ``Norte Chico'' to ``Sur.'' However, there is an
-increasing trend in ``Austral.'' The degree of the trend is stronger at
-higher time scales. The SSI indicates that in ``Norte Grande,'' there
-are surfaces that have increased in the southwest and in the northeast
-have decreased, and are shown for all time scales. Similar to SPI and
+the drought indices from short- to long-term scales. SPI and SPEI have a
+decreasing trend from ``Norte Chico'' to ``Sur'', but an increasing
+trend in ``Austral''. The degree of the trend is larger at higher time
+scales. In ``Norte Grande'', the SSI increased in the southwest and
+decreased in the northeast, for all time scales. Similar to SPI and
 SPEI, SSI decreases at higher time scales. EDDI showed a positive trend
-for the whole of continental Chile, with a higher trend toward the north
-and a descending gradient toward the south. The degree of trend
-increases at higher time scales.
+for the whole of continental Chile, with a higher slope toward the north
+and a descending gradient toward the south. The slope of trend increases
+at higher time scales.
 
@@
 
-\begin{figure}
-
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}
-
-}
+\begin{figure*}[!ht]
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}
 
@@
 \subcaption{\label{fig-trendDI-1}SPI (Standardized Precipitation Index)}
+
 \end{minipage}%
 %
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
+\begin{minipage}{0.50\linewidth}
+
+\centering{
 
@@
 
-}
-
-\subcaption{\label{fig-trendDI-2}SPEI (Standardized Precipitation
-Evapotranspiration Index)}
+\subcaption{\label{fig-trendDI-2}EDDI (Evaporative Demand Drought
+Index)}
+
 \end{minipage}%
 \newline
-\begin{minipage}[t]{0.50\linewidth}
-
-{\centering
-
-\raisebox{-\height}{
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}
+
+}
+
+\subcaption{\label{fig-trendDI-3}SPEI (Standardized Precipitation
+Evapotranspiration Index)}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
 
@@
 
-}
-
-\subcaption{\label{fig-trendDI-3}EDDI (Evaporative Demand Drought
-Index)}
+\subcaption{\label{fig-trendDI-4}SSI (Standardized Soil Moisture Index)}
+
 \end{minipage}%
@@
 
-\end{figure}
+\end{figure*}%
 
@@
 
-{\centering \includegraphics{../output/figs/trend_macrozone_drought_indices.png}
+\centering{
+
+\includegraphics{../output/figs/trend_macrozone_drought_indices.png}
 
@@
 
-\end{figure}
-
-The Figure~\ref{fig-trendDIMacro} displays the averaged aggregation per
-macrozone, the drought index, and the timescale. The macrozones that
-reached the lowest trend for SPI, SPEI, and SSI are ``Norte Chico'' and
-``Centro,'' where the indices also decrease at longer time scales. This
-may potentially be explained by the prolonged reduction in precipitation
+\end{figure}%
+
+The Figure~\ref{fig-trendDIMacro} displays the aggregated trend per
+macrozone, drought index, and timescale. The macrozones that reached the
+lowest trend for SPI, SPEI, and SSI are ``Norte Chico'' and ``Centro'',
+where the indices also decrease at longer time scales. This may
+potentially be explained by the prolonged reduction in precipitation
 that has affected the hydrological system in Chile. At 36 months, it
 reaches trends between -0.03 and -0.04 (z-score) per decade for SPI,
-SPEI, and SSI. For ``Sur,'' the behavior is similar, decreasing at
+SPEI, and SSI. For ``Sur'', the behavior is similar, decreasing at
 longer scales and having between -0.016 and -0.025 per decade for SPI,
@@
 EDDI (0.042 per decade), and ``Centro'' has the lowest for SPI and SPEI.
-In ``Norte Grande'' and ``Norte Chico,'' which are in a semi-arid
+In ``Norte Grande'' and ``Norte Chico'', which are in a semi-arid
 climate, it is evident that the EDDI has an effect on the difference
@@
 
-\hypertarget{interaction-of-land-cover-and-drought-1}{%
 \subsection{Interaction of land cover and
-drought}\label{interaction-of-land-cover-and-drought-1}}
-
-\hypertarget{land-cover-change-1}{%
-\subsubsection{Land cover change}\label{land-cover-change-1}}
+drought}\label{interaction-of-land-cover-and-drought-1}
+
+\subsubsection{Land cover change}\label{land-cover-change-1}
 
@@
 
-{\centering \includegraphics{../output/figs/proportion_landcover_macroclass_2001-2022.png}
+\centering{
+
+\includegraphics{../output/figs/proportion_landcover_macroclass_2001-2022.png}
 
@@
 
-\end{figure}
+\end{figure}%
 
@@
 largest type of vegetation, with 137,085 \(km^2\), is forest. Grassland
-has 74,247 \(km^2\), savanna 55,206 \(km^2\), shrubland 25,341 \(km^2\),
-and cropland 3,146 \(km^2\) (Table \ref{tab-landcoverSurf}). The
-macrozones with major changes for 2001--2022 were ``Centro,'' ``Sur,''
-and ``Austral,'' with 36\%, 31\%, and 34\% of their surface changing the
-type of land cover, respectively (Figure~\ref{fig-studyArea} and Table
-\ref{tab-landcoverTrend}). Figure~\ref{fig-LCprop} shows the summary of
-the proportion of surface per land cover class and macrozone, derived
-from the persistence mask over continental Chile.
+has 74,247 \(km^2\) , savanna 55,206 \(km^2\), shrubland 25,341
+\(km^2\), and cropland 3,146 \(km^2\) (Table \ref{tab-landcoverSurf}).
+The macrozones with major changes for 2001--2022 were ``Centro'',
+``Sur'', and ``Austral'', with 36\%, 31\%, and 34\% of their surface
+changing the type of land cover, respectively
+(Figure~\ref{fig-studyArea} and Table \ref{tab-landcoverTrend}).
+Figure~\ref{fig-LCprop} shows the variation for 2001--2022 in the
+proportion of surface per land cover class and macrozone, derived from
+the persistence mask over continental Chile.
 
@@
 111 \(km^2 yr^{-1}\) and a reduction in the class savanna of 70
-\(km^2 yr^{-1}\). In the ``Centro'' and ``Sur,'' there are changes with
+\(km^2 yr^{-1}\). In the ``Centro'' and ``Sur'', there are changes with
 an important reduction in savanna with 136 and 319 \(km^2 yr^{-1}\),
@@
 change for more dense vegetation types. The area under cultivation
-(croplands) appears to be shifting from the ``Centro'' to the ``Sur.''
+(croplands) appears to be shifting from the ``Centro'' to the ``Sur''.
 Also, there is a high increase in forest (397 \(km^2 yr^{-1}\)) in the
-``Sur,'' seemingly replacing the savanna lost (Table
+``Sur'', seemingly replacing the savanna lost (Table
 \ref{tab-landcoverTrend}).
 
-\hypertarget{relationship-between-drought-indices-and-land-cover-change}{%
 \subsubsection{Relationship between drought indices and land cover
-change}\label{relationship-between-drought-indices-and-land-cover-change}}
+change}\label{relationship-between-drought-indices-and-land-cover-change}
+
+\begin{table}[!ht]
+\caption{The five most important trends of drought indices in estimating the landcover trend per land cover type and the $R^2$ reached by each random forest model.
+The white cells indicate that the landcover class is not significant in terms of surface area.}
+\label{tab-landcoverTrendRF}
+\includegraphics[width=0.5\textwidth]{../output/figs/table_r2_most_important_variables.png}
+\end{table}
 
@@
 
-{\centering \includegraphics{../output/figs/bars_relative_importance_RF.png}
-
-}
-
-\caption{\label{fig-varImportance}cdcdcdc}
-
-\end{figure}
-
-\begin{table}[!ht]
-\caption{The five most important trends of drought indices in estimating the landcover trend per land cover type and the r-squared (rsq) reached by each random forest model.}
-\label{tab-landcoverTrendRF}
-\includegraphics[]{../output/figs/table_r2_most_important_variables.png}
-\end{table}
-
-The random forest models for estimating the land cover trend based on
-the trends in drought indices reach an \(R^2\) between 0.32 and 0.39 for
-the land cover types (Table \ref{tab-landcoverTrendRF}) excluding
-cropland. It is more likely that short- and medium-term increases in AED
-(EDDI-6 and EDDI-12) and short-term precipitation deficits (SPI-6 and
-SPEI-6) are associated to changes in grassland and bare land. The
-short-term increase of AED (EDDI-3 and EDDI-6) and the longer duration
-of the precipitation deficit (SPI-24, SPI-36, and SPEI-36) were the most
-important variables that correlated with changes in shrubland. The
-changes in savanna are associated with a short- and long-term increase
-in AED and a three-year precipitation deficit (SPI-36). The increase in
-cumulative AED from 12 to 36 months is the strongest associated variable
-that contributes to changes in forests, followed by the reduction of
-soil moisture over six and 36 months. The supplementary material in
-Section S3 provides further details about the variable's importance.
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/bars_relative_importance_RF.png}
+
+}
+
+\caption{\label{fig-varImportance}Relative importance of drought indices
+for explaining the trend in landcover change across the five macrozones
+in Chile. SPI, Standardized Precipitation Index; EDDI, Evaporative
+Demand Drought Index; SPEI, Standardized Precipitation
+Evapotranspiration Index; SSI, Standardized Soil Moisture Index. The
+numbers beside the drough index correspond to the time scales.}
+
+\end{figure}%
+
+Table \ref{tab-landcoverTrendRF} shows the drought indices that are the
+most important variables in the random forest models, together with the
+\(R^2\) reached. The random forest models reached an \(R^2\) between
+0.12 and 0.41 for the land cover types and macrozones. The model shows
+the highest \(R^2\) for shrublands (0.28 to 0.42) and the lowest \(R^2\)
+for croplands (0.16 to 0.20) across all macrozones.
+
+Figure~\ref{fig-varImportance} shows the three most important variables
+for the models across the five macrozones and per landcover type. For
+shurublands, the SPI of long-term and short-term SSI were the most
+relevant drought indices within the five macrozones. We showed that the
+trend in short-term EDDI (1--6 months) and long-term SPI (24- and
+36-months) affected grasslands and forest changes. We showed that trends
+in SPI-36 and long-term trends in EDDI (12 to 36 months) were associated
+with changes in savannas. The changes in barren land are shown to be
+limited to the changes in the short-term AED (1 to 6 months). Changes in
+croplands have not been linked to drought across all macrozones. The
+supplementary material in Section S3 provides further details about the
+variable's importance.
 
@@
 
-{\centering \includegraphics{../output/figs/points_landcover_drought_indices_trend_and_time_scale.png}
+\centering{
+
+\includegraphics{../output/figs/points_landcover_drought_indices_trend_and_time_scale.png}
 
@@
 land cover change (y-axis) and the trend in drought indices (x-axis) for
-the five macrozones. Vertical panels correspond to 1, 3, 6, 12, 24, and
-36 months of the time scale by drought index. Horizontal panels show
-each drought index}
-
-\end{figure}
+the five macrozones. Vertical panels correspond to short (6 months) and
+long (36 months) time scales. Horizontal panels show the drought indices
+SPI, EDDI, and SSI.}
+
+\end{figure}%
 
 Figure~\ref{fig-TrendsLandDrought} shows the connection between the SPI,
-EDDI, and SSI drought indices and changes in land cover. Forest in the
-``Sur,'' shrubland and grassland in ``Centro,'' barren land in ``Norte
-Chico,'' and savanna in ``Austral'' showed an increase in land cover
-extent, which was associated with an increase in EDDI. Savanna in
-``Centro,'' ``Sur,'' and ``Norte Chico'' decreases with the increase in
-EDDI. The SPI and SSI showed similar behavior regarding the trend in
-land cover type. A decrease in SPI and SSI is associated with an
-increase in the surface in shrubland and grassland in ``Centro,'' forest
-in ``Sur,'' and barren land in ``Norte Chico,'' as well as a decrease
-trend in savanna in ``Norte Chico,'' ``Centro,'' and ``Sur.''
-
-\hypertarget{drought-impacts-on-vegetation-productivity-within-land-cover}{%
+EDDI, and SSI drought indices for time scales of 6 and 36 months against
+changes in land cover. Forest in the ``Sur'', shrubland and grassland in
+``Centro'', barren land in ``Norte Chico'', and savanna in ``Austral''
+showed an increase in land cover extent, which was associated with an
+increase in EDDI. Savanna in ``Centro'', ``Sur'', and ``Norte Chico''
+decreases with the increase in EDDI. The SPI and SSI showed similar
+behavior regarding the trend in land cover type. A decrease in SPI and
+SSI is associated with an increase in the surface in shrubland and
+grassland in ``Centro'', forest in ``Sur'', and barren land in ``Norte
+Chico'', as well as a decrease trend in savanna in ``Norte Chico'',
+``Centro'', and ``Sur''.
+
 \subsection{Drought impacts on vegetation productivity within land
-cover}\label{drought-impacts-on-vegetation-productivity-within-land-cover}}
-
-\hypertarget{trends-in-vegetation-productivity}{%
+cover}\label{drought-impacts-on-vegetation-productivity-within-land-cover}
+
 \subsubsection{Trends in vegetation
-productivity}\label{trends-in-vegetation-productivity}}
+productivity}\label{trends-in-vegetation-productivity}
 
@@
 
-{\centering \includegraphics{../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png}
+\centering{
+
+\includegraphics{../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png}
 
@@
 
-\end{figure}
+\end{figure}%
+
+Figure~\ref{fig-zcNDVI_var} shows a spatial map of trends in zcNDVI
+(Figure~\ref{fig-zcNDVI_var}a). In ``Norte Grande'', vegetation
+productivity, as per the z-index, exhibits a yearly increase of 0.027
+for grassland and 0.032 for shrubland. In ``Norte Chico'', savanna has
+the lowest trend slope of -0.062, cropland -0.047, shrubland -0.042, and
+grassland -0.037. In ``Centro'', shrubland reaches -0.07, savanna
+-0.031, cropland -0.024, forest -0.017, and grassland -0.005 per decade.
+This decrease in productivity could be associated either with a
+reduction in vegetation surface, a decrease in biomass, or browning.
 
@@
 
-In Figure~\ref{fig-zcNDVI_var} it is showed the spatial map of trends in
-zcNDVI (Figure~\ref{fig-zcNDVI_var}a). In ``Norte Grande,'' vegetation
-productivity, as per the z-index, exhibits a yearly increase of 0.027
-for grassland and 0.032 for shrubland. In the ``Norte Chico,'' savanna
-has the lowest trend of -0.062, cropland -0.047, shrubland -0.042, and
-grassland -0.037. In ``Centro,'' shrubland reaches -0.07, savanna
--0.031, cropland -0.024, forest -0.017, and grassland -0.005 per decade.
-This decrease in productivity could be associated either with a
-reduction in vegetation surface, a decrease in biomass, or browning.
-
-\hypertarget{correlation-between-vegetation-productivity-and-drought-indices}{%
 \subsubsection{Correlation between vegetation productivity and drought
-indices}\label{correlation-between-vegetation-productivity-and-drought-indices}}
+indices}\label{correlation-between-vegetation-productivity-and-drought-indices}
 
@@
 
-{\centering \includegraphics{../output/figs/mapa_cor_selec_indices_zcNDVI6.png}
+\centering{
+
+\includegraphics{../output/figs/mapa_cor_selec_indices_zcNDVI6.png}
 
@@
 
-{\centering \includegraphics{../output/figs/mapa_cor_r_indices_zcNDVI6.png}
+\centering{
+
+\includegraphics{../output/figs/mapa_cor_r_indices_zcNDVI6.png}
 
@@
 
-\end{figure}
-
-Figure~\ref{fig-corTimeScale} shows the highest coefficient of
-determination (r-squared, or rsq) found in the regression analysis
-between zcNDVI and different drought indicators over time scales of 1,
-3, 6, 12, 24, and 36 months. The spatial variation of time scales
-reached per index is mostly for time scales above 12 months. In the case
-of SSI, the predominant scales are 6 and 12 months. For all indices, to
-the north, the time scales are higher and diminish toward the south
-until the south part of ``Austral,'' where they increase. In
-Figure~\ref{fig-corPerson}, the map of Pearson correlation values (r) is
-shown. The EDDI reached correlations above 0.5 between ``Norte Chico''
-and ``Sur.'' The correlation changes from negative to positive toward
-the Andes Mountains and to the sea, just as in the northern part of
-``Austral.'' The SPI and SPEI have similar results, with the higher
-values in ``Norte Chico'' and ``Centro'' being higher than 0.6.
-Following a similar spatial pattern as EDDI but with an opposite sign.
-The SSI showed to be the index that has a major spatial extension with a
-higher correlation. It has a similar correlation to SPI and SPEI for
-``Norte Chico'' and ``Sur,'' but has improvements for ``Sur.''
+\end{figure}%
+
+Figure~\ref{fig-corTimeScale} shows the time scales that reached the
+highest r-squared in the regression analysis between zcNDVI and
+different drought indicators over time scales of 1, 3, 6, 12, 24, and 36
+months. The spatial variation of time scales reached per index is mostly
+for time scales above 12 months. In the case of SSI, the predominant
+scales are 6 and 12 months. For all indices, to the north, the time
+scales are higher and diminish toward the south until the south part of
+``Austral'', where they increase. In Figure~\ref{fig-corPerson}, the map
+of Pearson correlation values (r) is shown. The EDDI reached
+correlations above 0.5 between ``Norte Chico'' and ``Sur''. The
+correlation changes from negative to positive toward the Andes Mountains
+and to the sea, just as in the northern part of ``Austral''. The SPI and
+SPEI have similar results, with the higher values in ``Norte Chico'' and
+``Centro'' being higher than 0.6. Following a similar spatial pattern as
+EDDI but with an opposite sign. The SSI showed to be the index that has
+a major spatial extension with a higher correlation. It has a similar
+correlation to SPI and SPEI for ``Norte Chico'' and ``Centro'', but for
+``Sur'' the correlation is higher with SSI.
 
@@
 Figure~\ref{fig-corTimeScale} and Figure~\ref{fig-corPerson}. According
-to what is shown, forests seem to be the most resistant to drought.
-Showing that only ``Centro'' is slightly (rsq = 0.25) impacted by a
+to what is shown, forests is likely to be the most resistant to drought.
+Showing that only ``Centro'' is slightly (\(R^2\) = 0.25) impacted by a
 12-month soil moisture deficit (SSI-12). In the ``Norte Chico'' and to a
-lesser extent in the ``Norte Grande,'' it is evident that a SSI-12 with
-a rsq = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with rsq
-= 0.28 and 0.34, respectively) have an impact on grasslands. However,
-this type was unaffected by soil moisture, water supply, or demand in
-macrozones further south. The types that show to be most affected by
-variation in climate conditions are shrublands, savannas, and croplands.
-For savannas in ``Norte Chico,'' the SSI-12 and SPI-24 reached an rsq of
-0.74 and 0.58, respectively. This value decreases to the south, but the
-SSI-12 is still the variable explaining more of the variation in
-vegetation productivity (rsq = 0.45 in ``Centro'' and 0.2 in ``Sur'').
-In the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain
-between 45\% and 66\% of the variability in ``Norte Chico.'' The type of
-land most impacted by climatic variation was shrubland, where soil
-moisture explained 59\% and precipitation, 37\%, in ``Norte Chico'' and
-``Centro,'' with SSI-12 being the most relevant variable, then SPI-36 in
-``Norte Chico'' and SPI-24 in ``Sur.''
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\hypertarget{vegetation-water-demand-and-its-relation-to-drought}{%
+lesser extent in the ``Norte Grande'', it is evident that a SSI-12 with
+a \(R^2\) = 0.45 and a decrease in water supply (SPI-36 and SPEI-24 with
+\(R^2\) = 0.28 and 0.34, respectively) have an impact on grasslands.
+However, this type was unaffected by soil moisture, water supply, or
+demand in macrozones further south. The types that show to be most
+affected by variation in climate conditions are shrublands, savannas,
+and croplands. For savannas in ``Norte Chico'', the SSI-12 and SPI-24
+reached an
+\(R^2" of 0.74 and 0.58, respectively. This value decreases to the south, but the SSI-12 is still the variable explaining more of the variation in vegetation productivity (\)R\^{}2\$
+= 0.45 in ``Centro'' and 0.2 in ``Sur''). In the case of croplands, the
+SPEI-12, SPI-36, and SSI-12 explain between 45\% and 66\% of the
+variability in ``Norte Chico''. The type of land most impacted by
+climatic variation was shrubland, where soil moisture explained 59\% and
+precipitation, 37\%, in ``Norte Chico'' and ``Centro'', with SSI-12
+being the most relevant variable, then SPI-36 in ``Norte Chico'' and
+SPI-24 in ``Sur''.
+
+\section{Discussion}\label{discussion}
+
 \subsection{Vegetation water demand and its relation to
-drought}\label{vegetation-water-demand-and-its-relation-to-drought}}
-
-In our study, we took into account the variation in vegetation
-productivity in Chile, specifically in areas without any changes in land
-cover, to prevent any misleading conclusions about the increase in water
-demand due to land cover change. Our results show a contrasting
-perspective regarding the evidence provided by
-\citet{Vicente-Serrano2022} on a global scale, who indicates that the
-increase in drought is led by an increase in agricultural land, which in
-turn increases water demand.
+drought}\label{vegetation-water-demand-and-its-relation-to-drought}
+
+In our study, we considered the variation in vegetation productivity in
+Chile, specifically in areas without any changes in land cover, to
+prevent any misleading conclusions about the increase in water demand
+due to land cover change. Our results show a contrasting perspective
+regarding the evidence provided by \citet{Vicente-Serrano2022} on a
+global scale, who indicates that the increase in drought is led by an
+increase in agricultural land, which in turn increases water demand.
 
@@
 with the cultivated land. Evidence suggests that higher-water-demanding
-crops have replaced croplands in the Petorca basin (central Chile),
-leading to an increase in water abstraction
+crops have replaced less demanding crops in the Petorca basin (central
+Chile), leading to an increase in water abstraction
 \citep{Munoz2020, Duran2020}. Nonetheless, at this scale of analysis,
@@
 The long-scale trends (e.g., 36 months) demonstrate the impact of
-human-induced climate change on water availability in Chile, potentially
-due to an intense hydrological drought stemming from the ongoing
-precipitation deficit and rising AED. But it is likely that in zones
-most affected by drought, the main cause is not an increase in
-vegetation water demand due to an intensification of cultivated land
-(e.g., an increase in irrigated crops) like in other parts of the globe
+climate change on water availability in Chile, potentially due to an
+intense hydrological drought stemming from the ongoing precipitation
+deficit and rising AED. But it is likely that in zones most affected by
+drought, the main cause is not an increase in vegetation water demand
+due to an intensification of cultivated land (e.g., an increase in
+irrigated crops) like in other parts of the globe
 \citep{Vicente-Serrano2020}. North-central Chile has experienced a
@@
 
-\hypertarget{sensitivity-of-land-cover-vegetation-to-short--and-long-term-drought}{%
 \subsection{Sensitivity of land cover vegetation to short- and long-term
-drought}\label{sensitivity-of-land-cover-vegetation-to-short--and-long-term-drought}}
+drought}\label{sensitivity-of-land-cover-vegetation-to-short--and-long-term-drought}
 
@@
 the type most resistant to drought, and shrublands, savannas, and
-croplands have higher sensitivity.
+croplands have higher sensitivities.
 
@@
 case, the SPEI-12 was the one that most impacted the croplands in
-``Norte Chico'' and ``Centro.'' In general, most studies show that
+``Norte Chico'' and ``Centro''. In general, most studies show that
 croplands are most sensitive to short-term drought (\textless{} SPI-6)
@@
 precipitation deficits have an impact on soil water, so less water is
-available for plant growth. However, we found that in ``Norte Chico,''
+available for plant growth. However, we found that in ``Norte Chico'',
 an SPI-36 and SPEI-12 had a higher impact, which are associated with
-long-term water deficit, and in ``Centro,'' an SPI-12 and SPEI-12. Thus,
+long-term water deficit, and in ``Centro'', an SPI-12 and SPEI-12. Thus,
 we hypothesize that this impact could be attributed to the hydrological
@@
 which in turn is impacted by long-term deficits, and consequently, the
-vegetation is more dependent on groundwater. In ``Sur'' and ``Austral,''
+vegetation is more dependent on groundwater. In ``Sur'' and ``Austral'',
 the correlations between drought indices and vegetation productivity
@@
 possible reason for this is that the most resistant types, forest and
-grassland, predominate south of ``Centro.'' Also, drought episodes have
+grassland, predominate south of ``Centro''. Also, drought episodes have
 been less frequent and intense and have had a lower impact on water
@@
 
-According to \citet{Senf2020}, severe drought conditions in Europe are a
-significant cause of tree mortality. However, we discovered that forests
-are the most resilient land cover class to drought, with less variation
-in drought indices. Supporting this is \citet{Fathi-Taperasht2022}, who
-assert that Indian forests are the most drought-resistant and recover
-rapidly. Similarly, the work of \citet{Wu2024}, who analyzed vegetation
-loss and recovery in response to meteorological drought in the humid
-subtropical Pearl River basin in China, indicates that forests showed
-higher drought resistance. Using Vegetation Optical Depth (VOD), kNDVI,
-and EVI, \citet{Xiao2023} tested the resistance of ecosystems and found
-that ecosystems with more forests are better able to handle severe
-droughts than croplands. They attribute the difference to a deeper
-rooting depth for trees, a higher water storage capacity, and different
-water use strategies between forest and cropland \citep{Xiao2023}.
-
 In central Chile, \citet{Venegas2022} observed a significant decline in
 the overall growth of sclerophyllous moist forests (mediterranean
-forests), which they attributed to increased drought conditions. In
-another study, \citet{Fuentes2021} evaluated water scarcity and land
+forests), which they attributed to increased drought conditions.
+However, we found that forests are the most resilient land cover class
+to drought, with less variation in drought indices. In the ``Sur'',
+there is a large domain of planted forests that have replaced native
+vegetation since the 1970s
+\citep{Heilmayr2016, Heilmayr2020, Miranda2017}, impacting biodiversity
+and ecosystem services \citep{Rodriguez2018}. It has recently been shown
+that these planted forests are responding positively to climate change,
+and it is expected that they will benefit from future climate scenarios
+\citep{Carrasco2022}. Further, the forests of the ``Austral'' region
+correspond to Patagonian ecosystems, mainly native forests dominated by
+tree species of wide niche breadth . Overall, these forests have been
+more affected by the increase in temperature than by the reduction in
+moisture \citep{Fajardo2023, Holz2018}. These responses have caused
+stabilized tree growth, linked to more frequent warm autumns
+\citep{Gibson2022}. It has also been observed that these forests have
+shown resistance to drought episodes \citep{Fajardo2023}, which might be
+attributable to their relatively low growth rate. Supporting this is
+\citet{Fathi-Taperasht2022}, who assert that Indian forests are the most
+drought-resistant and recover rapidly. Similarly, the work of
+\citet{Wu2024}, who analyzed vegetation loss and recovery in response to
+meteorological drought in the humid subtropical Pearl River basin in
+China, indicates that forests showed higher drought resistance. Using
+Vegetation Optical Depth (VOD), kNDVI, and EVI, \citet{Xiao2023} tested
+the resistance of ecosystems and found that ecosystems with more forests
+are better able to handle severe droughts than croplands. They attribute
+the difference to a deeper rooting depth for trees, a higher water
+storage capacity, and different water use strategies between forest and
+cropland \citep{Xiao2023}.
+
+In another study, \citet{Fuentes2021} evaluated water scarcity and land
 cover change in Chile between 29° and 39° south latitude. They used the
@@
 
-\hypertarget{vegetation-productivity-and-drought}{%
 \subsection{Vegetation productivity and
-drought}\label{vegetation-productivity-and-drought}}
+drought}\label{vegetation-productivity-and-drought}
 
@@
 irrigated rather than rainfed croplands, soil moisture has better skills
-than SPI and SPEI for estimating gross primary production (GPP) . Also,
+than SPI and SPEI for estimating gross primary production (GPP). Also,
 \citet{Zhou2021} indicate that the monthly scaled Standardized Water
@@
 they seem to react in a shorter time, with six months (SSI-6) in
-``Centro.'' This variation for croplands could be related to the fact
-that in ``Norte Chico,'' the majority of crops are irrigated, but to the
+``Centro''. This variation for croplands could be related to the fact
+that in ``Norte Chico'', the majority of crops are irrigated, but to the
 south there is a higher proportion of rainfed agriculture, which is most
 dependent on the short-term availability of water. Rather, in ``Norte
-Chico,'' the orchards are more dependent on irrigation, which in turn
-depends on the availability of storage water in dams of groundwater
+Chico'', the orchards are more dependent on irrigation, which in turn
+depends on the availability of storage water in dams or groundwater
 reservoirs, which are affected by long-term drought (e.g., SPI-36).
 
-\hypertarget{drought-information-to-aid-in-adaptation}{%
 \subsection{Drought information to aid in
-adaptation}\label{drought-information-to-aid-in-adaptation}}
+adaptation}\label{drought-information-to-aid-in-adaptation}
 
@@
 Drought Observatory for Agriculture and Biodiversity of Chile (ODES)
-(\url{https://odes-chile.org/app/unidades})
+\url{https://odes-chile.org/app/unidades}
 \citep{Zambrano2023b, Kunst2023}. This initiative ensures the
-availability of extensive climate data, facilitating the development of
-adaptive strategies that are both responsive to the realities of
-different regions and grounded in the latest scientific understanding.
-The proactive sharing and updating of such data underscores its key role
-in enabling policymakers to craft adaptive measures that are finely
-tuned to the diverse and evolving landscapes of drought impacts.
-Furthermore, the recently promulgated law about climate change in Chile
-(Law 21.455, https://www.bcn.cl/leychile/navegar?idNorma=1177286), which
-aims to implement sectoral adaptation plans for agriculture, forests,
-and biodiversity, could benefit from this information.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+availability and easy accessibility of extensive climate data,
+facilitating the development of adaptive strategies that are both
+responsive to the realities of different regions and grounded in the
+latest scientific understanding. The proactive sharing and updating of
+such data underscores its key role in enabling policymakers to craft
+adaptive measures that are finely tuned to the diverse and evolving
+landscapes of drought impacts. Furthermore, the recently promulgated law
+about climate change in Chile (Law 21.455,
+\url{https://www.bcn.cl/leychile/navegar?idNorma=1177286}), which aims
+to implement sectoral adaptation plans for agriculture, forests, and
+biodiversity, could benefit from this information.
+
+\section{Conclusion}\label{conclusion}
 
@@
 
-We model about 20--30\% of the trends in land cover types, such as
-forest, grassland, shrubland, and savanna, based on drought indicators
-across Chile. There is no evidence that drought alters cropland surface
-area. The increase in AED is the most important variable explaining the
-variability in the change in land cover, followed by a reduction in
-precipitation and soil moisture.
+Using Random Forest and variable importance metrics, we assessed how
+land cover trends relate to drought indices and found that drought
+indices across Chile could explain about 12--41\% of the trends in land
+cover types. Drought explains a higher variance in changes in
+shrublands, followed by savanna, grasslands, and forest. The changes in
+croplands showed the lowest association with drought. We found that the
+short- to long-term trend in AED was the most important variable that
+could partially explain the observed land cover trend, followed by
+long-term trends in precipitation and short-term trends in soil
+moisture.
 
 The trends in drought indices are accompanied by multiple land cover
-changes in the country, most notably an increase of forest in ``Sur,''
-of shrubland and grassland in ``Centro,'' and of savanna in ``Centro''
-and ``Sur.'' In ``Norte Chico'' and ``Centro,'' the croplands have been
-declining in surface, whereas in ``Sur,'' there is an increase in
+changes in the country, most notably an increase of forest in ``Sur'',
+of shrubland and grassland in ``Centro'', and of savanna in ``Centro''
+and ``Sur''. In ``Norte Chico'' and ``Centro'', the croplands have been
+declining in surface, whereas in ``Sur'', there is an increase in
 cultivated land.
@@
 by anomalies in accumulated precipitation over one to two years. The
-variation in AED seems to intensify the drought impact on vegetation
-productivity.
-
-The results of this study provide insightful information that would
-assist in developing adaptation measures for Chilean ecosystems to cope
-with climate change and drought. Information that could be used in the
-scope of the national law on climate change, which seeks to implement
+variation in AED is likely to exacerbate the drought's impact on
+vegetation productivity.
+
+The results of this study provide insightful information that can assist
+in developing adaptation measures for Chilean ecosystems to cope with
+climate change and drought. This information could be used in the scope
+of the national law on climate change, which seeks to implement
 adaptation strategies for agriculture, forests, and biodiversity.
 
-\hypertarget{acknowledgment}{%
-\section{Acknowledgment}\label{acknowledgment}}
+\section{Acknowledgment}\label{acknowledgment}
 

### 2024-05-09 scripts/33_analysis_trend_drought_indices_and_landcover_cuencas.R

@@ -149,15 +149,20 @@ tabla <- map_df(1:6,\(i){
 library(broom.helpers)
 
 paleta <- read.csv('data/processed_data/paleta_colores_landcover.csv') |> 
-  dplyr::filter(Name %in% c('Forest','Shrubland','Savanna','Grassland','Cropland','Barren land'))
+  dplyr::filter(Name %in% c('Forest','Shrubland','Savanna','Grassland','Cropland','Barren land')) |> 
+  mutate(Name = case_when(Name == 'Barren land' ~ "Barren Land",
+                          .default = Name))
 
 colors <-  rgb(paleta$R,paleta$G,paleta$B,maxColorValue = 255)
 #attr(colors,'names') <- paleta$class
-paleta$Name[6] <- 'Barren_land'
+#paleta$Name[6] <- 'Barren_land'
 attr(colors,'names') <- paleta$Name
 
 tabla |>
-  mutate(type = factor(type,levels=paleta$Name),
+  mutate(
+    type = case_when(type == 'Barren_land' ~ 'Barren Land',
+                     .default = type),
+    type = factor(type,levels=paleta$Name),
          macrozona = factor(macrozona,levels=c("norte grande", "norte chico",
                                                'zona central','zona sur','zona austral'),
                             labels = c('Norte Grande','Norte Chico','Centro','Sur','Austral')),
@@ -168,9 +173,17 @@ tabla |>
                           macrozona == "Austral" & type == "Cropland" ~ NA,
                           .default = Mean)
   ) |> 
+  mutate(Variable = 
+           factor(Variable,
+             levels = rev(c(
+                 paste0('SPI-',c(12,24,36)),
+                 paste0('EDDI-',c(1,3,6,12,24,36)),
+                 paste0('SPEI-',c(3,24,36)),
+                 paste0('SSI-',c(1,3,36))
+                 )))) |>
   group_by(type,macrozona) |> 
   mutate(rel_imp = Mean*10e2) |> 
-  slice_max(rel_imp, n= 1) |> 
+  slice_max(rel_imp, n= 3) |> 
   ggplot(aes(Variable,rel_imp)) +
   geom_col(aes(fill = type),position = 'dodge') +
   scale_y_continuous(expand = c(0,0)) +
@@ -178,16 +191,17 @@ tabla |>
                     values=colors) +
[…]

### 2024-05-10 article/supplementary.qmd
update supplementary

@@
 title: "Supplementary Material"
-author: "Francisco Zambrano et al."
+subtitle: "The effects of drought on land cover change and vegetation productivity in continental Chile"
+author:
+  - name: Francisco Zambrano
+    email: francisco.zambrano@umayor.cl
+    affiliations: 
+        - id: umayor
+          name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad, Universidad Mayor
+          city: Santiago
+          country: 7500994, Chile.
+        - id: odes
+          name: Observatorio de Sequía para la Agricultura y la Biodiversidad de Chile (ODES), Universidad Mayor
+          city: Santiago
+          country: 7500994, Chile.
+    attributes:
+        corresponding: true
+    #note: This is the first author footnote.
+  - name: Anton Vrieling
+    #email: a.vrieling@utwente.nl
+    affiliations:
+        - id: itc
+          name: Faculty of Geo-Information Science and Earth, University of Twente
+          city:  Enschede
+          country: The Netherlands.
+  - name: Francisco Meza
+    affiliations:
+        - id: agrouc
+          name: Facultad de Agronomía y Sistemas Naturales, Pontificia Universidad Católica de Chile. 
+          city: Santiago
+          country: Chile.
+        - id: instuc
+          name: Instituto para el Desarrollo Sustentable. Pontificia Universidad Católica de Chile
+          city: Santiago
+          country: Chile.
+        - id: cambioglobal
+          name: Centro Interdisciplinario de Cambio Global, Pontificia Universidad Católica de Chile
+          city: Santiago
+          country: Chile.
+  - name: Iongel Duran-Llacer
+    affiliations:
+        - id: umayor-io
+          name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Universidad Mayor,
+          city: Santiago
+          country: 7500994, Chile.
+  - name: Francisco Fernández
+    #email: francisco.fernandez@uss.cl
+    affiliations:
+        - id: uss
+          name: Center of Economics for Sustainable Development (CEDES), Faculty of Economics and Government, Universidad San Sebastian
+          city: Santiago
+          country: Chile.
+        - id: capes
+          name: Center of Applied Ecology and Sustainability (CAPES)
+          city: Santiago
+          country: Chile.
+  - name: Alejandro Venegas-González
+    affiliations:
+        - id: ica3
+          name: Instituto de Ciencias Agroalimentarias, Animales y Ambientales (ICA3), Universidad de O'Higgins
+          city: San Fernando
+          country: Chile.
+  - name: Nicolas Raab
+    affiliations:
+        - id: agrouc
+  - name: Dylan Craven
+    affiliations:
+        - id: gema
+          department: GEMA Center for Genomics, Ecology & Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750
+          city: Santiago
+          country: Chile.
+        - id: do
+          name: Data Observatory Foundation
+          city: Santiago
+          country: Chile.
 date: "`r Sys.Date()`"
+include-before-body:
+  - text: |
+      \renewcommand\thesection{S\arabic{section}}
+      \renewcommand\thefigure{S\arabic{figure}}
+      \renewcommand\thetable{S\arabic{table}}
 bibliography: references.bib
 format:
-  pdf:
-    number-sections: true
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      name: Journal Name
+      formatting: preprint
+      layout: onecolumn
+      model: 3p
+      cite-style: authoryear
+      natbiboptions: numbers
+    include-in-header: 
+      text: \usepackage{lineno}\linenumbers
+            \usepackage{multirow}
+            \usepackage{lscape}
+            \newcommand{\blandscape}{\begin{landscape}}
+            \newcommand{\elandscape}{\end{landscape}}
 editor: source
 ---
+
+```{r setup}
+#| include: false
+library(tidyverse)
+library(gt)
+```
 
@@
 We used 277 stations located throughout Chile.
-For in-situ soil moisture, we used a private soil network that is owned by the agricultural enterprise Garces Fruit, which has 99 stations in Central Chile, located in fields with cherry fruit crops.
-The sensors are installed at 30, 60, and 90cm and are a Advanced Soil Moisture Sensor model Teros 12 from METER Group ([https://metergroup.com/products/teros-12/](https://metergroup.com/products/teros-12/)).
-To avoid comaring ERA5L with in-situ soil moisture levels caused by irrigation, which are not captured by ERA5L, we used daily data for the year 2022 and the months outside the growing season, May to September.
+For _in-situ_ soil moisture, we used a private soil network that is owned by the agricultural enterprise Garces Fruit (https://garcesfruit.com/), which has 99 stations in Central Chile, located in fields with cherry fruit crops.
+The sensors are installed at 30, 60, and 90cm and are a Advanced Soil Moisture Sensor model Teros 12 from the METER Group ([https://metergroup.com/products/teros-12/](https://metergroup.com/products/teros-12/)).
+To avoid comparing ERA5L with _in-situ_ soil moisture levels caused by irrigation, which are not captured by ERA5L, we used daily data for the year 2022 and the months outside the growing season, May to September.
 
@@
 
-The average performance metrics of ERA5L over the 266 weather stations were in the case of monthly temperature: $ubRMSE=1.06\,°C$, $MAE=1.131\,°C$, and $CC=0.963$, showing a good agreement, low error, and low overestimation.
+The average performance metrics of ERA5L over the 266 weather stations is shown in @table_metrics_era5
+
+<!-- were in the case of monthly temperature: $ubRMSE=1.06\,°C$, $MAE=1.131\,°C$, and $CC=0.963$, showing a good agreement, low error, and low overestimation.
 For cumulative monthly precipitation, $MAE=28.1\,mm$, $bias=1.93$, and $CC=0.845$, showing a high correlation and a 93% bias and being overestimated by ERA5L.
+-->
+
 In the case of the 97 soil moisture stations, we averaged for the three depths (30, 60, and 90cm) and then compared it with the volumetric water content in the top 100cm of the soil derived from ERA5L.
-For this case, we made a daily comparison to determine the performance metrics per station, then we averaged over all stations, having a $CC=0.71$, $RMSE=0.174$, $MAE=0.167$, and $bias=1.74$.
-The ERA5 soil moisture overestimate is 74%, but there is some good correlation.
+For this case, we made a daily comparison to determine the performance metrics per station, then we averaged the measured soil moisture at the three levels and over all stations.
+
+```{r table_metrics_era5}
+#| echo: false
+#| tbl-cap: "Metrics of performance for the climatic and soil moisture variables from ERA5-Land with data from 266 stations in Central Chile."
+tribble(~Variable,~ubRMSE,~MAE,~Bias,~CC,
+        "Temperature","1.06°C","1.131°C","","0.96",
+        "Precipitation","","28.1mm","1.93","0.845",
+        "Soil Moisture (1m)","0.174 cc/cc","0.167 cc/cc","1.74", "0.71") |> 
+  gt()
+```
 
@@
 
-To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 Collection 6.1 from MODIS.
+To analyze the LULCC, we used the IGBP scheme from the MCD12Q1 Collection 6.1 from MODIS.
 This product has a yearly frequency from 2001 to 2022.
@@
 
-To validate the land cover obtained, we compare the macroclasses with those of a more detailed land cover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile).
+To assess the accuracy of the regrouped MCD12Q1 land cover maps, we compared the macroclasses with those of a more detailed land cover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile).
 The latter has a spatial resolution of 30 m and three hierarchy levels of defined classes; from those, we used level 1, which fits with the macroclass land cover.
-We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from the land cover macrolcasses to compare with LCChile.
+We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from the land cover macroclasses to compare with LCChile.
 For this comparison, we used the following procedure:
@@
 ii) we took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values from the other two years as predictions.
-iii) we derivede a confusion matrix with the classes extracted from the 1000 points for LCChile, IGBP2013, and IGBP2014; and calculate the performance metrics of accuracy and F1.
+iii) we derived a confusion matrix with the classes extracted from the 1000 points for LCChile, IGBP2013, and IGBP2014; and calculated the performance metrics of accuracy and F1.
 
@@
 
-For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands (Table \ref{tab-landIGBPmacro}).
+Our results showed a global accuracy of ~0.82 and a F1 score of ~0.66 between the high resolution landcover (LCChile) and the landcover IGBP from MODIS.
+Thus, for vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands (Table \ref{tab-landIGBPmacro}).
 
@@
 \centering
-\includegraphics[width = .5\textwidth]{../output/figs/tabla_landcover_macroclass.png}
+\includegraphics[width = .3\textwidth]{../output/figs/tabla_landcover_macroclass.png}
 \end{table}
@@
 
-@fig-varImportance shows the ranking of variable importance obtained with the ten folds used in the resampling of the Random Forest model per land cover type.
+@fig-RF_importance shows the ranking of variable importance obtained with the ten folds used in the resampling of the Random Forest model per land cover type.
 
-::: {#fig-varImportance layout-ncol=3}
+```{r fig-RF_importance}
+#| echo: false
+#| layout-ncol: 3
+#| fig-pos: "!ht"
+#| fig-env: "figure*"
+#| out-width: 50%
+#| fig-cap: "Error bar for the variables importance obtained from the 10 resamples folded by the Random Forest used to model land cover trend using as predictors the trend in drought indices of water supply, water demand, soil moisture, and vegetation productivity."
+#| fig-subcap: 
+#| - "Forest"
+#| - "Cropland"
+#| - "Grassland" 
+#| - "Savanna"
+#| - "Shrubland"
+#| - "Barren Land"
 
-![Forest](../output/figs/fig_errorbar_resample_random_forest_trends_Forest.png){#fig-varimp_forest}
-
-![Cropland](../output/figs/fig_errorbar_resample_random_forest_trends_Cropland.png){#fig-varimp_crops}
-
-![Grassland](../output/figs/fig_errorbar_resample_random_forest_trends_Grassland.png){#fig-varimp_grass}
-
-![Savanna](../output/figs/fig_errorbar_resample_random_forest_trends_Savanna.png){#fig-varimp_savanna}
-
-![Shrubland](../output/figs/fig_errorbar_resample_random_forest_trends_Shrubland.png){#fig-varimp_shrubs}
-
-![Barren land](../output/figs/fig_errorbar_resample_random_forest_trends_Barren_land.png){#fig-varimp_barren}
-
-Error bar for the variables importance obtained from the 10 resamples folded by the Random Forest used to model land cover trend using as predictors the trend in drought indices of water supply, water demand, soil moisture, and vegetation productivity.
-
-:::
+knitr::include_graphics("../output/figs/fig_errorbar_resample_random_forest_trends_Forest.png")
+knitr::include_graphics("../output/figs/fig_errorbar_resample_random_forest_trends_Cropland.png")
+knitr::include_graphics("../output/figs/fig_errorbar_resample_random_forest_trends_Grassland.png")
+knitr::include_graphics("../output/figs/fig_errorbar_resample_random_forest_trends_Savanna.png")
+knitr::include_graphics("../output/figs/fig_errorbar_resample_random_forest_trends_Shrubland.png")
+knitr::include_graphics("../output/figs/fig_errorbar_resample_random_forest_trends_Barren_land.png")
+```
 
@@
 #| echo: false
-#| out-width: 75%
+#| out-width: 50%
 #| fig-pos: '!ht'
@@
 
-# References
+\newpage
+
+# References {-}
 

### 2024-05-10 article/supplementary.tex
update supplementary

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint,
  3p,
  onecolumn]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\usepackage{colortbl}
\usepackage{array}
\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Supplementary Material},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{Supplementary Material \\\large{The effects of drought on land
cover change and vegetation productivity in continental Chile} }
\author[1,2]{Francisco Zambrano%
\corref{cor1}%
}
 \ead{francisco.zambrano@umayor.cl} 
\author[3]{Anton Vrieling%
%
}

\author[4,5,6]{Francisco Meza%
%
}

\author[7]{Iongel Duran-Llacer%
%
}

\author[8,9]{Francisco Fernández%
%
}

\author[10]{Alejandro Venegas-González%
%
}

\author[4]{Nicolas Raab%
%
}

\author[12,13]{Dylan Craven%
%
}

\affiliation[1]{organization={Hémera Centro de Observación de la Tierra,
Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y
Sustentabilidad, Universidad Mayor},city={Santiago},country={7500994,
Chile.},countrysep={,},postcodesep={}}
\affiliation[2]{organization={Observatorio de Sequía para la Agricultura
y la Biodiversidad de Chile (ODES), Universidad
Mayor},city={Santiago},country={7500994,
Chile.},countrysep={,},postcodesep={}}
\affiliation[3]{organization={Faculty of Geo-Information Science and
Earth, University of Twente},city={Enschede},country={The
Netherlands.},countrysep={,},postcodesep={}}
\affiliation[4]{organization={Facultad de Agronomía y Sistemas
Naturales, Pontificia Universidad Católica de
Chile.},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
\affiliation[5]{organization={Instituto para el Desarrollo Sustentable.
Pontificia Universidad Católica de
Chile},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
\affiliation[6]{organization={Centro Interdisciplinario de Cambio
Global, Pontificia Universidad Católica de
Chile},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
\affiliation[7]{organization={Hémera Centro de Observación de la Tierra,
Facultad de Ciencias, Universidad
Mayor,},city={Santiago},country={7500994,
Chile.},countrysep={,},postcodesep={}}
\affiliation[8]{organization={Center of Economics for Sustainable
Development (CEDES), Faculty of Economics and Government, Universidad
San
Sebastian},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
\affiliation[9]{organization={Center of Applied Ecology and
Sustainability
(CAPES)},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
\affiliation[10]{organization={Instituto de Ciencias Agroalimentarias,
Animales y Ambientales (ICA3), Universidad de O'Higgins},city={San
Fernando},country={Chile.},countrysep={,},postcodesep={}}
\affiliation[11]{organization={},,postcodesep={}}
\affiliation[12]{organization={, GEMA Center for Genomics, Ecology \&
Environment, Universidad Mayor, Camino La Pirámide Huechuraba
5750},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
\affiliation[13]{organization={Data Observatory
Foundation},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}

\cortext[cor1]{Corresponding author}

\end{frontmatter}
    \renewcommand\thesection{S\arabic{section}}
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{S\arabic{table}}

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi

\hypertarget{validation-of-era5l-variables}{%
\section{Validation of ERA5L
variables}\label{validation-of-era5l-variables}}

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

We compared the ERA5L \citep{MunozSabater2021} variables for monthly
mean temperature, total precipitation, and volumetric soil water content
against values retrieved by weather stations. For temperature and
precipitation, we used data from the weather network from the Ministry
of Agriculture of Chile (www.agromet.com) between 2015 and 2023. We used
277 stations located throughout Chile. For \emph{in-situ} soil moisture,
we used a private soil network that is owned by the agricultural
enterprise Garces Fruit (https://garcesfruit.com/), which has 99
stations in Central Chile, located in fields with cherry fruit crops.
The sensors are installed at 30, 60, and 90cm and are a Advanced Soil
Moisture Sensor model Teros 12 from the METER Group
(\url{https://metergroup.com/products/teros-12/}). To avoid comparing
ERA5L with \emph{in-situ} soil moisture levels caused by irrigation,
which are not captured by ERA5L, we used daily data for the year 2022
and the months outside the growing season, May to September.

We selected the following metrics:

\[MAE = \frac{1}{n}\sum |{E-S}|\] \[Bias = \frac{\sum E}{\sum S}\]
\[ubRMSE =\sqrt{\frac{\sum{ \left[ (E_i-\overline{E})-(S_i-\overline{S}) \right ] ^2}}{n}}\]

\[CC = \frac{\sum (S_i-\bar{S})(E_i-\bar{E})}{\sqrt{(Si-\bar{S})^2(E_i-\bar{E})^2}}\]

\(MAE\): mean absolute error\\
\(bias\): bias\\
\(ubRMSE\): unbiassed root mean squared error\\
\(CC\): coefficient of correlation\\
\(S\): value of the variable measure by the weather station\\
\(E\): value of the variable measure by ERA5L

\hypertarget{results}{%
\subsection{Results}\label{results}}

The average performance metrics of ERA5L over the 266 weather stations
is shown in \citet{table_metrics_era5}

In the case of the 97 soil moisture stations, we averaged for the three
depths (30, 60, and 90cm) and then compared it with the volumetric water
content in the top 100cm of the soil derived from ERA5L. For this case,
we made a daily comparison to determine the performance metrics per
station, then we averaged the measured soil moisture at the three levels
and over all stations.

\begin{longtable}{lllrr}
\caption{Metrics of performance for the climatic and soil moisture variables from
ERA5-Land with data from 266 stations in Central Chile.}\tabularnewline

\toprule
Variable & ubRMSE & MAE & Bias & CC \\ 
\midrule\addlinespace[2.5pt]
Temperature & 1.06°C & 1.131°C &  & 0.96 \\ 
Precipitation &  & 28.1mm & 1.93 & 0.845 \\ 
Soil Moisture (1m) & 0.174 cc/cc & 0.167 cc/cc & 1.74 & 0.71 \\ 
\bottomrule
\end{longtable}

\hypertarget{land-cover-macroclasses-and-validation}{%
\section{Land cover macroclasses and
validation}\label{land-cover-macroclasses-and-validation}}

\hypertarget{methods-1}{%
\subsection{Methods}\label{methods-1}}

To analyze the LULCC, we used the IGBP scheme from the MCD12Q1
Collection 6.1 from MODIS. This product has a yearly frequency from 2001
to 2022. The IGBP defines 17 classes; we regrouped these into ten
macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9
to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands,
13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies.

To assess the accuracy of the regrouped MCD12Q1 land cover maps, we
compared the macroclasses with those of a more detailed land cover map
made by \citet{Zhao2016} for Chile with samples acquired in the years
2013--2014 (LCChile). The latter has a spatial resolution of 30 m and
three hierarchy levels of defined classes; from those, we used level 1,
which fits with the macroclass land cover. We chose the years 2013
(IGBP2013) and 2014 (IGBP2014) from the land cover macroclasses to
compare with LCChile. For this comparison, we used the following
procedure:

\begin{enumerate}
\def\labelenumi{\roman{enumi})}
\tightlist
\item
  we resampled LCChile to the spatial resolution (500m) of the land
  cover macroclasses using the majority method,
\item
  we took a random sample of 1000 points within continental Chile and
  extracted the classes that fell within each point for LCChile,
  IGBP2013, and IGBP2014; we considered the point extracted from LCChile
  as the truth and the values from the other two years as predictions.
\item
  we derived a confusion matrix with the classes extracted from the 1000
  points for LCChile, IGBP2013, and IGBP2014; and calculated the
  performance metrics of accuracy and F1.
\end{enumerate}

\[Accuracy = \frac{TP+TN}{TP+TN+FP+FN}=\frac{correct\,\, classifications}{all\,\, classifications}\]
\[F1=\frac{2\cdot TP}{2\cdot TP + FP +FN}\]

where \(TP\) and \(FN\) refer to true positive and false negative,
correctly classified classes; \(TN\) and \(FP\) to true negative and
false positive, wrongly classified classes.

\hypertarget{results-1}{%
\subsection{Results}\label{results-1}}

Our results showed a global accuracy of \textasciitilde0.82 and a F1
score of \textasciitilde0.66 between the high resolution landcover
(LCChile) and the landcover IGBP from MODIS. Thus, for vegetation, we
obtained and use hereafter five macroclasses of land cover from IGBP
MODIS: forest, shrubland, savanna, grassland, and croplands (Table
\ref{tab-landIGBPmacro}).

\begin{table}[!ht]
\caption{Landcover clases from IGBP MODIS and the corresponding macroclasses.}
\label{tab-landIGBPmacro}
\centering
\includegraphics[width = .3\textwidth]{../output/figs/tabla_landcover_macroclass.png}
\end{table}

\hypertarget{relationship-between-drought-indices-and-land-cover-change}{%
\section{Relationship between drought indices and land cover
change}\label{relationship-between-drought-indices-and-land-cover-change}}

Figure~\ref{fig-RF_importance} shows the ranking of variable importance
obtained with the ten folds used in the resampling of the Random Forest
model per land cover type.

\begin{figure}

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{../output/figs/fig_errorbar_resample_random_forest_trends_Forest.png}

}

}

\subcaption{\label{fig-RF_importance-1}Forest}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{../output/figs/fig_errorbar_resample_random_forest_trends_Cropland.png}

}

}

\subcaption{\label{fig-RF_importance-2}Cropland}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{../output/figs/fig_errorbar_resample_random_forest_trends_Grassland.png}

}

}

\subcaption{\label{fig-RF_importance-3}Grassland}
\end{minipage}%
\newline
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{../output/figs/fig_errorbar_resample_random_forest_trends_Savanna.png}

}

}

\subcaption{\label{fig-RF_importance-4}Savanna}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{../output/figs/fig_errorbar_resample_random_forest_trends_Shrubland.png}

}

}

\subcaption{\label{fig-RF_importance-5}Shrubland}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{../output/figs/fig_errorbar_resample_random_forest_trends_Barren_land.png}

}

}

\subcaption{\label{fig-RF_importance-6}Barren Land}
\end{minipage}%

\caption{\label{fig-RF_importance}Error bar for the variables importance
obtained from the 10 resamples folded by the Random Forest used to model
land cover trend using as predictors the trend in drought indices of
water supply, water demand, soil moisture, and vegetation productivity.}

\end{figure}

\hypertarget{trend-of-vegetation-productivity}{%
\section{Trend of vegetation
productivity}\label{trend-of-vegetation-productivity}}

Figure~\ref{fig-hetmaptrendzcNDVI} shows the average trend for zcNDVI
for 2000-2023 per macrozone and landcover macroclass.

\begin{figure}[!ht]

{\centering \includegraphics[width=0.5\textwidth,height=\textheight]{../output/figs/heatmap_trends_zcNDVI_macro_landcover.png}

}

\caption{\label{fig-hetmaptrendzcNDVI}Heatmap of trends in zcNDVI for
2000 to 2023 per macrozone and landcover macroclass.}

\end{figure}

\hypertarget{vegetation-productivity}{%
\section{Vegetation productivity}\label{vegetation-productivity}}

We analyzed the correlation of zcNDVI for time scales of 1, 3, 6, and 12
months versus net primary production (NPP). We obtained both the zcNDVI
from MOD13A3.061 and the NPP from MOD17A3HGF.061, using MODIS products.
We used the zcNDVI in December to correlate with the annual NPP.
Figure~\ref{fig-r2_npp_zcNDVI} shows a map of the r-squared correlation
between zcNDVI and NPP, and Figure~\ref{fig-hetmap_npp_zcndvi} shows the
aggregated values per macrozone.

\begin{figure}[!ht]

{\centering \includegraphics[width=0.5\textwidth,height=\textheight]{../output/figs/map_r2_NPP_vs_zcNDVI.png}

}

\caption{\label{fig-r2_npp_zcNDVI}Spatial variation of the r-squared
values obtained from the yearly correlation of zcNDVI of 1, 3, 6, and 12
months with the net primary productivity (NPP) for continental Chile.}

\end{figure}

\begin{figure}[!ht]

{\centering \includegraphics[width=0.5\textwidth,height=\textheight]{../output/figs/heatmap_r2_npp_vs_zcndvi_per_landcover.png}

}

\caption{\label{fig-hetmap_npp_zcndvi}A heatmap showing the r-squared
values obtained from the yearly correlation of zcNDVI of 1, 3, 6, and 12
months with the net primary productivity (NPP) for continental Chile.}

\end{figure}

\newpage

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\renewcommand{\bibsection}{}
\bibliography{references.bib}

\end{document}

### 2024-06-25 article/article_first_version/paper_drought_Chile2023.qmd
actualizar repositorio0+

--- 2024-05-09 article/paper_drought_Chile2023.qmd
@@
   - name: Anton Vrieling
-    email: a.vrieling@utwente.nl
+    #email: a.vrieling@utwente.nl
     affiliations:
@@
   - name: Francisco Fernández
-    email: francisco.fernandez@uss.cl
+    #email: francisco.fernandez@uss.cl
     affiliations:
@@
 This product has been previously used for studies of drought and land cover in Chile [@Fuentes2021;@Zambrano2018].
-We regrouped the 17 classes into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies (Table S1).
+We regrouped the 17 classes into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies (see Table S2).
 Thus, we have a land cover raster time series with the ten macroclasses for 2001 and 2023.
@@
 The National Research and Development Agency of Chile (ANID) funded this study through the drought emergency project FSEQ210022, Fondecyt Iniciación N°11190360, Fondecyt Postdoctorado N°3230678, and Fondecyt Regular N°1210526.
+The companies Garces Fruit and CER Analytics provided access to the _in-situ_ soil moisture data used in this work.
 

### 2024-06-25 article/article_first_version/paper_drought_Chile2023.tex
actualizar repositorio0+

--- 2024-05-09 article/paper_drought_Chile2023.tex
@@
 \makeatletter
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
 \makeatother
 \makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
 \makeatother
@@
 \bibliographystyle{elsarticle-harv}
-\usepackage{bookmark}
-
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 }
- \ead{a.vrieling@utwente.nl} 
+
 \author[4,5,6]{Francisco Meza%
@@
 }
+
 \author[10]{Alejandro Venegas-González%
@@
 Sustentabilidad, Universidad Mayor},city={Santiago},country={7500994,
-Chile.}}
+Chile.},countrysep={,},postcodesep={}}
 \affiliation[2]{organization={Observatorio de Sequía para la Agricultura
@@
 Mayor},city={Santiago},country={7500994,
-Chile.}}
+Chile.},countrysep={,},postcodesep={}}
 \affiliation[3]{organization={Faculty of Geo-Information Science and
 Earth, University of Twente},city={Enschede},country={The
-Netherlands.}}
+Netherlands.},countrysep={,},postcodesep={}}
 \affiliation[4]{organization={Facultad de Agronomía y Sistemas
 Naturales, Pontificia Universidad Católica de
-Chile.},city={Santiago},country={Chile.}}
+Chile.},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
 \affiliation[5]{organization={Instituto para el Desarrollo Sustentable.
 Pontificia Universidad Católica de
-Chile},city={Santiago},country={Chile.}}
+Chile},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
 \affiliation[6]{organization={Centro Interdisciplinario de Cambio
 Global, Pontificia Universidad Católica de
-Chile},city={Santiago},country={Chile.}}
+Chile},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
 \affiliation[7]{organization={Hémera Centro de Observación de la Tierra,
@@
 Mayor,},city={Santiago},country={7500994,
-Chile.}}
+Chile.},countrysep={,},postcodesep={}}
 \affiliation[8]{organization={Center of Economics for Sustainable
@@
 San
-Sebastian},city={Santiago},country={Chile.}}
+Sebastian},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
 \affiliation[9]{organization={Center of Applied Ecology and
 Sustainability
-(CAPES)},city={Santiago},country={Chile.}}
+(CAPES)},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
 \affiliation[10]{organization={Instituto de Ciencias Agroalimentarias,
 Animales y Ambientales (ICA3), Universidad de O'Higgins},city={San
-Fernando},country={Chile.}}
-\affiliation[11]{organization={, GEMA Center for Genomics, Ecology \&
+Fernando},country={Chile.},countrysep={,},postcodesep={}}
+\affiliation[11]{organization={},,postcodesep={}}
+\affiliation[12]{organization={, GEMA Center for Genomics, Ecology \&
 Environment, Universidad Mayor, Camino La Pirámide Huechuraba
-5750},city={Santiago},country={Chile.}}
-\affiliation[12]{organization={Data Observatory
-Foundation},city={Santiago},country={Chile.}}
+5750},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
+\affiliation[13]{organization={Data Observatory
+Foundation},city={Santiago},country={Chile.},countrysep={,},postcodesep={}}
 
@@
 \end{frontmatter}
-
     \captionsetup{justification=raggedright,singlelinecheck=false}
 
-\section{Introduction}\label{introduction}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, sharp corners, boxrule=0pt, interior hidden]}{\end{tcolorbox}}\fi
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 
-\section{Study area}\label{study-area}
+\hypertarget{study-area}{%
+\section{Study area}\label{study-area}}
 
@@
 
-\centering{
-
-\includegraphics{../output/figs/map_study_con_landcover.png}
+{\centering \includegraphics{../output/figs/map_study_con_landcover.png}
 
@@
 
-\end{figure*}%
-
-\section{Materials and Methods}\label{materials-and-methods}
-
-\subsection{Data}\label{data}
-
+\end{figure*}
+
+\hypertarget{materials-and-methods}{%
+\section{Materials and Methods}\label{materials-and-methods}}
+
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
+
+\hypertarget{gridded-meteorological-and-vegetation-data}{%
 \subsubsection{Gridded meteorological and vegetation
-data}\label{gridded-meteorological-and-vegetation-data}
+data}\label{gridded-meteorological-and-vegetation-data}}
 
@@
 
-\begin{table}
-
-\centering{
-
-[!ht]
-
+\begin{table}[!ht]
+\caption{Description of the satellite and reanalysis data used}
+\label{tab-desEOD}
 \scriptsize
@@
 {\raggedright *Calculated from maximum and minimum temperatures derived from ERA5L with Eq. \ref{eq-AED}. \par}
-
-}
-
-\caption{\label{tab-desEOD}Description of the satellite and reanalysis data used}
-
-\end{table}%
+\end{table}
 
@@
 
+\hypertarget{short--to-long-term-drought-trends}{%
 \subsection{Short- to long-term drought
-trends}\label{short--to-long-term-drought-trends}
-
+trends}\label{short--to-long-term-drought-trends}}
+
+\hypertarget{atmospheric-evaporative-demand-aed}{%
 \subsubsection{Atmospheric Evaporative Demand
-(AED)}\label{atmospheric-evaporative-demand-aed}
+(AED)}\label{atmospheric-evaporative-demand-aed}}
 
@@
 
-\begin{equation}\phantomsection\label{eq-AED}{AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}}\end{equation}
+\begin{equation}\protect\hypertarget{eq-AED}{}{AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}}\label{eq-AED}\end{equation}
 
@@
 
-\begin{equation}\phantomsection\label{eq-Ra}{R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]}\end{equation}
+\begin{equation}\protect\hypertarget{eq-Ra}{}{R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]}\label{eq-Ra}\end{equation}
 
@@
 
+\hypertarget{non-parametric-calculation-of-drought-indices}{%
 \subsubsection{Non-parametric calculation of drought
-indices}\label{non-parametric-calculation-of-drought-indices}
+indices}\label{non-parametric-calculation-of-drought-indices}}
 
@@
 
-\begin{equation}\phantomsection\label{eq-sumvar}{A^s_i = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  }\end{equation}
+\begin{equation}\protect\hypertarget{eq-sumvar}{}{A^s_i = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  }\label{eq-sumvar}\end{equation}
 
@@
 
-\begin{equation}\phantomsection\label{eq-probPai}{P(A^s_i) = \frac{i-0.33}{n+0.33'}}\end{equation}
+\begin{equation}\protect\hypertarget{eq-probPai}{}{P(A^s_i) = \frac{i-0.33}{n+0.33'}}\label{eq-probPai}\end{equation}
 
@@
 
-\begin{equation}\phantomsection\label{eq-DI}{DI(A^s_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}}\end{equation}
+\begin{equation}\protect\hypertarget{eq-DI}{}{DI(A^s_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}}\label{eq-DI}\end{equation}
 
@@
 
-\subsubsection{Trend of drought indices}\label{trend-of-drought-indices}
+\hypertarget{trend-of-drought-indices}{%
+\subsubsection{Trend of drought
+indices}\label{trend-of-drought-indices}}
 
@@
 
+\hypertarget{interaction-of-land-cover-and-drought}{%
 \subsection{Interaction of land cover and
-drought}\label{interaction-of-land-cover-and-drought}
-
-\subsubsection{Land cover change}\label{land-cover-change}
+drought}\label{interaction-of-land-cover-and-drought}}
+
+\hypertarget{land-cover-change}{%
+\subsubsection{Land cover change}\label{land-cover-change}}
 
@@
 
+\hypertarget{relationship-between-land-cover-and-drought-trends}{%
 \subsubsection{Relationship between land cover and drought
-trends}\label{relationship-between-land-cover-and-drought-trends}
+trends}\label{relationship-between-land-cover-and-drought-trends}}
 
@@
 
+\hypertarget{drought-impacts-on-vegetation-productivity}{%
 \subsection{Drought impacts on vegetation
-productivity}\label{drought-impacts-on-vegetation-productivity}
+productivity}\label{drought-impacts-on-vegetation-productivity}}
 
@@
 
-\subsection{Software}\label{software}
+\hypertarget{software}{%
+\subsection{Software}\label{software}}
 
@@
 
-\section{Results}\label{results}
-
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\hypertarget{short--to-long-term-drought-trends-1}{%
 \subsection{Short- to long-term drought
-trends}\label{short--to-long-term-drought-trends-1}
+trends}\label{short--to-long-term-drought-trends-1}}
 
@@
 
-\begin{figure*}[!ht]
-
-\begin{minipage}{0.50\linewidth}
-
-\centering{
+\begin{figure}
+
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
 
@@
 
+}
+
 \subcaption{\label{fig-trendDI-1}SPI (Standardized Precipitation Index)}
-
 \end{minipage}%
 %
-\begin{minipage}{0.50\linewidth}
-
-\centering{
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
 
 \includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}
+
+}
 
@@
 Index)}
-
 \end{minipage}%
 \newline
-\begin{minipage}{0.50\linewidth}
-
-\centering{
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
 
 \includegraphics{../output/figs/trend_raster_SPEI_1981-2023.png}
+
+}
 
@@
 Evapotranspiration Index)}
-
 \end{minipage}%
 %
-\begin{minipage}{0.50\linewidth}
-
-\centering{
+\begin{minipage}[t]{0.50\linewidth}
+
+{\centering
+
+\raisebox{-\height}{
 
@@
 
+}
+
 \subcaption{\label{fig-trendDI-4}SSI (Standardized Soil Moisture Index)}
-
 \end{minipage}%
@@
 
-\end{figure*}%
+\end{figure}
 
@@
 
-\centering{
-
-\includegraphics{../output/figs/trend_macrozone_drought_indices.png}
+{\centering \includegraphics{../output/figs/trend_macrozone_drought_indices.png}
 
@@
 
-\end{figure}%
+\end{figure}
 
@@
 
+\hypertarget{interaction-of-land-cover-and-drought-1}{%
 \subsection{Interaction of land cover and
-drought}\label{interaction-of-land-cover-and-drought-1}
-
-\subsubsection{Land cover change}\label{land-cover-change-1}
+drought}\label{interaction-of-land-cover-and-drought-1}}
+
+\hypertarget{land-cover-change-1}{%
+\subsubsection{Land cover change}\label{land-cover-change-1}}
 
@@
 
-\centering{
-
-\includegraphics{../output/figs/proportion_landcover_macroclass_2001-2022.png}
+{\centering \includegraphics{../output/figs/proportion_landcover_macroclass_2001-2022.png}
 
@@
 
+\hypertarget{relationship-between-drought-indices-and-land-cover-change}{%
 \subsubsection{Relationship between drought indices and land cover
-change}\label{relationship-between-drought-indices-and-land-cover-change}
+change}\label{relationship-between-drought-indices-and-land-cover-change}}
 
@@
 
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/bars_relative_importance_RF.png}
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{../output/figs/bars_relative_importance_RF.png}
 
@@
 
-\centering{
-
-\includegraphics{../output/figs/points_landcover_drought_indices_trend_and_time_scale.png}
+{\centering \includegraphics{../output/figs/points_landcover_drought_indices_trend_and_time_scale.png}
 
@@
 
+\hypertarget{drought-impacts-on-vegetation-productivity-within-land-cover}{%
 \subsection{Drought impacts on vegetation productivity within land
-cover}\label{drought-impacts-on-vegetation-productivity-within-land-cover}
-
+cover}\label{drought-impacts-on-vegetation-productivity-within-land-cover}}
+
+\hypertarget{trends-in-vegetation-productivity}{%
 \subsubsection{Trends in vegetation
-productivity}\label{trends-in-vegetation-productivity}
+productivity}\label{trends-in-vegetation-productivity}}
 
@@
 
-\centering{
-
-\includegraphics{../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png}
+{\centering \includegraphics{../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png}
 
@@
 
+\hypertarget{correlation-between-vegetation-productivity-and-drought-indices}{%
 \subsubsection{Correlation between vegetation productivity and drought
-indices}\label{correlation-between-vegetation-productivity-and-drought-indices}
+indices}\label{correlation-between-vegetation-productivity-and-drought-indices}}
 
@@
 
-\centering{
-
-\includegraphics{../output/figs/mapa_cor_selec_indices_zcNDVI6.png}
+{\centering \includegraphics{../output/figs/mapa_cor_selec_indices_zcNDVI6.png}
 
@@
 
-\centering{
-
-\includegraphics{../output/figs/mapa_cor_r_indices_zcNDVI6.png}
+{\centering \includegraphics{../output/figs/mapa_cor_r_indices_zcNDVI6.png}
 
@@
 
-\section{Discussion}\label{discussion}
-
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+\hypertarget{vegetation-water-demand-and-its-relation-to-drought}{%
 \subsection{Vegetation water demand and its relation to
-drought}\label{vegetation-water-demand-and-its-relation-to-drought}
+drought}\label{vegetation-water-demand-and-its-relation-to-drought}}
 
@@
 
+\hypertarget{sensitivity-of-land-cover-vegetation-to-short--and-long-term-drought}{%
 \subsection{Sensitivity of land cover vegetation to short- and long-term
-drought}\label{sensitivity-of-land-cover-vegetation-to-short--and-long-term-drought}
+drought}\label{sensitivity-of-land-cover-vegetation-to-short--and-long-term-drought}}
 
@@
 
+\hypertarget{vegetation-productivity-and-drought}{%
 \subsection{Vegetation productivity and
-drought}\label{vegetation-productivity-and-drought}
+drought}\label{vegetation-productivity-and-drought}}
 
@@
 
+\hypertarget{drought-information-to-aid-in-adaptation}{%
 \subsection{Drought information to aid in
-adaptation}\label{drought-information-to-aid-in-adaptation}
+adaptation}\label{drought-information-to-aid-in-adaptation}}
 
@@
 
-\section{Conclusion}\label{conclusion}
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
@@
 
-\section{Acknowledgment}\label{acknowledgment}
+\hypertarget{acknowledgment}{%
+\section{Acknowledgment}\label{acknowledgment}}
 
@@
 Iniciación N°11190360, Fondecyt Postdoctorado N°3230678, and Fondecyt
-Regular N°1210526.
+Regular N°1210526. The companies Garces Fruit and CER Analytics provided
+access to the \emph{in-situ} soil moisture data used in this work.
 

### 2024-06-25 article/article_first_version/supplementary.qmd
actualizar repositorio0+

--- 2024-05-10 article/supplementary.qmd
@@
 
+# Data
+
+```{=latex}
+\begin{table}[!ht]
+\caption{Description of the satellite and reanalysis data used}
+\label{tab-desEOD}
+\scriptsize
+\centering
+\begin{tabular}{p{0.1\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
+\hline
+\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
+\hline
+\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
+         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
+         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
+         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
+ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
+        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
+         & MCD12Q1.061 & land cover IGBP scheme & & 2001-2022 & ~ & land cover \\ 
+\hline
+\end{tabular}
+{\raggedright *Calculated from maximum and minimum temperatures derived from ERA5L with Eq. \ref{eq-AED}. \par}
+\end{table}
+```
+
 # Validation of ERA5L variables
@@
 
-The average performance metrics of ERA5L over the 266 weather stations is shown in @table_metrics_era5
+The average performance metrics of ERA5L over the weather and soil moisture stations are shown in @tbl-metricsera5.
+In the case of the 97 soil moisture stations, we averaged for the three depths (30, 60, and 90cm) and then compared it with the volumetric water content in the top 100cm of the soil derived from ERA5L.
+For this case, we made a daily comparison to determine the performance metrics per station, then we averaged the measured soil moisture at the three levels and over all stations.
 
@@
 
-In the case of the 97 soil moisture stations, we averaged for the three depths (30, 60, and 90cm) and then compared it with the volumetric water content in the top 100cm of the soil derived from ERA5L.
-For this case, we made a daily comparison to determine the performance metrics per station, then we averaged the measured soil moisture at the three levels and over all stations.
-
-```{r table_metrics_era5}
+```{r}
 #| echo: false
-#| tbl-cap: "Metrics of performance for the climatic and soil moisture variables from ERA5-Land with data from 266 stations in Central Chile."
+#| label: tbl-metricsera5
+#| tbl-cap: "Metrics of performance for the climatic and soil moisture variables from ERA5-Land with data from 277 stations in Central Chile for temperature and precipitation and 99 stations of soil moisture."
 tribble(~Variable,~ubRMSE,~MAE,~Bias,~CC,
-        "Temperature","1.06°C","1.131°C","","0.96",
-        "Precipitation","","28.1mm","1.93","0.845",
+        "Temperature","1.06 °C","1.13 °C","1.02","0.96",
+        "Precipitation","35.5 mm","28.1 mm","1.93","0.85",
         "Soil Moisture (1m)","0.174 cc/cc","0.167 cc/cc","1.74", "0.71") |> 
@@
 \centering
-\includegraphics[width = .3\textwidth]{../output/figs/tabla_landcover_macroclass.png}
+\includegraphics[width = .5\textwidth]{../output/figs/tabla_landcover_macroclass.png}
 \end{table}
@@
 #| echo: false
-#| layout-ncol: 3
+#| layout-nrow: 2
 #| fig-pos: "!ht"
 #| fig-env: "figure*"
-#| out-width: 50%
+#| out-width: 100%
 #| fig-cap: "Error bar for the variables importance obtained from the 10 resamples folded by the Random Forest used to model land cover trend using as predictors the trend in drought indices of water supply, water demand, soil moisture, and vegetation productivity."
@@
 #| echo: false
-#| out-width: 50%
+#| out-width: 75%
 #| fig-pos: '!ht'
@@
 
+# Relationship of drought indices and vegetation productivity
+
+```{r}
+#| echo: false
+#| out-width: 100%
+#| fig-pos: '!ht'
+#| fig-cap: "Summary per land cover and study region regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
+The numbers in each cell indicate the time scale that reached the maximum correlation for the land cover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.
+Cells without values indicate that the land cover type was not significant in that macrozone."
+knitr::include_graphics('../output/figs/tabla_r_cor_macro_indice.png')
+```
+
 \newpage

### 2024-06-25 article/article_first_version/supplementary.tex
actualizar repositorio0+

--- 2024-05-10 article/supplementary.tex
@@
 \end{frontmatter}
+
 \renewcommand\thesection{S\arabic{section}}
@@
 
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable]}{\end{tcolorbox}}\fi
+\section{Data}\label{data}
 
-\hypertarget{validation-of-era5l-variables}{%
+\begin{table}
+
+\centering{
+
+[!ht]
+
+\scriptsize
+\centering
+\begin{tabular}{p{0.1\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
+\hline
+\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
+\hline
+\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
+         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
+         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
+         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
+ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
+        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
+         & MCD12Q1.061 & land cover IGBP scheme & & 2001-2022 & ~ & land cover \\ 
+\hline
+\end{tabular}
+{\raggedright *Calculated from maximum and minimum temperatures derived from ERA5L with Eq. \ref{eq-AED}. \par}
+
+}
+
+\caption{\label{tab-desEOD}Description of the satellite and reanalysis data used}
+
+\end{table}%
+
 \section{Validation of ERA5L
-variables}\label{validation-of-era5l-variables}}
+variables}\label{validation-of-era5l-variables}
 
-\hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
+\subsection{Methods}\label{methods}
 
@@
 
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
+\subsection{Results}\label{results}
 
-The average performance metrics of ERA5L over the 266 weather stations
-is shown in \citet{table_metrics_era5}
-
-In the case of the 97 soil moisture stations, we averaged for the three
-depths (30, 60, and 90cm) and then compared it with the volumetric water
-content in the top 100cm of the soil derived from ERA5L. For this case,
-we made a daily comparison to determine the performance metrics per
-station, then we averaged the measured soil moisture at the three levels
-and over all stations.
+The average performance metrics of ERA5L over the weather and soil
+moisture stations are shown in Table~\ref{tbl-metricsera5}. In the case
+of the 97 soil moisture stations, we averaged for the three depths (30,
+60, and 90cm) and then compared it with the volumetric water content in
+the top 100cm of the soil derived from ERA5L. For this case, we made a
+daily comparison to determine the performance metrics per station, then
+we averaged the measured soil moisture at the three levels and over all
+stations.
 
 \begin{longtable}{lllrr}
-\caption{Metrics of performance for the climatic and soil moisture variables from
-ERA5-Land with data from 266 stations in Central Chile.}\tabularnewline
+
+\caption{\label{tbl-metricsera5}Metrics of performance for the climatic
+and soil moisture variables from ERA5-Land with data from 277 stations
+in Central Chile for temperature and precipitation and 99 stations of
+soil moisture.}
+
+\tabularnewline
 
@@
 \midrule\addlinespace[2.5pt]
-Temperature & 1.06°C & 1.131°C &  & 0.96 \\ 
-Precipitation &  & 28.1mm & 1.93 & 0.845 \\ 
+Temperature & 1.06 °C & 1.13 °C & 1.02 & 0.96 \\ 
+Precipitation & 35.5 mm & 28.1 mm & 1.93 & 0.85 \\ 
 Soil Moisture (1m) & 0.174 cc/cc & 0.167 cc/cc & 1.74 & 0.71 \\ 
 \bottomrule
+
 \end{longtable}
 
-\hypertarget{land-cover-macroclasses-and-validation}{%
 \section{Land cover macroclasses and
-validation}\label{land-cover-macroclasses-and-validation}}
+validation}\label{land-cover-macroclasses-and-validation}
 
-\hypertarget{methods-1}{%
-\subsection{Methods}\label{methods-1}}
+\subsection{Methods}\label{methods-1}
 
@@
 
-\hypertarget{results-1}{%
-\subsection{Results}\label{results-1}}
+\subsection{Results}\label{results-1}
 
@@
 \centering
-\includegraphics[width = .3\textwidth]{../output/figs/tabla_landcover_macroclass.png}
+\includegraphics[width = .5\textwidth]{../output/figs/tabla_landcover_macroclass.png}
 \end{table}
 
-\hypertarget{relationship-between-drought-indices-and-land-cover-change}{%
 \section{Relationship between drought indices and land cover
-change}\label{relationship-between-drought-indices-and-land-cover-change}}
+change}\label{relationship-between-drought-indices-and-land-cover-change}
 
@@
 
-\begin{figure}
+\begin{figure*}[!ht]
 
-\begin{minipage}[t]{0.33\linewidth}
+\begin{minipage}{0.33\linewidth}
 
-{\centering
+\centering{
 
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/fig_errorbar_resample_random_forest_trends_Forest.png}
-
-}
+\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/fig_errorbar_resample_random_forest_trends_Forest.png}
 
@@
 \subcaption{\label{fig-RF_importance-1}Forest}
+
 \end{minipage}%
 %
-\begin{minipage}[t]{0.33\linewidth}
+\begin{minipage}{0.33\linewidth}
 
-{\centering
+\centering{
 
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/fig_errorbar_resample_random_forest_trends_Cropland.png}
-
-}
+\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/fig_errorbar_resample_random_forest_trends_Cropland.png}
 
@@
 \subcaption{\label{fig-RF_importance-2}Cropland}
+
 \end{minipage}%
 %
-\begin{minipage}[t]{0.33\linewidth}
+\begin{minipage}{0.33\linewidth}
 
-{\centering
+\centering{
 
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/fig_errorbar_resample_random_forest_trends_Grassland.png}
-
-}
+\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/fig_errorbar_resample_random_forest_trends_Grassland.png}
 
@@
 \subcaption{\label{fig-RF_importance-3}Grassland}
+
 \end{minipage}%
 \newline
-\begin{minipage}[t]{0.33\linewidth}
+\begin{minipage}{0.33\linewidth}
 
-{\centering
+\centering{
 
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/fig_errorbar_resample_random_forest_trends_Savanna.png}
-
-}
+\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/fig_errorbar_resample_random_forest_trends_Savanna.png}
 
@@
 \subcaption{\label{fig-RF_importance-4}Savanna}
+
 \end{minipage}%
 %
-\begin{minipage}[t]{0.33\linewidth}
+\begin{minipage}{0.33\linewidth}
 
-{\centering
+\centering{
 
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/fig_errorbar_resample_random_forest_trends_Shrubland.png}
-
-}
+\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/fig_errorbar_resample_random_forest_trends_Shrubland.png}
 
@@
 \subcaption{\label{fig-RF_importance-5}Shrubland}
+
 \end{minipage}%
 %
-\begin{minipage}[t]{0.33\linewidth}
+\begin{minipage}{0.33\linewidth}
 
-{\centering
+\centering{
 
-\raisebox{-\height}{
-
-\includegraphics{../output/figs/fig_errorbar_resample_random_forest_trends_Barren_land.png}
-
-}
+\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/fig_errorbar_resample_random_forest_trends_Barren_land.png}
 
@@
 \subcaption{\label{fig-RF_importance-6}Barren Land}
+
 \end{minipage}%
@@
 
-\end{figure}
+\end{figure*}%
 
-\hypertarget{trend-of-vegetation-productivity}{%
 \section{Trend of vegetation
-productivity}\label{trend-of-vegetation-productivity}}
+productivity}\label{trend-of-vegetation-productivity}
 
@@
 
-{\centering \includegraphics[width=0.5\textwidth,height=\textheight]{../output/figs/heatmap_trends_zcNDVI_macro_landcover.png}
+\centering{
+
+\includegraphics[width=0.75\textwidth,height=\textheight]{../output/figs/heatmap_trends_zcNDVI_macro_landcover.png}
 
@@
 
-\end{figure}
+\end{figure}%
 
-\hypertarget{vegetation-productivity}{%
-\section{Vegetation productivity}\label{vegetation-productivity}}
+\section{Vegetation productivity}\label{vegetation-productivity}
 
@@
 
-{\centering \includegraphics[width=0.5\textwidth,height=\textheight]{../output/figs/map_r2_NPP_vs_zcNDVI.png}
+\centering{
+
+\includegraphics[width=0.75\textwidth,height=\textheight]{../output/figs/map_r2_NPP_vs_zcNDVI.png}
 
@@
 
-{\centering \includegraphics[width=0.5\textwidth,height=\textheight]{../output/figs/heatmap_r2_npp_vs_zcndvi_per_landcover.png}
+\centering{
+
+\includegraphics[width=0.5\textwidth,height=\textheight]{../output/figs/heatmap_r2_npp_vs_zcndvi_per_landcover.png}
 
@@
 
-\end{figure}
+\end{figure}%
+
+\section{Relationship of drought indices and vegetation
+productivity}\label{relationship-of-drought-indices-and-vegetation-productivity}
+
+\begin{figure}[H]
+
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{../output/figs/tabla_r_cor_macro_indice.png}
+
+}
+
+\caption{Summary per land cover and study region regarding the
+correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and
+SSI for time scales of 1, 3, 6, 12, 24, and 36. The numbers in each cell
+indicate the time scale that reached the maximum correlation for the
+land cover and macrozone, and the color indicates the strength of the
+r-squared obtained with the index and the time scale. Cells without
+values indicate that the land cover type was not significant in that
+macrozone.}
+
+\end{figure}%
 
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}

### 2024-06-25 article/article_nature/article_nature.qmd
actualizar repositorio0+

---
title: Shifts in water supply and demand drive land cover change across Chile
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-nature
    classoption:
      - numbered
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
  html:
    code-fold: true
    html-math-method: katex
  docx: default
# The `csl` style will be used with LaTeX-based rendering of`nature-pdf`
# if cite-method is `citeproc`. This is also the default if cite-method
# is not specified.
csl: _extensions/christopherkenny/nature/csl/springer-vancouver.csl
author:
  - name: Francisco Zambrano
    email: francisco.zambrano@umayor.cl
    affiliations: 
        - name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad, Universidad Mayor
          id: 1
          city: Santiago
          country: 7500994, Chile.
        - id: 2
          name: Observatorio de Sequía para la Agricultura y la Biodiversidad de Chile (ODES), Universidad Mayor
          city: Santiago
          country: 7500994, Chile.
    attributes:
        corresponding: true
    #note: This is the first author footnote.
  - name: Anton Vrieling
    #email: a.vrieling@utwente.nl
    affiliations:
        - id: 3
          name: Faculty of Geo-Information Science and Earth, University of Twente
          city:  Enschede
          country: The Netherlands.
  - name: Francisco Meza
    affiliations:
        - id: 4
          name: Facultad de Agronomía y Sistemas Naturales, Pontificia Universidad Católica de Chile. 
          city: Santiago
          country: Chile.
        - id: 5
          name: Instituto para el Desarrollo Sustentable. Pontificia Universidad Católica de Chile
          city: Santiago
          country: Chile.
        - id: 6
          name: Centro Interdisciplinario de Cambio Global, Pontificia Universidad Católica de Chile
          city: Santiago
          country: Chile.
  - name: Iongel Duran-Llacer
    affiliations:
        - id: 7
          name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Universidad Mayor,
          city: Santiago
          country: 7500994, Chile.
  - name: Francisco Fernández
    #email: francisco.fernandez@uss.cl
    affiliations:
        - id: 8
          name: Center of Economics for Sustainable Development (CEDES), Faculty of Economics and Government, Universidad San Sebastian
          city: Santiago
          country: Chile.
        - id: 9
          name: Center of Applied Ecology and Sustainability (CAPES)
          city: Santiago
          country: Chile.
  - name: Alejandro Venegas-González
    affiliations:
        - id: 10
          name: Instituto de Ciencias Agroalimentarias, Animales y Ambientales (ICA3), Universidad de O'Higgins
          city: San Fernando
          country: Chile.
  - name: Nicolas Raab
    affiliations:
        - id: 4
  - name: Dylan Craven
    affiliations:
        - id: 11
          department: GEMA Center for Genomics, Ecology & Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750
          city: Santiago
          country: Chile.
        - id: 12
          name: Data Observatory Foundation
          city: Santiago
          country: Chile.
abstract: |
  Globally, droughts are becoming longer, more frequent and more severe, and their impacts are multidimensional. These impacts typically extend beyond the water balance as  long-term, cumulative changes in the water balance can lead to regime shifts in land use. Here, we assess the effects of temporal changes in water supply and demand on vegetation productivity and land cover change over multiple time scales in continental Chile, which has experienced a severe drought over the last 20 years. Across most of continental Chile, we found a persistent decreasing trend in water supply and an increasing trend in water demand since 1981, trends that intensify over longer time scales. This long-term decrease in water availability has led to a decrease in vegetation productivity, especially in central and southern Chile. Our models suggest that increasing drought severity has led to shifts in land use towards more drought-tolerant land cover types, such as shrublands. We also found evidence that human perceptions of prolonged drought can indirectly lead to large-scale changes in land use. Our results suggest that long-term climate change may lead to regime shifts in land cover that can be mitigated by context-specific adaptation strategies.
keywords: [drought, land cover, water demand, water supply]
bibliography: references.bib  
---

# Introduction {#sec-intro}

Across many regions of the world, droughts are becoming longer and more frequent and severe [@Miranda2023;@IPCC2023], impacting ecosystems via tree mortality[@Cheng2024] and productivity [@Miranda2023] and inducing shifts in land cover and use [@Crausbay2017]. However, identifying drought events is surprisingly idiosyncratic due to the varying criteria used for classification. Droughts can be classified as either 1) meteorological, i.e., when precipitation in a specific period falls below mean precipitation values observed over multiple years (usually more than 30 years); 2) hydrological, i.e., when precipitation anomalies last for long periods (months to years) and affect water systems; 3) agricultural, i.e. when precipitation deficits negatively impact plant health, leading to decreases in crop or pasture productivity [@Wilhite1985]; or 4) ecological, i.e., when precipitation deficits negatively affect the provisioning of ecosystem services and trigger feedbacks in natural or human systems [@Crausbay2017]. Such feedbacks include drought impacts on human decision making and activities, which can lead to land-use change [@VanLoon2016;@AghaKouchak2021], which may have cascading effects on biodiversity and ecosystem services (e.g., @Lawler2014;@Newbold2015).

Despite the high degree of confidence in the impacts of rising temperatures on the extent, frequency, and severity of agricultural and ecological droughts [@IPCC2023], which are likely to increase even if global warming stabilizes at 1.5°–2°C, the severity of meteorological droughts has been remarkably stable globally over the past century [@Vicente-Serrano2022;@Kogan2020]. In the few regions where drought severity has increased over this period (1900-2000), rising temperatures have increased atmospheric evaporative demand (AED), which has been associated with increases in agricultural land area [@Vicente-Serrano2022]. Thus, rising water demand may reflect parallel changes in land use - primarily agriculture - that can exacerbate the effects of meteorological droughts on ecosystems.

From 1960 to 2019, land-use change has impacted approximately one-third of the Earth’s surface, which is four times more than previously thought[@Winkler2021]. Despite the considerable interest in land-use change dynamics [e.g., @Song2018;@Winkler2021], the direction and magnitude of  drought impacts on land cover change and vegetation productivity remain uncertain [@Chen2022;@Akinyemi2021;@Peng2017]. While meteorological droughts are responsible for approximately 37% of land cover change and variability in vegetation productivity globally [@Peng2017], there is little support for the idea that meteorological droughts affect soil moisture[@Chen2022]. However, the evidence supporting these results is derived from only one drought index, Standardized Precipitation Evapotranspiration Index [SPEI, @Vicente-Serrano2010], which combines a proxy for water supply -  precipitation - with a proxy for water demand - AED - at one time scale (12 months). The use of only one time scale may bias results of drought impacts towards ecosystems dominated by plant growth forms such as grasses and herbs that respond more rapidly to drought stress (< 12 months). This is because  physiological differences among and within dominant plant growth forms may increase (or decrease) tolerance of drought stress [@Craine2013;@McDowell2022]. For example, trees growing in more arid ecosystems typically respond over longer time scales than those in more humid ecosystems [@Vicente-Serrano2014]..

Expanding analyses to include multiple dimensions of droughts can provide complementary insights into the Earth’s water balance - and its impacts - over multiple time scales. Yet, the World Meteorological Organization recommends the use of a single drought index for monitoring droughts [@WMO2012], i.e., the multi-scale Standardized Precipitation Index [SPI; @McKee1993], but is limited in that it only looks at  water supply in the form of precipitation. The SPEI builds upon SPI by incorporating the effects of  temperature on droughts, and is now used widely for drought monitoring [e.g., @Gebrechorkos2023;@Liu2024]. To better disentangle the effects of precipitation from those of temperature [@Vicente-Serrano2020], as well as to capture droughts in terms of water demand, AED has been integrated into the Evaporative Demand Drought Index [EDDI, @McEvoy2016], which is particularly effective at detecting the rapid onset or intensification of droughts. Indices derived from soil moisture products, such as  the Soil Moisture Deficit Index [SDMI, @Narasimhan2005], the Soil Moisture Agricultural Drought Index [SMADI, @Souza2021], and the Standardized Soil Moisture Index [SSI, @AghaKouchak2014;@AghaKouchak2015] also monitor water supply and are thought to better capture water availability for crops, thus providing more relevant information for evaluating agricultural droughts. In turn, ecological droughts, which capture the joint impacts of precipitation and temperature on natural and productive ecosystems via variation in net primary productivity  [@Camps-Valls2021;@Paruelo2016;@Helman2014], are usually monitored with the Normalized Difference Vegetation Index (NDVI) and derived anomaly indices, e.g., zcNDVI [@Zambrano2018]. However, none of the aforementioned drought indices directly or indirectly consider the broader impacts of droughts on human decisions and activities - particularly land-use change, which is critical to developing a more holistic view of climate change impacts.

Here, we analyze the multi-dimensional impacts of drought on water supply and demand, net primary productivity, and land-use change across terrestrial ecosystems in continental Chile. Chile's diverse climate and ecosystems [@Beck2023;@Luebert2022] make it an ideal natural laboratory for assessing the dynamic interactions between climate and ecosystems, and potential impacts on land-use change. Additionally, large parts of Chile have experienced severe drought conditions that have significantly affected vegetation and water storage in recent years; north-central Chile has faced a persistent precipitation deficit (or “mega-drought”) since 2010 [@Garreaud2017], which has broadly impacted native forests [e.g., @Miranda2020;@UrrutiaJalabert2018;@Venegas2018]  and agricultural productivity [e.g., @Zambrano2016;@Zambrano2018;@Zambrano2023]. There is also growing evidence that this “mega-drought” has impacted farmers' decision making, shifting to crop systems with shorter rotations and lower capital costs [@Zuniga2021]. Given the persistent water deficit associated with the "mega-drought" and its cascading effects on the hydrological system [@Boisier2018], it is critical to assess multiple time scales that account for the cumulative impacts of this extreme event over several years. We therefore aim to assess: i) short- to long-term time trends in multi-scalar drought indices that capture variation in the components of water balance, i.e., water supply and demand; ii) temporal changes in land-use cover and vegetation productivity, and iii) drought impacts on vegetation productivity and land-use change across continental Chile.

Here, we analyze the multi-dimensional impacts of drought on water supply and demand, net primary productivity, and land-use change across terrestrial ecosystems in continental Chile. Chile's diverse climate and ecosystems [@Beck2023;@Luebert2022] make it an ideal natural laboratory for assessing the dynamic interactions between climate and ecosystems, and potential impacts on land-use change. Additionally, large parts of Chile have experienced severe droughts conditions that have significantly affected vegetation and water storage in recent years; north-central Chile has faced a persistent precipitation deficit (or "mega-drought") since 2010 [@Garreaud2017], which has broadly impacted native forests [e.g., @Miranda2020;@UrrutiaJalabert2018;@Venegas2018] and agricultural productivity [e.g., @Zambrano2016;@Zambrano2018;@Zambrano2023]. There is also growing evidence that this "mega-drought" has impacted farmers' decision making, who now opt for crops with shorter rotations and lower capital costs [@Zuniga2021]. Given the persistent water deficit associated with the "mega-drought" and its cascading effects on the hydrological system [@Boisier2018], it is critical to assess multiple time scales that account for the cumulative impacts of this extreme event over several years. We therefore aim to assess: i) short- to long-term time trends in multi-scalar drought indices that capture variation in the components of water balance, i.e., water supply and demand; ii) temporal changes in land-use cover and vegetation productivity, and iii) drought impacts on vegetation productivity and land-use change across continental Chile.

# Results

## Decreases in water supply and increases in water demand strengthen over longer time scales

We observed  a decrease in SPI, SPEI, and SSI - proxies largely associated with water supply - from north to south in continental Chile, with the exception of the southernmost region ("Austral"), a trend that became more pronounced over longer time scales (@fig-trendDIMacro). In contrast, we found that EDDI - a proxy for atmospheric water demand - showed a positive trend across Chile, with a sharper increase over time scales in the north than in the south.  In general, these results suggest that declines in precipitation have reduced water supply, while increases in temperature have increased water demand over the past four decades.

```{r fig-trendDIMacro}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "**Drought severity increases over longer time scales across most of continental Chile.** We evaluated temporal shifts in drought severity over multiple time scales for indices associated with water supply (SPI, SPEI,  SSI) and demand (EDDI) across continental Chile for 1981-2023. SPI is the standardized precipitation index, SPEI is the Standardized Precipitation Evapotranspiration Index , SSI is the Standardized Soil Moisture Index, and EDDI is the Evaporative Demand Drought Index. Drought indices were aggregated per region for visualization."
#| out-width: "\\textwidth"
knitr::include_graphics('figs/trend_macrozone_drought_indices.png')
```

## Vegetation productivity decreased in northern and central Chile

Despite evidence of increasing drought severity across Chile, we found contrasting temporal trends in vegetation productivity (@fig-zcNDVI_var). In the two southernmost regions ('Sur' and 'Austral') and one northern region ('Norte Grande'), vegetation productivity increased over the last 23 years, while in two more central regions ('Centro' and 'Norte Chico') it decreased over the same period (@fig-zcNDVI_var). In central Chile, vegetation productivity was lowest from 2019 to 2022, which could be due to either a decrease in vegetation area, a loss of biomass or browning.

```{r fig-zcNDVI_var}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "**Central and northern Chile have experienced the greatest decline in vegetation productivity over the last two decades.** Spatial (a) and temporal (b) variation in vegetation productivity across continental Chile for 2000-2023. Vegetation productivity was estimated as standardized vegetation productivity (zcNDVI).  Green corresponds to areas with a positive temporal trend in zcNDVI, red corresponds to a negative temporal trend in zcNDVI, and gray corresponds to areas that did not change over time. Temporal trends in zcNDVI were estimated with the non-parametric modified Mann-Kendall test for serially correlated data.  "
#| out-width: "\\textwidth"
knitr::include_graphics("figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png")
```

## Cropland and forest cover shifting southwards

During the same period, we also observed significant changes in land cover across continental Chile (@fig-temp_var_landcover).  In northern Chile ("Norte Grande" and "Norte Chico"), the cover of croplands (-12 $km^2 yr^{-1}$) and savannas (-70 $km^2 yr^{-1}$) decreased, while the cover of barren lands increased significantly (111 $km^2 yr^{-1}$) and the cover of forests, grasslands and shrublands did not change (0 $km^2 yr^{-1}$). In central Chile ("Centro"), croplands (-22 $km^2 yr^{-1}$) and savannas (-136 $km^2 yr^{-1}$) experienced a strong decrease in cover, but shrublands (146 $km^2 yr^{-1}$), grasslands (83 $km^2 yr^{-1}$), and barren lands (23 $km^2 yr^{-1}$) increased, and forests did not change (0 $km^2 yr^{-1}$). In contrast, in southern Chile ("Sur"), forest cover (397 $km^2 yr^{-1}$) and cropland cover (38 $km^2 yr^{-1}$) increased over time, with only savanna cover decreasing (-319 $km^2 yr^{-1}$). In the southernmost region ("Austral"), only savanna cover increased (172 $km^2 yr^{-1}$), while barren land (-93 $km^2 yr^{-1}$) and shrubland (-37 $km^2 yr^{-1}$) cover decreased. These changes in land cover suggest that agricultural cover is shifting further south, from northern and central Chile to southern Chile, where savannas are apparently being rapidly replaced by native and planted forests.

```{r fig-temp_var_landcover}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "**Land cover is shifting dynamically across continental Chile.** Temporal trends in absolute (a) and relative (b) land cover across continental Chile for 2001-2022. Temporal change in land cover for each class was estimated with Sen’s  slope; zero values indicate no change, while red and blue points indicate maximum and minimum values, respectively. Land cover classes with no values did not have statistically significant changes in area over the study period. Relative land cover change was estimated within each study region."
#| out-width: "\\textwidth"
knitr::include_graphics("figs/temporal_variation_and_trend_landcover_macroclass_2001-2022.png")
```

## Vegetation productivity most strongly impacted by drought in south-central Chile

We found that temporal variation in vegetation productivity was usually best explained by drought indices with time scales greater than 12 months (@fig-map_cor_r_indices). For all drought indices, the time scales with the strongest correlation with vegetation productivity were longer towards northern Chile and shorter towards southern Chile, with the exception of the southernmost region ("Austral"). Especially in south-central Chile ("Centro" and "Sur"), the time scales with the strongest correlation with vegetation productivity were concentrated in the Coastal and Andean mountain ranges.  However, the areas where vegetation was most affected by drought, i.e. where correlations were positive for SPI, SPEI and SSI and negative for EDDI, were located in south-central Chile, but not necessarily in either of the two mountain ranges. While the spatial variation in the relationship between drought intensity and vegetation productivity was consistent across drought indices, the drought index that captures water supply via soil moisture (Standardized Soil Moisture Index; SSI) tended to show a stronger correlation with vegetation productivity over larger areas than the other drought indices.

Our analysis also revealed that water demand and supply differentially affected the time scales at which vegetation productivity of land cover types within each region was most impacted by drought (@fig-map_selec_time_scales_indices /Table SSX). In northern Chile, all land cover types exhibited stronger correlations with drought indices associated with water supply, i.e. SPI, SPEI, and SSI, at shorter time scales (12 or 14 months) than those associated with water demand, AED (36 months). In central Chile, we observed a similar pattern for shrublands and savannas, and found that vegetation productivity of shrublands, savannas, and croplands were generally more affected by changes in water supply than grasslands, croplands, or forests. In southern Chile, vegetation productivity within land cover types was less affected by variation in water supply or demand - and at shorter timescales than in other regions. Notably, vegetation productivity of native and planted forests was weakly correlated with drought indices (r < 0.2) at relatively long time scales, particularly in central Chile.

```{r fig-map_cor_r_indices}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "**Drought impacts on vegetation productivity shift from north to south across continental Chile.** Pearson's correlation coefficient was used to estimate the direction and magnitude of the relationship between drought severity and vegetation productivity for each index. We show Pearson correlation coefficients for the time scale (3 - 36 months) at which they reach their maximum value. Areas in white indicate no statistically significant correlation. "
#| out-width: "\\textwidth"
knitr::include_graphics("figs/mapa_cor_r_indices_zcNDVI6.png")
```

```{r fig-map_selec_time_scales_indices}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "**Drought impacts on vegetation productivity are higher over longer time scales across continental Chile.** Spatial variation in the time scale (3-36 months) at which drought impacts on vegetation productivity are most severe across continental Chile.  White spaces indicate no significant correlation between vegetation productivity and drought severity."
#| out-width: "\\textwidth"
knitr::include_graphics("figs/mapa_cor_selec_indices_zcNDVI6.png")
```

## Drought transforms land cover distribution

Our random forest models show that drought indices explain between 22-48% of the variation in land cover change across continental Chile, with the exception of croplands whose variation was weakly affected by drought (@fig-rel_import_RF; 11-20%). Moreover, these results highlight the importance of considering water supply and demand, as drought indices associated with both aspects of water balance had high importance values across most study regions and land cover types. The variation in the time scale of drought indices within study regions also suggests that different types of vegetation are not equally sensitive to droughts of similar intensities. For example, changes in savanna and shrubland cover were associated with longer time scales in most regions, while changes in forest cover in central and southern Chile were associated with shorter time scales. Our results also show that drought severity was associated with the magnitude and direction of land cover change (@fig-parcial_variation). More specifically, we found that decreases in precipitation (SPI-6) and soil moisture (SSI-36) and increases in atmospheric evaporative demand (EDDI-6 and EDDI-36) at multiple time scales are associated with non-linear decreases in grassland across continental Chile and forest cover from central to southern Chile. In contrast, shrubland increased non-linearly in response to decreases in precipitation (SPI-6 and SPI-36) and soil moisture (SSI-6 and SSI–36) and increases in atmospheric evaporative demand (EDDI-6 and EDDI-36) across central and northern Chile. Savanna cover responded weakly to changes in precipitation across continental Chile, but exhibited more pronounced non-linear declines in response to increasing atmospheric evaporative demand (EDDI-6 and EDDI-36) across most study regions. Cropland cover, not surprisingly, varied weakly in response to changes in either water supply or demand.

```{r fig-rel_import_RF}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "**Shifts in water supply and demand drive land cover change across multiple time scales.** Variable importance of multi-scalar drought indices for explaining land cover change in five study regions across continental Chile. Variable importance was estimated with Random Forest models fitted for each combination of study region and land cover type. SPI is the Standardized Precipitation index, SPEI is the Standardized Precipitation Evapotranspiration Index , SSI is the Standardized Soil Moisture Index, and EDDI is the Evaporative Demand Drought Index. The numbers next to the drought index correspond to the time scales in months (1- 36)"
#| out-width: "\\textwidth"
knitr::include_graphics("figs/bars_relative_importance_RF.png")
```

```{r fig-parcial_variation}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "**Drought severity drives land cover change, but not for all cover types.** Response of land cover change in response to water demand and supply across multiple time scales and study regions in continental Chile. SPI is the standardized precipitation index, SPEI is the Standardized Precipitation Evapotranspiration Index, SSI is the Standardized Soil Moisture Index, and EDDI is the Evaporative Demand Drought Index.  Numbers next to the drought index correspond to the time scales in months (1- 36). Fitted lines are smoothed response curves across river basins in each region estimated with Random Forest models."
#| out-width: "\\textwidth"
knitr::include_graphics("figs/RF_most_important_variables_trend_LC_vs_trend_DI.png")
```

# Discussion

## Temporal trends in water supply and demand

With the exception of the southernmost region, we found a significant decreasing trend in water supply (SPI, SPEI, and SSI) over the past four decades across continental Chile and is strongest in northern and central Chile [@Boisier2018;@Sarricolea2019]. Our results reveal that decreases in water supply increased over longer time scales, which is consistent with a progressive intensification of drought severity across much of Chile, as has been observed in other regions experiencing long-term droughts[@Rashid2019;@Miro2023]. In parallel, we observed an increased water demand (EDDI) due to rising air temperatures, which also strengthened over longer time scales. Taken together, our results provide multiple lines of evidence that continental Chile has experienced a sustained drying trend due to a concurrent decrease in precipitation and increase in atmospheric evaporative demand [@Pascoa2021].

With the exception of the southernmost region, we found a significant decreasing trend in water supply (SPI, SPEI, and SSI) over the past four decades across continental Chile and is strongest in northern and central Chile [@Boisier2018;@Sarricolea2019]. Our results reveal that decreases in water supply increased over longer time scales, which is consistent with a progressive intensification of drought severity across much of Chile, as has been observed in other regions experiencing long-term droughts [@Rashid2019;@Miro2023]. In parallel, we observed an increased water demand (EDDI) due to rising air temperatures, which also strengthened over longer time scales. Taken together, our results provide multiple lines of evidence that continental Chile has experienced a sustained drying trend due to a concurrent decrease in precipitation and increase in atmospheric evaporative demand [@Pascoa2021].

## Temporal trends in vegetation productivity

The consequences of the persistent drying trend for ecosystems throughout continental Chile are manifold. First, the prolonged hydrological drought, i.e. precipitation deficit, has reduced groundwater storage  [SSI, @Taucare2024], leading to a steady decline in vegetation productivity (zcNDVI) since 2000 across northern and central Chile, reaching its lowest level between 2020 and 2022. This decline was most strongly associated with declines in soil moisture, as has been reported for natural and productive ecosystems [@Nicolai-Shaw2017;@Jiang2020;@Zhou2021]. Second, the strong coupling between vegetation productivity and soil moisture over longer time scales [@Bonan2008] that we observed provides a more direct physiological explanation for the sharp decline in forest growth and productivity in central Chile [e.g., @Miranda2023;@Venegas2022], as the dominant woody vegetation (e.g., trees, shrubs) in this region is likely to obtain water from deeper in the soil profile than herbs, grasses, or agricultural crops[@Oliviera2005]. Moreover, the strengthening of the correlation between vegetation productivity and water supply (SPI, SPEI, SSI) or demand (EDDI) over multiple time scales (up to 36 months) and across land cover types (@fig-map_selec_time_scales_indices) - demonstrates the impacts of climate change on the water balance in Chile. Impacts may extend beyond vegetation productivity, as reduced soil moisture in central Chile and the western United States has increased wildfire activity [@Holden2018;@Gonzalez2018], which is a growing concern in Chile and may be further exacerbated by extensive plantations of highly flammable tree species, e.g., Eucalyptus spp. and Pinus spp.[@Bowman2019]. Third, we found that the decline in the vegetation productivity of croplands is due to a decrease in the water supply to a greater extent than to an increase in water demand [@Quiring2010], despite evidence that more water-intensive crops have replaced less water-intensive crops in central Chile, leading to an increase in water extraction from rivers or groundwater [@Munoz2020;@Duran2020].

## Drought impacts  on land cover

We found evidence that temporal decreases in water supply and decreases in water demand are driving shifts not only in vegetation productivity but also in land cover across most of continental Chile. Forest and grassland cover were particularly sensitive to changes in the water balance over short and long temporal scales, which is consistent with recent studies showing that progressive, long-term water deficits in central Chile have triggered forest browning and declines in native forest productivity [@Miranda2020;@Venegas2022;@Miranda2023]. While our analysis do not distinguish between  native and planted forests, the latter of which are considered to be more drought tolerant in central and southern Chile [@Carrasco2022], we show that forest area declines more sharply in response to increasing water demand due to rising temperatures (EDDI) than decreasing water supply [e.g., SPI, SSI, @Fajardo2019;@Holz2018], which may have cascading impacts on multiple facets of forest diversity [@Segovia2021;@Sabatini2022]. Our results extend the results of these studies by showing that drought-induced forest cover decline has extended beyond central Chile to the southernmost region of continental Chile. This is noteworthy because declines in vegetation productivity in southern Chile - a region whose water balance is typically projected to be less affected by climate change than central and northern Chile [@Breda2020] -  have only manifested since 2022 (@fig-zcNDVI_var). Moreover, our results provide evidence that, in addition to forest cover, other land cover types have been affected by water deficits, particularly grasslands, despite physiological differences between dominant plant growth forms [e.g., trees, shrubs, C3 and C4 grasses, @Craine2013;@McDowell2022]. Our results therefore suggest that multiple land cover types could be vulnerable to regime shifts towards more drought tolerant land cover types [@Scheffer2001;@Martinez-Vilalta2016], such as shrublands, whose cover increased non-linearly in response to increasing drought severity. In central Chile, for example, the increase in shrubland cover could be due to drought-induced decreases in savanna or cropland cover. Changes in cropland cover may not be a direct consequence of drought (@fig-parcial_variation), but rather an indirect one, possibly reflecting the decision of resource-poor farmers to migrate to regions with more abundant water resources or to change economic activity [@AghaKouchak2021;@Hermans2021]. In contrast, the increase in shrubland cover due to a decrease in savanna cover may be ecological, as shrubs may  be more drought tolerant than other growth forms  [@Eldridge2011;@Gotmark2016].

Overall, our results show that long-term declines in water supply and demand have induced widespread, multi-dimensional impacts on the vegetation productivity and on the extent of land cover types. While prolonged droughts may directly cause shifts to more drought-tolerant land cover types, such as shrublands, they may also influence land cover change through human decision making and activities. This study extends current understanding of drought impacts by demonstrating how their multidimensionality emerges over multiple time scales and across land cover types, which can contribute to developing context-specific adaptation strategies for agriculture, biodiversity conservation, and natural resource management.

# Materials and Methods

## Study area

Continental Chile has a diverse climate, with strong gradients from north to south and east to west [@Aceituno2021] (@fig-studyArea a), which, together with its complex topography, determine its ecosystem diversity [@Garreaud2009;@Luebert2022] (@fig-studyArea c).  We divided Chile into five regions: "Norte Grande" (17°34'–25°42'S), "Norte Chico" (25°42'-32°8'S), "Centro" (32°08'-36°12'S), "Sur" (36°12'-43°48'S), and "Austral" (43°48'-56°00'S). "Norte Grande" and "Norte Chico" are predominantly arid with hot (Bwh in the Koppen-Geiger classification) and cold (Bwk) temperatures. Towards the south of "Norte Chico", the climate changes to an arid steppe with cold temperatures (Bsk). In these two northern regions, the land is mostly bare, with a small area covered by shrublands and grasslands. In the "Centro" region and the northern half of "Sur",  the climate is mostly Mediterranean, with warm to hot summers (Csa and Csb). Land cover in the "Centro" region consists of a significant amount of shrublands and savannas (50%), followed by grasslands (16%), forests (8%), and croplands (5%). The south of "Sur" and the north of the "Austral" region have a mostly oceanic climate (Cfb). Those zones have a large area of forests and grasslands. The southern part of the country has a tundra climate, while "Austral" is a cold, semi-arid area covered by grasslands and  forests, and, to a lesser extent, savannas.

```{r fig-studyArea}
#| echo: false
#| fig-pos: "!ht"
#| fig-env: 'figure*'
#| fig-cap: '(a) Chile with the Koppen-Geiger climate classes and the five macrozones “Norte Grande”, “Norte Chico”, “Centro”, “Sur”, and “Austral”. (b) Topography reference map. (c) land cover classes for 2022. (d) Persistent land cover classes (> 80%) for 2001-2022.'
#| out-width: "\\textwidth"
knitr::include_graphics("figs/map_study_con_landcover.png")
```

## Data

### Gridded meteorological and vegetation data

To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the product MOD13A3 Collection 6.1 from MODIS . MOD13A3 provides vegetation indices with a 1 km spatial resolution and monthly frequency [@Didan2015]. 
For soil moisture, water supply, and water demand variables, we used ERA5-Land (ERA5L) (ECMWF Reanalysis version 5 over land) [@MunozSabater2021], a reanalysis dataset that provides atmospheric and land variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage. We selected total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100 cm of depth (layer 1 to layer 3; Table SSX).

## Short- to long-term drought trends

### Atmospheric Evaporative Demand (AED)

To compute drought indices that use water demand, it is necessary to first calculate AED. To do this, we employed the Hargreaves method [@Hargreaves1994;@Hargreaves1985] by applying the following equation:

$$AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}$$ {#eq-AED}

where $Ra$ $(MJ\,m^2\, day^{-1})$ is extraterrestrial radiation; $T$, $T_{max}$, and $T_{min}$ are mean, maximum, and minimum temperature $(°C)$ at 2m. For calculating $Ra$ we used the coordinate of the latitud of the centroid of each pixel as follow:

$$R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]$$ {#eq-Ra}

where:

$Ra$: extraterrestrial radiation $[MJ\, m^{-2} day-1]$,  
$G_{sc}$: solar constant = 0.0820 $[MJ\,m^{-2} min^{-1}]$,    
$d_r$: inverse relative distance Earth-Sun,  
$\omega_s$ sunset hour angle  $[rad]$,  
$\phi$: latitude $[rad]$,  
$\delta$: solar declination $[rad]$.

We selected the method of Hargreaves to estimate AED because of its simplicity, as it only requires temperature and extraterrestrial radiation, and because access to the data needed for alternative methods (e.g., Penman-Montieth) is often limited [@Vicente-Serrano2014].

###  Drought indices

To derive the drought indices of water supply and demand we used the ERA5L dataset and the MODIS product (MOD13A3[@Didan2015b]), with a monthly frequency for 1981–2023 and 2000–2023, respectively. Drought indices capture historical anomalies of water supply and demand. To quantify each anomaly, the common practice is to derive it following a statistical parametric method in which it is assumed that the statistical distribution of the data is known [@Heim2002]. The use of an erroneous statistical distribution that does not fit the data is usually the highest source of uncertainty [@Laimighofer2022]. In the case of Chile, due to its high degree of climatic variability, it is difficult to choose a statistical distribution that can be used across its entire extent. We therefore use a non-parametric method for the calculation of the drought indices, following @Farahmand2015.

For monitoring water supply, we used the Standardized Precipitation Index (SPI; @McKee1993), which relies on precipitation data. To evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI; @Hobbins2016 and  @McEvoy2016), which is based on the AED. To consider the combined effect of water supply and demand, we selected the SPEI [@Vicente-Serrano2010]. For SPEI, an auxiliary variable $D=P−AED$ is calculated. Soil moisture is the main driver of vegetation productivity, particularly in semi-arid regions [@Li2022]. Hence, we used the Standardized Soil Moisture Index (SSI) to monitor soil moisture (SM) [@Hao2013]. For the SSI, we used the average soil moisture from ERA5L at a depth of 1m. All calculated indices are multi-scalar and can be used for the analysis of short- to long-term droughts.

To derive the drought indices, we first calculate the sum of the variables with regard to the time scale(s). In this case, for generalization purposes, we will use $V$, referring to variables P, AED, D,  and SM (Table  SSX). We accumulated each over the time series of values (months), and for the time scales s:

$$A^s_i = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  $$ {#eq-sumvar}

The $A^s_i$ corresponds to a moving window (convolution) that sums the variable for time scales $s$. This is summed over $s$ months, starting from the most recent month ($n$) back in time until month $n-s+1$. For example, using as a variable the precipitation, a period of twelve months ($n$), and a time scale of three months ($s$):

$$
\begin{split}
A^3_1 &= P_{oct} +P_{nov} +P_{dic} \\
\vdots\,\,\, &= \,\,\,\vdots + \,\,\,\vdots + \,\,\,\vdots \\
A^3_{10} &= P_{jan}+P_{feb} +P_{mar}
\end{split}
$$

Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i^s$ to derive the $P(A_i^s)$ probabilities across a period of interest:

$$P(A^s_i) = \frac{i-0.33}{n+0.33'}$$ {#eq-probPai}

An inverse normal approximation [@Abramowitz1968] obtains the empirically derived probabilities once the variable cumulates over time for the scale $s$. Thus, the drought indices $SPI$, $SPEI$, $EDDI$, and $SSI$ are  obtained following the equation:

$$DI(A^s_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}$$ {#eq-DI}

$DI$ is referring to the drought index calculated for the variable $V$. The values for the constats are: $C_0 = 2.515517$, $C_1 = 0.802853$, $C_2 = 0.010328$, $d_1 = 1.432788$, $d_2 = 0.189269$, and $d3 = 0.001308$. For $P(A^s_i) \leq 0.5$, W=$\sqrt{-2\cdot ln(P(A^s_i))}$ , and for $P(A^s_i) > 0.5$, replace $P(A^s_i)$ with $1-P(A^s_i)$ and reverse the sign of $DI(A^s_i)$.

The drought indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months at a monthly frequency for 1981–2023.

###  Temporal trends of drought indices

To determine if there are statistically significant positive or negative temporal trends for the drought indices, we used the non-parametric modified Mann-Kendall test for serially correlated data [@Yue2004]. To determine the magnitude of the trend, we used Sen's slope [@Sen1968]. Sen's slope  is less affected by outliers than parametric ordinary least squares regression, and as a non-parametric method it is not influenced by the distribution of the data. We applied both methods for SPI, EDDI, SPEI, and SSI and six time scales, resulting in a total of 24 trends. We then aggregated temporal trends for each region and land cover type.

## Vegetation productivity

We also used the MODIS product to calculate vegetation productivity, and calculated anomalies in NDVI using  zcNDVI [@Zambrano2018], which was derived from the monthly time series of NDVI, with Equations @eq-sumvar and @eq-DI. For vegetation productivity, we selected the time scale that best correlates with annual net primary productivity (NPP) across continental Chile. For this purpose, we calculated zcNDVI for time scales of 1, 3, 6, and 12 months (from December) and compared it with the annual NPP. We obtained NPP from MOD17A3HGF [@Running2019]. We chose to use six months because the $R^2$ between zcNDVI and NPP reaches its highest value at six months, obtaining an $R^2$ of 0.31 for forest and 0.72 for shrubland (Supplementary Information Section S5). We subsequently used zcNDVI with a time scale of 6 months and calculated it at a monthly frequency for 2000–2023.

### Drought impacts on vegetation productivity

For each land cover, we analyzed the trend of vegetation productivity. To this end, we identified areas within each land cover macro-class that are persistent over time, to reduce the possibility that trends in vegetation productivity may be influenced by changes in  land cover. We examined the correlation between drought indices and vegetation productivity across land cover types to determine to the extent to which soil moisture and water demand and supply affect vegetation productivity.

We estimated pixel-to-pixel Pearson's correlations between drought indices at time scales of 1, 3, 6, 12, 24, and 36 months with zcNDVI.  We extracted the Pearson correlation coefficient corresponding to the time scale with the highest value. For each index, we then generated two raster maps: 1) a raster with values of the time scales and drought index that reached the maximum correlation, and 2) a raster with the magnitude of the correlation between the drought index and vegetation productivity.

## Drought impacts on land cover change

### Land cover change

To analyze land cover change, we used the classification scheme of the International Geosphere-Biosphere Programme (IGBP) from the product MCD12Q1 Collection 6.1 from MODIS. The MCD12Q1 product is produced for each year from 2001 to 2022 and defines 17 classes (see Table Sx). Following the FAO classification [@FAO2022], we considered native and planted forests as "forests", which represent natural and productive ecosystems dominated by large trees.  To analyze the land cover change, we use the IGBP scheme from the MCD12Q1 Collection 6.1 from MODIS. We regrouped the 17 classes into ten macro-classes, as follows: 1-4 to forests (native forest and plantations), 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 as water (Table S1). This resulted in a time series of land cover with ten macro-classes for 2001 and 2023. We validated the land cover macro-classes using a high resolution (30m ) land cover map for 2013-2014 [@Zhao2016]. Our results showed a global accuracy of ~0.82 and a F1 score of ~0.66 (Supplementary Information, S2).

We calculated the area for each land cover class in the five study regions for 2001–2022. We then estimated the temporal change in area for each land cover type and macro-class, and determined the statistical significance and magnitude of the trend as described above.
To assess how water demand and supply, and soil moisture affect the variation in vegetation productivity across various land cover types, we avoid analyzing areas that experienced major land cover changes in the 2001–2022 period. To assess how zcNDVI varied irrespective of land cover change, we developed a persistence mask for land cover, which only retains pixels for which the macro-class remained the same for at least 80% of the 22 years (@fig-studyArea d).

### Relationship between land cover and drought trends

To identify which drought indices and time scales have a major impact on changes in land cover type, we examined the relationship between the trend in land cover classes and the trend in drought indices. We performed the analysis at the sub-basin scale, using 469 basins, which have a surface area between 0.0746 and 24,000 $km^2$ and a median area of 1,249 $km^2$. For each basin, we calculated the trend per land cover type, considering the proportion of the type relative to the total surface of the basin. For each basin we extracted the average trend of all drought indices and at time scales of 1, 3, 6, 12, 24, and 36 months. Also, we extracted the average trend in zcNDVI.

We modeled trends in land cover type per macroclass with the aim of assessing how land cover trends relate to drought indices. We used the random forest method [@Ho1995], which employs multiple decision trees, allowing for classification and regression. Some advantages include the ability to find non-linear relationships, reduce overfitting, and derive variable importance. We included the four drought indices at each time scale and zcNDVI for a total of 25 predictors and built six random forest models, one for each land cover and  region. We trained each model with 1000 trees  using a resampling strategy with cross-validation. To this end, we used cross-validation to evaluate model fit using ten folds then calculated $R^2$, root mean square error (RMSE), and variable importance. Variable importance identifies which variables have a higher contribution to explaining model variation. We calculated variable importance by permuting out-of-bag (OOB) data per tree and calculating the mean standard error of the OOB data. After permuting each predictor variable, we repeated the process for the remaining variables. We repeated this process ten times per fold to assess  model fit.
Finally, we visually explored the relationship between drought indices and changes in land cover. To do this, we compared the relative changes in land cover surface with the drought indices of six and thirty-six months.

## Software

For the downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, R [@R2023]. For downloading ERA5L, we used the {ecmwfr} package [@Hufkens2019]. For processing raster data, we used {terra} [@Hijmans2023] and {stars} [@Pebesma2023]. For managing vectorial data, we used {sf} [@Pebesma2018]. For the calculation of AED, we used {SPEI} [@Bergueria2023]. For mapping, we used {tmap} [@Tennekes2018]. For data analysis and visualization, the suite {tidyverse} [@Wickham2019] was used. For the random forest modeling, we used the {tidymodels} [@Kuhn2020] and {ranger} [@Wright2017] packages.

# Acknowledgments {-}

The National Research and Development Agency of Chile (ANID) funded this study through the drought emergency project FSEQ210022, Fondecyt Iniciación N°11190360, Fondecyt Postdoctorado N°3230678, and Fondecyt Regular N°1210526.

# References
<!-- the pattern below controls the placement of the references -->
:::{#refs}
:::

### 2024-06-25 article/article_nature/article_nature.tex
actualizar repositorio0+

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sn-nature,
  numbered]{sn-jnl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

%%%% Standard Packages

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%

%%%%

\raggedbottom
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Shifts in water supply and demand drive land cover change across Chile},
  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
  pdfkeywords={drought, land cover, water demand, water supply},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Shifts in water supply and demand drive land cover change across
Chile]{Shifts in water supply and demand drive land cover change across
Chile}

% author setup
\author*[1,2]{\fnm{Francisco} \sur{Zambrano}}\email{francisco.zambrano@umayor.cl}\author[3]{\fnm{Anton} \sur{Vrieling}}\author[4,5,6]{\fnm{Francisco} \sur{Meza}}\author[7]{\fnm{Iongel} \sur{Duran-Llacer}}\author[8,9]{\fnm{Francisco} \sur{Fernández}}\author[10]{\fnm{Alejandro} \sur{Venegas-González}}\author[4]{\fnm{Nicolas} \sur{Raab}}\author[11,12]{\fnm{Dylan} \sur{Craven}}
% affil setup
\affil[1]{, \orgname{Hémera Centro de Observación de la Tierra, Facultad
de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad,
Universidad Mayor}}
\affil[2]{, \orgname{Observatorio de Sequía para la Agricultura y la
Biodiversidad de Chile (ODES), Universidad Mayor}}
\affil[3]{, \orgname{Faculty of Geo-Information Science and Earth,
University of Twente}}
\affil[4]{, \orgname{Facultad de Agronomía y Sistemas Naturales,
Pontificia Universidad Católica de Chile.}}
\affil[5]{, \orgname{Instituto para el Desarrollo Sustentable.
Pontificia Universidad Católica de Chile}}
\affil[6]{, \orgname{Centro Interdisciplinario de Cambio Global,
Pontificia Universidad Católica de Chile}}
\affil[7]{, \orgname{Hémera Centro de Observación de la Tierra, Facultad
de Ciencias, Universidad Mayor,}}
\affil[8]{, \orgname{Center of Economics for Sustainable Development
(CEDES), Faculty of Economics and Government, Universidad San
Sebastian}}
\affil[9]{, \orgname{Center of Applied Ecology and Sustainability
(CAPES)}}
\affil[10]{, \orgname{Instituto de Ciencias Agroalimentarias, Animales y
Ambientales (ICA3), Universidad de O'Higgins}}
\affil[4]{}
\affil[11]{\orgdiv{GEMA Center for Genomics, Ecology \& Environment,
Universidad Mayor, Camino La Pirámide Huechuraba 5750}}
\affil[12]{, \orgname{Data Observatory Foundation}}

% abstract

\abstract{Globally, droughts are becoming longer, more frequent and more
severe, and their impacts are multidimensional. These impacts typically
extend beyond the water balance as long-term, cumulative changes in the
water balance can lead to regime shifts in land use. Here, we assess the
effects of temporal changes in water supply and demand on vegetation
productivity and land cover change over multiple time scales in
continental Chile, which has experienced a severe drought over the last
20 years. Across most of continental Chile, we found a persistent
decreasing trend in water supply and an increasing trend in water demand
since 1981, trends that intensify over longer time scales. This
long-term decrease in water availability has led to a decrease in
vegetation productivity, especially in central and southern Chile. Our
models suggest that increasing drought severity has led to shifts in
land use towards more drought-tolerant land cover types, such as
shrublands. We also found evidence that human perceptions of prolonged
drought can indirectly lead to large-scale changes in land use. Our
results suggest that long-term climate change may lead to regime shifts
in land cover that can be mitigated by context-specific adaptation
strategies.}

% keywords
\keywords{drought,  land cover,  water demand,  water supply}

\begin{document}
\maketitle

\section{Introduction}\label{sec-intro}

Across many regions of the world, droughts are becoming longer and more
frequent and severe \citep{Miranda2023, IPCC2023}, impacting ecosystems
via tree mortality\citep{Cheng2024} and productivity \citep{Miranda2023}
and inducing shifts in land cover and use \citep{Crausbay2017}. However,
identifying drought events is surprisingly idiosyncratic due to the
varying criteria used for classification. Droughts can be classified as
either 1) meteorological, i.e., when precipitation in a specific period
falls below mean precipitation values observed over multiple years
(usually more than 30 years); 2) hydrological, i.e., when precipitation
anomalies last for long periods (months to years) and affect water
systems; 3) agricultural, i.e.~when precipitation deficits negatively
impact plant health, leading to decreases in crop or pasture
productivity \citep{Wilhite1985}; or 4) ecological, i.e., when
precipitation deficits negatively affect the provisioning of ecosystem
services and trigger feedbacks in natural or human systems
\citep{Crausbay2017}. Such feedbacks include drought impacts on human
decision making and activities, which can lead to land-use change
\citep{VanLoon2016, AghaKouchak2021}, which may have cascading effects
on biodiversity and ecosystem services (e.g.,
\citet{Lawler2014};\citet{Newbold2015}).

Despite the high degree of confidence in the impacts of rising
temperatures on the extent, frequency, and severity of agricultural and
ecological droughts \citep{IPCC2023}, which are likely to increase even
if global warming stabilizes at 1.5°--2°C, the severity of
meteorological droughts has been remarkably stable globally over the
past century \citep{Vicente-Serrano2022, Kogan2020}. In the few regions
where drought severity has increased over this period (1900-2000),
rising temperatures have increased atmospheric evaporative demand (AED),
which has been associated with increases in agricultural land area
\citep{Vicente-Serrano2022}. Thus, rising water demand may reflect
parallel changes in land use - primarily agriculture - that can
exacerbate the effects of meteorological droughts on ecosystems.

From 1960 to 2019, land-use change has impacted approximately one-third
of the Earth's surface, which is four times more than previously
thought\citep{Winkler2021}. Despite the considerable interest in
land-use change dynamics \citep[e.g.,][]{Song2018, Winkler2021}, the
direction and magnitude of drought impacts on land cover change and
vegetation productivity remain uncertain
\citep{Chen2022, Akinyemi2021, Peng2017}. While meteorological droughts
are responsible for approximately 37\% of land cover change and
variability in vegetation productivity globally \citep{Peng2017}, there
is little support for the idea that meteorological droughts affect soil
moisture\citep{Chen2022}. However, the evidence supporting these results
is derived from only one drought index, Standardized Precipitation
Evapotranspiration Index \citep[SPEI,][]{Vicente-Serrano2010}, which
combines a proxy for water supply - precipitation - with a proxy for
water demand - AED - at one time scale (12 months). The use of only one
time scale may bias results of drought impacts towards ecosystems
dominated by plant growth forms such as grasses and herbs that respond
more rapidly to drought stress (\textless{} 12 months). This is because
physiological differences among and within dominant plant growth forms
may increase (or decrease) tolerance of drought stress
\citep{Craine2013, McDowell2022}. For example, trees growing in more
arid ecosystems typically respond over longer time scales than those in
more humid ecosystems \citep{Vicente-Serrano2014}..

Expanding analyses to include multiple dimensions of droughts can
provide complementary insights into the Earth's water balance - and its
impacts - over multiple time scales. Yet, the World Meteorological
Organization recommends the use of a single drought index for monitoring
droughts \citep{WMO2012}, i.e., the multi-scale Standardized
Precipitation Index {[}SPI; \citet{McKee1993}{]}, but is limited in that
it only looks at water supply in the form of precipitation. The SPEI
builds upon SPI by incorporating the effects of temperature on droughts,
and is now used widely for drought monitoring
\citep[e.g.,][]{Gebrechorkos2023, Liu2024}. To better disentangle the
effects of precipitation from those of temperature
\citep{Vicente-Serrano2020}, as well as to capture droughts in terms of
water demand, AED has been integrated into the Evaporative Demand
Drought Index \citep[EDDI,][]{McEvoy2016}, which is particularly
effective at detecting the rapid onset or intensification of droughts.
Indices derived from soil moisture products, such as the Soil Moisture
Deficit Index \citep[SDMI,][]{Narasimhan2005}, the Soil Moisture
Agricultural Drought Index \citep[SMADI,][]{Souza2021}, and the
Standardized Soil Moisture Index
\citep[SSI,][]{AghaKouchak2014, AghaKouchak2015} also monitor water
supply and are thought to better capture water availability for crops,
thus providing more relevant information for evaluating agricultural
droughts. In turn, ecological droughts, which capture the joint impacts
of precipitation and temperature on natural and productive ecosystems
via variation in net primary productivity
\citep{Camps-Valls2021, Paruelo2016, Helman2014}, are usually monitored
with the Normalized Difference Vegetation Index (NDVI) and derived
anomaly indices, e.g., zcNDVI \citep{Zambrano2018}. However, none of the
aforementioned drought indices directly or indirectly consider the
broader impacts of droughts on human decisions and activities -
particularly land-use change, which is critical to developing a more
holistic view of climate change impacts.

Here, we analyze the multi-dimensional impacts of drought on water
supply and demand, net primary productivity, and land-use change across
terrestrial ecosystems in continental Chile. Chile's diverse climate and
ecosystems \citep{Beck2023, Luebert2022} make it an ideal natural
laboratory for assessing the dynamic interactions between climate and
ecosystems, and potential impacts on land-use change. Additionally,
large parts of Chile have experienced severe drought conditions that
have significantly affected vegetation and water storage in recent
years; north-central Chile has faced a persistent precipitation deficit
(or ``mega-drought'') since 2010 \citep{Garreaud2017}, which has broadly
impacted native forests
\citep[e.g.,][]{Miranda2020, UrrutiaJalabert2018, Venegas2018} and
agricultural productivity
\citep[e.g.,][]{Zambrano2016, Zambrano2018, Zambrano2023}. There is also
growing evidence that this ``mega-drought'' has impacted farmers'
decision making, shifting to crop systems with shorter rotations and
lower capital costs \citep{Zuniga2021}. Given the persistent water
deficit associated with the ``mega-drought'' and its cascading effects
on the hydrological system \citep{Boisier2018}, it is critical to assess
multiple time scales that account for the cumulative impacts of this
extreme event over several years. We therefore aim to assess: i) short-
to long-term time trends in multi-scalar drought indices that capture
variation in the components of water balance, i.e., water supply and
demand; ii) temporal changes in land-use cover and vegetation
productivity, and iii) drought impacts on vegetation productivity and
land-use change across continental Chile.

Here, we analyze the multi-dimensional impacts of drought on water
supply and demand, net primary productivity, and land-use change across
terrestrial ecosystems in continental Chile. Chile's diverse climate and
ecosystems \citep{Beck2023, Luebert2022} make it an ideal natural
laboratory for assessing the dynamic interactions between climate and
ecosystems, and potential impacts on land-use change. Additionally,
large parts of Chile have experienced severe droughts conditions that
have significantly affected vegetation and water storage in recent
years; north-central Chile has faced a persistent precipitation deficit
(or ``mega-drought'') since 2010 \citep{Garreaud2017}, which has broadly
impacted native forests
\citep[e.g.,][]{Miranda2020, UrrutiaJalabert2018, Venegas2018} and
agricultural productivity
\citep[e.g.,][]{Zambrano2016, Zambrano2018, Zambrano2023}. There is also
growing evidence that this ``mega-drought'' has impacted farmers'
decision making, who now opt for crops with shorter rotations and lower
capital costs \citep{Zuniga2021}. Given the persistent water deficit
associated with the ``mega-drought'' and its cascading effects on the
hydrological system \citep{Boisier2018}, it is critical to assess
multiple time scales that account for the cumulative impacts of this
extreme event over several years. We therefore aim to assess: i) short-
to long-term time trends in multi-scalar drought indices that capture
variation in the components of water balance, i.e., water supply and
demand; ii) temporal changes in land-use cover and vegetation
productivity, and iii) drought impacts on vegetation productivity and
land-use change across continental Chile.

\section{Results}\label{results}

\subsection{Decreases in water supply and increases in water demand
strengthen over longer time
scales}\label{decreases-in-water-supply-and-increases-in-water-demand-strengthen-over-longer-time-scales}

We observed a decrease in SPI, SPEI, and SSI - proxies largely
associated with water supply - from north to south in continental Chile,
with the exception of the southernmost region (``Austral''), a trend
that became more pronounced over longer time scales
(Figure~\ref{fig-trendDIMacro}). In contrast, we found that EDDI - a
proxy for atmospheric water demand - showed a positive trend across
Chile, with a sharper increase over time scales in the north than in the
south. In general, these results suggest that declines in precipitation
have reduced water supply, while increases in temperature have increased
water demand over the past four decades.

\begin{figure}[!ht]

\centering{

\includegraphics{figs/trend_macrozone_drought_indices.png}

}

\caption{\label{fig-trendDIMacro}\textbf{Drought severity increases over
longer time scales across most of continental Chile.} We evaluated
temporal shifts in drought severity over multiple time scales for
indices associated with water supply (SPI, SPEI, SSI) and demand (EDDI)
across continental Chile for 1981-2023. SPI is the standardized
precipitation index, SPEI is the Standardized Precipitation
Evapotranspiration Index , SSI is the Standardized Soil Moisture Index,
and EDDI is the Evaporative Demand Drought Index. Drought indices were
aggregated per region for visualization.}

\end{figure}%

\subsection{Vegetation productivity decreased in northern and central
Chile}\label{vegetation-productivity-decreased-in-northern-and-central-chile}

Despite evidence of increasing drought severity across Chile, we found
contrasting temporal trends in vegetation productivity
(Figure~\ref{fig-zcNDVI_var}). In the two southernmost regions (`Sur'
and `Austral') and one northern region (`Norte Grande'), vegetation
productivity increased over the last 23 years, while in two more central
regions (`Centro' and `Norte Chico') it decreased over the same period
(Figure~\ref{fig-zcNDVI_var}). In central Chile, vegetation productivity
was lowest from 2019 to 2022, which could be due to either a decrease in
vegetation area, a loss of biomass or browning.

\begin{figure}[!ht]

\centering{

\includegraphics{figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png}

}

\caption{\label{fig-zcNDVI_var}\textbf{Central and northern Chile have
experienced the greatest decline in vegetation productivity over the
last two decades.} Spatial (a) and temporal (b) variation in vegetation
productivity across continental Chile for 2000-2023. Vegetation
productivity was estimated as standardized vegetation productivity
(zcNDVI). Green corresponds to areas with a positive temporal trend in
zcNDVI, red corresponds to a negative temporal trend in zcNDVI, and gray
corresponds to areas that did not change over time. Temporal trends in
zcNDVI were estimated with the non-parametric modified Mann-Kendall test
for serially correlated data.}

\end{figure}%

\subsection{Cropland and forest cover shifting
southwards}\label{cropland-and-forest-cover-shifting-southwards}

During the same period, we also observed significant changes in land
cover across continental Chile (Figure~\ref{fig-temp_var_landcover}). In
northern Chile (``Norte Grande'' and ``Norte Chico''), the cover of
croplands (-12 \(km^2 yr^{-1}\)) and savannas (-70 \(km^2 yr^{-1}\))
decreased, while the cover of barren lands increased significantly (111
\(km^2 yr^{-1}\)) and the cover of forests, grasslands and shrublands
did not change (0 \(km^2 yr^{-1}\)). In central Chile (``Centro''),
croplands (-22 \(km^2 yr^{-1}\)) and savannas (-136 \(km^2 yr^{-1}\))
experienced a strong decrease in cover, but shrublands (146
\(km^2 yr^{-1}\)), grasslands (83 \(km^2 yr^{-1}\)), and barren lands
(23 \(km^2 yr^{-1}\)) increased, and forests did not change (0
\(km^2 yr^{-1}\)). In contrast, in southern Chile (``Sur''), forest
cover (397 \(km^2 yr^{-1}\)) and cropland cover (38 \(km^2 yr^{-1}\))
increased over time, with only savanna cover decreasing (-319
\(km^2 yr^{-1}\)). In the southernmost region (``Austral''), only
savanna cover increased (172 \(km^2 yr^{-1}\)), while barren land (-93
\(km^2 yr^{-1}\)) and shrubland (-37 \(km^2 yr^{-1}\)) cover decreased.
These changes in land cover suggest that agricultural cover is shifting
further south, from northern and central Chile to southern Chile, where
savannas are apparently being rapidly replaced by native and planted
forests.

\begin{figure}[!ht]

\centering{

\includegraphics{figs/temporal_variation_and_trend_landcover_macroclass_2001-2022.png}

}

\caption{\label{fig-temp_var_landcover}\textbf{Land cover is shifting
dynamically across continental Chile.} Temporal trends in absolute (a)
and relative (b) land cover across continental Chile for 2001-2022.
Temporal change in land cover for each class was estimated with Sen's
slope; zero values indicate no change, while red and blue points
indicate maximum and minimum values, respectively. Land cover classes
with no values did not have statistically significant changes in area
over the study period. Relative land cover change was estimated within
each study region.}

\end{figure}%

\subsection{Vegetation productivity most strongly impacted by drought in
south-central
Chile}\label{vegetation-productivity-most-strongly-impacted-by-drought-in-south-central-chile}

We found that temporal variation in vegetation productivity was usually
best explained by drought indices with time scales greater than 12
months (Figure~\ref{fig-map_cor_r_indices}). For all drought indices,
the time scales with the strongest correlation with vegetation
productivity were longer towards northern Chile and shorter towards
southern Chile, with the exception of the southernmost region
(``Austral''). Especially in south-central Chile (``Centro'' and
``Sur''), the time scales with the strongest correlation with vegetation
productivity were concentrated in the Coastal and Andean mountain
ranges. However, the areas where vegetation was most affected by
drought, i.e.~where correlations were positive for SPI, SPEI and SSI and
negative for EDDI, were located in south-central Chile, but not
necessarily in either of the two mountain ranges. While the spatial
variation in the relationship between drought intensity and vegetation
productivity was consistent across drought indices, the drought index
that captures water supply via soil moisture (Standardized Soil Moisture
Index; SSI) tended to show a stronger correlation with vegetation
productivity over larger areas than the other drought indices.

Our analysis also revealed that water demand and supply differentially
affected the time scales at which vegetation productivity of land cover
types within each region was most impacted by drought
(Figure~\ref{fig-map_selec_time_scales_indices} /Table SSX). In northern
Chile, all land cover types exhibited stronger correlations with drought
indices associated with water supply, i.e.~SPI, SPEI, and SSI, at
shorter time scales (12 or 14 months) than those associated with water
demand, AED (36 months). In central Chile, we observed a similar pattern
for shrublands and savannas, and found that vegetation productivity of
shrublands, savannas, and croplands were generally more affected by
changes in water supply than grasslands, croplands, or forests. In
southern Chile, vegetation productivity within land cover types was less
affected by variation in water supply or demand - and at shorter
timescales than in other regions. Notably, vegetation productivity of
native and planted forests was weakly correlated with drought indices (r
\textless{} 0.2) at relatively long time scales, particularly in central
Chile.

\begin{figure}[!ht]

\centering{

\includegraphics{figs/mapa_cor_r_indices_zcNDVI6.png}

}

\caption{\label{fig-map_cor_r_indices}\textbf{Drought impacts on
vegetation productivity shift from north to south across continental
Chile.} Pearson's correlation coefficient was used to estimate the
direction and magnitude of the relationship between drought severity and
vegetation productivity for each index. We show Pearson correlation
coefficients for the time scale (3 - 36 months) at which they reach
their maximum value. Areas in white indicate no statistically
significant correlation.}

\end{figure}%

\begin{figure}[!ht]

\centering{

\includegraphics{figs/mapa_cor_selec_indices_zcNDVI6.png}

}

\caption{\label{fig-map_selec_time_scales_indices}\textbf{Drought
impacts on vegetation productivity are higher over longer time scales
across continental Chile.} Spatial variation in the time scale (3-36
months) at which drought impacts on vegetation productivity are most
severe across continental Chile. White spaces indicate no significant
correlation between vegetation productivity and drought severity.}

\end{figure}%

\subsection{Drought transforms land cover
distribution}\label{drought-transforms-land-cover-distribution}

Our random forest models show that drought indices explain between
22-48\% of the variation in land cover change across continental Chile,
with the exception of croplands whose variation was weakly affected by
drought (Figure~\ref{fig-rel_import_RF}; 11-20\%). Moreover, these
results highlight the importance of considering water supply and demand,
as drought indices associated with both aspects of water balance had
high importance values across most study regions and land cover types.
The variation in the time scale of drought indices within study regions
also suggests that different types of vegetation are not equally
sensitive to droughts of similar intensities. For example, changes in
savanna and shrubland cover were associated with longer time scales in
most regions, while changes in forest cover in central and southern
Chile were associated with shorter time scales. Our results also show
that drought severity was associated with the magnitude and direction of
land cover change (Figure~\ref{fig-parcial_variation}). More
specifically, we found that decreases in precipitation (SPI-6) and soil
moisture (SSI-36) and increases in atmospheric evaporative demand
(EDDI-6 and EDDI-36) at multiple time scales are associated with
non-linear decreases in grassland across continental Chile and forest
cover from central to southern Chile. In contrast, shrubland increased
non-linearly in response to decreases in precipitation (SPI-6 and
SPI-36) and soil moisture (SSI-6 and SSI--36) and increases in
atmospheric evaporative demand (EDDI-6 and EDDI-36) across central and
northern Chile. Savanna cover responded weakly to changes in
precipitation across continental Chile, but exhibited more pronounced
non-linear declines in response to increasing atmospheric evaporative
demand (EDDI-6 and EDDI-36) across most study regions. Cropland cover,
not surprisingly, varied weakly in response to changes in either water
supply or demand.

\begin{figure}[!ht]

\centering{

\includegraphics{figs/bars_relative_importance_RF.png}

}

\caption{\label{fig-rel_import_RF}\textbf{Shifts in water supply and
demand drive land cover change across multiple time scales.} Variable
importance of multi-scalar drought indices for explaining land cover
change in five study regions across continental Chile. Variable
importance was estimated with Random Forest models fitted for each
combination of study region and land cover type. SPI is the Standardized
Precipitation index, SPEI is the Standardized Precipitation
Evapotranspiration Index , SSI is the Standardized Soil Moisture Index,
and EDDI is the Evaporative Demand Drought Index. The numbers next to
the drought index correspond to the time scales in months (1- 36)}

\end{figure}%

\begin{figure}[!ht]

\centering{

\includegraphics{figs/RF_most_important_variables_trend_LC_vs_trend_DI.png}

}

\caption{\label{fig-parcial_variation}\textbf{Drought severity drives
land cover change, but not for all cover types.} Response of land cover
change in response to water demand and supply across multiple time
scales and study regions in continental Chile. SPI is the standardized
precipitation index, SPEI is the Standardized Precipitation
Evapotranspiration Index, SSI is the Standardized Soil Moisture Index,
and EDDI is the Evaporative Demand Drought Index. Numbers next to the
drought index correspond to the time scales in months (1- 36). Fitted
lines are smoothed response curves across river basins in each region
estimated with Random Forest models.}

\end{figure}%

\section{Discussion}\label{discussion}

\subsection{Temporal trends in water supply and
demand}\label{temporal-trends-in-water-supply-and-demand}

With the exception of the southernmost region, we found a significant
decreasing trend in water supply (SPI, SPEI, and SSI) over the past four
decades across continental Chile and is strongest in northern and
central Chile \citep{Boisier2018, Sarricolea2019}. Our results reveal
that decreases in water supply increased over longer time scales, which
is consistent with a progressive intensification of drought severity
across much of Chile, as has been observed in other regions experiencing
long-term droughts\citep{Rashid2019, Miro2023}. In parallel, we observed
an increased water demand (EDDI) due to rising air temperatures, which
also strengthened over longer time scales. Taken together, our results
provide multiple lines of evidence that continental Chile has
experienced a sustained drying trend due to a concurrent decrease in
precipitation and increase in atmospheric evaporative demand
\citep{Pascoa2021}.

With the exception of the southernmost region, we found a significant
decreasing trend in water supply (SPI, SPEI, and SSI) over the past four
decades across continental Chile and is strongest in northern and
central Chile \citep{Boisier2018, Sarricolea2019}. Our results reveal
that decreases in water supply increased over longer time scales, which
is consistent with a progressive intensification of drought severity
across much of Chile, as has been observed in other regions experiencing
long-term droughts \citep{Rashid2019, Miro2023}. In parallel, we
observed an increased water demand (EDDI) due to rising air
temperatures, which also strengthened over longer time scales. Taken
together, our results provide multiple lines of evidence that
continental Chile has experienced a sustained drying trend due to a
concurrent decrease in precipitation and increase in atmospheric
evaporative demand \citep{Pascoa2021}.

\subsection{Temporal trends in vegetation
productivity}\label{temporal-trends-in-vegetation-productivity}

The consequences of the persistent drying trend for ecosystems
throughout continental Chile are manifold. First, the prolonged
hydrological drought, i.e.~precipitation deficit, has reduced
groundwater storage \citep[SSI,][]{Taucare2024}, leading to a steady
decline in vegetation productivity (zcNDVI) since 2000 across northern
and central Chile, reaching its lowest level between 2020 and 2022. This
decline was most strongly associated with declines in soil moisture, as
has been reported for natural and productive ecosystems
\citep{Nicolai-Shaw2017, Jiang2020, Zhou2021}. Second, the strong
coupling between vegetation productivity and soil moisture over longer
time scales \citep{Bonan2008} that we observed provides a more direct
physiological explanation for the sharp decline in forest growth and
productivity in central Chile \citep[e.g.,][]{Miranda2023, Venegas2022},
as the dominant woody vegetation (e.g., trees, shrubs) in this region is
likely to obtain water from deeper in the soil profile than herbs,
grasses, or agricultural crops\citep{Oliviera2005}. Moreover, the
strengthening of the correlation between vegetation productivity and
water supply (SPI, SPEI, SSI) or demand (EDDI) over multiple time scales
(up to 36 months) and across land cover types
(Figure~\ref{fig-map_selec_time_scales_indices}) - demonstrates the
impacts of climate change on the water balance in Chile. Impacts may
extend beyond vegetation productivity, as reduced soil moisture in
central Chile and the western United States has increased wildfire
activity \citep{Holden2018, Gonzalez2018}, which is a growing concern in
Chile and may be further exacerbated by extensive plantations of highly
flammable tree species, e.g., Eucalyptus spp. and Pinus
spp.\citep{Bowman2019}. Third, we found that the decline in the
vegetation productivity of croplands is due to a decrease in the water
supply to a greater extent than to an increase in water demand
\citep{Quiring2010}, despite evidence that more water-intensive crops
have replaced less water-intensive crops in central Chile, leading to an
increase in water extraction from rivers or groundwater
\citep{Munoz2020, Duran2020}.

\subsection{Drought impacts on land
cover}\label{drought-impacts-on-land-cover}

We found evidence that temporal decreases in water supply and decreases
in water demand are driving shifts not only in vegetation productivity
but also in land cover across most of continental Chile. Forest and
grassland cover were particularly sensitive to changes in the water
balance over short and long temporal scales, which is consistent with
recent studies showing that progressive, long-term water deficits in
central Chile have triggered forest browning and declines in native
forest productivity \citep{Miranda2020, Venegas2022, Miranda2023}. While
our analysis do not distinguish between native and planted forests, the
latter of which are considered to be more drought tolerant in central
and southern Chile \citep{Carrasco2022}, we show that forest area
declines more sharply in response to increasing water demand due to
rising temperatures (EDDI) than decreasing water supply \citep[e.g.,
SPI, SSI,][]{Fajardo2019, Holz2018}, which may have cascading impacts on
multiple facets of forest diversity \citep{Segovia2021, Sabatini2022}.
Our results extend the results of these studies by showing that
drought-induced forest cover decline has extended beyond central Chile
to the southernmost region of continental Chile. This is noteworthy
because declines in vegetation productivity in southern Chile - a region
whose water balance is typically projected to be less affected by
climate change than central and northern Chile \citep{Breda2020} - have
only manifested since 2022 (Figure~\ref{fig-zcNDVI_var}). Moreover, our
results provide evidence that, in addition to forest cover, other land
cover types have been affected by water deficits, particularly
grasslands, despite physiological differences between dominant plant
growth forms \citep[e.g., trees, shrubs, C3 and C4
grasses,][]{Craine2013, McDowell2022}. Our results therefore suggest
that multiple land cover types could be vulnerable to regime shifts
towards more drought tolerant land cover types
\citep{Scheffer2001, Martinez-Vilalta2016}, such as shrublands, whose
cover increased non-linearly in response to increasing drought severity.
In central Chile, for example, the increase in shrubland cover could be
due to drought-induced decreases in savanna or cropland cover. Changes
in cropland cover may not be a direct consequence of drought
(Figure~\ref{fig-parcial_variation}), but rather an indirect one,
possibly reflecting the decision of resource-poor farmers to migrate to
regions with more abundant water resources or to change economic
activity \citep{AghaKouchak2021, Hermans2021}. In contrast, the increase
in shrubland cover due to a decrease in savanna cover may be ecological,
as shrubs may be more drought tolerant than other growth forms
\citep{Eldridge2011, Gotmark2016}.

Overall, our results show that long-term declines in water supply and
demand have induced widespread, multi-dimensional impacts on the
vegetation productivity and on the extent of land cover types. While
prolonged droughts may directly cause shifts to more drought-tolerant
land cover types, such as shrublands, they may also influence land cover
change through human decision making and activities. This study extends
current understanding of drought impacts by demonstrating how their
multidimensionality emerges over multiple time scales and across land
cover types, which can contribute to developing context-specific
adaptation strategies for agriculture, biodiversity conservation, and
natural resource management.

\section{Materials and Methods}\label{materials-and-methods}

\subsection{Study area}\label{study-area}

Continental Chile has a diverse climate, with strong gradients from
north to south and east to west \citep{Aceituno2021}
(Figure~\ref{fig-studyArea} a), which, together with its complex
topography, determine its ecosystem diversity
\citep{Garreaud2009, Luebert2022} (Figure~\ref{fig-studyArea} c). We
divided Chile into five regions: ``Norte Grande'' (17°34'--25°42'S),
``Norte Chico'' (25°42'-32°8'S), ``Centro'' (32°08'-36°12'S), ``Sur''
(36°12'-43°48'S), and ``Austral'' (43°48'-56°00'S). ``Norte Grande'' and
``Norte Chico'' are predominantly arid with hot (Bwh in the
Koppen-Geiger classification) and cold (Bwk) temperatures. Towards the
south of ``Norte Chico'', the climate changes to an arid steppe with
cold temperatures (Bsk). In these two northern regions, the land is
mostly bare, with a small area covered by shrublands and grasslands. In
the ``Centro'' region and the northern half of ``Sur'', the climate is
mostly Mediterranean, with warm to hot summers (Csa and Csb). Land cover
in the ``Centro'' region consists of a significant amount of shrublands
and savannas (50\%), followed by grasslands (16\%), forests (8\%), and
croplands (5\%). The south of ``Sur'' and the north of the ``Austral''
region have a mostly oceanic climate (Cfb). Those zones have a large
area of forests and grasslands. The southern part of the country has a
tundra climate, while ``Austral'' is a cold, semi-arid area covered by
grasslands and forests, and, to a lesser extent, savannas.

\begin{figure*}[!ht]

\centering{

\includegraphics{figs/map_study_con_landcover.png}

}

\caption{\label{fig-studyArea}(a) Chile with the Koppen-Geiger climate
classes and the five macrozones ``Norte Grande'', ``Norte Chico'',
``Centro'', ``Sur'', and ``Austral''. (b) Topography reference map. (c)
land cover classes for 2022. (d) Persistent land cover classes
(\textgreater{} 80\%) for 2001-2022.}

\end{figure*}%

\subsection{Data}\label{data}

\subsubsection{Gridded meteorological and vegetation
data}\label{gridded-meteorological-and-vegetation-data}

To derive a proxy for vegetation productivity, we used the Normalized
Difference Vegetation Index (NDVI) from the product MOD13A3 Collection
6.1 from MODIS . MOD13A3 provides vegetation indices with a 1 km spatial
resolution and monthly frequency \citep{Didan2015}. For soil moisture,
water supply, and water demand variables, we used ERA5-Land (ERA5L)
(ECMWF Reanalysis version 5 over land) \citep{MunozSabater2021}, a
reanalysis dataset that provides atmospheric and land variables since
1950. It has a spatial resolution of 0.1° (9 km), hourly frequency, and
global coverage. We selected total precipitation, maximum and minimum
temperature at 2 meters, and volumetric soil water layers between 0 and
100 cm of depth (layer 1 to layer 3; Table SSX).

\subsection{Short- to long-term drought
trends}\label{short--to-long-term-drought-trends}

\subsubsection{Atmospheric Evaporative Demand
(AED)}\label{atmospheric-evaporative-demand-aed}

To compute drought indices that use water demand, it is necessary to
first calculate AED. To do this, we employed the Hargreaves method
\citep{Hargreaves1994, Hargreaves1985} by applying the following
equation:

\begin{equation}\phantomsection\label{eq-AED}{AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}}\end{equation}

where \(Ra\) \((MJ\,m^2\, day^{-1})\) is extraterrestrial radiation;
\(T\), \(T_{max}\), and \(T_{min}\) are mean, maximum, and minimum
temperature \((°C)\) at 2m. For calculating \(Ra\) we used the
coordinate of the latitud of the centroid of each pixel as follow:

\begin{equation}\phantomsection\label{eq-Ra}{R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]}\end{equation}

where:

\(Ra\): extraterrestrial radiation \([MJ\, m^{-2} day-1]\),\\
\(G_{sc}\): solar constant = 0.0820 \([MJ\,m^{-2} min^{-1}]\),\\
\(d_r\): inverse relative distance Earth-Sun,\\
\(\omega_s\) sunset hour angle \([rad]\),\\
\(\phi\): latitude \([rad]\),\\
\(\delta\): solar declination \([rad]\).

We selected the method of Hargreaves to estimate AED because of its
simplicity, as it only requires temperature and extraterrestrial
radiation, and because access to the data needed for alternative methods
(e.g., Penman-Montieth) is often limited \citep{Vicente-Serrano2014}.

\subsubsection{Drought indices}\label{drought-indices}

To derive the drought indices of water supply and demand we used the
ERA5L dataset and the MODIS product (MOD13A3\citep{Didan2015b}), with a
monthly frequency for 1981--2023 and 2000--2023, respectively. Drought
indices capture historical anomalies of water supply and demand. To
quantify each anomaly, the common practice is to derive it following a
statistical parametric method in which it is assumed that the
statistical distribution of the data is known \citep{Heim2002}. The use
of an erroneous statistical distribution that does not fit the data is
usually the highest source of uncertainty \citep{Laimighofer2022}. In
the case of Chile, due to its high degree of climatic variability, it is
difficult to choose a statistical distribution that can be used across
its entire extent. We therefore use a non-parametric method for the
calculation of the drought indices, following \citet{Farahmand2015}.

For monitoring water supply, we used the Standardized Precipitation
Index (SPI; \citet{McKee1993}), which relies on precipitation data. To
evaluate water demand, we chose the Evaporative Demand Drought Index
(EDDI; \citet{Hobbins2016} and \citet{McEvoy2016}), which is based on
the AED. To consider the combined effect of water supply and demand, we
selected the SPEI \citep{Vicente-Serrano2010}. For SPEI, an auxiliary
variable \(D=P−AED\) is calculated. Soil moisture is the main driver of
vegetation productivity, particularly in semi-arid regions
\citep{Li2022}. Hence, we used the Standardized Soil Moisture Index
(SSI) to monitor soil moisture (SM) \citep{Hao2013}. For the SSI, we
used the average soil moisture from ERA5L at a depth of 1m. All
calculated indices are multi-scalar and can be used for the analysis of
short- to long-term droughts.

To derive the drought indices, we first calculate the sum of the
variables with regard to the time scale(s). In this case, for
generalization purposes, we will use \(V\), referring to variables P,
AED, D, and SM (Table SSX). We accumulated each over the time series of
values (months), and for the time scales s:

\begin{equation}\phantomsection\label{eq-sumvar}{A^s_i = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  }\end{equation}

The \(A^s_i\) corresponds to a moving window (convolution) that sums the
variable for time scales \(s\). This is summed over \(s\) months,
starting from the most recent month (\(n\)) back in time until month
\(n-s+1\). For example, using as a variable the precipitation, a period
of twelve months (\(n\)), and a time scale of three months (\(s\)):

\[
\begin{split}
A^3_1 &= P_{oct} +P_{nov} +P_{dic} \\
\vdots\,\,\, &= \,\,\,\vdots + \,\,\,\vdots + \,\,\,\vdots \\
A^3_{10} &= P_{jan}+P_{feb} +P_{mar}
\end{split}
\]

Then, we used the empirical Tukey plotting position \citep{Wilks2011}
over \(A_i^s\) to derive the \(P(A_i^s)\) probabilities across a period
of interest:

\begin{equation}\phantomsection\label{eq-probPai}{P(A^s_i) = \frac{i-0.33}{n+0.33'}}\end{equation}

An inverse normal approximation \citep{Abramowitz1968} obtains the
empirically derived probabilities once the variable cumulates over time
for the scale \(s\). Thus, the drought indices \(SPI\), \(SPEI\),
\(EDDI\), and \(SSI\) are obtained following the equation:

\begin{equation}\phantomsection\label{eq-DI}{DI(A^s_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}}\end{equation}

\(DI\) is referring to the drought index calculated for the variable
\(V\). The values for the constats are: \(C_0 = 2.515517\),
\(C_1 = 0.802853\), \(C_2 = 0.010328\), \(d_1 = 1.432788\),
\(d_2 = 0.189269\), and \(d3 = 0.001308\). For \(P(A^s_i) \leq 0.5\),
W=\(\sqrt{-2\cdot ln(P(A^s_i))}\) , and for \(P(A^s_i) > 0.5\), replace
\(P(A^s_i)\) with \(1-P(A^s_i)\) and reverse the sign of \(DI(A^s_i)\).

The drought indices were calculated for time scales of 1, 3, 6, 12, 24,
and 36 months at a monthly frequency for 1981--2023.

\subsubsection{Temporal trends of drought
indices}\label{temporal-trends-of-drought-indices}

To determine if there are statistically significant positive or negative
temporal trends for the drought indices, we used the non-parametric
modified Mann-Kendall test for serially correlated data \citep{Yue2004}.
To determine the magnitude of the trend, we used Sen's slope
\citep{Sen1968}. Sen's slope is less affected by outliers than
parametric ordinary least squares regression, and as a non-parametric
method it is not influenced by the distribution of the data. We applied
both methods for SPI, EDDI, SPEI, and SSI and six time scales, resulting
in a total of 24 trends. We then aggregated temporal trends for each
region and land cover type.

\subsection{Vegetation productivity}\label{vegetation-productivity}

We also used the MODIS product to calculate vegetation productivity, and
calculated anomalies in NDVI using zcNDVI \citep{Zambrano2018}, which
was derived from the monthly time series of NDVI, with Equations
Equation~\ref{eq-sumvar} and Equation~\ref{eq-DI}. For vegetation
productivity, we selected the time scale that best correlates with
annual net primary productivity (NPP) across continental Chile. For this
purpose, we calculated zcNDVI for time scales of 1, 3, 6, and 12 months
(from December) and compared it with the annual NPP. We obtained NPP
from MOD17A3HGF \citep{Running2019}. We chose to use six months because
the \(R^2\) between zcNDVI and NPP reaches its highest value at six
months, obtaining an \(R^2\) of 0.31 for forest and 0.72 for shrubland
(Supplementary Information Section S5). We subsequently used zcNDVI with
a time scale of 6 months and calculated it at a monthly frequency for
2000--2023.

\subsubsection{Drought impacts on vegetation
productivity}\label{drought-impacts-on-vegetation-productivity}

For each land cover, we analyzed the trend of vegetation productivity.
To this end, we identified areas within each land cover macro-class that
are persistent over time, to reduce the possibility that trends in
vegetation productivity may be influenced by changes in land cover. We
examined the correlation between drought indices and vegetation
productivity across land cover types to determine to the extent to which
soil moisture and water demand and supply affect vegetation
productivity.

We estimated pixel-to-pixel Pearson's correlations between drought
indices at time scales of 1, 3, 6, 12, 24, and 36 months with zcNDVI. We
extracted the Pearson correlation coefficient corresponding to the time
scale with the highest value. For each index, we then generated two
raster maps: 1) a raster with values of the time scales and drought
index that reached the maximum correlation, and 2) a raster with the
magnitude of the correlation between the drought index and vegetation
productivity.

\subsection{Drought impacts on land cover
change}\label{drought-impacts-on-land-cover-change}

\subsubsection{Land cover change}\label{land-cover-change}

To analyze land cover change, we used the classification scheme of the
International Geosphere-Biosphere Programme (IGBP) from the product
MCD12Q1 Collection 6.1 from MODIS. The MCD12Q1 product is produced for
each year from 2001 to 2022 and defines 17 classes (see Table Sx).
Following the FAO classification \citep{FAO2022}, we considered native
and planted forests as ``forests'', which represent natural and
productive ecosystems dominated by large trees. To analyze the land
cover change, we use the IGBP scheme from the MCD12Q1 Collection 6.1
from MODIS. We regrouped the 17 classes into ten macro-classes, as
follows: 1-4 to forests (native forest and plantations), 5-7 to
shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14
to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 as
water (Table S1). This resulted in a time series of land cover with ten
macro-classes for 2001 and 2023. We validated the land cover
macro-classes using a high resolution (30m ) land cover map for
2013-2014 \citep{Zhao2016}. Our results showed a global accuracy of
\textasciitilde0.82 and a F1 score of \textasciitilde0.66 (Supplementary
Information, S2).

We calculated the area for each land cover class in the five study
regions for 2001--2022. We then estimated the temporal change in area
for each land cover type and macro-class, and determined the statistical
significance and magnitude of the trend as described above. To assess
how water demand and supply, and soil moisture affect the variation in
vegetation productivity across various land cover types, we avoid
analyzing areas that experienced major land cover changes in the
2001--2022 period. To assess how zcNDVI varied irrespective of land
cover change, we developed a persistence mask for land cover, which only
retains pixels for which the macro-class remained the same for at least
80\% of the 22 years (Figure~\ref{fig-studyArea} d).

\subsubsection{Relationship between land cover and drought
trends}\label{relationship-between-land-cover-and-drought-trends}

To identify which drought indices and time scales have a major impact on
changes in land cover type, we examined the relationship between the
trend in land cover classes and the trend in drought indices. We
performed the analysis at the sub-basin scale, using 469 basins, which
have a surface area between 0.0746 and 24,000 \(km^2\) and a median area
of 1,249 \(km^2\). For each basin, we calculated the trend per land
cover type, considering the proportion of the type relative to the total
surface of the basin. For each basin we extracted the average trend of
all drought indices and at time scales of 1, 3, 6, 12, 24, and 36
months. Also, we extracted the average trend in zcNDVI.

We modeled trends in land cover type per macroclass with the aim of
assessing how land cover trends relate to drought indices. We used the
random forest method \citep{Ho1995}, which employs multiple decision
trees, allowing for classification and regression. Some advantages
include the ability to find non-linear relationships, reduce
overfitting, and derive variable importance. We included the four
drought indices at each time scale and zcNDVI for a total of 25
predictors and built six random forest models, one for each land cover
and region. We trained each model with 1000 trees using a resampling
strategy with cross-validation. To this end, we used cross-validation to
evaluate model fit using ten folds then calculated \(R^2\), root mean
square error (RMSE), and variable importance. Variable importance
identifies which variables have a higher contribution to explaining
model variation. We calculated variable importance by permuting
out-of-bag (OOB) data per tree and calculating the mean standard error
of the OOB data. After permuting each predictor variable, we repeated
the process for the remaining variables. We repeated this process ten
times per fold to assess model fit. Finally, we visually explored the
relationship between drought indices and changes in land cover. To do
this, we compared the relative changes in land cover surface with the
drought indices of six and thirty-six months.

\subsection{Software}\label{software}

For the downloading, processing, and analysis of the spatio-temporal
data, we used the open source software for statistical computing and
graphics, R \citep{R2023}. For downloading ERA5L, we used the \{ecmwfr\}
package \citep{Hufkens2019}. For processing raster data, we used
\{terra\} \citep{Hijmans2023} and \{stars\} \citep{Pebesma2023}. For
managing vectorial data, we used \{sf\} \citep{Pebesma2018}. For the
calculation of AED, we used \{SPEI\} \citep{Bergueria2023}. For mapping,
we used \{tmap\} \citep{Tennekes2018}. For data analysis and
visualization, the suite \{tidyverse\} \citep{Wickham2019} was used. For
the random forest modeling, we used the \{tidymodels\} \citep{Kuhn2020}
and \{ranger\} \citep{Wright2017} packages.

\section*{Acknowledgments}\label{acknowledgments}
\addcontentsline{toc}{section}{Acknowledgments}

The National Research and Development Agency of Chile (ANID) funded this
study through the drought emergency project FSEQ210022, Fondecyt
Iniciación N°11190360, Fondecyt Postdoctorado N°3230678, and Fondecyt
Regular N°1210526.

\section{References}\label{references}

\renewcommand{\bibsection}{}
\bibliography{references.bib}

\end{document}

### 2024-06-25 article/article_nature/supplementary_material.qmd
actualizar repositorio0+

--- 2024-06-25 article/article_first_version/supplementary.qmd
@@
 title: "Supplementary Material"
-subtitle: "The effects of drought on land cover change and vegetation productivity in continental Chile"
+subtitle: "Shifts in water supply and demand drive land cover change across Chile"
 author:
@@
 
+# Results
+
+```{r}
+#| echo: false
+#| out-width: 100%
+#| tbl-pos: '!ht'
+#| tbl-cap: "Summary per land cover and study region regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
+The numbers in each cell indicate the time scale that reached the maximum correlation for the land cover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.
+Cells without values indicate that the land cover type was not significant in that macrozone."
+knitr::include_graphics('figs/tabla_r_cor_macro_indice.png')
+```
+
 # Data
@@
 \centering
-\includegraphics[width = .5\textwidth]{../output/figs/tabla_landcover_macroclass.png}
+\includegraphics[width = .5\textwidth]{figs/tabla_landcover_macroclass.png}
 \end{table}
@@
 #| out-width: 100%
-#| fig-cap: "Error bar for the variables importance obtained from the 10 resamples folded by the Random Forest used to model land cover trend using as predictors the trend in drought indices of water supply, water demand, soil moisture, and vegetation productivity."
+#| fig-cap: "Error bar for the variables importance obtained from the 10 resamples folded by the random forest used to model land cover trend using as predictors the trend in drought indices of water supply, water demand, soil moisture, and vegetation productivity."
 #| fig-subcap: 
@@
 #| - "Barren Land"
-
-knitr::include_graphics("../output/figs/fig_errorbar_resample_random_forest_trends_Forest.png")
-knitr::include_graphics("../output/figs/fig_errorbar_resample_random_forest_trends_Cropland.png")
-knitr::include_graphics("../output/figs/fig_errorbar_resample_random_forest_trends_Grassland.png")
-knitr::include_graphics("../output/figs/fig_errorbar_resample_random_forest_trends_Savanna.png")
-knitr::include_graphics("../output/figs/fig_errorbar_resample_random_forest_trends_Shrubland.png")
-knitr::include_graphics("../output/figs/fig_errorbar_resample_random_forest_trends_Barren_land.png")
+knitr::include_graphics("figs/fig_errorbar_resample_random_forest_trends_Forest_norte grande.png")
+knitr::include_graphics("figs/fig_errorbar_resample_random_forest_trends_Cropland_norte grande.png")
+knitr::include_graphics("figs/fig_errorbar_resample_random_forest_trends_Grassland_norte grande.png")
+knitr::include_graphics("figs/fig_errorbar_resample_random_forest_trends_Savanna_norte grande.png")
+knitr::include_graphics("figs/fig_errorbar_resample_random_forest_trends_Shrubland_norte grande.png")
+knitr::include_graphics("figs/fig_errorbar_resample_random_forest_trends_Barren_land_norte grande.png")
 ```
@@
 #| fig-cap: "Heatmap of trends in zcNDVI for 2000 to 2023 per macrozone and landcover macroclass."
-knitr::include_graphics('../output/figs/heatmap_trends_zcNDVI_macro_landcover.png')
+knitr::include_graphics('figs/heatmap_trends_zcNDVI_macro_landcover.png')
 ```
@@
 #| fig-cap: "Spatial variation of the r-squared values obtained from the yearly correlation of zcNDVI of 1, 3, 6, and 12 months with the net primary productivity (NPP) for continental Chile."
-knitr::include_graphics('../output/figs/map_r2_NPP_vs_zcNDVI.png')
+knitr::include_graphics('figs/map_r2_NPP_vs_zcNDVI.png')
 ```
@@
 #| fig-cap: "A heatmap showing the r-squared values obtained from the yearly correlation of zcNDVI of 1, 3, 6, and 12 months with the net primary productivity (NPP) for continental Chile."
-knitr::include_graphics('../output/figs/heatmap_r2_npp_vs_zcndvi_per_landcover.png')
-```
-
-# Relationship of drought indices and vegetation productivity
-
-```{r}
-#| echo: false
-#| out-width: 100%
-#| fig-pos: '!ht'
-#| fig-cap: "Summary per land cover and study region regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36.
-The numbers in each cell indicate the time scale that reached the maximum correlation for the land cover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.
-Cells without values indicate that the land cover type was not significant in that macrozone."
-knitr::include_graphics('../output/figs/tabla_r_cor_macro_indice.png')
+knitr::include_graphics('figs/heatmap_r2_npp_vs_zcndvi_per_landcover.png')
 ```

### 2024-06-25 article/article_nature/supplementary_material.tex
actualizar repositorio0+

--- 2024-06-25 article/article_first_version/supplementary.tex
@@
 \begin{frontmatter}
-\title{Supplementary Material \\\large{The effects of drought on land
-cover change and vegetation productivity in continental Chile} }
+\title{Supplementary Material \\\large{Shifts in water supply and demand
+drive land cover change across Chile} }
 \author[1,2]{Francisco Zambrano%
@@
 
+\section{Results}\label{results}
+
+\includegraphics[width=1\textwidth,height=\textheight]{figs/tabla_r_cor_macro_indice.png}
+
 \section{Data}\label{data}
@@
 
-\subsection{Results}\label{results}
+\subsection{Results}\label{results-1}
 
@@
 
-\subsection{Results}\label{results-1}
+\subsection{Results}\label{results-2}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/fig_errorbar_resample_random_forest_trends_Forest.png}
+\includegraphics[width=1\textwidth,height=\textheight]{figs/fig_errorbar_resample_random_forest_trends_Forest_norte grande.png}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/fig_errorbar_resample_random_forest_trends_Cropland.png}
+\includegraphics[width=1\textwidth,height=\textheight]{figs/fig_errorbar_resample_random_forest_trends_Cropland_norte grande.png}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/fig_errorbar_resample_random_forest_trends_Grassland.png}
+\includegraphics[width=1\textwidth,height=\textheight]{figs/fig_errorbar_resample_random_forest_trends_Grassland_norte grande.png}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/fig_errorbar_resample_random_forest_trends_Savanna.png}
+\includegraphics[width=1\textwidth,height=\textheight]{figs/fig_errorbar_resample_random_forest_trends_Savanna_norte grande.png}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/fig_errorbar_resample_random_forest_trends_Shrubland.png}
+\includegraphics[width=1\textwidth,height=\textheight]{figs/fig_errorbar_resample_random_forest_trends_Shrubland_norte grande.png}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{../output/figs/fig_errorbar_resample_random_forest_trends_Barren_land.png}
+\includegraphics[width=1\textwidth,height=\textheight]{figs/fig_errorbar_resample_random_forest_trends_Barren_land_norte grande.png}
 
@@
 \caption{\label{fig-RF_importance}Error bar for the variables importance
-obtained from the 10 resamples folded by the Random Forest used to model
+obtained from the 10 resamples folded by the random forest used to model
 land cover trend using as predictors the trend in drought indices of
@@
 
-\includegraphics[width=0.75\textwidth,height=\textheight]{../output/figs/heatmap_trends_zcNDVI_macro_landcover.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{figs/heatmap_trends_zcNDVI_macro_landcover.png}
 
@@
 
-\includegraphics[width=0.75\textwidth,height=\textheight]{../output/figs/map_r2_NPP_vs_zcNDVI.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{figs/map_r2_NPP_vs_zcNDVI.png}
 
@@
 
-\includegraphics[width=0.5\textwidth,height=\textheight]{../output/figs/heatmap_r2_npp_vs_zcndvi_per_landcover.png}
+\includegraphics[width=0.5\textwidth,height=\textheight]{figs/heatmap_r2_npp_vs_zcndvi_per_landcover.png}
 
@@
 months with the net primary productivity (NPP) for continental Chile.}
-
-\end{figure}%
-
-\section{Relationship of drought indices and vegetation
-productivity}\label{relationship-of-drought-indices-and-vegetation-productivity}
-
-\begin{figure}[H]
-
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{../output/figs/tabla_r_cor_macro_indice.png}
-
-}
-
-\caption{Summary per land cover and study region regarding the
-correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and
-SSI for time scales of 1, 3, 6, 12, 24, and 36. The numbers in each cell
-indicate the time scale that reached the maximum correlation for the
-land cover and macrozone, and the color indicates the strength of the
-r-squared obtained with the index and the time scale. Cells without
-values indicate that the land cover type was not significant in that
-macrozone.}
 

### 2024-11-05 article/article_nature/article_nature.qmd
actualizar repo

@@
 
+To derive the drought indices of water supply and demand we used the ERA5L dataset and the MODIS product, with a monthly frequency for 1981–2023 and 2000–2023, respectively.
+Drought indices capture historical anomalies of water supply and demand.
+To quantify each anomaly, the common practice is to derive it following a statistical parametric method in which it is assumed that the statistical distribution of the data is known [@Heim2002].
+The use of an erroneous statistical distribution that does not fit the data is usually the highest source of uncertainty [@Laimighofer2022].
+In the case of Chile, due to its high degree of climatic variability, it is difficult to choose a proper distribution without previous research that could be applicable throughout its entire extent.
+We therefore use a non-parametric method for the calculation of the drought indices, following @Farahmand2015.
+
 For monitoring water supply, we used the Standardized Precipitation Index (SPI; @McKee1993), which relies on precipitation data.

### 2024-11-05 conference/IGARSS2024.qmd
actualizar repo

@@
 bibliography: references.bib
+csl: ieee.csl
 date: 2024-01-05
@@
 
-We evaluated the trend of the drought indices SPI, SPEI, EDDI, and zcSM; and their relation to vegetation productivity.
+We evaluated the trend of the drought indices SPI, SPEI, EDDI, and zcSM (standardized anomaly of cumulative soil moisture at one meter); and their relation to vegetation productivity.
 Our goal is to assess the impact on vegetation of the climate variables of water demand and supply and soil moisture.
@@
 (c) Land cover classes for 2022.
-(d) Persistent land cover classes (> 80%) for 2001-2022'
+(d) Persistent land cover classes (> 80%) for 2001-2022.
+The legend in (d) denotes both (c) and (d).'
 #| out-width: "\\textwidth"
@@
 
+# **Discusion**
+
+In our study, we considered the variation in vegetation productivity in Chile, specifically in areas without any changes in land cover, to prevent any misleading conclusions about the increase in water demand due to land cover change.
+Our results show a contrasting perspective regarding the evidence provided by @Vicente-Serrano2022 on a global scale, who indicates that the increase in drought is led by an increase in agricultural land, which in turn increases water demand.
+
+Our results indicate that except for the southern part of the country, the SPI, SPEI, and SSI (water supply) showed declining trends, while the EDDI (water demand) increased across continental Chile.
+The trends in water demand and supply were stronger as the time scales increased, indicating a long-term reduction in water supply (except for the southern part) and an increase in water demand by the atmosphere.
+Also, we found that there has been a significant declining trend in vegetation productivity (zcNDVI) since 2000 for the north-central part of the country, which reached its lowest level between 2020 and 2022 and has impacted natural and cultivated land.
+Further, croplands showed a decrease in surface area for the north-central region, while barren land increased.
+We link these changes to a decrease in the water demand from vegetation because, despite the increase in AED, the surface area for water-demanding vegetation is declining as well as the vegetation productivitybiomass production.
+However, some questions arise regarding what is occurring with the cultivated land.
+Evidence suggests that higher-water-demanding crops have replaced less demanding crops in the Petorca basin (central Chile), leading to an increase in water abstraction [@Munoz2020;@Duran2020].
+Nonetheless, at this scale of analysis, the effect of higher crop water demand on drought is minor compared to the decrease in water supply and increase in AED over all land cover types.
+
+The long-scale trends (e.g., 36 months) demonstrate the impact of climate change on water availability in Chile, potentially due to an intense hydrological drought stemming from the ongoing precipitation deficit and rising AED.
+But it is likely that in zones most affected by drought, the main cause is not an increase in vegetation water demand due to an intensification of cultivated land (e.g., an increase in irrigated crops) like in other parts of the globe (Vicente‐Serrano et al.
+2020).
+North-central Chile has experienced a decline in vegetation productivity across land cover types, which is primarily attributable to variations in water supply and soil moisture.
+An increase in water demand, led by an increase in the surface area of irrigated crops or the change to more water-demanding crops, could strengthen this trend, however, it escapes the scope of this study.
+Future work should focus on the regions where the drought has been more severe and has a high proportion of irrigated crops to get insight on the real impact of irrigation on ecosystems in those zones.
+
 # **Conlusion**

### 2024-11-05 R/trend_func.R

@@ -7,7 +7,21 @@ trend_func <- function(x,start = c(1981,1),end = c(2023,11),frecuency =12){
     res <- mk.test(t)
     mk_res <- ifelse(res$p.value < 0.05,1,NA)
     ss <- sens.slope(t)
-    mk_ss <- ifelse(ss$p.value < 0.05,ss$estimates,0)
+    mk_ss <- ifelse(ss$p.value < 0.05,ss$estimates,NA)
+    out <- c(mk_res,mk_ss)
+  } else out <- c(NA,NA)
+  return(out)
+}
+
+trend_func2 <- function(x){
+  x <- as.numeric(x)
+  
+  p <- sum(!is.na(x))/length(x)
+  
+  if (p > 0.5){
+    res <- mmky(x)
+    mk_res <- ifelse(res[2] < 0.05,1,NA)
+    mk_ss <- ifelse(res[2] < 0.05,res[7],0)
     out <- c(mk_res,mk_ss)
   } else out <- c(NA,NA)
   return(out)

### 2024-11-05 R/updt_modis.R

@@ -2,7 +2,10 @@
 # by Francisco Zambrano Bigiarini (frzambra@gmail.com)
 # July 2018
 
-updt_modis <- function(type,product,dir_out){
+updt_modis <- function(type,product,dir_out,token = NULL){
+  
+  if (is.null(token)) token <- Sys.getenv('token_nasa')
+  
   URL <- paste0("https://e4ftl01.cr.usgs.gov/",type,"/",product,'/')
 
   htmlContent <- RCurl::getURL(URL)
@@ -36,17 +39,18 @@ updt_modis <- function(type,product,dir_out){
     files <- files[filesInd]
     dir.create(paste0(dir_out,product,'/',dir,'/'))
     
-    purrr::map(files,function(name){
-      tryCatch(
-      httr::GET(url = paste0(URL,dir,'/',name),
-          config = httr::config(connecttimeout = 120),
-          httr::authenticate("frzambra@gmail.com","Traplozx398#"),
-          write_disk(paste0(dir_out,product,'/',dir,'/',name),
-                     overwrite=TRUE),
-          progress()),
-      error = function(e) print(paste('Error:',e))
+    out <- purrr::map(files,function(name){
+        tryCatch(
+        paste0(URL,dir,'/',name) |> 
+          httr2::request() |> 
+          httr2::req_auth_bearer_token(token) |> 
+          httr2::req_perform(path = paste0(dir_out,product,'/',dir,'/',name),
+                             verbosity = 0)
       )
-    })
+    }, .progress = list(
+      type = "iterator", 
+      format = "Downloading {cli::pb_bar} {cli::pb_percent}",
+      clear = TRUE))
   })
 }
 

### 2024-11-05 scripts/01_update_MODIS_products.R

@@ -7,56 +7,8 @@ library(XML)
 library(httr)
 library(curl)
 
-type <- c('MOLA','MOLT','MOTA')[2]
-product <- 'MOD17A3HGF.061'
-dir_out <- '/mnt/md0/raster_raw/MODIS/'
+type <- c('MOLA','MOLT','MOTA')[3]
+product <- 'MCD12Q1.061'
+dir_out <- '/mnt/discoB/raw/MODIS/'
 
 updt_modis(type,product,dir_out)
-
-URL <- paste0("https://e4ftl01.cr.usgs.gov/",type,"/",product,'/')
-
-htmlContent <- getURL(URL)
-htmlTree <- htmlParse(htmlContent)
-links <- xpathSApply(htmlTree, "//a/@href")
-
-datesWeb <- as.character(unlist(regmatches(links,gregexpr("[0-9]{4}.[0-9]{2}.[0-9]{2}",links))))
-
-if (dir.exists(file.path(dir.out,product))){
-  datesDir <- list.files(file.path(dir.out,product))
-  } else {
-    dir.create(file.path(dir.out,product))
-    datesDir <- character(0)
-}
-
-datesUpdate <- datesWeb[!datesWeb %in% datesDir]
-
-lapply(datesUpdate, function(dir){
-  htmlContent2 <- getURL(paste0(URL,dir,'/'))
-  htmlTree2 <- htmlParse(htmlContent2)
-  links2 <- xpathSApply(htmlTree2, "//a/@href")
-  
-  files <- unlist(regmatches(links2,gregexpr(paste0(substr(product,1,7),".*hdf$"),links2)))
-  filesInd <- grep(files,pattern='.*(h11v10|h11v11|h11v12|h12v12|h12v13|h13v13|h13v14|h14v14).*.hdf$')
-  files <- files[filesInd]
-  dir.create(paste0(dir.out,product,'/',dir,'/'))
-  
-  lapply(files,function(name){
-    tryCatch(
-    GET(url = paste0(URL,dir,'/',name),
-        config = httr::config(connecttimeout = 120),
-        authenticate("frzambra@gmail.com","29H1kc12"),
-        write_disk(paste0(dir.out,product,'/',dir,'/',name),
-                   overwrite=TRUE),
-        progress()),
-    error = function(e) print(paste('Error:',e))
-    )
-    # curl_download(paste0(URL,dir,'/',name),
[…]

### 2024-11-05 scripts/02_mosaic_chile_modis_products.R

@@ -4,10 +4,10 @@
 
 require(purrr)
 
-dir_in <- '/mnt/md0/raster_raw/MODIS/MOD17A3HGF.061/'
-dir_out <- '/mnt/md0/raster_procesada/MODIS/NPP.MOD17A3HGF.061/'
+dir_in <- '/mnt/discoB/raw/MODIS/MOD13A3.061/'
+dir_out <- '/mnt/discoB/processed/MODIS/NDVI.MOD13A3.061/'
 
-mosaic_modisFromHDF(dir_in,dir_out,lyr=2)
+mosaic_modisFromHDF(dir_in,dir_out,lyr=1)
 
 datesUpdate <- dates2Update(dir.out,dirs)
 

### 2024-11-05 scripts/14_IGBP_persistencia.R

@@ -2,8 +2,8 @@
 # by Francisco Zambrano (frzambra@gmail.com)
 # April 2021
 
-dir <- '/mnt/md0/raster_procesada/MODIS/'
-dir.out <- '/mnt/md0/raster_procesada/MODIS_derived/IGBP.MCD12Q1.061/'
+dir <- '/mnt/discoB/processed/MODIS/'
+dir.out <- '/mnt/discoB/processed/MODIS/IGBP.pers.MCD12Q1.061/'
 
 library(terra)
 library(rnaturalearth)
@@ -62,9 +62,9 @@ igbpRecl[igbpRecl == 0] <- NA
 writeRaster(igbpRecl,paste0(dir.out,'IGBP',prop,'_reclassified.tif'),
             wopt=list(datatype='INT1U'))
 
-# landcover del último año (2021)
+# landcover del último año (2023)
 
 igbp2022 <- classify(igbp[[22]],rcl=classes,include.lowest = TRUE)
 igbp2022[igbp2022 == 0] <- NA
-writeRaster(igbp2022,paste0(dir.out,'IGBP_2022_reclassified.tif'),
+writeRaster(igbp2022,paste0(dir.out,'IGBP_2023_reclassified.tif'),
             wopt=list(datatype='INT1U'))

### 2024-11-05 scripts/15_IGBP_variacion_tendencia_2001-2021.R

@@ -1,82 +1,0 @@
-# Analyzing landcover variation through Chile
-# August 2021
-# by frzambra
-
-dir <- '/mnt/md0/raster_procesada/MODIS/'
-
-paleta <- read.csv('data/processed_data/paleta_colores_landcover.csv')
-colors <-  rgb(paleta$R,paleta$G,paleta$B,maxColorValue = 255)
-attr(colors,'names') <- paleta$Name
-
-library(terra)
-library(rnaturalearth)
-library(sf)
-
-chl <- ne_countries(country = 'chile',returnclass = 'sf') |> 
-  st_transform(32719)
-
-igbp <- rast(list.files(paste0(dir,'IGBP.MCD12Q1.061'),pattern='*.tif$',full.names=TRUE))
-igbp <- crop(igbp,chl)
-
-classes <- matrix(c(1:17,rep(1,5),rep(2,2),rep(3,2),4,5,6,7,6,8,9,10),ncol=2)
-
-igbpRecl <- classify(igbp,rcl=classes,include.lowest = TRUE)
-igbpRecl[igbpRecl == 0] <- NA
-
-zones <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
-  st_transform(32719) |> 
-  mutate(macrozona = c('Norte Chico','Norte Grande','Austral','Centro','Sur'))
-
-data <- terra::extract(igbpRecl,zones)
-names(data) <- c('zone',2001:2022)
-
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-
-data |> 
-  pivot_longer(-zone,values_to = 'LC_type',names_to = 'year') |> 
-  filter(LC_type %in% c(1:4,6,9)) |>  
-  mutate(class2 = LC_type,
-         year = as.numeric(year),
-         zone = factor(zone,levels=c(2,1,4,5,3),labels = zones$macrozona[c(2,1,4,5,3)]),
-         LC_type = factor(LC_type,levels=paleta$class,labels=paleta$Name)) -> data2 
-
-data2 |> 
-  group_by(zone) |>  
-  filter(year == 2002) |>  
-  summarize(ntot = n()) ->dataSum
-
-data2 |> 
-  group_by(zone,year,LC_type) |>  
-  summarise(n=n()) |> 
-  left_join(dataSum,by='zone') |>  
-  mutate(prop = n/ntot,sup_km2=n*0.2146587) -> data3
-
[…]

### 2024-11-05 scripts/15_IGBP_variacion_tendencia_2001-2023.R

@@ -1,0 +1,111 @@
+# Analyzing landcover variation through Chile
+# August 2021
+# by frzambra
+
+dir <- '/mnt/md0/raster_procesada/MODIS/'
+
+paleta <- read.csv('data/processed_data/paleta_colores_landcover.csv')
+colors <-  rgb(paleta$R,paleta$G,paleta$B,maxColorValue = 255)
+attr(colors,'names') <- paleta$Name
+
+library(terra)
+library(rnaturalearth)
+library(sf)
+
+chl <- ne_countries(country = 'chile',returnclass = 'sf') |> 
+  st_transform(32719)
+
+igbp <- rast(list.files(paste0(dir,'IGBP.MCD12Q1.061'),pattern='*.tif$',full.names=TRUE))
+igbp <- crop(igbp,chl)
+
+classes <- matrix(c(1:17,rep(1,5),rep(2,2),rep(3,2),4,5,6,7,6,8,9,10),ncol=2)
+
+igbpRecl <- classify(igbp,rcl=classes,include.lowest = TRUE)
+igbpRecl[igbpRecl == 0] <- NA
+
+ecoregions <- read_sf('data/processed_data/spatial/ecoregiones_2017.gpkg') |> 
+  st_transform(32719) 
+
+data <- terra::extract(igbpRecl,ecoregions)
+names(data) <- c('zone',2001:2023)
+
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+
+data |> 
+  pivot_longer(-zone,values_to = 'LC_type',names_to = 'year') |> 
+  filter(LC_type %in% c(1:4,6,9)) |>  
+  mutate(class2 = LC_type,
+         year = as.numeric(year),
+         eco = factor(zone,labels = ecoregions$ECO_NAME),
+         LC_type = factor(LC_type,levels=paleta$class,labels=paleta$Name)) -> data2 
+
+data2 |> 
+  group_by(eco) |>  
+  filter(year == 2002) |>  
+  summarize(ntot = n()) ->dataSum
+
+data2 |> 
+  dplyr::select(-zone) |> 
+  group_by(eco,year,LC_type) |>  
+  summarise(n=n()) |> 
+  left_join(dataSum,by='eco') |>  
+  mutate(prop = n/ntot,sup_km2=n*0.2146587) -> data3
+
+saveRDS(data3,'data/processed_data/timeseries_landcover_ecoregion_LC_class_2001-2022.rds')  
+
[…]

### 2024-11-05 scripts/16_map_study_landcover.R

@@ -18,12 +18,12 @@ chl <- st_crop(chl,ext)
 set_defaults(map_token = "pk.eyJ1IjoiZnJ6YW1icmEiLCJhIjoiY2tqdmw5Z3QxMDZyZjJydG54M2RobWMyeSJ9.rl8_KzhiKaV0wgsLL2Y1WQ")
 
 #macrozonas
-zones <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg','marcozonas_chile')
+ecoregions <- read_sf('data/processed_data/spatial/ecoregiones_2017.gpkg') |> 
+  st_transform(32719)
 
 #centroides de las macrozonas
-zones <- zones |> st_centroid() |> st_coordinates() |> cbind(zones)
-
-divisiones <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg','divisiones_macrozona')
+ecoregiones <- ecoregions |> st_centroid() |> st_coordinates() |> cbind(ecoregions) |> 
+  select(X,Y,ECO_NAME,geom)
 
 bm <- basemap_raster(st_as_sf(chl),map_service = 'carto',map_type="light")
 bm_topo <- basemap_raster(st_as_sf(chl),'osm','topographic')
@@ -44,8 +44,6 @@ attr(colors,'names') <- paleta$class
 m2 <- tm_shape(bm) +
   tm_rgb() +
   tm_credits('(c)',position = c('left','bottom'),fontface = 'bold') +
-  # tm_shape(divisiones) + 
-  # tm_lines(lwd=1,col='black',lty='solid',alpha=.6)+
   tm_shape(igbp2022) +
   tm_raster(style = 'cat',
             palette = colors,title = 'Landcover classes',
@@ -74,7 +72,14 @@ m1 <- tm_shape(bm_topo) +
   #tm_text('macrozona',just='top',xmod=c(-2,-2,2,-2,-2),size=.5) +
   tm_credits('(b)',position = c('left','bottom'),fontface = 'bold')
 
-mapArr <- tmap_arrange(m1,m2,m3,asp=0.23)
+map_eco <- tm_shape(bm) +
+  tm_rgb() + 
+  tm_shape(ecoregions) +
+  tm_fill(col = 'ECO_NAME',palette = ecoregions$COLOR,title = 'Ecoregions',alpha = .7) +
+  tm_borders(lwd=1.5) +
+  tm_layout(legend.width = 1)
+
[…]

### 2024-11-05 scripts/17_extrarer_correlations_macro_landcover.R

@@ -20,9 +20,10 @@ files <- dir_ls(dir,regexp = 'tif$')
 cors <- rast(files)
 cors <- resample(cors,lc)
 
-macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
+eco <- read_sf('data/processed_data/spatial/ecoregiones_2017.gpkg') |> 
   st_transform(32719) |> 
-  mutate(macrozona = c('norte chico','norte grande','austral','centro','sur'))
+  filter(ECO_NAME != "Rock and Ice") |> 
+  mutate(ECO_NAME = fct(ECO_NAME,levels = c("Atacama desert","Chilean Matorral","Valdivian temperate forests","Magellanic subpolar forests","Patagonian steppe"))) 
 
 library(tidyverse)
 getmode <- function(v) {
@@ -33,28 +34,27 @@ getmode <- function(v) {
 #obtener los indices y escalas de tiempo en dónde se alcanza la máxima correlación
 cors1 <- subset(cors,seq(1,8,2))
 data_ind <- map_df(1:5,function(i){
-  cors_m <- mask(cors1,macro[i,])
-  lc_m <- mask(lc,macro[i,])
+  cors_m <- mask(cors1,eco[i,])
+  lc_m <- mask(lc,eco[i,])
   cors_df <- zonal(cors_m,lc_m,getmode)
   names(cors_df)[2:5] <- c('EDDI','SPI','SPEI','SSI')
   cors_df |> 
     pivot_longer(-1) |> 
     rename(indice = name) |> 
     mutate(clase = factor(lyr.1,levels = paleta$class, labels = paleta$Name),
-           macro = macro[i,]$macrozona) |> 
-    select(macro,indice,clase,value) 
+           eco = eco[i,]$ECO_NAME) |> 
+    select(eco,indice,clase,value) 
 })
 
 data_ind_4gt <- data_ind |> 
   filter(clase %in% c('Forest','Shrubland','Savanna','Grassland','Cropland','Barren land')) |> 
   mutate(value = ceiling(value),
-         value = factor(value,levels = 1:6,labels = c(1,3,6,12,24,36)),
[…]

### 2024-11-05 scripts/19_crear_gt_table_var_landcover_macro.R

@@ -3,8 +3,13 @@ library(gtExtras)
 library(tidyverse)
 library(kableExtra)
 
-dataLCV_ts <- readRDS('data/processed_data/timeseries_landcover_zone_LCclass_2001-2022.rds')  
-dataLCV_trend <- readRDS('data/processed_data/trends_landcover_2001-2022.rds') 
+dataLCV_ts <- readRDS('data/processed_data/timeseries_landcover_ecoregion_LC_class_2001-2023.rds')  |> 
+  filter(eco != "Rock and Ice") |> 
+  mutate(eco =fct(as.character(eco),levels = c("Atacama desert","Chilean Matorral","Valdivian temperate forests","Magellanic subpolar forests","Patagonian steppe")))
+
+dataLCV_trend <- readRDS('data/processed_data/trends_landcover_2001-2022.rds') |> 
+  filter(eco != "Rock and Ice") |> 
+  mutate(eco =fct(as.character(eco),levels = c("Atacama desert","Chilean Matorral","Valdivian temperate forests","Magellanic subpolar forests","Patagonian steppe")))
 
 dataSpark <- c('Shrubland', 'Savanna', 'Grassland', 'Barren land','Forest','Cropland') %>% 
   map(function(class){
@@ -12,8 +17,8 @@ dataSpark <- c('Shrubland', 'Savanna', 'Grassland', 'Barren land','Forest','Crop
       complete(LC_type) |> 
       ungroup() |> 
       filter(LC_type == class) |>   
-      group_by(zone,LC_type) |> 
-      select(zone,LC_type,sup_km2) |> 
+      group_by(eco,LC_type) |> 
+      select(eco,LC_type,sup_km2) |> 
       group_split() |> 
       setNames(c('a','b','c','d','e')) |> 
       map(function(x) pull(x,sup_km2))
@@ -22,7 +27,8 @@ dataSpark <- c('Shrubland', 'Savanna', 'Grassland', 'Barren land','Forest','Crop
 
 
 dataLCV_trend |> 
-  pivot_longer(-zone,names_to = 'cropland_class') |> 
+  select(-eco) |> 
+  pivot_longer(-ECO_NAME,names_to = 'cropland_class') |> 
   write_rds('data/processed_data/trend_landcover_change_classXzone_2001_2022.rds')
 
[…]

### 2024-11-05 scripts/21_landcover_per_macrozone.R

@@ -19,10 +19,12 @@ names(t) <- pal$Name
 lc <- rast('/mnt/md0/raster_procesada/MODIS_derived/IGBP.MCD12Q1.061/IGBP80_reclassified.tif') |> 
   project('EPSG:4326')
 
-macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
-  dplyr::mutate(macrozona = c('Norte Chico','Norte Grande','Austral','Centro','Sur'))
+ecoregions <- read_sf('data/processed_data/spatial/ecoregiones_2017.gpkg') |> 
+  filter(ECO_NAME != "Rock and Ice") |> 
+  mutate(ECO_NAME = fct(ECO_NAME,levels = c("Atacama desert","Chilean Matorral","Valdivian temperate forests","Magellanic subpolar forests","Patagonian steppe")))
+  
 
-LC_data <- terra::extract(lc,macro)
+LC_data <- terra::extract(lc,ecoregions,bind = TRUE)
 
 library(tidyverse)
 
@@ -32,24 +34,22 @@ attr(colors,'names') <- paleta$Name
 
 LC_data |> 
   mutate(LC_type = factor(lyr.1,levels=paleta$class,labels=paleta$Name),
-         zone = factor(ID,levels=1:5,labels=macro$macrozona),
-         zone = fct_relevel(zone, "Norte Grande", "Norte Chico",
-                            'Centro','Sur','Austral')
-         ) |> 
-  select(LC_type,zone) |> 
+         ECO_NAME = factor(as.character(ID),labels = c("Atacama desert", "Chilean Matorral", "Magellanic subpolar forests","Patagonian steppe", "Valdivian temperate forests"))) |> 
+  select(LC_type,ECO_NAME) |> 
   na.omit() |>  
-  group_by(zone,LC_type)|> 
+  group_by(ECO_NAME,LC_type)|> 
   summarize(ntot = n()) -> summ_data2
 write_rds(summ_data2,'data/processed_data/surface_landcover_macrozone.rds')
 #tabla con superficie
 #
+library(gt)
 summ_data2 |> 
   mutate(surface = 21.5*ntot*0.01) |> 
   select(-ntot) |> 
   mutate(surface = case_when(surface <=20~NA,
                            .default = surface)) |> 
   pivot_wider(names_from = LC_type,values_from=surface) |> 
[…]

### 2024-11-05 scripts/23_analysis_trend_drought_indices_and_landcover_macroXlandcover.R

@@ -90,9 +90,10 @@ data_trend_di |>
   geom_tile() +
   scale_fill_viridis_b(name = 'Trend of zcNDVI')  +
   scale_y_discrete(limits=rev) +
+  geom_text(aes(label = round(`zcNDVI-6`,3))) +
   theme_bw()+
   theme(axis.title = element_blank())
-ggsave('output/figs/heatmap_trends_zcNDVI_macro_landcover.png',scale=2)
+ggsave('output/figs/heatmap_trends_zcNDVI_macro_landcover.png',scale=1.2,width=6,height=3)
   
 data_model <- data_ana[,c(1:6,9:12,15:18,21:24,29,31)] |> 
   left_join(lc_surf) |> 

### 2024-11-05 scripts/25_trend_drought_indices.R

@@ -6,8 +6,7 @@ library(rnaturalearth)
 library(fs)
 
 chl_b <- ne_countries(country='chile',scale = 'medium',returnclass = 'sf')
-macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
-  mutate(macrozona = c('Norte Chico','Norte Grande','Austral','Centro','Sur'))
+ecoregions <- read_sf('data/processed_data/spatial/ecoregiones_2017.gpkg')
 
 ## Time series of zcNDVI for macrozones
 dir <- '/mnt/md0/raster_procesada/MODIS_derived/zcNDVI/zcNDVI-6'
@@ -23,15 +22,15 @@ lc_pers <- resample(lc_pers,zcndvi)
 lc_pers <- resample(lc_pers,zcndvi,method = 'mode')
 zcndvi <- mask(zcndvi,lc_pers)
 
-macro_mask <- rasterize(st_transform(macro,32719),zcndvi[[1]],field ='macrozona')
+eco_mask <- rasterize(st_transform(ecoregions,32719),zcndvi[[1]],field ='ECO_NAME')
 
-df_zcndvi <- zonal(zcndvi,macro_mask,fun = 'mean',na.rm = TRUE)
+df_zcndvi <- zonal(zcndvi,eco_mask,fun = 'mean',na.rm = TRUE)
 df_zcndvi |> 
-  pivot_longer(-macrozona) |> 
+  pivot_longer(-ECO_NAME) |>
+  filter(ECO_NAME != "Rock and Ice" ) |> 
   mutate(dates = str_extract(name,'[0-9]{4}-[0-9]{2}-[0-9]{2}'),
          dates = ymd(dates),
-         macrozona =str_to_title(macrozona),
-         macrozona = fct(macrozona,c('Norte Grande','Norte Chico','Centro','Sur','Austral'))) |> 
+         ECO_NAME = fct(ECO_NAME,levels = c("Atacama desert","Chilean Matorral","Valdivian temperate forests","Magellanic subpolar forests","Patagonian steppe"))) |> 
   select(-name) |> 
   drop_na() |> 
   ggplot(aes(dates,value)) +
@@ -43,11 +42,11 @@ df_zcndvi |>
   scale_x_date(date_breaks = "2 years", date_labels = "%Y",limits = c(as.Date("2000-01-01"),as.Date("2023-04-01")),expand = c(0,0)) +
[…]

### 2024-11-05 scripts/26_calculate_man_kendall_trend_on_rasters.R

@@ -4,29 +4,30 @@ library(sf)
 library(fs)
 library(glue)
 library(purrr)
+library(modifiedmk)
 
 source('R/trend_func.R')
 dir <- '/mnt/md0/raster_procesada/MODIS/'
 
-indices <- c('EDDI','SPI','SPEI','zcSM','zcNDVI')
+indices <- c('EDDI','SPI','SPEI','zcSM','zcNDVI','zcET')
 scales <- c(1,3,6,12,24,36)
-scales_ndvi <- c(1,3,6,12)
 
 map(indices,\(index){
-  if( index == 'zcNDVI') scales <- scales_ndvi
+  if( index == 'zcNDVI') scales <- c(1,3,6,12)
+  if (index == 'zcET') scales <- 12
   
   map(scales,\(scale){
-      if (index == 'zcNDVI'){
-        dir <- glue::glue('/mnt/md0/raster_procesada/MODIS_derived/zcNDVI/{index}-{scale}')
+      if (index == 'zcNDVI' | index == 'zcET'){
+        dir <- glue::glue('/mnt/md0/raster_procesada/MODIS_derived/{index}/{index}-{scale}')
       } else dir <- glue::glue('/mnt/md0/raster_procesada/ERA5-Land_tiff/sequia/{index}/{index}-{scale}')
       
-      files <- dir_ls(dir,regexp = 'tif$')
+      files <- dir_ls(dir,regexp = paste0('(',paste(2000:2023,collapse = '|'),').*tif$'))
       index_ts <- rast(files)
       cl <- snow::makeCluster(80,"SOCK")
-      snow::clusterExport(cl, list = c('trend_func','mk.test','sens.slope'))
-      trend <- app(index_ts,trend_func,cores = cl,start = c(2022,2),end =c(2024,2))
+      snow::clusterExport(cl, list = c('trend_func2','mmky'))
+      trend <- app(index_ts,trend_func2,cores = cl)
       snow::stopCluster(cl)
-      writeRaster(trend,glue('/mnt/md0/raster_procesada/analysis/trends/trend_mann_kendall_{index}-{scale}.tif'),overwrite = TRUE)
+      writeRaster(trend,glue('/mnt/md0/raster_procesada/analysis/trends/trend_mann_kendall_mod_{index}-{scale}.tif'),overwrite = TRUE)
       
     })
 })

### 2024-11-05 scripts/29_mapas_correlaciones_indices_zcNDVI.R

@@ -13,8 +13,8 @@ lc[lc %in% c(5,7:10)] <- NA
 dir <- '/mnt/md0/raster_procesada/analysis/correlations/'
 files <- dir_ls(dir,type = 'file',regexp = 'tif$')
 
-chl <- ne_countries(country = 'chile',returnclass = 'sf') |> st_transform(32719)
-
+chl <- ne_countries(country = 'chile',scale='medium',returnclass = 'sf') |> st_transform(32719)
+macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg')
 cors <- rast(files)
 
 cors <- crop(cors,lc)
@@ -24,7 +24,7 @@ cors_i <- subset(cors,seq(1,8,2))
 cors_r <- subset(cors,seq(2,8,2))
 
 set_defaults(map_token = "pk.eyJ1IjoiZnJ6YW1icmEiLCJhIjoiY2tqdmw5Z3QxMDZyZjJydG54M2RobWMyeSJ9.rl8_KzhiKaV0wgsLL2Y1WQ")
-bm <- basemap_raster(chl,map_service = 'osm',map_type="streets")
+bm <- basemap_raster(chl,map_service = 'carto',map_type="light_no_labels")
 
 names(cors_i) <- 1:4
 
@@ -38,8 +38,8 @@ map_i <- tm_shape(bm) +
             palette = '-magma',
             title =  'Time-scale (months)',
             legend.is.portrait = FALSE) +
-  # tm_shape(chl) +
-  # tm_borders() +
+  tm_shape(chl) +
+  tm_borders() +
   tm_facets(nrow = 1) + 
   tm_layout(
     panel.labels = c('EDDI','SPEI','SPI','SSI'),
@@ -59,6 +59,8 @@ map_r <- tm_shape(bm) +
             title = 'r',
             legend.hist = TRUE,
             legend.reverse = TRUE) +
+  tm_shape(chl) + 
+  tm_borders() +
   tm_facets(free.scales	= FALSE) + 
   tm_layout(
     panel.labels = c('EDDI','SPI','SPEI','SSI'),

### 2024-11-05 scripts/32_extraer_tendencias_indices_cuencas_landcover.R

@@ -7,32 +7,40 @@ library(tidyverse)
 cuencas <- read_sf('data/processed_data/spatial/subcuencas_bna.gpkg') 
 
 dir <- '/mnt/md0/raster_procesada/analysis/trends'
-files <- dir_ls(dir,regexp = 'mann_kendall.*tif$')
+files <- dir_ls(dir,regexp = 'mann_kendall_mod.*tif$')
+files <- files[-19]  #elimina ET
 indices <- str_remove(str_remove(str_extract(files,'l_.*\\.'),'l_'),'\\.')
 indices <- indices[-19:-22]
 trends <- rast(names(files[-19:-22]))
-trends <- trends[[seq(2,48,2)]]
+#trends <- trends[[seq(2,48,2)]]
+
+ind <- sapply(1:24,\(i) rep(i,2)) |> as.numeric()
+trends <- tapp(trends,ind,prod)
 names(trends) <- indices
 
-df_scales_1 <- terra::extract(trends[[7:24]],cuencas,fun =\(x){
-  table(x) |> sort(decreasing = TRUE) -> x
-  as.numeric(names(x[1]))
-})
+df_scales_1 <- terra::extract(trends[[7:24]],cuencas,mean,na.rm = TRUE)
 
-df_scales_2 <- terra::extract(trends[[1:6]],cuencas,fun =\(x){
-  table(x) |> sort() -> x
-  as.numeric(names(x[1]))
-})
+#fun =\(x){
+  # table(x) |> sort(decreasing = TRUE) -> x
+  # as.numeric(names(x[1]))
+#})
+
+df_scales_2 <- terra::extract(trends[[1:6]],cuencas,mean,na.rm = TRUE)
+#fun =\(x){
+#   table(x) |> sort() -> x
+#   as.numeric(names(x[1]))
+# })
 
 #zcNDVI
 trends <- rast(names(files[22]))
 trends <- trends[[2]]
 names(trends) <- 'zcNDVI-6'
 cuencas <- st_transform(cuencas,crs(trends))
-df_scales_3 <- terra::extract(trends,cuencas,fun =\(x){
-  table(x) |> sort(decreasing = TRUE) -> x
-  as.numeric(names(x[1]))
-})
+df_scales_3 <- terra::extract(trends,cuencas,mean,na.rm = TRUE)
+# fun =\(x){
+#   table(x) |> sort(decreasing = TRUE) -> x
+#   as.numeric(names(x[1]))
+# })
[…]

### 2024-11-05 scripts/33_analysis_trend_drought_indices_and_landcover_cuencas.R

@@ -72,17 +72,19 @@ df <- map(1:6,\(i){
     rf_wflow <- workflow(as.formula(paste0(vars[i],'~ .')), rf_spec)
     rf_fit <- fit(rf_wflow, df_train)
     
-    # df_test |> 
-    #   predict(rf_fit,df_test) |> 
-    #   bind_cols(trend = df_test$Cropland) |> 
-    #   ggplot(aes(.pred,trend)) + geom_point()
+    # predict(rf_fit,new_data =df_test) |>
+    #   bind_cols(trend = df_test$Shrubland) |>
+    #   ggplot(aes(.pred,trend)) + geom_point() + geom_abline(col='green') +
+    #   ggtitle(paste0('Macrozona: ',macro,' \n Uso de suelo: ',vars[t]))
     # 
     # # 
     #  augment(rf_fit, new_data = df_test) %>%
     #    metrics(as.name(vars[i]),.pred)
     # # 
-    #  augment(rf_fit, new_data = df_test) %>%
-    #    metrics(vars[i],.pred)
+    print(macro)
+    print(vars[t])
+    augment(rf_fit, new_data = df_test) %>%
+        metrics(vars[i],.pred) |> print()
     #  
     # library(vip)
     # rf_fit |> 
@@ -181,6 +183,7 @@ tabla |>
                  paste0('SPEI-',c(3,24,36)),
                  paste0('SSI-',c(1,3,36))
                  )))) |>
+  drop_na() |> 
   group_by(type,macrozona) |> 
   mutate(rel_imp = Mean*10e2) |> 
   slice_max(rel_imp, n= 3) |> 
@@ -191,17 +194,18 @@ tabla |>
                     values=colors) +
   labs(y = 'Relative variable importance') +
   coord_flip() +
+  guides(fill = guide_legend(nrow = 1)) +
   facet_grid(.~macrozona) +
   theme_bw() +
   theme(strip.background = element_rect('white'),
         legend.background = element_rect('transparent'),
         legend.key.size = unit(0.4, "cm"),
-        legend.title = element_text(size=10),
+        legend.title = element_text(size=9),
         legend.text = element_text(size=8),
         panel.grid = element_blank(),
         axis.title.y = element_blank(),
-        legend.position = c(.15,.25))
-ggsave('output/figs/bars_relative_importance_RF.png',height = 2.5,width=12,scale = 1.3)
+        legend.position = 'bottom')
[…]

### 2024-11-05 scripts/34_trend_zcNDVI_macrozona_y_landcover.R

@@ -1,3 +1,6 @@
+library(tidyverse)
+library(fs)
+
 #Trend zcNDVi-6 por amcrozona y landcover 
 
 #datos para landcover
@@ -13,8 +16,7 @@ names(trends) <- 'zcNDVI-6'
 
 
 
-macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
-  mutate(macrozona = c('Norte Chico','Norte Grande','Austral','Centro','Sur')) |> 
+ecoregions <- read_sf('data/processed_data/spatial/ecoregiones_2017.gpkg') |> 
   st_transform(32719)
 
 # ahora por tipo de landcover
@@ -25,17 +27,23 @@ lc_pers <- resample(lc_pers,trends,method = 'near')
 lc_pers[lc_pers %in% c(5,7:10)] <- NA
 
 data <-  map_df(1:5,\(i){
-  macro_n <- macro[i,]
+  macro_n <- ecoregions[i,]
   trends_n <- mask(trends,macro_n)
   lc_pers_n <- mask(lc_pers,macro_n)
   trends_df <- zonal(trends_n,lc_pers_n,'mean',na.rm = TRUE)
   trends_df |> 
-    mutate(macro = factor(macro_n$macrozona,levels = c('Norte Grande','Norte Chico','Centro','Sur','Austral')),
-           clase = factor(lyr.1,levels = paleta$class, labels = paleta$Name)) |> 
+    mutate(ECO_NAME = macro_n$ECO_NAME) |> 
     select(-lyr.1)
 })
 
 data_final <- data |> 
-  pivot_wider(names_from = clase,values_from = `zcNDVI-6`) |> 
+  pivot_wider(names_from = ECO_NAME,values_from = `zcNDVI-6`) |> 
   arrange(macro) |> 
   select(c(1,6,5,4,3,2)) 
+
+data |> 
+  ggplot(aes(clase,macro,fill = `zcNDVI-6`)) + 
+  geom_tile() +
+  scale_y_discrete(limits=rev) +
+  scale_fill_viridis_c(name = 'Trend of zcNDVI') + 
+  theme_bw()

### 2024-11-05 scripts/37_validacion_era5_temp_pre.R

@@ -109,10 +109,11 @@ data_temp |>
   summarise(
     ubrmse = ubrmse(temp.x,temp.y),
     mae = mae_vec(temp.x,temp.y),
-    cor = cor(temp.x,temp.y)
+    cor = cor(temp.x,temp.y),
+    bias = sum(temp.x)/sum(temp.y)
   ) |> 
   filter(ubrmse < 4) |>
-  summarize(across(ubrmse:cor,\(x) mean(x)))
+  summarize(across(ubrmse:bias,\(x) mean(x)))
   
   
   ggplot(aes(ubrmse)) + 
@@ -159,7 +160,7 @@ data_pre |>
   drop_na() |> 
   summarise(
      mae = mae_vec(pre.y,pre.x),
-     rmse = rmse_vec(pre.y,pre.x),
+     ubrmse = ubrmse(pre.y,pre.x),
      bias =sum(pre.y)/sum(pre.x),
      cor  = cor(pre.y,pre.x)
    ) |> 

### 2024-11-05 scripts/38_analysis_NPP_zcNDVI.R

@@ -82,6 +82,8 @@ r2_land |>
   mutate(landcover = paleta$Name) |> 
   write_rds('data/processed_data/r2_npp_vs_zcndvi.rds')
 
+r2_land <- read_rds('data/processed_data/r2_npp_vs_zcndvi.rds')
+
 r2_land |> 
   mutate(landcover = paleta$Name) |> 
   dplyr::filter(landcover %in% c('Forest','Shrubland','Savanna','Grassland','Cropland','Barren land')) |> 
@@ -95,7 +97,7 @@ r2_land |>
   scale_fill_viridis_c(name = 'r-squared') + 
   theme_bw() + 
   theme(axis.title = element_blank())
-ggsave('output/figs/heatmap_r2_npp_vs_zcndvi_per_landcover.png',scale=2)  
+ggsave('output/figs/heatmap_r2_npp_vs_zcndvi_per_landcover.png',scale=2,width=3,height=2)  
 
 macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> st_transform(32719)
 df_cov <- extract(cov,macro)

### 2024-11-05 scripts/41_graficos_variaciones_parciales.R

@@ -1,0 +1,218 @@
+library(tidyverse)
+library(tidymodels)
+library(probably)
+
+cuenca_macro <- read_rds('data/processed_data/cuencas_macrozona.rds')
+
+data_trend_lc <- read_rds('data/processed_data/trends_landcover_cuencas_2001-2022.rds') |> 
+  pivot_longer(-cuenca) |> 
+  rename(landcover = value,class = name) |> 
+  replace_na(list(landcover = 0)) |> 
+  pivot_wider(names_from = 'class',values_from = 'landcover')
+
+data_trend_di <- read_rds('data/processed_data/df_trends_indices_cuencas.rds') |>
+  mutate(index = case_when(
+    index == 'zcSM-1' ~ 'SSI-1',
+    index == 'zcSM-3' ~ 'SSI-3',
+    index == 'zcSM-6' ~ 'SSI-6',
+    index == 'zcSM-12' ~ 'SSI-12',
+    index == 'zcSM-24' ~ 'SSI-24',
+    index == 'zcSM-36' ~ 'SSI-36',
+    .default = index)
+  ) |> 
+  replace_na(list(trend = 0)) |> 
+  pivot_wider(names_from = 'index',values_from = 'trend') |> 
+  rename(cuenca = COD_SUBC )
+
+data_model <- data_trend_di |> 
+  full_join(data_trend_lc) |> 
+  full_join(cuenca_macro,by = c('cuenca' = 'COD_SUBC')) |> 
+  relocate(macrozona,.after=cuenca) |> 
+  relocate(Latitude,.after=macrozona) |> 
+  drop_na()
+
+d1 <- pivot_longer(data_model,`SPEI-1`:`EDDI-6`,names_to ='indices',values_to ='val_index') 
+d2 <- pivot_longer(data_model,Shrubland:Forest,names_to ='type',values_to = 'trend_lc') 
+
+data <- full_join(d1,d2,by=c('cuenca','macrozona')) |> 
+  select(cuenca:macrozona,indices,type,val_index,type,trend_lc)
+
+paleta <- read.csv('data/processed_data/paleta_colores_landcover.csv') |> 
+  dplyr::filter(Name %in% c('Forest','Shrubland','Savanna','Grassland','Cropland','Barren land')) |> 
+  mutate(Name = case_when(Name == 'Barren land' ~ "Barren Land",
+                          .default = Name))
+
[…]

### 2024-11-05 scripts/42_mapa_de_subcuencas.R

@@ -1,0 +1,32 @@
+library(sf)
+library(tmap)
+library(tidyverse)
+library(rnaturalearth)
+
+chl <- ne_countries(country= 'chile',scale = 'small',returnclass = 'sf') |> 
+  st_transform(32719)
+bb <- st_bbox(chl)
+bb[1] <- -550000
+subcuencas <- read_sf('data/processed_data/spatial/subcuencas_bna.gpkg')
+macro <- read_sf('data/processed_data/spatial/macrozonas_chile.gpkg') |> 
+  st_transform(32719)
+subcuencas <- st_crop(subcuencas,bb)
+
+subcuencas2 <- st_intersection(subcuencas,macro)
+macro$macrozona <- c('Norte Chico','Norte Grande','Austral','Centro','Sur')
+map <- tm_shape(subcuencas) + 
+  tm_borders() +
+  tm_shape(macro) +
+  tm_text('macrozona',size=.6,xmod = -3) +
+  tm_borders(col='red') +
+  tm_layout(frame = FALSE)
+tmap_save(map,'output/figs/mapa_subcuencas.png',scale=1.7)
+
+library(units)
+subcuencas2 <- subcuencas2 |> 
+  mutate(area = st_area(geom) |> set_units(km2))
+
+subcuencas2 |> 
+  st_drop_geometry() |> 
+  group_by(macrozona) |> 
+  summarize(n = n(),max_area = max(area),med_area = median(area), min_area = min(area))

### 2024-11-05 scripts/43_preparar_ecoregiones_chile.R

@@ -1,0 +1,15 @@
+library(sf)
+library(rnaturalearth)
+
+eco2017 <- read_sf('data/processed_data/spatial/ecoregiones_2017.gpkg')
+chl <- ne_countries(country = 'chile',scale = 'medium',returnclass = 'sf')
+
+bb <- st_bbox(chl)
+bb[1] <- -77
+
+bb_pol <- bb |> st_as_sfc() 
+eco2017_chl <- st_intersection(eco2017,bb_pol) |> 
+  st_intersection(chl)
+plot(eco2017_chl["ECO_NAME"])
+
+write_sf(eco2017_chl,'data/processed_data/spatial/ecoregiones_2017.gpkg')

### 2024-11-05 scripts/44_descargar_nasa_marble.R

@@ -1,0 +1,27 @@
+library(blackmarbler)
+library(geodata)
+library(tidyterra)
+library(ggplot2)
+
+token<- Sys.getenv('token_nasa')
+
+chl <- gadm('chl',path = tempdir())
+
+nasa_ln <- lapply(2012:2023,\(year){
+  bm_raster(roi_sf = chl,
+            product_id = "VNP46A4",
+            date = year,
+            bearer = token,
+            check_all_tiles_exist = FALSE)
+})
+
+nasa_ln[[1]] <- crop(nasa_ln[[1]],ext(nasa_ln[[2]])) 
+
+nasa_ln_join <- rast(nasa_ln)
+nasa_ln_join <- mask(nasa_ln_join,chl)
+
+dir_out <- '/mnt/md0/raster_raw/VNP46A4/'
+
+lapply(seq_along(2012:2023),\(i){
+  writeRaster(subset(nasa_ln_join,i),paste0(dir_out,'VNP46A4_Chile_',names(nasa_ln_join[[i]]),'.tif'))
+})

### 2024-11-05 scripts/45_mask_resample_1km_ET.R

@@ -1,0 +1,26 @@
+library(sf)
+library(rnaturalearth)
+library(terra)
+library(fs)
+library(purrr)
+
+chl <- ne_countries(country='chile',scale = 'medium',returnclass = 'sf') |> 
+  st_transform(32719)
+
+dir <- '/mnt/discoB/processed/MODIS/ET.MOD16A2GF.061_monthly/'
+files <- dir_ls(dir,regexp = 'tif$')
+
+ndvi <- rast('/mnt/discoB/processed/MODIS/NDVI.MOD13A3.061/chl_2016245_1_km_monthly_NDVI.tif') |> 
+  mask(chl) |> 
+  trim()
+
+dir_out <- '/mnt/discoB/processed/MODIS/ET.MOD16A2GF.061_monthly_1km/'
+map(files,\(file){
+  rast(file) |> 
+    mask(chl) |> 
+    trim() |> 
+    resample(ndvi) |> 
+    writeRaster(filename = paste0(dir_out,basename(file[1])))
+})
+
+

### 2024-11-05 scripts/46_modis_et8_2_monthly.R

@@ -1,0 +1,23 @@
+library(fs)
+library(terra)
+library(tidyverse)
+
+dir <- '/mnt/discoB/processed/MODIS/ET.MOD16A2GF.061/'
+files <- dir_ls(dir,regexp = 'tif$')
+dates <- as.Date(str_extract(files,'[0-9]{7}'),'%Y%j')
+
+et <- rast(files)
+
+ind <- paste0(year(dates),str_pad(month(dates),3,'left',pad='0'))
+
+system.time(
+  et_mes <- tapp(et,ind,sum,na.rm = TRUE)
+)
+
+dir_out <- '/mnt/discoB/processed/MODIS/ET.MOD16A2GF.061_monthly/'
+dates_mes <- seq(ymd("20000101"),ymd("20231201"),by='1 month') |> format("%Y%m%d")
+
+lapply(seq_along(dates_mes),\(i){
+  writeRaster(subset(et_mes,i),paste0(dir_out,'ET.MOD16A2GF.061_monthly_',dates_mes[i],'.tif'),overwrite = TRUE)
+})
+

### 2025-01-12 article/article_scientific_reports/mansucript_11012025.qmd
actualizar manuscrito quarto

---
title: Shifts in water supply and demand drive land cover change across Chile
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-nature
    classoption:
      - numbered
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
  html:
    code-fold: true
    html-math-method: katex
  docx: default
# The `csl` style will be used with LaTeX-based rendering of`nature-pdf`
# if cite-method is `citeproc`. This is also the default if cite-method
# is not specified.
csl: _extensions/christopherkenny/nature/csl/springer-vancouver.csl
author:
  - name: Francisco Zambrano
    email: francisco.zambrano@umayor.cl
    affiliations: 
        - name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad, Universidad Mayor
          id: 1
          city: Santiago
          country: 7500994, Chile.
        - id: 2
          name: Observatorio de Sequía para la Agricultura y la Biodiversidad de Chile (ODES), Universidad Mayor
          city: Santiago
          country: 7500994, Chile.
    attributes:
        corresponding: true
    #note: This is the first author footnote.
  - name: Anton Vrieling
    #email: a.vrieling@utwente.nl
    affiliations:
        - id: 3
          name: Faculty of Geo-Information Science and Earth, University of Twente
          city:  Enschede
          country: The Netherlands.
  - name: Francisco Meza
    affiliations:
        - id: 4
          name: Facultad de Agronomía y Sistemas Naturales, Pontificia Universidad Católica de Chile. 
          city: Santiago
          country: Chile.
        - id: 5
          name: Instituto para el Desarrollo Sustentable. Pontificia Universidad Católica de Chile
          city: Santiago
          country: Chile.
        - id: 6
          name: Centro Interdisciplinario de Cambio Global, Pontificia Universidad Católica de Chile
          city: Santiago
          country: Chile.
  - name: Iongel Duran-Llacer
    affiliations:
        - id: 7
          name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Universidad Mayor,
          city: Santiago
          country: 7500994, Chile.
  - name: Francisco Fernández
    #email: francisco.fernandez@uss.cl
    affiliations:
        - id: 8
          name: Center of Economics for Sustainable Development (CEDES), Faculty of Economics and Government, Universidad San Sebastian
          city: Santiago
          country: Chile.
        - id: 9
          name: Center of Applied Ecology and Sustainability (CAPES)
          city: Santiago
          country: Chile.
  - name: Alejandro Venegas-González
    affiliations:
        - id: 10
          name: Instituto de Ciencias Agroalimentarias, Animales y Ambientales (ICA3), Universidad de O'Higgins
          city: San Fernando
          country: Chile.
  - name: Nicolas Raab
    affiliations:
        - id: 4
  - name: Dylan Craven
    affiliations:
        - id: 11
          department: GEMA Center for Genomics, Ecology & Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750
          city: Santiago
          country: Chile.
        - id: 12
          name: Data Observatory Foundation
          city: Santiago
          country: Chile.
abstract: |
  Globally, droughts are becoming longer, more frequent, and more severe, and their impacts are multidimensional. These impacts typically extend beyond the water balance, as long-term, cumulative changes in the water balance can lead to regime shifts in land cover. Here, we assess the effects of temporal changes in water supply and demand over multiple time scales on vegetation productivity and land cover changes in continental Chile, which has experienced a severe drought since 2010. Over most of continental Chile, we have observed a persistent negative trend in water supply and a positive trend in atmospheric water demand since 2000. However, in water-limited zones, we have observed a negative trend in the vegetation's water demand. All the trends intensify over longer time scales. This long-term decrease in water availability has led to a decrease in vegetation productivity, especially for the Chilean matorral and the Valdivian temperate forest. Drought indices of soil moisture and actual evapotranspiration from short- to mid-term time scales primarily explain this decrease. Besides, our models indicate that the intensity of the drought explains up to 78% of the changes in shrubland land covers, decreasing to 40% for forests, but the change in cropland surface is explained by the surface burned.  Long-term climate change may cause regime shifts in land cover, which context-specific adaptation strategies can mitigate.
keywords: [drought, land cover, water demand, water supply]
bibliography: referencias.bib  
---

# Introduction {#sec-intro}

Across many regions of the world, droughts are becoming longer, more frequent, and more severe[@Miranda2023; @IPCC2023], impacting ecosystems via tree mortality[@Cheng2024], reducing vegetation productivity[@Miranda2023] and inducing shifts in land use and cover [@Crausbay2017]. However, identifying drought events is idiosyncratic due to the varying criteria used for classification. Droughts can be classified as 1) meteorological, i.e., when precipitation in a specific period falls below mean precipitation values observed over multiple years [@Mishra2010] (usually more than 30 years); 2) hydrological, i.e., when precipitation anomalies last for long periods (months to years) and affect the  hydrological system [@VanLoon2015;@VanLoon2016b] (e.g., streamflows, reservoirs and groundwater); 3) agricultural, i.e. when precipitation deficits negatively impact plant health, leading to decreases in crop or pasture productivity [@Wilhite1985]; or 4) ecological, i.e., when water availability negatively affect the provisioning of ecosystem services and trigger feedbacks in natural or human systems [@Crausbay2017]. Such feedbacks include drought impacts on human decision making and activities, which can lead to land-cover change [@VanLoon2016;@AghaKouchak2021], which may have cascading effects on biodiversity and ecosystem services [e.g., @Lawler2014;@Newbold2015]. Despite the high degree of confidence in the impacts of rising temperatures on the extent, frequency, and severity of agricultural and ecological droughts [@IPCC2023], which are likely to increase even if global warming stabilizes at 1.5°–2°C, the severity of meteorological droughts has been remarkably stable globally over the past century [@Vicente-Serrano2022;@Kogan2020]. A global study analyzing drought severity trends from 1980 to 2020 reveals that in a few regions (some mid-latitudinal and subtropical areas), rising temperatures during the warm season have increased atmospheric evaporative demand (AED), leading to an increase in agricultural land area [@Vicente-Serrano2022]. Thus, rising atmospheric water demand may reflect parallel changes in land cover—primarily agriculture—that can exacerbate the effects of meteorological droughts on ecosystems.

Expanding analyses to include multiple dimensions of droughts can provide complementary insights into the Earth’s water balance - and its impacts - over multiple time scales. Yet, the World Meteorological Organization recommends the use of a single drought index for monitoring droughts [@WMO2012], i.e., the multi-scale Standardized Precipitation Index (SPI; [@McKee1993]), which is limited in that it only considers  water supply in the form of precipitation. The Standardized Precipitation Evapotranspiration Index (SPEI; [@Vicente-Serrano2020]) builds upon SPI by incorporating the effects of  temperature on drought, and is now used widely for drought monitoring [e.g., @Gebrechorkos2023;@Liu2024]. Indices derived from soil moisture products [@Narasimhan2005; @Souza2021], such as the Standardized Soil Moisture Index [SSI, @AghaKouchak2014;@AghaKouchak2015] also monitor water supply and are thought to better capture water availability for crops, thus providing more relevant information for evaluating agricultural droughts. To disentangle the effects of precipitation from those of temperature [@Vicente-Serrano2020], as well as to capture droughts in terms of water atmospheric demand, AED has been integrated into the Evaporative Demand Drought Index [EDDI, @McEvoy2016], which is particularly effective at detecting the rapid onset or intensification of droughts. To quantify vegetation water demand, we must use the actual evapotranspiration, or the amount of water removed from a surface by evaporation and transpiration; the Standardized Evapotranspiration Index (SETI; [@Yang2019]) can be used for this purpose. In turn, ecological droughts, which capture the joint effects of precipitation and temperature in modifying natural and productive ecosystems [@Camps-Valls2021;@Paruelo2016;@Helman2014], are complex to measure and can therefore be monitored using multiple drought indices that capture the multiple dimensions of drought, e.g., precipitation, temperature, evapotranspiration, and AED. Although such an approach accounts for the joint effects of changes in natural and productive ecosystems, its potential impacts on land cover change have been largely unexplored.

From 1960 to 2019, land-use change has impacted approximately one-third of the Earth’s surface, which is four times more than previously thought [@Winkler2021]. Despite the considerable interest in land-use change dynamics [e.g., @Song2018;@Winkler2021], the direction and magnitude of  drought impacts on land cover change and vegetation productivity remain uncertain [@Chen2022;@Akinyemi2021;@Peng2017]. Meteorological droughts are responsible for approximately 37% of land cover change and variability in vegetation productivity globally [@Peng2017]. However, the evidence supporting these results is derived from only one drought index, SPEI, which combines a proxy for water supply -  precipitation - with a proxy for water demand - AED - at one time scale (12 months). The use of only one time scale may bias results of drought impacts towards ecosystems dominated by plant growth forms such as grasses and herbs that respond more rapidly to drought stress (< 12 months). This is because  physiological differences among and within dominant plant growth forms may increase (or decrease) tolerance of drought stress [@Craine2013;@McDowell2022]. For example, trees growing in more arid ecosystems typically respond over longer time scales than those in more humid ecosystems [@Vicente-Serrano2014]. Another source of uncertainty regarding drought impacts on land cover change and vegetation productivity are extrinsic factors, such as large-scale public policy (e.g., national and international reforestation initiatives), agricultural practices (e.g., clearing forest for soybean or oil palm), and rural and urban land use planning [@Karabulut2023].

To deepen current knowledge on the multidimensional impacts of drought on the temporal dynamics of natural and productive ecosystems, we evaluate temporal changes in water supply and demand, net primary productivity, and land-cover change across terrestrial ecosystems in continental Chile for 2000-2023. Chile’s diverse climate and ecosystems [@Beck2023;@Luebert2022] make it an ideal natural laboratory for assessing the dynamic interactions between climate and ecosystems, and potential impacts on land-cover change. Additionally, large parts of Chile have experienced severe drought conditions that have significantly affected vegetation and water storage in recent years; north-central Chile has faced a persistent precipitation deficit (or "mega-drought") since 2010 [@Garreaud2017], which has broadly impacted native forests [e.g., @Miranda2020;@UrrutiaJalabert2018;@Venegas2018] and agricultural productivity [e.g., @Zambrano2016;@Zambrano2018;@Zambrano2023]. However, the effects of this prolonged extreme drought may also extend to changes in land cover, altering the provision of key ecosystem services and agricultural production. Here, we aim to assess: short- to long-term time trends in multi-scalar drought indices that capture variation in the components of water balance, i.e., water supply (SPI, SPEI, SSI) and demand (EDDI, SETI) and their impacts on vegetation productivity and land-cover change across continental Chile. By doing so, we aim to deepen understanding of how water supply and demand affect vegetation productivity and influence changes in the surface of land cover in a severely drought-impacted region. This may enhance understanding of the alterations, due to drought, to ecosystems in various regions across the world.

# Materials and Methods

## Study area

Continental Chile has a diverse climate, with strong environmental gradients from north to south and east to west [@Aceituno2021] (Fig. 1a), which, together with its complex topography (Fig. 1b), determine its ecosystem diversity [@Garreaud2009;@Luebert2022] (Fig. 1c). We therefore divided Chile into ecoregions [@Dinerstein2017], which are regions that share similar geography and ecology, and have comparable levels of precipitation and solar radiation. There are seven ecoregions: Atacama desert, Central Andean dry puna, Southern Andean steppe, Chilean Matorral, Valdivian temperate forests, Magellanic subpolar forests, and Patagonian steppe. The Atacama desert is predominantly arid with hot (Bwh in the Koppen-Geiger classification) and cold (Bwk) temperatures, as well as the northern part of the Chilean Matorral. Most of the land in these two northern regions is bare, except for a small area where shrublands and grasslands are present. With an annual rainfall of less than 400 mm, the Central Andean dry puna ecoregion has low, yet highly seasonal precipitation with an eight-month dry season, low temperatures (Bwk) and is dominated by grasslands, shrublands, and savanna. The climate of the Southern Andean steppe ecoregion is cold desert (BWk), with most precipitation occurring in the winter. There is little vegetation in this ecoregion because the plants have adapted to its windy, dry, and cold climate. In central Chile, the climate of the Chilean Matorral  changes to that of an arid steppe with cold temperatures (Bsk). Then, towards the center-south of the country, the climate of the Chilean Matorral changes to a Mediterranean climate, with warm to hot summers (Csa and Csb).  Land cover in this ecoregion consists of a significant amount of shrublands and savannas. The Valdivian temperate forests have a mostly oceanic climate (Cfb) and a large area of forests and grasslands. The Magellanic subpolar forests have a tundra climate. Lastly, the Patagonian steppe has high aridity, cold temperatures (Bsk), and primarily consists of grasslands.

## Data

### Gridded meteorological and vegetation data

To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the MOD13A3 Collection 6.1 product derived from the MODIS (Moderate-Resolution Imaging Spectroradiometer) sensor onboard the Terra satellite. MOD13A3 provides vegetation indices with a 1 km spatial resolution and monthly frequency [@Didan2015]. We also utilized the MOD16A2 collection 6.1 [@Running2017] product from MODIS to gauge the water consumption of vegetation. This product gives us monthly actual evapotranspiration (ET) and atmospheric evaporative demand (AED) with a ~500m spatial resolution. For soil moisture, water supply, and water demand variables, we used ERA5-Land (ERA5L; ECMWF Reanalysis version 5 over land) [@MunozSabater2021], a reanalysis dataset that provides atmospheric and land variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage. We selected total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100 cm of depth (layer 1 to layer 3; Supplementary Materials and Methods, Supplementary Tables 2 and 4).

### Gridded indicators for land use

To account for the impacts of human activity on land cover change, we obtained data on road density [@Meijer2018] and night lights for the period 2012–2023 [@Roman2018]. These products are frequently used in the literature to quantify the human footprint (e.g., [@Halpern2015;@Halpern2008]) or biodiversity threats (e.g., [@Bowler2020;@Kennedy2020]).To capture changes on land cover due to fires, we calculated the total burned area for 2002-2023 [@Chen2023].

## Short- to long-term drought trends

### Atmospheric Evaporative Demand (AED)

To compute drought indices that use water demand, it is necessary to first calculate AED. To do this, we employed the Hargreaves method [@Hargreaves1994;@Hargreaves1985] by applying the following equation:

$$AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}$$ {#eq-AED}

where $Ra$ $(MJ\,m^2\, day^{-1})$ is extraterrestrial radiation; $T$, $T_{max}$, and $T_{min}$ are mean, maximum, and minimum temperature $(°C)$ at 2m. For calculating $Ra$ we used the coordinate of the latitud of the centroid of each pixel as follow:

$$R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]$$ {#eq-Ra}

where:

$Ra$: extraterrestrial radiation $[MJ\, m^{-2} day-1]$,  
$G_{sc}$: solar constant = 0.0820 $[MJ\,m^{-2} min^{-1}]$,    
$d_r$: inverse relative distance Earth-Sun,  
$\omega_s$ sunset hour angle  $[rad]$,  
$\phi$: latitude $[rad]$,  
$\delta$: solar declination $[rad]$.

We selected the method of Hargreaves to estimate AED because of its simplicity, as it only requires temperature and extraterrestrial radiation, and because access to the data needed for alternative methods (e.g., Penman-Montieth) is often limited [@Vicente-Serrano2014].

### Drought indices

To derive the drought indices of water supply and demand we used the ERA5L dataset and the MODIS product (MOD13A3 [@Didan2015b]), with a monthly frequency for 2000–2023. Drought indices capture historical anomalies of water supply and demand. To quantify each anomaly, the common practice is to derive it following a statistical parametric method in which it is assumed that the statistical distribution of the data is known [@Heim2002]. The use of an erroneous statistical distribution that does not fit the data is usually the highest source of uncertainty [@Laimighofer2022]. In the case of Chile, due to its high degree of climatic variability, it is difficult to choose a statistical distribution that can be used across its entire extent. We therefore use a non-parametric method for the calculation of the drought indices, following [@Farahmand2015].

For monitoring water supply, we used the Standardized Precipitation Index (SPI;  [@McKee1993]), which only uses precipitation data. To evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI; [@Hobbins2016] and  [@McEvoy2016]), which is based on AED, and the Standardized Evapotranspiration Index (SETI; [@Yang2019]), which quantifies actual evapotranspiration, i.e. the amount of water removed from a surface due to evaporation and transpiration. To quantify the combined effect of water supply and demand, we estimated SPEI [@Vicente-Serrano2010]. For SPEI, we calculated an auxiliary variable (D) with the following formula:

$$D = P-AED$$
where P is precipitation. Soil moisture is often considered to be the main driver of vegetation productivity, particularly in semi-arid regions [@Li2022]. Hence, we used the Standardized Soil Moisture Index (SSI) to estimate soil moisture (SM) [@Hao2013]. For SSI, we used the average soil moisture from ERA5L at a depth of 1m. All calculated indices are multi-scalar and can be used for the analysis of short- to long-term droughts.

To derive the drought indices, we first calculate the sum of the variables with regard to the time scale(s). In this case, for generalization purposes, we will use $V$, referring to variables P, AED, D,  and SM (Table  SSX). We accumulated each over the time series of values (months), and for the time scales s:

$$A^s_i = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  $$ {#eq-sumvar}

The $A^s_i$ corresponds to a moving window (convolution) that sums the variable for time scales $s$. This is summed over $s$ months, starting from the most recent month ($n$) back in time until month $n-s+1$. For example, using as a variable the precipitation, a period of twelve months ($n$), and a time scale of three months ($s$):

$$
\begin{split}
A^3_1 &= P_{oct} +P_{nov} +P_{dic} \\
\vdots\,\,\, &= \,\,\,\vdots + \,\,\,\vdots + \,\,\,\vdots \\
A^3_{10} &= P_{jan}+P_{feb} +P_{mar}
\end{split}
$$

Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i^s$ to derive the $P(A_i^s)$ probabilities across a period of interest:

$$P(A^s_i) = \frac{i-0.33}{n+0.33'}$$ {#eq-probPai}

An inverse normal approximation [@Abramowitz1968] obtains the empirically derived probabilities once the variable cumulates over time for the scale $s$. Thus, the drought indices $SPI$, $SPEI$, $EDDI$, and $SSI$ are  obtained following the equation:

$$DI(A^s_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}$$ {#eq-DI}

$DI$ is referring to the drought index calculated for the variable $V$. The values for the constats are: $C_0 = 2.515517$, $C_1 = 0.802853$, $C_2 = 0.010328$, $d_1 = 1.432788$, $d_2 = 0.189269$, and $d3 = 0.001308$. For $P(A^s_i) \leq 0.5$, W=$\sqrt{-2\cdot ln(P(A^s_i))}$ , and for $P(A^s_i) > 0.5$, replace $P(A^s_i)$ with $1-P(A^s_i)$ and reverse the sign of $DI(A^s_i)$.

The drought indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months at a monthly frequency for 2000–2023.

### Temporal trends of drought indices

To determine if there are statistically significant positive or negative temporal trends for the drought indices, we used the non-parametric modified Mann-Kendall test for serially correlated data [@Yue2004]. To determine the magnitude of the trend, we used Sen’s slope [@Sen1968]. Sen's slope is less affected by outliers than parametric ordinary least squares (OLS) regression, and as a non-parametric method, it is not influenced by the distribution of the data. We applied both methods for SPI, EDDI, SPEI, SETI, and SSI and six time scales, resulting in a total of 30 trends. We then aggregated temporal trends for each ecoregion and land cover type.

### Vegetation productivity

We also used the MODIS product to calculate vegetation productivity, and calculated anomalies of cumulative NDVI using  zcNDVI [@Zambrano2018], which was derived from the monthly time series of NDVI, with Equations 2 and 4. For vegetation productivity, we selected the time scale that best correlates with annual net primary productivity (NPP) across continental Chile. For this purpose, we calculated zcNDVI for time scales of 1, 3, 6, and 12 months (from December) and compared it with the annual NPP. We obtained NPP from MOD17A3HGF90. We chose to use six months because the $R^2$ between zcNDVI and NPP reaches its highest value at six months, obtaining an $R^2$ of 0.31 for forest and 0.72 for shrubland (Supplementary Information Section S2). We subsequently used zcNDVI with a time scale of 6 months and calculated it at a monthly frequency for 2000–2023.

## Drought impacts on vegetation productivity

For each land cover, we analyzed the trend of vegetation productivity. To this end, we identified areas within each land cover that are persistent over time to reduce the possibility that trends in vegetation productivity may be influenced by changes in  land cover. We examined the correlation between drought indices and vegetation productivity across land cover types to determine to the extent to which soil moisture and water demand and supply affect vegetation productivity.

We estimated pixel-to-pixel Pearson’s correlations between drought indices at time scales of 1, 3, 6, 12, 24, and 36 months with zcNDVI.  We extracted the Pearson correlation coefficient corresponding to the time scale with the highest value. For each index, we then generated two maps: 1) a raster with values of the time scales and drought index that reached the maximum correlation, and 2) a raster with the magnitude of the correlation between the drought index and vegetation productivity.

## Drought impacts on land cover change

### Land cover change

To analyze land cover change, we used the classification scheme of the International Geosphere-Biosphere Programme (IGBP) from the product MCD12Q1 Collection 6.1 from MODIS. The MCD12Q1 product is produced for each year from 2001 to 2023 and defines 17 classes (see Table S1). Following the FAO classification [@FAO2022], we classified native and planted forests as "forests", which represent natural and productive ecosystems dominated by large trees. To analyze the land cover change, we use the IGBP scheme from the MCD12Q1 product. We regrouped the 17 classes into ten macro-classes, as follows: 1-4 to forests (native forest and plantations), 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 as water (Table S3). This resulted in a time series of land cover with ten macro-classes for 2001-2023. We validated the land cover macro-classes using a high resolution (30 m) land cover map for 2013-2014 [@Zhao2016]. Our results showed a global accuracy of ~0.82 and a F1 score of ~0.66 (Supplementary Information, S2).

We calculated the area for each land cover class in the five study regions for 2001–2023. We then estimated the temporal change in area for each land cover type and macro-class, and determined the statistical significance (p-value < 0.05) and magnitude of the trend as described above.

To assess how water demand and supply, and soil moisture affect the variation in vegetation productivity across various land cover types, we avoid analyzing areas that experienced major land cover changes in the 2001–2023 period. To assess how zcNDVI varied irrespective of land cover change, we developed a persistence mask for land cover, which only retains pixels for which the macro-class remained the same for at least 80% of the 23 years (Fig. 1d).

## Relationship between land cover and drought trends

To identify which drought indices and time scales have a major impact on changes in land cover type, we examined the relationships between the temporal trends in the surface of land cover classes, drought indices, road density, burned area, and night lights, and  for each ecoregion. We performed the analysis at the sub-basin scale, using 485 river basins, which have a surface area between 0.906 and 24,408 km2 and a median area of 1,249 km2 (Supplementary Fig. S8/Table S5). For each basin, we calculated the trend per land cover, considering the proportion of the type relative to the total surface of the basin. For each basin we extracted the average trend of all drought indices and at time scales of 1, 3, 6, 12, 24, and 36 months. In the case of burned area, we used as variables the total and the trend of burned area for 2002-2023, and for night lights we used the average and the trend of nightlights for 2012-2023.

Prior to modelling relationships between trends in land cover and drought indices, we assessed multi-collinearity among explanatory variables, i.e., drought indices, road density, night lights, and burned area, with the variance inflation factor (VIF).  We analyzed the VIF for all drought indices at time scales of 1, 3, 6, 12, 24, and 36 separately because each index has a strong correlation across time scales. As VIF values greater than five may affect the interpretation of model results [@Dorman2013], we therefore excluded SPI from all subsequent models.

To assess the relationship between land cover trends and drought indices, we modeled trends in the surface of land cover types. We made a regression analysis using the random forest method93, which employs multiple decision trees. Some advantages of random forest include the ability to find non-linear relationships, reduce overfitting, and derive variable importance. We incorporated the trends of the five drought indices (SPI, SPEI, EDDI, SETI, and SSI), the trends of night lights (2) and burned area (2), the road density, for a total of ten predictors. We then constructed random forest models for each time scale (1, 3, 6, 12, 24, and 36) and each land cover class (forest, grassland, shrubland, savanna, cropland, and barren land), resulting in a total of 36 RF models. We trained each model using 1000 trees, setting the minimum number of nodes per decision tree at five and the number of predictors per split (boosting) to the square root of the total number of predictors. To account for uncertainty, we trained all the models ten times using a resampling strategy (ten folds) in a cross-validation scheme. Finally, we evaluate model fit by calculating R², root mean square error (RMSE), and variable importance. Variable importance identifies which variables have a higher contribution to explaining model variation. We calculated variable importance by permuting out-of-bag (OOB) data per tree and calculating the mean standard error of the OOB data. After permuting each predictor variable, we repeated the process for the remaining variables. We repeated this process ten times per fold to assess  model fit.

Finally, we visually explored the relationship between drought indexes and changes in land cover across sub basins within Chile.  To achieve this, we compared the relative changes in land cover surface with the drought indices, burned area, nighlights, and road density and other variables for the time scale that were deemed more significant in the random forest model.

## Software

For downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, R [@R2023]. For downloading ERA5L, we used the {ecmwfr} package [@Hufkens2019]. For processing raster data, we used {terra} [@Hijmans2023] and {stars} [@Pebesma2023]. For managing vectorial data, we used {sf} [@Pebesma2018]. For the calculation of AED, we used {SPEI} [@Bergueria2023]. For mapping, we used {tmap} [@Tennekes2018]. For data analysis and visualization, the suite {tidyverse} [@Wickham2019] was used. For the random forest modeling, we used the {tidymodels} [@Kuhn2020] and {ranger} [@Wright2017] packages.

# Results

## The Chilean matorral and Patagonian steppe increase atmospheric water demand but decrease vegetation evapotranspiration

For the Atacama desert and the Central Andean Puna, we found a positive temporal trend for drought indices of water supply (i.e., SPI, SSI), atmospheric water demand (i.e., EDDI), and vegetation water demand (i.e., SETI). For the Chilean matorral and Patagonian steppe, EDDI presents the higher positive trend and lower negative trends in SPI, SPEI, SSI, and SETI. This reflects a critical scenario of drought, where a rise in temperature increases atmospheric water demand, but vegetation cannot increase evapotranspiration due to a lack of water availability. In the Southern Andean steppe, there is a positive trend in AED (i.e., EDDI), but a negative in water supply (i.e., SPI, SPEI, SSI). The vegetation water demand (i.e., SETI) has negative trends, but it is increasing at higher time scales. The Valdivian temperate forests show a negative trend in water supply (i.e., SPI, SPEI, and SSI) and a positive trend in both AED and ET, as shown by the EDDI and SETI, respectively. Here, an increase in AED implies an increase in ET, likely due to a greater availability of water, unlike in the Chilean Matorral and Patagonian steppe. The vegetation water demand (SETI) in the Magellanic subpolar forests does not exhibit a significant trend over any given time scale. The AED and water supply present a positive trend. The trends of drought indices in the Patagonian steppe exhibit a similar behavior to the Chilean matorral, albeit to a lesser extent. Generally, the majority of the indices indicate that the trend (positive or negative) intensifies over longer time periods. (Fig. 2)

## Vegetation productivity has strongly decreased in the Chilean matorral and the Patagonian steppe.

We found contrasting temporal trends in vegetation productivity for 2000-2023 (Fig. 3). The Atacama desert does not exhibit significant trends over time. The Chilean Matorral, Patagonian steppe, and the Southern Andean steppe exhibit negative trends of -0.023, -0.016, and -0.006 (z-score/per decade), respectively.  In contrast, the Central Andean dry puna, Valdivian temperate forests, and Central Andean dry puna show positive trends ranging from 0.01 to 0.03 (z-score/per decade). The Chilean matorral was at its lowest point from 2019 to 2022, while the Patagonian steppe has experienced an increasingly negative trend since 2022.

Forest, savanna, and shrubland exhibit the highest change in surface area across ecoregions
We also observed significant changes in land cover surfaces across continental Chile (Fig. 4). The forest surface has increased in the Chilean matorral and in the Valdivian temperate forest at rates of 78 and 316 km² yr⁻¹, respectively. Grassland surface has diminished in the Southern Andean steppe (-19 km² yr⁻¹) and has increased in the Patagonian steppe (90 km² yr⁻¹). Savanna exhibits a decrease in the Chilean matorral of -271 km² yr⁻¹ and the Valdivian temperate forest of -276 km² yr⁻¹, but an increase at a rate of 133 km² yr⁻¹ in the Magellanic subpolar forest. Shrubland has the highest increase in the Chilean matorral (160 km² yr⁻¹). Barren land has increased in the Central Andean dry puna (36 km² yr⁻¹) and the Southern Andean steppe (50 km² yr⁻¹), but has diminished in the Magellanic subpolar forest (-81 km² yr⁻¹). 
## Drought impact on vegetation productivity are strongest in the Chilean matorral and Valdivian temperate forest

Our results indicate that drought impacts on vegetation productivity are highest in the Chilean Matorral and Valdivian temperate forests across all land cover types, except forest (Fig. 5, Fig. S5 and Table 1). For time scales of 6 and 12 months, SETI and SSI have the strongest positive correlation with vegetation productivity among the land cover types. Next, we found that grassland and savanna in the Patagonian steppe had higher correlations with SPI and SSI over 12 months. Further, there is a positive relationship between the vegetation in the Atacama desert and drought indices of 12 months of water supply and vegetation water demand. However, there is a negative relationship between the vegetation and atmospheric water demand over 12 months. All drought indices show a positive correlation with the vegetation in the Central Andean dry puna, particularly the drought indices of water supply (SPI, SPEI, and SSI) at time scales of 24 and vegetation water demand (SETI) at time scales of 36 months. For the Southern Andean steppe the SETI of 24 months showed the highest correlation with savannas, followed by the EDDI of 24 months. 
Our analysis also revealed that water demand and supply differentially affected the time scales at which vegetation productivity of land cover types within each region was most impacted by drought (Fig. 5, Fig. S5 and Table 1). While the spatial variation in the relationship between drought intensity and vegetation productivity was consistent across drought indices, the drought indices that captures water supply via soil moisture (Standardized Soil Moisture Index; SSI), and via vegetation water demand (Standardized Evapotranspiration Index, SETI)  tended to show a stronger correlation with vegetation productivity over larger areas than the other drought indices (Fig. 5).

## Drought strongly impacts land cover distribution for shrubland

Our random forest models show that drought indices explain between 32-79% of the variation in land cover change across continental Chile (Fig. 6). Moreover, these results highlight the importance of considering water supply and demand, as drought indices associated with both aspects of the water balance had high importance values across most ecoregions and land cover types. The variation in the time scale of drought indices that provide the strongest correlation with vegetation productivity also suggests that different types of vegetation are not equally sensitive to droughts of similar intensities (Table 1). 
The RF models show that the drought indices are explaining 71-78% of the variability in land cover surface change for shrublands. Second, the RF models explain approximately 58-78% of the variability in change on croplands. In the case of other land cover types, the RF models account for approximately 33-59% of the variability, with drought indices explaining less for the forest type (Fig. 6).

Our model has the highest r-squared for shrublands, followed by croplands, and barren land (Fig. 6). Ecoregions most frequently observed the variables SETI and SSI, which significantly influenced surface changes for each land cover (Fig. 7). In cropland, the primary factor is not the drought indices, but rather the total surface or trend of the burned area. The nightlight variable, which reflects urban areas, primarily explains the change in the surface of barren land, followed by the drought indices SPEI at time scales of 3 and 6 months (Fig. 7).

Our results also show that drought intensity was associated with the magnitude and direction of land cover change (Fig. 8).  We observe that shrublands are sensitive to both increases and decreases in SETI and SSI, reaching a point of equilibrium around a normal climatic situation (drought index = 0). This could potentially stem from the fact that favorable water supply conditions alter the type and quantity of vegetation. Conversely, a reduction in water supply results in the shrublands losing their vegetation and transforming into bare soil. Thereby, both situations alter the land cover type. Changes in the burned surface impact the cropland surface; an increase in the burned area leads to an increase in the agricultural surface. Agricultural areas likely replace the burned area, explaining this phenomenon. In the case of bare soil, values that indicate low urban development (nightlights) are associated with a larger surface of bare soil. However, when the variable's range extends beyond the normal urban situation (zero value), an increase in bare soil correlates with a fluctuating increase in urban development. The relationship between SETI and SPEI in grasslands is opposite; an increase in SPEI leads to an increase in grassland area, while an increase in water demand by grasslands, as reflected by SETI, results in a decrease in grassland area.

# Discussion

## Temporal trends in water supply and demand

We discovered that the Atacama desert, Central Andean dry puna, and the Magellanic subpolar forests experience an increase in water supply (SPI, SSI), as well as an increase in atmospheric and vegetation water demand (EDDI, SETI). However, in the Magellanic subpolar forests, there is no significant increase or decrease in SETI across time scales. Also, we found a significant decreasing trend in water supply (SPI, SPEI, and SSI) across the Southern Andean steppe, Chilean Matorral [@Boisier2018;@Sarricolea2019], Valdivian temperate forests, and Patagonian steppe, accompanied by an increase in atmospheric water demand (EDDI). Our results indicate that the trend of water supply and atmospheric demand tends to decrease or increase more strongly over longer time scales—a trend that is consistent with the progressive intensification of drought severity across much of Chile, as observed in other regions facing long-term droughts [@Rashid2019;@Miro2023]. Simultaneously, we observed a divergent trend between the EDDI and SETI in various ecoregions. In the majority of ecoregions, a rise in atmospheric water demand (EDDI) typically leads to a rise in vegetation water demand (SETI). However, in the zones most affected by drought (Fig. 3 and Fig. 5), where water availability is limited, an increase in atmospheric water demand results in a decrease in vegetation water demand. The Chilean matorral and the Patagonian steppe are where this is most evident. When combined, our findings demonstrate a persistent drying trend in the Chilean matorral, the Patagonian steppe, and the Southern Andean steppe.  We attribute this trend to a simultaneous decrease in precipitation and an increase in atmospheric evaporative demand, which leads to a decrease in the water demand for vegetation in water-limited zones[@Pascoa2021].

## Temporal trends in vegetation productivity

The consequences of the persistent drying trend for ecosystems throughout continental Chile are manifold. First, the prolonged hydrological drought, i.e., precipitation deficit, has reduced groundwater storage (SSI, [@Taucare2024]), leading to a steady decline in vegetation productivity (zcNDVI) since 2000 across the Chilean Matorral, Patagonian steppe, and Southern Andean steppe, reaching its lowest level between 2020 and 2022, which could be due to either a decrease in vegetation area, a loss of biomass or browning in forest ecosystems. Reports for natural and productive ecosystems [@Nicolai-Shaw2017;@Jiang2020;@Zhou2021] most strongly associate this decline with declines in soil moisture and increases in evapotranspiration. Second, the sharp decline in vegetation productivity in the Chilean Matorral and Valdivian temperate forests showed that grasses and shrublands respond to shifts in water supply on higher time scales (12 months), in comparison to savanna and croplands (6 months). Also, in the Valdivian temperate forest that has a higher surface with forest trees, soil moisture (SSI) and vegetation water demand (SETI) at higher time scales, 12 and 36 months, respectively, evidence stronger correlations. Which is consistent with recent studies showing that progressive, long-term water deficits in central Chile have triggered forest browning and declines in native forest productivity  [e.g., @Miranda2023;@Miranda2020;@Venegas2022]. While our analysis do not distinguish between  native and planted forests, the latter of which are considered to be more drought tolerant in central and southern Chile [@Carrasco2022], we show that forest area declines more sharply in response to increasing water demand due to rising temperatures (EDDI) than decreasing water supply (e.g., SPI, SSI; refs. [@Fajardo2019;@Holz2018]), which may have cascading impacts on multiple facets of forest diversity [@Segovia2021;@Sabatini2022]

Moreover, the strengthening of the correlation between vegetation productivity and water supply (SPI, SPEI, SSI) or demand (EDDI, SETI) over multiple time scales (up to 36 months) and across land cover types (Fig. 5) - demonstrates the impacts of climate change on the water balance in Chile. Impacts may extend beyond vegetation productivity, as reduced soil moisture in central Chile and the western United States has increased wildfire activity[@Holden2018;@Gonzalez2018], which is a growing concern in Chile and may be further exacerbated by extensive plantations of highly flammable tree species, e.g., Eucalyptus spp. and Pinus spp. [@Bowman2019]. Third, we found that the decline in the vegetation productivity of croplands is due to a decrease in the water supply and vegetation water demand to a greater extent than to an increase in atmospheric water demand [@Quiring2010], causing a decline in water availability. This is consistent with evidence that more water-intensive crops have replaced less water-intensive crops in central Chile, leading to an increase in water extraction from rivers or groundwater [@Munoz2020;@Duran2020].

## Drought impacts  on land cover

We found evidence that temporal decreases in water supply (SPEI, SSI) and decreases in vegetation water demand (SETI) are driving shifts not only in vegetation productivity but also in land cover across most of continental Chile. Shrubland covers were the land cover type most sensitive to changes in the water balance over short and long temporal scales. In contrast, forest surfaces are the type less affected by drought. Moreover, our results provide evidence that, in addition to shrublands cover, other land cover types have been affected by water deficits, particularly savannas and grasslands. Our results therefore suggest that multiple land cover types could be vulnerable to regime shifts towards more drought tolerant land cover types [@Scheffer2001;@Martinez-Vilalta2016], such as forest, whose cover increased non-linearly in response to increasing drought intensity. Changes in cropland cover are not a direct consequence of drought (Fig. 7), but rather an indirect one, likely occupying burned areas and  possibly reflecting the decision of resource-poor farmers to migrate to regions with more abundant water resources or to change economic activity [@AghaKouchak2021;@Hermans2021]. In contrast, the increase in shrubland cover due to a decrease in savanna cover may be ecological, as shrubs may  be more drought tolerant than other growth forms[@Eldridge2011;@Gotmark2016].

We run each model ten times to account for the uncertainty in its performance. In the supplementary material (Fig. S12-S17), we show the error bar of r-squared; despite the inherent uncertainty, the analysis reinforces our results regarding how drought impacts land cover type.

Overall, our results show that long-term declines in water supply and demand have induced widespread, multi-dimensional impacts on the vegetation productivity and on the extent of land cover types. While prolonged droughts may directly cause shifts to more drought-tolerant land cover types, such as shrublands, they may also influence land cover change through human decision making and activities. This study extends current understanding of drought impacts by demonstrating how their multidimensionality emerges over multiple time scales and across land cover types, which can contribute to developing context-specific adaptation strategies for agriculture, biodiversity conservation, and natural resource management.

## Study limitations

This study has evaluated the impact of water supply and demand on vegetation productivity and how land use change influences it, with some important limitations to highlight. One of the major limitations of the study is the consideration of secondary information. For instance, we used estimates of water supply and demand, such as ERA5L and MODIS, which, despite their improved estimation capacity, suffer from biases and uncertainties [@Gomis2023;@Clelland2024] in different areas or climatic conditions. In this study, we compared the ERA5L data with climatic stations (see Table S2) to verify bias and uncertainty, but future studies will need to correct the products for a more precise measurement. We used zcNDVI [@Zambrano2018] (MODIS) as a proxy for vegetation productivity, which has proven to be a good estimate of NPP (see Fig. S1 and S2), but its quality varies between different types of vegetation.

To define land cover, products were used that estimate the classes using classification models, which are subject to quality errors that must be taken into account [@Stehman2019;@Verburg2011]. In addition, in our case we used macro classes of land cover, where, for example, the different types of forests were grouped (e.g., monoculture, native forest). This approach hinders our ability to understand the effects of drought on the various subclasses within each land use class. When it comes to cultivated areas, we cannot distinguish between rainfed and irrigated areas using macro classes. However, in this study, we aimed to provide a broad overview at a national scale, with the goal of focusing on a more detailed level in future research.

Other studies have analyzed how human decisions impact land use change, such as population increase [@Kleemann2017]. In our case, we attempted to consider this by incorporating road density, night lights, and burned areas. However, human decisions are complex to consider from the point of view of earth observation. On the other hand, earth observation tools can analyze land cover change, but each type's use depends on social and economic factors, which are challenging to quantify [@Rindfuss2004;@Jenner2024] and necessitate the integration of social, natural, and geographic information sciences.

# Data availability

The codes generated during the current study are available in the GitHub repository, https://github.com/FSEQ210022/drought_vegetation. The datasets generated and/or analyzed during the current study are available in the Zenodo repository, https://doi.org/10.5281/zenodo.10359547.

# Acknowledgments {-}

The National Research and Development Agency of Chile (ANID) funded this study through the drought emergency project FSEQ210022, Fondecyt Iniciación N°11190360, Fondecyt Postdoctorado N°3230678, and Fondecyt Regular N°1210526.

### 2025-01-12 article/article_scientific_reports/mansucript_11012025.tex
actualizar manuscrito quarto

--- 2025-01-12 article/article_scientific_reports/mansucript_11012025.qmd
@@
----
-title: Shifts in water supply and demand drive land cover change across Chile
-format:
-  nature-pdf:
-    # journal.cite-style is included in the tex file but ignored by pandoc if 
-    # cite-method is not `natbib`.
-    journal:
-      cite-style: sn-nature
-    classoption:
-      - numbered
-    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
-    # to use the bst styles from the upstream template.
-    cite-method: natbib
-    keep-tex: true
-  html:
-    code-fold: true
-    html-math-method: katex
-  docx: default
-# The `csl` style will be used with LaTeX-based rendering of`nature-pdf`
-# if cite-method is `citeproc`. This is also the default if cite-method
-# is not specified.
-csl: _extensions/christopherkenny/nature/csl/springer-vancouver.csl
-author:
-  - name: Francisco Zambrano
-    email: francisco.zambrano@umayor.cl
-    affiliations: 
-        - name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad, Universidad Mayor
-          id: 1
-          city: Santiago
-          country: 7500994, Chile.
-        - id: 2
-          name: Observatorio de Sequía para la Agricultura y la Biodiversidad de Chile (ODES), Universidad Mayor
-          city: Santiago
-          country: 7500994, Chile.
-    attributes:
-        corresponding: true
-    #note: This is the first author footnote.
-  - name: Anton Vrieling
-    #email: a.vrieling@utwente.nl
-    affiliations:
-        - id: 3
-          name: Faculty of Geo-Information Science and Earth, University of Twente
-          city:  Enschede
-          country: The Netherlands.
-  - name: Francisco Meza
-    affiliations:
-        - id: 4
-          name: Facultad de Agronomía y Sistemas Naturales, Pontificia Universidad Católica de Chile. 
-          city: Santiago
-          country: Chile.
-        - id: 5
-          name: Instituto para el Desarrollo Sustentable. Pontificia Universidad Católica de Chile
-          city: Santiago
-          country: Chile.
-        - id: 6
-          name: Centro Interdisciplinario de Cambio Global, Pontificia Universidad Católica de Chile
-          city: Santiago
-          country: Chile.
-  - name: Iongel Duran-Llacer
-    affiliations:
-        - id: 7
-          name: Hémera Centro de Observación de la Tierra, Facultad de Ciencias, Universidad Mayor,
-          city: Santiago
-          country: 7500994, Chile.
-  - name: Francisco Fernández
-    #email: francisco.fernandez@uss.cl
-    affiliations:
-        - id: 8
-          name: Center of Economics for Sustainable Development (CEDES), Faculty of Economics and Government, Universidad San Sebastian
-          city: Santiago
-          country: Chile.
-        - id: 9
-          name: Center of Applied Ecology and Sustainability (CAPES)
-          city: Santiago
-          country: Chile.
-  - name: Alejandro Venegas-González
-    affiliations:
-        - id: 10
-          name: Instituto de Ciencias Agroalimentarias, Animales y Ambientales (ICA3), Universidad de O'Higgins
-          city: San Fernando
-          country: Chile.
-  - name: Nicolas Raab
-    affiliations:
-        - id: 4
-  - name: Dylan Craven
-    affiliations:
-        - id: 11
-          department: GEMA Center for Genomics, Ecology & Environment, Universidad Mayor, Camino La Pirámide Huechuraba 5750
-          city: Santiago
-          country: Chile.
-        - id: 12
-          name: Data Observatory Foundation
-          city: Santiago
-          country: Chile.
-abstract: |
-  Globally, droughts are becoming longer, more frequent, and more severe, and their impacts are multidimensional.
-These impacts typically extend beyond the water balance, as long-term, cumulative changes in the water balance can lead to regime shifts in land cover.
-Here, we assess the effects of temporal changes in water supply and demand over multiple time scales on vegetation productivity and land cover changes in continental Chile, which has experienced a severe drought since 2010.
-Over most of continental Chile, we have observed a persistent negative trend in water supply and a positive trend in atmospheric water demand since 2000.
-However, in water-limited zones, we have observed a negative trend in the vegetation's water demand.
-All the trends intensify over longer time scales.
-This long-term decrease in water availability has led to a decrease in vegetation productivity, especially for the Chilean matorral and the Valdivian temperate forest.
-Drought indices of soil moisture and actual evapotranspiration from short- to mid-term time scales primarily explain this decrease.
-Besides, our models indicate that the intensity of the drought explains up to 78% of the changes in shrubland land covers, decreasing to 40% for forests, but the change in cropland surface is explained by the surface burned.
-Long-term climate change may cause regime shifts in land cover, which context-specific adaptation strategies can mitigate.
-keywords: [drought, land cover, water demand, water supply]
-bibliography: referencias.bib  
----
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%
+\documentclass[
+  sn-nature,
+  numbered]{sn-jnl}
 
-# Introduction {#sec-intro}
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% Make \paragraph and \subparagraph free-standing
+\makeatletter
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+\makeatother
 
-Across many regions of the world, droughts are becoming longer, more frequent, and more severe[@Miranda2023; @IPCC2023], impacting ecosystems via tree mortality[@Cheng2024], reducing vegetation productivity[@Miranda2023] and inducing shifts in land use and cover [@Crausbay2017].
-However, identifying drought events is idiosyncratic due to the varying criteria used for classification.
-Droughts can be classified as 1) meteorological, i.e., when precipitation in a specific period falls below mean precipitation values observed over multiple years [@Mishra2010] (usually more than 30 years); 2) hydrological, i.e., when precipitation anomalies last for long periods (months to years) and affect the  hydrological system [@VanLoon2015;@VanLoon2016b] (e.g., streamflows, reservoirs and groundwater); 3) agricultural, i.e.
-when precipitation deficits negatively impact plant health, leading to decreases in crop or pasture productivity [@Wilhite1985]; or 4) ecological, i.e., when water availability negatively affect the provisioning of ecosystem services and trigger feedbacks in natural or human systems [@Crausbay2017].
-Such feedbacks include drought impacts on human decision making and activities, which can lead to land-cover change [@VanLoon2016;@AghaKouchak2021], which may have cascading effects on biodiversity and ecosystem services [e.g., @Lawler2014;@Newbold2015].
-Despite the high degree of confidence in the impacts of rising temperatures on the extent, frequency, and severity of agricultural and ecological droughts [@IPCC2023], which are likely to increase even if global warming stabilizes at 1.5°–2°C, the severity of meteorological droughts has been remarkably stable globally over the past century [@Vicente-Serrano2022;@Kogan2020].
-A global study analyzing drought severity trends from 1980 to 2020 reveals that in a few regions (some mid-latitudinal and subtropical areas), rising temperatures during the warm season have increased atmospheric evaporative demand (AED), leading to an increase in agricultural land area [@Vicente-Serrano2022].
-Thus, rising atmospheric water demand may reflect parallel changes in land cover—primarily agriculture—that can exacerbate the effects of meteorological droughts on ecosystems.
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 
-Expanding analyses to include multiple dimensions of droughts can provide complementary insights into the Earth’s water balance - and its impacts - over multiple time scales.
-Yet, the World Meteorological Organization recommends the use of a single drought index for monitoring droughts [@WMO2012], i.e., the multi-scale Standardized Precipitation Index (SPI; [@McKee1993]), which is limited in that it only considers  water supply in the form of precipitation.
-The Standardized Precipitation Evapotranspiration Index (SPEI; [@Vicente-Serrano2020]) builds upon SPI by incorporating the effects of  temperature on drought, and is now used widely for drought monitoring [e.g., @Gebrechorkos2023;@Liu2024].
-Indices derived from soil moisture products [@Narasimhan2005; @Souza2021], such as the Standardized Soil Moisture Index [SSI, @AghaKouchak2014;@AghaKouchak2015] also monitor water supply and are thought to better capture water availability for crops, thus providing more relevant information for evaluating agricultural droughts.
-To disentangle the effects of precipitation from those of temperature [@Vicente-Serrano2020], as well as to capture droughts in terms of water atmospheric demand, AED has been integrated into the Evaporative Demand Drought Index [EDDI, @McEvoy2016], which is particularly effective at detecting the rapid onset or intensification of droughts.
-To quantify vegetation water demand, we must use the actual evapotranspiration, or the amount of water removed from a surface by evaporation and transpiration; the Standardized Evapotranspiration Index (SETI; [@Yang2019]) can be used for this purpose.
-In turn, ecological droughts, which capture the joint effects of precipitation and temperature in modifying natural and productive ecosystems [@Camps-Valls2021;@Paruelo2016;@Helman2014], are complex to measure and can therefore be monitored using multiple drought indices that capture the multiple dimensions of drought, e.g., precipitation, temperature, evapotranspiration, and AED.
-Although such an approach accounts for the joint effects of changes in natural and productive ecosystems, its potential impacts on land cover change have been largely unexplored.
+%%%% Standard Packages
 
-From 1960 to 2019, land-use change has impacted approximately one-third of the Earth’s surface, which is four times more than previously thought [@Winkler2021].
-Despite the considerable interest in land-use change dynamics [e.g., @Song2018;@Winkler2021], the direction and magnitude of  drought impacts on land cover change and vegetation productivity remain uncertain [@Chen2022;@Akinyemi2021;@Peng2017].
-Meteorological droughts are responsible for approximately 37% of land cover change and variability in vegetation productivity globally [@Peng2017].
-However, the evidence supporting these results is derived from only one drought index, SPEI, which combines a proxy for water supply -  precipitation - with a proxy for water demand - AED - at one time scale (12 months).
-The use of only one time scale may bias results of drought impacts towards ecosystems dominated by plant growth forms such as grasses and herbs that respond more rapidly to drought stress (< 12 months).
-This is because  physiological differences among and within dominant plant growth forms may increase (or decrease) tolerance of drought stress [@Craine2013;@McDowell2022].
-For example, trees growing in more arid ecosystems typically respond over longer time scales than those in more humid ecosystems [@Vicente-Serrano2014].
-Another source of uncertainty regarding drought impacts on land cover change and vegetation productivity are extrinsic factors, such as large-scale public policy (e.g., national and international reforestation initiatives), agricultural practices (e.g., clearing forest for soybean or oil palm), and rural and urban land use planning [@Karabulut2023].
+\usepackage{graphicx}%
+\usepackage{multirow}%
+\usepackage{amsmath,amssymb,amsfonts}%
+\usepackage{amsthm}%
+\usepackage{mathrsfs}%
+\usepackage[title]{appendix}%
+\usepackage{xcolor}%
+\usepackage{textcomp}%
+\usepackage{manyfoot}%
+\usepackage{booktabs}%
+\usepackage{algorithm}%
+\usepackage{algorithmicx}%
+\usepackage{algpseudocode}%
+\usepackage{listings}%
 
-To deepen current knowledge on the multidimensional impacts of drought on the temporal dynamics of natural and productive ecosystems, we evaluate temporal changes in water supply and demand, net primary productivity, and land-cover change across terrestrial ecosystems in continental Chile for 2000-2023.
-Chile’s diverse climate and ecosystems [@Beck2023;@Luebert2022] make it an ideal natural laboratory for assessing the dynamic interactions between climate and ecosystems, and potential impacts on land-cover change.
-Additionally, large parts of Chile have experienced severe drought conditions that have significantly affected vegetation and water storage in recent years; north-central Chile has faced a persistent precipitation deficit (or "mega-drought") since 2010 [@Garreaud2017], which has broadly impacted native forests [e.g., @Miranda2020;@UrrutiaJalabert2018;@Venegas2018] and agricultural productivity [e.g., @Zambrano2016;@Zambrano2018;@Zambrano2023].
-However, the effects of this prolonged extreme drought may also extend to changes in land cover, altering the provision of key ecosystem services and agricultural production.
-Here, we aim to assess: short- to long-term time trends in multi-scalar drought indices that capture variation in the components of water balance, i.e., water supply (SPI, SPEI, SSI) and demand (EDDI, SETI) and their impacts on vegetation productivity and land-cover change across continental Chile.
-By doing so, we aim to deepen understanding of how water supply and demand affect vegetation productivity and influence changes in the surface of land cover in a severely drought-impacted region.
-This may enhance understanding of the alterations, due to drought, to ecosystems in various regions across the world.
+%%%%
 
-# Materials and Methods
+\raggedbottom
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
+\fi
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
+\fi
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
+\else
+  \newcommand\listtablename{List of Tables}
+\fi
+\ifdefined\figurename
+  \renewcommand*\figurename{Figure}
+\else
+  \newcommand\figurename{Figure}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
 
-## Study area
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\usepackage[]{natbib}
+\bibliographystyle{plainnat}
+\usepackage{bookmark}
 
-Continental Chile has a diverse climate, with strong environmental gradients from north to south and east to west [@Aceituno2021] (Fig.
-1a), which, together with its complex topography (Fig.
-1b), determine its ecosystem diversity [@Garreaud2009;@Luebert2022] (Fig.
-1c).
-We therefore divided Chile into ecoregions [@Dinerstein2017], which are regions that share similar geography and ecology, and have comparable levels of precipitation and solar radiation.
-There are seven ecoregions: Atacama desert, Central Andean dry puna, Southern Andean steppe, Chilean Matorral, Valdivian temperate forests, Magellanic subpolar forests, and Patagonian steppe.
-The Atacama desert is predominantly arid with hot (Bwh in the Koppen-Geiger classification) and cold (Bwk) temperatures, as well as the northern part of the Chilean Matorral.
-Most of the land in these two northern regions is bare, except for a small area where shrublands and grasslands are present.
-With an annual rainfall of less than 400 mm, the Central Andean dry puna ecoregion has low, yet highly seasonal precipitation with an eight-month dry season, low temperatures (Bwk) and is dominated by grasslands, shrublands, and savanna.
-The climate of the Southern Andean steppe ecoregion is cold desert (BWk), with most precipitation occurring in the winter.
-There is little vegetation in this ecoregion because the plants have adapted to its windy, dry, and cold climate.
-In central Chile, the climate of the Chilean Matorral  changes to that of an arid steppe with cold temperatures (Bsk).
-Then, towards the center-south of the country, the climate of the Chilean Matorral changes to a Mediterranean climate, with warm to hot summers (Csa and Csb).
-Land cover in this ecoregion consists of a significant amount of shrublands and savannas.
-The Valdivian temperate forests have a mostly oceanic climate (Cfb) and a large area of forests and grasslands.
-The Magellanic subpolar forests have a tundra climate.
-Lastly, the Patagonian steppe has high aridity, cold temperatures (Bsk), and primarily consists of grasslands.
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={Shifts in water supply and demand drive land cover change across Chile},
+  pdfauthor={Francisco Zambrano; Anton Vrieling; Francisco Meza; Iongel Duran-Llacer; Francisco Fernández; Alejandro Venegas-González; Nicolas Raab; Dylan Craven},
+  pdfkeywords={drought, land cover, water demand, water supply},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
 
-## Data
+\title[Shifts in water supply and demand drive land cover change across
+Chile]{Shifts in water supply and demand drive land cover change across
+Chile}
 
-### Gridded meteorological and vegetation data
+% author setup
+\author*[1,2]{\fnm{Francisco} \sur{Zambrano}}\email{francisco.zambrano@umayor.cl}\author[3]{\fnm{Anton} \sur{Vrieling}}\author[4,5,6]{\fnm{Francisco} \sur{Meza}}\author[7]{\fnm{Iongel} \sur{Duran-Llacer}}\author[8,9]{\fnm{Francisco} \sur{Fernández}}\author[10]{\fnm{Alejandro} \sur{Venegas-González}}\author[4]{\fnm{Nicolas} \sur{Raab}}\author[11,12]{\fnm{Dylan} \sur{Craven}}
+% affil setup
+\affil[1]{, \orgname{Hémera Centro de Observación de la Tierra, Facultad
+de Ciencias, Escuela de Ingeniería en Medio Ambiente y Sustentabilidad,
+Universidad Mayor}}
+\affil[2]{, \orgname{Observatorio de Sequía para la Agricultura y la
+Biodiversidad de Chile (ODES), Universidad Mayor}}
+\affil[3]{, \orgname{Faculty of Geo-Information Science and Earth,
+University of Twente}}
+\affil[4]{, \orgname{Facultad de Agronomía y Sistemas Naturales,
+Pontificia Universidad Católica de Chile.}}
+\affil[5]{, \orgname{Instituto para el Desarrollo Sustentable.
+Pontificia Universidad Católica de Chile}}
+\affil[6]{, \orgname{Centro Interdisciplinario de Cambio Global,
+Pontificia Universidad Católica de Chile}}
+\affil[7]{, \orgname{Hémera Centro de Observación de la Tierra, Facultad
+de Ciencias, Universidad Mayor,}}
+\affil[8]{, \orgname{Center of Economics for Sustainable Development
+(CEDES), Faculty of Economics and Government, Universidad San
+Sebastian}}
+\affil[9]{, \orgname{Center of Applied Ecology and Sustainability
+(CAPES)}}
+\affil[10]{, \orgname{Instituto de Ciencias Agroalimentarias, Animales y
+Ambientales (ICA3), Universidad de O'Higgins}}
+\affil[4]{}
+\affil[11]{\orgdiv{GEMA Center for Genomics, Ecology \& Environment,
+Universidad Mayor, Camino La Pirámide Huechuraba 5750}}
+\affil[12]{, \orgname{Data Observatory Foundation}}
 
-To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the MOD13A3 Collection 6.1 product derived from the MODIS (Moderate-Resolution Imaging Spectroradiometer) sensor onboard the Terra satellite.
-MOD13A3 provides vegetation indices with a 1 km spatial resolution and monthly frequency [@Didan2015].
-We also utilized the MOD16A2 collection 6.1 [@Running2017] product from MODIS to gauge the water consumption of vegetation.
-This product gives us monthly actual evapotranspiration (ET) and atmospheric evaporative demand (AED) with a ~500m spatial resolution.
-For soil moisture, water supply, and water demand variables, we used ERA5-Land (ERA5L; ECMWF Reanalysis version 5 over land) [@MunozSabater2021], a reanalysis dataset that provides atmospheric and land variables since 1950.
-It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage.
-We selected total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100 cm of depth (layer 1 to layer 3; Supplementary Materials and Methods, Supplementary Tables 2 and 4).
+% abstract
 
-### Gridded indicators for land use
+\abstract{Globally, droughts are becoming longer, more frequent, and
+more severe, and their impacts are multidimensional. These impacts
+typically extend beyond the water balance, as long-term, cumulative
+changes in the water balance can lead to regime shifts in land cover.
+Here, we assess the effects of temporal changes in water supply and
+demand over multiple time scales on vegetation productivity and land
+cover changes in continental Chile, which has experienced a severe
+drought since 2010. Over most of continental Chile, we have observed a
+persistent negative trend in water supply and a positive trend in
+atmospheric water demand since 2000. However, in water-limited zones, we
+have observed a negative trend in the vegetation's water demand. All the
+trends intensify over longer time scales. This long-term decrease in
+water availability has led to a decrease in vegetation productivity,
+especially for the Chilean matorral and the Valdivian temperate forest.
+Drought indices of soil moisture and actual evapotranspiration from
+short- to mid-term time scales primarily explain this decrease. Besides,
+our models indicate that the intensity of the drought explains up to
+78\% of the changes in shrubland land covers, decreasing to 40\% for
+forests, but the change in cropland surface is explained by the surface
+burned. Long-term climate change may cause regime shifts in land cover,
+which context-specific adaptation strategies can mitigate.}
 
-To account for the impacts of human activity on land cover change, we obtained data on road density [@Meijer2018] and night lights for the period 2012–2023 [@Roman2018].
-These products are frequently used in the literature to quantify the human footprint (e.g., [@Halpern2015;@Halpern2008]) or biodiversity threats (e.g., [@Bowler2020;@Kennedy2020]).To capture changes on land cover due to fires, we calculated the total burned area for 2002-2023 [@Chen2023].
+% keywords
+\keywords{drought,  land cover,  water demand,  water supply}
 
-## Short- to long-term drought trends
+\begin{document}
+\maketitle
 
-### Atmospheric Evaporative Demand (AED)
+\section{Introduction}\label{sec-intro}
 
-To compute drought indices that use water demand, it is necessary to first calculate AED.
-To do this, we employed the Hargreaves method [@Hargreaves1994;@Hargreaves1985] by applying the following equation:
+Across many regions of the world, droughts are becoming longer, more
+frequent, and more severe\citep{Miranda2023, IPCC2023}, impacting
+ecosystems via tree mortality\citep{Cheng2024}, reducing vegetation
+productivity\citep{Miranda2023} and inducing shifts in land use and
+cover \citep{Crausbay2017}. However, identifying drought events is
+idiosyncratic due to the varying criteria used for classification.
+Droughts can be classified as 1) meteorological, i.e., when
+precipitation in a specific period falls below mean precipitation values
+observed over multiple years \citep{Mishra2010} (usually more than 30
+years); 2) hydrological, i.e., when precipitation anomalies last for
+long periods (months to years) and affect the hydrological system
+\citep{VanLoon2015, VanLoon2016b} (e.g., streamflows, reservoirs and
+groundwater); 3) agricultural, i.e.~when precipitation deficits
+negatively impact plant health, leading to decreases in crop or pasture
+productivity \citep{Wilhite1985}; or 4) ecological, i.e., when water
+availability negatively affect the provisioning of ecosystem services
+and trigger feedbacks in natural or human systems \citep{Crausbay2017}.
+Such feedbacks include drought impacts on human decision making and
+activities, which can lead to land-cover change
+\citep{VanLoon2016, AghaKouchak2021}, which may have cascading effects
+on biodiversity and ecosystem services
+\citep[e.g.,][]{Lawler2014, Newbold2015}. Despite the high degree of
+confidence in the impacts of rising temperatures on the extent,
+frequency, and severity of agricultural and ecological droughts
+\citep{IPCC2023}, which are likely to increase even if global warming
+stabilizes at 1.5°--2°C, the severity of meteorological droughts has
+been remarkably stable globally over the past century
+\citep{Vicente-Serrano2022, Kogan2020}. A global study analyzing drought
+severity trends from 1980 to 2020 reveals that in a few regions (some
+mid-latitudinal and subtropical areas), rising temperatures during the
+warm season have increased atmospheric evaporative demand (AED), leading
+to an increase in agricultural land area \citep{Vicente-Serrano2022}.
+Thus, rising atmospheric water demand may reflect parallel changes in
+land cover---primarily agriculture---that can exacerbate the effects of
+meteorological droughts on ecosystems.
 
-$$AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}$$ {#eq-AED}
+Expanding analyses to include multiple dimensions of droughts can
+provide complementary insights into the Earth's water balance - and its
+impacts - over multiple time scales. Yet, the World Meteorological
+Organization recommends the use of a single drought index for monitoring
+droughts \citep{WMO2012}, i.e., the multi-scale Standardized
+Precipitation Index (SPI; \citep{McKee1993}), which is limited in that
+it only considers water supply in the form of precipitation. The
+Standardized Precipitation Evapotranspiration Index (SPEI;
+\citep{Vicente-Serrano2020}) builds upon SPI by incorporating the
+effects of temperature on drought, and is now used widely for drought
+monitoring \citep[e.g.,][]{Gebrechorkos2023, Liu2024}. Indices derived
+from soil moisture products \citep{Narasimhan2005, Souza2021}, such as
+the Standardized Soil Moisture Index
+\citep[SSI,][]{AghaKouchak2014, AghaKouchak2015} also monitor water
+supply and are thought to better capture water availability for crops,
+thus providing more relevant information for evaluating agricultural
+droughts. To disentangle the effects of precipitation from those of
+temperature \citep{Vicente-Serrano2020}, as well as to capture droughts
+in terms of water atmospheric demand, AED has been integrated into the
+Evaporative Demand Drought Index \citep[EDDI,][]{McEvoy2016}, which is
+particularly effective at detecting the rapid onset or intensification
+of droughts. To quantify vegetation water demand, we must use the actual
+evapotranspiration, or the amount of water removed from a surface by
+evaporation and transpiration; the Standardized Evapotranspiration Index
+(SETI; \citep{Yang2019}) can be used for this purpose. In turn,
+ecological droughts, which capture the joint effects of precipitation
+and temperature in modifying natural and productive ecosystems
+\citep{Camps-Valls2021, Paruelo2016, Helman2014}, are complex to measure
+and can therefore be monitored using multiple drought indices that
+capture the multiple dimensions of drought, e.g., precipitation,
+temperature, evapotranspiration, and AED. Although such an approach
+accounts for the joint effects of changes in natural and productive
+ecosystems, its potential impacts on land cover change have been largely
+unexplored.
 
-where $Ra$ $(MJ\,m^2\, day^{-1})$ is extraterrestrial radiation; $T$, $T_{max}$, and $T_{min}$ are mean, maximum, and minimum temperature $(°C)$ at 2m.
-For calculating $Ra$ we used the coordinate of the latitud of the centroid of each pixel as follow:
+From 1960 to 2019, land-use change has impacted approximately one-third
+of the Earth's surface, which is four times more than previously thought
+\citep{Winkler2021}. Despite the considerable interest in land-use
+change dynamics \citep[e.g.,][]{Song2018, Winkler2021}, the direction
+and magnitude of drought impacts on land cover change and vegetation
+productivity remain uncertain \citep{Chen2022, Akinyemi2021, Peng2017}.
+Meteorological droughts are responsible for approximately 37\% of land
+cover change and variability in vegetation productivity globally
+\citep{Peng2017}. However, the evidence supporting these results is
+derived from only one drought index, SPEI, which combines a proxy for
+water supply - precipitation - with a proxy for water demand - AED - at
+one time scale (12 months). The use of only one time scale may bias
+results of drought impacts towards ecosystems dominated by plant growth
+forms such as grasses and herbs that respond more rapidly to drought
+stress (\textless{} 12 months). This is because physiological
+differences among and within dominant plant growth forms may increase
+(or decrease) tolerance of drought stress
+\citep{Craine2013, McDowell2022}. For example, trees growing in more
+arid ecosystems typically respond over longer time scales than those in
+more humid ecosystems \citep{Vicente-Serrano2014}. Another source of
+uncertainty regarding drought impacts on land cover change and
+vegetation productivity are extrinsic factors, such as large-scale
+public policy (e.g., national and international reforestation
+initiatives), agricultural practices (e.g., clearing forest for soybean
+or oil palm), and rural and urban land use planning
+\citep{Karabulut2023}.
 
-$$R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]$$ {#eq-Ra}
+To deepen current knowledge on the multidimensional impacts of drought
+on the temporal dynamics of natural and productive ecosystems, we
+evaluate temporal changes in water supply and demand, net primary
+productivity, and land-cover change across terrestrial ecosystems in
+continental Chile for 2000-2023. Chile's diverse climate and ecosystems
+\citep{Beck2023, Luebert2022} make it an ideal natural laboratory for
+assessing the dynamic interactions between climate and ecosystems, and
+potential impacts on land-cover change. Additionally, large parts of
+Chile have experienced severe drought conditions that have significantly
+affected vegetation and water storage in recent years; north-central
+Chile has faced a persistent precipitation deficit (or ``mega-drought'')
+since 2010 \citep{Garreaud2017}, which has broadly impacted native
+forests \citep[e.g.,][]{Miranda2020, UrrutiaJalabert2018, Venegas2018}
+and agricultural productivity
+\citep[e.g.,][]{Zambrano2016, Zambrano2018, Zambrano2023}. However, the
+effects of this prolonged extreme drought may also extend to changes in
+land cover, altering the provision of key ecosystem services and
+agricultural production. Here, we aim to assess: short- to long-term
+time trends in multi-scalar drought indices that capture variation in
+the components of water balance, i.e., water supply (SPI, SPEI, SSI) and
+demand (EDDI, SETI) and their impacts on vegetation productivity and
+land-cover change across continental Chile. By doing so, we aim to
+deepen understanding of how water supply and demand affect vegetation
+productivity and influence changes in the surface of land cover in a
+severely drought-impacted region. This may enhance understanding of the
+alterations, due to drought, to ecosystems in various regions across the
+world.
+
+\section{Materials and Methods}\label{materials-and-methods}
+
+\subsection{Study area}\label{study-area}
+
+Continental Chile has a diverse climate, with strong environmental
+gradients from north to south and east to west \citep{Aceituno2021}
+(Fig. 1a), which, together with its complex topography (Fig. 1b),
+determine its ecosystem diversity \citep{Garreaud2009, Luebert2022}
+(Fig. 1c). We therefore divided Chile into ecoregions
+\citep{Dinerstein2017}, which are regions that share similar geography
+and ecology, and have comparable levels of precipitation and solar
+radiation. There are seven ecoregions: Atacama desert, Central Andean
+dry puna, Southern Andean steppe, Chilean Matorral, Valdivian temperate
+forests, Magellanic subpolar forests, and Patagonian steppe. The Atacama
+desert is predominantly arid with hot (Bwh in the Koppen-Geiger
+classification) and cold (Bwk) temperatures, as well as the northern
+part of the Chilean Matorral. Most of the land in these two northern
+regions is bare, except for a small area where shrublands and grasslands
+are present. With an annual rainfall of less than 400 mm, the Central
+Andean dry puna ecoregion has low, yet highly seasonal precipitation
+with an eight-month dry season, low temperatures (Bwk) and is dominated
+by grasslands, shrublands, and savanna. The climate of the Southern
+Andean steppe ecoregion is cold desert (BWk), with most precipitation
+occurring in the winter. There is little vegetation in this ecoregion
+because the plants have adapted to its windy, dry, and cold climate. In
+central Chile, the climate of the Chilean Matorral changes to that of an
+arid steppe with cold temperatures (Bsk). Then, towards the center-south
+of the country, the climate of the Chilean Matorral changes to a
+Mediterranean climate, with warm to hot summers (Csa and Csb). Land
+cover in this ecoregion consists of a significant amount of shrublands
+and savannas. The Valdivian temperate forests have a mostly oceanic
+climate (Cfb) and a large area of forests and grasslands. The Magellanic
+subpolar forests have a tundra climate. Lastly, the Patagonian steppe
+has high aridity, cold temperatures (Bsk), and primarily consists of
+grasslands.
+
+\subsection{Data}\label{data}
+
+\subsubsection{Gridded meteorological and vegetation
+data}\label{gridded-meteorological-and-vegetation-data}
+
+To derive a proxy for vegetation productivity, we used the Normalized
+Difference Vegetation Index (NDVI) from the MOD13A3 Collection 6.1
+product derived from the MODIS (Moderate-Resolution Imaging
+Spectroradiometer) sensor onboard the Terra satellite. MOD13A3 provides
+vegetation indices with a 1 km spatial resolution and monthly frequency
+\citep{Didan2015}. We also utilized the MOD16A2 collection 6.1
+\citep{Running2017} product from MODIS to gauge the water consumption of
+vegetation. This product gives us monthly actual evapotranspiration (ET)
+and atmospheric evaporative demand (AED) with a \textasciitilde500m
+spatial resolution. For soil moisture, water supply, and water demand
+variables, we used ERA5-Land (ERA5L; ECMWF Reanalysis version 5 over
+land) \citep{MunozSabater2021}, a reanalysis dataset that provides
+atmospheric and land variables since 1950. It has a spatial resolution
+of 0.1° (9 km), hourly frequency, and global coverage. We selected total
+precipitation, maximum and minimum temperature at 2 meters, and
+volumetric soil water layers between 0 and 100 cm of depth (layer 1 to
+layer 3; Supplementary Materials and Methods, Supplementary Tables 2 and
+4).
+
+\subsubsection{Gridded indicators for land
+use}\label{gridded-indicators-for-land-use}
+
+To account for the impacts of human activity on land cover change, we
+obtained data on road density \citep{Meijer2018} and night lights for
+the period 2012--2023 \citep{Roman2018}. These products are frequently
+used in the literature to quantify the human footprint (e.g.,
+\citep{Halpern2015, Halpern2008}) or biodiversity threats (e.g.,
+\citep{Bowler2020, Kennedy2020}).To capture changes on land cover due to
+fires, we calculated the total burned area for 2002-2023
+\citep{Chen2023}.
+
+\subsection{Short- to long-term drought
+trends}\label{short--to-long-term-drought-trends}
+
+\subsubsection{Atmospheric Evaporative Demand
+(AED)}\label{atmospheric-evaporative-demand-aed}
+
+To compute drought indices that use water demand, it is necessary to
+first calculate AED. To do this, we employed the Hargreaves method
+\citep{Hargreaves1994, Hargreaves1985} by applying the following
+equation:
+
+\begin{equation}\phantomsection\label{eq-AED}{AED = 0.0023\cdot Ra\cdot (T+17.8)\cdot (T_{max}-T_{min})^{0.5}}\end{equation}
+
+where \(Ra\) \((MJ\,m^2\, day^{-1})\) is extraterrestrial radiation;
+\(T\), \(T_{max}\), and \(T_{min}\) are mean, maximum, and minimum
+temperature \((°C)\) at 2m. For calculating \(Ra\) we used the
+coordinate of the latitud of the centroid of each pixel as follow:
+
+\begin{equation}\phantomsection\label{eq-Ra}{R_a = \frac{14,400}{\pi}\cdot G_{sc}\cdot d_r \left[\omega_s\cdot sin(\phi)\cdot sin(\delta)+cos(\phi)\cdot cos(\delta)\cdot sin(\omega_s) \right]}\end{equation}
 
@@
 
-$Ra$: extraterrestrial radiation $[MJ\, m^{-2} day-1]$,  
-$G_{sc}$: solar constant = 0.0820 $[MJ\,m^{-2} min^{-1}]$,    
-$d_r$: inverse relative distance Earth-Sun,  
-$\omega_s$ sunset hour angle  $[rad]$,  
-$\phi$: latitude $[rad]$,  
-$\delta$: solar declination $[rad]$.
+\(Ra\): extraterrestrial radiation \([MJ\, m^{-2} day-1]\),\\
+\(G_{sc}\): solar constant = 0.0820 \([MJ\,m^{-2} min^{-1}]\),\\
+\(d_r\): inverse relative distance Earth-Sun,\\
+\(\omega_s\) sunset hour angle \([rad]\),\\
+\(\phi\): latitude \([rad]\),\\
+\(\delta\): solar declination \([rad]\).
 
-We selected the method of Hargreaves to estimate AED because of its simplicity, as it only requires temperature and extraterrestrial radiation, and because access to the data needed for alternative methods (e.g., Penman-Montieth) is often limited [@Vicente-Serrano2014].
+We selected the method of Hargreaves to estimate AED because of its
+simplicity, as it only requires temperature and extraterrestrial
+radiation, and because access to the data needed for alternative methods
+(e.g., Penman-Montieth) is often limited \citep{Vicente-Serrano2014}.
 
-### Drought indices
+\subsubsection{Drought indices}\label{drought-indices}
 
-To derive the drought indices of water supply and demand we used the ERA5L dataset and the MODIS product (MOD13A3 [@Didan2015b]), with a monthly frequency for 2000–2023.
-Drought indices capture historical anomalies of water supply and demand.
-To quantify each anomaly, the common practice is to derive it following a statistical parametric method in which it is assumed that the statistical distribution of the data is known [@Heim2002].
-The use of an erroneous statistical distribution that does not fit the data is usually the highest source of uncertainty [@Laimighofer2022].
-In the case of Chile, due to its high degree of climatic variability, it is difficult to choose a statistical distribution that can be used across its entire extent.
-We therefore use a non-parametric method for the calculation of the drought indices, following [@Farahmand2015].
+To derive the drought indices of water supply and demand we used the
+ERA5L dataset and the MODIS product (MOD13A3 \citep{Didan2015b}), with a
+monthly frequency for 2000--2023. Drought indices capture historical
+anomalies of water supply and demand. To quantify each anomaly, the
+common practice is to derive it following a statistical parametric
+method in which it is assumed that the statistical distribution of the
+data is known \citep{Heim2002}. The use of an erroneous statistical
+distribution that does not fit the data is usually the highest source of
+uncertainty \citep{Laimighofer2022}. In the case of Chile, due to its
+high degree of climatic variability, it is difficult to choose a
+statistical distribution that can be used across its entire extent. We
+therefore use a non-parametric method for the calculation of the drought
+indices, following \citep{Farahmand2015}.
 
-For monitoring water supply, we used the Standardized Precipitation Index (SPI;  [@McKee1993]), which only uses precipitation data.
-To evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI; [@Hobbins2016] and  [@McEvoy2016]), which is based on AED, and the Standardized Evapotranspiration Index (SETI; [@Yang2019]), which quantifies actual evapotranspiration, i.e.
-the amount of water removed from a surface due to evaporation and transpiration.
-To quantify the combined effect of water supply and demand, we estimated SPEI [@Vicente-Serrano2010].
-For SPEI, we calculated an auxiliary variable (D) with the following formula:
+For monitoring water supply, we used the Standardized Precipitation
+Index (SPI; \citep{McKee1993}), which only uses precipitation data. To
+evaluate water demand, we chose the Evaporative Demand Drought Index
+(EDDI; \citep{Hobbins2016} and \citep{McEvoy2016}), which is based on
+AED, and the Standardized Evapotranspiration Index (SETI;
+\citep{Yang2019}), which quantifies actual evapotranspiration, i.e.~the
+amount of water removed from a surface due to evaporation and
+transpiration. To quantify the combined effect of water supply and
+demand, we estimated SPEI \citep{Vicente-Serrano2010}. For SPEI, we
+calculated an auxiliary variable (D) with the following formula:
 
-$$D = P-AED$$
-where P is precipitation.
-Soil moisture is often considered to be the main driver of vegetation productivity, particularly in semi-arid regions [@Li2022].
-Hence, we used the Standardized Soil Moisture Index (SSI) to estimate soil moisture (SM) [@Hao2013].
-For SSI, we used the average soil moisture from ERA5L at a depth of 1m.
-All calculated indices are multi-scalar and can be used for the analysis of short- to long-term droughts.
+\[D = P-AED\] where P is precipitation. Soil moisture is often
+considered to be the main driver of vegetation productivity,
+particularly in semi-arid regions \citep{Li2022}. Hence, we used the
+Standardized Soil Moisture Index (SSI) to estimate soil moisture (SM)
+\citep{Hao2013}. For SSI, we used the average soil moisture from ERA5L
+at a depth of 1m. All calculated indices are multi-scalar and can be
+used for the analysis of short- to long-term droughts.
 
-To derive the drought indices, we first calculate the sum of the variables with regard to the time scale(s).
-In this case, for generalization purposes, we will use $V$, referring to variables P, AED, D,  and SM (Table  SSX).
-We accumulated each over the time series of values (months), and for the time scales s:
+To derive the drought indices, we first calculate the sum of the
+variables with regard to the time scale(s). In this case, for
+generalization purposes, we will use \(V\), referring to variables P,
+AED, D, and SM (Table SSX). We accumulated each over the time series of
+values (months), and for the time scales s:
 
-$$A^s_i = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  $$ {#eq-sumvar}
+\begin{equation}\phantomsection\label{eq-sumvar}{A^s_i = \sum_{i=n-s-i+2}^{n-i+1} V_i\,\, \forall\, i\geq n-s+1  }\end{equation}
 
-The $A^s_i$ corresponds to a moving window (convolution) that sums the variable for time scales $s$.
-This is summed over $s$ months, starting from the most recent month ($n$) back in time until month $n-s+1$.
-For example, using as a variable the precipitation, a period of twelve months ($n$), and a time scale of three months ($s$):
+The \(A^s_i\) corresponds to a moving window (convolution) that sums the
+variable for time scales \(s\). This is summed over \(s\) months,
+starting from the most recent month (\(n\)) back in time until month
+\(n-s+1\). For example, using as a variable the precipitation, a period
+of twelve months (\(n\)), and a time scale of three months (\(s\)):
 
-$$
+\[
 \begin{split}
@@
 \end{split}
-$$
+\]
 
-Then, we used the empirical Tukey plotting position [@Wilks2011] over $A_i^s$ to derive the $P(A_i^s)$ probabilities across a period of interest:
+Then, we used the empirical Tukey plotting position \citep{Wilks2011}
+over \(A_i^s\) to derive the \(P(A_i^s)\) probabilities across a period
+of interest:
 
-$$P(A^s_i) = \frac{i-0.33}{n+0.33'}$$ {#eq-probPai}
+\begin{equation}\phantomsection\label{eq-probPai}{P(A^s_i) = \frac{i-0.33}{n+0.33'}}\end{equation}
 
-An inverse normal approximation [@Abramowitz1968] obtains the empirically derived probabilities once the variable cumulates over time for the scale $s$.
-Thus, the drought indices $SPI$, $SPEI$, $EDDI$, and $SSI$ are  obtained following the equation:
+An inverse normal approximation \citep{Abramowitz1968} obtains the
+empirically derived probabilities once the variable cumulates over time
+for the scale \(s\). Thus, the drought indices \(SPI\), \(SPEI\),
+\(EDDI\), and \(SSI\) are obtained following the equation:
 
-$$DI(A^s_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}$$ {#eq-DI}
+\begin{equation}\phantomsection\label{eq-DI}{DI(A^s_i) = W - \frac{C_0+C_1\cdot W + c_2 \cdot W^2}{1+d_1\cdot W +d_2\cdot W^2 +d_3\cdot W^3}}\end{equation}
 
-$DI$ is referring to the drought index calculated for the variable $V$.
-The values for the constats are: $C_0 = 2.515517$, $C_1 = 0.802853$, $C_2 = 0.010328$, $d_1 = 1.432788$, $d_2 = 0.189269$, and $d3 = 0.001308$.
-For $P(A^s_i) \leq 0.5$, W=$\sqrt{-2\cdot ln(P(A^s_i))}$ , and for $P(A^s_i) > 0.5$, replace $P(A^s_i)$ with $1-P(A^s_i)$ and reverse the sign of $DI(A^s_i)$.
+\(DI\) is referring to the drought index calculated for the variable
+\(V\). The values for the constats are: \(C_0 = 2.515517\),
+\(C_1 = 0.802853\), \(C_2 = 0.010328\), \(d_1 = 1.432788\),
+\(d_2 = 0.189269\), and \(d3 = 0.001308\). For \(P(A^s_i) \leq 0.5\),
+W=\(\sqrt{-2\cdot ln(P(A^s_i))}\) , and for \(P(A^s_i) > 0.5\), replace
+\(P(A^s_i)\) with \(1-P(A^s_i)\) and reverse the sign of \(DI(A^s_i)\).
 
-The drought indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months at a monthly frequency for 2000–2023.
+The drought indices were calculated for time scales of 1, 3, 6, 12, 24,
+and 36 months at a monthly frequency for 2000--2023.
 
-### Temporal trends of drought indices
+\subsubsection{Temporal trends of drought
+indices}\label{temporal-trends-of-drought-indices}
 
-To determine if there are statistically significant positive or negative temporal trends for the drought indices, we used the non-parametric modified Mann-Kendall test for serially correlated data [@Yue2004].
-To determine the magnitude of the trend, we used Sen’s slope [@Sen1968].
-Sen's slope is less affected by outliers than parametric ordinary least squares (OLS) regression, and as a non-parametric method, it is not influenced by the distribution of the data.
-We applied both methods for SPI, EDDI, SPEI, SETI, and SSI and six time scales, resulting in a total of 30 trends.
-We then aggregated temporal trends for each ecoregion and land cover type.
+To determine if there are statistically significant positive or negative
+temporal trends for the drought indices, we used the non-parametric
+modified Mann-Kendall test for serially correlated data \citep{Yue2004}.
+To determine the magnitude of the trend, we used Sen's slope
+\citep{Sen1968}. Sen's slope is less affected by outliers than
+parametric ordinary least squares (OLS) regression, and as a
+non-parametric method, it is not influenced by the distribution of the
+data. We applied both methods for SPI, EDDI, SPEI, SETI, and SSI and six
+time scales, resulting in a total of 30 trends. We then aggregated
+temporal trends for each ecoregion and land cover type.
 
-### Vegetation productivity
+\subsubsection{Vegetation productivity}\label{vegetation-productivity}
 
-We also used the MODIS product to calculate vegetation productivity, and calculated anomalies of cumulative NDVI using  zcNDVI [@Zambrano2018], which was derived from the monthly time series of NDVI, with Equations 2 and 4.
-For vegetation productivity, we selected the time scale that best correlates with annual net primary productivity (NPP) across continental Chile.
-For this purpose, we calculated zcNDVI for time scales of 1, 3, 6, and 12 months (from December) and compared it with the annual NPP.
-We obtained NPP from MOD17A3HGF90.
-We chose to use six months because the $R^2$ between zcNDVI and NPP reaches its highest value at six months, obtaining an $R^2$ of 0.31 for forest and 0.72 for shrubland (Supplementary Information Section S2).
-We subsequently used zcNDVI with a time scale of 6 months and calculated it at a monthly frequency for 2000–2023.
+We also used the MODIS product to calculate vegetation productivity, and
+calculated anomalies of cumulative NDVI using zcNDVI
+\citep{Zambrano2018}, which was derived from the monthly time series of
+NDVI, with Equations 2 and 4. For vegetation productivity, we selected
+the time scale that best correlates with annual net primary productivity
+(NPP) across continental Chile. For this purpose, we calculated zcNDVI
+for time scales of 1, 3, 6, and 12 months (from December) and compared
+it with the annual NPP. We obtained NPP from MOD17A3HGF90. We chose to
+use six months because the \(R^2\) between zcNDVI and NPP reaches its
+highest value at six months, obtaining an \(R^2\) of 0.31 for forest and
+0.72 for shrubland (Supplementary Information Section S2). We
+subsequently used zcNDVI with a time scale of 6 months and calculated it
+at a monthly frequency for 2000--2023.
 
-## Drought impacts on vegetation productivity
+\subsection{Drought impacts on vegetation
+productivity}\label{drought-impacts-on-vegetation-productivity}
 
 For each land cover, we analyzed the trend of vegetation productivity.
-To this end, we identified areas within each land cover that are persistent over time to reduce the possibility that trends in vegetation productivity may be influenced by changes in  land cover.
-We examined the correlation between drought indices and vegetation productivity across land cover types to determine to the extent to which soil moisture and water demand and supply affect vegetation productivity.
+To this end, we identified areas within each land cover that are
+persistent over time to reduce the possibility that trends in vegetation
+productivity may be influenced by changes in land cover. We examined the
+correlation between drought indices and vegetation productivity across
+land cover types to determine to the extent to which soil moisture and
+water demand and supply affect vegetation productivity.
 
-We estimated pixel-to-pixel Pearson’s correlations between drought indices at time scales of 1, 3, 6, 12, 24, and 36 months with zcNDVI.
-We extracted the Pearson correlation coefficient corresponding to the time scale with the highest value.
-For each index, we then generated two maps: 1) a raster with values of the time scales and drought index that reached the maximum correlation, and 2) a raster with the magnitude of the correlation between the drought index and vegetation productivity.
+We estimated pixel-to-pixel Pearson's correlations between drought
+indices at time scales of 1, 3, 6, 12, 24, and 36 months with zcNDVI. We
+extracted the Pearson correlation coefficient corresponding to the time
+scale with the highest value. For each index, we then generated two
+maps: 1) a raster with values of the time scales and drought index that
+reached the maximum correlation, and 2) a raster with the magnitude of
+the correlation between the drought index and vegetation productivity.
 
-## Drought impacts on land cover change
+\subsection{Drought impacts on land cover
+change}\label{drought-impacts-on-land-cover-change}
 
-### Land cover change
+\subsubsection{Land cover change}\label{land-cover-change}
 
-To analyze land cover change, we used the classification scheme of the International Geosphere-Biosphere Programme (IGBP) from the product MCD12Q1 Collection 6.1 from MODIS.
-The MCD12Q1 product is produced for each year from 2001 to 2023 and defines 17 classes (see Table S1).
-Following the FAO classification [@FAO2022], we classified native and planted forests as "forests", which represent natural and productive ecosystems dominated by large trees.
-To analyze the land cover change, we use the IGBP scheme from the MCD12Q1 product.
-We regrouped the 17 classes into ten macro-classes, as follows: 1-4 to forests (native forest and plantations), 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 as water (Table S3).
-This resulted in a time series of land cover with ten macro-classes for 2001-2023.
-We validated the land cover macro-classes using a high resolution (30 m) land cover map for 2013-2014 [@Zhao2016].
-Our results showed a global accuracy of ~0.82 and a F1 score of ~0.66 (Supplementary Information, S2).
+To analyze land cover change, we used the classification scheme of the
+International Geosphere-Biosphere Programme (IGBP) from the product
+MCD12Q1 Collection 6.1 from MODIS. The MCD12Q1 product is produced for
+each year from 2001 to 2023 and defines 17 classes (see Table S1).
+Following the FAO classification \citep{FAO2022}, we classified native
+and planted forests as ``forests'', which represent natural and
+productive ecosystems dominated by large trees. To analyze the land
+cover change, we use the IGBP scheme from the MCD12Q1 product. We
+regrouped the 17 classes into ten macro-classes, as follows: 1-4 to
+forests (native forest and plantations), 5-7 to shrublands, 8-9 to
+savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13
+as urban, 15 as snow and ice, 16 as barren, and 17 as water (Table S3).
+This resulted in a time series of land cover with ten macro-classes for
+2001-2023. We validated the land cover macro-classes using a high
+resolution (30 m) land cover map for 2013-2014 \citep{Zhao2016}. Our
+results showed a global accuracy of \textasciitilde0.82 and a F1 score
+of \textasciitilde0.66 (Supplementary Information, S2).
 
-We calculated the area for each land cover class in the five study regions for 2001–2023.
-We then estimated the temporal change in area for each land cover type and macro-class, and determined the statistical significance (p-value < 0.05) and magnitude of the trend as described above.
+We calculated the area for each land cover class in the five study
+regions for 2001--2023. We then estimated the temporal change in area
+for each land cover type and macro-class, and determined the statistical
+significance (p-value \textless{} 0.05) and magnitude of the trend as
+described above.
 
-To assess how water demand and supply, and soil moisture affect the variation in vegetation productivity across various land cover types, we avoid analyzing areas that experienced major land cover changes in the 2001–2023 period.
-To assess how zcNDVI varied irrespective of land cover change, we developed a persistence mask for land cover, which only retains pixels for which the macro-class remained the same for at least 80% of the 23 years (Fig.
-1d).
+To assess how water demand and supply, and soil moisture affect the
+variation in vegetation productivity across various land cover types, we
+avoid analyzing areas that experienced major land cover changes in the
+2001--2023 period. To assess how zcNDVI varied irrespective of land
+cover change, we developed a persistence mask for land cover, which only
+retains pixels for which the macro-class remained the same for at least
+80\% of the 23 years (Fig. 1d).
 
-## Relationship between land cover and drought trends
+\subsection{Relationship between land cover and drought
+trends}\label{relationship-between-land-cover-and-drought-trends}
 
-To identify which drought indices and time scales have a major impact on changes in land cover type, we examined the relationships between the temporal trends in the surface of land cover classes, drought indices, road density, burned area, and night lights, and  for each ecoregion.
-We performed the analysis at the sub-basin scale, using 485 river basins, which have a surface area between 0.906 and 24,408 km2 and a median area of 1,249 km2 (Supplementary Fig.
-S8/Table S5).
-For each basin, we calculated the trend per land cover, considering the proportion of the type relative to the total surface of the basin.
-For each basin we extracted the average trend of all drought indices and at time scales of 1, 3, 6, 12, 24, and 36 months.
-In the case of burned area, we used as variables the total and the trend of burned area for 2002-2023, and for night lights we used the average and the trend of nightlights for 2012-2023.
+To identify which drought indices and time scales have a major impact on
+changes in land cover type, we examined the relationships between the
+temporal trends in the surface of land cover classes, drought indices,
+road density, burned area, and night lights, and for each ecoregion. We
+performed the analysis at the sub-basin scale, using 485 river basins,
+which have a surface area between 0.906 and 24,408 km2 and a median area
+of 1,249 km2 (Supplementary Fig. S8/Table S5). For each basin, we
+calculated the trend per land cover, considering the proportion of the
+type relative to the total surface of the basin. For each basin we
+extracted the average trend of all drought indices and at time scales of
+1, 3, 6, 12, 24, and 36 months. In the case of burned area, we used as
+variables the total and the trend of burned area for 2002-2023, and for
+night lights we used the average and the trend of nightlights for
+2012-2023.
 
-Prior to modelling relationships between trends in land cover and drought indices, we assessed multi-collinearity among explanatory variables, i.e., drought indices, road density, night lights, and burned area, with the variance inflation factor (VIF).
-We analyzed the VIF for all drought indices at time scales of 1, 3, 6, 12, 24, and 36 separately because each index has a strong correlation across time scales.
-As VIF values greater than five may affect the interpretation of model results [@Dorman2013], we therefore excluded SPI from all subsequent models.
+Prior to modelling relationships between trends in land cover and
+drought indices, we assessed multi-collinearity among explanatory
+variables, i.e., drought indices, road density, night lights, and burned
+area, with the variance inflation factor (VIF). We analyzed the VIF for
+all drought indices at time scales of 1, 3, 6, 12, 24, and 36 separately
+because each index has a strong correlation across time scales. As VIF
+values greater than five may affect the interpretation of model results
+\citep{Dorman2013}, we therefore excluded SPI from all subsequent
+models.
 
-To assess the relationship between land cover trends and drought indices, we modeled trends in the surface of land cover types.
-We made a regression analysis using the random forest method93, which employs multiple decision trees.
-Some advantages of random forest include the ability to find non-linear relationships, reduce overfitting, and derive variable importance.
-We incorporated the trends of the five drought indices (SPI, SPEI, EDDI, SETI, and SSI), the trends of night lights (2) and burned area (2), the road density, for a total of ten predictors.
-We then constructed random forest models for each time scale (1, 3, 6, 12, 24, and 36) and each land cover class (forest, grassland, shrubland, savanna, cropland, and barren land), resulting in a total of 36 RF models.
-We trained each model using 1000 trees, setting the minimum number of nodes per decision tree at five and the number of predictors per split (boosting) to the square root of the total number of predictors.
-To account for uncertainty, we trained all the models ten times using a resampling strategy (ten folds) in a cross-validation scheme.
-Finally, we evaluate model fit by calculating R², root mean square error (RMSE), and variable importance.
-Variable importance identifies which variables have a higher contribution to explaining model variation.
-We calculated variable importance by permuting out-of-bag (OOB) data per tree and calculating the mean standard error of the OOB data.
-After permuting each predictor variable, we repeated the process for the remaining variables.
-We repeated this process ten times per fold to assess  model fit.
+To assess the relationship between land cover trends and drought
+indices, we modeled trends in the surface of land cover types. We made a
+regression analysis using the random forest method93, which employs
+multiple decision trees. Some advantages of random forest include the
+ability to find non-linear relationships, reduce overfitting, and derive
+variable importance. We incorporated the trends of the five drought
+indices (SPI, SPEI, EDDI, SETI, and SSI), the trends of night lights (2)
+and burned area (2), the road density, for a total of ten predictors. We
+then constructed random forest models for each time scale (1, 3, 6, 12,
+24, and 36) and each land cover class (forest, grassland, shrubland,
+savanna, cropland, and barren land), resulting in a total of 36 RF
+models. We trained each model using 1000 trees, setting the minimum
+number of nodes per decision tree at five and the number of predictors
+per split (boosting) to the square root of the total number of
+predictors. To account for uncertainty, we trained all the models ten
+times using a resampling strategy (ten folds) in a cross-validation
+scheme. Finally, we evaluate model fit by calculating R², root mean
+square error (RMSE), and variable importance. Variable importance
+identifies which variables have a higher contribution to explaining
+model variation. We calculated variable importance by permuting
+out-of-bag (OOB) data per tree and calculating the mean standard error
+of the OOB data. After permuting each predictor variable, we repeated
+the process for the remaining variables. We repeated this process ten
+times per fold to assess model fit.
 
-Finally, we visually explored the relationship between drought indexes and changes in land cover across sub basins within Chile.
-To achieve this, we compared the relative changes in land cover surface with the drought indices, burned area, nighlights, and road density and other variables for the time scale that were deemed more significant in the random forest model.
+Finally, we visually explored the relationship between drought indexes
+and changes in land cover across sub basins within Chile. To achieve
+this, we compared the relative changes in land cover surface with the
+drought indices, burned area, nighlights, and road density and other
+variables for the time scale that were deemed more significant in the
+random forest model.
 
-## Software
+\subsection{Software}\label{software}
 
-For downloading, processing, and analysis of the spatio-temporal data, we used the open source software for statistical computing and graphics, R [@R2023].
-For downloading ERA5L, we used the {ecmwfr} package [@Hufkens2019].
-For processing raster data, we used {terra} [@Hijmans2023] and {stars} [@Pebesma2023].
-For managing vectorial data, we used {sf} [@Pebesma2018].
-For the calculation of AED, we used {SPEI} [@Bergueria2023].
-For mapping, we used {tmap} [@Tennekes2018].
-For data analysis and visualization, the suite {tidyverse} [@Wickham2019] was used.
-For the random forest modeling, we used the {tidymodels} [@Kuhn2020] and {ranger} [@Wright2017] packages.
+For downloading, processing, and analysis of the spatio-temporal data,
+we used the open source software for statistical computing and graphics,
+R \citep{R2023}. For downloading ERA5L, we used the \{ecmwfr\} package
+\citep{Hufkens2019}. For processing raster data, we used \{terra\}
+\citep{Hijmans2023} and \{stars\} \citep{Pebesma2023}. For managing
+vectorial data, we used \{sf\} \citep{Pebesma2018}. For the calculation
+of AED, we used \{SPEI\} \citep{Bergueria2023}. For mapping, we used
+\{tmap\} \citep{Tennekes2018}. For data analysis and visualization, the
+suite \{tidyverse\} \citep{Wickham2019} was used. For the random forest
+modeling, we used the \{tidymodels\} \citep{Kuhn2020} and \{ranger\}
+\citep{Wright2017} packages.
 
-# Results
+\section{Results}\label{results}
 
-## The Chilean matorral and Patagonian steppe increase atmospheric water demand but decrease vegetation evapotranspiration
+\subsection{The Chilean matorral and Patagonian steppe increase
+atmospheric water demand but decrease vegetation
+evapotranspiration}\label{the-chilean-matorral-and-patagonian-steppe-increase-atmospheric-water-demand-but-decrease-vegetation-evapotranspiration}
 
-For the Atacama desert and the Central Andean Puna, we found a positive temporal trend for drought indices of water supply (i.e., SPI, SSI), atmospheric water demand (i.e., EDDI), and vegetation water demand (i.e., SETI).
-For the Chilean matorral and Patagonian steppe, EDDI presents the higher positive trend and lower negative trends in SPI, SPEI, SSI, and SETI.
-This reflects a critical scenario of drought, where a rise in temperature increases atmospheric water demand, but vegetation cannot increase evapotranspiration due to a lack of water availability.
-In the Southern Andean steppe, there is a positive trend in AED (i.e., EDDI), but a negative in water supply (i.e., SPI, SPEI, SSI).
-The vegetation water demand (i.e., SETI) has negative trends, but it is increasing at higher time scales.
-The Valdivian temperate forests show a negative trend in water supply (i.e., SPI, SPEI, and SSI) and a positive trend in both AED and ET, as shown by the EDDI and SETI, respectively.
-Here, an increase in AED implies an increase in ET, likely due to a greater availability of water, unlike in the Chilean Matorral and Patagonian steppe.
-The vegetation water demand (SETI) in the Magellanic subpolar forests does not exhibit a significant trend over any given time scale.
-The AED and water supply present a positive trend.
-The trends of drought indices in the Patagonian steppe exhibit a similar behavior to the Chilean matorral, albeit to a lesser extent.
-Generally, the majority of the indices indicate that the trend (positive or negative) intensifies over longer time periods.
-(Fig.
-2)
+For the Atacama desert and the Central Andean Puna, we found a positive
+temporal trend for drought indices of water supply (i.e., SPI, SSI),
+atmospheric water demand (i.e., EDDI), and vegetation water demand
+(i.e., SETI). For the Chilean matorral and Patagonian steppe, EDDI
+presents the higher positive trend and lower negative trends in SPI,
+SPEI, SSI, and SETI. This reflects a critical scenario of drought, where
+a rise in temperature increases atmospheric water demand, but vegetation
+cannot increase evapotranspiration due to a lack of water availability.
+In the Southern Andean steppe, there is a positive trend in AED (i.e.,
+EDDI), but a negative in water supply (i.e., SPI, SPEI, SSI). The
+vegetation water demand (i.e., SETI) has negative trends, but it is
+increasing at higher time scales. The Valdivian temperate forests show a
+negative trend in water supply (i.e., SPI, SPEI, and SSI) and a positive
+trend in both AED and ET, as shown by the EDDI and SETI, respectively.
+Here, an increase in AED implies an increase in ET, likely due to a
+greater availability of water, unlike in the Chilean Matorral and
+Patagonian steppe. The vegetation water demand (SETI) in the Magellanic
+subpolar forests does not exhibit a significant trend over any given
+time scale. The AED and water supply present a positive trend. The
+trends of drought indices in the Patagonian steppe exhibit a similar
+behavior to the Chilean matorral, albeit to a lesser extent. Generally,
+the majority of the indices indicate that the trend (positive or
+negative) intensifies over longer time periods. (Fig. 2)
 
-## Vegetation productivity has strongly decreased in the Chilean matorral and the Patagonian steppe.
+\subsection{Vegetation productivity has strongly decreased in the
+Chilean matorral and the Patagonian
+steppe.}\label{vegetation-productivity-has-strongly-decreased-in-the-chilean-matorral-and-the-patagonian-steppe.}
 
-We found contrasting temporal trends in vegetation productivity for 2000-2023 (Fig.
-3).
-The Atacama desert does not exhibit significant trends over time.
-The Chilean Matorral, Patagonian steppe, and the Southern Andean steppe exhibit negative trends of -0.023, -0.016, and -0.006 (z-score/per decade), respectively.
-In contrast, the Central Andean dry puna, Valdivian temperate forests, and Central Andean dry puna show positive trends ranging from 0.01 to 0.03 (z-score/per decade).
-The Chilean matorral was at its lowest point from 2019 to 2022, while the Patagonian steppe has experienced an increasingly negative trend since 2022.
+We found contrasting temporal trends in vegetation productivity for
+2000-2023 (Fig. 3). The Atacama desert does not exhibit significant
+trends over time. The Chilean Matorral, Patagonian steppe, and the
+Southern Andean steppe exhibit negative trends of -0.023, -0.016, and
+-0.006 (z-score/per decade), respectively. In contrast, the Central
+Andean dry puna, Valdivian temperate forests, and Central Andean dry
+puna show positive trends ranging from 0.01 to 0.03 (z-score/per
+decade). The Chilean matorral was at its lowest point from 2019 to 2022,
+while the Patagonian steppe has experienced an increasingly negative
+trend since 2022.
 
-Forest, savanna, and shrubland exhibit the highest change in surface area across ecoregions
-We also observed significant changes in land cover surfaces across continental Chile (Fig.
-4).
-The forest surface has increased in the Chilean matorral and in the Valdivian temperate forest at rates of 78 and 316 km² yr⁻¹, respectively.
-Grassland surface has diminished in the Southern Andean steppe (-19 km² yr⁻¹) and has increased in the Patagonian steppe (90 km² yr⁻¹).
-Savanna exhibits a decrease in the Chilean matorral of -271 km² yr⁻¹ and the Valdivian temperate forest of -276 km² yr⁻¹, but an increase at a rate of 133 km² yr⁻¹ in the Magellanic subpolar forest.
-Shrubland has the highest increase in the Chilean matorral (160 km² yr⁻¹).
-Barren land has increased in the Central Andean dry puna (36 km² yr⁻¹) and the Southern Andean steppe (50 km² yr⁻¹), but has diminished in the Magellanic subpolar forest (-81 km² yr⁻¹).
-## Drought impact on vegetation productivity are strongest in the Chilean matorral and Valdivian temperate forest
+Forest, savanna, and shrubland exhibit the highest change in surface
+area across ecoregions We also observed significant changes in land
+cover surfaces across continental Chile (Fig. 4). The forest surface has
+increased in the Chilean matorral and in the Valdivian temperate forest
+at rates of 78 and 316 km² yr⁻¹, respectively. Grassland surface has
+diminished in the Southern Andean steppe (-19 km² yr⁻¹) and has
+increased in the Patagonian steppe (90 km² yr⁻¹). Savanna exhibits a
+decrease in the Chilean matorral of -271 km² yr⁻¹ and the Valdivian
+temperate forest of -276 km² yr⁻¹, but an increase at a rate of 133 km²
+yr⁻¹ in the Magellanic subpolar forest. Shrubland has the highest
+increase in the Chilean matorral (160 km² yr⁻¹). Barren land has
+increased in the Central Andean dry puna (36 km² yr⁻¹) and the Southern
+Andean steppe (50 km² yr⁻¹), but has diminished in the Magellanic
+subpolar forest (-81 km² yr⁻¹). \#\# Drought impact on vegetation
+productivity are strongest in the Chilean matorral and Valdivian
+temperate forest
 
-Our results indicate that drought impacts on vegetation productivity are highest in the Chilean Matorral and Valdivian temperate forests across all land cover types, except forest (Fig.
-5, Fig.
-S5 and Table 1).
-For time scales of 6 and 12 months, SETI and SSI have the strongest positive correlation with vegetation productivity among the land cover types.
-Next, we found that grassland and savanna in the Patagonian steppe had higher correlations with SPI and SSI over 12 months.
-Further, there is a positive relationship between the vegetation in the Atacama desert and drought indices of 12 months of water supply and vegetation water demand.
-However, there is a negative relationship between the vegetation and atmospheric water demand over 12 months.
-All drought indices show a positive correlation with the vegetation in the Central Andean dry puna, particularly the drought indices of water supply (SPI, SPEI, and SSI) at time scales of 24 and vegetation water demand (SETI) at time scales of 36 months.
-For the Southern Andean steppe the SETI of 24 months showed the highest correlation with savannas, followed by the EDDI of 24 months.
-Our analysis also revealed that water demand and supply differentially affected the time scales at which vegetation productivity of land cover types within each region was most impacted by drought (Fig.
-5, Fig.
-S5 and Table 1).
-While the spatial variation in the relationship between drought intensity and vegetation productivity was consistent across drought indices, the drought indices that captures water supply via soil moisture (Standardized Soil Moisture Index; SSI), and via vegetation water demand (Standardized Evapotranspiration Index, SETI)  tended to show a stronger correlation with vegetation productivity over larger areas than the other drought indices (Fig.
-5).
+Our results indicate that drought impacts on vegetation productivity are
+highest in the Chilean Matorral and Valdivian temperate forests across
+all land cover types, except forest (Fig. 5, Fig. S5 and Table 1). For
+time scales of 6 and 12 months, SETI and SSI have the strongest positive
+correlation with vegetation productivity among the land cover types.
+Next, we found that grassland and savanna in the Patagonian steppe had
+higher correlations with SPI and SSI over 12 months. Further, there is a
+positive relationship between the vegetation in the Atacama desert and
+drought indices of 12 months of water supply and vegetation water
+demand. However, there is a negative relationship between the vegetation
+and atmospheric water demand over 12 months. All drought indices show a
+positive correlation with the vegetation in the Central Andean dry puna,
+particularly the drought indices of water supply (SPI, SPEI, and SSI) at
+time scales of 24 and vegetation water demand (SETI) at time scales of
+36 months. For the Southern Andean steppe the SETI of 24 months showed
+the highest correlation with savannas, followed by the EDDI of 24
+months. Our analysis also revealed that water demand and supply
+differentially affected the time scales at which vegetation productivity
+of land cover types within each region was most impacted by drought
+(Fig. 5, Fig. S5 and Table 1). While the spatial variation in the
+relationship between drought intensity and vegetation productivity was
+consistent across drought indices, the drought indices that captures
+water supply via soil moisture (Standardized Soil Moisture Index; SSI),
+and via vegetation water demand (Standardized Evapotranspiration Index,
+SETI) tended to show a stronger correlation with vegetation productivity
+over larger areas than the other drought indices (Fig. 5).
 
-## Drought strongly impacts land cover distribution for shrubland
+\subsection{Drought strongly impacts land cover distribution for
+shrubland}\label{drought-strongly-impacts-land-cover-distribution-for-shrubland}
 
-Our random forest models show that drought indices explain between 32-79% of the variation in land cover change across continental Chile (Fig.
-6).
-Moreover, these results highlight the importance of considering water supply and demand, as drought indices associated with both aspects of the water balance had high importance values across most ecoregions and land cover types.
-The variation in the time scale of drought indices that provide the strongest correlation with vegetation productivity also suggests that different types of vegetation are not equally sensitive to droughts of similar intensities (Table 1).
-The RF models show that the drought indices are explaining 71-78% of the variability in land cover surface change for shrublands.
-Second, the RF models explain approximately 58-78% of the variability in change on croplands.
-In the case of other land cover types, the RF models account for approximately 33-59% of the variability, with drought indices explaining less for the forest type (Fig.
-6).
+Our random forest models show that drought indices explain between
+32-79\% of the variation in land cover change across continental Chile
+(Fig. 6). Moreover, these results highlight the importance of
+considering water supply and demand, as drought indices associated with
+both aspects of the water balance had high importance values across most
+ecoregions and land cover types. The variation in the time scale of
+drought indices that provide the strongest correlation with vegetation
+productivity also suggests that different types of vegetation are not
+equally sensitive to droughts of similar intensities (Table 1). The RF
+models show that the drought indices are explaining 71-78\% of the
+variability in land cover surface change for shrublands. Second, the RF
+models explain approximately 58-78\% of the variability in change on
+croplands. In the case of other land cover types, the RF models account
+for approximately 33-59\% of the variability, with drought indices
+explaining less for the forest type (Fig. 6).
 
-Our model has the highest r-squared for shrublands, followed by croplands, and barren land (Fig.
-6).
-Ecoregions most frequently observed the variables SETI and SSI, which significantly influenced surface changes for each land cover (Fig.
-7).
-In cropland, the primary factor is not the drought indices, but rather the total surface or trend of the burned area.
-The nightlight variable, which reflects urban areas, primarily explains the change in the surface of barren land, followed by the drought indices SPEI at time scales of 3 and 6 months (Fig.
-7).
+Our model has the highest r-squared for shrublands, followed by
+croplands, and barren land (Fig. 6). Ecoregions most frequently observed
+the variables SETI and SSI, which significantly influenced surface
+changes for each land cover (Fig. 7). In cropland, the primary factor is
+not the drought indices, but rather the total surface or trend of the
+burned area. The nightlight variable, which reflects urban areas,
+primarily explains the change in the surface of barren land, followed by
+the drought indices SPEI at time scales of 3 and 6 months (Fig. 7).
 
-Our results also show that drought intensity was associated with the magnitude and direction of land cover change (Fig.
-8).
-We observe that shrublands are sensitive to both increases and decreases in SETI and SSI, reaching a point of equilibrium around a normal climatic situation (drought index = 0).
-This could potentially stem from the fact that favorable water supply conditions alter the type and quantity of vegetation.
-Conversely, a reduction in water supply results in the shrublands losing their vegetation and transforming into bare soil.
-Thereby, both situations alter the land cover type.
-Changes in the burned surface impact the cropland surface; an increase in the burned area leads to an increase in the agricultural surface.
-Agricultural areas likely replace the burned area, explaining this phenomenon.
-In the case of bare soil, values that indicate low urban development (nightlights) are associated with a larger surface of bare soil.
-However, when the variable's range extends beyond the normal urban situation (zero value), an increase in bare soil correlates with a fluctuating increase in urban development.
-The relationship between SETI and SPEI in grasslands is opposite; an increase in SPEI leads to an increase in grassland area, while an increase in water demand by grasslands, as reflected by SETI, results in a decrease in grassland area.
+Our results also show that drought intensity was associated with the
+magnitude and direction of land cover change (Fig. 8). We observe that
+shrublands are sensitive to both increases and decreases in SETI and
+SSI, reaching a point of equilibrium around a normal climatic situation
+(drought index = 0). This could potentially stem from the fact that
+favorable water supply conditions alter the type and quantity of
+vegetation. Conversely, a reduction in water supply results in the
+shrublands losing their vegetation and transforming into bare soil.
+Thereby, both situations alter the land cover type. Changes in the
+burned surface impact the cropland surface; an increase in the burned
+area leads to an increase in the agricultural surface. Agricultural
+areas likely replace the burned area, explaining this phenomenon. In the
+case of bare soil, values that indicate low urban development
+(nightlights) are associated with a larger surface of bare soil.
+However, when the variable's range extends beyond the normal urban
+situation (zero value), an increase in bare soil correlates with a
+fluctuating increase in urban development. The relationship between SETI
+and SPEI in grasslands is opposite; an increase in SPEI leads to an
+increase in grassland area, while an increase in water demand by
+grasslands, as reflected by SETI, results in a decrease in grassland
+area.
 
-# Discussion
+\section{Discussion}\label{discussion}
 
-## Temporal trends in water supply and demand
+\subsection{Temporal trends in water supply and
+demand}\label{temporal-trends-in-water-supply-and-demand}
 
-We discovered that the Atacama desert, Central Andean dry puna, and the Magellanic subpolar forests experience an increase in water supply (SPI, SSI), as well as an increase in atmospheric and vegetation water demand (EDDI, SETI).
-However, in the Magellanic subpolar forests, there is no significant increase or decrease in SETI across time scales.
-Also, we found a significant decreasing trend in water supply (SPI, SPEI, and SSI) across the Southern Andean steppe, Chilean Matorral [@Boisier2018;@Sarricolea2019], Valdivian temperate forests, and Patagonian steppe, accompanied by an increase in atmospheric water demand (EDDI).
-Our results indicate that the trend of water supply and atmospheric demand tends to decrease or increase more strongly over longer time scales—a trend that is consistent with the progressive intensification of drought severity across much of Chile, as observed in other regions facing long-term droughts [@Rashid2019;@Miro2023].
-Simultaneously, we observed a divergent trend between the EDDI and SETI in various ecoregions.
-In the majority of ecoregions, a rise in atmospheric water demand (EDDI) typically leads to a rise in vegetation water demand (SETI).
-However, in the zones most affected by drought (Fig.
-3 and Fig.
-5), where water availability is limited, an increase in atmospheric water demand results in a decrease in vegetation water demand.
-The Chilean matorral and the Patagonian steppe are where this is most evident.
-When combined, our findings demonstrate a persistent drying trend in the Chilean matorral, the Patagonian steppe, and the Southern Andean steppe.
-We attribute this trend to a simultaneous decrease in precipitation and an increase in atmospheric evaporative demand, which leads to a decrease in the water demand for vegetation in water-limited zones[@Pascoa2021].
+We discovered that the Atacama desert, Central Andean dry puna, and the
+Magellanic subpolar forests experience an increase in water supply (SPI,
+SSI), as well as an increase in atmospheric and vegetation water demand
+(EDDI, SETI). However, in the Magellanic subpolar forests, there is no
+significant increase or decrease in SETI across time scales. Also, we
+found a significant decreasing trend in water supply (SPI, SPEI, and
+SSI) across the Southern Andean steppe, Chilean Matorral
+\citep{Boisier2018, Sarricolea2019}, Valdivian temperate forests, and
+Patagonian steppe, accompanied by an increase in atmospheric water
+demand (EDDI). Our results indicate that the trend of water supply and
+atmospheric demand tends to decrease or increase more strongly over
+longer time scales---a trend that is consistent with the progressive
+intensification of drought severity across much of Chile, as observed in
+other regions facing long-term droughts \citep{Rashid2019, Miro2023}.
+Simultaneously, we observed a divergent trend between the EDDI and SETI
+in various ecoregions. In the majority of ecoregions, a rise in
+atmospheric water demand (EDDI) typically leads to a rise in vegetation
+water demand (SETI). However, in the zones most affected by drought
+(Fig. 3 and Fig. 5), where water availability is limited, an increase in
+atmospheric water demand results in a decrease in vegetation water
+demand. The Chilean matorral and the Patagonian steppe are where this is
+most evident. When combined, our findings demonstrate a persistent
+drying trend in the Chilean matorral, the Patagonian steppe, and the
+Southern Andean steppe. We attribute this trend to a simultaneous
+decrease in precipitation and an increase in atmospheric evaporative
+demand, which leads to a decrease in the water demand for vegetation in
+water-limited zones\citep{Pascoa2021}.
 
-## Temporal trends in vegetation productivity
+\subsection{Temporal trends in vegetation
+productivity}\label{temporal-trends-in-vegetation-productivity}
 
-The consequences of the persistent drying trend for ecosystems throughout continental Chile are manifold.
-First, the prolonged hydrological drought, i.e., precipitation deficit, has reduced groundwater storage (SSI, [@Taucare2024]), leading to a steady decline in vegetation productivity (zcNDVI) since 2000 across the Chilean Matorral, Patagonian steppe, and Southern Andean steppe, reaching its lowest level between 2020 and 2022, which could be due to either a decrease in vegetation area, a loss of biomass or browning in forest ecosystems.
-Reports for natural and productive ecosystems [@Nicolai-Shaw2017;@Jiang2020;@Zhou2021] most strongly associate this decline with declines in soil moisture and increases in evapotranspiration.
-Second, the sharp decline in vegetation productivity in the Chilean Matorral and Valdivian temperate forests showed that grasses and shrublands respond to shifts in water supply on higher time scales (12 months), in comparison to savanna and croplands (6 months).
-Also, in the Valdivian temperate forest that has a higher surface with forest trees, soil moisture (SSI) and vegetation water demand (SETI) at higher time scales, 12 and 36 months, respectively, evidence stronger correlations.
-Which is consistent with recent studies showing that progressive, long-term water deficits in central Chile have triggered forest browning and declines in native forest productivity  [e.g., @Miranda2023;@Miranda2020;@Venegas2022].
-While our analysis do not distinguish between  native and planted forests, the latter of which are considered to be more drought tolerant in central and southern Chile [@Carrasco2022], we show that forest area declines more sharply in response to increasing water demand due to rising temperatures (EDDI) than decreasing water supply (e.g., SPI, SSI; refs.
-[@Fajardo2019;@Holz2018]), which may have cascading impacts on multiple facets of forest diversity [@Segovia2021;@Sabatini2022]
+The consequences of the persistent drying trend for ecosystems
+throughout continental Chile are manifold. First, the prolonged
+hydrological drought, i.e., precipitation deficit, has reduced
+groundwater storage (SSI, \citep{Taucare2024}), leading to a steady
+decline in vegetation productivity (zcNDVI) since 2000 across the
+Chilean Matorral, Patagonian steppe, and Southern Andean steppe,
+reaching its lowest level between 2020 and 2022, which could be due to
+either a decrease in vegetation area, a loss of biomass or browning in
+forest ecosystems. Reports for natural and productive ecosystems
+\citep{Nicolai-Shaw2017, Jiang2020, Zhou2021} most strongly associate
+this decline with declines in soil moisture and increases in
+evapotranspiration. Second, the sharp decline in vegetation productivity
+in the Chilean Matorral and Valdivian temperate forests showed that
+grasses and shrublands respond to shifts in water supply on higher time
+scales (12 months), in comparison to savanna and croplands (6 months).
+Also, in the Valdivian temperate forest that has a higher surface with
+forest trees, soil moisture (SSI) and vegetation water demand (SETI) at
+higher time scales, 12 and 36 months, respectively, evidence stronger
+correlations. Which is consistent with recent studies showing that
+progressive, long-term water deficits in central Chile have triggered
+forest browning and declines in native forest productivity
+\citep[e.g.,][]{Miranda2023, Miranda2020, Venegas2022}. While our
+analysis do not distinguish between native and planted forests, the
+latter of which are considered to be more drought tolerant in central
+and southern Chile \citep{Carrasco2022}, we show that forest area
+declines more sharply in response to increasing water demand due to
+rising temperatures (EDDI) than decreasing water supply (e.g., SPI, SSI;
+refs. \citep{Fajardo2019, Holz2018}), which may have cascading impacts
+on multiple facets of forest diversity \citep{Segovia2021, Sabatini2022}
 
-Moreover, the strengthening of the correlation between vegetation productivity and water supply (SPI, SPEI, SSI) or demand (EDDI, SETI) over multiple time scales (up to 36 months) and across land cover types (Fig.
-5) - demonstrates the impacts of climate change on the water balance in Chile.
-Impacts may extend beyond vegetation productivity, as reduced soil moisture in central Chile and the western United States has increased wildfire activity[@Holden2018;@Gonzalez2018], which is a growing concern in Chile and may be further exacerbated by extensive plantations of highly flammable tree species, e.g., Eucalyptus spp.
-and Pinus spp.
-[@Bowman2019].
-Third, we found that the decline in the vegetation productivity of croplands is due to a decrease in the water supply and vegetation water demand to a greater extent than to an increase in atmospheric water demand [@Quiring2010], causing a decline in water availability.
-This is consistent with evidence that more water-intensive crops have replaced less water-intensive crops in central Chile, leading to an increase in water extraction from rivers or groundwater [@Munoz2020;@Duran2020].
+Moreover, the strengthening of the correlation between vegetation
+productivity and water supply (SPI, SPEI, SSI) or demand (EDDI, SETI)
+over multiple time scales (up to 36 months) and across land cover types
+(Fig. 5) - demonstrates the impacts of climate change on the water
+balance in Chile. Impacts may extend beyond vegetation productivity, as
+reduced soil moisture in central Chile and the western United States has
+increased wildfire activity\citep{Holden2018, Gonzalez2018}, which is a
+growing concern in Chile and may be further exacerbated by extensive
+plantations of highly flammable tree species, e.g., Eucalyptus spp. and
+Pinus spp. \citep{Bowman2019}. Third, we found that the decline in the
+vegetation productivity of croplands is due to a decrease in the water
+supply and vegetation water demand to a greater extent than to an
+increase in atmospheric water demand \citep{Quiring2010}, causing a
+decline in water availability. This is consistent with evidence that
+more water-intensive crops have replaced less water-intensive crops in
+central Chile, leading to an increase in water extraction from rivers or
+groundwater \citep{Munoz2020, Duran2020}.
 
-## Drought impacts  on land cover
+\subsection{Drought impacts on land
+cover}\label{drought-impacts-on-land-cover}
 
-We found evidence that temporal decreases in water supply (SPEI, SSI) and decreases in vegetation water demand (SETI) are driving shifts not only in vegetation productivity but also in land cover across most of continental Chile.
-Shrubland covers were the land cover type most sensitive to changes in the water balance over short and long temporal scales.
-In contrast, forest surfaces are the type less affected by drought.
-Moreover, our results provide evidence that, in addition to shrublands cover, other land cover types have been affected by water deficits, particularly savannas and grasslands.
-Our results therefore suggest that multiple land cover types could be vulnerable to regime shifts towards more drought tolerant land cover types [@Scheffer2001;@Martinez-Vilalta2016], such as forest, whose cover increased non-linearly in response to increasing drought intensity.
+We found evidence that temporal decreases in water supply (SPEI, SSI)
+and decreases in vegetation water demand (SETI) are driving shifts not
+only in vegetation productivity but also in land cover across most of
+continental Chile. Shrubland covers were the land cover type most
+sensitive to changes in the water balance over short and long temporal
+scales. In contrast, forest surfaces are the type less affected by
+drought. Moreover, our results provide evidence that, in addition to
+shrublands cover, other land cover types have been affected by water
+deficits, particularly savannas and grasslands. Our results therefore
+suggest that multiple land cover types could be vulnerable to regime
+shifts towards more drought tolerant land cover types
+\citep{Scheffer2001, Martinez-Vilalta2016}, such as forest, whose cover
+increased non-linearly in response to increasing drought intensity.
 Changes in cropland cover are not a direct consequence of drought (Fig.
-7), but rather an indirect one, likely occupying burned areas and  possibly reflecting the decision of resource-poor farmers to migrate to regions with more abundant water resources or to change economic activity [@AghaKouchak2021;@Hermans2021].
-In contrast, the increase in shrubland cover due to a decrease in savanna cover may be ecological, as shrubs may  be more drought tolerant than other growth forms[@Eldridge2011;@Gotmark2016].
+7), but rather an indirect one, likely occupying burned areas and
+possibly reflecting the decision of resource-poor farmers to migrate to
+regions with more abundant water resources or to change economic
+activity \citep{AghaKouchak2021, Hermans2021}. In contrast, the increase
+in shrubland cover due to a decrease in savanna cover may be ecological,
+as shrubs may be more drought tolerant than other growth
+forms\citep{Eldridge2011, Gotmark2016}.
 
-We run each model ten times to account for the uncertainty in its performance.
-In the supplementary material (Fig.
-S12-S17), we show the error bar of r-squared; despite the inherent uncertainty, the analysis reinforces our results regarding how drought impacts land cover type.
+We run each model ten times to account for the uncertainty in its
+performance. In the supplementary material (Fig. S12-S17), we show the
+error bar of r-squared; despite the inherent uncertainty, the analysis
+reinforces our results regarding how drought impacts land cover type.
 
-Overall, our results show that long-term declines in water supply and demand have induced widespread, multi-dimensional impacts on the vegetation productivity and on the extent of land cover types.
-While prolonged droughts may directly cause shifts to more drought-tolerant land cover types, such as shrublands, they may also influence land cover change through human decision making and activities.
-This study extends current understanding of drought impacts by demonstrating how their multidimensionality emerges over multiple time scales and across land cover types, which can contribute to developing context-specific adaptation strategies for agriculture, biodiversity conservation, and natural resource management.
+Overall, our results show that long-term declines in water supply and
+demand have induced widespread, multi-dimensional impacts on the
+vegetation productivity and on the extent of land cover types. While
+prolonged droughts may directly cause shifts to more drought-tolerant
+land cover types, such as shrublands, they may also influence land cover
+change through human decision making and activities. This study extends
+current understanding of drought impacts by demonstrating how their
+multidimensionality emerges over multiple time scales and across land
+cover types, which can contribute to developing context-specific
+adaptation strategies for agriculture, biodiversity conservation, and
+natural resource management.
 
-## Study limitations
+\subsection{Study limitations}\label{study-limitations}
 
-This study has evaluated the impact of water supply and demand on vegetation productivity and how land use change influences it, with some important limitations to highlight.
-One of the major limitations of the study is the consideration of secondary information.
-For instance, we used estimates of water supply and demand, such as ERA5L and MODIS, which, despite their improved estimation capacity, suffer from biases and uncertainties [@Gomis2023;@Clelland2024] in different areas or climatic conditions.
-In this study, we compared the ERA5L data with climatic stations (see Table S2) to verify bias and uncertainty, but future studies will need to correct the products for a more precise measurement.
-We used zcNDVI [@Zambrano2018] (MODIS) as a proxy for vegetation productivity, which has proven to be a good estimate of NPP (see Fig.
-S1 and S2), but its quality varies between different types of vegetation.
+This study has evaluated the impact of water supply and demand on
+vegetation productivity and how land use change influences it, with some
+important limitations to highlight. One of the major limitations of the
+study is the consideration of secondary information. For instance, we
+used estimates of water supply and demand, such as ERA5L and MODIS,
+which, despite their improved estimation capacity, suffer from biases
+and uncertainties \citep{Gomis2023, Clelland2024} in different areas or
+climatic conditions. In this study, we compared the ERA5L data with
+climatic stations (see Table S2) to verify bias and uncertainty, but
+future studies will need to correct the products for a more precise
+measurement. We used zcNDVI \citep{Zambrano2018} (MODIS) as a proxy for
+vegetation productivity, which has proven to be a good estimate of NPP
+(see Fig. S1 and S2), but its quality varies between different types of
+vegetation.
 
-To define land cover, products were used that estimate the classes using classification models, which are subject to quality errors that must be taken into account [@Stehman2019;@Verburg2011].
-In addition, in our case we used macro classes of land cover, where, for example, the different types of forests were grouped (e.g., monoculture, native forest).
-This approach hinders our ability to understand the effects of drought on the various subclasses within each land use class.
-When it comes to cultivated areas, we cannot distinguish between rainfed and irrigated areas using macro classes.
-However, in this study, we aimed to provide a broad overview at a national scale, with the goal of focusing on a more detailed level in future research.
+To define land cover, products were used that estimate the classes using
+classification models, which are subject to quality errors that must be
+taken into account \citep{Stehman2019, Verburg2011}. In addition, in our
+case we used macro classes of land cover, where, for example, the
+different types of forests were grouped (e.g., monoculture, native
+forest). This approach hinders our ability to understand the effects of
+drought on the various subclasses within each land use class. When it
+comes to cultivated areas, we cannot distinguish between rainfed and
+irrigated areas using macro classes. However, in this study, we aimed to
+provide a broad overview at a national scale, with the goal of focusing
+on a more detailed level in future research.
 
-Other studies have analyzed how human decisions impact land use change, such as population increase [@Kleemann2017].
-In our case, we attempted to consider this by incorporating road density, night lights, and burned areas.
-However, human decisions are complex to consider from the point of view of earth observation.
-On the other hand, earth observation tools can analyze land cover change, but each type's use depends on social and economic factors, which are challenging to quantify [@Rindfuss2004;@Jenner2024] and necessitate the integration of social, natural, and geographic information sciences.
+Other studies have analyzed how human decisions impact land use change,
+such as population increase \citep{Kleemann2017}. In our case, we
+attempted to consider this by incorporating road density, night lights,
+and burned areas. However, human decisions are complex to consider from
+the point of view of earth observation. On the other hand, earth
+observation tools can analyze land cover change, but each type's use
+depends on social and economic factors, which are challenging to
+quantify \citep{Rindfuss2004, Jenner2024} and necessitate the
+integration of social, natural, and geographic information sciences.
 
-# Data availability
+\section{Data availability}\label{data-availability}
 
-The codes generated during the current study are available in the GitHub repository, https://github.com/FSEQ210022/drought_vegetation.
-The datasets generated and/or analyzed during the current study are available in the Zenodo repository, https://doi.org/10.5281/zenodo.10359547.
+The codes generated during the current study are available in the GitHub
+repository, https://github.com/FSEQ210022/drought\_vegetation. The
+datasets generated and/or analyzed during the current study are
+available in the Zenodo repository,
+https://doi.org/10.5281/zenodo.10359547.
 
-# Acknowledgments {-}
+\section*{Acknowledgments}\label{acknowledgments}
+\addcontentsline{toc}{section}{Acknowledgments}
 
-The National Research and Development Agency of Chile (ANID) funded this study through the drought emergency project FSEQ210022, Fondecyt Iniciación N°11190360, Fondecyt Postdoctorado N°3230678, and Fondecyt Regular N°1210526.
+The National Research and Development Agency of Chile (ANID) funded this
+study through the drought emergency project FSEQ210022, Fondecyt
+Iniciación N°11190360, Fondecyt Postdoctorado N°3230678, and Fondecyt
+Regular N°1210526.
+
+  \bibliography{referencias.bib}
+
+\end{document}

### 2025-04-08 article/article_scientific_reports/mansucript_11012025.qmd
actualización paper

@@
 In turn, ecological droughts, which capture the joint effects of precipitation and temperature in modifying natural and productive ecosystems [@Camps-Valls2021;@Paruelo2016;@Helman2014], are complex to measure and can therefore be monitored using multiple drought indices that capture the multiple dimensions of drought, e.g., precipitation, temperature, evapotranspiration, and AED.
-Although such an approach accounts for the joint effects of changes in natural and productive ecosystems, its potential impacts on land cover change have been largely unexplored.
+Although such an approach accounts for the joint effects of changes in natural and productive ecosystems, its potential impacts on land cover change have been largely unexplored [@Tollerud2018;@Marumbwa2020].
 
@@
 
-To derive the drought indices of water supply and demand we used the ERA5L dataset and the MODIS product (MOD13A3 [@Didan2015b]), with a monthly frequency for 2000–2023.
+To derive the drought indices of water supply and demand we used the ERA5L dataset, with a monthly frequency for 2000–2023.
 Drought indices capture historical anomalies of water supply and demand.
@@
 
-We also used the MODIS product to calculate vegetation productivity, and calculated anomalies of cumulative NDVI using  zcNDVI [@Zambrano2018], which was derived from the monthly time series of NDVI, with Equations 2 and 4.
+We also used the MODIS product (MOD13A3[@Didan2015]) to calculate vegetation productivity, and calculated anomalies of cumulative NDVI using  zcNDVI [@Zambrano2018], which was derived from the monthly time series of NDVI, with Equations 2 and 4.
 For vegetation productivity, we selected the time scale that best correlates with annual net primary productivity (NPP) across continental Chile.
@@
 We discovered that the Atacama desert, Central Andean dry puna, and the Magellanic subpolar forests experience an increase in water supply (SPI, SSI), as well as an increase in atmospheric and vegetation water demand (EDDI, SETI).
-However, in the Magellanic subpolar forests, there is no significant increase or decrease in SETI across time scales.
-Also, we found a significant decreasing trend in water supply (SPI, SPEI, and SSI) across the Southern Andean steppe, Chilean Matorral [@Boisier2018;@Sarricolea2019], Valdivian temperate forests, and Patagonian steppe, accompanied by an increase in atmospheric water demand (EDDI).
-Our results indicate that the trend of water supply and atmospheric demand tends to decrease or increase more strongly over longer time scales—a trend that is consistent with the progressive intensification of drought severity across much of Chile, as observed in other regions facing long-term droughts [@Rashid2019;@Miro2023].
-Simultaneously, we observed a divergent trend between the EDDI and SETI in various ecoregions.
+However, in the Magellanic subpolar forests, we found no evidence  of either a significant increase or decrease in SETI across time scales.
+Also, we found a significant decreasing trend in water supply (SPI, SPEI, and SSI) across the Southern Andean steppe, Chilean Matorral[@Boisier2018;@Sarricolea2019], Valdivian temperate forests, and Patagonian steppe, accompanied by an increase in atmospheric water demand (EDDI).
+Our results indicate that temporal trends of water supply and atmospheric demand tend to decrease or increase more strongly over longer time scales,a trend that is consistent with the progressive intensification of drought severity across much of Chile, and that has been observed in other regions facing long-term droughts[@Rashid2019;@Miro2023].
+Simultaneously, we observed a divergent trend between EDDI and SETI.
 In the majority of ecoregions, a rise in atmospheric water demand (EDDI) typically leads to a rise in vegetation water demand (SETI).
-However, in the zones most affected by drought (Fig.
+However, in the ecoregions most affected by drought (Fig.
 3 and Fig.
-5), where water availability is limited, an increase in atmospheric water demand results in a decrease in vegetation water demand.
-The Chilean matorral and the Patagonian steppe are where this is most evident.
-When combined, our findings demonstrate a persistent drying trend in the Chilean matorral, the Patagonian steppe, and the Southern Andean steppe.
-We attribute this trend to a simultaneous decrease in precipitation and an increase in atmospheric evaporative demand, which leads to a decrease in the water demand for vegetation in water-limited zones[@Pascoa2021].
+5), i.e., the Chilean matorral, and the Patagonian steppe , we found that an increase in atmospheric water demand results in a decrease in the water demand of vegetation.
+Together, our findings demonstrate a persistent drying trend in the Chilean Matorral, the Patagonian steppe, and the Southern Andean steppe.
+We attribute this trend to a simultaneous decrease in precipitation and an increase in atmospheric evaporative demand, leading to a decrease in the water demand by vegetation in water-limited areas[@Pascoa2021].
 
@@
 The consequences of the persistent drying trend for ecosystems throughout continental Chile are manifold.
-First, the prolonged hydrological drought, i.e., precipitation deficit, has reduced groundwater storage (SSI, [@Taucare2024]), leading to a steady decline in vegetation productivity (zcNDVI) since 2000 across the Chilean Matorral, Patagonian steppe, and Southern Andean steppe, reaching its lowest level between 2020 and 2022, which could be due to either a decrease in vegetation area, a loss of biomass or browning in forest ecosystems.
-Reports for natural and productive ecosystems [@Nicolai-Shaw2017;@Jiang2020;@Zhou2021] most strongly associate this decline with declines in soil moisture and increases in evapotranspiration.
-Second, the sharp decline in vegetation productivity in the Chilean Matorral and Valdivian temperate forests showed that grasses and shrublands respond to shifts in water supply on higher time scales (12 months), in comparison to savanna and croplands (6 months).
-Also, in the Valdivian temperate forest that has a higher surface with forest trees, soil moisture (SSI) and vegetation water demand (SETI) at higher time scales, 12 and 36 months, respectively, evidence stronger correlations.
-Which is consistent with recent studies showing that progressive, long-term water deficits in central Chile have triggered forest browning and declines in native forest productivity  [e.g., @Miranda2023;@Miranda2020;@Venegas2022].
-While our analysis do not distinguish between  native and planted forests, the latter of which are considered to be more drought tolerant in central and southern Chile [@Carrasco2022], we show that forest area declines more sharply in response to increasing water demand due to rising temperatures (EDDI) than decreasing water supply (e.g., SPI, SSI; refs.
-[@Fajardo2019;@Holz2018]), which may have cascading impacts on multiple facets of forest diversity [@Segovia2021;@Sabatini2022]
+First, the prolonged hydrological drought, i.e., precipitation deficit, has reduced groundwater storage (SSI; ref.
+[@Taucare2024]), leading to a steady decline in vegetation productivity (zcNDVI) since 2000 across the Patagonian steppe, the Southern Andean steppe, and the Chilean Matorral, which reached its lowest level between 2020 and 2022 and could be due to either a decrease in vegetation area, a loss of biomass, or browning in forest ecosystems.
+Recent studies examining  natural and productive ecosystems[@Nicolai-Shaw2017;@Jiang2020;@Zhou2021] have attributed the decline in vegetation productivity with declines in soil moisture and increases in evapotranspiration.
+Second, the sharp decline in vegetation productivity in the Chilean Matorral and Valdivian temperate forest ecoregions showed that grasslands and shrublands respond to shifts in water supply over longer time scales (12 months) than savannas and croplands (6 months).
+Also, in the Valdivian temperate forest ecoregion, which has a large forested area, vegetation productivity responded to soil moisture (SSI) and vegetation water demand (SETI) most strongly at 12 and 36 months, respectively.
+This result is consistent with recent studies showing that progressive, long-term water deficits in central Chile have triggered forest browning and declines in native forest productivity [@Miranda2023;@Miranda2020;@Venegas2022].
+While our analysis do not distinguish between  native and planted forests, the latter of which are considered to be more drought tolerant in central and southern Chile[@Carrasco2022], we show that forest area declines more sharply in response to increasing water demand due to rising temperatures (EDDI) than decreasing water supply (e.g., SPI, SSI; refs.
+[@Fajardo2019;@Holz2018]), which may have cascading impacts on multiple facets of forest diversity[@Segovia2021;@Sabatini2022].
 
 Moreover, the strengthening of the correlation between vegetation productivity and water supply (SPI, SPEI, SSI) or demand (EDDI, SETI) over multiple time scales (up to 36 months) and across land cover types (Fig.
-5) - demonstrates the impacts of climate change on the water balance in Chile.
-Impacts may extend beyond vegetation productivity, as reduced soil moisture in central Chile and the western United States has increased wildfire activity[@Holden2018;@Gonzalez2018], which is a growing concern in Chile and may be further exacerbated by extensive plantations of highly flammable tree species, e.g., Eucalyptus spp.
-and Pinus spp.
-[@Bowman2019].
-Third, we found that the decline in the vegetation productivity of croplands is due to a decrease in the water supply and vegetation water demand to a greater extent than to an increase in atmospheric water demand [@Quiring2010], causing a decline in water availability.
+5) demonstrates the impacts of climate change on the water balance in Chile.
+These impacts may extend beyond vegetation productivity, as reduced soil moisture in central Chile and the western United States has increased wildfire activity[@Holden2018;@Gonzalez2018], which is a growing concern in Chile and may be further exacerbated by extensive plantations of highly flammable tree species, e.g., Eucalyptus spp.
+and Pinus spp.[@Bowman2019].
+Lastly, we found that the decline in the vegetation productivity of croplands is largely due to a decrease in the water supply and vegetation water demand to a greater extent than to an increase in atmospheric water demand[@Quiring2010], causing a decline in water availability.
 This is consistent with evidence that more water-intensive crops have replaced less water-intensive crops in central Chile, leading to an increase in water extraction from rivers or groundwater [@Munoz2020;@Duran2020].
@@
 
-We found evidence that temporal decreases in water supply (SPEI, SSI) and decreases in vegetation water demand (SETI) are driving shifts not only in vegetation productivity but also in land cover across most of continental Chile.
-Shrubland covers were the land cover type most sensitive to changes in the water balance over short and long temporal scales.
-In contrast, forest surfaces are the type less affected by drought.
-Moreover, our results provide evidence that, in addition to shrublands cover, other land cover types have been affected by water deficits, particularly savannas and grasslands.
-Our results therefore suggest that multiple land cover types could be vulnerable to regime shifts towards more drought tolerant land cover types [@Scheffer2001;@Martinez-Vilalta2016], such as forest, whose cover increased non-linearly in response to increasing drought intensity.
-Changes in cropland cover are not a direct consequence of drought (Fig.
-7), but rather an indirect one, likely occupying burned areas and  possibly reflecting the decision of resource-poor farmers to migrate to regions with more abundant water resources or to change economic activity [@AghaKouchak2021;@Hermans2021].
-In contrast, the increase in shrubland cover due to a decrease in savanna cover may be ecological, as shrubs may  be more drought tolerant than other growth forms[@Eldridge2011;@Gotmark2016].
-
-We run each model ten times to account for the uncertainty in its performance.
-In the supplementary material (Fig.
-S12-S17), we show the error bar of r-squared; despite the inherent uncertainty, the analysis reinforces our results regarding how drought impacts land cover type.
-
-Overall, our results show that long-term declines in water supply and demand have induced widespread, multi-dimensional impacts on the vegetation productivity and on the extent of land cover types.
-While prolonged droughts may directly cause shifts to more drought-tolerant land cover types, such as shrublands, they may also influence land cover change through human decision making and activities.
-This study extends current understanding of drought impacts by demonstrating how their multidimensionality emerges over multiple time scales and across land cover types, which can contribute to developing context-specific adaptation strategies for agriculture, biodiversity conservation, and natural resource management.
+We found evidence that temporal decreases in water supply (SPEI, SSI) and decreases in vegetation water demand (SETI) are driving shifts not only in vegetation productivity but also in temporal trends of land cover change across most of continental Chile.
+Despite differences in drought tolerance (e.g., shrublands, grasslands, and savannas), our results provide evidence that  the area of most land cover types dominated by vegetation has been affected by water deficits, albeit to varying degrees (Fig.
+8).
+Additionally, our results suggest that water deficits, to a greater extent than factors associated with human activity, have affected temporal trends in land cover change for most land  cover types.
+Further, we found that temporal changes in cropland cover may not be a direct consequence of drought (Fig.
+7), but rather an indirect one, likely associated with forest fires and possibly the decisions of resource-poor farmers to migrate to regions with more abundant water resources or to change economic activity[@AghaKouchak2021;@Hermans2021].
+The reason for the non-linear increases in forest area in response to burned area across most ecoregions (Fig.
+8) is unclear, as it could be due to forest recovery[@UrrutiaJalabert2018] or the establishment of forest plantations [@Smith2021].
 
@@
 
-This study has evaluated the impact of water supply and demand on vegetation productivity and how land use change influences it, with some important limitations to highlight.
-One of the major limitations of the study is the consideration of secondary information.
-For instance, we used estimates of water supply and demand, such as ERA5L and MODIS, which, despite their improved estimation capacity, suffer from biases and uncertainties [@Gomis2023;@Clelland2024] in different areas or climatic conditions.
-In this study, we compared the ERA5L data with climatic stations (see Table S2) to verify bias and uncertainty, but future studies will need to correct the products for a more precise measurement.
-We used zcNDVI [@Zambrano2018] (MODIS) as a proxy for vegetation productivity, which has proven to be a good estimate of NPP (see Fig.
+Our analysis of the impacts of water supply and demand on vegetation productivity and land cover change has some important limitations that need to be highlighted..
+One of the principal limitations of this study is the use of secondary information.
+For instance, we used estimates of water supply and demand, such as ERA5L and MODIS, which, despite their improved estimation capacity, suffer from biases and uncertainties[@Gomis2023;@Clelland2024] in different areas or climatic conditions.
+In this study, we compared the ERA5L data with climatic stations (see Table S2) to verify bias and uncertainty, but future studies that aim for more focalized analysis will need to improve the precision of these products.
+We used zcNDVI[@Zambrano2018] (MODIS) as a proxy for vegetation productivity, which has proven to be a good estimate of NPP (see Fig.
 S1 and S2), but its quality varies between different types of vegetation.
 
-To define land cover, products were used that estimate the classes using classification models, which are subject to quality errors that must be taken into account [@Stehman2019;@Verburg2011].
-In addition, in our case we used macro classes of land cover, where, for example, the different types of forests were grouped (e.g., monoculture, native forest).
-This approach hinders our ability to understand the effects of drought on the various subclasses within each land use class.
-When it comes to cultivated areas, we cannot distinguish between rainfed and irrigated areas using macro classes.
-However, in this study, we aimed to provide a broad overview at a national scale, with the goal of focusing on a more detailed level in future research.
+A second limitation is that we used products that estimate land cover types using classification models, which are subject to quality errors that must be taken into account[@Stehman2019;@Verburg2011].
+In addition, in our case we used macro classes of land cover, where, for example, the different types of forests (e.g., monoculture, native forest) were pooled into the same land cover type.
+This approach may hinder our ability to understand the effects of drought on the various subclasses within each land cover class.
+In terms of croplands, we could not distinguish between rainfed and irrigated areas using macro classes.
+However, in this study, we aimed to provide a broad overview at a large spatial scale, but acknowledge that using sub-classes of land cover types at finer spatial resolutions may help to better understand underlying mechanisms.
 
-Other studies have analyzed how human decisions impact land use change, such as population increase [@Kleemann2017].
-In our case, we attempted to consider this by incorporating road density, night lights, and burned areas.
-However, human decisions are complex to consider from the point of view of earth observation.
-On the other hand, earth observation tools can analyze land cover change, but each type's use depends on social and economic factors, which are challenging to quantify [@Rindfuss2004;@Jenner2024] and necessitate the integration of social, natural, and geographic information sciences.
+In our analysis of the impacts of drought intensity on temporal trends of land cover change, we integrated proxies for human activity that also may affect land cover change.
+However, attributing land cover change to human activity and decisions is complex when using earth observation tools.
+While earth observation tools can analyze land cover change, whether a land cover type changes likely depends on a multitude of social and economic factors that are challenging to quantify[@Rindfuss2004;@Jenner2024] and necessitate the integration of social, natural, and geographic information sciences.
+
+# Conclusion
+
+Overall, our results show that long-term declines in water supply and demand have consistently induced widespread, multi-dimensional impacts on the vegetation productivity and on the temporal trends of changes in land cover types across a broad range of ecoregions.
+While prolonged droughts may directly cause shifts to more drought-tolerant land cover types, such as shrublands, we also found that areas affected by fires were associated with increases in the area of forests and croplands, highlighting the importance of socio-economic factors in shaping land use change dynamics.
+Our study extends current understanding of drought impacts by demonstrating how their multidimensionality emerges over multiple time scales and across land cover types, which can contribute to developing context-specific adaptation strategies for agriculture, biodiversity conservation, and natural resource management.
 

### 2025-04-08 article/article_scientific_reports/mansucript_11012025.tex
actualización paper

@@
 ecosystems, its potential impacts on land cover change have been largely
-unexplored.
+unexplored \citep{Tollerud2018, Marumbwa2020}.
 
@@
 To derive the drought indices of water supply and demand we used the
-ERA5L dataset and the MODIS product (MOD13A3 \citep{Didan2015b}), with a
-monthly frequency for 2000--2023. Drought indices capture historical
-anomalies of water supply and demand. To quantify each anomaly, the
-common practice is to derive it following a statistical parametric
-method in which it is assumed that the statistical distribution of the
-data is known \citep{Heim2002}. The use of an erroneous statistical
-distribution that does not fit the data is usually the highest source of
-uncertainty \citep{Laimighofer2022}. In the case of Chile, due to its
-high degree of climatic variability, it is difficult to choose a
-statistical distribution that can be used across its entire extent. We
-therefore use a non-parametric method for the calculation of the drought
-indices, following \citep{Farahmand2015}.
+ERA5L dataset, with a monthly frequency for 2000--2023. Drought indices
+capture historical anomalies of water supply and demand. To quantify
+each anomaly, the common practice is to derive it following a
+statistical parametric method in which it is assumed that the
+statistical distribution of the data is known \citep{Heim2002}. The use
+of an erroneous statistical distribution that does not fit the data is
+usually the highest source of uncertainty \citep{Laimighofer2022}. In
+the case of Chile, due to its high degree of climatic variability, it is
+difficult to choose a statistical distribution that can be used across
+its entire extent. We therefore use a non-parametric method for the
+calculation of the drought indices, following \citep{Farahmand2015}.
 
@@
 
-We also used the MODIS product to calculate vegetation productivity, and
-calculated anomalies of cumulative NDVI using zcNDVI
-\citep{Zambrano2018}, which was derived from the monthly time series of
-NDVI, with Equations 2 and 4. For vegetation productivity, we selected
-the time scale that best correlates with annual net primary productivity
-(NPP) across continental Chile. For this purpose, we calculated zcNDVI
-for time scales of 1, 3, 6, and 12 months (from December) and compared
-it with the annual NPP. We obtained NPP from MOD17A3HGF90. We chose to
-use six months because the \(R^2\) between zcNDVI and NPP reaches its
-highest value at six months, obtaining an \(R^2\) of 0.31 for forest and
-0.72 for shrubland (Supplementary Information Section S2). We
-subsequently used zcNDVI with a time scale of 6 months and calculated it
-at a monthly frequency for 2000--2023.
+We also used the MODIS product (MOD13A3\citep{Didan2015}) to calculate
+vegetation productivity, and calculated anomalies of cumulative NDVI
+using zcNDVI \citep{Zambrano2018}, which was derived from the monthly
+time series of NDVI, with Equations 2 and 4. For vegetation
+productivity, we selected the time scale that best correlates with
+annual net primary productivity (NPP) across continental Chile. For this
+purpose, we calculated zcNDVI for time scales of 1, 3, 6, and 12 months
+(from December) and compared it with the annual NPP. We obtained NPP
+from MOD17A3HGF90. We chose to use six months because the \(R^2\)
+between zcNDVI and NPP reaches its highest value at six months,
+obtaining an \(R^2\) of 0.31 for forest and 0.72 for shrubland
+(Supplementary Information Section S2). We subsequently used zcNDVI with
+a time scale of 6 months and calculated it at a monthly frequency for
+2000--2023.
 
@@
 SSI), as well as an increase in atmospheric and vegetation water demand
-(EDDI, SETI). However, in the Magellanic subpolar forests, there is no
-significant increase or decrease in SETI across time scales. Also, we
-found a significant decreasing trend in water supply (SPI, SPEI, and
-SSI) across the Southern Andean steppe, Chilean Matorral
-\citep{Boisier2018, Sarricolea2019}, Valdivian temperate forests, and
-Patagonian steppe, accompanied by an increase in atmospheric water
-demand (EDDI). Our results indicate that the trend of water supply and
-atmospheric demand tends to decrease or increase more strongly over
-longer time scales---a trend that is consistent with the progressive
-intensification of drought severity across much of Chile, as observed in
-other regions facing long-term droughts \citep{Rashid2019, Miro2023}.
-Simultaneously, we observed a divergent trend between the EDDI and SETI
-in various ecoregions. In the majority of ecoregions, a rise in
-atmospheric water demand (EDDI) typically leads to a rise in vegetation
-water demand (SETI). However, in the zones most affected by drought
-(Fig. 3 and Fig. 5), where water availability is limited, an increase in
-atmospheric water demand results in a decrease in vegetation water
-demand. The Chilean matorral and the Patagonian steppe are where this is
-most evident. When combined, our findings demonstrate a persistent
-drying trend in the Chilean matorral, the Patagonian steppe, and the
-Southern Andean steppe. We attribute this trend to a simultaneous
-decrease in precipitation and an increase in atmospheric evaporative
-demand, which leads to a decrease in the water demand for vegetation in
-water-limited zones\citep{Pascoa2021}.
+(EDDI, SETI). However, in the Magellanic subpolar forests, we found no
+evidence of either a significant increase or decrease in SETI across
+time scales. Also, we found a significant decreasing trend in water
+supply (SPI, SPEI, and SSI) across the Southern Andean steppe, Chilean
+Matorral\citep{Boisier2018, Sarricolea2019}, Valdivian temperate
+forests, and Patagonian steppe, accompanied by an increase in
+atmospheric water demand (EDDI). Our results indicate that temporal
+trends of water supply and atmospheric demand tend to decrease or
+increase more strongly over longer time scales,a trend that is
+consistent with the progressive intensification of drought severity
+across much of Chile, and that has been observed in other regions facing
+long-term droughts\citep{Rashid2019, Miro2023}. Simultaneously, we
+observed a divergent trend between EDDI and SETI. In the majority of
+ecoregions, a rise in atmospheric water demand (EDDI) typically leads to
+a rise in vegetation water demand (SETI). However, in the ecoregions
+most affected by drought (Fig. 3 and Fig. 5), i.e., the Chilean
+matorral, and the Patagonian steppe , we found that an increase in
+atmospheric water demand results in a decrease in the water demand of
+vegetation. Together, our findings demonstrate a persistent drying trend
+in the Chilean Matorral, the Patagonian steppe, and the Southern Andean
+steppe. We attribute this trend to a simultaneous decrease in
+precipitation and an increase in atmospheric evaporative demand, leading
+to a decrease in the water demand by vegetation in water-limited
+areas\citep{Pascoa2021}.
 
@@
 hydrological drought, i.e., precipitation deficit, has reduced
-groundwater storage (SSI, \citep{Taucare2024}), leading to a steady
+groundwater storage (SSI; ref. \citep{Taucare2024}), leading to a steady
 decline in vegetation productivity (zcNDVI) since 2000 across the
-Chilean Matorral, Patagonian steppe, and Southern Andean steppe,
-reaching its lowest level between 2020 and 2022, which could be due to
-either a decrease in vegetation area, a loss of biomass or browning in
-forest ecosystems. Reports for natural and productive ecosystems
-\citep{Nicolai-Shaw2017, Jiang2020, Zhou2021} most strongly associate
-this decline with declines in soil moisture and increases in
-evapotranspiration. Second, the sharp decline in vegetation productivity
-in the Chilean Matorral and Valdivian temperate forests showed that
-grasses and shrublands respond to shifts in water supply on higher time
-scales (12 months), in comparison to savanna and croplands (6 months).
-Also, in the Valdivian temperate forest that has a higher surface with
-forest trees, soil moisture (SSI) and vegetation water demand (SETI) at
-higher time scales, 12 and 36 months, respectively, evidence stronger
-correlations. Which is consistent with recent studies showing that
-progressive, long-term water deficits in central Chile have triggered
-forest browning and declines in native forest productivity
-\citep[e.g.,][]{Miranda2023, Miranda2020, Venegas2022}. While our
-analysis do not distinguish between native and planted forests, the
+Patagonian steppe, the Southern Andean steppe, and the Chilean Matorral,
+which reached its lowest level between 2020 and 2022 and could be due to
+either a decrease in vegetation area, a loss of biomass, or browning in
+forest ecosystems. Recent studies examining natural and productive
+ecosystems\citep{Nicolai-Shaw2017, Jiang2020, Zhou2021} have attributed
+the decline in vegetation productivity with declines in soil moisture
+and increases in evapotranspiration. Second, the sharp decline in
+vegetation productivity in the Chilean Matorral and Valdivian temperate
+forest ecoregions showed that grasslands and shrublands respond to
+shifts in water supply over longer time scales (12 months) than savannas
+and croplands (6 months). Also, in the Valdivian temperate forest
+ecoregion, which has a large forested area, vegetation productivity
+responded to soil moisture (SSI) and vegetation water demand (SETI) most
+strongly at 12 and 36 months, respectively. This result is consistent
+with recent studies showing that progressive, long-term water deficits
+in central Chile have triggered forest browning and declines in native
+forest productivity \citep{Miranda2023, Miranda2020, Venegas2022}. While
+our analysis do not distinguish between native and planted forests, the
 latter of which are considered to be more drought tolerant in central
-and southern Chile \citep{Carrasco2022}, we show that forest area
+and southern Chile\citep{Carrasco2022}, we show that forest area
 declines more sharply in response to increasing water demand due to
@@
 refs. \citep{Fajardo2019, Holz2018}), which may have cascading impacts
-on multiple facets of forest diversity \citep{Segovia2021, Sabatini2022}
+on multiple facets of forest diversity\citep{Segovia2021, Sabatini2022}.
 
@@
 over multiple time scales (up to 36 months) and across land cover types
-(Fig. 5) - demonstrates the impacts of climate change on the water
-balance in Chile. Impacts may extend beyond vegetation productivity, as
+(Fig. 5) demonstrates the impacts of climate change on the water balance
+in Chile. These impacts may extend beyond vegetation productivity, as
 reduced soil moisture in central Chile and the western United States has
@@
 plantations of highly flammable tree species, e.g., Eucalyptus spp. and
-Pinus spp. \citep{Bowman2019}. Third, we found that the decline in the
-vegetation productivity of croplands is due to a decrease in the water
-supply and vegetation water demand to a greater extent than to an
-increase in atmospheric water demand \citep{Quiring2010}, causing a
+Pinus spp.\citep{Bowman2019}. Lastly, we found that the decline in the
+vegetation productivity of croplands is largely due to a decrease in the
+water supply and vegetation water demand to a greater extent than to an
+increase in atmospheric water demand\citep{Quiring2010}, causing a
 decline in water availability. This is consistent with evidence that
@@
 and decreases in vegetation water demand (SETI) are driving shifts not
-only in vegetation productivity but also in land cover across most of
-continental Chile. Shrubland covers were the land cover type most
-sensitive to changes in the water balance over short and long temporal
-scales. In contrast, forest surfaces are the type less affected by
-drought. Moreover, our results provide evidence that, in addition to
-shrublands cover, other land cover types have been affected by water
-deficits, particularly savannas and grasslands. Our results therefore
-suggest that multiple land cover types could be vulnerable to regime
-shifts towards more drought tolerant land cover types
-\citep{Scheffer2001, Martinez-Vilalta2016}, such as forest, whose cover
-increased non-linearly in response to increasing drought intensity.
-Changes in cropland cover are not a direct consequence of drought (Fig.
-7), but rather an indirect one, likely occupying burned areas and
-possibly reflecting the decision of resource-poor farmers to migrate to
-regions with more abundant water resources or to change economic
-activity \citep{AghaKouchak2021, Hermans2021}. In contrast, the increase
-in shrubland cover due to a decrease in savanna cover may be ecological,
-as shrubs may be more drought tolerant than other growth
-forms\citep{Eldridge2011, Gotmark2016}.
-
-We run each model ten times to account for the uncertainty in its
-performance. In the supplementary material (Fig. S12-S17), we show the
-error bar of r-squared; despite the inherent uncertainty, the analysis
-reinforces our results regarding how drought impacts land cover type.
-
-Overall, our results show that long-term declines in water supply and
-demand have induced widespread, multi-dimensional impacts on the
-vegetation productivity and on the extent of land cover types. While
-prolonged droughts may directly cause shifts to more drought-tolerant
-land cover types, such as shrublands, they may also influence land cover
-change through human decision making and activities. This study extends
-current understanding of drought impacts by demonstrating how their
-multidimensionality emerges over multiple time scales and across land
-cover types, which can contribute to developing context-specific
-adaptation strategies for agriculture, biodiversity conservation, and
-natural resource management.
+only in vegetation productivity but also in temporal trends of land
+cover change across most of continental Chile. Despite differences in
+drought tolerance (e.g., shrublands, grasslands, and savannas), our
+results provide evidence that the area of most land cover types
+dominated by vegetation has been affected by water deficits, albeit to
+varying degrees (Fig. 8). Additionally, our results suggest that water
+deficits, to a greater extent than factors associated with human
+activity, have affected temporal trends in land cover change for most
+land cover types. Further, we found that temporal changes in cropland
+cover may not be a direct consequence of drought (Fig. 7), but rather an
+indirect one, likely associated with forest fires and possibly the
+decisions of resource-poor farmers to migrate to regions with more
+abundant water resources or to change economic
+activity\citep{AghaKouchak2021, Hermans2021}. The reason for the
+non-linear increases in forest area in response to burned area across
+most ecoregions (Fig. 8) is unclear, as it could be due to forest
+recovery\citep{UrrutiaJalabert2018} or the establishment of forest
+plantations \citep{Smith2021}.
 
@@
 
-This study has evaluated the impact of water supply and demand on
-vegetation productivity and how land use change influences it, with some
-important limitations to highlight. One of the major limitations of the
-study is the consideration of secondary information. For instance, we
-used estimates of water supply and demand, such as ERA5L and MODIS,
-which, despite their improved estimation capacity, suffer from biases
-and uncertainties \citep{Gomis2023, Clelland2024} in different areas or
+Our analysis of the impacts of water supply and demand on vegetation
+productivity and land cover change has some important limitations that
+need to be highlighted.. One of the principal limitations of this study
+is the use of secondary information. For instance, we used estimates of
+water supply and demand, such as ERA5L and MODIS, which, despite their
+improved estimation capacity, suffer from biases and
+uncertainties\citep{Gomis2023, Clelland2024} in different areas or
 climatic conditions. In this study, we compared the ERA5L data with
 climatic stations (see Table S2) to verify bias and uncertainty, but
-future studies will need to correct the products for a more precise
-measurement. We used zcNDVI \citep{Zambrano2018} (MODIS) as a proxy for
-vegetation productivity, which has proven to be a good estimate of NPP
-(see Fig. S1 and S2), but its quality varies between different types of
-vegetation.
-
-To define land cover, products were used that estimate the classes using
-classification models, which are subject to quality errors that must be
-taken into account \citep{Stehman2019, Verburg2011}. In addition, in our
-case we used macro classes of land cover, where, for example, the
-different types of forests were grouped (e.g., monoculture, native
-forest). This approach hinders our ability to understand the effects of
-drought on the various subclasses within each land use class. When it
-comes to cultivated areas, we cannot distinguish between rainfed and
-irrigated areas using macro classes. However, in this study, we aimed to
-provide a broad overview at a national scale, with the goal of focusing
-on a more detailed level in future research.
-
-Other studies have analyzed how human decisions impact land use change,
-such as population increase \citep{Kleemann2017}. In our case, we
-attempted to consider this by incorporating road density, night lights,
-and burned areas. However, human decisions are complex to consider from
-the point of view of earth observation. On the other hand, earth
-observation tools can analyze land cover change, but each type's use
-depends on social and economic factors, which are challenging to
-quantify \citep{Rindfuss2004, Jenner2024} and necessitate the
-integration of social, natural, and geographic information sciences.
+future studies that aim for more focalized analysis will need to improve
+the precision of these products. We used zcNDVI\citep{Zambrano2018}
+(MODIS) as a proxy for vegetation productivity, which has proven to be a
+good estimate of NPP (see Fig. S1 and S2), but its quality varies
+between different types of vegetation.
+
+A second limitation is that we used products that estimate land cover
+types using classification models, which are subject to quality errors
+that must be taken into account\citep{Stehman2019, Verburg2011}. In
+addition, in our case we used macro classes of land cover, where, for
+example, the different types of forests (e.g., monoculture, native
+forest) were pooled into the same land cover type. This approach may
+hinder our ability to understand the effects of drought on the various
+subclasses within each land cover class. In terms of croplands, we could
+not distinguish between rainfed and irrigated areas using macro classes.
+However, in this study, we aimed to provide a broad overview at a large
+spatial scale, but acknowledge that using sub-classes of land cover
+types at finer spatial resolutions may help to better understand
+underlying mechanisms.
+
+In our analysis of the impacts of drought intensity on temporal trends
+of land cover change, we integrated proxies for human activity that also
+may affect land cover change. However, attributing land cover change to
+human activity and decisions is complex when using earth observation
+tools. While earth observation tools can analyze land cover change,
+whether a land cover type changes likely depends on a multitude of
+social and economic factors that are challenging to
+quantify\citep{Rindfuss2004, Jenner2024} and necessitate the integration
+of social, natural, and geographic information sciences.
+
+\section{Conclusion}\label{conclusion}
+
+Overall, our results show that long-term declines in water supply and
+demand have consistently induced widespread, multi-dimensional impacts
+on the vegetation productivity and on the temporal trends of changes in
+land cover types across a broad range of ecoregions. While prolonged
+droughts may directly cause shifts to more drought-tolerant land cover
+types, such as shrublands, we also found that areas affected by fires
+were associated with increases in the area of forests and croplands,
+highlighting the importance of socio-economic factors in shaping land
+use change dynamics. Our study extends current understanding of drought
+impacts by demonstrating how their multidimensionality emerges over
+multiple time scales and across land cover types, which can contribute
+to developing context-specific adaptation strategies for agriculture,
+biodiversity conservation, and natural resource management.
 
